---
title: "00 - Run the sequence-alignment steps"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output: html_document
params:
  steps: [2, 3]         # count matrices and the read-trimming record. "all" also runs step 01 (HISAT2 + StringTie)
  stop_on_fail: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs the numbered scripts in `01_code/` in order, each in its own R process, with an HTML
report and a log per step in `03_analyses/knit_html/` (git-ignored) and a `run_log.csv`.

| step | script | what it does |
|---|---|---|
| 01 | `01_hisat_stringtie.Rmd` | HISAT2 alignment and StringTie quantification of the trimmed Tag-seq reads; writes `03_analyses/hisat/` and its `sample_list.txt` |
| 02 | `02_prepDE.Rmd` | gene and transcript count matrices (`03_analyses/prepDE/`), read by `06_differential-expression` |
| 03 | `03_read_trimming.Rmd` | how the trimmed reads were made: read retention from the committed FastQC tables (`03_analyses/read_trimming/`); the recipe check runs only with its own `online: true` |

By default steps 02 and 03 run. Step 01 needs the trimmed reads, the genome and HISAT2 /
StringTie, none of which are in this repository (see `../README.md`); with its outputs
present, step 02 counts from the StringTie GTFs, and without them it rebuilds the gene matrix
from the committed transcript matrix. Set `steps: "all"` on a machine that has the inputs.

```{r run}
suppressPackageStartupMessages(library(here))
source(file.path(here::here(), "..", "tools", "run_steps.R"))
html_dir <- here::here("03_analyses", "knit_html")
run_log <- run_steps(here::here("01_code"), html_dir, steps = params$steps,
                     stop_on_fail = isTRUE(params$stop_on_fail))
finish_run(run_log, html_dir, "Sequence-alignment")
```
