---
title: "00 - Run the sequence-alignment steps"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output: html_document
params:
  online: false         # true: also steps 04 to 06 and step 08's alignment (they download the Iso-Seq transcriptome, the genome and the reads, and run the aligners)
  minimap2: "minimap2"  # step 04
  bin: ""               # steps 06 and 08: folder with hisat2, hisat2-build, samtools, stringtie and featureCounts ("" = PATH)
  threads: 4
  steps: "default"      # "default" (02, 03, 07, 08; with online: true, 02 to 08), "all" (01 to 08), or step numbers, e.g. [7]
  stop_on_fail: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs the numbered scripts in `01_code/` in order, each in its own R process, with an HTML
report and a log per step in `03_analyses/knit_html/` (git-ignored) and a `run_log.csv`.

| step | script | what it does |
|---|---|---|
| 01 | `01_hisat_stringtie.Rmd` | HISAT2 alignment and StringTie quantification of the trimmed Tag-seq reads (HPC); writes `03_analyses/hisat/` and its `sample_list.txt` |
| 02 | `02_prepDE.Rmd` | gene and transcript count matrices of the HPC alignment (`03_analyses/prepDE/`): the previous record, read by `04_iso-seq-transcriptome` step 04 for comparison |
| 03 | `03_read_trimming.Rmd` | how the trimmed reads were made: read retention from the committed FastQC tables (`03_analyses/read_trimming/`); the recipe check runs only with the script's own `online: true` (knit by hand; it needs Perl, the FASTX-Toolkit and cutadapt) |
| 04 | `04_isoform_gene_map.Rmd` | each Iso-Seq isoform's genome gene (`03_analyses/isoform-gene-map/`), from a minimap2 alignment of the transcriptome to the genome; online; about 1.5 hours and 9 GB of memory |
| 05 | `05_augmented_annotation.Rmd` | the RefSeq annotation with Iso-Seq-extended 3' ends (`ext3`) and with the Iso-Seq models added (`full`) (`03_analyses/augmented-annotation/`); online (step 04's alignment and the genome); a few minutes |
| 06 | `06_genome_recount.Rmd` | the reads realigned (HISAT2) and counted on the RefSeq, `ext3` and `full` annotations with StringTie + prepDE and featureCounts (`03_analyses/genome-recount/`); online; about 15 minutes for the index plus 4 minutes per library |
| 07 | `07_count_matrix_of_record.Rmd` | the count matrix of record (`03_analyses/featurecounts/`): step 06's featureCounts counts on `ext3`, with prepDE's row names; read by `06_differential-expression` |
| 08 | `08_mitogenome_counts.Rmd` | the mitochondrial genes counted on the mitochondrial genome alone (`03_analyses/mitogenome/`), read by `06_differential-expression` step 13; aligns with `online: true`, otherwise summarises the committed counts |

By default steps 02, 03, 07 and 08 run, from committed files. Step 01 needs the trimmed reads,
the genome and HISAT2 / StringTie on the HPC (see `../README.md`); with its outputs present,
step 02 counts from the StringTie GTFs, and without them it rebuilds the gene matrix from the
committed transcript matrix. Steps 04 to 06 download the Iso-Seq transcriptome (owl), the genome
and its annotation (NCBI) and the trimmed reads (gannet); set `online: true`, with `minimap2`
and `bin` pointing at the binaries (minimap2 2.31, HISAT2 2.2.1, StringTie 2.2.1, Subread
2.1.1, samtools 1.24 were used, from bioconda), to rebuild them and step 08's counts. Their
large intermediate files are git-ignored and reused when present. `steps: "all"` runs 01 to 08
on a machine that has every input.

```{r run}
suppressPackageStartupMessages(library(here))
source(file.path(here::here(), "..", "tools", "run_steps.R"))
html_dir <- here::here("03_analyses", "knit_html")
steps <- if (!identical(params$steps, "default")) params$steps else if (isTRUE(params$online)) 2:8 else c(2, 3, 7, 8)
tool_params <- list("04" = list(minimap2 = params$minimap2, threads = params$threads),
                    "05" = list(threads = params$threads),
                    "06" = list(bin = params$bin, threads = params$threads),
                    "08" = list(online = isTRUE(params$online), bin = params$bin, threads = params$threads))
run_log <- run_steps(here::here("01_code"), html_dir, steps = steps, step_params = tool_params,
                     stop_on_fail = isTRUE(params$stop_on_fail))
finish_run(run_log, html_dir, "Sequence-alignment")
```
