---
title: "06 - Realign the Tag-seq reads and count them on the three annotations (option B)"
output: html_document
params:
  bin: ""                  # folder holding hisat2 (2.2.1), samtools, stringtie (2.2.1) and featureCounts; "" = PATH
  threads: 4
  max_libraries: .Inf      # process at most this many new libraries per knit, then stop (for batches)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

The previous record's counts (steps 01 and 02) came from HISAT2 alignments made on an HPC
server and not kept, so option B realigns the reads here and counts them on each of step 05's
annotations. Only the annotation should differ between the counts compared in
`04_iso-seq-transcriptome` step 04, so the alignment is shared and the RefSeq annotation is
counted again as the control.

- **Reads:** the 131 trimmed, lane-merged files the record aligned
  (`*_L099_R1_cmb.trim.fastq.gz`, gannet `.../byssus-exp-analysis/data/raw-trimmed/`), each
  downloaded, processed and deleted.
- **Alignment:** HISAT2 2.2.1 (Kim et al. 2019, *Nature Biotechnology* 37:907) with the
  record's settings (defaults and `--dta`). The record's index was built with the RefSeq splice
  sites and exons (`hisat2-build --ss --exon`), which needs far more memory than the 15 GB
  machine used here (HISAT2's manual quotes about 160 GB for a human-sized genome), so the
  index here is the plain genome and the same splice sites are given at alignment time
  (`--known-splicesite-infile`, from NCBI's GTF of the same annotation with HISAT2's
  `hisat2_extract_splice_sites.py`, as the record made them). The alignments are close but not
  identical to the record's; the control count measures how close.
- **StringTie + prepDE:** StringTie 2.2.1 (Pertea et al. 2015, *Nature Biotechnology* 33:290)
  `-e -G <annotation>` as in the record, reduced at once to prepDE's gene counts (read length
  75) by `_prepde_sample.R`, which is checked below against step 02's port (`_prepde.R`) on one library.
  Reference transcripts StringTie files under its own `STRG.N` gene IDs, which change between
  libraries, are named by their transcript (see `_prepde_sample.R`).
- **featureCounts:** Subread's featureCounts (Liao, Smyth and Shi 2014, *Bioinformatics*
  30:923) on each annotation's exons (`.saf`), sense strand (`-s 1`, as 3' Tag-seq reads are),
  one count per uniquely aligned read; reads aligned to several places, or overlapping more
  than one gene, are left out (the defaults). This changes the counting method as well as the
  annotation, so `04_iso-seq-transcriptome` step 04 compares each counter with its own RefSeq control.

Each library is processed by `_recount_library.sh` (alignment, the six counts, the alignment
deleted at the end), into `03_analyses/genome-recount/libraries/<library>/` (git-ignored).
`max_libraries` lets the step run in batches; it combines the counts once all 131 are done.

Writes, in `03_analyses/genome-recount/`:

- `<stringtie|featurecounts>_<refseq|ext3|full>_gene_counts.csv.gz`: genes x libraries
  (column names as in `06`'s count matrix, `T001F`);
- `mapping_summary.csv`: per library, reads, HISAT2's alignment (unique, multiple, overall
  rate) beside the record's rate, featureCounts's assigned reads and StringTie's total count on
  each annotation;
- `checks.csv`: the per-library counter against step 02's port, and the control library against
  the record's StringTie output for the same library;
- `RUN_provenance.txt`: tool versions and settings, input MD5s.

and, git-ignored: `hisat2_index/` (about 2 GB, 15 minutes), `refseq_splice_sites.txt`,
`hisat2-build.log` and `libraries/`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "pipeline_checks.R"))  # provenance_lines()
suppressPackageStartupMessages(library(data.table))
tool <- function(x) if (nzchar(params$bin)) file.path(params$bin, x) else unname(Sys.which(x))
for (x in c("hisat2", "hisat2-build", "samtools", "stringtie", "featureCounts", "hisat2_extract_splice_sites.py", "python3"))
  if (!nzchar(tool(x)) || !file.exists(tool(x))) stop(x, " not found; set params$bin")
rscript <- file.path(R.home("bin"), "Rscript")
helper  <- here::here("01_code", "_recount_library.sh")
ann_files <- file.path(ann_dir, paste0("annotation_", rep(c("refseq", "ext3", "full"), each = 2), c(".gff", ".saf")))
stopifnot(all(file.exists(ann_files)))                      # step 05 first
idx_prefix <- file.path(recount_dir, "hisat2_index", "PNRI_Mtr1.1.1")
ss_file    <- file.path(recount_dir, "refseq_splice_sites.txt")
gtf_gz     <- file.path(dat, paste0(ASSEMBLY, "_genomic.gtf.gz"))
lib_root   <- file.path(recount_dir, "libraries")
dir.create(lib_root, showWarnings = FALSE)
```

# Index and splice sites

```{r index}
if (!file.exists(paste0(idx_prefix, ".1.ht2"))) {
  dir.create(dirname(idx_prefix), showWarnings = FALSE)
  fa <- tempfile(fileext = ".fa")
  stopifnot(system(paste("gzip -dc", shQuote(genome_fa), ">", shQuote(fa))) == 0)
  stopifnot(system2(tool("hisat2-build"), c("-p", params$threads, shQuote(fa), shQuote(idx_prefix)),
                    stdout = file.path(recount_dir, "hisat2-build.log"), stderr = file.path(recount_dir, "hisat2-build.log")) == 0)
  unlink(fa)
}
if (!file.exists(ss_file)) {
  if (!file.exists(gtf_gz)) download.file(paste0(NCBI_DIR, basename(gtf_gz)), gtf_gz, mode = "wb")
  ncbi_md5 <- fread(cmd = paste("curl -sS", shQuote(paste0(NCBI_DIR, "md5checksums.txt"))), header = FALSE,
                    col.names = c("md5", "file"))[, file := basename(file)]
  stopifnot(unname(tools::md5sum(gtf_gz)) == ncbi_md5[file == basename(gtf_gz), md5])
  stopifnot(system(paste("gzip -dc", shQuote(gtf_gz), "|", shQuote(tool("python3")),
                         shQuote(tool("hisat2_extract_splice_sites.py")), "- >", shQuote(ss_file))) == 0)
}
cat(length(readLines(ss_file)), "RefSeq splice sites\n")
```

# Libraries

```{r libraries}
listing <- readLines(url(READS_URL), warn = FALSE)
files <- unique(regmatches(listing, regexpr("[A-H][0-9]+-T[0-9]+[A-Z]*_S[0-9]+_L099_R1_cmb\\.trim\\.fastq\\.gz", listing)))
lib_of <- function(f) sub("^[A-H][0-9]+-", "", sub("_S[0-9]+_L099.*$", "", f))
stopifnot(length(files) == 131, !anyDuplicated(lib_of(files)))
done <- function() sum(file.exists(file.path(lib_root, lib_of(files), "done")))
n_new <- 0
for (f in files) {
  out <- file.path(lib_root, lib_of(f))
  if (file.exists(file.path(out, "done"))) next
  if (n_new >= params$max_libraries) break
  unlink(out, recursive = TRUE)                             # a library a stopped run left half done
  reads <- tempfile(fileext = ".fastq.gz")
  ok <- FALSE
  for (i in 1:4) { if (download.file(paste0(READS_URL, f), reads, mode = "wb", quiet = TRUE) == 0) { ok <- TRUE; break }; Sys.sleep(2^i) }
  if (!ok) stop("download failed: ", f)
  rc <- system2("bash", c(shQuote(helper), shQuote(reads), shQuote(out), shQuote(idx_prefix), shQuote(ss_file),
                          shQuote(ann_dir), shQuote(dirname(tool("hisat2"))), shQuote(rscript), params$threads))
  unlink(reads)
  if (rc != 0) stop("_recount_library.sh failed for ", f)
  n_new <- n_new + 1
}
cat(n_new, "libraries processed in this run;", done(), "of 131 done\n")
all_done <- done() == length(files)
```

# Count matrices

```{r combine, eval = all_done}
libs <- sort(lib_of(files))
read_counts <- function(counter, a) {
  long <- rbindlist(lapply(libs, function(l) {
    d <- fread(file.path(lib_root, l, paste0(counter, "_", a, ".tsv.gz")), sep = "\t", quote = "", header = FALSE,
               col.names = c("gene", "count"), colClasses = c("character", "numeric"))   # names hold "|": say the separator
    d[, library := l]
  }))
  stopifnot(!anyDuplicated(long[, .(gene, library)]))
  w <- dcast(long, gene ~ library, value.var = "count", fill = 0)
  setcolorder(w, c("gene", libs))
  w
}
mats <- list()
for (counter in c("stringtie", "featurecounts")) for (a in c("refseq", "ext3", "full")) {
  m <- read_counts(counter, a)
  fwrite(m, file.path(recount_dir, paste0(counter, "_", a, "_gene_counts.csv.gz")))
  mats[[paste(counter, a)]] <- m
}

## mapping summary
hs <- rbindlist(lapply(libs, function(l) {
  x <- readLines(file.path(lib_root, l, "hisat2.log"))
  num <- function(p) as.numeric(sub("^\\s*([0-9]+).*$", "\\1", grep(p, x, value = TRUE)[1]))
  data.table(library = l, reads = num("reads; of these"), aligned_once = num("aligned exactly 1 time"),
             aligned_multi = num("aligned >1 times"),
             overall_rate = as.numeric(sub("%.*$", "", grep("overall alignment rate", x, value = TRUE))))
}))
rec <- fread(here::here("03_analyses", "hisat", "multiqc_data", "multiqc_bowtie2.txt"))
rec[, library := sub("^[A-H][0-9]+-", "", sub("_S[0-9]+_hisat$", "", Sample))]
hs <- merge(hs, rec[, .(library, record_reads = total_reads, record_rate = overall_alignment_rate)], by = "library", all.x = TRUE)
for (a in c("refseq", "ext3", "full")) {
  fc <- rbindlist(lapply(libs, function(l) {
    s <- fread(file.path(lib_root, l, paste0("featurecounts_", a, ".summary")))
    data.table(library = l, v = s[Status == "Assigned"][[2]])
  }))
  hs[fc, (paste0("featurecounts_assigned_", a)) := i.v, on = "library"]
  st <- mats[[paste("stringtie", a)]]
  hs[data.table(library = libs, v = colSums(st[, ..libs])), (paste0("stringtie_total_", a)) := i.v, on = "library"]
}
stopifnot(nrow(hs) == 131, all(hs$reads == hs$record_reads))  # the same reads as the record
fwrite(hs, file.path(recount_dir, "mapping_summary.csv"))
knitr::kable(hs[, .(median_rate = median(overall_rate), median_record_rate = median(record_rate),
                    median_rate_minus_record = median(overall_rate - record_rate))])
```

# Checks

On one library (T014F, whose StringTie table the record kept), the alignment is repeated and
counted on the RefSeq annotation: the per-library counter against step 02's port (prepDE's
exact names), and the control against the record's StringTie output for the same library
(transcript by transcript) and the record's count matrix.

```{r checks, eval = all_done}
check_lib <- files[lib_of(files) == "T014F"]
tmp <- tempfile(); dir.create(tmp)
download.file(paste0(READS_URL, check_lib), file.path(tmp, "reads.fq.gz"), mode = "wb", quiet = TRUE)
stopifnot(system(sprintf("%s -p %s --dta --known-splicesite-infile %s -x %s -U %s 2> /dev/null | %s sort -@ 2 -m 1G -o %s - && %s index %s",
                         tool("hisat2"), params$threads, shQuote(ss_file), shQuote(idx_prefix), shQuote(file.path(tmp, "reads.fq.gz")),
                         tool("samtools"), shQuote(file.path(tmp, "aln.bam")), tool("samtools"), shQuote(file.path(tmp, "aln.bam")))) == 0)
stopifnot(system2(tool("stringtie"), c("-e", "-p", params$threads, "-G", shQuote(file.path(ann_dir, "annotation_refseq.gff")),
                                       "-o", shQuote(file.path(tmp, "st.gtf")), shQuote(file.path(tmp, "aln.bam")))) == 0)
source(here::here("01_code", "_prepde_sample.R"))
source(here::here("01_code", "_prepde.R"))
fast <- prepde_sample(file.path(tmp, "st.gtf"), stable_strg = FALSE)
writeLines(paste("S", file.path(tmp, "st.gtf")), file.path(tmp, "list.txt"))
prepde(file.path(tmp, "list.txt"), gene_out = file.path(tmp, "g.csv"), transcript_out = file.path(tmp, "t.csv"))
port <- fread(file.path(tmp, "g.csv"), sep = ",")
fp <- merge(as.data.table(fast), port, by.x = "gene", by.y = "gene_id")
## the control against the record, transcript by transcript (the record's t_data.ctab is T014F's)
ctab_rec <- fread(here::here("03_analyses", "hisat", "t_data.ctab"))
st <- fread(cmd = sprintf("grep -v '^#' %s | awk -F'\\t' '$3==\"transcript\"' | cut -f9", shQuote(file.path(tmp, "st.gtf"))),
            sep = "\t", header = FALSE, col.names = "attr")
st[, `:=`(t_name = sub('.*transcript_id "([^"]+)".*', "\\1", attr), cov = as.numeric(sub('.*cov "([^"]+)".*', "\\1", attr)))]
tc <- merge(ctab_rec[, .(t_name, cov_record = cov)], st[, .(t_name, cov)], by = "t_name")
## and gene by gene against the record's count matrix (both in prepDE's names)
rec_csv <- here::here("03_analyses", "prepDE", "gene_count_matrix.csv")
gm <- fread(rec_csv, sep = ",", select = c("gene_id", grep("T014F", names(fread(rec_csv, sep = ",", nrows = 0)), value = TRUE)))
setnames(gm, c("gene", "record"))
gg <- merge(gm, as.data.table(fast), by = "gene", all = TRUE)
gg[is.na(record), record := 0][is.na(count), count := 0]
g10 <- gg[record >= 10 | count >= 10]
checks <- data.table(
  check = c("per-library counter = 05 port (genes, all counts equal)",
            "control vs record, transcripts compared", "control vs record, transcripts with identical coverage",
            "control vs record, Spearman of transcript coverage",
            "control vs record, total gene count (control / record)",
            "control vs record, genes with >= 10 counts in either", "control vs record, of those within 10%",
            "control vs record, Spearman of gene counts (genes >= 10)",
            "genes named only by a StringTie STRG ID in the record (not matched)"),
  value = c(sprintf("%d genes, %s", nrow(fp), nrow(fp) == nrow(port) && all(fp$count == fp$S)),
            nrow(tc), sum(abs(tc$cov - tc$cov_record) < 1e-6), round(cor(tc$cov, tc$cov_record, method = "spearman"), 4),
            round(sum(gg$count) / sum(gg$record), 4), nrow(g10),
            round(mean(abs(g10$count - g10$record) <= 0.1 * pmax(g10$record, 1)), 3),
            round(cor(g10$count, g10$record, method = "spearman"), 4),
            sum(grepl("^STRG\\.[0-9]+$", gm$gene))))
stopifnot(nrow(fp) == nrow(port), all(fp$count == fp$S))
fwrite(checks, file.path(recount_dir, "checks.csv"))
unlink(tmp, recursive = TRUE)
knitr::kable(checks)
```

```{r provenance, eval = all_done}
ver <- function(x, args = "--version") {                       # first line of the version text, without the binary's path
  v <- trimws(system2(tool(x), args, stdout = TRUE, stderr = TRUE)); sub("^\\S*/", "", v[nzchar(v)][1])
}
settings <- c("[06_genome_recount.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              paste0("hisat2               : ", ver("hisat2"), "; index: plain genome (hisat2-build, no --ss/--exon); align: --dta --known-splicesite-infile (RefSeq, hisat2_extract_splice_sites.py)"),
              paste0("samtools             : ", ver("samtools")),
              paste0("stringtie            : ", ver("stringtie"), "; -e -G <annotation>; prepDE gene counts, read length 75 (_prepde_sample.R)"),
              paste0("featureCounts        : ", ver("featureCounts", "-v"), "; -s 1 -F SAF; unique alignments, reads on one gene only"),
              paste0("reads                : ", READS_URL, " (131 libraries)"),
              paste0("libraries            : ", done(), " of 131"))
writeLines(provenance_lines(settings, c(genome_fa, gtf_gz, ann_files, helper, here::here("01_code", "_prepde_sample.R"),
                                        here::here("01_code", "06_genome_recount.Rmd")),
                            repo_root, packages = "data.table"),
           file.path(recount_dir, "RUN_provenance.txt"))
```
