---
title: "02_7 - DESeq2 sensitivity: foot contrasts adjusted for byssal secretion state"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Why this script exists, and why it is off by default

`01_7-secretion_state.Rmd` labels every foot sample `on` / `off` for the byssal
plaque-protein module (mfp-2, mfp-4, foot proteins 10/12/15, tyrosinase-like 1): whether the
animal was actively secreting a thread when sampled. About a third of the day-3 animals are
"on", spread across every arm, and the module carries thousands of CPM, so it is a large
binary source of variance inside every foot contrast that the primary design
(`~ treatment`) leaves in the residual.

This script refits the three foot stressor contrasts (OA, OW, DO vs the day-3 control) with
the state as a nuisance covariate, `~ secretion_state + treatment`, and reports how the DEG
lists move. The module genes themselves are excluded from the DEG counts of the adjusted fit
(they define the covariate, so their "adjusted" p is meaningless).

**Gate.** `USE_SECRETION_STATE` below is `FALSE` by default and the analysis chunks do not
run until it is `TRUE`. The primary results of record stay unadjusted, because the state is
measured on the same sample as the contrast it would adjust: removing it also removes any
genuine stressor effect on the secretion rate. Turn it on to answer "what does the stressor
do beyond secretion state", and report both fits, never only the adjusted one.

```{r config}
USE_SECRETION_STATE <- FALSE   # TRUE = run the adjusted fits; FALSE = knit the note only
SECRETION_COVARIATE <- "state" # "state" (on/off factor) or "score" (continuous module score)
PADJ_CUTOFF <- 0.05
source(here::here("01_code", "_paths.R"))
suppressPackageStartupMessages({ library(tidyverse) })
STATE_FILE <- file.path(dat, "secretion_state.csv")
OUT <- file.path(deg, "sensitivity_secretion")
if (!USE_SECRETION_STATE) {
  cat("USE_SECRETION_STATE = FALSE: the secretion-state sensitivity fits are switched off.\n",
      "Set USE_SECRETION_STATE <- TRUE at the top of this script to run them (needs",
      STATE_FILE, "from 01_7-secretion_state.Rmd).\n")
}
```

```{r load, eval = USE_SECRETION_STATE}
suppressPackageStartupMessages({ library(DESeq2); library(apeglm) })
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)
if (!file.exists(STATE_FILE)) stop("Run 01_7-secretion_state.Rmd first: ", STATE_FILE, " is missing.")
state <- read_csv(STATE_FILE, show_col_types = FALSE)
master <- as.matrix(read.csv(file.path(dat, "gene_count_matrix_clean.csv"), row.names = 1, check.names = FALSE))
loc_of <- sub(".*\\|", "", sub("^gene-", "", rownames(master)))
## the genes that define the covariate (SECRETION_MODULE and COMPANION_GENES in 01_7); they
## are kept in the fit but removed from the adjusted DEG counts
MODULE_LOCS <- c("LOC134719757", "LOC134709805", "LOC134711496", "LOC134711495", "LOC134719452",
                 "LOC134686938", "LOC134718845", "LOC134716385")
module_rows <- rownames(master)[loc_of %in% MODULE_LOCS]
ti <- read.csv(file.path(dat, "treatmentinfo_clean.csv"), row.names = 1, stringsAsFactors = FALSE)

read_treatmentinfo <- function(path) {
  d <- read.table(path, header = TRUE, sep = " ", stringsAsFactors = FALSE)
  d <- d[, c("sample.1", "tissue", "treatment", "day")]; rownames(d) <- d$sample.1; d
}
primary_siggene <- function(trt) {
  f <- list.files(file.path(deg, "Foot"), pattern = sprintf("^F%s_TC_siggene.*\\.csv$", trt), full.names = TRUE)
  if (!length(f)) return(character(0))
  read.table(f[1], header = TRUE, sep = " ", stringsAsFactors = FALSE)$gene
}
```

# Adjusted fits

```{r fit, eval = USE_SECRETION_STATE}
fit_contrast <- function(trt) {
  cd <- read_treatmentinfo(file.path(deg, "Foot", sprintf("F%s_TC_treatmentinfo.csv", trt)))
  cd$treatment <- factor(cd$treatment, levels = c("control", trt))
  st <- state[match(cd$sample.1, state$sample), ]
  if (anyNA(st$secretion_state)) stop(trt, ": sample(s) without a secretion state: ",
                                      paste(cd$sample.1[is.na(st$secretion_state)], collapse = ", "))
  cd$secretion_state <- factor(st$secretion_state, levels = c("off", "on"))
  cd$secretion_score <- st$secretion_score
  cm <- master[, cd$sample.1, drop = FALSE]
  design <- if (SECRETION_COVARIATE == "score") ~ secretion_score + treatment else ~ secretion_state + treatment
  cat(sprintf("\n== F%s: %d samples; on/off by arm:\n", trt, ncol(cm)))
  print(table(cd$treatment, cd$secretion_state))
  dds  <- DESeqDataSetFromMatrix(cm, cd, design)
  dds  <- DESeq(dds, quiet = TRUE)
  keep <- rowSums(counts(dds) >= 10) >= ncol(cm) / 3
  ddsf <- dds[keep, ]
  coefn <- sprintf("treatment_%s_vs_control", trt)
  stopifnot(coefn %in% resultsNames(ddsf))
  r <- lfcShrink(ddsf, coef = coefn, type = "apeglm", quiet = TRUE)
  full <- data.frame(gene = rownames(r), as.data.frame(r), row.names = NULL) %>%
    mutate(is_module_gene = gene %in% module_rows)
  sig  <- full %>% filter(!is.na(padj), padj < PADJ_CUTOFF, !is_module_gene)
  write.table(full, file.path(OUT, sprintf("F%s_TC_secstate_apeglm.csv", trt)),  row.names = FALSE)
  write.table(sig,  file.path(OUT, sprintf("F%s_TC_secstate_siggene.csv", trt)), row.names = FALSE)
  ## the covariate's own effect, for the record (how much of the transcriptome the state moves)
  rs <- results(ddsf, name = if (SECRETION_COVARIATE == "score") "secretion_score" else "secretion_state_on_vs_off")
  prim <- primary_siggene(trt)
  tibble(code = paste0("F", trt), treatment = trt, covariate = SECRETION_COVARIATE,
         n_samples = ncol(cm), n_on = sum(cd$secretion_state == "on"),
         genes_after_filter = sum(keep),
         n_DEG_adjusted = nrow(sig), n_DEG_primary = length(prim),
         n_shared = length(intersect(sig$gene, prim)),
         n_only_adjusted = length(setdiff(sig$gene, prim)),
         n_only_primary  = length(setdiff(prim, sig$gene)),
         n_DEG_secretion_covariate = sum(!is.na(rs$padj) & rs$padj < PADJ_CUTOFF),
         module_genes_excluded = sum(full$is_module_gene))
}
cmp <- bind_rows(lapply(c("OA", "OW", "DO"), fit_contrast))
write_csv(cmp, file.path(OUT, "secretion_sensitivity_DEG_counts.csv"))
knitr::kable(cmp)
writeLines(c(paste0("generated : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
             paste0("design    : ~ secretion_", SECRETION_COVARIATE, " + treatment (foot TC contrasts)"),
             paste0("state file: ", STATE_FILE),
             "filter    : >= 10 counts in >= 1/3 of samples; apeglm; padj < 0.05; module genes excluded from DEG counts",
             paste0("R / DESeq2 / apeglm : ", R.version.string, " / ", as.character(packageVersion("DESeq2")),
                    " / ", as.character(packageVersion("apeglm")))),
           file.path(OUT, "RUN_provenance.txt"))
```

# Reading the comparison

`n_DEG_primary` is the unadjusted list of record; `n_DEG_adjusted` the same contrast with
the secretion state in the model (module genes removed). `n_DEG_secretion_covariate` is how
many genes the state itself moves at `padj < 0.05` within that contrast, which is the size of
the nuisance the primary design leaves in the residual. If `n_only_primary` is large, part
of the primary list was riding on the on/off split between arms; if `n_only_adjusted` is
large, the state was masking stressor effects. Either way the two lists, not one, are the
result.
