---
title: "04 - Join DEG lists to the genome annotation"
output: html_document
date: "2024-02-20 (rewritten 2026-09-16)"
---

```{r paths, include=FALSE}
source(here::here("01_code", "_paths.R"))
```

# What this does

For each treatment-vs-treatment-control contrast, join the DESeq2 DEG list
(`<X>_TC_siggene*.csv`, one row per gene, `padj < 0.05`) to the BLAST/UniProt/GO master
table (`blast/03_analyses/genome-foot/LOC_GO_list.txt`) and write three files to
`03_analyses/DEG_lists/GOterms_genome/`:

- `<X>_sigs_merged.csv`: every DEG with every annotation row it matched (a gene with several
  BLAST hits appears several times, so **this is not a gene count**);
- `<X>_sigs_ID.csv`: the merged rows that carry an annotation;
- `<X>_sigs_unID.csv`: DEGs with no annotation (one row each).

It also writes `03_analyses/DEG_lists/DEG_join_summary.csv` with, per contrast, the number
of DEGs (genes), merged rows, annotated and unannotated genes, and how many of the DEGs are
mitochondrial. Quote `n_DEG_genes` from that file, not the merged-row count.

## Join key and the mitochondrial genes

Nuclear genes are named `gene-LOC134721619|LOC134721619` in the count matrix and the master
table keys them by the `LOC` identifier. The 13 mitochondrial protein-coding genes are named
`gene-ND2|ND2`, `gene-CYTB|CYTB`, and so on; they carry no `LOC` identifier, and the master
table keys them by that symbol (`ND2`, `CYTB`, ...) because the BLAST step took the gene
symbol from the CDS FASTA header. The earlier version of this script extracted only
`LOC\\d+` as the key and, for the gill-OA contrast only, dropped every DEG whose key came
back `NA`, which removed the five mitochondrial DEGs in that contrast (ND1, ND2, ND6, CYTB,
ATP6) and defined that DEG set differently from the other five. The key now falls back to
the symbol, so those genes join to their annotation like any other, in every contrast.

The join is `left_join(..., na_matches = "never")`: the master table has about 16,800 rows
with an empty key (isoform hits already represented under another transcript of the same
gene), and a default join would attach all of them to any DEG whose key were missing.

```{r load_libraries}
suppressPackageStartupMessages({ library(tidyverse); library(stringr) })
```

```{r masterID}
# Output of blastx + UniProt + GO, one row per (transcript, hit)
masterID <- read.delim(file.path(blast_go, "LOC_GO_list.txt"), sep = "\t", header = TRUE)
# Drop the three trailing helper columns (V11, V2.x, uniprot_accession.y)
masterID <- masterID[, -((ncol(masterID) - 2):ncol(masterID))]
colnames(masterID) <- c("transcript", "database", "uniprot_accession", "geneID", "species",
                        "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart",
                        "send", "evalue", "bitscore", "reviewed", "entryname", "protein_name",
                        "gene_name", "organism", "prot_length", "GO_ID", "GO_BP", "GO_MF",
                        "GO_CC", "Genename_orderedlocus", "Induction", "Pathwaycommons",
                        "Reactome", "Unipathway", "Interacts_with", "LOC_ID")
masterID$LOC_ID[masterID$LOC_ID == ""] <- NA
cat("Master table:", nrow(masterID), "rows;", sum(is.na(masterID$LOC_ID)), "with no key;",
    dplyr::n_distinct(na.omit(masterID$LOC_ID)), "distinct keys\n")
```

```{r key}
# Join key for a count-matrix gene name: the LOC identifier when there is one, otherwise the
# gene symbol between "gene-" and "|" (mitochondrial genes). NA only if neither parses.
deg_key <- function(gene) {
  loc <- str_extract(gene, "LOC\\d+")
  sym <- str_match(gene, "^gene-([^|]+)\\|")[, 2]
  dplyr::coalesce(loc, sym)
}
is_mito <- function(key) !grepl("^LOC\\d+$", key)
```

# Contrasts

```{r contrasts}
contrasts <- tribble(
  ~code,  ~dir,   ~siggene,
  "FDO",  "Foot", "FDO_TC_siggene_apeglm.csv",
  "GDO",  "Gill", "GDO_TC_siggene_apeglm.csv",
  "FOA",  "Foot", "FOA_TC_siggene.csv",
  "GOA",  "Gill", "GOA_TC_siggene.csv",
  "FOW",  "Foot", "FOW_TC_siggene.csv",
  "GOW",  "Gill", "GOW_TC_siggene_apeglm.csv"
)
out_dir <- file.path(deg, "GOterms_genome")
dir.create(out_dir, showWarnings = FALSE)
```

# Join and write

```{r join}
join_contrast <- function(code, dir, siggene) {
  sig <- read.delim(file.path(deg, dir, siggene), sep = " ", header = TRUE) %>%
    mutate(LOC_ID = deg_key(gene))
  if (anyNA(sig$LOC_ID))
    stop(code, ": ", sum(is.na(sig$LOC_ID)), " DEG(s) with an unparseable gene name: ",
         paste(head(sig$gene[is.na(sig$LOC_ID)]), collapse = ", "))
  if (anyDuplicated(sig$gene) > 0)
    stop(code, ": duplicated gene names in ", siggene)

  merged <- left_join(sig, masterID, by = "LOC_ID", na_matches = "never",
                      relationship = "many-to-many")
  unID <- merged[is.na(merged$geneID), ]
  ID   <- merged[!is.na(merged$geneID), ]

  write_csv(merged, file.path(out_dir, sprintf("%s_sigs_merged.csv", code)))
  write_csv(unID,   file.path(out_dir, sprintf("%s_sigs_unID.csv", code)))
  write_csv(ID,     file.path(out_dir, sprintf("%s_sigs_ID.csv", code)))

  tibble(code = code, tissue = dir,
         n_DEG_genes         = nrow(sig),                       # the number to report
         n_merged_rows       = nrow(merged),                    # one row per gene x hit
         n_ID_rows           = nrow(ID),
         n_unID_rows         = nrow(unID),
         n_genes_annotated   = dplyr::n_distinct(ID$gene),
         n_genes_unannotated = dplyr::n_distinct(unID$gene),
         n_mito_DEGs         = sum(is_mito(sig$LOC_ID)),
         mito_DEGs           = paste(sig$LOC_ID[is_mito(sig$LOC_ID)], collapse = ";"))
}

summary_tbl <- pmap_dfr(contrasts, join_contrast)
stopifnot(all(summary_tbl$n_genes_annotated + summary_tbl$n_genes_unannotated == summary_tbl$n_DEG_genes))
write_csv(summary_tbl, file.path(deg, "DEG_join_summary.csv"))
knitr::kable(summary_tbl)
```

`n_DEG_genes` is what the manuscript should report as the number of DEGs. `n_merged_rows`
is larger wherever a gene matched more than one BLAST hit; the earlier draft of the results
section quoted those row counts (409, 251, 830 gill; 460, 201, 92 foot) as DEG numbers.
