---
title: "01 - HISAT2 alignment and StringTie quantification (HPC)"
author: "Steven Roberts"
date: "`r format(Sys.time(), '%d %B, %Y')`"  
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
library(DT)
library(Biostrings)
library(tm)
library(pheatmap)
library(DESeq2)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center", # Align plots to the center
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Differentially Expressed Genes
 


# Reads

```{r, engine='bash', eval=TRUE}
ls /home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis/data/raw-trimmed/
```


# Genome


```{r, engine='bash'}
cd ../02_data

/home/shared/datasets download genome accession GCF_036588685.1 --include gff3,gtf,rna,cds,protein,genome,seq-report
```

```{bash}
cd ../02_data 
unzip ncbi_dataset.zip
```

```{r, engine='bash', eval=TRUE}
ls ../02_data/ncbi_dataset/data/GCF_036588685.1

```




# Hisat

```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2_extract_exons.py \
../02_data/ncbi_dataset/data/GCF_036588685.1/genomic.gtf \
> ../03_analyses/hisat/m_exon.tab
```

```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2_extract_splice_sites.py \
../02_data/ncbi_dataset/data/GCF_036588685.1/genomic.gtf \
> ../03_analyses/hisat/m_spice_sites.tab
```




```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2-build \
../02_data/ncbi_dataset/data/GCF_036588685.1/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fna \
../03_analyses/hisat/PNRI_Mtr1.1.1.index \
--exon ../03_analyses/hisat/m_exon.tab \
--ss ../03_analyses/hisat/m_spice_sites.tab \
-p 32 \
../02_data/ncbi_dataset/data/GCF_036588685.1/genomic.gtf \
2> ../03_analyses/hisat/hisat2-build_stats.txt
```


H9-T014F_S171_L099_R1_cmb.trim.fastq.gz

/home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis/data/raw-trimmed/

```{r, engine='bash', eval=TRUE}
find /home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis/data/raw-trimmed/*_L099_R1_cmb.trim.fastq.gz | xargs basename -s _L099_R1_cmb.trim.fastq.gz | xargs -I{} echo {}
```





```{r, engine='bash'}
find /home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis/data/raw-trimmed/*_L099_R1_cmb.trim.fastq.gz | xargs -I{} basename -s _L099_R1_cmb.trim.fastq.gz {} | xargs -I{} sh -c '/home/shared/hisat2-2.2.1/hisat2 \
-x ../03_analyses/hisat/PNRI_Mtr1.1.1.index \
--dta \
-p 32 \
-U /home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis/data/raw-trimmed/{}_L099_R1_cmb.trim.fastq.gz \
-S ../03_analyses/hisat/{}.sam \
> ../03_analyses/hisat/{}_hisat.stdout 2> ../03_analyses/hisat/{}_hisat.stderr'
```


Explanation
xargs -I{}: This option allows you to replace {} in the command with the output from the previous command (i.e., basename). It's used twice: first, to strip the suffix from the filenames, and second, to construct and execute the hisat2 command.

sh -c: This is used to execute a complex command within xargs. It's necessary because the output redirection (>, 2>) is shell functionality, and without sh -c, xargs wouldn't handle it correctly.




```{r, engine='bash'}
for samfile in ../03_analyses/hisat/*.sam; do
  bamfile="${samfile%.sam}.bam"
  sorted_bamfile="${samfile%.sam}.sorted.bam"
  /home/shared/samtools-1.12/samtools view -bS -@ 20 "$samfile" > "$bamfile"
  /home/shared/samtools-1.12/samtools sort -@ 20 "$bamfile" -o "$sorted_bamfile"
  /home/shared/samtools-1.12/samtools index -@ 20 "$sorted_bamfile"
done
```

,
```{r, engine='bash'}
rm ../03_analyses/hisat/*sam
```


```{r, engine='bash'}
ls ../03_analyses/hisat/*sorted.bam | wc -l
```




# Stringtie

`-B` writes the Ballgown tables (`*.ctab`) into the folder of the `-o` file, so each sample
gets its own folder; with a shared folder every sample overwrote the previous one's tables.
(The `.ctab` files committed in `../03_analyses/hisat/` are the last sample's, H9-T014F_S171.)




```{r, engine='bash'}
find ../03_analyses/hisat/*sorted.bam \
| xargs basename -s .sorted.bam | xargs -I{} \
sh -c 'mkdir -p ../03_analyses/hisat/{} && /home/shared/stringtie-2.2.1.Linux_x86_64/stringtie \
-p 36 \
-eB \
-G ../02_data/ncbi_dataset/data/GCF_036588685.1/genomic.gff \
-o ../03_analyses/hisat/{}/{}.gtf \
../03_analyses/hisat/{}.sorted.bam'
```



# Sample list for prepDE

One line per sample, `<sample ID> <path to its GTF>`, read by `02_prepDE.Rmd` (the R port of
StringTie's `prepDE.py`; the original was run with its default read length, 75). The sample ID
is the read file's base name (e.g. `A1-T001F_S100`), as in the original count matrices.

```{r, engine='bash'}
for d in ../03_analyses/hisat/*/; do
  s=$(basename "$d")
  [ -f "$d$s.gtf" ] && echo "$s $(cd "$d" && pwd)/$s.gtf"
done > ../03_analyses/hisat/sample_list.txt
wc -l ../03_analyses/hisat/sample_list.txt
```

Compare genome with gff...

```{bash}
head -30 ../02_data/ncbi_dataset/data/GCF_036588685.1/genomic.gff
```


```{bash}
head ../02_data/ncbi_dataset/data/GCF_036588685.1/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fna
```





```{bash}
head ../03_analyses/hisat/*119.gtf

```







```{r, engine='bash'}
eval "$(/opt/anaconda/anaconda3/bin/conda shell.bash hook)"
conda activate
which multiqc


multiqc ../03_analyses/hisat/ \
-o ../03_analyses/hisat/
```




```{bash}
head ../03_analyses/hisat/*gtf
```