---
title: "02 - Gene and transcript count matrices from StringTie (R port of prepDE.py)"
output: html_document
params:
  read_length: 75     # prepDE.py's default, which the original run used (verified below)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# What this does

Turns the per-sample StringTie output of `01_hisat_stringtie.Rmd` into the gene and transcript
count matrices of the previous record (the count matrices of record until 2026-10-02, when
step 07's featureCounts matrix replaced them), with `prepde()` in
`_prepde.R`, an R port of StringTie's `prepDE.py3`:

    count = ceiling(cov * transcript length / read length)

summed over a gene's transcripts. The port reproduces `prepDE.py3` byte for byte (tested on
synthetic GTFs with edge cases, both input modes and three read lengths, and on this genome's
full transcript set rebuilt from the committed Ballgown tables).

Two ways to run, chosen automatically:

1. **From StringTie GTFs** (`03_analyses/hisat/sample_list.txt` exists; written by script 01):
   runs `prepde()` and writes both matrices.
2. **From the committed transcript matrix** (no GTFs; the original ones are lost). The
   transcript matrix `03_analyses/prepDE/transcript_count_matrix.csv` is the original prepDE
   output, so the gene matrix is rebuilt by summing transcripts into genes. The transcript-to-
   gene map comes from the reference transcript table `03_analyses/hisat/t_data.ctab`, plus
   `02_data/strg_gene_ids.csv` for the 284 transcripts StringTie gave its own gene IDs (see
   `_derive_strg_gene_ids.R`).

Files are written with LF line ends, as everything in this repository (prepDE.py3 writes
CRLF; the values are identical). Either way the output is `03_analyses/prepDE/gene_count_matrix.csv`,
which steps 06 and 07 (comparisons), `05_differential-expression` step 13 and
`06_iso-seq-transcriptome` step 04 read as the previous record.

```{r paths}
source(here::here("01_code", "_prepde.R"))
out_dir     <- here::here("03_analyses", "prepDE")
gene_csv    <- file.path(out_dir, "gene_count_matrix.csv")
tx_csv      <- file.path(out_dir, "transcript_count_matrix.csv")
sample_list <- here::here("03_analyses", "hisat", "sample_list.txt")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
mode <- if (file.exists(sample_list)) "gtf" else "transcript_matrix"
cat("mode:", mode, "\n")
```

# 1. From StringTie GTFs

```{r from-gtf, eval = (mode == "gtf")}
prepde(sample_list, gene_out = gene_csv, transcript_out = tx_csv,
       read_length = params$read_length, verbose = TRUE, eol = "\n")
```

# 2. From the committed transcript matrix

```{r from-transcripts, eval = (mode == "transcript_matrix")}
stopifnot(file.exists(tx_csv))
td <- read.delim(here::here("03_analyses", "hisat", "t_data.ctab"))

## prepDE's gene ID: gene_id|gene_name, else gene_id, else the transcript ID
gid <- ifelse(td$gene_name != ".", paste0(td$gene_id, "|", td$gene_name), td$gene_id)
no_gene <- td$gene_id == "."
gid[no_gene] <- td$t_name[no_gene]
tx_to_gene <- setNames(gid, td$t_name)

## the 284 transcripts StringTie gave its own gene IDs (STRG.<n>|LOC<n>)
strg <- read.csv(here::here("02_data", "strg_gene_ids.csv"))
stopifnot(all(strg$transcript_id %in% td$t_name[no_gene]))
tx_to_gene[strg$transcript_id] <- strg$gene_id

prepde_genes_from_transcripts(tx_csv, tx_to_gene, gene_csv, eol = "\n")
```

# Checks

```{r checks}
g  <- read.csv(gene_csv, check.names = FALSE)
tx <- read.csv(tx_csv, check.names = FALSE)
cat("genes:", nrow(g), "| transcripts:", nrow(tx), "| samples:", ncol(g) - 1, "\n")
stopifnot(identical(names(g)[-1], names(tx)[-1]),
          !anyDuplicated(g$gene_id),
          sum(g[, -1]) == sum(tx[, -1]))                  # every transcript count lands in one gene

## The read length: in the committed Ballgown tables (one sample, H9-T014F_S171, the last one
## StringTie wrote before the per-sample folders fix), ceiling(cov * length / 75) reproduces
## that sample's column of the transcript matrix exactly; other read lengths do not.
td <- read.delim(here::here("03_analyses", "hisat", "t_data.ctab"))
if ("H9-T014F_S171" %in% names(tx)) {
  pred <- as.integer(ceiling(td$cov * td$length / params$read_length))
  obs  <- tx[["H9-T014F_S171"]][match(td$t_name, tx$transcript_id)]
  cat("read length", params$read_length, ": H9-T014F_S171 transcripts reproduced",
      sum(pred == obs, na.rm = TRUE), "of", nrow(td), "\n")
  stopifnot(all(pred == obs))
}
```
