---
title: "08 - Mitochondrial reads counted on the mitochondrial genome alone"
output: html_document
params:
  online: false            # true: download the 131 trimmed libraries, align and count (about 1 hour); false: summarise the committed counts
  bin: ""                  # folder holding hisat2 (2.2.1), hisat2-build, samtools and featureCounts; "" = PATH
  threads: 4
  max_libraries: .Inf      # process at most this many new libraries per knit (for batches)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Counts the reads of the 12 mitochondrial protein genes (and the rRNAs and tRNAs) by aligning
each library's trimmed reads to the mitochondrial genome alone, NC_007687.1
(`02_data/mitogenome_NC_007687.1.fa`, its 38 features in `02_data/mitogenome_genes.saf`, both
cut once from the RefSeq assembly by `_derive_mitogenome.R`). `05_differential-expression`
step 13 tests these counts, per protein, against the nuclear genes of the count matrix of
record.

**Why not the genome alignment.** In the genome alignment the mitochondrial reads compete with
293 annotated nuclear copies of the mitochondrial genes (`tools/mt_encoded.R`) and with
unannotated mitochondrial-like sequence, and HISAT2 reports at most five placements of a read.
In a test on 10 libraries (2026-10-02, outside the pipeline, alignments not kept), 72 to 93% of
the reads on mitochondrial loci had five placements and only 41 to 55% of their weight fell on
mitochondrial loci. Twelve of the 59 animals also carry mitochondrial haplotypes that differ
from the reference at fixed positions (every read of the animal, in every tissue), so their
reads match the copies as well as the mitogenome; `05` step 13 assigns the groups. On the
mitogenome alone each read has one place to go. Against the previous count (each protein's
mitogenome gene and copies summed in the StringTie + prepDE matrix; `05 .../mitochondrial/mt_counts_vs_previous.csv`),
the 12 proteins get about 2.0 times the reads in reference-like animals and 2.4 to 2.5 times
in the divergent ones; COX1, where the haplotypes differ most, 1.5 against 3.3 to 3.7 times;
ND3, which has no annotated copy, 10 times or more. Reads transcribed from nuclear copies, if
any, are counted as mitochondrial here, as summing the copies did before.

**How.** Each library (gannet `.../byssus-exp-analysis/data/raw-trimmed/`, the reads every
count rests on) is downloaded, aligned and deleted by `_mitogenome_library.sh`:

- HISAT2 2.2.1 (Kim et al. 2019, *Nature Biotechnology* 37:907) against the mitogenome, no
  spliced alignment, default scoring (minimum score -0.2 x read length, about two mismatches in
  a 63-base read): **of record**. Again with a permissive minimum score (-0.6 x read length,
  about six mismatches): a sensitivity setting for the divergent haplotypes.
- featureCounts (Subread 2.1.1; Liao, Smyth and Shi 2014, *Bioinformatics* 30:923) on the 38
  features, sense strand (`-s 1`); the few reads that overlap two adjacent features or align to
  two places are shared fractionally (`-M -O --fraction`).

With `online: false` (the pipeline) the step only summarises the committed counts.

Writes, in `03_analyses/mitogenome/`:

- `mitogenome_gene_counts.csv`: the 38 features x 131 libraries (default scoring), the counts
  step 13 tests;
- `mitogenome_gene_counts_permissive.csv`: the same with the permissive score;
- `mapping_summary.csv`: per library, reads, mean read length, reads aligned to the mitogenome
  (both settings), and reads on the 12 protein genes and on the rRNAs;
- `RUN_provenance.txt`: tool versions, settings and input MD5s (online runs);
- `libraries/<library>/` (git-ignored): the per-library counts and logs.

```{r paths}
repo_root <- normalizePath(file.path(here::here(), ".."))
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # provenance_lines()
suppressPackageStartupMessages(library(data.table))
mito_fa  <- here::here("02_data", "mitogenome_NC_007687.1.fa")
mito_saf <- here::here("02_data", "mitogenome_genes.saf")
out_dir  <- here::here("03_analyses", "mitogenome")
lib_root <- file.path(out_dir, "libraries")
READS_URL <- "https://gannet.fish.washington.edu/panopea/PSMFC-mytilus-byssus-pilot/byssus-exp-analysis/data/raw-trimmed/"
PROTEINS <- c("ND1", "ND2", "ND3", "ND4", "ND4L", "ND5", "ND6", "CYTB", "COX1", "COX2", "COX3", "ATP6")
dir.create(lib_root, showWarnings = FALSE, recursive = TRUE)
saf <- fread(mito_saf)
stopifnot(nrow(saf) == 38, all(PROTEINS %in% saf$GeneID))
```

# Align and count (online)

```{r align, eval = isTRUE(params$online)}
tool <- function(x) if (nzchar(params$bin)) file.path(params$bin, x) else unname(Sys.which(x))
for (x in c("hisat2", "hisat2-build", "samtools", "featureCounts"))
  if (!nzchar(tool(x)) || !file.exists(tool(x))) stop(x, " not found; set params$bin")
idx <- file.path(tempdir(), "mitogenome")
stopifnot(system2(tool("hisat2-build"), c("-q", shQuote(mito_fa), shQuote(idx)), stdout = FALSE, stderr = FALSE) == 0)
helper <- here::here("01_code", "_mitogenome_library.sh")
listing <- readLines(url(READS_URL), warn = FALSE)
files <- unique(regmatches(listing, regexpr("[A-H][0-9]+-T[0-9]+[A-Z]*_S[0-9]+_L099_R1_cmb\\.trim\\.fastq\\.gz", listing)))
lib_of <- function(f) sub("^[A-H][0-9]+-", "", sub("_S[0-9]+_L099.*$", "", f))
stopifnot(length(files) == 131, !anyDuplicated(lib_of(files)))
n_new <- 0
for (f in files) {
  out <- file.path(lib_root, lib_of(f))
  if (file.exists(file.path(out, "done"))) next
  if (n_new >= params$max_libraries) break
  unlink(out, recursive = TRUE)                              # a library a stopped run left half done
  reads <- tempfile(fileext = ".fastq.gz")
  ok <- FALSE
  for (i in 1:4) { if (download.file(paste0(READS_URL, f), reads, mode = "wb", quiet = TRUE) == 0) { ok <- TRUE; break }; Sys.sleep(2^i) }
  if (!ok) stop("download failed: ", f)
  rc <- system2("bash", c(shQuote(helper), shQuote(reads), shQuote(out), shQuote(idx), shQuote(mito_saf),
                          shQuote(dirname(tool("hisat2"))), params$threads))
  unlink(reads)
  if (rc != 0) stop("_mitogenome_library.sh failed for ", f)
  n_new <- n_new + 1
}
libs <- sort(lib_of(files))
cat(n_new, "libraries processed in this run;", sum(file.exists(file.path(lib_root, libs, "done"))), "of 131 done\n")
```

```{r combine, eval = isTRUE(params$online) && all(file.exists(file.path(lib_root, libs, "done")))}
read_counts <- function(v) {
  long <- rbindlist(lapply(libs, function(l) fread(file.path(lib_root, l, paste0("counts_", v, ".tsv")), header = FALSE,
                                                    col.names = c("gene", "count"))[, library := l]))
  w <- dcast(long, gene ~ library, value.var = "count", fill = 0)
  w <- w[match(saf$GeneID, gene)]
  setcolorder(w, c("gene", libs))
  stopifnot(!anyNA(w$gene))
  w
}
cnt <- read_counts("default"); cnt_p <- read_counts("permissive")
fwrite(cnt, file.path(out_dir, "mitogenome_gene_counts.csv"))
fwrite(cnt_p, file.path(out_dir, "mitogenome_gene_counts_permissive.csv"))
aligned <- function(l, v) {
  x <- readLines(file.path(lib_root, l, paste0("hisat2_", v, ".log")))
  sum(as.numeric(sub("^\\s*([0-9]+).*$", "\\1", grep("aligned (exactly 1|>1) time", x, value = TRUE))))
}
rd <- rbindlist(lapply(libs, function(l) fread(file.path(lib_root, l, "reads.tsv"), header = FALSE,
                                                col.names = c("reads", "mean_length"))[, library := l]))
ms <- rd[, .(library, reads, mean_length,
             aligned = vapply(library, aligned, 0, v = "default"),
             aligned_permissive = vapply(library, aligned, 0, v = "permissive"))]
ms[, aligned_pct := round(100 * aligned / reads, 2)]
ms[, protein_reads := colSums(as.matrix(cnt[gene %in% PROTEINS, ..libs]))[library]]
ms[, rRNA_reads := colSums(as.matrix(cnt[gene %in% saf$GeneID[saf$End - saf$Start > 500 & !saf$GeneID %in% PROTEINS], ..libs]))[library]]
fwrite(ms, file.path(out_dir, "mapping_summary.csv"))
hv <- system2(tool("hisat2"), "--version", stdout = TRUE)[1]
fv <- system2(tool("featureCounts"), "-v", stdout = TRUE, stderr = TRUE)
fv <- trimws(grep("featureCounts v", fv, value = TRUE)[1])
writeLines(provenance_lines(
  c("04 step 08: mitochondrial reads on the mitogenome alone",
    paste0("run                  : ", format(Sys.time(), "%Y-%m-%d %H:%M %Z")),
    paste0("hisat2               : ", sub("^.*version ", "", hv), "; --no-spliced-alignment; default score (of record) and --score-min L,0,-0.6"),
    paste0("featureCounts        : ", fv, "; -s 1 -M -O --fraction -F SAF"),
    paste0("reads                : ", READS_URL, " (131 libraries)")),
  c(here::here("01_code", "08_mitogenome_counts.Rmd"), here::here("01_code", "_mitogenome_library.sh"), mito_fa, mito_saf),
  repo_root, packages = "data.table"), file.path(out_dir, "RUN_provenance.txt"))
```

# Summary

```{r summary}
cnt <- fread(file.path(out_dir, "mitogenome_gene_counts.csv"))
ms  <- fread(file.path(out_dir, "mapping_summary.csv"))
stopifnot(ncol(cnt) == 132, identical(cnt$gene, saf$GeneID), nrow(ms) == 131)
cat("reads aligned to the mitogenome: median", median(ms$aligned_pct), "% of the library (range",
    paste(range(ms$aligned_pct), collapse = " to "), ")\n")
knitr::kable(data.table(protein = PROTEINS,
                        median_count = apply(as.matrix(cnt[match(PROTEINS, gene), -1]), 1, median)),
             caption = "reads per library on each protein gene (default scoring)")
```
