---
title: "01 - Gene expression x byssal thread mechanics correlation (paired samples)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"   # "F" foot (default) or "G" gill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# Purpose

A subset of *M. trossulus* that were mechanically tested for byssal plaque attachment
were also sampled for Tag-seq. This script links those two datasets at the level of the
individual animal and asks: **among animals in the same arm with the same measured
baseline, does foot (or gill) gene expression at the end of the exposure covary with the
animal's day-3 plaque attachment?**

The estimand is one slope pooled over the arms: the change in the day-3 level (log scale, so
exp(slope) is a fold change per VST unit) per unit of VST expression, holding arm and the animal's measured baseline constant. The common
slope is an assumption: with no expression x arm term it is the variance-weighted average of
the within-arm slopes. With 10 to 12 animals per arm the design cannot tell a stress-specific
association from a general one. It does not estimate whether expression explains the arm
effects (mediation): each arm is one tank, library plate and RNA isolation date are grouped
by arm, and expression was sampled after the tested threads were made.

It produces:

1. A paired-sample manifest (exactly which animals/samples were used).
2. Per-gene association statistics: one baseline-adjusted regression (ANCOVA) of each
   animal's day-3 level of a mechanical metric on its expression, controlling for treatment
   and for the animal's own pre-exposure level, with multiple-testing correction.
3. Two gene universes: a biologically motivated **candidate set** (byssal / collagen /
   HSP / HIF / peroxidase / aminoacyl-tRNA-synthetase genes) and the **data-driven DEG
   union** (genes already found to respond to any stressor).
4. Diagnostic plots (added-variable plots of the top hits, candidate-gene heatmap).

> **READ THIS FIRST - power.** The paired set is small, even with the full thread dataset and
> the day-3 control arm included; the baseline-adjusted test uses only the animals with a
> baseline, and within any single arm n is smaller still. This analysis is
> **exploratory and hypothesis-generating, not confirmatory.** Treat raw p-values as ranking
> signals, expect few or no hits to survive FDR correction, and do not write causal language
> from it. Tissue was sampled destructively at the end of the exposure, after the tested
> day-3 threads were secreted. An association can reflect an individual state that persisted
> from secretion to sampling, a lagged effect, reverse direction or a common cause; it is not
> a demonstrated mechanism. Use "associated within arm, adjusted for baseline", not
> "predicts".

> **The search is declared here, not in the write-up.** `METRICS` below names the metrics
> that form the confirmatory family (candidate genes x primary metrics); every other metric
> is labelled `exploratory` in every output table (`tier` column), and `q_family` is the BH
> correction over each gene set x tier family, so the primary family carries its own
> search-corrected q. Every gene tested here has already been analysed once on these
> animals, so no family is confirmatory in the strict sense; `tier = primary` names the
> family whose q is the decision column.

> **Which metric to read first.** **Mean peak force and plaque area** are tested separately
> as well as through their ratio, since adhesion mixes the two
> (`02_thread-strength/01_code/05_decompose_adhesion.Rmd` decomposes it the same way) and a gene
> associated with plaque size can be invisible against the ratio. `mean_force` and
> `pad_area` are the primary targets here; `max_force` (the animal's strongest thread) and
> `adhesion_kpa` (the material-strength readout) are exploratory. Extension is not analysed:
> the length of distal thread under test differed between pulls. Every metric is tested as the animal's day-3 level with its
> own measured baseline level as a covariate. The baseline is measured with error, so the
> adjustment for where the animal started is only partial.

---

# 0. Configuration

Set these two paths to wherever the repos live on your machine. The script reads thread
data from the **main** repo and expression data from the **expression** repo.

```{r config}
## ---- analysis options ---------------------------------------------------------------
## Tissue comes from the knit parameter (`params$tissue`, default "F") so the same script
## serves foot and gill; the driver 00_run_gene_mechanics_by_tissue.Rmd renders both.
## When run chunk-by-chunk outside a knit, `params` may not exist; fall back to foot.
TISSUE <- if (exists("params") && !is.null(params$tissue)) toupper(params$tissue) else "F"
stopifnot(TISSUE %in% c("F", "G"))
MIN_COUNT     <- 5            # gene kept if count > MIN_COUNT in >= 1/3 of paired samples
FDR_ALPHA     <- 0.10         # BH threshold for flagging
## Detection floor. A gene whose VST value sits at the floor (zero counts) in many paired
## animals gives an association driven by presence/absence, not by expression level. Every
## tested gene gets frac_at_floor and floor_flag (exclude > FLOOR_EXCLUDE, caution >
## FLOOR_CAUTION); with EXCLUDE_FLOOR_GENES = TRUE the `exclude` genes are dropped from the
## candidate and DEG-union families BEFORE testing, so BH runs on the family that can be
## interpreted. They stay listed (floor_flag, tested = FALSE) and they still contribute to
## script 02's module scores.
FLOOR_EXCLUDE <- 0.40
FLOOR_CAUTION <- 0.20
EXCLUDE_FLOOR_GENES <- TRUE
INCLUDE_CONTROL_ARM <- TRUE   # TRUE = the day-3 treatment control animals (T126-T137, the
                              #        control of record) are part of the paired set, as a
                              #        fourth treatment level. They have foot RNA, day-3
                              #        threads and baseline threads. They share the pooled
                              #        slope; they are not a separate null (caveat 6 below).
                              # FALSE = stressor arms only.

## ---- mechanical metrics and the reported test ----------------------------------------
## One test per gene x metric: an analysis of covariance of the animal's day-3 level on
## its expression, with the arm and the animal's OWN pre-exposure
## level as covariates:
##     level_day3 ~ expression + treatment + level_baseline
## Every metric is modelled on the LOG scale (multiplicative, right-skewed quantities).
## `column` is the thread-level column of the thread summary and `agg` its per-animal summary,
## as in 02_thread-strength/01_code/_ancova.R (THREAD_METRICS): `mean` is the geometric mean
## of the threads, `max` the largest thread. `mean_force` and `max_force` both summarise each
## thread's peak force (`max_force` of the trace). Animals without baseline threads are not in
## the fits.
## `tier`: mean force and area are the pre-declared confirmatory family (candidate genes x
## primary metrics); max force and adhesion (the ratio of force and area) are exploratory.
## Every output table carries the tier, and `q_family` corrects within each gene set x tier.
## Order here is the reporting order. Force and area first, deliberately (see above).
METRICS <- data.frame(
  metric = c("mean_force",          "pad_area",           "max_force",              "adhesion_kpa"),
  column = c("max_force",           "pad_area",           "max_force",              "adhesion_kpa"),
  agg    = c("mean",                "mean",               "max",                    "mean"),
  scale  = c("log",                 "log",                "log",                    "log"),
  tier   = c("primary",             "primary",            "exploratory",            "exploratory"),
  label  = c("Mean peak force (N)", "Plaque area (mm^2)", "Maximum peak force (N)", "Adhesion (kPa)"),
  stringsAsFactors = FALSE)
metrics         <- METRICS$metric
PRIMARY_METRICS <- METRICS$metric[METRICS$tier == "primary"]
scale_of <- function(m) METRICS$scale[match(m, METRICS$metric)]
tier_of  <- function(m) METRICS$tier[match(m, METRICS$metric)]
label_of <- function(m) METRICS$label[match(m, METRICS$metric)]
on_scale <- function(m, v) if (scale_of(m) == "log") log(v) else v    # a value on its model scale

## ---- candidate gene keywords (regex, case-insensitive) ------------------------------
## Grouped by biology; edit freely. Matched against UniProt "protein_name".
CANDIDATE_KEYWORDS <- paste(
  # byssal structural / adhesive proteins (plaque proteins, thread core, cuticle)
  "byss", "foot protein", "Mefp", "adhesive plaque", "polyphenolic",
  "byssal protein-", "preCol", "precollagen", "byssal EP", "ACDC", "byssal tyrosinase",
  "thread matrix protein", "nongradient byssal", "YGH-rich protein",
  "byssal peroxidase", "byssal HSP", "byssal calumenin", "byssal protease inhibitor",
  "protease inhibitor-like protein-1", "C1q-domain-containing protein-1",
  "TSP_1 domain containing protein-1",
  # collagens / preCols (byssal thread core)
  "collagen",
  # plaque-curing / cross-linking enzymes
  "tyrosinase", "peroxidase", "amine oxidase", "catechol oxidase", "lysyl oxidase",
  # proteostasis / heat-shock
  "heat shock", "Hsp", "stress-70", "BiP", "chaperone",
  "endoplasmic reticulum chaperone", "hypoxia up-regulated",
  # hypoxia signalling
  "hypoxia-inducible", "Egl nine", "prolyl 4-hydroxylase", "prolyl hydroxylase",
  # translational machinery (aminoacyl-tRNA synthetases)
  "--tRNA", "tRNA ligase", "aminoacyl",
  # oxidative stress / detox
  "superoxide", "glutathione", "catalase", "peroxiredoxin", "dual oxidase",
  sep = "|"
)
```


```{r}
# Get the correct Bioconductor repository link for your R version
# bioc_version <- BiocManager::version()
# bioc_repo <- paste0("https://bioconductor.org/packages/", bioc_version, "/bioc")
# 
# # Install DESeq2 directly as a standard package 
# install.packages("GenomeInfoDbData", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))
# install.packages("DESeq2", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))

```




```{r libs}
## Load Packages
## DESeq2 first, tidyverse LAST: S4Vectors/IRanges (pulled in by DESeq2) mask dplyr's
## rename(), count(), first() and desc(). Loading tidyverse after them restores dplyr's.
load.lib <- c("DESeq2","readxl","openxlsx","here","tidyverse")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for(lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)


`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a
```

```{r paths}
## Resolve the repository root (the folder holding the analysis subfolders), robust to whether
## this is knit from gene-mechanics-correlation.Rproj or the top-level project.
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("05_differential-expression", "02_thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing 05_differential-expression/ and 02_thread-strength/.")
}
repo_root   <- find_repo_root()
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # warn_unless(), provenance
source(file.path(repo_root, "tools", "plot_style.R"))        # shared colours, short_name()
source(file.path(repo_root, "tools", "gene_ids.R"))          # gene_key()
DE_DIR      <- file.path(repo_root, "05_differential-expression")          # expression-side inputs
COUNT_FILE  <- file.path(DE_DIR, "03_analyses", "count_matrix", "gene_count_matrix_clean.csv")
TI_FILE     <- file.path(DE_DIR, "03_analyses", "count_matrix", "treatmentinfo_clean.csv")
## The thread summary written by 02_thread-strength script 03, and the per-animal response
## classification written by 02_thread-strength script 05.
THREAD_FILE   <- file.path(repo_root, "02_thread-strength", "03_analyses", "03_assemble-thread-summary",
                           "thread-summary.xlsx")
RESPONSE_FILE <- file.path(repo_root, "02_thread-strength", "03_analyses", "05_decompose-adhesion",
                           "mussel_response_classification.csv")
OUT_DIR     <- file.path(repo_root, "09_gene-mechanics-correlation", "03_analyses", "gene_mechanics")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
## ---- annotation map and candidate universe ------------------------------------------
## CANDIDATE_ANNOTATION:
##   "genome" (default) = every gene in the count matrix is annotated with its best UniProt
##            hit (highest bitscore) from the genome-wide BLAST table MASTER_ANNOT_FILE, so
##            any EXPRESSED gene whose name matches CANDIDATE_KEYWORDS is a candidate.
##   "TC_DEG" = only genes that carry a name in the treatment-vs-control DEG tables
##            (GOterms_genome/*_sigs_ID.csv) can be candidates, plus the byssal structural
##            genes. This makes "already a DEG in some contrast" a hidden entry condition
##            for the candidate family.
## The two sources agree on the name for 947 of the 969 DEG-annotated genes; the 22 that
## differ are LOCs whose transcripts hit different proteins, and the genome map takes the
## highest-scoring hit (the same rule script 04 uses).
## CANDIDATE_MAX_EVALUE / CANDIDATE_MIN_PIDENT: BLAST-quality floor for a gene to be a
## candidate under "genome" (the map itself keeps every hit, with pident and evalue columns).
## On this table every keyword match already has evalue < 1e-10; pident >= 40 is script
## 04's HI_CONF_ONLY rule.
CANDIDATE_ANNOTATION <- "genome"
CANDIDATE_MAX_EVALUE <- 1e-10
CANDIDATE_MIN_PIDENT <- 0
MASTER_ANNOT_FILE <- file.path(repo_root, "03_blast", "03_analyses", "genome-foot-sprot2026_03", "LOC_GO_list.txt")
BYSSAL_STRUCTURAL_REGEX <- paste("foot protein", "adhesive plaque", "polyphenolic", "byssal protein-",
                                 "preCol", "precollagen", "byssal EP", "ACDC", "byssal tyrosinase",
                                 "thread matrix protein", "nongradient byssal", "YGH-rich protein",
                                 sep = "|")

if (!file.exists(THREAD_FILE))
  stop("Thread summary not found:\n  ", THREAD_FILE,
       "\nRun 02_thread-strength scripts 02 and 03; script 03 writes it to\n",
       "03_analyses/03_assemble-thread-summary/thread-summary.xlsx.")

## Canonicalize a mussel tag to one 3-digit form (T55 -> T055) so the thread side and the
## expression side cannot split on zero-padding when joined.
norm_id <- function(x) {
  x   <- as.character(x)
  pre <- gsub("[0-9].*$", "", x)                                        # leading letters
  num <- suppressWarnings(as.integer(gsub("[^0-9].*$", "", gsub("^[^0-9]+", "", x))))
  ifelse(is.na(num), NA_character_, sprintf("%s%03d", pre, num))
}

## Read the thread summary under either the current or the older column names, and
## make sure `phase` exists. thread_trt = what the THREAD was built in; mussel_trt = the arm
## the ANIMAL was assigned to; phase = lab / pre / post. The older `treatment == "control"`
## convention pooled baseline threads with the day-3 control arm and is not used here.
read_thread_summary <- function(path) {
  d <- readxl::read_excel(path, sheet = "data")
  ren <- c(mussel_ID = "mussel", thread_num = "thread", treatment = "thread_trt")
  for (from in names(ren)) {
    to <- ren[[from]]
    if (from %in% names(d) && !to %in% names(d)) names(d)[names(d) == from] <- to
  }
  need <- c("mussel", "thread_trt", "mussel_trt", "max_force", "pad_area")
  if (length(setdiff(need, names(d))) > 0)
    stop("thread-summary.xlsx is missing: ", paste(setdiff(need, names(d)), collapse = ", "))
  if (!"phase" %in% names(d))
    d$phase <- dplyr::case_when(d$thread_trt == "lab_control" ~ "lab",
                                d$thread_trt == "baseline"    ~ "pre",
                                TRUE                          ~ "post")
  d %>%
    mutate(mussel       = norm_id(mussel),
           adhesion_kpa = max_force / pad_area * 1000) %>%   # recomputed, never trusted from the sheet
    filter(!is.na(adhesion_kpa))
}
```

---

# 1. Build the paired-sample table

Each animal gets one value per metric and timepoint: the **geometric mean** of its plaques
for mean force, area and adhesion (the mean on the log scale the model uses) and its largest
thread peak force for max force; the same summary of its pre-exposure threads is its
baseline. The
Tag-seq treatment assignment (the authoritative one) comes from the sample table written by
`05_differential-expression` (`03_analyses/count_matrix/treatmentinfo_clean.csv`). With `INCLUDE_CONTROL_ARM = TRUE` the day-3 control animals enter as their own
treatment level; the day-0 lab-reference animals never do (they have no day-3 threads).

```{r paired}
## ---- thread mechanics: one value per animal at day 3 and at baseline ------------------
thr_raw <- read_thread_summary(THREAD_FILE)

## per-animal summary of each metric's thread column: the geometric mean (agg = "mean"; the
## mean on the log scale the model uses) or the largest thread (agg = "max")
column_of <- function(m) METRICS$column[match(m, METRICS$metric)]
agg_of    <- function(m) METRICS$agg[match(m, METRICS$metric)]
summarise_animal <- function(d, prefix = "") {
  out <- d %>% group_by(mussel) %>% summarise(n = n(), .groups = "drop")
  for (m in metrics) {
    col <- column_of(m)
    v <- d %>% group_by(mussel) %>%
      summarise(val = if (agg_of(m) == "max") max(.data[[col]], na.rm = TRUE)
                      else exp(mean(log(.data[[col]]), na.rm = TRUE)), .groups = "drop")
    out[[paste0(prefix, m)]] <- v$val[match(out$mussel, v$mussel)]
  }
  names(out)[names(out) == "n"] <- paste0("n_plaques", if (nzchar(prefix)) "_base" else "")
  out
}

arm_of <- thr_raw %>% filter(phase == "post") %>% distinct(mussel, treatment = mussel_trt)
stopifnot(!any(duplicated(arm_of$mussel)))                        # one arm per animal
post_stress <- thr_raw %>% filter(phase == "post") %>% summarise_animal() %>%
  left_join(arm_of, by = "mussel") %>% relocate(treatment, .after = mussel)

## pre-exposure baseline: the same animals' own threads before the exposure began
baseline <- thr_raw %>% filter(phase == "pre") %>% summarise_animal(prefix = "base_")

## ---- Tag-seq treatment assignment (definitive), for the chosen tissue ----------------
## One row per library in the clean count matrix. Foot means the phenol gland to the tip of
## the foot; the day-0 libraries of the rest of the foot (IDs ending FX) are dropped here.
read_treatmentinfo <- function(tissue) {
  if (!file.exists(TI_FILE)) stop("Sample table not found: ", TI_FILE,
                                  "\nRun 05_differential-expression/01_code/01_clean_count_matrix.Rmd.")
  read.csv(TI_FILE, colClasses = "character") %>%
    filter(tissue == .env$tissue, region != "foot (without phenol gland)") %>%   # .env: the file has a `tissue` column
    transmute(sample_id = sample.1,
              mussel    = norm_id(sub(paste0(.env$tissue, "$"), "", sample.1)),
              treatment,
              day       = as.integer(day))
}
arms_kept <- if (INCLUDE_CONTROL_ARM) c("control", "OA", "OW", "DO") else c("OA", "OW", "DO")
fti <- read_treatmentinfo(TISSUE) %>%
  filter(day == 3, treatment %in% arms_kept) %>%       # day-3 animals only
  distinct(mussel, treatment)

## ---- assemble paired table ----------------------------------------------------------
paired <- post_stress %>%
  inner_join(fti, by = c("mussel", "treatment")) %>%   # the join also checks the two
  left_join(baseline, by = "mussel") %>%               # sources agree on the arm
  mutate(
    sample    = paste0(mussel, TISSUE),                # expression column name for chosen tissue
    treatment = factor(treatment, levels = arms_kept)
  )

## Arm disagreement between the thread key and the Tag-seq metadata would drop an animal
## silently at the inner_join above; make that loud.
arm_clash <- post_stress %>%
  inner_join(fti %>% dplyr::rename(rna_treatment = treatment), by = "mussel") %>%
  dplyr::filter(treatment != rna_treatment)
if (nrow(arm_clash) > 0) print(arm_clash)
warn_unless(nrow(arm_clash) == 0, nrow(arm_clash), " animal(s) carry different arms in ",
              "thread-summary and the 05 sample table (", TISSUE, ") and are excluded: ",
              paste(arm_clash$mussel, collapse = ", "))

## ---- per-animal response classification from 02_thread-strength script 05 ------------
## Carried in the manifest for inspection (which animals weakened relative to control);
## it enters no model here.
if (file.exists(RESPONSE_FILE)) {
  resp <- read_csv(RESPONSE_FILE, show_col_types = FALSE) %>%
    mutate(mussel = norm_id(mussel)) %>%
    select(mussel, response_class, response_score)
  paired <- paired %>% left_join(resp, by = "mussel")
} else {
  warn_unless(FALSE, "Response classification not found (", RESPONSE_FILE,
                "); run 02_thread-strength/01_code/05_decompose_adhesion.Rmd. Continuing without it.")
}

## An animal can be in treatmentinfo without a column in the count matrix (a library that
## was dropped at QC). Drop it here, loudly, rather than fail at the VST step.
cm_cols <- colnames(read.csv(COUNT_FILE, nrows = 1, check.names = FALSE))
no_expr <- paired %>% filter(!sample %in% cm_cols)
if (nrow(no_expr) > 0) {
  message(nrow(no_expr), " paired animal(s) have no ", TISSUE,
          " column in the count matrix and are dropped: ",
          paste(no_expr$mussel, collapse = ", "))
  paired <- paired %>% filter(sample %in% cm_cols)
}

## the nuisance terms every regression below controls for; script 02 reads the same list
COVARIATES <- "treatment"

cat("Paired animals:", nrow(paired), "\n")
print(table(paired$treatment))
cat("With baseline threads (the animals in every ANCOVA fit):",
    sum(!is.na(paired$base_mean_force)), "\n")
print(table(paired$treatment[!is.na(paired$base_mean_force)]))
if ("response_class" %in% names(paired)) print(table(paired$treatment, paired$response_class))
write_csv(paired, file.path(OUT_DIR, sprintf("paired_sample_manifest_%s.csv", TISSUE)))

## ---- invariants: the animals in the fits, and agreement with 02_thread-strength -------
## (1) The set of animals that enter every ANCOVA fit (paired, with a baseline) must equal
##     02_data/expected_animals.csv for this tissue. Every other day-3 animal is listed there
##     with the reason it is out, so a wrong join shows up as a
##     difference here instead of silently changing n.
EXPECTED_FILE <- file.path(repo_root, "09_gene-mechanics-correlation", "02_data", "expected_animals.csv")
tissue_col <- ifelse(TISSUE == "F", "foot", "gill")
in_fit <- paired$mussel[!is.na(paired$base_mean_force)]
if (file.exists(EXPECTED_FILE)) {
  expected <- read_csv(EXPECTED_FILE, show_col_types = FALSE)
  exp_fit  <- expected$mussel[expected[[paste0(tissue_col, "_in_fit")]]]
  recon <- tibble(mussel = sort(union(exp_fit, in_fit))) %>%
    mutate(expected = mussel %in% exp_fit, observed = mussel %in% in_fit) %>%
    filter(expected != observed)
  write_csv(recon, file.path(OUT_DIR, sprintf("animal_reconciliation_%s.csv", TISSUE)))
  if (nrow(recon) > 0) print(recon)
  warn_unless(nrow(recon) == 0, "The animals in the ", tissue_col, " fits differ from ",
                "02_data/expected_animals.csv (", nrow(recon), " animal(s); see ",
                "animal_reconciliation_", TISSUE, ".csv). Update the expected file only with a ",
                "documented reason.")
} else {
  warn_unless(FALSE, "Not found: ", EXPECTED_FILE, "; the animal set is not checked.")
}

## (2) Each animal's day-3 and baseline values must equal the ones 02_thread-strength's ANCOVA
##     used (DATA_ancova_animals.csv, scripts 04 and 05), so the two pipelines cannot drift apart.
##     This checks consistency between the codes, not whether the traces are right.
TS_ANIMALS <- file.path(repo_root, "02_thread-strength", "03_analyses",
                        c("04_analyze-thread-strength", "05_decompose-adhesion"), "DATA_ancova_animals.csv")
if (all(file.exists(TS_ANIMALS))) {
  ts_an <- bind_rows(lapply(TS_ANIMALS, read_csv, show_col_types = FALSE)) %>%
    filter(metric %in% metrics) %>% mutate(mussel = norm_id(mussel))
  cmp <- bind_rows(lapply(metrics, function(m) {
    ours <- tibble(mussel  = paired$mussel,
                   gm_y    = on_scale(m, paired[[m]]),
                   gm_base = on_scale(m, paired[[paste0("base_", m)]]))
    ts_an %>% filter(metric == m) %>% inner_join(ours, by = "mussel")
  })) %>%
    filter(!is.na(gm_base)) %>%
    mutate(diff = pmax(abs(y - gm_y), abs(baseline - gm_base)))
  bad <- cmp %>% filter(diff > 1e-8)
  if (nrow(bad) > 0) print(bad %>% select(metric, mussel, y, gm_y, baseline, gm_base))
  warn_unless(nrow(bad) == 0, nrow(bad), " animal x metric value(s) differ from 02_thread-strength's ",
                "DATA_ancova_animals.csv; rerun 02_thread-strength scripts 04 and 05 on the same thread table.")
} else {
  warn_unless(FALSE, "02_thread-strength DATA_ancova_animals.csv not found; run 02_thread-strength ",
                "scripts 04 and 05 first. Cross-pipeline agreement is not checked.")
}
```

---

# 2. Load counts and VST-normalise

Variance-stabilising transformation is estimated across **all** samples of the chosen
tissue (more stable dispersion estimates), then we subset to the paired animals. Gene IDs
in the matrix look like `gene-LOC134696364|LOC134696364`; we key on the `LOC######` id
(`gene_key()`, `tools/gene_ids.R`, shared with 05 and 07).

```{r counts}
cm <- read.csv(COUNT_FILE, row.names = 1, check.names = FALSE)

## keep only the chosen tissue's libraries: IDs ending F (foot, phenol gland to tip) or G
## (gill); "F$" leaves out the day-0 FX libraries of the rest of the foot
tissue_cols <- if (TISSUE == "F") {
  grep("F$", colnames(cm), value = TRUE)
} else {
  grep("G$", colnames(cm), value = TRUE)
}
cm_tissue <- cm[, tissue_cols, drop = FALSE]

## minimal colData (treatment per sample, for dispersion modelling)
mussel_of  <- norm_id(sub(paste0(TISSUE, "$"), "", colnames(cm_tissue)))
all_meta <- read_treatmentinfo(TISSUE)
coldata <- data.frame(mussel = mussel_of) %>%
  left_join(all_meta %>% distinct(mussel, treatment, day), by = "mussel") %>%
  mutate(grp = ifelse(is.na(treatment), "unk", paste0(treatment, "_d", day)))
rownames(coldata) <- colnames(cm_tissue)
coldata$grp <- factor(coldata$grp)

## DESeq2 + VST
dds <- DESeqDataSetFromMatrix(round(as.matrix(cm_tissue)), coldata, design = ~ 1)
dds <- dds[rowSums(counts(dds)) > 0, ]
vsd  <- vst(dds, blind = TRUE)
vmat <- assay(vsd)                        # rownames = full gene IDs, same order as counts(dds)
rownames(vmat) <- gene_key(rownames(vmat))   # -> LOC key (unique across the matrix)

stopifnot(all(paired$sample %in% colnames(vmat)))
vp       <- vmat[,        paired$sample, drop = FALSE]
counts_p <- counts(dds)[, paired$sample, drop = FALSE]   # identical row order to vp
keep <- rowSums(counts_p > MIN_COUNT) >= ceiling(nrow(paired) / 3)
vp <- vp[keep, , drop = FALSE]
## the mitochondrial genome's genes and their copies on unplaced scaffolds are analysed on their
## own (05_differential-expression step 13; script 05 here tests their summed signal), so they
## enter neither gene universe
MT_FILE <- file.path(DE_DIR, "03_analyses", "count_matrix", "mitochondrial_loci.csv")
mt_loci <- read.csv(MT_FILE, stringsAsFactors = FALSE)$LOC_ID
cat("Mitochondrial loci passing the expression filter, left out:", sum(rownames(vp) %in% mt_loci), "\n")
vp <- vp[!rownames(vp) %in% mt_loci, , drop = FALSE]
cat("Genes tested (after expression filter):", nrow(vp), "\n")
if (anyDuplicated(rownames(vp))) stop("duplicate gene keys among tested genes: ",
                                      paste(head(rownames(vp)[duplicated(rownames(vp))]), collapse = ", "))

## detection floor per gene: fraction of paired animals at the VST minimum (zero counts)
floor_val   <- min(vp)
frac_floor  <- rowMeans(abs(vp - floor_val) < 1e-6)
floor_flag_of <- function(ids) {
  fr <- unname(frac_floor[match(ids, rownames(vp))])
  tibble(LOC_ID = ids, frac_at_floor = round(fr, 3),
         floor_flag = dplyr::case_when(fr > FLOOR_EXCLUDE ~ "exclude", fr > FLOOR_CAUTION ~ "caution", TRUE ~ "ok"),
         tested = !(EXCLUDE_FLOOR_GENES & fr > FLOOR_EXCLUDE))
}
```

---

# 3. Annotation map (portable LOC -> protein name)

The map is the genome-wide BLAST table, best UniProt hit per LOC by bitscore
(`CANDIDATE_ANNOTATION = "genome"`), with the hit's `pident` and `evalue` carried along so a
weak annotation is visible wherever a name is printed. The treatment-vs-control DEG tables
(`GOterms_genome/*_sigs_ID.csv`) are read as well: `in_TC_DEG_annotation` marks the genes
they name, and with `CANDIDATE_ANNOTATION = "TC_DEG"` only those (plus the byssal structural
genes) can enter the candidate set.

```{r annot}
## names carried by the TC DEG tables (the CANDIDATE_ANNOTATION = "TC_DEG" universe)
sig_files <- list.files(file.path(DE_DIR, "03_analyses", "DEG_lists", "GOterms_genome"),
                        pattern = "_sigs_ID\\.csv$", full.names = TRUE)
warn_unless(length(sig_files) == 6, "Expected 6 *_sigs_ID.csv files in GOterms_genome/, found ",
              length(sig_files), ".")
annot_deg <- map_dfr(sig_files, ~ {
  d <- tryCatch(read.csv(.x, stringsAsFactors = FALSE), error = function(e) NULL)
  ok <- !is.null(d) && all(c("LOC_ID", "protein_name") %in% names(d))
  warn_unless(ok, "Could not read LOC_ID / protein_name from ", basename(.x), "; skipped.")
  if (!ok) return(NULL)
  d %>% select(LOC_ID, protein_name)
}) %>% distinct(LOC_ID, .keep_all = TRUE) %>% as_tibble()
cat("LOCs named in the TC DEG tables:", nrow(annot_deg), "\n")

## genome-wide map: best hit per LOC
if (!file.exists(MASTER_ANNOT_FILE))
  stop("Master annotation not found: ", MASTER_ANNOT_FILE,
       "\nIt is the source of the annotation map (CANDIDATE_ANNOTATION = \"genome\").")
master <- read_delim(MASTER_ANNOT_FILE, delim = "\t", show_col_types = FALSE,
                     col_select = c("gene", "pident", "evalue", "bitscore", "Protein.names")) %>%
  ## key: the LOC id, or the gene symbol for the 13 mitochondrial genes (ND1, COX1, ...),
  ## which is how the count matrix names them too
  transmute(LOC_ID = na_if(gene_key(trimws(as.character(gene))), ""),
            protein_name = as.character(Protein.names),
            blast_pident = suppressWarnings(as.numeric(pident)),
            blast_evalue = suppressWarnings(as.numeric(evalue)),
            bitscore     = suppressWarnings(as.numeric(bitscore))) %>%
  filter(!is.na(LOC_ID), !is.na(protein_name), nzchar(protein_name)) %>%
  arrange(desc(bitscore)) %>% distinct(LOC_ID, .keep_all = TRUE) %>%           # best hit per LOC
  select(-bitscore)
cat("LOCs with a genome-wide best hit:", nrow(master), "\n")

if (CANDIDATE_ANNOTATION == "genome") {
  annot <- master %>%
    mutate(annotation_source = "genome_blast",
           in_TC_DEG_annotation = LOC_ID %in% annot_deg$LOC_ID)
  cand_ok <- with(annot, blast_evalue <= CANDIDATE_MAX_EVALUE & blast_pident >= CANDIDATE_MIN_PIDENT)
} else if (CANDIDATE_ANNOTATION == "TC_DEG") {
  ## the DEG-table names, plus the byssal structural genes from the genome map (they are not
  ## DEGs in any treatment-vs-control contrast, so the DEG tables never reach them)
  byssal_add <- master %>%
    filter(grepl(BYSSAL_STRUCTURAL_REGEX, protein_name, ignore.case = TRUE),
           LOC_ID %in% rownames(vp), !LOC_ID %in% annot_deg$LOC_ID) %>%
    mutate(annotation_source = "genome_blast_byssal")
  annot <- bind_rows(annot_deg %>% mutate(annotation_source = "TC_DEG_sigs") %>%
                       left_join(master %>% select(LOC_ID, blast_pident, blast_evalue), by = "LOC_ID"),
                     byssal_add) %>%
    mutate(in_TC_DEG_annotation = annotation_source == "TC_DEG_sigs")
  cand_ok <- rep(TRUE, nrow(annot))
} else stop("CANDIDATE_ANNOTATION must be \"genome\" or \"TC_DEG\"")
annot$blast_ok <- !is.na(cand_ok) & cand_ok   # passes the BLAST-quality floor; script 02's modules use it too
cat("Annotation map (", CANDIDATE_ANNOTATION, "):", nrow(annot), "LOCs, of which", sum(annot$blast_ok),
    "pass the BLAST-quality floor\n")

## candidate gene set = annotated, expressed genes whose protein name matches the keyword list
candidates <- annot %>% filter(blast_ok) %>%
  filter(grepl(CANDIDATE_KEYWORDS, protein_name, ignore.case = TRUE)) %>%
  filter(LOC_ID %in% rownames(vp)) %>%
  mutate(byssal_structural = grepl(BYSSAL_STRUCTURAL_REGEX, protein_name, ignore.case = TRUE))
candidates <- candidates %>% left_join(floor_flag_of(candidates$LOC_ID), by = "LOC_ID")
cat("Candidate (byssal/stress) genes present & expressed:", nrow(candidates),
    "( byssal structural:", sum(candidates$byssal_structural),
    "; named in the TC DEG tables:", sum(candidates$in_TC_DEG_annotation),
    "; at the detection floor:", sum(candidates$floor_flag == "exclude"),
    if (EXCLUDE_FLOOR_GENES) "excluded from the tested family" else "kept", ")\n")
cat("Candidate genes tested:", sum(candidates$tested), "\n")
print(as.data.frame(candidates %>% select(LOC_ID, protein_name, blast_pident, in_TC_DEG_annotation,
                                          byssal_structural, frac_at_floor, floor_flag, tested) %>%
                      mutate(protein_name = substr(protein_name, 1, 50))))
```

```{r exports}
## ---- handoff CSVs for downstream scripts (02 reads these; 03 reads annotation_map) ---
## paired_sample_manifest_<T>.csv was written above; these add the rest so 02 can run on its
## own instead of sharing this R session's objects.
write_csv(data.frame(LOC_ID = rownames(vp), vp, check.names = FALSE),
          file.path(OUT_DIR, sprintf("vst_paired_%s.csv", TISSUE)))            # genes x paired samples (VST)
write_csv(annot,      file.path(OUT_DIR, "annotation_map.csv"))                # LOC_ID -> protein_name
write_csv(candidates, file.path(OUT_DIR, sprintf("candidate_genes_%s.csv", TISSUE)))  # candidate set
write_csv(METRICS %>%                                                            # so 02 fits the same
            mutate(arms                = paste(arms_kept, collapse = ";"),        # model on the same
                   covariates          = paste(COVARIATES, collapse = ";"),       # metrics and scales
                   exclude_floor_genes = EXCLUDE_FLOOR_GENES),
          file.path(OUT_DIR, sprintf("metrics_config_%s.csv", TISSUE)))
```

---

# 4. DEG union (data-driven gene universe)

Union of significant genes from all stressor contrasts for this tissue (treatment-control
baseline). These are genes that already responded to stress; the question is whether,
among the paired animals, their expression *level* tracks attachment.

```{r deg_union}
deg_dir <- file.path(DE_DIR, "03_analyses", "DEG_lists", ifelse(TISSUE == "F", "Foot", "Gill"))
deg_files <- file.path(deg_dir, sprintf("%s%s_TC_siggene.csv", TISSUE, c("OA", "OW", "DO")))  # one per stressor
for (f in deg_files) warn_unless(file.exists(f), "Missing DEG list ", basename(f), " in ", deg_dir, ".")
deg_files <- deg_files[file.exists(deg_files)]
deg_union <- map(deg_files, ~ {
  d <- tryCatch(read.csv(.x, stringsAsFactors = FALSE), error = function(e) NULL)
  warn_unless(!is.null(d), "Could not read ", basename(.x), "; its genes are missing from the DEG union.")
  if (is.null(d)) return(character(0))
  g <- d[[grep("gene", names(d), ignore.case = TRUE)[1]]]
  gene_key(g)
}) %>% unlist() %>% unique()
deg_union <- intersect(deg_union, rownames(vp))
deg_union_tbl <- floor_flag_of(deg_union)
cat("DEG-union genes present & expressed:", length(deg_union),
    "| at the detection floor:", sum(deg_union_tbl$floor_flag == "exclude"),
    "| tested:", sum(deg_union_tbl$tested), "\n")

## the floor table for both universes, so downstream scripts and the report can read it
write_csv(bind_rows(candidates %>% transmute(LOC_ID, gene_set = "candidate", frac_at_floor, floor_flag, tested),
                    deg_union_tbl %>% filter(!LOC_ID %in% candidates$LOC_ID) %>%
                      transmute(LOC_ID, gene_set = "DEG_union", frac_at_floor, floor_flag, tested)) %>%
            left_join(annot %>% select(LOC_ID, protein_name), by = "LOC_ID") %>%
            mutate(n_samples = nrow(paired), n_at_floor = round(frac_at_floor * nrow(paired))) %>%
            select(LOC_ID, protein_name, gene_set, n_samples, n_at_floor, frac_at_floor, floor_flag, tested) %>%
            arrange(desc(frac_at_floor), LOC_ID),
          file.path(OUT_DIR, sprintf("detection_floor_flags_%s.csv", TISSUE)))
```

---

# 5. Association testing

One reported test per gene x metric, an analysis of covariance on the animal's day-3 level:

    level_day3 ~ expression + treatment + level_baseline

Every metric enters on the log scale on both sides of the model: the log of the per-animal
geometric mean, or of the largest thread for `max_force` (`scale` column). `slope` is the coefficient on VST
expression with the arm and the animal's own baseline held constant and `p_lm` its t-test;
`baseline_slope` is the coefficient on the baseline. `partial_r` is the same test expressed
as a partial correlation, t / sqrt(t^2 + residual df): the correlation of expression and
day-3 level with arm, state and baseline held constant; its sign and p are those of `slope`.
BH correction is applied **within each gene set x metric**
(`q_lm`) and **within each gene set x tier** (`q_family`); the candidate x primary family is
the declared confirmatory one.

```{r assoc_fns}
## the ANCOVA data for one gene x metric: response and baseline on the metric's model scale,
## complete cases only (animals without baseline threads drop here)
ancova_data <- function(loc, metric) {
  dat <- data.frame(y    = on_scale(metric, paired[[metric]]),
                    x    = as.numeric(vp[match(loc, rownames(vp)), paired$sample]),
                    base = on_scale(metric, paired[[paste0("base_", metric)]]),
                    paired[, COVARIATES, drop = FALSE])
  dat[stats::complete.cases(dat), , drop = FALSE]
}
ANCOVA_FORMULA <- reformulate(c("x", COVARIATES, "base"), response = "y")

test_gene <- function(loc, metric) {
  dat <- ancova_data(loc, metric)
  if (nrow(dat) < 6) return(NULL)
  fm <- tryCatch(lm(ANCOVA_FORMULA, data = dat), error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!all(c("x", "base") %in% rownames(co))) return(NULL)
  tibble(
    LOC_ID         = loc,
    metric         = metric,
    scale          = scale_of(metric),
    tier           = tier_of(metric),
    n              = nrow(dat),
    slope          = co["x", "Estimate"],
    se             = co["x", "Std. Error"],
    p_lm           = co["x", "Pr(>|t|)"],
    baseline_slope = co["base", "Estimate"],
    partial_r      = co["x", "t value"] / sqrt(co["x", "t value"]^2 + fm$df.residual),
    ## marginal 95% CI for the partial correlation (Fisher z, SE 1/sqrt(df.residual - 1));
    ## a per-test interval, not simultaneous over the family
    partial_r_lo   = tanh(atanh(partial_r) - qnorm(0.975) / sqrt(fm$df.residual - 1)),
    partial_r_hi   = tanh(atanh(partial_r) + qnorm(0.975) / sqrt(fm$df.residual - 1))
  )
}

run_set <- function(loci, set_name) {
  res <- map_dfr(metrics, function(mt) map_dfr(loci, ~ test_gene(.x, mt)))
  res %>%
    group_by(metric) %>% mutate(q_lm = p.adjust(p_lm, "BH")) %>% ungroup() %>%      # within metric
    group_by(tier)   %>% mutate(q_family = p.adjust(p_lm, "BH")) %>% ungroup() %>%  # within tier family
    left_join(annot, by = "LOC_ID") %>%
    mutate(gene_set = set_name) %>%
    arrange(factor(metric, levels = metrics), p_lm)
}
```

```{r run}
res_cand <- run_set(candidates$LOC_ID[candidates$tested], "candidate") %>%
  left_join(candidates %>% select(LOC_ID, frac_at_floor, floor_flag, byssal_structural), by = "LOC_ID")
res_deg  <- run_set(deg_union_tbl$LOC_ID[deg_union_tbl$tested], "DEG_union") %>%
  left_join(deg_union_tbl %>% select(LOC_ID, frac_at_floor, floor_flag), by = "LOC_ID")

write_csv(res_cand, file.path(OUT_DIR, sprintf("assoc_candidate_%s.csv", TISSUE)))
write_csv(res_deg,  file.path(OUT_DIR, sprintf("assoc_DEGunion_%s.csv",  TISSUE)))

cat("\n================ CANDIDATE GENES: top associations per metric ================\n")
res_cand %>%
  mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_lm, n = 6) %>% ungroup() %>%
  mutate(across(c(slope, baseline_slope, partial_r), ~round(., 3)),
         across(c(p_lm, q_lm, q_family), ~signif(., 2))) %>%
  select(metric, tier, n, LOC_ID, protein_name, slope, p_lm, q_lm, q_family, partial_r) %>%
  print(n = 24)

## the declared confirmatory family: candidate genes x primary metrics
fam <- res_cand %>% filter(tier == "primary")
cat(sprintf("\nCandidate x primary family (%s): %d tests, %d with q_family < %s; smallest p = %s\n",
            paste(PRIMARY_METRICS, collapse = ", "), nrow(fam), sum(fam$q_family < FDR_ALPHA),
            FDR_ALPHA, signif(min(fam$p_lm), 2)))

cat("\n================ DEG-UNION: genes passing BH <", FDR_ALPHA, "================\n")
hits <- res_deg %>% filter(q_lm < FDR_ALPHA) %>% arrange(factor(metric, levels = metrics), p_lm)
if (nrow(hits) == 0) {
  cat("None survive FDR at this sample size (expected). Inspect raw-p ranking below.\n")
  res_deg %>% mutate(metric = factor(metric, levels = metrics)) %>%
    group_by(metric) %>% slice_min(p_lm, n = 5) %>% ungroup() %>%
    mutate(across(c(slope, partial_r), ~round(., 3)), across(c(p_lm, q_lm, q_family), ~signif(., 2))) %>%
    select(metric, LOC_ID, protein_name, slope, p_lm, q_lm, q_family, partial_r) %>%
    as.data.frame() %>% head(24) %>% print()
} else {
  print(hits, n = nrow(hits))
}
```

---

# 6. Plots

## 6.1 Candidate-gene correlation heatmap

ANCOVA partial correlation (`partial_r`: arm and baseline held constant) of candidate genes
against each metric's day-3 level. With the genome-wide candidate set the panel shows the
`HEATMAP_MAX_GENES` genes with the strongest association on any metric (smallest `p_lm`);
the full table is `assoc_candidate_<T>.csv`.

```{r heatmap, fig.height=7}
HEATMAP_MAX_GENES <- 60
hm_keep <- res_cand %>% group_by(LOC_ID) %>% summarise(best_p = min(p_lm), .groups = "drop") %>%
  slice_min(best_p, n = HEATMAP_MAX_GENES, with_ties = FALSE) %>% pull(LOC_ID)
hm <- res_cand %>%
  filter(LOC_ID %in% hm_keep) %>%
  mutate(label = paste0(short_name(protein_name, 34), " (", LOC_ID, ")")) %>%
  select(label, metric, partial_r) %>%
  pivot_wider(names_from = metric, values_from = partial_r) %>%
  arrange(desc(abs(mean_force)))

hm_long <- hm %>% pivot_longer(-label, names_to = "metric", values_to = "r") %>%
  mutate(metric = factor(metric, levels = metrics))
short_label <- function(m) sub(" \\(.*\\)$", "", label_of(m))    # "Plaque area (mm^2)" -> "Plaque area"
ggplot(hm_long, aes(metric, factor(label, levels = rev(hm$label)), fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = DIRECTION_COLORS[["down"]], mid = "white", high = DIRECTION_COLORS[["up"]],
                       midpoint = 0, limits = c(-1, 1)) +   # the repository's direction colours
  scale_x_discrete(labels = short_label, position = "top") +
  labs(x = NULL, y = NULL, fill = "partial r\n(ANCOVA)",
       title = "Candidate gene expression vs plaque mechanics",
       subtitle = sprintf("%s, n = %d animals with baseline; top %d of %d candidate genes selected by smallest p (effects biased away from zero)",
                          ifelse(TISSUE == "F", "foot", "gill"), sum(!is.na(paired$base_mean_force)),
                          nrow(hm), nrow(candidates))) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 9, angle = 20, hjust = 0),
        plot.title.position = "plot")
ggsave(file.path(OUT_DIR, sprintf("candidate_heatmap_%s.png", TISSUE)),
       width = 7.5, height = 8.5, dpi = 150, bg = "white")   # theme_minimal has no background: force one
```

## 6.2 Added-variable plots for the strongest candidate hits (vs mean peak force)

Each panel shows exactly what the ANCOVA tests. The day-3 level (y) and expression (x) are
each residualised on arm and baseline, and the line through the origin has the tested slope
(the Frisch-Waugh-Lovell identity: the slope of the residuals equals the ANCOVA coefficient).
Points are coloured by arm. The genes are **selected by smallest p**, so their effects are
biased away from zero; the panels show the fit, not independent evidence.

```{r scatter, fig.height=8}
## residuals of y and of expression on the nuisance terms (arm [+ state] + baseline)
av_data <- function(loc, metric) {
  d  <- ancova_data(loc, metric)
  nf <- reformulate(c(COVARIATES, "base"))
  tibble(x_res     = residuals(lm(update(nf, x ~ .), data = d)),
         y_res     = residuals(lm(update(nf, y ~ .), data = d)),
         treatment = d$treatment)
}
av_label <- function(metric)
  sprintf("%s%s | arm, baseline", if (scale_of(metric) == "log") "log " else "", label_of(metric))

top_hits <- res_cand %>% filter(metric == "mean_force") %>%
  slice_min(p_lm, n = 6, with_ties = FALSE)

plot_df <- pmap_dfr(top_hits %>% select(LOC_ID, slope, p_lm), function(LOC_ID, slope, p_lm) {
  av_data(LOC_ID, "mean_force") %>%
    mutate(facet = sprintf("%s\n%s, p = %s", short_name(annot$protein_name[match(LOC_ID, annot$LOC_ID)], 34),
                           LOC_ID, signif(p_lm, 2)),
           slope = slope)
}) %>% mutate(facet = factor(facet, levels = unique(facet)))

arm_cols <- TREATMENT_COLORS      # the repository's treatment colours; shared by 6.2 and 6.3
ggplot(plot_df, aes(x_res, y_res)) +
  geom_hline(yintercept = 0, color = "grey85") + geom_vline(xintercept = 0, color = "grey85") +
  geom_abline(aes(intercept = 0, slope = slope), color = "grey30", linewidth = 0.7) +
  geom_point(aes(fill = treatment), shape = 21, size = 2.4, stroke = 0.5, color = "black") +
  facet_wrap(~ facet, scales = "free") +
  scale_fill_manual(values = arm_cols, name = "Arm") +
  labs(x = "VST expression | arm, baseline", y = av_label("mean_force"),
       title = "Top candidate genes vs mean peak force: added-variable plots",
       subtitle = "Selected by smallest p (effects biased away from zero); line = tested ANCOVA slope") +
  theme_bw(base_size = 11)
ggsave(file.path(OUT_DIR, sprintf("top_candidate_scatter_%s.png", TISSUE)),
       width = 9, height = 7, dpi = 150, bg = "white")
```

## 6.3 The strongest candidate hit for each metric, added-variable plots

One panel per metric: the single best candidate gene for that metric, drawn as in 6.2.
Colouring by arm shows whether one arm carries the within-arm slope; the arm-mean shifts
themselves are removed by the model, so they cannot appear here.

```{r scatter_best, fig.height=8}
best_per_metric <- res_cand %>%
  mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  arrange(metric)

best_df <- pmap_dfr(best_per_metric %>% select(LOC_ID, metric, slope, p_lm, q_lm, q_family),
                    function(LOC_ID, metric, slope, p_lm, q_lm, q_family) {
  m <- as.character(metric)
  av_data(LOC_ID, m) %>%
    mutate(panel = sprintf("%s (%s)\n%s\np = %s, q = %s, q_family = %s", av_label(m), tier_of(m),
                           short_name(annot$protein_name[match(LOC_ID, annot$LOC_ID)], 34),
                           signif(p_lm, 2), signif(q_lm, 2), signif(q_family, 2)),
           slope = slope)
}) %>% mutate(panel = factor(panel, levels = unique(panel)))

ggplot(best_df, aes(x_res, y_res)) +
  geom_hline(yintercept = 0, color = "grey85") + geom_vline(xintercept = 0, color = "grey85") +
  geom_abline(aes(intercept = 0, slope = slope), color = "grey30", linewidth = 0.7) +
  geom_point(aes(fill = treatment), shape = 21, size = 2.6, stroke = 0.5, color = "black") +
  facet_wrap(~ panel, scales = "free", ncol = 2) +
  scale_fill_manual(values = arm_cols, name = "Arm") +
  labs(x = "VST expression | arm, baseline", y = "Day-3 level | arm, baseline (model scale)",
       title = sprintf("Best candidate gene per metric (%s, n = %d with baseline): added-variable plots",
                       ifelse(TISSUE == "F", "foot", "gill"), sum(!is.na(paired$base_mean_force))),
       subtitle = "Selected by smallest p (effects biased away from zero); line = tested ANCOVA slope") +
  theme_bw(base_size = 10) + theme(strip.text = element_text(size = 7.5))
ggsave(file.path(OUT_DIR, sprintf("best_hit_per_metric_scatter_%s.png", TISSUE)),
       width = 9, height = 8.5, dpi = 150, bg = "white")
```

---

# 7. Session info & interpretation notes

```{r notes, echo=FALSE, results='asis'}
cat("
**How to read the outputs**

- `assoc_candidate_*.csv` / `assoc_DEGunion_*.csv`: one row per gene x metric from the
  ANCOVA `level_day3 ~ expression + treatment + level_baseline`.
  `slope` (with `se`) is the change in the day-3 level per unit VST expression with the arm
  and the animal's own baseline held constant, in log units, so exp(slope) is the
  multiplicative change per VST unit. `p_lm` is its
  t-test; `q_lm` is BH within the gene set x metric and `q_family` BH within the gene set x
  tier (candidate x primary is the declared confirmatory family); `baseline_slope` is the
  coefficient on the baseline level; `partial_r` is the same test as a partial correlation,
  with a marginal 95% interval (`partial_r_lo`, `partial_r_hi`).
- A negative slope/partial_r means higher expression goes with *weaker* attachment.
- Any table or figure that shows the genes with the smallest p is a selection: their
  effects are biased away from zero. Read the intervals and the family q, not the ranking.

**Caveats baked into the design (do not skip in the manuscript):**

1. Only the paired animals with baseline threads are in the fits (listed with reasons in
   `02_data/expected_animals.csv`), and within-arm n is small. The treatment term removes
   arm-mean shifts in expression and attachment, so the slope is a within-arm association,
   pooled over arms under a common-slope assumption.
2. Multiple testing: the candidate set (the keyword matches under the genome-wide map) x
   PRIMARY_METRICS is the declared confirmatory family (`tier = primary`,
   `q_family`); the remaining metrics and the DEG-union (hundreds) are exploratory /
   screening. Quote q beside p, never p alone.
3. Baseline adjustment: the covariate is the animal's own pre-exposure level, the same
   summary on the same scale. Unlike a change score it adds no baseline noise to the
   response, but the baseline is measured with error, so the adjustment is only partial
   (Frost and Thompson 2000, J R Stat Soc A 163:173-189). Baseline by future arm is
   described in 02_thread-strength script 04 (`DESC_baseline_by_arm.csv`). Animals without
   baseline threads are not in the fits.
4. Timing: tissue was sampled at the end of the exposure, after the tested day-3 threads
   were secreted. The association is not predictive or causal; it may reflect a persistent
   individual state, a lag, reverse direction or a common cause.
5. `mean_force` and `pad_area` are the declared primary metrics; `adhesion_kpa` is their
   ratio, and a response in plaque area alone can be invisible in it. Read force and area
   first. `max_force`, the animal's strongest thread, rises with the number of threads
   pulled; at baseline some animals have only one or two.
6. With `INCLUDE_CONTROL_ARM = TRUE` the control animals are a fourth stratum of the same
   pooled slope, not a null that a stress signal must beat. With 10 to 12 animals per arm
   the design cannot tell a stress-specific association from a general one.
")
sessionInfo()
```

# 8. Provenance

`RUN_provenance_<T>.txt` records the configuration behind every file in
`03_analyses/gene_mechanics/`: the settings, the code commit that ran and whether any tracked
file differed from it, the R and package versions, and an MD5 of every input, all with
repository-relative paths. Script 02 appends its own block.

```{r provenance}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
fmt_tab <- function(v) paste(sprintf("%s %d", names(table(v)), as.integer(table(v))), collapse = ", ")
settings <- c(
  "[01_gene_mechanics_correlation.Rmd]",
  paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
  paste0("tissue               : ", TISSUE),
  paste0("INCLUDE_CONTROL_ARM  : ", INCLUDE_CONTROL_ARM),
  paste0("reported test        : ANCOVA level_day3 ~ expression + ", paste(COVARIATES, collapse = " + "),
         " + level_baseline (log scale; per-animal geometric mean, largest thread for max_force)"),
  paste0("metrics [scale, tier]: ", paste(sprintf("%s [%s, %s]", METRICS$metric, METRICS$scale, METRICS$tier),
                                          collapse = ", ")),
  paste0("PRIMARY_METRICS      : ", paste(PRIMARY_METRICS, collapse = ", ")),
  paste0("MIN_COUNT / FDR_ALPHA: ", MIN_COUNT, " / ", FDR_ALPHA),
  paste0("paired animals       : ", nrow(paired), " (", fmt_tab(paired$treatment), ")"),
  paste0("in the fits          : ", sum(!is.na(paired$base_mean_force)), " (",
         fmt_tab(paired$treatment[!is.na(paired$base_mean_force)]), ")"),
  paste0("genes passing expression filter: ", nrow(vp)),
  paste0("CANDIDATE_ANNOTATION : ", CANDIDATE_ANNOTATION, " (evalue <= ", CANDIDATE_MAX_EVALUE,
         ", pident >= ", CANDIDATE_MIN_PIDENT, ")"),
  paste0("candidate genes      : ", nrow(candidates), " (byssal structural ", sum(candidates$byssal_structural),
         "; named in TC DEG tables ", sum(candidates$in_TC_DEG_annotation), ")"),
  paste0("EXCLUDE_FLOOR_GENES  : ", EXCLUDE_FLOOR_GENES, " (exclude > ", FLOOR_EXCLUDE, ", caution > ", FLOOR_CAUTION, ")"),
  paste0("candidates tested    : ", sum(candidates$tested), " (", sum(!candidates$tested), " at the detection floor)"),
  paste0("DEG union tested     : ", sum(deg_union_tbl$tested), " of ", nrow(deg_union_tbl)),
  paste0("family sizes (tests) : candidate x primary ", sum(res_cand$tier == "primary"),
         ", candidate x exploratory ", sum(res_cand$tier == "exploratory"),
         ", DEG union ", nrow(res_deg)))
inputs <- c(THREAD_FILE, RESPONSE_FILE, EXPECTED_FILE, TS_ANIMALS,
            COUNT_FILE, TI_FILE, MT_FILE,
            MASTER_ANNOT_FILE, sig_files, deg_files,
            file.path(repo_root, "09_gene-mechanics-correlation", "01_code", "01_gene_mechanics_correlation.Rmd"),
            file.path(repo_root, "tools", c("pipeline_checks.R", "gene_ids.R")))
writeLines(provenance_lines(settings, inputs, repo_root, packages = c("DESeq2", "dplyr", "readr")), prov)
cat(readLines(prov), sep = "\n")
```
