---
title: "02 - Gene x mechanics: expanded analysis (modules, diagnostics)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"   # "F" foot (default) or "G" gill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# What this adds to script 01

Script 01 fits the reported test, one baseline-adjusted regression (ANCOVA) per gene x
metric on the per-animal values:

    level_day3 ~ expression + treatment + level_baseline

with every metric on the log scale (per-animal geometric means of the plaques, or the largest
thread for `max_force`). The animal is the unit, so there is no pseudoreplication,
but every animal counts the same whether it built one plaque or five. This script keeps
that test as the reported one and adds what a reader needs to trust or discount a hit:

- **A. Module eigengenes** - collapse co-regulated pathways (HIF, oxidative, HSP,
  tRNA-synthetase, byssal/collagen, and the byssal structural module) into one score each
  and put the score through the same ANCOVA, cutting the multiple-testing burden from
  hundreds of genes to six modules.
- **B. Diagnostics** - echoes the detection-floor flags, refits the reported ANCOVA for the
  best hits without their most influential animal (Cook's distance), and writes
  `best_hits_<T>.csv`: one row per metric and gene set with the reported p, both q's, the
  floor flag and the influence flag side by side.

Every test here is the per-animal ANCOVA script 01 fits; nothing else is fitted.

> **Prerequisite:** run script 01 first. This script does not share 01's R session; it
> reads 01's CSV handoffs (`paired_sample_manifest_<T>`, `vst_paired_<T>`, `annotation_map`,
> `candidate_genes_<T>`, `metrics_config_<T>`,
> `assoc_candidate_<T>`, `assoc_DEGunion_<T>`, `detection_floor_flags_<T>`) from
> `03_analyses/gene_mechanics/`.

```{r setup_inputs}
suppressPackageStartupMessages({ library(tidyverse); library(here) })

## ---- config -------------------------------------------------------------------------
## Tissue comes from the knit parameter (`params$tissue`, default "F") so the same script
## serves foot and gill; the driver 00_run_gene_mechanics_by_tissue.Rmd renders both.
## When run chunk-by-chunk outside a knit, `params` may not exist; fall back to foot.
TISSUE <- if (exists("params") && !is.null(params$tissue)) toupper(params$tissue) else "F"
stopifnot(TISSUE %in% c("F", "G"))
FDR_ALPHA        <- 0.10
N_INFLUENCE_HITS <- 3     # candidate hits per metric that get the leave-one-animal-out refit
## The detection-floor filter (FLOOR_EXCLUDE / EXCLUDE_FLOOR_GENES) lives in script 01, which
## writes detection_floor_flags_<T>.csv and marks each candidate `tested`; this script works
## on the same family script 01 tested.

## The metrics, their model scale (log / raw) and tier (primary / exploratory), the treatment
## levels and the nuisance covariates are read from script 01's metrics_config handoff, so
## the two scripts fit the same model on the same animals.

## ---- read script 01's CSV handoffs (run 01 first) ----------------------------------
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("05_differential-expression", "02_thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing 05_differential-expression/ and 02_thread-strength/.")
}
repo_root <- find_repo_root()
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # warn_unless(), provenance
IN_DIR  <- file.path(repo_root, "09_gene-mechanics-correlation", "03_analyses", "gene_mechanics")
OUT_DIR <- IN_DIR                                  # 02 writes alongside 01's outputs
need <- function(f) { p <- file.path(IN_DIR, f)
  if (!file.exists(p)) stop("Missing ", f, " - run script 01 first (it writes this)."); p }

mcfg <- read_csv(need(sprintf("metrics_config_%s.csv", TISSUE)), show_col_types = FALSE)
if (!all(c("metric", "scale", "tier", "arms", "covariates") %in% names(mcfg)))
  stop("metrics_config_", TISSUE, ".csv predates the ANCOVA version of script 01; re-run script 01.")
metrics         <- mcfg$metric
PRIMARY_METRICS <- mcfg$metric[mcfg$tier == "primary"]
arms_kept       <- strsplit(mcfg$arms[1], ";", fixed = TRUE)[[1]]
COVARIATES      <- strsplit(mcfg$covariates[1], ";", fixed = TRUE)[[1]]   # "treatment"
scale_of <- function(m) mcfg$scale[match(m, mcfg$metric)]
tier_of  <- function(m) mcfg$tier[match(m, mcfg$metric)]
on_scale <- function(m, v) if (scale_of(m) == "log") log(v) else v    # a value on its model scale

paired <- read_csv(need(sprintf("paired_sample_manifest_%s.csv", TISSUE)), show_col_types = FALSE) %>%
  mutate(treatment = factor(treatment, levels = arms_kept))
candidates_all <- read_csv(need(sprintf("candidate_genes_%s.csv", TISSUE)), show_col_types = FALSE)
candidates <- candidates_all %>% filter(tested)
floor_tbl  <- read_csv(need(sprintf("detection_floor_flags_%s.csv", TISSUE)), show_col_types = FALSE) %>%
  select(LOC_ID, frac_at_floor, floor_flag)
annot      <- read_csv(need("annotation_map.csv"), show_col_types = FALSE)
vp_df <- read_csv(need(sprintf("vst_paired_%s.csv", TISSUE)), show_col_types = FALSE)
vp <- as.matrix(vp_df[, -1]); rownames(vp) <- vp_df[[1]]   # LOC_ID column -> matrix rownames
stopifnot(all(paired$sample %in% colnames(vp)))
vp <- vp[, paired$sample, drop = FALSE]                      # column order = manifest order
expr_of <- function(loc) as.numeric(vp[match(loc, rownames(vp)), paired$sample])

## script 01's reported test, one row per gene x metric, for the two gene universes
res_cand <- read_csv(need(sprintf("assoc_candidate_%s.csv", TISSUE)), show_col_types = FALSE)
if (!all(c("p_lm", "q_lm", "q_family", "baseline_slope") %in% names(res_cand)))
  stop("assoc_candidate_", TISSUE, ".csv predates the ANCOVA version of script 01; re-run script 01.")
deg_union_file <- file.path(IN_DIR, sprintf("assoc_DEGunion_%s.csv", TISSUE))
warn_unless(file.exists(deg_union_file), "Missing ", basename(deg_union_file),
            "; the DEG-union rows of the influence and best-hits tables are skipped. Run script 01.")
res_deg <- if (file.exists(deg_union_file)) read_csv(deg_union_file, show_col_types = FALSE) else NULL

## ---- the reported ANCOVA, exactly as script 01 fits it --------------------------------
## y and base on the metric's model scale; complete cases only (animals without baseline
## threads drop here); `mussel` rides along so the influence block can name an animal.
ancova_data <- function(x, metric) {          # x = one value per animal, manifest order
  dat <- data.frame(y      = on_scale(metric, paired[[metric]]),
                    x      = x,
                    base   = on_scale(metric, paired[[paste0("base_", metric)]]),
                    mussel = paired$mussel,
                    paired[, COVARIATES, drop = FALSE])
  dat[stats::complete.cases(dat) & is.finite(dat$y) & is.finite(dat$base), , drop = FALSE]
}
ANCOVA_FORMULA <- reformulate(c("x", COVARIATES, "base"), response = "y")
fit_ancova <- function(x, metric) {
  dat <- ancova_data(x, metric)
  if (nrow(dat) < 6) return(NULL)
  fm <- tryCatch(lm(ANCOVA_FORMULA, data = dat), error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!all(c("x", "base") %in% rownames(co))) return(NULL)
  list(fit = fm, data = dat, n = nrow(dat),
       slope = co["x", "Estimate"], se = co["x", "Std. Error"], p = co["x", "Pr(>|t|)"])
}

cat("Paired animals:", nrow(paired), "| in the fits (with baseline):", sum(!is.na(paired$base_mean_force)),
    "| arms:", paste(arms_kept, collapse = ", "),
    "\nMetrics [scale, tier]:", paste(sprintf("%s [%s, %s]", mcfg$metric, mcfg$scale, mcfg$tier), collapse = ", "),
    "\nNuisance covariates:", paste(COVARIATES, collapse = " + "),
    "| candidates tested:", nrow(candidates), "of", nrow(candidates_all),
    "| DEG union tested:", if (is.null(res_deg)) 0L else dplyr::n_distinct(res_deg$LOC_ID), "\n")
```

---

# A. Module eigengenes (power-preserving)

Each module = PC1 of its members' VST values (per mussel), oriented so higher score = higher
module expression. Six modules x four metrics is FDR-tractable, and a coordinated pathway
score is biologically closer to "is the HIF response engaged" than any single gene. The
score goes through the same ANCOVA as a single gene.

`byssal_structural` is the plaque/thread protein module (foot proteins, preCols, byssal
EP/ACDC): the structural proteins of the plaque and thread.
It is kept separate from the broader `byssal_collagen` regex (which also catches every
collagen and every "byss*" annotation) so that the direct test of "these genes track thread
building, not strength" is its own row.

```{r modules}
modules <- list(
  HIF_hypoxia       = "hypoxia-inducible|Egl nine|prolyl 4-hydroxylase|prolyl hydroxylase|hypoxia up-regulated",
  oxidative_detox   = "superoxide|glutathione|catalase|peroxiredoxin|dual oxidase|peroxidase",
  HSP_proteostasis  = "heat shock|Hsp|stress-70|chaperone|BiP|peptidyl-prolyl",
  tRNA_translation  = "--tRNA|tRNA ligase|aminoacyl",
  byssal_collagen   = "collagen|byss|foot protein|adhesive plaque|polyphenolic|thread matrix protein|YGH-rich protein|protease inhibitor-like protein-1|C1q-domain-containing protein-1|TSP_1 domain containing protein-1",
  byssal_structural = "foot protein|adhesive plaque|polyphenolic|byssal protein-|preCol|precollagen|byssal EP|ACDC|thread matrix protein|nongradient byssal|YGH-rich protein"
)

if (!"blast_ok" %in% names(annot)) annot$blast_ok <- TRUE   # older script-20 handoff
module_score <- function(rx) {
  locs <- annot %>% filter(blast_ok, grepl(rx, protein_name, ignore.case = TRUE)) %>% pull(LOC_ID)
  locs <- intersect(locs, rownames(vp))
  if (length(locs) < 3) return(NULL)
  z <- t(scale(t(vp[locs, paired$sample, drop = FALSE])))   # standardise each gene
  z <- z[stats::complete.cases(z), , drop = FALSE]
  pc <- prcomp(t(z), center = TRUE, scale. = FALSE)
  sc <- pc$x[, 1]
  if (suppressWarnings(cor(sc, colMeans(vp[locs, paired$sample]))) < 0) sc <- -sc
  list(score = sc, n_genes = length(locs), members = locs,
       var_expl = summary(pc)$importance[2, 1])
}
mod_scores <- purrr::compact(map(modules, module_score))
write_csv(map_dfr(names(mod_scores), ~ tibble(module = .x, LOC_ID = mod_scores[[.x]]$members)) %>%
            left_join(annot, by = "LOC_ID"),
          file.path(OUT_DIR, sprintf("module_members_%s.csv", TISSUE)))

## the same ANCOVA as script 01, on the module score
mod_fits <- imap_dfr(mod_scores, function(ms, nm) {
  map_dfr(metrics, function(mt) {
    a <- fit_ancova(ms$score, mt)
    if (is.null(a)) return(NULL)
    tibble(module = nm, n_genes = ms$n_genes, pc1_var = round(ms$var_expl, 2),
           metric = mt, scale = scale_of(mt), tier = tier_of(mt), n = a$n,
           slope = a$slope, se = a$se, p_lm = a$p)
  })
}) %>%
  group_by(metric) %>% mutate(q_lm = p.adjust(p_lm, "BH")) %>% ungroup() %>%
  group_by(tier)   %>% mutate(q_family = p.adjust(p_lm, "BH")) %>% ungroup() %>%
  arrange(p_lm)
write_csv(mod_fits, file.path(OUT_DIR, sprintf("module_associations_%s.csv", TISSUE)))

cat("MODULE-level associations (ANCOVA on the module score; q_lm = BH across the ",
    n_distinct(mod_fits$module), " modules within each metric, q_family = BH within each tier):\n", sep = "")
mod_fits %>% mutate(across(c(slope), ~round(., 4)),
                    across(c(p_lm, q_lm, q_family), ~signif(., 2))) %>%
  select(module, n_genes, pc1_var, metric, tier, n, slope, p_lm, q_lm, q_family) %>%
  as.data.frame() %>% print()
```

---

# B. Diagnostics: detection floor and influence; the best-hits table

Genes where many paired samples sit at the VST minimum are effectively absent/zero-inflated;
their correlations are driven by detection, not biology. Script 01 flags them and, by
default, keeps them out of the tested family; the flags are echoed here. Also report the
most influential animal per top hit (Cook's distance) so you know if one mussel drives a fit.

The tables are written to `03_analyses/gene_mechanics/` so the report and any downstream
filter can use them without re-running this script:

- `detection_floor_flags_<T>.csv` (written by script 01): one row per gene in the candidate
  set and the DEG union, with the fraction of paired samples at the VST floor, a
  `floor_flag` of `exclude` (> 0.4), `caution` (> 0.2) or `ok`, and `tested` (FALSE for the
  `exclude` genes when script 01's `EXCLUDE_FLOOR_GENES` is TRUE: they are out of the family
  before BH and appear in no association table).
- `influence_top_hits_<T>.csv`: the three best candidate hits and the best DEG-union hit per
  metric (ranked by the reported `p_lm`) with the most influential animal, its Cook's
  distance, the slope and p after dropping that animal, and an `influence_flag` (`fragile`
  when the hit loses p < 0.05 or its slope moves by more than half).
- `best_hits_<T>.csv`: the single best hit per metric for each gene set (candidate, module,
  DEG union), with `p_lm`, `q_lm`, `q_family`, the floor flag and the influence flag in one
  row. Nothing should be quoted from a p alone.

```{r diagnostics}
## detection floor: computed and applied by script 01 (detection_floor_flags_<T>.csv); shown
## here so the report carries it beside the influence table
floor_flag <- read_csv(need(sprintf("detection_floor_flags_%s.csv", TISSUE)), show_col_types = FALSE)
cat("Detection-floor summary (", nrow(floor_flag), " genes in the two universes; `tested` = in the family):\n", sep = "")
print(table(gene_set = floor_flag$gene_set, floor_flag = floor_flag$floor_flag, tested = floor_flag$tested))

## influence: the best hits per metric, which mussel has the largest Cook's distance in the
## reported ANCOVA, and what the fit looks like without it
top_hits <- bind_rows(
  res_cand %>% group_by(metric) %>% slice_min(p_lm, n = N_INFLUENCE_HITS, with_ties = FALSE) %>% ungroup() %>%
    transmute(gene_set = "candidate", metric, LOC_ID, p_lm, q_lm, q_family),
  if (!is.null(res_deg)) res_deg %>% group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
    transmute(gene_set = "DEG_union", metric, LOC_ID, p_lm, q_lm, q_family))

infl <- pmap_dfr(top_hits, function(gene_set, metric, LOC_ID, p_lm, q_lm, q_family) {
  a <- fit_ancova(expr_of(LOC_ID), metric)
  if (is.null(a)) return(NULL)
  cd   <- cooks.distance(a$fit)
  drop <- which.max(cd)
  fm_drop <- lm(ANCOVA_FORMULA, data = a$data[-drop, ])
  b_all <- a$slope; b_drop <- unname(coef(fm_drop)["x"])
  p_drop <- summary(fm_drop)$coefficients["x", "Pr(>|t|)"]
  tibble(gene_set, metric, LOC_ID,
         p_lm = signif(p_lm, 3), q_lm = signif(q_lm, 3), q_family = signif(q_family, 3),
         n                       = a$n,
         most_influential_mussel = a$data$mussel[drop],
         max_cooks_D             = round(max(cd), 3),
         cook_threshold_4n       = round(4 / a$n, 3),
         ## the max of n Cook's distances is expected to exceed 4/n in most fits of this
         ## size, so that screen alone is weak; D > 1 is the conventional "one point moves
         ## the fit" level (Cook & Weisberg 1982)
         cooks_flag              = case_when(max(cd) > 1       ~ "D>1",
                                             max(cd) > 4 / a$n ~ "D>4/n",
                                             TRUE              ~ "ok"),
         slope                   = signif(b_all, 4),
         slope_without_mussel    = signif(b_drop, 4),
         p_without_mussel        = signif(p_drop, 3),
         slope_change_frac       = round(abs(b_drop - b_all) / abs(b_all), 3),
         ## "fragile": the result depends on one animal (loses p < 0.05, or the slope moves
         ## by more than half) - report such a hit with the leave-one-out slope beside it
         influence_flag          = ifelse((p_lm < 0.05 & p_drop >= 0.05) |
                                            abs(b_drop - b_all) / abs(b_all) > 0.5,
                                          "fragile", "robust"))
}) %>%
  left_join(annot %>% select(LOC_ID, protein_name, blast_pident), by = "LOC_ID") %>%
  mutate(tier = tier_of(metric),
         metric = factor(metric, levels = metrics),
         gene_set = factor(gene_set, levels = c("candidate", "DEG_union"))) %>%
  arrange(gene_set, metric, p_lm) %>%
  mutate(metric = as.character(metric), gene_set = as.character(gene_set)) %>%
  select(gene_set, metric, tier, LOC_ID, protein_name, n, p_lm, q_lm, q_family, everything())
write_csv(infl, file.path(OUT_DIR, sprintf("influence_top_hits_%s.csv", TISSUE)))

cat("\nMost influential animal for the top", N_INFLUENCE_HITS, "candidate hits and the best DEG-union hit per metric",
    "('fragile' = loses p < 0.05 or slope moves > 50% when that animal is dropped):\n")
infl %>% select(gene_set, metric, LOC_ID, p_lm, q_lm, most_influential_mussel, max_cooks_D,
                cooks_flag, slope, slope_without_mussel, p_without_mussel, influence_flag) %>%
  as.data.frame() %>% print()

## ---- best hit per metric and gene set, everything on one row --------------------------
best_cand <- res_cand %>% group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  transmute(gene_set = "candidate", metric, tier, feature = LOC_ID, protein_name,
            n, slope, p_lm, q_lm, q_family)
best_mod <- mod_fits %>% group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  transmute(gene_set = "module", metric, tier, feature = module,
            protein_name = paste0(n_genes, "-gene module"),
            n, slope, p_lm, q_lm, q_family)
best_deg <- if (!is.null(res_deg)) res_deg %>% group_by(metric) %>%
  slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  transmute(gene_set = "DEG_union", metric, tier, feature = LOC_ID, protein_name,
            n, slope, p_lm, q_lm, q_family) else NULL
best_hits <- bind_rows(best_cand, best_mod, best_deg) %>%
  left_join(floor_tbl %>% select(feature = LOC_ID, floor_flag), by = "feature") %>%
  left_join(infl %>% select(feature = LOC_ID, metric, most_influential_mussel, influence_flag),
            by = c("feature", "metric")) %>%
  mutate(metric = factor(metric, levels = metrics),
         gene_set = factor(gene_set, levels = c("candidate", "module", "DEG_union"))) %>%
  arrange(gene_set, metric) %>% mutate(metric = as.character(metric), gene_set = as.character(gene_set)) %>%
  select(gene_set, metric, tier, feature, protein_name, n, slope, p_lm, q_lm, q_family,
         floor_flag, most_influential_mussel, influence_flag)
write_csv(best_hits, file.path(OUT_DIR, sprintf("best_hits_%s.csv", TISSUE)))

cat("\nBest hit per metric and gene set (the reported p beside both q's):\n")
best_hits %>% mutate(across(c(slope), ~signif(., 3)),
                     across(c(p_lm, q_lm, q_family), ~signif(., 2)),
                     protein_name = substr(protein_name, 1, 30)) %>%
  select(gene_set, metric, tier, feature, protein_name, n, slope, p_lm, q_lm, q_family,
         floor_flag, influence_flag) %>%
  as.data.frame() %>% print()
```

---

# C. Provenance

Script 01 writes `RUN_provenance_<T>.txt` with its settings, the code commit and input
hashes; this block replaces any earlier script-21 block in that file with this run's, so
re-running script 02 alone never stacks stale blocks.

```{r provenance}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
old  <- if (file.exists(prov)) readLines(prov) else character(0)
cut  <- grep("^\\[02_gene_mechanics_expanded\\.Rmd\\]", old)
if (length(cut)) old <- old[seq_len(cut[1] - 1)]
while (length(old) && !nzchar(old[length(old)])) old <- old[-length(old)]
settings <- c(
  "[02_gene_mechanics_expanded.Rmd]",
  paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
  paste0("reported test        : script 01's ANCOVA (assoc_candidate_", TISSUE, ".csv, assoc_DEGunion_", TISSUE, ".csv)"),
  paste0("metrics [scale, tier]: ", paste(sprintf("%s [%s, %s]", mcfg$metric, mcfg$scale, mcfg$tier), collapse = ", ")),
  paste0("primary metrics      : ", paste(PRIMARY_METRICS, collapse = ", ")),
  paste0("nuisance covariates  : ", paste(COVARIATES, collapse = " + ")),
  paste0("modules              : ", paste(sprintf("%s (%d)", names(mod_scores), sapply(mod_scores, `[[`, "n_genes")), collapse = ", ")),
  paste0("FDR_ALPHA            : ", FDR_ALPHA),
  paste0("candidates tested    : ", nrow(candidates), " of ", nrow(candidates_all), " (floor filter in script 01)"),
  paste0("N_INFLUENCE_HITS     : ", N_INFLUENCE_HITS))
inputs <- file.path(IN_DIR, c(sprintf(c("metrics_config_%s.csv", "paired_sample_manifest_%s.csv",
                                        "candidate_genes_%s.csv", "detection_floor_flags_%s.csv",
                                        "vst_paired_%s.csv", "assoc_candidate_%s.csv",
                                        "assoc_DEGunion_%s.csv"), TISSUE), "annotation_map.csv"))
inputs <- c(inputs, file.path(repo_root, "09_gene-mechanics-correlation", "01_code", "02_gene_mechanics_expanded.Rmd"))
writeLines(c(old, "", provenance_lines(settings, inputs, repo_root, packages = c("dplyr", "readr"))), prov)
cat(readLines(prov), sep = "\n")
```
