---
title: "05 - Thread mechanics against DEG sets, enriched GO terms and mitochondrial expression"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"                       # "F" foot (default) or "G" gill
  ontologies: ["BP", "MF", "CC"]    # GO ontologies whose enriched terms are tested
  min_genes: 3                      # smallest gene set scored
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 9)
```

# What this does

Scripts 01 and 02 test single genes and six keyword modules. This script asks the same
question of the gene sets the differential expression and the GO enrichment found:
**within an arm, and with the animal's own baseline held constant, do animals whose
stress-responsive genes are expressed more strongly make weaker or stronger plaques?**

Three kinds of set, all for this script's tissue and the treatment-control (TC) contrasts of
record:

1. **DEG sets**: the up- and the down-regulated DEGs of each stressor
   (`06_differential-expression`, `<T><X>_TC_siggene.csv`), six sets;
2. **enriched GO terms**: every term topGO `weight01` enriched in one of those runs
   (`07_enrichment/03_analyses/02_topgo/topgo_enriched.csv`), scored on the DEGs that put it
   there (the term's genes in that run). `fdr_supported` marks the terms a method with
   false-discovery control also enriched (topGO `classic`, goseq or clusterProfiler, BH < 0.05;
   `06_method-comparison/consensus_terms_TC_<ontology>.csv`); outside those, topGO's lists
   are exploratory;
3. **mitochondrial expression**: the reads of the 12 mitochondrial protein genes (the
   mitochondrial genome's genes and their copies on unplaced scaffolds) as a share of the
   library's nuclear reads, on the log scale, from `06 .../mitochondrial/mt_share_by_sample.csv`.
   These loci are left out of every gene set above.

**Score.** A set's score is the first principal component of its member genes' VST values
(standardised per gene), oriented so that a higher score means higher expression of the set,
exactly as script 02 scores its modules; members are the set's genes that passed script 01's
expression filter, and a set needs at least `min_genes` of them.

**Test.** The score goes through the per-animal ANCOVA every script here fits,
`level_day3 ~ score + treatment + level_baseline`, for each metric of script 01
(`metrics_config_<T>.csv`). `partial_r` is the score's partial correlation with the day-3
level. BH correction runs in two families, the whole-response scores (the six DEG sets and the
mitochondrial share) and the GO terms: within family and metric (`q_lm`) and within family
and tier (`q_family`).

**How to read it.** Each set was selected because its genes differ between arms, and the
treatment term removes those arm differences, so a slope here is an association among
animals of the same arm. Sets share genes (a GO term is a subset of its run's DEG set, and
parent and child terms overlap), so their tests are not independent. With 10 to 12 animals
per arm this is exploratory, as script 01 explains; quote q beside p.

Writes, in `03_analyses/go_mechanics/`:

- `mechanics_sets_<T>.csv`: one row per set: type, run, ontology, GO term, genes scored and
  their IDs, variance explained by the score, FDR support;
- `mechanics_set_associations_<T>.csv`: one row per set and metric: correction family, n,
  slope, partial r with a 95% interval, p, `q_lm`, `q_family`;
- `go_mechanics_<T>.png`: the partial correlations of every set with every metric;
- `RUN_provenance_<T>.txt`.

```{r inputs}
suppressPackageStartupMessages({ library(tidyverse) })
TISSUE <- if (exists("params") && !is.null(params$tissue)) toupper(params$tissue) else "F"
stopifnot(TISSUE %in% c("F", "G"))
ONTOLOGIES <- if (exists("params") && !is.null(params$ontologies)) unlist(params$ontologies) else c("BP", "MF", "CC")
MIN_GENES  <- if (exists("params") && !is.null(params$min_genes)) params$min_genes else 3
FDR_ALPHA  <- 0.10                          # as scripts 01 and 02

find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/", mustWork = FALSE)
  for (i in 1:8) {
    if (all(dir.exists(file.path(d, c("06_differential-expression", "02_thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing 06_differential-expression/ and 02_thread-strength/.")
}
repo_root <- find_repo_root()
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # warn_unless(), provenance
source(file.path(repo_root, "tools", "plot_style.R"))        # shared colours
source(file.path(repo_root, "tools", "gene_ids.R"))          # gene_key()
IN_DIR  <- file.path(repo_root, "09_gene-mechanics-correlation", "03_analyses", "gene_mechanics")
OUT_DIR <- file.path(repo_root, "09_gene-mechanics-correlation", "03_analyses", "go_mechanics")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
DE  <- file.path(repo_root, "06_differential-expression", "03_analyses")
GO  <- file.path(repo_root, "07_enrichment", "03_analyses")
need <- function(f) { if (!file.exists(f)) stop("Missing ", f, " - run the step that writes it first."); f }

## script 01's handoffs: the animals, their expression and the metrics, as scripts 01 and 02 use them
mcfg <- read_csv(need(file.path(IN_DIR, sprintf("metrics_config_%s.csv", TISSUE))), show_col_types = FALSE)
metrics    <- mcfg$metric
arms_kept  <- strsplit(mcfg$arms[1], ";", fixed = TRUE)[[1]]
COVARIATES <- strsplit(mcfg$covariates[1], ";", fixed = TRUE)[[1]]
scale_of <- function(m) mcfg$scale[match(m, mcfg$metric)]
tier_of  <- function(m) mcfg$tier[match(m, mcfg$metric)]
label_of <- function(m) sub(" \\(.*\\)$", "", mcfg$label[match(m, mcfg$metric)])
on_scale <- function(m, v) if (scale_of(m) == "log") log(v) else v
paired <- read_csv(need(file.path(IN_DIR, sprintf("paired_sample_manifest_%s.csv", TISSUE))), show_col_types = FALSE) %>%
  mutate(treatment = factor(treatment, levels = arms_kept))
vp_df <- read_csv(need(file.path(IN_DIR, sprintf("vst_paired_%s.csv", TISSUE))), show_col_types = FALSE)
vp <- as.matrix(vp_df[, -1]); rownames(vp) <- vp_df[[1]]
stopifnot(all(paired$sample %in% colnames(vp)))
vp <- vp[, paired$sample, drop = FALSE]
key <- gene_key                                # tools/gene_ids.R: "gene-LOC1|LOC1" -> "LOC1"
cat("Paired animals:", nrow(paired), "| genes passing the expression filter:", nrow(vp), "\n")
```

# The sets

```{r sets}
tissue_dir <- ifelse(TISSUE == "F", "Foot", "Gill")
codes <- paste0(TISSUE, c("OA", "OW", "DO"), "_TC")
run_label <- function(code, direction)
  sprintf("%s %s, %s", tissue_dir, substr(code, 2, 3), ifelse(direction == "up", "up", "down"))

## 1. DEG sets
deg_files <- file.path(DE, "DEG_lists", tissue_dir, paste0(codes, "_siggene.csv"))
warn_unless(all(file.exists(deg_files)), "Missing DEG list(s): ", paste(basename(deg_files[!file.exists(deg_files)]), collapse = ", "))
deg_sets <- bind_rows(lapply(codes[file.exists(deg_files)], function(cd) {
  d <- read.csv(file.path(DE, "DEG_lists", tissue_dir, paste0(cd, "_siggene.csv")), stringsAsFactors = FALSE)
  bind_rows(lapply(c("up", "down"), function(dir) {
    g <- key(d$gene[if (dir == "up") d$log2FoldChange > 0 else d$log2FoldChange < 0])
    tibble(set_type = "DEG set", code = cd, direction = dir, ontology = NA_character_, GO.ID = NA_character_,
           Term = sprintf("all %s DEGs", ifelse(dir == "up", "up-regulated", "down-regulated")),
           genes_listed = length(g), members = list(intersect(g, rownames(vp))))
  }))
}))

## 2. enriched GO terms, scored on the DEGs that enriched them
tg_file <- file.path(GO, "02_topgo", "topgo_enriched.csv")
warn_unless(file.exists(tg_file), "Missing ", tg_file, "; run 07_enrichment first. GO terms are skipped.")
tg <- if (file.exists(tg_file)) read.csv(tg_file, stringsAsFactors = FALSE) else NULL
cons <- bind_rows(lapply(ONTOLOGIES, function(o) {
  f <- file.path(GO, "06_method-comparison", sprintf("consensus_terms_TC_%s.csv", o))
  if (file.exists(f)) read.csv(f, stringsAsFactors = FALSE) %>% transmute(code, direction, GO.ID, ontology = o, fdr_methods = methods)
}))
go_sets <- if (is.null(tg)) NULL else tg %>%
  filter(code %in% codes, ontology %in% ONTOLOGIES) %>%
  left_join(cons, by = c("code", "direction", "GO.ID", "ontology")) %>%
  transmute(set_type = "GO term", code, direction, ontology, GO.ID, Term,
            genes_listed = Significant, fdr_methods,
            members = lapply(strsplit(genes, ";", fixed = TRUE), function(g) intersect(key(g), rownames(vp))))
warn_unless(is.null(tg) || nrow(go_sets) > 0, "No enriched TC terms for tissue ", TISSUE, " in ", tg_file, ".")

sets <- bind_rows(deg_sets, go_sets) %>%
  mutate(n_genes = lengths(members),
         fdr_supported = case_when(set_type != "GO term" ~ NA,
                                   TRUE ~ !is.na(fdr_methods)),
         set_id = ifelse(set_type == "DEG set", paste(code, direction),
                         paste(code, direction, ontology, GO.ID)),
         run = run_label(code, direction)) %>%
  select(-any_of("fdr_methods"))
cat("Sets:", nrow(sets), "| with at least", MIN_GENES, "expressed genes:", sum(sets$n_genes >= MIN_GENES), "\n")
print(table(sets$set_type, scored = sets$n_genes >= MIN_GENES))
```

# Scores and the ANCOVA

```{r scores}
## first principal component of the standardised member genes, oriented to the set's mean
## expression; the same score script 02 gives its modules
set_score <- function(locs) {
  z <- t(scale(t(vp[locs, , drop = FALSE])))
  z <- z[stats::complete.cases(z), , drop = FALSE]
  if (nrow(z) < MIN_GENES) return(NULL)
  pc <- prcomp(t(z), center = TRUE, scale. = FALSE)
  sc <- pc$x[, 1]
  if (suppressWarnings(cor(sc, colMeans(vp[rownames(z), , drop = FALSE]))) < 0) sc <- -sc
  list(score = unname(sc), var_expl = summary(pc)$importance[2, 1])
}

## the ANCOVA of scripts 01 and 02 on one score
ANCOVA_FORMULA <- reformulate(c("x", COVARIATES, "base"), response = "y")
fit_score <- function(x, metric) {
  dat <- data.frame(y    = on_scale(metric, paired[[metric]]), x = x,
                    base = on_scale(metric, paired[[paste0("base_", metric)]]),
                    paired[, COVARIATES, drop = FALSE])
  dat <- dat[stats::complete.cases(dat) & is.finite(dat$y) & is.finite(dat$base), , drop = FALSE]
  if (nrow(dat) < 6) return(NULL)
  fm <- lm(ANCOVA_FORMULA, data = dat)
  co <- summary(fm)$coefficients
  t  <- co["x", "t value"]; df <- fm$df.residual
  r  <- t / sqrt(t^2 + df)
  tibble(metric = metric, tier = tier_of(metric), n = nrow(dat),
         slope = co["x", "Estimate"], se = co["x", "Std. Error"], p_lm = co["x", "Pr(>|t|)"],
         partial_r = r, partial_r_lo = tanh(atanh(r) - qnorm(0.975) / sqrt(df - 1)),
         partial_r_hi = tanh(atanh(r) + qnorm(0.975) / sqrt(df - 1)))
}

scored <- sets %>% filter(n_genes >= MIN_GENES)
scores <- lapply(scored$members, set_score)
keep <- !vapply(scores, is.null, TRUE)
scored <- scored[keep, ]; scores <- scores[keep]
scored$pc1_var <- round(vapply(scores, `[[`, 0, "var_expl"), 3)

## 3. mitochondrial expression: log of the mitochondrial protein reads as a share of nuclear reads
mt_file <- file.path(DE, "mitochondrial", "mt_share_by_sample.csv")
warn_unless(file.exists(mt_file), "Missing ", mt_file, "; run 06_differential-expression step 13. ",
            "The mitochondrial row is skipped.")
if (file.exists(mt_file)) {
  mt <- read.csv(mt_file, stringsAsFactors = FALSE)
  mt_score <- log(mt$mt_reads_pct_of_nuclear[match(paired$sample, mt$sample.1)])
  warn_unless(!anyNA(mt_score), "Mitochondrial share missing for some paired samples.")
  scored <- bind_rows(scored, tibble(set_type = "mitochondrial", code = NA, direction = NA,
                                     Term = "mitochondrial protein genes (% of nuclear reads, log)",
                                     n_genes = 12L, set_id = "mitochondrial", run = "all samples",
                                     members = list(character(0)), pc1_var = NA_real_))
  scores <- c(scores, list(list(score = mt_score)))
}

assoc <- bind_rows(lapply(seq_len(nrow(scored)), function(i)
  bind_rows(lapply(metrics, function(m) fit_score(scores[[i]]$score, m))) %>%
    mutate(set_id = scored$set_id[i], .before = 1))) %>%
  left_join(scored %>% select(set_id, set_type, run, code, direction, ontology, GO.ID, Term, n_genes,
                              fdr_supported), by = "set_id") %>%
  ## two correction families: the whole-response scores (six DEG sets and the mitochondrial
  ## share, which alone would be a family of one whose q is its p) and the GO terms
  mutate(bh_family = ifelse(set_type == "GO term", "GO terms", "DEG sets and mitochondrial")) %>%
  group_by(bh_family, metric) %>% mutate(q_lm = p.adjust(p_lm, "BH")) %>% ungroup() %>%
  group_by(bh_family, tier)   %>% mutate(q_family = p.adjust(p_lm, "BH")) %>% ungroup() %>%
  select(set_type, set_id, bh_family, run, code, direction, ontology, GO.ID, Term, n_genes, fdr_supported, metric,
         tier, n, slope, se, partial_r, partial_r_lo, partial_r_hi, p_lm, q_lm, q_family) %>%
  arrange(factor(set_type, levels = c("DEG set", "mitochondrial", "GO term")), factor(metric, levels = metrics), p_lm)

write_csv(sets %>% mutate(scored = set_id %in% scored$set_id,
                          members = vapply(members, paste, "", collapse = ";")) %>%
            left_join(scored %>% select(set_id, pc1_var), by = "set_id") %>%
            select(set_type, set_id, run, code, direction, ontology, GO.ID, Term, genes_listed, n_genes,
                   scored, pc1_var, fdr_supported, members),
          file.path(OUT_DIR, sprintf("mechanics_sets_%s.csv", TISSUE)))
write_csv(assoc, file.path(OUT_DIR, sprintf("mechanics_set_associations_%s.csv", TISSUE)))

cat("\nSmallest p per set type and metric:\n")
assoc %>% group_by(set_type, metric) %>% slice_min(p_lm, n = 2, with_ties = FALSE) %>% ungroup() %>%
  transmute(set_type, metric, run, Term = substr(Term, 1, 45), n_genes, partial_r = round(partial_r, 2),
            p_lm = signif(p_lm, 2), q_lm = signif(q_lm, 2), q_family = signif(q_family, 2)) %>%
  as.data.frame() %>% print()
cat("\nSet x metric tests with q_lm <", FDR_ALPHA, ":", sum(assoc$q_lm < FDR_ALPHA, na.rm = TRUE),
    "of", nrow(assoc), "\n")
```

# Figure

```{r figure}
## GO term names are cut, not edited: short_name() is for UniProt names, whose parentheses
## hold alternative names, while a GO name's parentheses are part of it
R_CAP <- 0.5   # colour scale limit; partial r beyond it is drawn at full colour
go_label <- function(term, width = 44)
  ifelse(nchar(term) <= width, term, paste0(trimws(substr(term, 1, width - 3)), "..."))
fig <- assoc %>%
  mutate(group = case_when(set_type == "DEG set" ~ "DEG sets",
                           set_type == "mitochondrial" ~ "Mito-\nchondrial",
                           TRUE ~ paste("GO", ontology)),
         group = factor(group, levels = c("DEG sets", "Mito-\nchondrial", paste("GO", ONTOLOGIES))),
         row = ifelse(set_type == "GO term",
                      sprintf("%s | %s%s", run, go_label(Term), ifelse(fdr_supported, " (FDR)", "")),
                      ifelse(set_type == "DEG set", sprintf("%s | %s", run, Term), Term)),
         metric = factor(metric, levels = metrics, labels = vapply(metrics, label_of, "")),
         mark = case_when(q_lm < FDR_ALPHA ~ "q < 0.1", p_lm < 0.05 ~ "p < 0.05", TRUE ~ ""))
## two terms whose shortened names coincide would share a row; give them their GO IDs
fig <- fig %>% group_by(group, row) %>%
  mutate(row = if (n_distinct(GO.ID) > 1) sprintf("%s [%s]", row, GO.ID) else row) %>% ungroup()
stopifnot(!anyDuplicated(distinct(fig, row, set_id)$row))
row_levels <- fig %>% distinct(group, row, code, direction) %>%
  arrange(group, match(code, codes), desc(direction == "up"), row) %>% pull(row) %>% unique()
fig$row <- factor(fig$row, levels = rev(row_levels))
p <- ggplot(fig, aes(metric, row, fill = partial_r)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(mark == "", "", ifelse(mark == "q < 0.1", "**", "*"))), size = 3, vjust = 0.75) +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  scale_fill_gradient2(low = DIRECTION_COLORS[["down"]], mid = "white", high = DIRECTION_COLORS[["up"]],
                       midpoint = 0, limits = c(-R_CAP, R_CAP), oob = scales::squish,
                       breaks = c(-R_CAP, -R_CAP / 2, 0, R_CAP / 2, R_CAP),
                       labels = c(paste0("-", R_CAP, " or less"), -R_CAP / 2, 0, R_CAP / 2, paste0(R_CAP, " or more")),
                       name = "partial r\n(ANCOVA)") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = sprintf("%s: thread mechanics against DEG sets, enriched GO terms and mitochondrial expression",
                       tissue_dir),
       caption = paste(strwrap(sprintf(paste(
         "Partial correlation of each set's score (first principal component of its genes' VST expression) with",
         "the day-3 level, arm and the animal's baseline held constant; n = %d animals. Red: higher set expression",
         "goes with a higher value; colour capped at |r| = %s. *: p < 0.05; **: BH q < %s within family (response-level sets; GO terms) and metric. (FDR) after a GO term:",
         "a method with FDR control (topGO classic, goseq or clusterProfiler, BH < 0.05) also enriched it.",
         "Sets were chosen for differing between arms; exploratory."),
         sum(!is.na(paired$base_mean_force)), R_CAP, FDR_ALPHA), width = 120), collapse = "\n")) +
  theme_psmfc(9) +
  theme(panel.grid = element_blank(), strip.text.y = element_text(angle = 0),
        plot.title.position = "plot", plot.caption = element_text(hjust = 0), plot.caption.position = "plot")
ggsave(file.path(OUT_DIR, sprintf("go_mechanics_%s.png", TISSUE)), p,
       width = 10, height = 2.2 + 0.17 * length(row_levels), dpi = 300, limitsize = FALSE)
p
```

# Provenance

```{r provenance}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
settings <- c(
  "[05_go_term_mechanics.Rmd]",
  paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
  paste0("tissue               : ", TISSUE),
  paste0("reported test        : ANCOVA level_day3 ~ score + ", paste(COVARIATES, collapse = " + "),
         " + level_baseline (script 01's metrics, scales and animals)"),
  paste0("metrics [tier]       : ", paste(sprintf("%s [%s]", mcfg$metric, mcfg$tier), collapse = ", ")),
  paste0("ontologies           : ", paste(ONTOLOGIES, collapse = ", ")),
  paste0("MIN_GENES / FDR_ALPHA: ", MIN_GENES, " / ", FDR_ALPHA),
  paste0("sets scored          : ", paste(sprintf("%s %d", names(table(scored$set_type)), as.integer(table(scored$set_type))),
                                          collapse = ", "), " (of ", nrow(sets) + file.exists(mt_file), " sets)"),
  paste0("animals in the fits  : ", sum(!is.na(paired$base_mean_force))))
inputs <- c(file.path(IN_DIR, sprintf(c("metrics_config_%s.csv", "paired_sample_manifest_%s.csv", "vst_paired_%s.csv"), TISSUE)),
            deg_files, tg_file,
            file.path(GO, "06_method-comparison", sprintf("consensus_terms_TC_%s.csv", ONTOLOGIES)),
            mt_file,
            file.path(repo_root, "09_gene-mechanics-correlation", "01_code", "05_go_term_mechanics.Rmd"),
            file.path(repo_root, "tools", c("pipeline_checks.R", "plot_style.R", "gene_ids.R")))
writeLines(provenance_lines(settings, inputs, repo_root, packages = c("dplyr", "readr")), prov)
cat(readLines(prov), sep = "\n")
```
