---
title: "06 - Expression suites: programs, modules and multi-gene prediction of thread mechanics"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"              # "F" foot (default) or "G" gill
  min_set_genes: 10        # smallest DEG set or GO term (script 05's sets) used as an axis
  n_var_genes: 4000        # genes with the largest within-arm variance, clustered into modules
  min_module: 30           # smallest co-expression module
  n_pcs: 10                # principal components of all expressed genes used as axes
  n_top_var: 2000          # genes with the largest variance, one of the prediction test's gene sets
  nperm: 100               # label permutations per prediction scenario
  cv_repeats: 3            # repeats of the 5-fold cross-validation
  cores: 4                 # parallel workers (socket cluster, so Windows too); 1 runs serially
  seed: 20261006
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

# What this does

Scripts 01, 02 and 05 test one gene, one keyword module or one gene set at a time. This script
asks whether **combinations** of genes ("suites") go with strong or weak attachment, and it
separates two questions that a single "expression profile" plot mixes up:

- **Across treatments**: do animals whose expression is in a given state attach more or less
  strongly, whatever the reason? Warming and hypoxia change both expression and thread
  strength, so an association here can be the treatment itself.
- **Within treatments**: among animals in the same arm, with each animal's own baseline held
  constant, does the suite track attachment? This is the question of scripts 01, 02 and 05,
  and the one that matters for screening individuals.

**Axes** (each one a score per animal):

1. **Programs**: script 05's DEG sets (the up- and the down-regulated DEGs of each stressor,
   tens to hundreds of genes) and its enriched GO terms, with at least `min_set_genes`
   expressed genes; scored exactly as script 05 scores them (first principal component of the
   standardised genes, oriented to the set's mean expression), which a check confirms;
2. **Co-expression modules**: the `n_var_genes` genes with the largest within-arm variance,
   arm removed, clustered by signed correlation (dissimilarity (1 - r) / 2, average linkage)
   and cut with the dynamic tree cut (Langfelder, Zhang and Horvath 2008, *Bioinformatics*
   24:719-720; `dynamicTreeCut`), modules of at least `min_module` genes, in the spirit of
   WGCNA (Langfelder and Horvath 2008, *BMC Bioinformatics* 9:559) without its soft threshold
   and topological overlap. A module's score is its first principal component within arms,
   applied to the expression including the arm differences, so that the score's arm means can
   be drawn; its within-arm part is the module's;
3. **Expression principal components**: the first `n_pcs` components of all expressed genes
   (script 01's filter), unsupervised.

**Tests**, for every axis and every metric of script 01 (`metrics_config_<T>.csv`):

- *within arms*: the ANCOVA of scripts 01, 02 and 05, `level_day3 ~ score + treatment +
  level_baseline`, partial r with a 95% interval, p; BH over all axes within metric (`q_lm`),
  within metric and axis type (`q_type`) and within tier (`q_family`);
- *across arms*: the same without the treatment term, `level_day3 ~ score + level_baseline`
  (`across_r`, `p_across`); an association here includes the treatment effect;
- *between arms*: the correlation of the four arm means of the score with the four arm means of
  the baseline-adjusted level (`between_r`, four points, descriptive only), and the share of
  the score's variance between arms (`eta2_arm`);
- *interaction*: whether the within-arm slope differs between arms (`p_interaction`).

**Prediction test.** For each primary metric, an elastic net (glmnet; Friedman, Hastie and
Tibshirani 2010, *J Stat Softw* 33(1):1-22; alpha 0.5, lambda by an inner 5-fold
cross-validation) predicts the day-3 level from three predictor sets (script 01's candidate
genes, the `n_top_var` most variable genes and all the axes above), across and within arms.
Covariates (the baseline, and the arm for the within question) are removed inside every
training fold. Out-of-sample Q2 is 1 - SSE(model) / SSE(training-fold mean) over `cv_repeats`
repeats of a 5-fold split balanced by arm; 0 means no better than the mean. The null is
`nperm` label permutations (across arms: all animals; within arms: within each arm), each run
through the same procedure; p = (1 + #null >= observed) / (1 + nperm). Every run is seeded from
`seed`, the scenario and the permutation, so results do not depend on the number of workers.
For reference, the same cross-validation predicts the across-arm level from the treatment alone.

**How to read it.** The programs were chosen because their genes differ between arms, so
arms separate along them by construction; the within-arm tests are the informative ones.
With 10 to 12 animals per arm, a within-arm association needs a partial r of about 0.41 to
reach p < 0.05 in one test with 80% power and about 0.6 after correcting for 100 tests, so
this screen can show strong suites, not rule out modest ones. Exploratory.

Writes, in `03_analyses/expression_suites/`:

- `suite_axes_<T>.csv`: one row per axis: type, label, genes, variance explained, arm means
  of the score, `eta2_arm`;
- `suite_axis_tests_<T>.csv`: one row per axis and metric: the tests above;
- `suite_axis_members_<T>.csv`: the genes of each axis (all members of programs and
  modules with their weight; the 50 genes with the largest loading for each component) with
  protein names;
- `suite_scores_<T>.csv`: one row per animal: arm, the metrics, the baseline-adjusted level as
  % of the control mean for each primary metric, and every axis score;
- `suite_prediction_<T>.csv` and `suite_prediction_null_<T>.csv`: the prediction test and
  its permutation nulls;
- `suites_between_within_<T>.png`: the DEG programs, arm means against within-arm slopes;
- `suites_axes_map_<T>.png`: every axis, across-arm against within-arm association;
- `suites_state_space_<T>.png`: animals in the space of the DEG programs and of the first two
  expression components, coloured by attachment;
- `suites_prediction_<T>.png`: the prediction test and the within-arm tests against chance;
- `RUN_provenance_<T>.txt`.

```{r inputs}
suppressPackageStartupMessages({ library(tidyverse); library(glmnet); library(dynamicTreeCut)
                                 library(patchwork); library(ggrepel); library(parallel) })
par_or <- function(name, default) if (exists("params") && !is.null(params[[name]])) params[[name]] else default
TISSUE      <- toupper(par_or("tissue", "F")); stopifnot(TISSUE %in% c("F", "G"))
MIN_SET     <- par_or("min_set_genes", 10)
N_VAR       <- par_or("n_var_genes", 4000)
MIN_MODULE  <- par_or("min_module", 30)
N_PCS       <- par_or("n_pcs", 10)
N_TOP_VAR   <- par_or("n_top_var", 2000)
NPERM       <- par_or("nperm", 100)
CV_REPEATS  <- par_or("cv_repeats", 3)
CORES       <- max(1L, as.integer(par_or("cores", 4)))
SEED        <- par_or("seed", 20261006)
FDR_ALPHA   <- 0.10                                   # as scripts 01, 02 and 05

find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/", mustWork = FALSE)
  for (i in 1:8) {
    if (all(dir.exists(file.path(d, c("05_differential-expression", "02_thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing 05_differential-expression/ and 02_thread-strength/.")
}
repo_root <- find_repo_root()
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # warn_unless(), provenance
source(file.path(repo_root, "tools", "plot_style.R"))        # shared colours
GM_DIR  <- file.path(repo_root, "09_gene-mechanics-correlation", "03_analyses", "gene_mechanics")
GO_DIR  <- file.path(repo_root, "09_gene-mechanics-correlation", "03_analyses", "go_mechanics")
OUT_DIR <- file.path(repo_root, "09_gene-mechanics-correlation", "03_analyses", "expression_suites")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
need <- function(f) { if (!file.exists(f)) stop("Missing ", f, " - run the step that writes it first."); f }
f_in <- c(metrics = sprintf("metrics_config_%s.csv", TISSUE), paired = sprintf("paired_sample_manifest_%s.csv", TISSUE),
          vst = sprintf("vst_paired_%s.csv", TISSUE), cand = sprintf("candidate_genes_%s.csv", TISSUE))
f_in <- setNames(file.path(GM_DIR, f_in), names(f_in))
f_in <- c(f_in, ann = file.path(GM_DIR, "annotation_map.csv"),
          sets = file.path(GO_DIR, sprintf("mechanics_sets_%s.csv", TISSUE)),
          sets_assoc = file.path(GO_DIR, sprintf("mechanics_set_associations_%s.csv", TISSUE)))
invisible(lapply(f_in, need))

mcfg <- read_csv(f_in[["metrics"]], show_col_types = FALSE)
metrics    <- mcfg$metric
primary    <- mcfg$metric[mcfg$tier == "primary"]
arms_kept  <- strsplit(mcfg$arms[1], ";", fixed = TRUE)[[1]]
COVARIATES <- strsplit(mcfg$covariates[1], ";", fixed = TRUE)[[1]]
stopifnot(identical(COVARIATES, "treatment"))
scale_of <- function(m) mcfg$scale[match(m, mcfg$metric)]
tier_of  <- function(m) mcfg$tier[match(m, mcfg$metric)]
label_of <- function(m) sub(" \\(.*\\)$", "", mcfg$label[match(m, mcfg$metric)])
on_scale <- function(m, v) if (scale_of(m) == "log") log(v) else v

paired <- read_csv(f_in[["paired"]], show_col_types = FALSE) %>% mutate(treatment = factor(treatment, levels = arms_kept))
vp_df <- read_csv(f_in[["vst"]], show_col_types = FALSE)
vp <- as.matrix(vp_df[, -1]); rownames(vp) <- vp_df[[1]]
stopifnot(all(paired$sample %in% colnames(vp)))
vp <- vp[, paired$sample, drop = FALSE]
X_all <- t(vp)                                          # animals x genes
arm <- paired$treatment
ann <- read_csv(f_in[["ann"]], show_col_types = FALSE) %>% select(LOC_ID, protein_name)
cand <- read_csv(f_in[["cand"]], show_col_types = FALSE) %>% filter(tested) %>% pull(LOC_ID) %>% intersect(colnames(X_all))
cat(sprintf("%s: %d paired animals, %d expressed genes, %d candidate genes\n", TISSUE, nrow(X_all), ncol(X_all), length(cand)))
```

# The axes

```{r axes}
## 1. programs: script 05's DEG sets and GO terms, scored as script 05 scores them
set_score <- function(locs, min_genes = 3) {
  z <- t(scale(t(vp[locs, , drop = FALSE])))
  z <- z[stats::complete.cases(z), , drop = FALSE]
  if (nrow(z) < min_genes) return(NULL)
  pc <- prcomp(t(z), center = TRUE, scale. = FALSE)
  sc <- pc$x[, 1]
  if (suppressWarnings(cor(sc, colMeans(vp[rownames(z), , drop = FALSE]))) < 0) sc <- -sc
  list(score = unname(sc), var_expl = summary(pc)$importance[2, 1], weight = pc$rotation[, 1] * sign(cor(pc$x[, 1], sc)))
}
sets05 <- read_csv(f_in[["sets"]], show_col_types = FALSE) %>%
  filter(scored, set_type %in% c("DEG set", "GO term"), n_genes >= MIN_SET)
prog <- lapply(strsplit(sets05$members, ";", fixed = TRUE), function(g) set_score(intersect(g, rownames(vp))))
keep <- !vapply(prog, is.null, TRUE); sets05 <- sets05[keep, ]; prog <- prog[keep]
axes <- tibble(axis_id = sets05$set_id,
               type = ifelse(sets05$set_type == "DEG set", "DEG program", "GO term"),
               label = ifelse(sets05$set_type == "DEG set", sets05$run, sprintf("%s: %s", sets05$run, sets05$Term)),
               n_genes = sets05$n_genes, var_expl = vapply(prog, `[[`, 0, "var_expl"))
S <- sapply(prog, `[[`, "score"); colnames(S) <- axes$axis_id
members <- bind_rows(lapply(seq_along(prog), function(i)
  tibble(axis_id = axes$axis_id[i], LOC_ID = names(prog[[i]]$weight), weight = unname(prog[[i]]$weight))))

## 2. co-expression modules within arms
R_arm <- resid(lm(X_all ~ arm))                          # expression with the arm means removed
sd_w <- apply(R_arm, 2, sd)
top_w <- names(sort(sd_w, decreasing = TRUE))[seq_len(min(N_VAR, ncol(X_all)))]
Zw <- sweep(R_arm[, top_w], 2, sd_w[top_w], "/")         # within-arm, standardised
Zraw <- sweep(sweep(X_all[, top_w], 2, colMeans(X_all[, top_w])), 2, sd_w[top_w], "/")  # same scale, arm differences kept
D <- (1 - cor(Zw)) / 2
tree <- hclust(as.dist(D), method = "average")
mod <- cutreeDynamic(tree, distM = D, deepSplit = 2, minClusterSize = MIN_MODULE, verbose = 0)
mods <- sort(setdiff(unique(mod), 0))
short <- function(x) ifelse(is.na(x) | x == "", NA, substr(sub(" \\(.*$", "", x), 1, 40))
for (k in mods) {
  g <- top_w[mod == k]
  pc <- prcomp(Zw[, g], center = TRUE, scale. = FALSE)
  w <- pc$rotation[, 1]; if (cor(as.vector(Zw[, g] %*% w), rowMeans(Zw[, g])) < 0) w <- -w
  id <- sprintf("module %02d", k)
  kme <- as.vector(cor(Zw[, g], Zw[, g] %*% w))
  hubs <- tibble(LOC_ID = g, kME = kme) %>% left_join(ann, by = "LOC_ID") %>% arrange(desc(kME))
  hub_names <- head(unique(na.omit(short(hubs$protein_name))), 3)
  axes <- bind_rows(axes, tibble(axis_id = id, type = "Co-expression module",
                                 label = sprintf("Module %02d: %s", k, paste(hub_names, collapse = "; ")),
                                 n_genes = length(g), var_expl = summary(pc)$importance[2, 1]))
  S <- cbind(S, setNames(data.frame(as.vector(Zraw[, g] %*% w)), id))
  members <- bind_rows(members, tibble(axis_id = id, LOC_ID = g, weight = unname(w), kME = kme))
}

## 3. principal components of all expressed genes
pca <- prcomp(X_all, center = TRUE, scale. = FALSE)
ve <- summary(pca)$importance[2, ]
for (k in seq_len(min(N_PCS, ncol(pca$x)))) {
  ld <- pca$rotation[, k]; top50 <- names(sort(abs(ld), decreasing = TRUE))[1:50]
  sg <- if (sum(ld[top50]) < 0) -1 else 1                # oriented so its largest loadings are mostly positive
  id <- sprintf("PC%d", k)
  axes <- bind_rows(axes, tibble(axis_id = id, type = "Expression PC",
                                 label = sprintf("Expression PC%d (%.1f%% of variance)", k, 100 * ve[k]),
                                 n_genes = ncol(X_all), var_expl = unname(ve[k])))
  S <- cbind(S, setNames(data.frame(sg * pca$x[, k]), id))
  members <- bind_rows(members, tibble(axis_id = id, LOC_ID = top50, weight = unname(sg * ld[top50])))
}
S <- as.matrix(S); rownames(S) <- paired$sample
S <- scale(S)                                             # every axis in SD units across animals
axes <- axes %>% mutate(eta2_arm = apply(S, 2, function(x) summary(lm(x ~ arm))$r.squared),
                        mean_control = colMeans(S[arm == "control", , drop = FALSE]),
                        mean_OA = colMeans(S[arm == "OA", , drop = FALSE]),
                        mean_OW = colMeans(S[arm == "OW", , drop = FALSE]),
                        mean_DO = colMeans(S[arm == "DO", , drop = FALSE]))
members <- members %>% left_join(ann, by = "LOC_ID")
cat("Axes:\n"); print(table(axes$type))
cat(sprintf("Modules: %d from %d genes (%d unassigned)\n", length(mods), length(top_w), sum(mod == 0)))
warn_unless(!anyNA(S), "Axis scores contain missing values.")
warn_unless(length(mods) >= 1, "No co-expression module was found.")
warn_unless(all(table(mod[mod != 0]) >= MIN_MODULE), "A module is smaller than min_module.")
```

# The axis tests

```{r tests}
fit_axis <- function(x, metric) {
  dat <- data.frame(y = on_scale(metric, paired[[metric]]), x = x,
                    base = on_scale(metric, paired[[paste0("base_", metric)]]), treatment = arm)
  dat <- dat[stats::complete.cases(dat) & is.finite(dat$y) & is.finite(dat$base), , drop = FALSE]
  fm <- lm(y ~ x + treatment + base, data = dat)
  co <- summary(fm)$coefficients; t <- co["x", "t value"]; df <- fm$df.residual; r <- t / sqrt(t^2 + df)
  fa <- lm(y ~ x + base, data = dat); ta <- summary(fa)$coefficients["x", "t value"]
  fi <- lm(y ~ x * treatment + base, data = dat)
  yadj <- resid(lm(y ~ base, data = dat))
  am <- tapply(dat$x, dat$treatment, mean); ay <- tapply(yadj, dat$treatment, mean)
  tibble(metric = metric, tier = tier_of(metric), n = nrow(dat),
         slope = co["x", "Estimate"], se = co["x", "Std. Error"], p_lm = co["x", "Pr(>|t|)"],
         partial_r = r, partial_r_lo = tanh(atanh(r) - qnorm(0.975) / sqrt(df - 1)),
         partial_r_hi = tanh(atanh(r) + qnorm(0.975) / sqrt(df - 1)),
         across_r = ta / sqrt(ta^2 + fa$df.residual), p_across = summary(fa)$coefficients["x", "Pr(>|t|)"],
         between_r = suppressWarnings(cor(am, ay)), p_interaction = anova(fm, fi)$`Pr(>F)`[2])
}
tests <- bind_rows(lapply(seq_len(ncol(S)), function(i)
  bind_rows(lapply(metrics, function(m) fit_axis(S[, i], m))) %>% mutate(axis_id = colnames(S)[i], .before = 1))) %>%
  left_join(axes %>% select(axis_id, type, label, n_genes), by = "axis_id") %>%
  group_by(metric) %>% mutate(q_lm = p.adjust(p_lm, "BH")) %>%
  group_by(metric, type) %>% mutate(q_type = p.adjust(p_lm, "BH")) %>%
  group_by(tier) %>% mutate(q_family = p.adjust(p_lm, "BH")) %>% ungroup() %>%
  relocate(type, label, n_genes, .after = axis_id)

## check: script 05's sets give script 05's within-arm partial correlations
a05 <- read_csv(f_in[["sets_assoc"]], show_col_types = FALSE) %>% select(set_id, metric, partial_r)
cmp <- tests %>% filter(type %in% c("DEG program", "GO term")) %>% select(axis_id, metric, partial_r) %>%
  left_join(a05, by = c("axis_id" = "set_id", "metric"), suffix = c("", "_05"))
warn_unless(nrow(cmp) > 0 && !anyNA(cmp$partial_r_05) && max(abs(cmp$partial_r - cmp$partial_r_05)) < 1e-8,
            "Program scores do not reproduce script 05's partial correlations (largest difference ",
            signif(max(abs(cmp$partial_r - cmp$partial_r_05), na.rm = TRUE), 3), ").")
cat("Within-arm tests with q_lm <", FDR_ALPHA, ":", sum(tests$q_lm < FDR_ALPHA), "of", nrow(tests), "\n")
tests %>% filter(tier == "primary") %>% group_by(type, metric) %>%
  summarise(axes = n(), lowest_p = signif(min(p_lm), 2), lowest_q = signif(min(q_lm), 2),
            p_below_05 = sum(p_lm < 0.05), max_abs_across_r = round(max(abs(across_r)), 2), .groups = "drop") %>%
  knitr::kable()
```

# Multi-gene prediction of held-out animals

```{r prediction}
resid_on <- function(Y, D, tr) {                          # residuals of Y on D, coefficients fitted on rows tr
  B <- qr.coef(qr(D[tr, , drop = FALSE]), Y[tr, , drop = FALSE]); B[is.na(B)] <- 0
  Y - D %*% B
}
cv_q2 <- function(X, y, D, strata, reps) {               # the RNG stream is set by the caller
  n <- length(y); sse_m <- sse_0 <- 0
  for (r in seq_len(reps)) {
    fold <- integer(n)
    for (s in sort(unique(strata))) { idx <- which(strata == s); fold[idx] <- sample(rep_len(1:5, length(idx))) }
    for (f in 1:5) {
      te <- which(fold == f); tr <- which(fold != f)
      Xr <- resid_on(X, D, tr); yr <- as.vector(resid_on(matrix(y), D, tr))
      keep <- which(apply(Xr[tr, , drop = FALSE], 2, sd) > 1e-8)
      fit <- glmnet::cv.glmnet(Xr[tr, keep, drop = FALSE], yr[tr], alpha = 0.5, foldid = sample(rep_len(1:5, length(tr))))
      p <- as.vector(predict(fit, Xr[te, keep, drop = FALSE], s = "lambda.min"))
      sse_m <- sse_m + sum((yr[te] - p)^2); sse_0 <- sse_0 + sum((yr[te] - mean(yr[tr]))^2)
    }
  }
  1 - sse_m / sse_0
}
top_var <- names(sort(apply(X_all, 2, var), decreasing = TRUE))[seq_len(min(N_TOP_VAR, ncol(X_all)))]
pred_sets <- list(candidates = X_all[, cand, drop = FALSE], top_variable = X_all[, top_var, drop = FALSE], axes = S)
SCEN <- list(); scen_tab <- tibble()
for (m in primary) for (ps in names(pred_sets)) for (tg in c("across", "within")) {
  ok <- is.finite(on_scale(m, paired[[m]])) & is.finite(on_scale(m, paired[[paste0("base_", m)]]))
  b <- on_scale(m, paired[[paste0("base_", m)]])[ok]
  SCEN[[length(SCEN) + 1]] <- list(X = pred_sets[[ps]][ok, , drop = FALSE], y = on_scale(m, paired[[m]])[ok],
    D = if (tg == "across") cbind(1, b) else model.matrix(~ arm[ok] + b), strata = as.integer(arm[ok]), within = tg == "within")
  scen_tab <- bind_rows(scen_tab, tibble(scenario = length(SCEN), metric = m, predictors = ps, target = tg,
                                         n = sum(ok), n_predictors = ncol(pred_sets[[ps]])))
}
tasks <- expand.grid(perm = 0:NPERM, scenario = seq_along(SCEN))
run_task <- function(i) {
  sc <- SCEN[[tasks$scenario[i]]]; set.seed(SEED + 10000L * tasks$scenario[i] + tasks$perm[i])
  y <- sc$y
  if (tasks$perm[i] > 0) {
    if (sc$within) for (s in unique(sc$strata)) { idx <- which(sc$strata == s); y[idx] <- y[idx][sample.int(length(idx))] }
    else y <- y[sample.int(length(y))]
  }
  cv_q2(sc$X, y, sc$D, sc$strata, CV_REPEATS)
}
t0 <- Sys.time()
## workers get the data once (clusterExport) and functions whose environment is the global one,
## so that no task carries this report's environment (with its large matrices) to a worker
ship <- function(f) { environment(f) <- globalenv(); f }
q2 <- if (CORES > 1) {
  cl <- makePSOCKcluster(CORES)
  tryCatch({
    invisible(clusterCall(cl, function(lp) { .libPaths(lp); suppressPackageStartupMessages(library(glmnet)); NULL }, .libPaths()))
    wk <- new.env()
    wk$SCEN <- SCEN; wk$tasks <- tasks; wk$SEED <- SEED; wk$CV_REPEATS <- CV_REPEATS
    wk$cv_q2 <- ship(cv_q2); wk$resid_on <- ship(resid_on); wk$run_task <- ship(run_task)
    clusterExport(cl, ls(wk), envir = wk)
    unlist(parLapplyLB(cl, seq_len(nrow(tasks)), ship(function(i) run_task(i))))
  }, finally = stopCluster(cl))
} else vapply(seq_len(nrow(tasks)), run_task, 0)
cat(sprintf("%d cross-validated fits on %d worker(s) in %.1f min\n", nrow(tasks), CORES, as.numeric(difftime(Sys.time(), t0, units = "mins"))))
tasks$q2 <- q2
## determinism: the first scenario's observed Q2, rerun serially in this session
warn_unless(isTRUE(all.equal(run_task(1), tasks$q2[1], tolerance = 1e-12)), "The prediction test is not reproducible across workers.")

## reference: the treatment alone, across arms
arm_q2 <- sapply(primary, function(m) {
  ok <- is.finite(on_scale(m, paired[[m]])) & is.finite(on_scale(m, paired[[paste0("base_", m)]]))
  y <- on_scale(m, paired[[m]])[ok]; b <- on_scale(m, paired[[paste0("base_", m)]])[ok]; a <- arm[ok]
  set.seed(SEED); n <- length(y); sse_m <- sse_0 <- 0
  for (r in seq_len(CV_REPEATS)) {
    fold <- integer(n); for (s in levels(a)) { idx <- which(a == s); fold[idx] <- sample(rep_len(1:5, length(idx))) }
    for (f in 1:5) { tr <- fold != f; te <- fold == f
      yr <- y - (cbind(1, b) %*% qr.coef(qr(cbind(1, b)[tr, ]), y[tr]))[, 1]
      p <- tapply(yr[tr], a[tr], mean)[as.character(a[te])]
      sse_m <- sse_m + sum((yr[te] - p)^2); sse_0 <- sse_0 + sum((yr[te] - mean(yr[tr]))^2) } }
  1 - sse_m / sse_0 })

pred <- scen_tab %>%
  left_join(tasks %>% group_by(scenario) %>%
              summarise(q2_obs = q2[perm == 0], null_median = median(q2[perm > 0]), null_lo = unname(quantile(q2[perm > 0], 0.025)),
                        null_hi = unname(quantile(q2[perm > 0], 0.975)), n_perm = sum(perm > 0),
                        p_perm = (1 + sum(q2[perm > 0] >= q2[perm == 0])) / (1 + sum(perm > 0)), .groups = "drop"),
            by = "scenario") %>%
  rename(q2 = q2_obs) %>%
  mutate(treatment_only_q2 = ifelse(target == "across", arm_q2[metric], NA_real_))
warn_unless(all(pred$n_perm == NPERM) && all(is.finite(pred$q2)), "A prediction scenario is incomplete.")
warn_unless(all(is.finite(arm_q2)), "The treatment-only reference failed.")
knitr::kable(pred %>% transmute(metric, predictors, target, n, n_predictors, q2 = round(q2, 3), null_median = round(null_median, 3),
                                null_hi = round(null_hi, 3), p_perm = round(p_perm, 3), treatment_only_q2 = round(treatment_only_q2, 3)))
```

# Outputs

```{r write}
strength_pct <- sapply(primary, function(m) {             # baseline-adjusted level, % of the control mean
  y <- on_scale(m, paired[[m]]); b <- on_scale(m, paired[[paste0("base_", m)]]); ok <- is.finite(y) & is.finite(b)
  out <- rep(NA_real_, length(y)); r <- resid(lm(y[ok] ~ b[ok]))
  out[ok] <- 100 * exp(r - mean(r[arm[ok] == "control"])); out })
colnames(strength_pct) <- paste0(primary, "_pct_control")
scores_out <- bind_cols(paired %>% select(sample, mussel, treatment, any_of(metrics), any_of(paste0("base_", metrics))),
                        as_tibble(strength_pct), as_tibble(S))
write_csv(axes, file.path(OUT_DIR, sprintf("suite_axes_%s.csv", TISSUE)))
write_csv(tests, file.path(OUT_DIR, sprintf("suite_axis_tests_%s.csv", TISSUE)))
write_csv(members %>% select(axis_id, LOC_ID, protein_name, weight, kME), file.path(OUT_DIR, sprintf("suite_axis_members_%s.csv", TISSUE)))
write_csv(scores_out, file.path(OUT_DIR, sprintf("suite_scores_%s.csv", TISSUE)))
write_csv(pred, file.path(OUT_DIR, sprintf("suite_prediction_%s.csv", TISSUE)))
write_csv(tasks %>% filter(perm > 0) %>% left_join(scen_tab %>% select(scenario, metric, predictors, target), by = "scenario") %>%
            select(metric, predictors, target, perm, q2), file.path(OUT_DIR, sprintf("suite_prediction_null_%s.csv", TISSUE)))
```

# Figures

```{r figures, fig.height = 9}
tissue_name <- TISSUE_LABELS[[TISSUE]]
arm_lab <- c(control = "Control", OA = "OA", OW = "OW", DO = "DO")
arm_fill_shape <- c(control = 21, OA = 22, OW = 24, DO = 23)
fmt_p <- function(p) ifelse(p < 0.001, formatC(p, format = "e", digits = 1), formatC(p, format = "fg", digits = 2))
strength_fill <- scale_fill_gradient(low = "#f0f0f0", high = THREAD_TRT_COLORS[["baseline"]], trans = "log10",
                                     name = "Mean peak force,\n% of control mean\n(baseline-adjusted)")

## 1. DEG programs: arm means against within-arm slopes, both primary metrics
progs <- axes %>% filter(type == "DEG program")
bw <- bind_rows(lapply(primary, function(m) bind_rows(lapply(seq_len(nrow(progs)), function(i)
  tibble(metric = m, axis_id = progs$axis_id[i], arm = arm, score = S[, progs$axis_id[i]], pct = strength_pct[, paste0(m, "_pct_control")]))))) %>%
  filter(is.finite(pct)) %>%
  left_join(progs %>% select(axis_id, label, n_genes), by = "axis_id") %>%
  mutate(panel = sprintf("%s (%d genes)", label, n_genes), metric_lab = vapply(metric, label_of, ""))
bw$panel <- factor(bw$panel, levels = unique(bw$panel[order(match(bw$axis_id, progs$axis_id))]))
bm <- bw %>% group_by(metric_lab, panel, arm) %>%
  summarise(x = mean(score), sx = sd(score) / sqrt(n()), ly = mean(log(pct)), sy = sd(log(pct)) / sqrt(n()), .groups = "drop") %>%
  mutate(y = exp(ly), lo = exp(ly - sy), hi = exp(ly + sy))
bl <- tests %>% filter(type == "DEG program", metric %in% primary) %>%
  left_join(bw %>% distinct(axis_id, panel, metric, metric_lab), by = c("axis_id", "metric")) %>%
  mutate(txt = sprintf("within arms r = %.2f, p = %s, q = %.2f", partial_r, fmt_p(p_lm), q_lm))
p1 <- ggplot(bw, aes(score, pct)) +
  geom_point(aes(colour = arm, shape = arm), size = 1.4, alpha = 0.5) +
  geom_smooth(aes(colour = arm, group = arm), method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.55) +
  geom_path(data = bm %>% arrange(panel, match(arm, arms_kept)), aes(x, y, group = interaction(panel, metric_lab)),
            colour = "grey35", linetype = "dashed", linewidth = 0.45) +
  geom_errorbar(data = bm, aes(x = x, ymin = lo, ymax = hi, colour = arm), inherit.aes = FALSE, width = 0, linewidth = 0.7) +
  geom_errorbar(data = bm, aes(y = y, xmin = x - sx, xmax = x + sx, colour = arm), inherit.aes = FALSE, width = 0, linewidth = 0.7, orientation = "y") +
  geom_point(data = bm, aes(x, y, fill = arm, shape = arm), inherit.aes = FALSE, colour = "black", size = 2.8, stroke = 0.4) +
  geom_text(data = bl, aes(x = -Inf, y = Inf, label = txt), hjust = -0.03, vjust = 1.3, size = 2.5, colour = "grey25") +
  facet_grid(metric_lab ~ panel, scales = "free") +
  scale_y_log10() +
  scale_colour_manual(values = TREATMENT_COLORS, labels = arm_lab, name = NULL) +
  scale_fill_manual(values = TREATMENT_COLORS, labels = arm_lab, name = NULL) +
  scale_shape_manual(values = arm_fill_shape, labels = arm_lab, name = NULL) +
  labs(x = "Program score (SD units; first principal component of the program's genes)",
       y = "Day-3 level, % of control mean (baseline-adjusted, log scale)",
       title = sprintf("%s: DEG programs between arms (large points, arm means +/- 1 SE, joined in arm order) and within arms (thin lines)", tissue_name),
       caption = "Within-arm r, p and q: ANCOVA level_day3 ~ score + treatment + level_baseline; q over every axis of this script and metric (suite_axis_tests).") +
  theme_psmfc(9) + theme(legend.position = "bottom", strip.text = element_text(face = "bold", size = 7.5),
                         plot.title.position = "plot", plot.caption = element_text(hjust = 0))
ggsave(file.path(OUT_DIR, sprintf("suites_between_within_%s.png", TISSUE)), p1,
       width = 2.3 + 2.2 * nrow(progs), height = 6.4, dpi = 220, limitsize = FALSE)

## 2. every axis: across-arm against within-arm association
type_shape <- c("DEG program" = 21, "GO term" = 22, "Co-expression module" = 24, "Expression PC" = 23)
type_fill  <- c("DEG program" = "black", "GO term" = "grey55", "Co-expression module" = "white", "Expression PC" = "grey85")
mp <- tests %>% filter(tier == "primary") %>% mutate(metric_lab = vapply(metric, label_of, ""))
lab2 <- mp %>% group_by(metric_lab) %>% slice_max(abs(across_r), n = 5, with_ties = FALSE) %>% ungroup() %>%
  mutate(txt = substr(label, 1, 40))
sumtxt <- mp %>% group_by(metric_lab) %>%
  summarise(txt = sprintf("%d axes; within arms lowest q = %.2f, %d with p < 0.05", n(), min(q_lm), sum(p_lm < 0.05)), .groups = "drop")
p2 <- ggplot(mp, aes(across_r, partial_r)) +
  geom_hline(yintercept = 0, colour = "grey70") + geom_vline(xintercept = 0, colour = "grey70") +
  geom_point(aes(shape = type, fill = type, size = n_genes), colour = "grey20", stroke = 0.35, alpha = 0.85) +
  geom_point(data = mp %>% filter(q_lm < FDR_ALPHA), aes(size = n_genes), shape = 1, colour = DIRECTION_COLORS[["up"]], stroke = 1.1, show.legend = FALSE) +
  geom_text_repel(data = lab2, aes(label = txt), size = 2.5, colour = "grey25", min.segment.length = 0, seed = 1, max.overlaps = 20) +
  geom_text(data = sumtxt, aes(x = -Inf, y = Inf, label = txt), hjust = -0.03, vjust = 1.4, size = 3, colour = "grey20") +
  facet_wrap(~ metric_lab) +
  scale_shape_manual(values = type_shape, name = NULL) + scale_fill_manual(values = type_fill, name = NULL) +
  scale_size_continuous(trans = "log10", range = c(1.2, 4.5), breaks = c(10, 100, 1000, 10000), labels = scales::comma, name = "genes") +
  coord_cartesian(xlim = c(-0.75, 0.75), ylim = c(-0.75, 0.75)) +
  labs(x = "Across arms: partial r of the score with the day-3 level, baseline held constant (includes the treatment effect)",
       y = "Within arms: partial r, arm and baseline held constant",
       title = sprintf("%s: every axis, across arms against within arms", tissue_name),
       caption = sprintf("Red ring: within-arm BH q < %.1f over every axis and metric. Labelled: the five axes most associated across arms.", FDR_ALPHA)) +
  theme_psmfc(9.5) + theme(plot.title.position = "plot", plot.caption = element_text(hjust = 0), legend.position = "right")
ggsave(file.path(OUT_DIR, sprintf("suites_axes_map_%s.png", TISSUE)), p2, width = 13, height = 6.2, dpi = 220)

## 3. state space: the DEG programs (principal components of their scores) and the expression PCs
P <- S[, progs$axis_id, drop = FALSE]; pp <- prcomp(P, center = TRUE, scale. = TRUE)
ve_p <- summary(pp)$importance[2, ]
sp <- tibble(arm = arm, a1 = pp$x[, 1], a2 = pp$x[, 2], pc1 = S[, "PC1"], pc2 = S[, "PC2"],
             pct = strength_pct[, paste0(primary[1], "_pct_control")]) %>% filter(is.finite(pct))
ld <- as_tibble(pp$rotation[, 1:2], rownames = "axis_id") %>% left_join(progs %>% select(axis_id, label), by = "axis_id") %>%
  mutate(s = 0.85 * max(abs(c(sp$a1, sp$a2))) / max(sqrt(PC1^2 + PC2^2)))
cent <- function(d, x, y) d %>% group_by(arm) %>% summarise(x = mean({{ x }}), y = mean({{ y }}), .groups = "drop")
state_plot <- function(d, x, y, xl, yl, title, arrows = NULL) {
  g <- ggplot(d, aes({{ x }}, {{ y }})) +
    geom_hline(yintercept = 0, colour = "grey90") + geom_vline(xintercept = 0, colour = "grey90")
  if (!is.null(arrows)) g <- g +
    geom_segment(data = arrows, aes(x = 0, y = 0, xend = PC1 * s, yend = PC2 * s), inherit.aes = FALSE,
                 arrow = arrow(length = unit(5, "pt")), colour = "grey45", linewidth = 0.4) +
    geom_text_repel(data = arrows, aes(x = PC1 * s, y = PC2 * s, label = label), inherit.aes = FALSE, size = 2.6, colour = "grey30", seed = 2)
  g + geom_point(aes(fill = pct, shape = arm), size = 2.6, colour = "grey25", stroke = 0.35) +
    geom_point(data = cent(d, {{ x }}, {{ y }}), aes(x, y, colour = arm), inherit.aes = FALSE, shape = 3, size = 4.5, stroke = 1.3, show.legend = FALSE) +
    strength_fill + scale_shape_manual(values = arm_fill_shape, labels = arm_lab, name = NULL) +
    scale_colour_manual(values = TREATMENT_COLORS) +
    guides(shape = guide_legend(override.aes = list(fill = "grey70"), order = 1)) +
    labs(x = xl, y = yl, title = title) + theme_psmfc(9.5) + theme(plot.title.position = "plot")
}
p3 <- state_plot(sp, a1, a2, sprintf("Programs axis 1 (%.0f%%)", 100 * ve_p[1]), sprintf("Programs axis 2 (%.0f%%)", 100 * ve_p[2]),
                 sprintf("%d DEG programs (arrows)", nrow(progs)), ld) |
  state_plot(sp, pc1, pc2, axes$label[axes$axis_id == "PC1"], axes$label[axes$axis_id == "PC2"],
             sprintf("All %s expressed genes", format(ncol(X_all), big.mark = ",")))
p3 <- p3 + plot_layout(guides = "collect") +
  plot_annotation(title = sprintf("%s: animals in expression space, coloured by attachment; crosses, arm means", tissue_name),
                  caption = with(tests %>% filter(metric == primary[1], type %in% c("DEG program", "Expression PC")),
                                 sprintf("Left: principal components of the %d DEG-program scores. Within arms, the lowest q of these programs and the expression components with %s is %.2f (suite_axis_tests_%s.csv).",
                                         nrow(progs), tolower(label_of(primary[1])), min(q_lm), TISSUE)))
ggsave(file.path(OUT_DIR, sprintf("suites_state_space_%s.png", TISSUE)), p3, width = 13, height = 6, dpi = 220)

## 4. prediction test and the within-arm tests against chance
pr <- pred %>% mutate(row = sprintf("%s | %s", label_of(metric), c(candidates = sprintf("candidates (%d)", length(cand)),
                                     top_variable = sprintf("%s most variable genes", format(length(top_var), big.mark = ",")),
                                     axes = sprintf("all axes (%d)", ncol(S)))[predictors]),
                      target = factor(target, levels = c("across", "within"),
                                      labels = c("Across arms (baseline removed)", "Within arms (arm and baseline removed)")))
p4a <- ggplot(pr, aes(y = row)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_errorbar(aes(xmin = null_lo, xmax = null_hi), width = 0, colour = "grey75", linewidth = 2.2, orientation = "y") +
  geom_point(aes(x = null_median), shape = 124, size = 3, colour = "grey45") +
  geom_point(aes(x = treatment_only_q2), shape = 23, size = 2.6, fill = TREATMENT_COLORS[["control"]], colour = "black", stroke = 0.3, na.rm = TRUE) +
  geom_point(aes(x = q2, fill = p_perm < 0.05), shape = 21, size = 2.8, colour = "black", stroke = 0.4) +
  facet_wrap(~ target) +
  scale_fill_manual(values = c(`TRUE` = "black", `FALSE` = "white"), labels = c(`TRUE` = "beats the permutations (p < 0.05)", `FALSE` = "within the permutation null"), name = NULL, drop = FALSE) +
  labs(x = "Out-of-sample Q2 (elastic net; 0 = no better than the training mean)", y = NULL,
       title = "Do multi-gene profiles predict held-out animals?",
       subtitle = sprintf("Circle: observed; grey bar: 95%% of %d permutations; grey diamond: the treatment alone as predictor", NPERM)) +
  theme_psmfc(9.5) + theme(legend.position = "bottom", plot.title.position = "plot")
qq <- tests %>% filter(tier == "primary") %>% group_by(type) %>% arrange(p_lm, .by_group = TRUE) %>%
  mutate(k = row_number(), n = n(), obs = -log10(p_lm), exp = -log10(k / (n + 1)),
         lo = -log10(qbeta(0.975, k, n - k + 1)), hi = -log10(qbeta(0.025, k, n - k + 1))) %>% ungroup()
p4b <- ggplot(qq, aes(exp, obs)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey90") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 1.3, colour = "grey20") +
  facet_wrap(~ type, nrow = 1, scales = "free_x") +
  labs(x = "expected -log10 p (no association)", y = "observed -log10 p", title = "Within-arm tests against chance, primary metrics",
       subtitle = "Band: 95% of what chance gives") +
  theme_psmfc(9.5) + theme(plot.title.position = "plot")
p4 <- (p4a / p4b) + plot_layout(heights = c(1, 0.9)) + plot_annotation(title = tissue_name)
ggsave(file.path(OUT_DIR, sprintf("suites_prediction_%s.png", TISSUE)), p4, width = 12, height = 9, dpi = 220)
p1
```

# Provenance

```{r provenance}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
settings <- c(
  "[06_expression_suites.Rmd]",
  paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
  paste0("tissue               : ", TISSUE),
  paste0("axes                 : ", paste(sprintf("%s %d", names(table(axes$type)), as.integer(table(axes$type))), collapse = ", ")),
  paste0("modules              : ", length(mods), " from the ", length(top_w), " genes with the largest within-arm variance (min ", MIN_MODULE, " genes)"),
  paste0("within-arm test      : ANCOVA level_day3 ~ score + treatment + level_baseline; across-arm: without treatment"),
  paste0("prediction           : elastic net (alpha 0.5, lambda.min of an inner 5-fold CV), ", CV_REPEATS, " x 5-fold CV balanced by arm, ",
         NPERM, " permutations per scenario, seed ", SEED, ", ", CORES, " worker(s)"),
  paste0("lowest within-arm q  : ", signif(min(tests$q_lm), 3), " (", sum(tests$q_lm < FDR_ALPHA), " of ", nrow(tests), " axis x metric tests below ", FDR_ALPHA, ")"),
  paste0("prediction Q2 range  : ", sprintf("%.3f to %.3f", min(pred$q2), max(pred$q2)), "; lowest permutation p ", signif(min(pred$p_perm), 3),
         "; treatment alone (across): ", paste(sprintf("%s %.3f", names(arm_q2), arm_q2), collapse = ", ")),
  paste0("animals              : ", nrow(X_all)))
inputs <- c(unname(f_in), file.path(repo_root, "09_gene-mechanics-correlation", "01_code", "06_expression_suites.Rmd"),
            file.path(repo_root, "tools", c("pipeline_checks.R", "plot_style.R")))
writeLines(provenance_lines(settings, inputs, repo_root, packages = c("glmnet", "dynamicTreeCut", "dplyr", "ggplot2", "patchwork", "ggrepel")), prov)
cat(readLines(prov), sep = "\n")
```
