---
title: "3_1_venn_diagrams"
author: "Matt George; mngeorge@uw.edu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE, cache = TRUE)
```

```{bash, engine.path="C:/Windows/System32/bash.exe"}
echo $PATH
```

### Load libraries
```{r load_libraries, inlcude = TRUE}

## clear
rm(list=ls())

# List of packages we want to install (run every time)
load.lib<-c("RColorBrewer","readxl","ggpubr","tidyverse","tibble","stringr","beepr","gplots","ggvenn")

# Select only the packages that aren't currently installed (run every time)
install.lib <- load.lib[!load.lib %in% installed.packages()]

# And finally we install the missing packages, including their dependency.
for(lib in install.lib) install.packages(lib,dependencies=TRUE)
# After the installation process completes, we load all packages.
sapply(load.lib,require,character=TRUE)

#Set ggplot theme
my_theme <- theme(line              = element_line(size=1.5),
                  rect              = element_rect(size=1.5),
                  text              = element_text(size=14,color="black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(), 
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size=16,color="black"),
                  axis.text.y       = element_text(size=16,color="black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color="black"),
                  axis.ticks.y      = element_line(color="black"),
                  # axis.line         = element_line(color = "black", size = 0.1),
                  panel.border      = element_rect(color = "black", fill=NA, size=1.5),
                  legend.key        = element_blank()) # removes background of legend bullets

blank_theme <-  theme_minimal()+
                theme(
                axis.title.x = element_blank(),
                axis.title.y = element_blank(),
                panel.border = element_blank(),
                panel.grid=element_blank(),
                axis.ticks = element_blank(),
                plot.title=element_text(size=14, face="bold")
                )
```

#Load data
```{r}
dat <- read_excel("data/venn_diagram.xlsx", sheet = "list", col_names = TRUE)

trip_C  <- na.omit(dat$trip_C)
dip_SS  <- na.omit(dat$dip_SS)
dip_MS  <- na.omit(dat$dip_MS)
trip_SS <- na.omit(dat$trip_SS)
trip_MS <- na.omit(dat$trip_MS)


```


```{r}
## Venn diagram

# venn_list <- gene_list_unique[,c(5:8)]
# venn(venn_list)

## Compare 4 treatments in one VENN
venn_list <- list(
              dip_MS    = dip_MS,
              dip_SS    = dip_SS, 
              trip_SS   = trip_SS,
              trip_MS   = trip_MS)


p1 <- ggvenn(venn_list , fill_alpha = 0.6, show_percentage = TRUE, text_size = 2.5, text_color = "white") + 
             scale_fill_manual(values = c("blue","royalblue1","orangered1","red")) + my_theme
p1

# Save plots
ggsave("analyses/DEG_venn_diagram.png",
       plot   = p1,
       dpi    = 600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")


```

```{r}
dat <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "up_down", col_names = TRUE)

dat$treatment <- factor(dat$treatment, levels = c("trip_C","dip_SS","dip_MS","trip_SS","trip_MS"), ordered = TRUE)


p1 <- ggplot(dat, aes(x=treatment, y=count, fill=direction)) +
      geom_bar(stat="identity", position="dodge") + 
      scale_fill_manual(values = c("steelblue2","orangered2")) +
      scale_x_discrete(expand = c(0,1)) +
      scale_y_continuous(limits = c(0,2800), breaks = seq(0,2500,500)) + my_theme

ggsave("analyses/up_down.png",
       plot   = p1,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 6,
       units  = "in")

``` 

Note: This code assumes that you have loaded the ggplot2 package. If you haven't, you can do so by running the command `install.packages("ggplot2")` and then `library(ggplot2)`. Also, if the variable names in your data frame are different, simply replace 'treatment', 'direction' and 'count' with the actual variable names in your data frame.


```


```{r}
## Compare ploidy across multiple stressor (v control condition)
venn_list_multi <- list(
                    dip_desiccation  = diploid_desiccation$geneID,
                    trip_desiccation = triploid_desiccation$geneID)

p2 <- ggvenn(venn_list_multi, fill_alpha = 0.6, show_percentage = TRUE, text_size = 2.5, text_color = "white") + 
             scale_fill_manual(values = c("blue","red"))
p2

## Compare ploidy across single stressor (v control condition)
venn_list_single <- list(
                    dip_heat         = diploid_heat$geneID,
                    trip_heat        = triploid_heat$geneID)

p3 <- ggvenn(venn_list_single, fill_alpha = 0.6, show_percentage = TRUE, text_size = 2.5, text_color = "white") + 
             scale_fill_manual(values = c("blue","red"))
p3

## Unique DEGs between diploid and triploid single stressor
diploid_heat_unique  <- anti_join(diploid_heat, triploid_heat, by = "geneID") %>% arrange(log2FoldChange)
triploid_heat_unique <- anti_join(triploid_heat, diploid_heat, by = "geneID") %>% arrange(log2FoldChange)
write.table(diploid_heat_unique, file = "output/filtered/HISAT2_multiqc_biplot/DEG_diploid_single_unique.csv", row.names = FALSE)
write.table(triploid_heat_unique, file = "output/filtered/HISAT2_multiqc_biplot/DEG_triploid_single_unique.csv", row.names = FALSE)

## Unique DEGs between diploid and triploid multi-stressor
diploid_desiccation_unique  <- anti_join(diploid_desiccation, triploid_desiccation, by = "geneID") %>% arrange(log2FoldChange)
triploid_desiccation_unique <- anti_join(triploid_desiccation, diploid_desiccation, by = "geneID") %>% arrange(log2FoldChange)
write.table(diploid_desiccation_unique, file = "output/filtered/HISAT2_multiqc_biplot/DEG_diploid_multi_unique.csv", row.names = FALSE)
write.table(triploid_desiccation_unique, file = "output/filtered/HISAT2_multiqc_biplot/DEG_triploid_multi_unique.csv", row.names = FALSE)

### Pie chart of unknown of up or down expression
dip_count_plus <- as.numeric(nrow(diploid_desiccation_unique[diploid_desiccation_unique$plus_minus == "1",]))
dip_count_minus <- as.numeric(nrow(diploid_desiccation_unique[diploid_desiccation_unique$plus_minus == "-1",]))
trip_count_plus <- as.numeric(nrow(triploid_desiccation_unique[triploid_desiccation_unique$plus_minus == "1",]))
trip_count_minus <- as.numeric(nrow(triploid_desiccation_unique[triploid_desiccation_unique$plus_minus == "-1",]))

bp1  <- ggplot(diploid_desiccation_unique, aes(x=factor(1),y=plus_minus, fill=as.character(plus_minus))) +
        geom_bar(width = 1, stat = "identity", show.legend = TRUE) +
        scale_fill_manual(values = c("indianred","lightskyblue")) + labs(NULL) +
        coord_polar("y", start=0) + blank_theme + theme(axis.title.x=element_blank(),axis.title.y=element_blank())
bp1

bp2  <- ggplot(triploid_desiccation_unique, aes(x=factor(1),y=plus_minus, fill=as.character(plus_minus))) +
        geom_bar(width = 1, stat = "identity", show.legend = TRUE) +
        scale_fill_manual(values = c("indianred","lightskyblue")) + labs(NULL) +
        coord_polar("y", start=0) + blank_theme + theme(axis.title.x=element_blank(),axis.title.y=element_blank())
bp2

bp3  <- ggplot(diploid_heat_unique, aes(x=factor(1),y=plus_minus, fill=as.character(plus_minus))) +
        geom_bar(width = 1, stat = "identity", show.legend = TRUE) +
        scale_fill_manual(values = c("indianred","lightskyblue")) + labs(NULL) +
        coord_polar("y", start=0) + blank_theme + theme(axis.title.x=element_blank(),axis.title.y=element_blank())
bp3

bp4  <- ggplot(triploid_heat_unique, aes(x=factor(1),y=plus_minus, fill=as.character(plus_minus))) +
        geom_bar(width = 1, stat = "identity", show.legend = TRUE) +
        scale_fill_manual(values = c("indianred","lightskyblue")) + labs(NULL) +
        coord_polar("y", start=0) + blank_theme + theme(axis.title.x=element_blank(),axis.title.y=element_blank())
bp4

### Characterized sigs bar chart
diploid_desiccation_unique_plot <- diploid_desiccation_unique %>% filter(cat == "characterized") %>% arrange(log2FoldChange)
diploid_desiccation_unique_plot$gene <- factor(diploid_desiccation_unique_plot$gene, levels = diploid_desiccation_unique_plot$gene, ordered = TRUE) 

triploid_desiccation_unique_plot <- triploid_desiccation_unique %>% filter(cat == "characterized") %>% arrange(desc(log2FoldChange))
triploid_desiccation_unique_plot$gene <- factor(triploid_desiccation_unique_plot$gene, levels = triploid_desiccation_unique_plot$gene, ordered = TRUE)

num_keep <- nrow(diploid_desiccation_unique_plot)
if(nrow(diploid_desiccation_unique_plot)>20){num_keep <- 10} else {num_keep <- nrow(diploid_desiccation_unique_plot)/2}
rows_keep <- c(seq(1:num_keep), seq((nrow(diploid_desiccation_unique_plot)-num_keep+1), nrow(diploid_desiccation_unique_plot)))

bp5 <- ggplot(diploid_desiccation_unique_plot[rows_keep,], 
              aes(x=diploid_desiccation_unique_plot$gene[rows_keep], y=log2FoldChange, fill=as.character(plus_minus))) +
              geom_bar(width = 1, stat = "identity", position = "dodge", color="black", size = 0.3, show.legend = F) +
              # scale_fill_discrete(values=c("salmon1","royalblue1")) +
              # scale_y_continuous(breaks = seq(-4, 8, by = 2), limits = c(-5.5,9)) +
              # scale_x_continuous(breaks = NULL) + xlab("") +
              coord_flip() + my_theme
bp5

num_keep <- nrow(triploid_desiccation_unique_plot)
rows_keep = seq(1:num_keep)
if(nrow(triploid_desiccation_unique_plot)>20){num_keep <- 10} else {num_keep <- nrow(triploid_desiccation_unique_plot)/2}
rows_keep <- c(seq(1:num_keep), seq((nrow(triploid_desiccation_unique_plot)-num_keep+1), nrow(triploid_desiccation_unique_plot)))

bp6 <- ggplot(triploid_desiccation_unique_plot[rows_keep,], 
              aes(x=triploid_desiccation_unique_plot$gene[rows_keep], y=log2FoldChange, fill=as.character(plus_minus))) +
              geom_bar(width = 1, stat = "identity", position = "dodge", color="black", size = 0.3, show.legend = F) +
              scale_fill_manual(values=c("salmon1","royalblue1")) +
              # scale_y_continuous(breaks = seq(-4, 8, by = 2), limits = c(-5.5,9)) +
              # scale_x_continuous(breaks = NULL) + xlab("") +
              coord_flip() + my_theme
bp6

# Save plots
ggsave("output/filtered/HISAT2_multiqc_biplot/DEG_venn_diagram.png",
       plot   = p1,
       dpi    = 600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/DEG_venn_diagram_multi_stressor.png",
       plot   = p2,
       dpi    = 600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/DEG_venn_diagram_single_stressor.png",
       plot   = p3,
       dpi    = 600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/diploid_desiccation_unique_DEG_pie.png",
       plot   = bp1,
       dpi    = 1600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/triploid_desiccation_unique_DEG_pie.png",
       plot   = bp2,
       dpi    = 1600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/diploid_heat_unique_DEG_pie.png",
       plot   = bp3,
       dpi    = 1600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/triploid_heat_unique_DEG_pie.png",
       plot   = bp4,
       dpi    = 1600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/diploid_desiccation_unique_DEG_bar.png",
       plot   = bp5,
       dpi    = 600,
       device = "png",
       width  = 6,
       height = 4,
       units  = "in")

ggsave("output/filtered/HISAT2_multiqc_biplot/triploid_desiccation_unique_DEG_bar.png",
       plot   = bp6,
       dpi    = 600,
       device = "png",
       width  = 6,
       height = 6,
       units  = "in")

```

```{r}

# nrow(control_ploidy[control_ploidy$plus_minus == '1', ])

total <- rbind(control_ploidy, 
               heat_ploidy,
               desiccation_ploidy,
               diploid_heat,
               diploid_desiccation,
               triploid_heat,
               triploid_desiccation)

total$analysis <- factor(total$analysis, levels = c('control_ploidy', 'heat_ploidy', 'desiccation_ploidy', 'diploid_heat',
                                                    'diploid_desiccation', 'triploid_heat', 'triploid_desiccation'), ordered = T)


total$plus_minus <- factor(total$plus_minus, levels = c('1','-1'), ordered = T)

bp5 <- ggplot(total, aes(x=analysis, y=count, fill=plus_minus)) +
        geom_bar(stat = "identity", show.legend = F) +
        scale_fill_manual(values = c("royalblue3","salmon3")) +
        scale_y_continuous(breaks = seq(0, 5000, by = 1000), limits = c(0,5500)) +
        my_theme

bp5

ggsave("output/filtered/HISAT2_multiqc_biplot/DEG_expression_barplot.png",
       plot   = bp5,
       dpi    = 600,
       device = "png",
       width  = 7,
       height = 5,
       units  = "in")

```
