---
title: "00 - Run the whole analysis"
author: "PSMFC Mytilus byssus pilot"
date: "`r Sys.Date()`"
output: html_document
params:
  stages: "all"              # "all", or folder numbers, e.g. ["06", "07"]
  stop_on_fail: true         # stop at the first failed stage
  cores: 4                   # parallel GO runs in 07 (1 on Windows)
  online_annotation: false   # true also refreshes 08's NCBI and OrthoDB tables
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs the analysis from thread mechanics and Tag-seq counts through differential expression,
GO enrichment and the gene-mechanics associations, by knitting each folder's own runner in
order, each in a fresh R process. Every runner leaves an HTML report and a log per script in
its folder's `03_analyses/knit_html/`; this file leaves one per stage, and `run_log.csv`, in
`knit_html/` at the repository root (all git-ignored).

| stage | runner | needs | about |
|---|---|---|---|
| 02 | `02_thread-strength/01_code/00_run_thread_strength.Rmd` | `01_mussel-measurements/` | 1 min |
| 05 | `05_sequence-alignment/01_code/00_run_sequence_alignment.Rmd` | committed StringTie and FastQC tables | 1 min |
| 06 | `06_differential-expression/01_code/00_run_differential_expression.Rmd` | 05 | 10 min |
| 04 | `04_iso-seq-transcriptome/01_code/00_run_isoseq.Rmd` | 06, committed Iso-Seq and recount gene counts | 12 min |
| 07 | `07_enrichment/01_code/00_run_enrichment.Rmd` | 06, 05, `03_blast/` | 8 min |
| 08 | `08_gene-annotation/01_code/00_run_gene_annotation.Rmd` | 06 | 1 min |
| 09 | `09_gene-mechanics-correlation/01_code/00_run_gene_mechanics_by_tissue.Rmd` | 02, 06, 07, `03_blast/` | 4 min |

Stage 04 runs after 06 because it repeats 06's contrasts on the Iso-Seq reference (step 04)
and on option B's recounts of the genome (step 07) and compares them with 06; both read
committed gene counts, so only those two steps run here.

Not run here, because they need data or machines outside this repository: the BLAST
annotation (`03_blast/`, HPC), the Iso-Seq steps 01-03 and 05-06 (`04_iso-seq-transcriptome/`,
which download the transcriptome, the genome and the reads and run minimap2, salmon, HISAT2,
StringTie and featureCounts; set that runner's `online: true`), HISAT2 + StringTie (`05_sequence-alignment` step 01, HPC), 05 step 03's
recipe check and, unless `online_annotation` is true, 08's NCBI and OrthoDB queries. Their
committed outputs are what the stages above read.

Knit this file from the repository root (`PSMFC-mytilus-byssus-pilot.Rproj`). The packages
each stage needs are listed in its folder's README.

```{r run}
suppressPackageStartupMessages(library(here))
repo_root <- here::here()
source(file.path(repo_root, "tools", "run_steps.R"))
html_dir <- file.path(repo_root, "knit_html")

stages <- data.frame(
  stage  = c("02", "05", "06", "04", "07", "08", "09"),
  runner = c("02_thread-strength/01_code/00_run_thread_strength.Rmd",
             "05_sequence-alignment/01_code/00_run_sequence_alignment.Rmd",
             "06_differential-expression/01_code/00_run_differential_expression.Rmd",
             "04_iso-seq-transcriptome/01_code/00_run_isoseq.Rmd",
             "07_enrichment/01_code/00_run_enrichment.Rmd",
             "08_gene-annotation/01_code/00_run_gene_annotation.Rmd",
             "09_gene-mechanics-correlation/01_code/00_run_gene_mechanics_by_tissue.Rmd"))
stage_params <- list(`07` = list(cores = params$cores),
                     `08` = list(online = isTRUE(params$online_annotation)))
want <- if (identical(params$stages, "all")) stages$stage else sprintf("%02d", as.integer(unlist(params$stages)))
stopifnot(all(want %in% stages$stage))

log <- list()
for (i in which(stages$stage %in% want)) {
  s <- stages[i, ]
  message(sprintf("==== stage %s: %s ====", s$stage, s$runner))
  t0  <- Sys.time()
  out <- sprintf("stage_%s.html", s$stage)
  ok  <- render_step(file.path(repo_root, s$runner), out, html_dir, stage_params[[s$stage]])
  if (!ok) message("FAILED; see ", file.path(html_dir, sub("[.]html$", ".log", out)),
                   " and the logs in that folder's 03_analyses/knit_html/")
  log[[length(log) + 1]] <- data.frame(stage = s$stage, runner = s$runner, ok = ok,
                                       minutes = round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1))
  if (!ok && isTRUE(params$stop_on_fail)) { message("Stopping (stop_on_fail = TRUE)."); break }
}
run_log <- do.call(rbind, log)
names(run_log)[names(run_log) == "runner"] <- "script"
finish_run(run_log, html_dir, "Pipeline")
```

# Where the results are

- Thread strength: `02_thread-strength/03_analyses/04_analyze-thread-strength/` and
  `05_decompose-adhesion/`.
- Differential expression: `06_differential-expression/03_analyses/DEG_lists/DEG_counts.csv`
  and `figures/`.
- Iso-Seq reference (sensitivity): `04_iso-seq-transcriptome/03_analyses/04_isoseq-de/`
  (`reference_agreement.csv`); option B, the genome recounted on Iso-Seq-augmented
  annotations (parallel analysis): `04_iso-seq-transcriptome/03_analyses/07_augmented-de/`
  (`deg_summary.csv`).
- GO enrichment: `07_enrichment/03_analyses/02_topgo/` (of record) and
  `06_method-comparison/consensus_terms_TC_BP.csv`.
- GO slims: `08_gene-annotation/03_analyses/goslims/`.
- Gene x mechanics: `09_gene-mechanics-correlation/03_analyses/gene_mechanics/`, with
  `RUN_provenance_<T>.txt` recording how many checks ran and which failed.
