---
title: "00 - Run the whole analysis"
author: "PSMFC Mytilus byssus pilot"
date: "`r Sys.Date()`"
output: html_document
params:
  stages: "all"              # "all", or folder numbers, e.g. ["05", "07"]
  stop_on_fail: true         # stop at the first failed stage
  cores: 4                   # parallel workers: GO runs in 07 (1 on Windows) and 09 step 06's prediction test
  online_annotation: false   # true also refreshes 08's NCBI and OrthoDB tables
  check_versions: true       # stop before the first stage unless R, Bioconductor and GO.db match renv.lock
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs the analysis from thread mechanics and Tag-seq counts through differential expression,
GO enrichment and the gene-mechanics associations, by knitting each folder's own runner in
order, each in a fresh R process. Every runner leaves an HTML report and a log per script in
its folder's `03_analyses/knit_html/`; this file leaves one per stage, and `run_log.csv`, in
`knit_html/` at the repository root (all git-ignored).

| stage | runner | needs | about |
|---|---|---|---|
| 02 | `02_thread-strength/01_code/00_run_thread_strength.Rmd` | `01_mussel-measurements/` | 1 min |
| 04 | `04_sequence-alignment/01_code/00_run_sequence_alignment.Rmd` | committed StringTie and FastQC tables, its committed featureCounts and mitogenome counts | 1 min |
| 05 | `05_differential-expression/01_code/00_run_differential_expression.Rmd` | 04, `03_blast/` | 10 min |
| 06 | `06_iso-seq-transcriptome/01_code/00_run_isoseq.Rmd` | 05, committed Iso-Seq and recount gene counts | 12 min |
| 07 | `07_enrichment/01_code/00_run_enrichment.Rmd` | 05, 04, `03_blast/` | 8 min |
| 08 | `08_gene-annotation/01_code/00_run_gene_annotation.Rmd` | 05 | 1 min |
| 09 | `09_gene-mechanics-correlation/01_code/00_run_gene_mechanics_by_tissue.Rmd` | 02, 05, 07, `03_blast/` | 4 min |

The stages run in folder order: every folder reads only lower-numbered folders.

The count matrix of record (featureCounts on the RefSeq annotation with Iso-Seq-extended 3'
ends) is built by `04_sequence-alignment` steps 04 to 06, which download the Iso-Seq
transcriptome, the genome and the reads, and so are not run here; stage 04 takes step 06's
committed counts (04 step 07). Stage 06 repeats 05's contrasts on the Iso-Seq reference (step
03) and on the six recounts of the genome (step 04), compares them with 05 and with the
previous record, and checks that its fit of the record reproduces 05's; both read committed
gene counts, so only those two steps run here.

Not run here, because they need data or machines outside this repository: the BLAST
annotation (`03_blast/`, HPC); HISAT2 + StringTie (`04_sequence-alignment` step 01, HPC);
`04_sequence-alignment` steps 04 to 06 and the mitochondrial alignment of its step 08 (they
download the transcriptome, the genome and the reads and run minimap2, HISAT2, StringTie and
featureCounts; set that runner's `online: true`; offline, step 08 summarises the committed
counts); 04 step 03's recipe check (its own `online: true`); the Iso-Seq steps 01 and 02
(`06_iso-seq-transcriptome/`, salmon; set that runner's `online: true`); and, unless
`online_annotation` is true, 08's NCBI and OrthoDB queries. Their committed outputs are what the
stages above read.

Knit this file from the repository root (`PSMFC-mytilus-byssus-pilot.Rproj`). The packages
each stage needs are listed in its folder's README; `renv.lock` records the library of the
runs of record. Before the first stage, `check_stack()` (`tools/run_steps.R`) compares this R
with `renv.lock` and stops within seconds, listing what differs, if R's minor version or
Bioconductor's is another, if a package the pipeline loads is not installed in any library on
R's path (it then prints `.libPaths()`), or, when stage 07 or 08 is to run, if `GO.db` does not
hold GO release 2026-01-23; without it a run on another R would stop only at stage 07, after half an
hour. It also reports `renv.lock` packages that are missing or at another version, without
stopping. `check_versions: false` skips it (outputs are then not of record). When a stage
fails, the error quotes the failed step's own error from its log.

```{r run}
suppressPackageStartupMessages(library(here))
repo_root <- here::here()
source(file.path(repo_root, "tools", "run_steps.R"))
html_dir <- file.path(repo_root, "knit_html")

stages <- data.frame(
  stage  = c("02", "04", "05", "06", "07", "08", "09"),
  runner = c("02_thread-strength/01_code/00_run_thread_strength.Rmd",
             "04_sequence-alignment/01_code/00_run_sequence_alignment.Rmd",
             "05_differential-expression/01_code/00_run_differential_expression.Rmd",
             "06_iso-seq-transcriptome/01_code/00_run_isoseq.Rmd",
             "07_enrichment/01_code/00_run_enrichment.Rmd",
             "08_gene-annotation/01_code/00_run_gene_annotation.Rmd",
             "09_gene-mechanics-correlation/01_code/00_run_gene_mechanics_by_tissue.Rmd"))
stage_params <- list(`07` = list(cores = params$cores), `09` = list(cores = params$cores),
                     `08` = list(online = isTRUE(params$online_annotation)))
want <- if (identical(params$stages, "all")) stages$stage else sprintf("%02d", as.integer(unlist(params$stages)))
stopifnot(all(want %in% stages$stage))
if (isTRUE(params$check_versions)) {
  source(file.path(repo_root, "tools", "pipeline_checks.R"))   # GO_RELEASE_OF_RECORD
  check_stack(file.path(repo_root, "renv.lock"), need_go = any(c("07", "08") %in% want))
}

log <- list()
for (i in which(stages$stage %in% want)) {
  s <- stages[i, ]
  message(sprintf("==== stage %s: %s ====", s$stage, s$runner))
  t0  <- Sys.time()
  out <- sprintf("stage_%s.html", s$stage)
  ok  <- render_step(file.path(repo_root, s$runner), out, html_dir, stage_params[[s$stage]])
  if (!ok) message("FAILED; see ", file.path(html_dir, sub("[.]html$", ".log", out)),
                   " and the logs in that folder's 03_analyses/knit_html/")
  log[[length(log) + 1]] <- data.frame(stage = s$stage, runner = s$runner, ok = ok,
                                       minutes = round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1),
                                       log = sub("[.]html$", ".log", out))
  if (!ok && isTRUE(params$stop_on_fail)) { message("Stopping (stop_on_fail = TRUE)."); break }
}
run_log <- do.call(rbind, log)
names(run_log)[names(run_log) == "runner"] <- "script"
finish_run(run_log, html_dir, "Pipeline")
```

# Where the results are

- Thread strength: `02_thread-strength/03_analyses/04_analyze-thread-strength/` and
  `05_decompose-adhesion/`.
- Differential expression: `05_differential-expression/03_analyses/DEG_lists/DEG_counts.csv`
  and `figures/`.
- Iso-Seq reference (sensitivity): `06_iso-seq-transcriptome/03_analyses/03_isoseq-de/`
  (`reference_agreement.csv`); option B, the genome recounted on Iso-Seq-augmented
  annotations (parallel analysis): `06_iso-seq-transcriptome/03_analyses/04_augmented-de/`
  (`deg_summary.csv`).
- GO enrichment: `07_enrichment/03_analyses/02_topgo/` (of record) and
  `06_method-comparison/consensus_terms_TC_BP.csv`.
- GO slims: `08_gene-annotation/03_analyses/goslims/`.
- Gene x mechanics: `09_gene-mechanics-correlation/03_analyses/gene_mechanics/`, with
  `RUN_provenance_<T>.txt` recording how many checks ran and which failed.
