---
title: "01 - Genome CDS against Swiss-Prot and the Mytilus foot proteins (BLAST annotation)"
output: html_document
params:
  run: false              # true: build the database, run blastx and rebuild the tables (lab server)
  work_dir: "/home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis"   # holds data/ and output/
  blast_bin: "/home/shared/ncbi-blast-2.15.0+/bin"
  threads: 20
  sprot_release: "2024_04"  # the release the 2024 database held (see "The BLAST database"); "2026_03" for the 2026 run
  foot_fasta: "uniprotkb_mytilus_foot.fasta"   # in 02_data/: the copy the 2024 run used; uniprotkb_mytilus_foot_2026_03.fasta for the 2026 run
  res_folder: "genome-foot"  # in 03_analyses/: the 2024 tables; genome-foot-sprot2026_03 for the 2026 run
  parts: 1                # blastx in this many resumable parts (1 = one run, as in 2024; see _blastx_parts.sh)
  jobs: 1                 # parts searched at a time, threads / jobs threads each
  as_2024: false          # true: keep the two quirks of the 2024 run (see "Reproducing the 2024 tables")
  blastx_tab: ""          # an existing blastx table to rebuild the tables from ("" = the one this run writes)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Annotates the genome's coding sequences: every CDS of RefSeq `GCF_036588685.1`
(`cds_from_genomic.fasta`) is searched with blastx (BLAST+ 2.15.0; Camacho et al. 2009,
*BMC Bioinformatics* 10:421) against UniProtKB/Swiss-Prot (`sprot_release`; see below) plus the UniProt
"mytilus foot" proteins (`02_data/uniprotkb_mytilus_foot.fasta`, the copy the 2024 run used),
e-value at most 1e-20, one target sequence per query. Each hit is joined to its UniProt record
(protein names, gene names, GO terms) and each CDS to its gene (`LOC` ID). It was run twice: in
2024 on the lab server (Swiss-Prot 2024_04, the committed 2024 foot proteins; outputs in
`03_analyses/genome-foot/`), and in October 2026 in the cloud environment (Swiss-Prot 2026_03,
the 196 foot proteins of that release's query plus 45 byssal proteins it misses,
`foot_fasta: uniprotkb_mytilus_foot_2026_03_byssal.fasta` (`_build_foot_fasta.R`), 200 parts on
4 threads; outputs in `03_analyses/genome-foot-sprot2026_03/`, with the blastx table
and `RUN_provenance.txt`). Writes, in `03_analyses/<res_folder>/`:

- `LOC_GO_list.txt`: one row per blastx hit (HSP): the CDS (`transcript`), the hit split into
  `sp`, `uniprot_accession`, `geneID` and `species`, the alignment columns, the UniProt record,
  and from `g.spid.txt` the CDS's gene (`gene`), the e-value of its first hit (`V11`) and the
  full subject ID (`V2.x`). Read by `04_sequence-alignment` steps 04 and 05 and
  `05_differential-expression` step 01 (the mitochondrial loci, `tools/mt_encoded.R`),
  `05_differential-expression` step 06, `06_iso-seq-transcriptome` step 04, `07_enrichment`
  step 01 and `09_gene-mechanics-correlation` steps 01 and 04;
- `g.spid.txt`: one row per gene and UniProt accession: the gene, the first hit's e-value, the
  subject ID, the CDS and the accession.

It ran on the lab server in June to August 2024; the committed tables are its outputs. Until
2026-10-03 the script did not run as written (it joined an object, `masterID`, that it never
made, used a misspelt name for the foot-protein table, read the UniProt downloads from a folder
they were not saved in, and loaded no packages). It is fixed here so that it does, and it
reproduces the committed tables (below). With `run: false` (the default) it only checks the
committed tables against the 2024 blastx table, if one is given.

```{r paths}
suppressPackageStartupMessages({ library(dplyr); library(stringr); library(tidyr) })
data_dir <- file.path(params$work_dir, "data")
out_dir  <- file.path(params$work_dir, "output")
cds_fa   <- file.path(data_dir, "ncbi_dataset", "data", "GCF_036588685.1", "cds_from_genomic.fasta")
rel      <- params$sprot_release
foot_fa  <- here::here("02_data", params$foot_fasta)                     # the foot proteins searched
sprot_fa <- file.path(data_dir, paste0("uniprot_sprot_r", rel, ".fasta"))
db       <- file.path(data_dir, paste0("swissprot_n_mytilusfoot_r", rel))   # 2024: swissprot_n_mytilusfoot_db
## the working files are named by release (the 2024 run wrote genome_n_foot_blastx.tab,
## uniprot2024.tsv and trembl_foot_info.tsv)
blastx   <- if (nzchar(params$blastx_tab)) params$blastx_tab else file.path(out_dir, paste0("genome_n_foot_blastx_r", rel, ".tab"))
sprot_info_tsv <- file.path(data_dir, paste0("uniprot_reviewed_r", rel, ".tsv"))     # reviewed UniProt records
foot_info_tsv  <- file.path(data_dir, paste0("uniprot_mytilus_foot_r", rel, ".tsv")) # the "mytilus foot" records
res_dir  <- here::here("03_analyses", params$res_folder)
repo_root <- normalizePath(file.path(here::here(), ".."))
## the bash chunks read these from the environment (knitr does not evaluate inline R code in a
## bash chunk)
Sys.setenv(DATA_DIR = data_dir, REL = rel, DB = db, SPROT_FA = sprot_fa, FOOT_FA = foot_fa, CDS_FA = cds_fa,
           BLAST_BIN = params$blast_bin, BLASTX = blastx, PARTS_SH = here::here("01_code", "_blastx_parts.sh"),
           THREADS = params$threads, PARTS = params$parts, JOBS = params$jobs,
           SPROT_INFO = sprot_info_tsv, FOOT_INFO = foot_info_tsv)
```

# The BLAST database and the search (lab server)

Swiss-Prot of the pinned release comes from UniProt's archive of previous releases
(`previous_releases/release-<release>/knowledgebase/uniprot_sprot-only<release>.tar.gz`, which
holds `uniprot_sprot.fasta.gz`; checked 2026-10-03), or, while the release is the current one
and not yet archived, from `current_release` (the chunk stops if the release is neither). The
database and its FASTA are made once; a rerun reuses them. The 2024 script downloaded
`current_release` and saved it as `uniprot_sprot_r2024_01.fasta`, but its hits match release
2024_04 (UniProt release notes: 24 July 2024) and no other: of the releases from 2023_05 to
2024_05, 2024_04 is the only one that holds all 10,740 subjects of the 2024 hits as written
(`sp|accession|entry name`, with the foot proteins); 2024_03 lacks 3 of the accessions and
2024_05 has renamed 15 entries, and every hit's alignment lies within its 2024_04 sequence. The
default `sprot_release` is therefore 2024_04. The foot proteins are the committed copy, because
the UniProt query returns other entries as UniProt changes (196 on 2026-10-03, against the 202
of the 2024 copy).

```{bash database, eval = params$run}
set -euo pipefail
cd "$DATA_DIR"
rel="$REL"
uni="https://ftp.uniprot.org/pub/databases/uniprot"
if [ ! -f "$DB.pin" ]; then
  if curl -sfI "${uni}/previous_releases/release-${rel}/knowledgebase/uniprot_sprot-only${rel}.tar.gz" > /dev/null; then
    curl -sS -o "uniprot_sprot-only${rel}.tar.gz" "${uni}/previous_releases/release-${rel}/knowledgebase/uniprot_sprot-only${rel}.tar.gz"
    tar -xzf "uniprot_sprot-only${rel}.tar.gz" uniprot_sprot.fasta.gz
  else
    cur=$(curl -sS "${uni}/current_release/knowledgebase/complete/reldate.txt" | sed -n 's/^UniProtKB\/Swiss-Prot Release \([0-9_]*\) .*$/\1/p')
    [ "$cur" = "$rel" ] || { echo "Swiss-Prot release ${rel} is neither archived nor current (current: ${cur})" >&2; exit 1; }
    curl -sS -o uniprot_sprot.fasta.gz "${uni}/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz"
  fi
  gunzip -c uniprot_sprot.fasta.gz > "$SPROT_FA"
  cat "$SPROT_FA" "$FOOT_FA" > "swissprot_n_mytilusfoot_r${rel}.fasta"
  "$BLAST_BIN/makeblastdb" -in "swissprot_n_mytilusfoot_r${rel}.fasta" -dbtype prot -out "$DB"
fi
"$BLAST_BIN/blastdbcmd" -db "$DB" -info | sed -n 1,3p
```

The search is `blastx -evalue 1E-20 -max_target_seqs 1 -outfmt 6` of every CDS against the
database, run by `_blastx_parts.sh`: with `parts: 1` as one command, as in 2024; with more, in
that many parts of the CDS FASTA (in order), each written whole or not at all, so that an
interrupted run resumes where it stopped (knit again, or run the helper with the same
arguments), and joined in order at the end. Each query's hits do not depend on the other
queries, so the parts give the table one run gives (checked on 99 CDSs, byte for byte).

```{bash blastx, eval = params$run}
bash "$PARTS_SH" "$BLAST_BIN" "$CDS_FA" "$DB" \
  "$BLASTX" $THREADS $PARTS $JOBS
```

# The UniProt records of the hits (lab server)

The two record tables of the 2024 run, with the same fields: the reviewed entries (Swiss-Prot)
and the foot proteins searched (in 2024 the entries of the "mytilus foot" query, which were the
proteins of its foot FASTA; now the records of every accession in `foot_fasta`, so that proteins
added to the query's, such as the 45 byssal proteins of the 2026 search, have theirs). UniProt's REST service serves the current release
only, so a new download differs from the 2024 one in its annotation: on 2026-10-03 (release
2026_03) the GO terms of 13,667 of the 16,675 genes' best hits had changed (28,150 gene-GO
pairs added, 31,946 removed). The column names are those of 2024 (checked the same day), so the
tables the analysis reads keep their layout.

```{bash uniprot, eval = params$run}
set -euo pipefail
cd "$DATA_DIR"
served=$(curl -sS -D - -o /dev/null "https://rest.uniprot.org/uniprotkb/search?query=accession:P69905&size=1" \
  | tr -d '\r' | sed -n 's/^[Xx]-[Uu]ni[Pp]rot-[Rr]elease: *//p')
[ "$served" = "$REL" ] || { echo "UniProt serves release ${served}; the records would not be of release $REL" >&2; exit 1; }
curl -sS -H "Accept: text/plain; format=tsv" "https://rest.uniprot.org/uniprotkb/stream?compressed=true&fields=accession%2Creviewed%2Cid%2Cprotein_name%2Cgene_names%2Corganism_name%2Clength%2Cgene_oln%2Ccc_interaction%2Ccc_induction%2Cgo_p%2Cgo_c%2Cgo_f%2Cgo_id%2Cxref_pathwaycommons%2Cxref_reactome%2Cxref_unipathway&format=tsv&query=%28*%29+AND+%28reviewed%3Atrue%29" \
  | gunzip -c > "$SPROT_INFO"
## the foot proteins' records, by accession: those of the FASTA searched (in 2024 the records of
## the "(mytilus foot)" query, which were the proteins of the FASTA then)
grep "^>" "$FOOT_FA" | cut -d"|" -f2 > foot_accessions.txt
rm -f "$FOOT_INFO" foot_acc_part_*
split -l 400 foot_accessions.txt foot_acc_part_
for f in foot_acc_part_*; do
  curl -sS "https://rest.uniprot.org/uniprotkb/accessions?accessions=$(paste -sd, "$f")&fields=accession%2Creviewed%2Cid%2Cprotein_name%2Cgene_names%2Corganism_name%2Clength%2Cgo_id%2Cgo_p%2Cgo_f%2Cgo_c%2Cgo%2Cgene_oln%2Ccc_induction%2Cxref_pathwaycommons%2Cxref_reactome%2Cxref_unipathway%2Ccc_interaction&format=tsv" > foot_part.tsv
  if [ -f "$FOOT_INFO" ]; then tail -n +2 foot_part.tsv >> "$FOOT_INFO"; else mv foot_part.tsv "$FOOT_INFO"; fi
done
rm -f foot_acc_part_* foot_part.tsv
echo "foot records: $(( $(wc -l < "$FOOT_INFO") - 1 )) of $(wc -l < foot_accessions.txt) foot proteins"
```

# The tables

`masterID` is the blastx table with the subject ID (`sp|Q5RA75|F135A_PONAB`) split into the
database (`sp`), the accession, and the entry name's two parts (`geneID`, `species`). The two
record tables are stacked, the foot query's first, without its combined GO column, and joined to
the hits by accession. `g.spid` gives each CDS its gene from the CDS FASTA's headers
(`[gene=LOC...]`) and keeps one row per gene and accession; it is joined to the hits by CDS.

```{r tables, eval = params$run}
read_blastx <- function(f) {
  b <- read.delim(f, header = FALSE, stringsAsFactors = FALSE)
  if (params$as_2024) b <- b[-1, ]                     # the 2024 run read the table with a header line
  b
}
## CDS ID and header description, as the 2024 run's perl one-liner tabulated them
h <- sub("^>", "", grep("^>", readLines(cds_fa), value = TRUE))
cdsftab <- data.frame(V1 = sub(" .*$", "", h), V2 = sub("^[^ ]* ?", "", h), stringsAsFactors = FALSE)

blast <- read.delim(blastx, header = FALSE, stringsAsFactors = FALSE)     # every hit, for g.spid
g.spid <- left_join(blast, cdsftab, by = "V1") %>%
  mutate(gene = str_extract(V2.y, "(?<=\\[gene=)\\w+")) %>%
  select(gene, V11, V2.x, V1) %>%
  mutate(uniprot_accession = str_extract(V2.x, "(?<=\\|)[^\\|]*(?=\\|)")) %>%
  distinct(gene, uniprot_accession, .keep_all = TRUE)
colnames(g.spid)[4] <- "transcript"

masterID <- read_blastx(blastx) %>%
  separate(V2, into = c("sp", "uniprot_accession", "entry_name"), sep = "\\|") %>%
  separate(entry_name, into = c("geneID", "species"), sep = "_", extra = "merge") %>%
  rename(transcript = V1, pident = V3, length = V4, mismatch = V5, gapopen = V6, qstart = V7,
         qend = V8, sstart = V9, send = V10, evalue = V11, bitscore = V12)

mytilus_foot_uniprot_info <- read.delim(foot_info_tsv, sep = "\t", header = TRUE)
uniprot_info <- read.delim(sprot_info_tsv, sep = "\t", header = TRUE)
mytilus_foot_uniprot_info <- subset(mytilus_foot_uniprot_info, select = -Gene.Ontology..GO.)
total_uniprot_info <- rbind(mytilus_foot_uniprot_info, uniprot_info) %>% rename(uniprot_accession = Entry)
if (!params$as_2024)                                   # one record per accession: Swiss-Prot's where both tables have it
  total_uniprot_info <- total_uniprot_info[!duplicated(total_uniprot_info$uniprot_accession, fromLast = TRUE), ]

full_blast <- left_join(masterID, total_uniprot_info, by = "uniprot_accession")
LOC_GO_blast <- left_join(full_blast, g.spid, by = "transcript")

dir.create(res_dir, showWarnings = FALSE, recursive = TRUE)
write.table(LOC_GO_blast, file = file.path(res_dir, "LOC_GO_list.txt"), sep = "\t", row.names = FALSE, quote = TRUE)
write.table(g.spid, file = file.path(res_dir, "g.spid.txt"), sep = "\t", row.names = FALSE, quote = TRUE)
```

# Reproducing the 2024 tables

The 2024 blastx table is kept on gannet
(`panopea/PSMFC-mytilus-byssus-pilot/byssus-exp-analysis/output/genome_n_foot_blastx.tab`, 41,432
hits, 2024-08-06), with the CDS FASTA (`.../data/ncbi_dataset/`). Rebuilt from them on
2026-10-03:

- the BLAST database was, by its hits, Swiss-Prot release 2024_04 plus the committed foot proteins
  (above). The `database` and `blastx` chunks, run with BLAST+ 2.15.0 on that database for a
  sample of 100 CDSs (the first of the table, 20 with several HSPs, 79 at random), reproduce the
  2024 table's 176 lines for them byte for byte, e-values (which depend on the database size)
  and order included. That sample took 4 minutes on 4 threads, so the whole search would take
  about a day and a half on 4 threads;
- `g.spid.txt` is byte-identical to the committed file (17,386 rows);
- `LOC_GO_list.txt` is byte-identical to the committed file (41,464 rows) with `as_2024: true`,
  when the two UniProt record tables are taken from the committed file itself (UniProt no longer
  serves the 2024 records). Two quirks of the 2024 run explain its row count: the blastx table
  was read with a header line, so its first hit (a second HSP of the same CDS and protein is
  kept) is missing; and four reviewed foot proteins (O96064, Q25434, Q25464, Q27409) are in both
  UniProt tables, so their 33 hits appear twice. Neither changes what the analysis uses, the
  best hit per gene. With `as_2024: false` (the default) the first hit is kept and each
  accession has one record.

The committed `V11` column is the e-value of the gene's first hit from the same search, not of
a second search.

The check below matches every blastx hit (HSP) to the rows of `LOC_GO_list.txt` on all twelve
blastx columns, which identify an HSP uniquely. On the committed tables and the 2024 blastx
table it gives 41,432 hits and 41,464 rows: 1 hit missing (the first), 0 rows that are not hits
and 33 rows repeated (the four accessions above). A run with `as_2024: false` gives 0, 0 and 0.

```{r check, eval = nzchar(params$blastx_tab) || file.exists(blastx)}
b <- read.delim(blastx, header = FALSE, stringsAsFactors = FALSE)
b$V2 <- sub("^[^|]*[|]([^|]*)[|].*$", "\\1", b$V2)             # the accession, as LOC_GO_list keeps it
loc <- read.delim(file.path(res_dir, "LOC_GO_list.txt"), stringsAsFactors = FALSE)
cols <- c("transcript", "uniprot_accession.x", "pident", "length", "mismatch", "gapopen", "qstart",
          "qend", "sstart", "send", "evalue", "bitscore")
hits <- do.call(paste, b)
rows <- do.call(paste, loc[, cols])
knitr::kable(data.frame(
  measure = c("blastx hits", "LOC_GO_list rows", "hits not in LOC_GO_list",
              "LOC_GO_list rows that are not hits", "LOC_GO_list rows repeated"),
  n = c(length(hits), length(rows), sum(!hits %in% rows), sum(!rows %in% hits), sum(duplicated(rows)))))
```

# Provenance

```{r provenance, eval = params$run}
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # provenance_lines()
blast_version <- system2(file.path(params$blast_bin, "blastx"), "-version", stdout = TRUE)[1]
db_info <- system2(file.path(params$blast_bin, "blastdbcmd"), c("-db", shQuote(db), "-info"), stdout = TRUE)
writeLines(provenance_lines(
  c("03 step 01: genome CDS blastx against Swiss-Prot and the Mytilus foot proteins",
    paste0("run                  : ", format(Sys.time(), "%Y-%m-%d %H:%M %Z")),
    paste0("Swiss-Prot release   : ", rel, " (UniProt records of the same release, REST)"),
    paste0("foot proteins        : 02_data/", params$foot_fasta, " (", sum(startsWith(readLines(foot_fa), ">")), " sequences)"),
    paste0("blastx               : ", blast_version, "; -evalue 1E-20 -max_target_seqs 1 -outfmt 6; ",
           params$parts, " part(s), ", params$threads, " threads"),
    paste0("database             : ", trimws(db_info[grepl("sequences;", db_info)])),
    paste0("hits (HSPs)          : ", nrow(read.delim(blastx, header = FALSE)))),
  c(here::here("01_code", "01_genome_blast.Rmd"), here::here("01_code", "_blastx_parts.sh"), foot_fa, cds_fa,
    blastx, sprot_info_tsv, foot_info_tsv), repo_root, packages = c("dplyr", "stringr", "tidyr")),
  file.path(res_dir, "RUN_provenance.txt"))
```
