---
title: "05 - Foot and byssal proteins searched again without the low-complexity filter"
output: html_document
params:
  run: false            # true: run the search (BLAST+ 2.15.0 and the genome CDS needed); false: rebuild the tables from the committed search
  blast_bin: "/usr/local/bin"
  cds_fasta: "04_sequence-alignment/02_data/cds_from_genomic.fasta"   # repo-relative; the genome CDS step 01 searched (git-ignored download)
  work_dir: ""          # where the search writes its table and database ("" = a folder in tempdir(); set it to resume)
  threads: 4
  pident_min: 70        # identity (%) a hit needs to be adopted
  full_db: ""           # optional, with run: true: step 01's BLAST database, to search the adopted genes' CDSs against it with the filter off
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Step 01 annotates every gene by its best blastx hit against Swiss-Prot 2026_03 plus 241 foot and
byssal proteins (`03_analyses/genome-foot-sprot2026_03/`). blastx masks low-complexity stretches
of the translated query by default (SEG; Wootton and Federhen 1993, *Computers & Chemistry*
17:149-163), and several mussel adhesive proteins are low-complexity throughout: mfp-3 and mfp-5
are short and made mostly of DOPA (hydroxylated tyrosine), glycine and basic residues (mfp-3:
Papov et al. 1995, *J Biol Chem* 270:20183-20192; mfp-5: Waite and Qin 2001, *Biochemistry*
40:2887-2893). Their genes got no hit: on 2026-10-06 the CDSs of the six genes this step
adopts, searched against step 01's database with step 01's settings, had best e-values of 7.5e-20
to 1.8e-9, all above the 1e-20 cut-off; with `-seg no` and nothing else changed, the same
database gives 5.6e-22 to 1.05e-28, each to a foot protein.

This step searches every CDS again with the filter off, against the 241 foot and byssal proteins
alone (`02_data/uniprotkb_mytilus_foot_2026_03_byssal.fasta`), with step 01's other settings
(`-evalue 1E-20 -max_target_seqs 1`) and `-dbsize` set to the total residues of step 01's
database, so that the e-values and the cut-off are those a search of the whole database would
give (checked on the six genes: the same proteins, bitscores and e-values). The search took
about 2 minutes on 4 threads (step 01's took about 31 hours).

With the filter off, low-complexity and repeated sequence also scores: of the 29 genes that have
no hit in step 01 and a hit here, 23 match at 29 to 45% identity, mostly through repeats: 13
collagen-like protein-5 or preCOL-NG (Gly-X-Y repeats), 4 mfp-1 (decapeptide repeats; RefSeq
names two of the genes "adhesive plaque matrix protein-like", at 36 and 37%), 3 mfp-4
(histidine-rich repeats) and 3 others; the other 6 match mfp-3, mfp-5, foot protein 9 and
thread matrix protein 2F at 70.5 to 86.7%, each a protein of the same genus. Of the 451 genes that have a hit in step 01, 142 score
higher here, 105 on the same protein (the filter had masked part of the query) and 37 on another:
8 on another variant of the same byssal protein (mfp-1, mfp-4, mfp-6, precollagen-P, preCOL-NG),
LOC134718845 (precollagen-D at 99% in step 01) on preCOL-NG at 63%, and 28 at 27 to 44% identity,
mostly on collagen-like, tyrosinase and EGF-repeat proteins; three of the 28 would become byssal
structural proteins (09 step 04's category: two EGF-repeat genes matching mfp-2 instead of notch
homolog 1 and fibropellin-1, one matching byssal EP-like protein 1 instead of HIP). A hit is therefore adopted only for a gene that
has no hit in step 01's search (no hit of record is replaced; scores with and without the filter
are not comparable) and only at an identity of at least `pident_min` (70%); any floor from 46% to
70% adopts the same six genes. Writes, in `03_analyses/genome-foot-sprot2026_03-noseg/`:

- `LOC_GO_list.txt`: the annotation the analysis reads: step 01's `LOC_GO_list.txt`, line for
  line, followed by the hits of the adopted genes' CDSs, built as step 01 builds its rows (the
  same columns in the same order, the hit joined to its UniProt record of release 2026_03 and
  each CDS to its gene);
- `noseg_genes.csv`: every gene with a hit in this search: its best hit here (highest bitscore),
  its best hit in step 01's search if it has one, whether the hit is adopted and why;
- `foot_noseg_blastx.tab`: the blastx table of this search; `foot_noseg_cds.tsv`: the FASTA
  header of every CDS in it and of every CDS step 01 found on a foot protein (so that the tables
  can be rebuilt without the genome CDS);
- `adopted_full_db_check.tsv` (with `full_db`): the adopted genes' CDSs searched against step 01's
  whole database with the filter off;
- `RUN_provenance.txt` (`run: true`): BLAST+ and its settings, the code commit, the input MD5s.

With `run: false` (the default) it rebuilds the tables from the committed search table and
headers and runs the checks. The UniProt records are those step 01 downloaded for the 241
proteins (`02_data/uniprot_mytilus_foot_r2026_03.tsv`, release 2026_03; its MD5 is checked
against step 01's `RUN_provenance.txt`).

```{r paths}
suppressPackageStartupMessages({ library(dplyr); library(stringr); library(tidyr) })
repo_root <- normalizePath(file.path(here::here(), ".."))
rec_dir   <- here::here("03_analyses", "genome-foot-sprot2026_03")       # step 01's search of 2026
res_dir   <- here::here("03_analyses", "genome-foot-sprot2026_03-noseg")
foot_fa   <- here::here("02_data", "uniprotkb_mytilus_foot_2026_03_byssal.fasta")
foot_tsv  <- here::here("02_data", "uniprot_mytilus_foot_r2026_03.tsv")
cds_fa    <- file.path(repo_root, params$cds_fasta)
work_dir  <- if (nzchar(params$work_dir)) params$work_dir else file.path(tempdir(), "byssal_noseg")
blastx_new <- file.path(work_dir, "foot_noseg_blastx.tab")
dir.create(res_dir, showWarnings = FALSE, recursive = TRUE)
## step 01's provenance gives its database size and the MD5s of the inputs the two steps share
prov01 <- readLines(file.path(rec_dir, "RUN_provenance.txt"))
dbsize <- as.numeric(gsub(",", "", sub("^.*; ([0-9,]+) total residues.*$", "\\1",
                                        grep("^database", prov01, value = TRUE))))
stopifnot(length(dbsize) == 1, is.finite(dbsize))
md5_01 <- local({
  m <- regmatches(prov01, regexec("^ +([0-9a-f]{32})  (.*)$", prov01))
  m <- m[lengths(m) == 3]
  setNames(vapply(m, `[`, "", 2), basename(vapply(m, `[`, "", 3)))
})
same_md5 <- function(f) unname(tools::md5sum(f)) == md5_01[[basename(f)]]
stopifnot("foot FASTA differs from step 01's" = same_md5(foot_fa),
          "foot records differ from step 01's download" = same_md5(foot_tsv))
if (params$run) stopifnot("genome CDS differs from step 01's" = same_md5(cds_fa))
cat("database size of step 01 (residues):", format(dbsize, big.mark = ","), "\n")
Sys.setenv(BLAST_BIN = params$blast_bin, CDS_FA = cds_fa, FOOT_FA = foot_fa, BLASTX_NEW = blastx_new,
           THREADS = params$threads, DBSIZE = sprintf("%.0f", dbsize), WORK_DIR = work_dir,
           HELPER = here::here("01_code", "_blastx_noseg.sh"))
```

# The search

`_blastx_noseg.sh` makes the database of the 241 proteins and runs
`blastx -evalue 1E-20 -max_target_seqs 1 -seg no -dbsize <step 01's residues> -outfmt 6` on every
CDS. It writes its table whole or not at all and does not search again if the table exists.

```{bash search, eval = params$run}
set -euo pipefail
mkdir -p "$WORK_DIR"
bash "$HELPER" "$BLAST_BIN" "$CDS_FA" "$FOOT_FA" "$BLASTX_NEW" "$THREADS" "$DBSIZE"
```

```{r keep-search, eval = params$run}
## the search table and the headers of the CDSs the tables need, beside the tables
stopifnot(file.copy(blastx_new, file.path(res_dir, "foot_noseg_blastx.tab"), overwrite = TRUE))
foot_acc <- sub("^>[^|]*[|]([^|]*)[|].*$", "\\1", grep("^>", readLines(foot_fa), value = TRUE))
b01 <- read.delim(file.path(rec_dir, "genome_n_foot_blastx.tab"), header = FALSE, stringsAsFactors = FALSE)
b05 <- read.delim(blastx_new, header = FALSE, stringsAsFactors = FALSE)
need <- union(b05$V1, b01$V1[sub("^[^|]*[|]([^|]*)[|].*$", "\\1", b01$V2) %in% foot_acc])
h <- sub("^>", "", grep("^>", readLines(cds_fa), value = TRUE))
hd <- data.frame(V1 = sub(" .*$", "", h), V2 = sub("^[^ ]* ?", "", h), stringsAsFactors = FALSE)
write.table(hd[hd$V1 %in% need, ], file.path(res_dir, "foot_noseg_cds.tsv"), sep = "\t",
            row.names = FALSE, quote = TRUE)
```

# The tables

`build_rows()` is step 01's table code (its `tables` chunk, `as_2024: false`), applied to a blastx
table, with the UniProt records of the foot proteins only (every subject here is one of them).
The check below applies it to step 01's own hits on the foot proteins that are TrEMBL entries
(for Swiss-Prot entries step 01 took the Swiss-Prot table's record) and compares the lines it
writes with step 01's `LOC_GO_list.txt`.

```{r tables}
cdsftab <- read.delim(file.path(res_dir, "foot_noseg_cds.tsv"), stringsAsFactors = FALSE)
records <- read.delim(foot_tsv, sep = "\t", header = TRUE)
records <- subset(records, select = -Gene.Ontology..GO.) %>% rename(uniprot_accession = Entry)
stopifnot(!anyDuplicated(records$uniprot_accession))

build_rows <- function(blast) {
  g.spid <- left_join(blast, cdsftab, by = "V1") %>%
    mutate(gene = str_extract(V2.y, "(?<=\\[gene=)\\w+")) %>%
    select(gene, V11, V2.x, V1) %>%
    mutate(uniprot_accession = str_extract(V2.x, "(?<=\\|)[^\\|]*(?=\\|)")) %>%
    distinct(gene, uniprot_accession, .keep_all = TRUE)
  colnames(g.spid)[4] <- "transcript"
  masterID <- blast %>%
    separate(V2, into = c("sp", "uniprot_accession", "entry_name"), sep = "\\|") %>%
    separate(entry_name, into = c("geneID", "species"), sep = "_", extra = "merge") %>%
    rename(transcript = V1, pident = V3, length = V4, mismatch = V5, gapopen = V6, qstart = V7,
           qend = V8, sstart = V9, send = V10, evalue = V11, bitscore = V12)
  left_join(left_join(masterID, records, by = "uniprot_accession"), g.spid, by = "transcript")
}
as_lines <- function(d) {        # the lines write.table() gives step 01's LOC_GO_list.txt
  f <- tempfile(); on.exit(unlink(f))
  write.table(d, f, sep = "\t", row.names = FALSE, quote = TRUE, col.names = FALSE)
  readLines(f)
}
gene_of <- function(cds) str_extract(cdsftab$V2[match(cds, cdsftab$V1)], "(?<=\\[gene=)\\w+")
best_by <- function(d, key) {   # one row per key: the highest bitscore, ties in table order
  d <- d[order(d[[key]], -d$bitscore, seq_len(nrow(d))), ]
  d[!duplicated(d[[key]]), ]
}

## step 01's table, as the analysis reads it (best hit per gene: highest bitscore, ties in table order)
rec_file  <- file.path(rec_dir, "LOC_GO_list.txt")
rec_lines <- readLines(rec_file)
rec <- read.delim(rec_file, stringsAsFactors = FALSE)
stopifnot(length(rec_lines) - 1 == nrow(rec))
rec_best <- best_by(rec[!is.na(rec$gene), c("gene", "uniprot_accession.x", "Protein.names", "pident", "bitscore")], "gene")

## check: step 01's rows on TrEMBL foot proteins, rebuilt
b01 <- read.delim(file.path(rec_dir, "genome_n_foot_blastx.tab"), header = FALSE, stringsAsFactors = FALSE)
tr_foot <- startsWith(b01$V2, "tr|") & sub("^[^|]*[|]([^|]*)[|].*$", "\\1", b01$V2) %in% records$uniprot_accession
rebuilt <- as_lines(build_rows(b01[tr_foot, ]))
rec_tr  <- rec_lines[-1][rec$sp == "tr" & rec$uniprot_accession.x %in% records$uniprot_accession]

## this search: best hit per gene, and whether it is adopted
b05 <- read.delim(file.path(res_dir, "foot_noseg_blastx.tab"), header = FALSE, stringsAsFactors = FALSE)
b05$gene <- gene_of(b05$V1)
hits <- data.frame(gene = b05$gene, cds = b05$V1, accession = sub("^[^|]*[|]([^|]*)[|].*$", "\\1", b05$V2),
                   pident = b05$V3, length = b05$V4, evalue = b05$V11, bitscore = b05$V12, stringsAsFactors = FALSE)
genes <- best_by(hits[!is.na(hits$gene), ], "gene") %>%
  left_join(count(hits, gene, name = "n_cds_hits"), by = "gene") %>%
  mutate(protein = records$Protein.names[match(accession, records$uniprot_accession)],
         organism = records$Organism[match(accession, records$uniprot_accession)],
         refseq_protein = str_extract(cdsftab$V2[match(cds, cdsftab$V1)], "(?<=\\[protein=)[^\\]]+")) %>%
  left_join(rec_best %>% rename(record_accession = uniprot_accession.x, record_protein = Protein.names,
                                record_pident = pident, record_bitscore = bitscore), by = "gene") %>%
  mutate(adopted = is.na(record_accession) & pident >= params$pident_min,
         reason = case_when(!is.na(record_accession) ~ "has a hit in step 01's search",
                            adopted ~ paste0("no hit in step 01's search; identity >= ", params$pident_min, "%"),
                            TRUE ~ paste0("no hit in step 01's search; identity < ", params$pident_min, "%"))) %>%
  arrange(desc(adopted), !is.na(record_accession), desc(pident))
adopted <- genes$gene[genes$adopted]

## the adopted genes' rows, built as step 01 builds its rows, after step 01's lines
new_rows <- build_rows(b05[b05$gene %in% adopted, setdiff(names(b05), "gene")])
stopifnot(identical(names(new_rows), names(rec)))
new_lines <- as_lines(new_rows)
out_file <- file.path(res_dir, "LOC_GO_list.txt")
writeLines(c(rec_lines, new_lines), out_file)
write.csv(genes, file.path(res_dir, "noseg_genes.csv"), row.names = FALSE)
knitr::kable(genes[genes$adopted, c("gene", "refseq_protein", "accession", "protein", "organism",
                                    "pident", "length", "evalue", "bitscore")], row.names = FALSE)
```

# Checks

```{r checks}
merged <- read.delim(out_file, stringsAsFactors = FALSE)
m_best <- best_by(merged[merged$gene %in% adopted, c("gene", "uniprot_accession.x", "bitscore")], "gene")
nohit_rest <- genes[is.na(genes$record_accession) & !genes$adopted, ]
checks <- data.frame(check = c(
  "step 01's rows on TrEMBL foot proteins, rebuilt by build_rows(), are its lines byte for byte",
  "every subject of this search is one of the 241 foot proteins",
  "no adopted gene has a row in step 01's table",
  "the merged table holds step 01's rows and then the adopted genes' rows",
  "each adopted gene's best row in the merged table is its adopted hit",
  "every adopted hit has e-value <= 1e-20 on step 01's database size"),
  pass = c(
    length(rebuilt) > 0 && identical(rebuilt, rec_tr),
    all(hits$accession %in% records$uniprot_accession),
    !any(adopted %in% rec$gene),
    nrow(merged) == nrow(rec) + nrow(new_rows) && identical(readLines(out_file)[seq_along(rec_lines)], rec_lines),
    nrow(m_best) == length(adopted) &&
      all(m_best$uniprot_accession.x == genes$accession[match(m_best$gene, genes$gene)] &
          m_best$bitscore == genes$bitscore[match(m_best$gene, genes$gene)]),
    all(genes$evalue[genes$adopted] <= 1e-20)))
knitr::kable(checks)
if (!all(checks$pass)) stop("check failed: ", paste(checks$check[!checks$pass], collapse = "; "))
knitr::kable(data.frame(measure = c(
  "hits (HSPs)", "CDSs with a hit", "genes with a hit", "  of them with a hit in step 01's search",
  "  with none there", paste0("  adopted (identity >= ", params$pident_min, "%)"),
  "highest identity among the genes with no hit in step 01 not adopted (%)",
  "step 01's rows on TrEMBL foot proteins rebuilt", "rows added to step 01's table"),
  n = c(nrow(b05), length(unique(b05$V1)), nrow(genes), sum(!is.na(genes$record_accession)),
        sum(is.na(genes$record_accession)), length(adopted),
        if (nrow(nohit_rest)) max(nohit_rest$pident) else NA, length(rebuilt), nrow(new_rows))))
```

# The adopted genes against step 01's whole database

With `full_db` (step 01's database, `swissprot_n_mytilusfoot_r2026_03`), the adopted genes' CDSs
are searched against it with step 01's settings and the filter off: their best hit should be the
same foot protein, with the same e-value, so that nothing in Swiss-Prot scores higher.

```{r full-db, eval = params$run && nzchar(params$full_db)}
q <- unique(b05$V1[b05$gene %in% adopted])
fa <- readLines(cds_fa); starts <- grep("^>", fa)
keep <- rep(sub("^>([^ ]*).*$", "\\1", fa[starts]) %in% q, diff(c(starts, length(fa) + 1)))
qf <- tempfile(fileext = ".fna"); writeLines(fa[keep], qf)
of <- tempfile(fileext = ".tab")
stopifnot(system2(file.path(params$blast_bin, "blastx"),
                  c("-query", qf, "-db", shQuote(params$full_db), "-out", of, "-evalue", "1E-20",
                    "-num_threads", params$threads, "-max_target_seqs", "1", "-seg", "no", "-outfmt", "6"),
                  stderr = FALSE) == 0)
full <- read.delim(of, header = FALSE, stringsAsFactors = FALSE)
full <- full[!duplicated(full$V1), c("V1", "V2", "V3", "V11", "V12")]
names(full) <- c("cds", "full_subject", "full_pident", "full_evalue", "full_bitscore")
foot1 <- b05[!duplicated(b05$V1) & b05$V1 %in% q, c("V1", "gene", "V2", "V3", "V11", "V12")]
names(foot1) <- c("cds", "gene", "foot_subject", "foot_pident", "foot_evalue", "foot_bitscore")
fc <- left_join(foot1, full, by = "cds") %>%
  mutate(same_protein = foot_subject == full_subject, same_evalue = foot_evalue == full_evalue)
write.table(fc, file.path(res_dir, "adopted_full_db_check.tsv"), sep = "\t", row.names = FALSE, quote = FALSE)
knitr::kable(fc)
if (!all(fc$same_protein %in% TRUE)) stop("an adopted CDS has a different best hit in step 01's database")
```

# Provenance

```{r provenance, eval = params$run}
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # provenance_lines()
blast_version <- system2(file.path(params$blast_bin, "blastx"), "-version", stdout = TRUE)[1]
writeLines(provenance_lines(
  c("03 step 05: genome CDS blastx against the foot and byssal proteins, low-complexity filter off",
    paste0("run                  : ", format(Sys.time(), "%Y-%m-%d %H:%M %Z")),
    paste0("foot proteins        : 02_data/uniprotkb_mytilus_foot_2026_03_byssal.fasta (",
           sum(startsWith(readLines(foot_fa), ">")), " sequences); records 02_data/uniprot_mytilus_foot_r2026_03.tsv (release 2026_03)"),
    paste0("blastx               : ", blast_version, "; -evalue 1E-20 -max_target_seqs 1 -seg no -dbsize ",
           sprintf("%.0f", dbsize), " -outfmt 6; ", params$threads, " threads"),
    paste0("hits (HSPs)          : ", nrow(b05), " (", nrow(genes), " genes)"),
    paste0("adopted              : ", length(adopted), " genes (no hit in step 01's search, identity >= ",
           params$pident_min, "%): ", paste(adopted, collapse = ", ")),
    paste0("rows added           : ", nrow(new_rows), " to step 01's ", nrow(rec)),
    paste0("full-database check  : ", if (nzchar(params$full_db))
      paste0(basename(params$full_db), ", same best protein for every adopted CDS") else "not run"),
    paste0("checks               : ", sum(checks$pass), " of ", nrow(checks), " passed")),
  c(here::here("01_code", "05_byssal_noseg_search.Rmd"), here::here("01_code", "_blastx_noseg.sh"), foot_fa,
    foot_tsv, cds_fa, file.path(rec_dir, "LOC_GO_list.txt"), file.path(rec_dir, "genome_n_foot_blastx.tab"),
    file.path(rec_dir, "RUN_provenance.txt"), file.path(res_dir, "foot_noseg_blastx.tab")),
  repo_root, packages = c("dplyr", "stringr", "tidyr")),
  file.path(res_dir, "RUN_provenance.txt"))
```
