---
title: "00 - Run the GO enrichment analysis"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output: html_document
params:
  steps: "all"          # "all", or step numbers, e.g. [2, 5]
  stop_on_fail: true    # stop at the first failed step
  cores: 4              # parallel runs in steps 02-04; the scripts use 1 on Windows
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs every numbered script in `01_code/` in order, each in its own R process, with an HTML
report and a log per step in `03_analyses/knit_html/` (git-ignored) and a `run_log.csv`.
Knit this file from inside `enrichment.Rproj`, or let the repository-level
`00_run_pipeline.Rmd` call it after `06_differential-expression`.

| step | script | writes to `03_analyses/` |
|---|---|---|
| 01 | `01_go_inputs.Rmd` | `01_go-inputs/`: gene annotation (best BLAST hit, GO by ontology, transcript length, mitochondrial flag) and gene-set sizes |
| 02 | `02_topgo.Rmd` | `02_topgo/`: topGO weight01 Fisher, the enrichment of record |
| 03 | `03_goseq.Rmd` | `03_goseq/`: goseq Wallenius with length bias, and the length-bias diagnostic |
| 04 | `04_clusterprofiler.Rmd` | `04_clusterprofiler/`: clusterProfiler `enricher`, compareCluster view |
| 05 | `05_rrvgo.Rmd` | `05_rrvgo/`: topGO terms reduced to clusters (REVIGO-style) |
| 06 | `06_method_comparison.Rmd` | `06_method-comparison/`: agreement between the methods, consensus terms |

Every method uses the same genes: for each of the 16 contrasts in
`06_differential-expression/03_analyses/DEG_lists/contrasts.csv`, the universe is the genes
DESeq2 tested (non-missing adjusted p) and up- and down-regulated DEGs (padj < 0.05) are
tested separately. Step 02 takes about five minutes on four cores; the rest about four
together. Step 05 needs `org.Hs.eg.db` (installed on first run if missing), for a column it
then replaces; see the script.

```{r run}
suppressPackageStartupMessages(library(here))
source(file.path(here::here(), "..", "tools", "run_steps.R"))
html_dir <- here::here("03_analyses", "knit_html")
par_steps <- setNames(rep(list(list(cores = params$cores)), 3), c("02", "03", "04"))
run_log <- run_steps(here::here("01_code"), html_dir, steps = params$steps, step_params = par_steps,
                     stop_on_fail = isTRUE(params$stop_on_fail))
finish_run(run_log, html_dir, "GO enrichment")
```

# After a run

Start with `06_method-comparison/consensus_terms_TC_BP.csv` (terms enriched by topGO and by
at least one other method) and the topGO dot plots in `02_topgo/`. topGO's `weight01`
p-values are not FDR-adjusted, by convention; in runs where no FDR-controlled method finds
any term, treat its list as exploratory. Terms marked as carried mostly by mitochondrially
encoded protein LOCs (Gill OA up) count one mitochondrial signal many times.
