---
title: "01 - GO enrichment inputs: annotation, gene lengths and gene sets"
output: html_document
params:
  families: ["TC", "FG"]   # contrast families to summarise (see 06 contrasts.csv)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Builds the one annotation table that every enrichment method in this folder uses, and
summarises the gene sets each contrast contributes.

1. **GO annotation per gene.** The genome-wide BLAST of the coding sequences against SwissProt
   and the Mytilus foot proteins (`03_blast/03_analyses/genome-foot/LOC_GO_list.txt`) can
   give several hits per LOC, one per coding sequence. The hit with the highest bitscore is
   kept, the rule `09_gene-mechanics-correlation` uses. Its GO IDs are trimmed (the table
   separates them with "; "), kept only if this `GO.db` knows them, and split by ontology.
2. **Length**, for goseq's length-bias correction: the median length of the gene's reference
   transcripts in `05_sequence-alignment/03_analyses/hisat/t_data.ctab`.
3. **No mitochondrial loci.** The mitochondrial genome's genes and their copies on unplaced
   scaffolds (`06 .../count_matrix/mitochondrial_loci.csv`) are left out of the genome DE in
   `06_differential-expression` and analysed there on their own (step 13). This script stops if
   any of them reaches a universe, which would mean 06 was not rerun: one mitochondrial signal
   counted many times dominated the GO results before (in Gill OA, 116 of 711 DEGs).
4. **Gene sets.** For each contrast the universe is every gene with an adjusted p in its
   apeglm table from `06_differential-expression` (independent filtering leaves the rest at
   NA, so they could never be DEGs); up- and down-regulated DEGs (padj < 0.05) are tested
   separately against that universe.

Writes, in `03_analyses/01_go-inputs/`:

- `gene_annotation.tsv`: one row per gene in any contrast's table: LOC ID, best-hit protein,
  median transcript length and the GO IDs by ontology (";"-separated);
- `gene_sets_summary.csv`: per contrast and direction, the universe and DEG counts and how many
  carry GO annotation;
- `RUN_provenance.txt`: the GO release in `GO.db`, the thresholds, the code commit, R and
  package versions for every method in this folder, and an MD5 checksum of each input.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(here::here("01_code", "_go_helpers.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))   # gene_key(): count-matrix name -> LOC key
suppressPackageStartupMessages({ library(GO.db); library(dplyr); library(tidyr) })
```

# Annotation

```{r annotation}
blast <- read.delim(file.path(blast_go, "LOC_GO_list.txt"), stringsAsFactors = FALSE, check.names = FALSE)
best <- blast %>% arrange(desc(bitscore)) %>% distinct(gene, .keep_all = TRUE) %>%
  transmute(LOC_ID = gene, protein_name = Protein.names,
            GO_raw = ifelse(is.na(Gene.Ontology.IDs), "", Gene.Ontology.IDs))

## GO IDs, trimmed and checked against GO.db
go_long <- best %>% select(LOC_ID, GO_raw) %>% separate_rows(GO_raw, sep = ";") %>%
  mutate(GO = trimws(GO_raw)) %>% filter(grepl("^GO:[0-9]{7}$", GO)) %>% distinct(LOC_ID, GO)
ont_of <- suppressMessages(AnnotationDbi::select(GO.db, unique(go_long$GO), "ONTOLOGY", "GOID"))
go_long <- go_long %>% left_join(dplyr::rename(ont_of, GO = GOID, ont = ONTOLOGY), by = "GO")
cat("GO IDs in the BLAST table:", n_distinct(go_long$GO), "| unknown to this GO.db (dropped):",
    n_distinct(go_long$GO[is.na(go_long$ont)]), "\n")
go_wide <- go_long %>% filter(!is.na(ont)) %>% arrange(GO) %>%
  group_by(LOC_ID, ont) %>% summarise(ids = paste(GO, collapse = ";"), .groups = "drop") %>%
  pivot_wider(names_from = ont, values_from = ids, names_prefix = "GO_")

## median reference-transcript length per gene. Reference genes without a transcript record
## (gene_id ".") appear under their gene record's name (t_name "gene-LOC..."); they are the
## genes the count matrix names "gene-LOC..." or "STRG.n|LOC...". StringTie-only genes
## (STRG.n) have no reference length and stay out of goseq.
td  <- read.delim(t_data, stringsAsFactors = FALSE)
len <- td %>% mutate(LOC_ID = gene_key(ifelse(gene_id != ".", gene_id, t_name))) %>%
  group_by(LOC_ID) %>% summarise(length = median(length), .groups = "drop")

## every gene that appears in any contrast's table
genes <- unique(unlist(lapply(contrasts$code, function(cd) gene_sets(cd)$universe)))
mt_loci <- read.csv(file.path(repo_root, "06_differential-expression", "03_analyses", "count_matrix",
                               "mitochondrial_loci.csv"), stringsAsFactors = FALSE)
if (any(genes %in% mt_loci$gene))
  stop(sum(genes %in% mt_loci$gene), " mitochondrial loci are in the 06 apeglm tables; rerun ",
       "06_differential-expression (its step 03 leaves them out) before this folder.")
annot <- data.frame(gene = genes, LOC_ID = gene_key(genes)) %>%
  left_join(best %>% select(LOC_ID, protein_name), by = "LOC_ID") %>%
  left_join(len, by = "LOC_ID") %>%
  left_join(go_wide, by = "LOC_ID") %>%
  mutate(across(starts_with("GO_"), ~ coalesce(.x, ""))) %>%
  select(gene, LOC_ID, protein_name, length, GO_BP, GO_MF, GO_CC) %>%
  arrange(gene)
write.table(annot, file.path(out_inputs, "gene_annotation.tsv"), sep = "\t", quote = FALSE,
            row.names = FALSE, na = "")
knitr::kable(data.frame(genes = nrow(annot), with_BLAST_hit = sum(!is.na(annot$protein_name)),
                        with_BP = sum(nzchar(annot$GO_BP)), with_MF = sum(nzchar(annot$GO_MF)),
                        with_CC = sum(nzchar(annot$GO_CC)), with_length = sum(!is.na(annot$length))),
             caption = "genes in any contrast")
```

# Gene sets

```{r sets}
runs <- runs_for(unlist(params$families))
summ <- bind_rows(lapply(seq_len(nrow(runs)), function(i) {
  s <- gene_sets(runs$code[i]); d <- s[[runs$direction[i]]]
  a <- annot[match(s$universe, annot$gene), ]
  data.frame(runs[i, ], universe = length(s$universe), universe_with_BP = sum(nzchar(a$GO_BP)),
             universe_with_length = sum(!is.na(a$length)), DEGs = length(d),
             DEGs_with_BP = sum(nzchar(annot$GO_BP[match(d, annot$gene)])))
}))
write.csv(summ, file.path(out_inputs, "gene_sets_summary.csv"), row.names = FALSE)
knitr::kable(summ)
```

# Provenance

GO results depend on the GO release in `GO.db` as much as on the code, so the release, the
package versions and an MD5 checksum of every input are written next to the inputs.

```{r provenance}
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # provenance_lines()
go_info <- GO.db::GO_dbInfo()
settings <- c(
  paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
  paste0("GO release           : ", go_info$value[go_info$name == "GOSOURCEDATE"],
         " (GO.db ", as.character(utils::packageVersion("GO.db")), ")"),
  paste0("DEG                  : padj < ", PADJ, " in the 06 apeglm table; up and down tested separately"),
  "universe             : genes with a non-missing padj in the contrast's apeglm table",
  paste0("terms tested         : ", MIN_GS, " or more annotated universe genes (topGO nodeSize); ",
         MIN_GS, " to ", MAX_GS, " for goseq and clusterProfiler"),
  paste0("families             : ", paste(unlist(params$families), collapse = ", ")),
  paste0("genes                : ", nrow(annot), " in any contrast's table; with GO BP ",
         sum(nzchar(annot$GO_BP)), ", MF ", sum(nzchar(annot$GO_MF)), ", CC ", sum(nzchar(annot$GO_CC))))
inputs <- c(file.path(blast_go, "LOC_GO_list.txt"), t_data, file.path(deg, "contrasts.csv"),
            file.path(deg, contrasts$dir, paste0(contrasts$code, "_apeglm.csv")),
            file.path(repo_root, "06_differential-expression", "03_analyses", "count_matrix", "mitochondrial_loci.csv"),
            here::here("01_code", c("01_go_inputs.Rmd", "_paths.R", "_go_helpers.R")),
            file.path(repo_root, "tools", c("pipeline_checks.R", "gene_ids.R")))
prov <- file.path(out_inputs, "RUN_provenance.txt")
writeLines(provenance_lines(settings, inputs, repo_root,
                            packages = c("topGO", "goseq", "clusterProfiler", "enrichplot", "rrvgo",
                                         "GOSemSim", "GO.db", "AnnotationDbi", "GSEABase",
                                         "org.Hs.eg.db")), prov)
cat(readLines(prov), sep = "\n")
```
