---
title: "01 - GO enrichment inputs: annotation, gene lengths and gene sets"
output: html_document
params:
  families: ["TC", "LC", "FG"]   # contrast families to summarise (see 06 contrasts.csv)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Builds the one annotation table that every enrichment method in this folder uses, and
summarises the gene sets each contrast contributes.

1. **GO annotation per gene.** The genome-wide BLAST of the coding sequences against SwissProt
   and the Mytilus foot proteins (`03_blast/03_analyses/genome-foot/LOC_GO_list.txt`) can
   give several hits per LOC, one per coding sequence. The hit with the highest bitscore is
   kept, the rule `09_gene-mechanics-correlation` uses. Its GO IDs are trimmed (the table
   separates them with "; "), kept only if this `GO.db` knows them, and split by ontology.
2. **Length**, for goseq's length-bias correction: the median length of the gene's reference
   transcripts in `05_sequence-alignment/03_analyses/hisat/t_data.ctab`.
3. **Mitochondrially encoded proteins.** LOCs whose best hit is one of the 13 mtDNA-encoded
   proteins (`tools/mt_encoded.R`) are flagged so later scripts can show how much of a term
   they carry. About 140 such LOCs exist, most on unplaced scaffolds; in Gill OA 116 of the
   711 DEGs are among them.
4. **Gene sets.** For each contrast the universe is every gene with an adjusted p in its
   apeglm table from `06_differential-expression` (independent filtering leaves the rest at
   NA, so they could never be DEGs); up- and down-regulated DEGs (padj < 0.05) are tested
   separately against that universe.

Writes, in `03_analyses/01_go-inputs/`:

- `gene_annotation.tsv`: one row per gene in any contrast's table: LOC ID, best-hit protein,
  median transcript length, the mitochondrial flag and the GO IDs by ontology (";"-separated);
- `gene_sets_summary.csv`: per contrast and direction, the universe and DEG counts, how many
  carry GO annotation, and how many DEGs are mitochondrially encoded protein LOCs.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(here::here("01_code", "_go_helpers.R"))
source(file.path(repo_root, "tools", "mt_encoded.R"))
suppressPackageStartupMessages({ library(GO.db); library(dplyr); library(tidyr) })
```

# Annotation

```{r annotation}
blast <- read.delim(file.path(blast_go, "LOC_GO_list.txt"), stringsAsFactors = FALSE, check.names = FALSE)
best <- blast %>% arrange(desc(bitscore)) %>% distinct(gene, .keep_all = TRUE) %>%
  transmute(LOC_ID = gene, protein_name = Protein.names,
            GO_raw = ifelse(is.na(Gene.Ontology.IDs), "", Gene.Ontology.IDs))

## GO IDs, trimmed and checked against GO.db
go_long <- best %>% select(LOC_ID, GO_raw) %>% separate_rows(GO_raw, sep = ";") %>%
  mutate(GO = trimws(GO_raw)) %>% filter(grepl("^GO:[0-9]{7}$", GO)) %>% distinct(LOC_ID, GO)
ont_of <- suppressMessages(AnnotationDbi::select(GO.db, unique(go_long$GO), "ONTOLOGY", "GOID"))
go_long <- go_long %>% left_join(dplyr::rename(ont_of, GO = GOID, ont = ONTOLOGY), by = "GO")
cat("GO IDs in the BLAST table:", n_distinct(go_long$GO), "| unknown to this GO.db (dropped):",
    n_distinct(go_long$GO[is.na(go_long$ont)]), "\n")
go_wide <- go_long %>% filter(!is.na(ont)) %>% arrange(GO) %>%
  group_by(LOC_ID, ont) %>% summarise(ids = paste(GO, collapse = ";"), .groups = "drop") %>%
  pivot_wider(names_from = ont, values_from = ids, names_prefix = "GO_")

## median reference-transcript length per gene
td  <- read.delim(t_data, stringsAsFactors = FALSE)
len <- td %>% filter(gene_id != ".") %>% mutate(LOC_ID = sub("^gene-", "", gene_id)) %>%
  group_by(LOC_ID) %>% summarise(length = median(length), .groups = "drop")

## every gene that appears in any contrast's table
genes <- unique(unlist(lapply(contrasts$code, function(cd) gene_sets(cd)$universe)))
annot <- data.frame(gene = genes, LOC_ID = sub(".*[|]", "", genes)) %>%
  left_join(best %>% select(LOC_ID, protein_name), by = "LOC_ID") %>%
  left_join(len, by = "LOC_ID") %>%
  left_join(go_wide, by = "LOC_ID") %>%
  mutate(mt_encoded = is_mt_encoded(protein_name),
         across(starts_with("GO_"), ~ coalesce(.x, ""))) %>%
  select(gene, LOC_ID, protein_name, length, mt_encoded, GO_BP, GO_MF, GO_CC) %>%
  arrange(gene)
write.table(annot, file.path(out_inputs, "gene_annotation.tsv"), sep = "\t", quote = FALSE,
            row.names = FALSE, na = "")
knitr::kable(data.frame(genes = nrow(annot), with_BLAST_hit = sum(!is.na(annot$protein_name)),
                        with_BP = sum(nzchar(annot$GO_BP)), with_MF = sum(nzchar(annot$GO_MF)),
                        with_CC = sum(nzchar(annot$GO_CC)), with_length = sum(!is.na(annot$length)),
                        mt_encoded = sum(annot$mt_encoded)), caption = "genes in any contrast")
```

# Gene sets

```{r sets}
runs <- runs_for(unlist(params$families))
summ <- bind_rows(lapply(seq_len(nrow(runs)), function(i) {
  s <- gene_sets(runs$code[i]); d <- s[[runs$direction[i]]]
  a <- annot[match(s$universe, annot$gene), ]
  data.frame(runs[i, ], universe = length(s$universe), universe_with_BP = sum(nzchar(a$GO_BP)),
             universe_with_length = sum(!is.na(a$length)), DEGs = length(d),
             DEGs_with_BP = sum(nzchar(annot$GO_BP[match(d, annot$gene)])),
             DEGs_mt_encoded = sum(annot$mt_encoded[match(d, annot$gene)]))
}))
write.csv(summ, file.path(out_inputs, "gene_sets_summary.csv"), row.names = FALSE)
knitr::kable(summ)
```
