---
title: "02 - GO enrichment with topGO (primary method)"
output: html_document
params:
  families: ["TC", "FG"]            # contrast families (05 contrasts.csv)
  ontologies: ["BP", "MF", "CC"]
  p_cutoff: 0.01                    # weight01 Fisher p for an enriched term
  cores: 4                          # parallel runs; set to 1 on Windows (no forking)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

The GO enrichment of record. For every contrast, direction (up, down) and ontology it fits
topGO (Alexa et al. 2006, Bioinformatics 22:1600) with the `weight01` algorithm and Fisher's
exact test:

- **genes**: the contrast's universe (genes DESeq2 gave an adjusted p) that carry a GO term
  of the ontology; a gene is "significant" if it is a DEG (padj < 0.05) in that direction.
  Up- and down-regulated genes are tested separately, each against the whole universe;
- **annotation**: each gene's direct GO IDs from `01_go-inputs/gene_annotation.tsv`; topGO
  propagates them up the GO graph itself;
- **terms tested**: at least 10 annotated genes (`nodeSize = 10`);
- **enriched**: `weight01` p < 0.01. `weight01` decorrelates the graph (a parent term is
  down-weighted by its significant children), so its p-values are not independent and are
  conventionally not FDR-adjusted (Alexa and Rahnenfuehrer, topGO vignette). The `classic`
  Fisher p (no decorrelation) is reported beside it.

Writes, in `03_analyses/02_topgo/`:

- `topgo_enriched.csv`: every enriched term of every run, with its significant genes;
- `topgo_all_terms_TC_<ontology>.csv`: every tested term of the TC runs (p-values only), one
  file per ontology, read by script 06;
- `topgo_run_summary.csv`: genes, DEGs, terms tested and enriched per run;
- `topgo_<family>_<ontology>_dotplot.png`: the five most significant terms of each run, for
  every family and ontology (BP, MF, CC), so every candidate figure exists.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(here::here("01_code", "_go_helpers.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({ library(topGO); library(GO.db); library(dplyr); library(ggplot2) })
annot <- read_annotation()
cores <- if (.Platform$OS.type == "windows") 1L else as.integer(params$cores)
runs  <- runs_for(unlist(params$families))
jobs  <- merge(runs, data.frame(ontology = unlist(params$ontologies)), by = NULL)
jobs  <- jobs[order(match(jobs$code, contrasts$code), jobs$direction != "up", jobs$ontology), ]
cat(nrow(jobs), "topGO runs on", cores, "core(s)\n")
```

# Run topGO

```{r topgo}
run_topgo <- function(code, direction, ontology) {
  s   <- gene_sets(code)
  g2g <- gene2go_direct(annot, ontology, s$universe)
  all_genes <- factor(as.integer(names(g2g) %in% s[[direction]]), levels = c(0, 1))
  names(all_genes) <- names(g2g)
  n_sig <- sum(all_genes == "1")
  summ <- data.frame(code, direction, ontology, genes_annotated = length(all_genes), DEGs_annotated = n_sig)
  if (n_sig < 2) return(list(summary = cbind(summ, terms_tested = 0L, enriched = 0L), terms = NULL))
  gd <- suppressMessages(new("topGOdata", ontology = ontology, allGenes = all_genes,
                             annot = annFUN.gene2GO, gene2GO = g2g, nodeSize = MIN_GS))
  w01 <- suppressMessages(runTest(gd, algorithm = "weight01", statistic = "fisher"))
  cls <- suppressMessages(runTest(gd, algorithm = "classic", statistic = "fisher"))
  ids <- names(score(w01))
  st  <- termStat(gd, ids)
  terms <- data.frame(code, direction, ontology, GO.ID = ids, Annotated = st$Annotated,
                      Significant = st$Significant, Expected = round(st$Expected, 3),
                      p_weight01 = unname(score(w01)[ids]), p_classic = unname(score(cls)[ids]))
  enr <- terms$GO.ID[terms$p_weight01 < params$p_cutoff]
  sig <- sigGenes(gd)
  in_term <- lapply(genesInTerm(gd, enr)[enr], intersect, sig)   # indexed by GO ID, not position
  stopifnot(identical(names(in_term), enr))
  terms$genes <- ""
  terms$genes[match(enr, terms$GO.ID)] <- vapply(in_term, function(g) paste(sort(g), collapse = ";"), "")
  list(summary = cbind(summ, terms_tested = nrow(terms), enriched = length(enr)), terms = terms)
}

t0  <- Sys.time()
res <- parallel::mclapply(seq_len(nrow(jobs)), function(i)
  run_topgo(jobs$code[i], jobs$direction[i], jobs$ontology[i]), mc.cores = cores)
stopifnot(!vapply(res, inherits, TRUE, what = "try-error"))
cat("topGO runs took", round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1), "min\n")

terms <- bind_rows(lapply(res, `[[`, "terms")) %>%
  left_join(distinct(runs, code, family), by = "code") %>%
  mutate(Term = go_term(GO.ID))
run_summary <- bind_rows(lapply(res, `[[`, "summary")) %>% left_join(distinct(runs, code, family), by = "code")
write.csv(run_summary, file.path(out_topgo, "topgo_run_summary.csv"), row.names = FALSE)

enriched <- terms %>% filter(p_weight01 < params$p_cutoff) %>%
  select(code, family, direction, ontology, GO.ID, Term, Annotated, Significant, Expected,
         p_weight01, p_classic, genes) %>%
  arrange(match(code, contrasts$code), desc(direction == "up"), ontology, p_weight01)
write.csv(enriched, file.path(out_topgo, "topgo_enriched.csv"), row.names = FALSE)
for (o in unlist(params$ontologies))
  write.csv(terms %>% filter(family == "TC", ontology == o) %>%
              select(code, direction, GO.ID, Annotated, Significant, p_weight01, p_classic),
            file.path(out_topgo, sprintf("topgo_all_terms_TC_%s.csv", o)), row.names = FALSE)
knitr::kable(run_summary %>% filter(ontology == "BP"), caption = "BP runs")
```

# Figures

```{r dotplots, fig.width = 10, fig.height = 9}
dotplot_family <- function(fam, ont, n_top = 5) {
  e <- enriched %>% filter(family == fam, ontology == ont)
  if (!nrow(e)) return(invisible(NULL))
  top <- e %>% group_by(code, direction) %>% slice_min(p_weight01, n = n_top, with_ties = FALSE) %>% ungroup()
  d <- e %>% filter(GO.ID %in% top$GO.ID) %>%
    transmute(code, direction, Term, p = p_weight01, n = Significant)
  print(go_dotplot(d, contrasts$code[contrasts$family == fam],
                   title = sprintf("topGO, %s: %s", ONT_NAME[[ont]], FAMILY_TITLE[[fam]]),
                   caption = sprintf(paste("The %d most significant terms of each contrast and direction",
                                           "(weight01 Fisher p < %s; a point only where the term is enriched).",
                                           "Universe: the contrast's tested genes. Colour capped at p = 1e-%d."),
                                     n_top, params$p_cutoff, P_CAP),
                   file = file.path(out_topgo, sprintf("topgo_%s_%s_dotplot.png", fam, ont)),
                   p_label = "-log10 p\n(weight01)"))
}
for (fam in unique(runs$family)) for (ont in unlist(params$ontologies)) dotplot_family(fam, ont)
```
