---
title: "03 - GO enrichment with goseq (comparison method)"
output: html_document
params:
  families: ["TC", "LC", "FG"]
  ontologies: ["BP", "MF", "CC"]
  padj_cutoff: 0.05
  cores: 4                          # set to 1 on Windows (no forking)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

goseq (Young et al. 2010, Genome Biology 11:R14), the method of the oyster Tag-seq template
(`template-oyster-pipeline/6_GO-annotation.Rmd`), run on the same gene sets as topGO
(script 02) for comparison:

- **genes**: the contrast's universe genes that carry a GO term of the ontology and have a
  length; up and down DEGs tested separately;
- **length bias**: a probability weighting function (PWF) of DE on the median reference
  transcript length is fitted per run (`nullp`) and the Wallenius approximation used. The
  template used the genomic span (start to end, introns included) instead; transcript
  length is what a read-count bias would follow. 3' Tag-seq should carry little length bias,
  and `goseq_pwf_TC_BP.png` shows how much is left;
- **annotation**: direct GO IDs plus all their ancestors (goseq tests the annotation as
  given, so without propagation a parent term would miss its children's genes);
- **terms tested**: 10 to 500 annotated genes, the window clusterProfiler (script 04) uses;
- **enriched**: Benjamini-Hochberg adjusted over-representation p < 0.05 within each run and
  ontology. The template called terms at an unadjusted p < 0.05, which, over thousands of
  terms, admits many false positives; the unadjusted p is reported too.

Writes, in `03_analyses/03_goseq/`: `goseq_enriched.csv`, `goseq_all_terms_TC_BP.csv`,
`goseq_run_summary.csv`, `goseq_pwf_TC_BP.png` and `goseq_<family>_BP_dotplot.png`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(here::here("01_code", "_go_helpers.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({ library(goseq); library(GO.db); library(dplyr); library(ggplot2) })
annot <- read_annotation()
mt    <- mt_genes(annot)
len   <- setNames(annot$length, annot$gene)
cores <- if (.Platform$OS.type == "windows") 1L else as.integer(params$cores)
runs  <- runs_for(unlist(params$families))
onts  <- unlist(params$ontologies)
## propagated annotation once per ontology, for every gene in any universe
prop  <- setNames(lapply(onts, function(o) gene2go_propagated(annot, o)), onts)
```

# Run goseq

```{r goseq}
run_goseq <- function(code, direction, ontology) {
  s  <- gene_sets(code)
  pg <- prop[[ontology]]
  genes <- intersect(s$universe, unique(pg$gene))
  genes <- genes[!is.na(len[genes])]
  de <- setNames(as.integer(genes %in% s[[direction]]), genes)
  summ <- data.frame(code, direction, ontology, genes_annotated = length(de), DEGs_annotated = sum(de))
  if (sum(de) < 2) return(list(summary = cbind(summ, terms_tested = 0L, enriched = 0L), terms = NULL, pwf = NULL))
  pwf <- suppressMessages(nullp(de, bias.data = len[genes], plot.fit = FALSE))
  g2c <- pg[pg$gene %in% genes, ]
  size <- table(g2c$GO)
  keep_terms <- names(size)[size >= MIN_GS & size <= MAX_GS]
  g2c <- g2c[g2c$GO %in% keep_terms, ]
  res <- suppressMessages(goseq(pwf, gene2cat = split(g2c$GO, g2c$gene), method = "Wallenius",
                                use_genes_without_cat = TRUE))
  ## use_genes_without_cat = TRUE keeps every annotated gene of the ontology in the test even
  ## if all its terms fall outside the size window; genes with no term of the ontology were
  ## never in `genes`
  res <- res[res$category %in% keep_terms, ]
  res$padj <- p.adjust(res$over_represented_pvalue, "BH")
  terms <- data.frame(code, direction, ontology, GO.ID = res$category, numDEInCat = res$numDEInCat,
                      numInCat = res$numInCat, p_over = res$over_represented_pvalue, padj = res$padj)
  enr <- terms$GO.ID[terms$padj < params$padj_cutoff]
  hits <- g2c[g2c$GO %in% enr & g2c$gene %in% names(de)[de == 1], ]
  by_term <- split(hits$gene, hits$GO)
  terms$genes <- ""; terms$n_mt_encoded <- NA_integer_
  terms$genes[match(enr, terms$GO.ID)] <- vapply(by_term[enr], function(g) paste(sort(g), collapse = ";"), "")
  terms$n_mt_encoded[match(enr, terms$GO.ID)] <- vapply(by_term[enr], function(g) sum(g %in% mt), 0L)
  list(summary = cbind(summ, terms_tested = nrow(terms), enriched = length(enr)), terms = terms,
       pwf = if (ontology == "BP") data.frame(code, direction, DE = pwf$DEgenes, length = pwf$bias.data, pwf = pwf$pwf))
}

jobs <- merge(runs, data.frame(ontology = onts), by = NULL)
jobs <- jobs[order(match(jobs$code, contrasts$code), jobs$direction != "up", jobs$ontology), ]
t0  <- Sys.time()
res <- parallel::mclapply(seq_len(nrow(jobs)), function(i)
  run_goseq(jobs$code[i], jobs$direction[i], jobs$ontology[i]), mc.cores = cores)
stopifnot(!vapply(res, inherits, TRUE, what = "try-error"))
cat(nrow(jobs), "goseq runs took", round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1), "min\n")

terms <- bind_rows(lapply(res, `[[`, "terms")) %>% left_join(distinct(runs, code, family), by = "code") %>%
  mutate(Term = go_term(GO.ID))
run_summary <- bind_rows(lapply(res, `[[`, "summary")) %>% left_join(distinct(runs, code, family), by = "code")
write.csv(run_summary, file.path(out_goseq, "goseq_run_summary.csv"), row.names = FALSE)
enriched <- terms %>% filter(padj < params$padj_cutoff) %>%
  select(code, family, direction, ontology, GO.ID, Term, numDEInCat, numInCat, p_over, padj, n_mt_encoded, genes) %>%
  arrange(match(code, contrasts$code), desc(direction == "up"), ontology, padj)
write.csv(enriched, file.path(out_goseq, "goseq_enriched.csv"), row.names = FALSE)
write.csv(terms %>% filter(family == "TC", ontology == "BP") %>%
            select(code, direction, GO.ID, numInCat, numDEInCat, p_over, padj),
          file.path(out_goseq, "goseq_all_terms_TC_BP.csv"), row.names = FALSE)
knitr::kable(run_summary %>% filter(ontology == "BP"), caption = "BP runs")
```

# Length bias

```{r pwf, fig.width = 10, fig.height = 6}
pwf <- bind_rows(lapply(res, `[[`, "pwf")) %>% filter(code %in% contrasts$code[contrasts$family == "TC"]) %>%
  mutate(run = factor(paste(contrast_label(code), direction),
                      levels = paste(rep(contrast_label(contrasts$code[contrasts$family == "TC"]), each = 2), c("up", "down"))))
binned <- pwf %>% group_by(run) %>% arrange(length, .by_group = TRUE) %>%
  mutate(bin = ceiling(seq_along(length) / 400)) %>%
  group_by(run, bin) %>% summarise(length = median(length), prop = mean(DE), .groups = "drop")
p <- ggplot(pwf, aes(length, pwf)) +
  geom_point(data = binned, aes(length, prop), colour = "grey55", size = 1) +
  geom_line(colour = "#2a78d6", linewidth = 0.8) +
  facet_wrap(~ run, ncol = 4, scales = "free_y") +
  scale_x_log10() +
  labs(x = "median reference transcript length (bp, log scale)", y = "proportion DE",
       title = "goseq probability weighting functions, TC contrasts (biological process genes)",
       caption = "Grey: proportion of DEGs in bins of 400 genes ordered by length. Blue: the fitted PWF used by the Wallenius test.") +
  theme_psmfc(9) + theme(plot.caption = element_text(hjust = 0), plot.caption.position = "plot",
                         plot.title.position = "plot")
ggsave(file.path(out_goseq, "goseq_pwf_TC_BP.png"), p, width = 10, height = 6.5, dpi = 300)
p
```

# Figures

```{r dotplots, fig.width = 10, fig.height = 9}
for (fam in unique(runs$family)) {
  e <- enriched %>% filter(family == fam, ontology == "BP")
  if (!nrow(e)) next
  top <- e %>% group_by(code, direction) %>% slice_min(padj, n = 5, with_ties = FALSE) %>% ungroup()
  d <- e %>% filter(GO.ID %in% top$GO.ID) %>% transmute(code, direction, Term, p = padj, n = numDEInCat, mt = n_mt_encoded)
  print(go_dotplot(d, contrasts$code[contrasts$family == fam],
                   title = sprintf("goseq, biological process: %s", FAMILY_TITLE[[fam]]),
                   caption = sprintf(paste("The 5 most significant terms of each contrast and direction (BH-adjusted",
                                           "Wallenius p < %s; a point only where the term is enriched). Terms of 10 to 500",
                                           "annotated genes; length-bias weighting by transcript length. Colour capped at 1e-%d."),
                                     params$padj_cutoff, P_CAP),
                   file = file.path(out_goseq, sprintf("goseq_%s_BP_dotplot.png", fam)),
                   p_label = "-log10 adjusted p"))
}
```
