---
title: "04 - GO enrichment with clusterProfiler (comparison across contrasts)"
output: html_document
params:
  families: ["TC", "LC", "FG"]
  ontologies: ["BP", "MF", "CC"]
  padj_cutoff: 0.05
  cores: 4                          # set to 1 on Windows (no forking)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Over-representation analysis with clusterProfiler (Yu et al. 2012, OMICS 16:284;
`enricher`, a one-sided hypergeometric test), and its compareCluster view of all contrasts
side by side:

- **genes**: as in scripts 02 and 03, the contrast's universe genes with a GO term of the
  ontology; up and down DEGs tested separately;
- **annotation**: direct GO IDs plus all their ancestors (`TERM2GENE`);
- **terms tested**: 10 to 500 annotated genes (clusterProfiler's defaults);
- **enriched**: Benjamini-Hochberg adjusted p < 0.05.

`compareCluster()` takes one universe for every gene cluster, but here each contrast has its
own universe. So `enricher()` runs once per contrast and direction with that contrast's
universe, and `merge_result()` combines the runs into the same `compareClusterResult` object
`compareCluster()` would return; `enrichplot::dotplot()` draws it.

Writes, in `03_analyses/04_clusterprofiler/`: `clusterprofiler_enriched.csv`,
`clusterprofiler_all_terms_TC_BP.csv`, `clusterprofiler_run_summary.csv` and
`clusterprofiler_<family>_BP_dotplot.png`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(here::here("01_code", "_go_helpers.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({
  library(clusterProfiler); library(enrichplot); library(GO.db); library(dplyr); library(ggplot2)
})
annot <- read_annotation()
mt    <- mt_genes(annot)
cores <- if (.Platform$OS.type == "windows") 1L else as.integer(params$cores)
runs  <- runs_for(unlist(params$families))
onts  <- unlist(params$ontologies)
prop  <- setNames(lapply(onts, function(o) gene2go_propagated(annot, o)), onts)
t2n   <- lapply(prop, function(p) data.frame(GO = unique(p$GO), name = go_term(unique(p$GO))))
```

# Run enricher

```{r enricher}
run_cp <- function(code, direction, ontology) {
  s   <- gene_sets(code)
  pg  <- prop[[ontology]]
  uni <- intersect(s$universe, unique(pg$gene))
  deg <- intersect(s[[direction]], uni)
  summ <- data.frame(code, direction, ontology, genes_annotated = length(uni), DEGs_annotated = length(deg))
  x <- if (length(deg) >= 2)
    enricher(deg, universe = uni, TERM2GENE = pg[pg$gene %in% uni, c("GO", "gene")],
             TERM2NAME = t2n[[ontology]], minGSSize = MIN_GS, maxGSSize = MAX_GS,
             pvalueCutoff = 1, qvalueCutoff = 1, pAdjustMethod = "BH")
  if (is.null(x) || !nrow(x@result))
    return(list(summary = cbind(summ, terms_tested = 0L, enriched = 0L), terms = NULL, obj = NULL))
  r <- x@result
  genes <- strsplit(r$geneID, "/", fixed = TRUE)
  terms <- data.frame(code, direction, ontology, GO.ID = r$ID, Count = r$Count,
                      GeneRatio = r$GeneRatio, BgRatio = r$BgRatio, p = r$pvalue, padj = r$p.adjust,
                      n_mt_encoded = vapply(genes, function(g) sum(g %in% mt), 0L),
                      genes = vapply(genes, function(g) paste(sort(g), collapse = ";"), ""))
  x@result <- r[r$p.adjust < params$padj_cutoff, ]
  list(summary = cbind(summ, terms_tested = nrow(r), enriched = nrow(x@result)), terms = terms,
       obj = if (nrow(x@result)) x)
}

jobs <- merge(runs, data.frame(ontology = onts), by = NULL)
jobs <- jobs[order(match(jobs$code, contrasts$code), jobs$direction != "up", jobs$ontology), ]
res  <- parallel::mclapply(seq_len(nrow(jobs)), function(i)
  run_cp(jobs$code[i], jobs$direction[i], jobs$ontology[i]), mc.cores = cores)
stopifnot(!vapply(res, inherits, TRUE, what = "try-error"))

terms <- bind_rows(lapply(res, `[[`, "terms")) %>% left_join(distinct(runs, code, family), by = "code") %>%
  mutate(Term = go_term(GO.ID))
run_summary <- bind_rows(lapply(res, `[[`, "summary")) %>% left_join(distinct(runs, code, family), by = "code")
write.csv(run_summary, file.path(out_cp, "clusterprofiler_run_summary.csv"), row.names = FALSE)
enriched <- terms %>% filter(padj < params$padj_cutoff) %>%
  select(code, family, direction, ontology, GO.ID, Term, Count, GeneRatio, BgRatio, p, padj, n_mt_encoded, genes) %>%
  arrange(match(code, contrasts$code), desc(direction == "up"), ontology, padj)
write.csv(enriched, file.path(out_cp, "clusterprofiler_enriched.csv"), row.names = FALSE)
write.csv(terms %>% filter(family == "TC", ontology == "BP") %>% select(code, direction, GO.ID, Count, p, padj),
          file.path(out_cp, "clusterprofiler_all_terms_TC_BP.csv"), row.names = FALSE)
knitr::kable(run_summary %>% filter(ontology == "BP"), caption = "BP runs")
```

# compareCluster view

```{r dotplots, fig.width = 11, fig.height = 9}
for (fam in unique(runs$family)) {
  idx <- which(jobs$ontology == "BP" & jobs$family == fam & !vapply(res, function(r) is.null(r$obj), TRUE))
  if (!length(idx)) next
  objs <- setNames(lapply(res[idx], `[[`, "obj"),
                   paste(contrast_label(jobs$code[idx]), jobs$direction[idx], sep = ", "))
  merged <- merge_result(objs)
  merged@compareClusterResult$Cluster <- factor(merged@compareClusterResult$Cluster, levels = names(objs))
  p <- suppressMessages(dotplot(merged, showCategory = 5, label_format = 60, font.size = 9) +
    scale_fill_gradient(low = "#1a1a1a", high = "#cfcfcf", name = "adjusted p")) +   # neutral: red and blue mean up and down
    labs(x = NULL, title = "clusterProfiler, biological process", subtitle = FAMILY_TITLE[[fam]],
         caption = paste(strwrap(sprintf(paste(
           "The 5 most significant terms of each contrast and direction (BH-adjusted hypergeometric p < %s).",
           "Runs with no enriched term are not shown. Universe: the contrast's tested genes; terms of 10 to 500 genes."),
           params$padj_cutoff), width = 120), collapse = "\n")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), plot.caption = element_text(hjust = 0),
          plot.caption.position = "plot", plot.title.position = "plot")
  n_terms <- length(unique(as.data.frame(merged)$Description[
    as.data.frame(merged) %>% group_by(Cluster) %>% mutate(r = rank(p.adjust, ties.method = "first")) %>% pull(r) <= 5]))
  ggsave(file.path(out_cp, sprintf("clusterprofiler_%s_BP_dotplot.png", fam)), p,
         width = 6.5 + 0.55 * length(objs), height = 2.8 + 0.2 * n_terms, dpi = 300, limitsize = FALSE)
  print(p)
}
```
