---
title: "06 - Do topGO, goseq and clusterProfiler agree?"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Compares the three methods on the treatment-control (TC) contrasts, in each ontology (BP, MF,
CC), the same genes and universes throughout:

| method | annotation | test | enriched |
|---|---|---|---|
| topGO `weight01` (script 02, of record) | direct, graph walked by topGO | Fisher, parents down-weighted by significant children | p < 0.01 |
| topGO `classic` (script 02) | direct, propagated by topGO | Fisher, each term on its own | BH p < 0.05 |
| goseq (script 03) | propagated | Wallenius, weighted by transcript length | BH p < 0.05 |
| clusterProfiler (script 04) | propagated | hypergeometric | BH p < 0.05 |

`classic` Fisher and clusterProfiler's hypergeometric test are the same test on the same
propagated annotation, so they should agree closely (clusterProfiler also drops terms over
500 genes); goseq differs from them by the length weighting; `weight01` differs by favouring
the most specific terms, so it calls fewer broad parents and more specific children, and its
overlap with the other three is expected to be partial.

For each run: the number of enriched terms per method, the Jaccard overlap of each pair's
enriched sets, and the Spearman correlation of their p-values over the terms both tested
(clusterProfiler lists only terms holding at least one DEG). Writes, in
`03_analyses/06_method-comparison/`, one set per ontology: `method_counts_TC_<ontology>.csv`,
`method_agreement_TC_<ontology>.csv` (per run and pair), `method_pair_summary_TC_<ontology>.csv`
(per pair), `consensus_terms_TC_<ontology>.csv` (terms enriched by `weight01` and by at least
one other method) and `method_comparison_TC_<ontology>.png`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(here::here("01_code", "_go_helpers.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(ggplot2); library(patchwork) })
ONTOLOGIES <- c("BP", "MF", "CC")
METHODS <- c(weight01 = "topGO weight01 (p < 0.01)", classic = "topGO classic (BH < 0.05)",
             goseq = "goseq (BH < 0.05)", clusterProfiler = "clusterProfiler (BH < 0.05)")
runs <- runs_for("TC")
read_long <- function(ont) {
  tg <- read.csv(file.path(out_topgo, sprintf("topgo_all_terms_TC_%s.csv", ont))) %>%
    group_by(code, direction) %>% mutate(padj_classic = p.adjust(p_classic, "BH")) %>% ungroup()
  gs <- read.csv(file.path(out_goseq, sprintf("goseq_all_terms_TC_%s.csv", ont)))
  cp <- read.csv(file.path(out_cp, sprintf("clusterprofiler_all_terms_TC_%s.csv", ont)))
  list(tg = tg, long = bind_rows(
    tg %>% transmute(code, direction, GO.ID, method = "weight01", p = p_weight01, enriched = p_weight01 < 0.01),
    tg %>% transmute(code, direction, GO.ID, method = "classic", p = p_classic, enriched = padj_classic < 0.05),
    gs %>% transmute(code, direction, GO.ID, method = "goseq", p = p_over, enriched = padj < 0.05),
    cp %>% transmute(code, direction, GO.ID, method = "clusterProfiler", p = p, enriched = padj < 0.05)))
}
```

# Counts, overlap and correlation

```{r compare}
compare_ontology <- function(ont) {
  x <- read_long(ont); long <- x$long; tg <- x$tg
  counts <- long %>% group_by(code, direction, method) %>% summarise(enriched = sum(enriched), .groups = "drop") %>%
    complete(nesting(code, direction), method = names(METHODS), fill = list(enriched = 0L))
  write.csv(counts %>% pivot_wider(names_from = method, values_from = enriched) %>%
              arrange(match(code, contrasts$code), desc(direction == "up")),
            file.path(out_cmp, sprintf("method_counts_TC_%s.csv", ont)), row.names = FALSE)

  pairs <- combn(names(METHODS), 2, simplify = FALSE)
  agree <- bind_rows(lapply(seq_len(nrow(runs)), function(i) bind_rows(lapply(pairs, function(pr) {
    a <- long %>% filter(code == runs$code[i], direction == runs$direction[i], method == pr[1])
    b <- long %>% filter(code == runs$code[i], direction == runs$direction[i], method == pr[2])
    ea <- a$GO.ID[a$enriched]; eb <- b$GO.ID[b$enriched]
    both <- inner_join(a, b, by = "GO.ID")
    data.frame(code = runs$code[i], direction = runs$direction[i], method_1 = pr[1], method_2 = pr[2],
               enriched_1 = length(ea), enriched_2 = length(eb), shared = length(intersect(ea, eb)),
               jaccard = if (length(ea) && length(eb)) length(intersect(ea, eb)) / length(union(ea, eb)) else NA,
               terms_compared = nrow(both),
               spearman = if (nrow(both) > 10) suppressWarnings(cor(both$p.x, both$p.y, method = "spearman")) else NA)
  }))))
  write.csv(agree, file.path(out_cmp, sprintf("method_agreement_TC_%s.csv", ont)), row.names = FALSE)
  pair_summary <- agree %>% group_by(method_1, method_2) %>%
    summarise(runs_both_enriched = sum(enriched_1 > 0 & enriched_2 > 0),
              runs_only_one = sum(xor(enriched_1 > 0, enriched_2 > 0)),
              median_jaccard_when_both = median(jaccard, na.rm = TRUE),
              median_spearman = median(spearman, na.rm = TRUE), .groups = "drop")
  write.csv(pair_summary, file.path(out_cmp, sprintf("method_pair_summary_TC_%s.csv", ont)), row.names = FALSE)

  consensus <- long %>% filter(enriched) %>% group_by(code, direction, GO.ID) %>%
    summarise(methods = paste(method[order(match(method, names(METHODS)))], collapse = ";"),
              n_methods = n(), .groups = "drop") %>%
    filter(grepl("weight01", methods), n_methods >= 2) %>%
    left_join(tg %>% select(code, direction, GO.ID, Annotated, Significant, p_weight01), by = c("code", "direction", "GO.ID")) %>%
    mutate(Term = go_term(GO.ID)) %>%
    arrange(match(code, contrasts$code), desc(direction == "up"), p_weight01) %>%
    select(code, direction, GO.ID, Term, Annotated, Significant, p_weight01, methods, n_methods)
  write.csv(consensus, file.path(out_cmp, sprintf("consensus_terms_TC_%s.csv", ont)), row.names = FALSE)
  list(counts = counts, agree = agree, pairs = pairs, pair_summary = pair_summary, consensus = consensus)
}
cmp <- setNames(lapply(ONTOLOGIES, compare_ontology), ONTOLOGIES)
for (o in ONTOLOGIES) {
  print(knitr::kable(cmp[[o]]$pair_summary, digits = 2,
                     caption = sprintf("%s: over the 12 TC runs; Jaccard only where both methods enriched a term", o)))
  print(knitr::kable(cmp[[o]]$consensus, digits = 6, caption = sprintf("%s consensus terms", o)))
}
```

# Figures

Panel A is a small multiple, one column per method in a single grey, so the figure needs no
method colours: four more hues could not be told apart for colour-vision deficiency without
reusing the treatment, direction or tissue colours, which mean something else everywhere
else in the repository.

```{r figure, fig.width = 13, fig.height = 6.5}
run_levels <- rev(paste(rep(contrast_label(unique(runs$code)), each = 2), c("up", "down")))
short <- c(weight01 = "weight01", classic = "classic", goseq = "goseq", clusterProfiler = "clusterProfiler")
draw_comparison <- function(ont) {
  x <- cmp[[ont]]
  cnt <- x$counts %>% mutate(run = factor(paste(contrast_label(code), direction), levels = run_levels),
                             method = factor(method, levels = names(METHODS), labels = unname(METHODS)))
  pa <- ggplot(cnt, aes(enriched, run)) +
    geom_col(fill = "grey45", width = 0.7) +
    geom_text(aes(label = ifelse(enriched > 0, enriched, "")), hjust = -0.25, size = 2.6) +
    facet_wrap(~ method, nrow = 1, labeller = label_wrap_gen(width = 16)) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.25))) +
    labs(x = sprintf("enriched %s terms", ont), y = NULL, title = "A. Enriched terms per method") +
    theme_psmfc(9) + theme(panel.grid.major.y = element_blank())
  jac <- x$agree %>% mutate(run = factor(paste(contrast_label(code), direction), levels = run_levels),
                            pair = factor(paste(short[method_1], "vs", short[method_2]),
                                          levels = unique(paste(short[sapply(x$pairs, `[`, 1)], "vs",
                                                                short[sapply(x$pairs, `[`, 2)]))))
  pb <- ggplot(jac, aes(pair, run)) +
    geom_tile(aes(fill = jaccard), colour = "white", linewidth = 0.6) +
    geom_text(aes(label = ifelse(is.na(jaccard), "-", sprintf("%.2f", jaccard))), size = 2.5) +
    scale_fill_gradient(low = "#f2f2f2", high = "#5a5a5a", limits = c(0, 1), na.value = "#ffffff", name = "Jaccard") +
    labs(x = NULL, y = NULL, title = "B. Overlap of enriched sets") +
    theme_psmfc(9) + theme(axis.text.x = element_text(angle = 35, hjust = 1), panel.grid = element_blank(),
                           axis.text.y = element_blank(), axis.ticks.y = element_blank())
  p <- (pa | pb) + plot_layout(widths = c(1.6, 1)) +
    plot_annotation(title = sprintf("GO enrichment methods compared: TC contrasts, %s", ONT_NAME[[ont]]),
                    caption = paste("Same genes and universes for every method. Jaccard = shared / either;",
                                    "'-': at least one of the two methods enriched no term in that run."),
                    theme = theme(plot.caption = element_text(hjust = 0)))
  ggsave(file.path(out_cmp, sprintf("method_comparison_TC_%s.png", ont)), p, width = 13, height = 6.5, dpi = 300)
  p
}
for (o in ONTOLOGIES) print(draw_comparison(o))
```
