---
title: "11-byssal_thread_by_sample"
output: html_document
date: "2024-12-12"
---
### Load libraries
```{r}
library(tidyverse)
library(ggplot2)
library(readr)
```

### Load in gene count matrix and gene loc/gene ID key
```{r}
gene_count_matrix_clean <- read_csv("/home/shared/8TB_HDD_02/graceleuchtenberger/byssus-exp-analysis/output/gene_count_matrix_clean.csv")

loc_go_blast <- read_delim("/home/shared/8TB_HDD_02/graceleuchtenberger/byssus-exp-analysis/output/LOC_GO_list.txt",delim = "\t")

colnames(loc_go_blast)[colnames(loc_go_blast) == "gene"] ="loc_id"
```

```{r}
gene_count_matrix_clean <- gene_count_matrix_clean %>%
  mutate(loc_id = sub(".*\\|", "", ...1)) %>% # Extract part after the pipe
  select(loc_id, everything()) 


# View the updated data frame
print(df)
```

```{r}
annot_GCM <- left_join(gene_count_matrix_clean, loc_go_blast[,c("loc_id", "Protein.names", "uniprot_accession.y", "Gene.Names", "Organism")], by= "loc_id")

annot_GCM <- annot_GCM %>%
  select(loc_id, Protein.names, uniprot_accession.y, Gene.Names, Organism, everything())
```

# Creating a gene count matrix of byssal thread genes
```{r}
foot_GCM <- annot_GCM %>%
  filter(grepl("foot", Protein.names, ignore.case = TRUE))

byss_GCM <- annot_GCM %>%
  filter(grepl("byss", Protein.names, ignore.case = TRUE))

foot_GCM <- foot_GCM %>%
  distinct(uniprot_accession.y, .keep_all = TRUE)

byss_GCM <- byss_GCM %>%
  distinct(uniprot_accession.y, .keep_all = TRUE)

foot_byss_GCM <- rbind(foot_GCM, byss_GCM)


write.csv(foot_byss_GCM , file = "/home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis/output/foot_byss_GCM.csv")
```

# Clean up foot-byss gene count matrix and join to treatment info
```{r}
foot_GCM_1 <- foot_byss_GCM %>%
  pivot_longer(!1:6, names_to = "sample", values_to = "count")

colnames(treatmentinfo.2)[colnames(treatmentinfo.2) == "sample.1"] ="sample"

foot_byss_GCM_2 <- left_join(foot_GCM_1, treatmentinfo.2, by = "sample")

write.csv(foot_byss_GCM_2, file = "/home/shared/8TB_HDD_02/graceleuchtenberger/Github/byssus-exp-analysis/output/foot_byss_GCM.csv")
```

# Read in foot-byssus gene count matrix
```{r}
foot_byss_GCM_2 <- read_csv("/home/shared/8TB_HDD_02/graceleuchtenberger/byssus-exp-analysis/output/foot_byss_GCM.csv")
```

# Graph the log transcript counts of foot and byssus genes in foot tissue
```{r}
foot_byss_gene_plot <- foot_byss_GCM_2 %>%
  filter(tissue == "F") %>%
  filter(day == 3) %>%
  ggplot(aes(x=Protein.names, y=log(count), fill=treatment))+
  geom_boxplot()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 1), 
                                "inches"))
foot_byss_gene_plot
  
ggsave(foot_byss_gene_plot, file = "/home/shared/8TB_HDD_02/graceleuchtenberger/byssus-exp-analysis/output/foot_byss_gene_plot.pdf")

```

# Graph the log transcript counts of foot and byssus genes in gill tissue
```{r}
gill_byss_gene_plot<- foot_byss_GCM_2 %>%
  filter(tissue == "G") %>%
  filter(day == 3) %>%
  ggplot(aes(x=Protein.names, y=log(count), fill=treatment))+
  geom_boxplot()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 1), 
                                "inches"))+
  ylim(1,10)

gill_byss_gene_plot


ggsave(gill_byss_gene_plot, file = "output/gill_byss_gene_plot.pdf")
```
### Structural foot protein graph

```{r}
genes_to_plot <- foot_byss_GCM_2 %>%
  filter(tissue == "F", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 0.1) %>%
  pull(Protein.names) 


genes_to_plot_dataframe <- foot_byss_GCM_2 %>%
  filter(tissue == "F", day == 3, treatment == "control") %>%
  group_by(Protein.names, Organism) %>%   # group by both
  summarise(max_count = max(count, na.rm = TRUE), .groups = "drop") %>%
  filter(max_count > 0.1) %>%
  select(Protein.names, Organism)

genes_to_plot <- genes_to_plot[c(-(2:7), -17)]

my_theme <- theme(line = element_line(linewidth=1.2), 
                  rect = element_rect(linewidth=1.2),
                  text = element_text(size=14,color="black"),
                  panel.background = element_blank(),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.text.x = element_text(size=14,color="black",angle=45, hjust =1),
                  axis.text.y = element_text(size=16,color="black"),
                  axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x = element_line(color="black"),
                  axis.ticks.y = element_line(color="black"),
                  # axis.line= element_line(color = "black", size = 0.1),
                  panel.border = element_rect(color = "black", fill=NA, linewidth=1.2),
                  legend.key = element_blank(), # removes background of legend bullets
                  legend.position = "right",
                  plot.margin = unit(c(0.5, 0.5, 0.5, 2), "inches"))

install.packages("ggtext")
library(ggtext)
foot_byss_GCM_2 %>%
  filter(tissue == "F", day == 3, Protein.names %in% genes_to_plot) %>%
  mutate(
    Protein.names = factor(
      Protein.names,
      levels = c(
        "ACDC (Byssal EP-like protein 1) (EP-protein-1)",
        "Foot protein 2",           # moved up
        "Foot protein-4 variant-1",           # moved up
        "Foot protein 10",
        "Foot protein 11",
        "Foot protein 12",
        "Foot protein 13",
        "Foot protein 14",
        "Foot protein 15",
        "Foot protein 17"
      )
    )
  ) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot(alpha = 5/10) +
  theme_classic() +
  labs(fill = "Treatment",
       x = "Structural foot proteins")+
  scale_fill_manual(labels = c("Treatment control", "Hypoxia", "Ocean acidification", "Ocean warming"), values = c("purple", "green", "red", "orange"))+
  my_theme+
  scale_x_discrete(labels = c(expression(paste("Foot protein 1 (", italic("M. coruscus"), ")")),expression(paste("Foot protein 2 (", italic("M. edulis"), ")")),"Foot protein-4 variant-1","Foot protein 10", "Foot protein 11", "Foot protein 12", "Foot protein 13", "Foot protein 14", "Foot protein 15", "Foot protein 17"))


genes_to_plot_list <- foot_byss_GCM_2 %>%
  filter(tissue == "F", day == 3, Protein.names %in% genes_to_plot)
```

### Structural foot protein graph in gill tissue
```{r}
genes_to_plot_G <- foot_byss_GCM_2 %>%
  filter(tissue == "G", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 0) %>%
  pull(Protein.names)

foot_byss_GCM_2 %>%
  filter(tissue == "G", day == 3, Protein.names %in% genes_to_plot_G) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.margin = unit(c(0.5, 0.5, 0.5, 1), "inches"))
```
### HSP isolation and graph
```{r}
HSP_GCM <- annot_GCM %>%
  filter(grepl("heat shock", Protein.names, ignore.case = TRUE))

HSP_GCM <- HSP_GCM %>%
  distinct(uniprot_accession.y, .keep_all = TRUE)

HSP_GCM_1 <- HSP_GCM %>%
  pivot_longer(!1:6, names_to = "sample", values_to = "count")

colnames(treatmentinfo)[colnames(treatmentinfo) == "sample.1"] ="sample"

HSP_GCM_2 <- left_join(HSP_GCM_1, treatmentinfo, by = "sample")

write.csv(HSP_GCM_2, file = "/home/shared/8TB_HDD_02/graceleuchtenberger/byssus-exp-analysis/output/HSP_GCM.csv")

```

# HSP graph in foot tissue
```{r}
genes_to_plot_hsp <- HSP_GCM_2 %>%
  filter(tissue == "F", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 0.1) %>%
  pull(Protein.names) 

my_theme <- theme(line = element_line(linewidth=1.2), 
                  rect = element_rect(linewidth=1.2),
                  text = element_text(size=14,color="black"),
                  panel.background = element_blank(),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.text.x = element_text(size=12,color="black",angle=45, hjust =1),
                  axis.text.y = element_text(size=14,color="black"),
                  axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x = element_line(color="black"),
                  axis.ticks.y = element_line(color="black"),
                  # axis.line= element_line(color = "black", size = 0.1),
                  panel.border = element_rect(color = "black", fill=NA, linewidth=1.2),
                  legend.key = element_blank(), # removes background of legend bullets
                  legend.position = "right",
                  plot.margin = unit(c(0.5, 0.5, 0.5, 2), "inches"))

genes_to_plot_hsp <- genes_to_plot_hsp[-8]

labels_hsp <- c(
  "10 kDa heat shock protein, mitochondrial", 
  "60 kDa heat shock protein, mitochondrial", 
  "97 kDa heat shock protein", 
  "Activator of 90 kDa heat shock protein ATPase homolog 1", 
  "DnaJ homologous subfamily A member 1 (Heat shock 40 kDa protein 4)", 
  "Endoplasmin (Heat shock protein 90 kDa beta member 1)", 
  "HSP70", 
  "Heat shock 70 kDa protein 12A", 
  "Heat shock 70 kDa protein 12B", 
  "Heat shock 70 kDa protein 13", 
  "Heat shock 70 kDa protein 14", 
  "Heat shock cognate 71 kDa protein", 
  "Heat shock factor protein", 
  "Heat shock factor-binding protein 1", 
  "Heat shock protein 70 A1", 
  "Heat shock protein 70 B2", 
  "Heat shock protein 75 kDa, mitochondrial", 
  "Heat shock protein HSP 90-alpha", 
  "Hsp70-binding protein 1 (HspBP1)", 
  "HSPB1-associated protein 1", 
  "Intraflagellar transport protein 25 homolog (Heat shock protein beta-11)", 
  "Stress-70 protein, mitochondrial (Heat shock 70 kDa protein 9)"
)

HSP_GCM_2 %>%
  filter(tissue == "F", day == 3, Protein.names %in% genes_to_plot_hsp) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot(alpha = 5/10) +
  theme_classic() +
  labs(fill = "Treatment",
       x = "Heat shock proteins")+
  scale_fill_manual(labels = c("Treatment control", "Hypoxia", "Ocean acidification", "Ocean warming"), values = c("purple", "green", "red", "orange"))+
  theme(axis.text.x = element_text(color="black",angle=45, hjust = 1),
        plot.margin = unit(c(0.5, 0.5, 0.5, 1.5), "inches"))+
  scale_x_discrete(labels = str_wrap(labels_hsp, width = 40))+
  my_theme

```
# HSP genes in gill tissue
```{r}
genes_to_plot_hspG <- HSP_GCM_2 %>%
  filter(tissue == "G", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 0.1) %>%
  pull(Protein.names) 

genes_to_plot_hspG <- genes_to_plot_hspG[-c(8,16)]

labels_hspG <- c(
  "10 kDa heat shock protein, mitochondrial", 
  "60 kDa heat shock protein, mitochondrial", 
  "97 kDa heat shock protein", 
  "Activator of 90 kDa heat shock protein ATPase homolog 1", 
  "DnaJ homologous subfamily A member 1 (Heat shock 40 kDa protein 4)", 
  "Endoplasmin (Heat shock protein 90 kDa beta member 1)", 
  "HSP70", 
  "Heat shock 70 kDa protein 12A", 
  "Heat shock 70 kDa protein 12B",
  "Heat shock 70 kDa protein 12B (Heat shock protein family A member 12B)",
  "Heat shock 70 kDa protein 13", 
  "Heat shock 70 kDa protein 14", 
  "Heat shock cognate 71 kDa protein", 
  "Heat shock factor protein",
  "Heat shock factor-binding protein 1", 
  "Heat shock protein 70 A1", 
  "Heat shock protein 70 B2", 
  "Heat shock protein 75 kDa, mitochondrial", 
  "Heat shock protein HSP 90-alpha", 
  "Hsp70-binding protein 1 (HspBP1)", 
  "HSPB1-associated protein 1", 
  "Intraflagellar transport protein 25 homolog (Heat shock protein beta-11)", 
  "Stress-70 protein, mitochondrial (Heat shock 70 kDa protein 9)"
)

HSP_GCM_2 %>%
  filter(tissue == "G", day == 3, Protein.names %in% genes_to_plot_hspG) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot(alpha = 5/10) +
  theme_classic() +
  labs(fill = "Treatment",
       x = "Heat shock proteins")+
  scale_fill_manual(labels = c("Treatment control", "Hypoxia", "Ocean acidification", "Ocean warming"), values = c("purple", "green", "red", "orange"))+
  theme(axis.text.x = element_text(color="black",angle=45, hjust = 1),
        plot.margin = unit(c(0.5, 0.5, 0.5, 2), "inches"))+
  scale_x_discrete(labels = str_wrap(labels_hspG, width = 40))+
  #scale_x_discrete(labels = function(x) str_wrap(x, width = 40))+
  my_theme
```
# HIF gene isolation from GCM
```{r}
### HIF isolation and graph

HIF_GCM <- annot_GCM %>%
  filter(grepl("hypoxia", Protein.names, ignore.case = TRUE))

HIF_GCM <- HIF_GCM %>%
  distinct(uniprot_accession.y, .keep_all = TRUE)

HIF_GCM_1 <- HIF_GCM %>%
  pivot_longer(!1:6, names_to = "sample", values_to = "count")


HIF_GCM_2 <- left_join(HIF_GCM_1, treatmentinfo, by = "sample")

write.csv(HIF_GCM_2, file = "/home/shared/8TB_HDD_02/graceleuchtenberger/byssus-exp-analysis/output/HIF_GCM.csv")


```

# Graphing HIF transcript counts in foot tissue
```{r}
genes_to_plot_hifF <- HIF_GCM_2 %>%
  filter(tissue == "F", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 0.1) %>%
  pull(Protein.names) 


genes_to_plot_hifF <- genes_to_plot_hifF[-1]

labels_hifF <- c(
  "Aryl hydrocarbon receptor nuclear translocator (Hypoxia-inducible factor 1-beta)",
  "Egl nine homologous 1 (Hypoxia-inducible factor prolyl hydroxylase 2)",
  "Endothelial PAS domain-containing protein 1 (Hypoxia-inducible factor 2-alpha)",
  "Hypoxia upregulated protein 1", 
  "Hypoxia-inducible factor 1-alpha",
  "Hypoxia-inducible factor 1-alpha inhibitor",
  "Transmembrane prolyl 4-hydroxylase (Hypoxia-inducible factor prolyl hydroxylase 4)"
)

HIF_GCM_2 %>%
  filter(tissue == "F", day == 3, Protein.names %in% genes_to_plot_hifF) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot(alpha = 5/10) +
  theme_classic() +
  labs(fill = "Treatment",
       x = "Hypoxia-inducible factor proteins")+
  scale_fill_manual(labels = c("Treatment control", "Hypoxia", "Ocean acidification", "Ocean warming"), values = c("purple", "green", "red", "orange"))+
  theme(axis.text.x = element_text(color="black",angle=45, hjust = 1),
        plot.margin = unit(c(0.5, 0.5, 0.5, 2), "inches"))+
  scale_x_discrete(labels = str_wrap(labels_hifF, width = 40))+
  my_theme
```

# Graphing HIF transcript counts in gill tissue
```{r}
genes_to_plot_hifG <- HIF_GCM_2 %>%
  filter(tissue == "G", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 0.1) %>%
  pull(Protein.names) 

genes_to_plot_hifG <- genes_to_plot_hifG[-1]

labels_hifG <- c(
  "Aryl hydrocarbon receptor nuclear translocator (Hypoxia-inducible factor 1-beta)",
  "Egl nine homologous 1 (Hypoxia-inducible factor prolyl hydroxylase 2)",
  "Endothelial PAS domain-containing protein 1 (Hypoxia-inducible factor 2-alpha)",
  "Hypoxia upregulated protein 1", 
  "Hypoxia-inducible factor 1-alpha",
  "Hypoxia-inducible factor 1-alpha inhibitor",
  "Transmembrane prolyl 4-hydroxylase (Hypoxia-inducible factor prolyl hydroxylase 4)"
)

HIF_GCM_2 %>%
  filter(tissue == "G", day == 3, Protein.names %in% genes_to_plot_hifG) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot(alpha = 5/10) +
  theme_classic() +
  labs(fill = "Treatment",
       x = "Hypoxia-inducible factor proteins")+
  scale_fill_manual(labels = c("Treatment control", "Hypoxia", "Ocean acidification", "Ocean warming"), values = c("purple", "green", "red", "orange"))+
  theme(axis.text.x = element_text(color="black",angle=45, hjust = 1),
        plot.margin = unit(c(0.5, 0.5, 0.5, 2), "inches"))+
  scale_x_discrete(labels = str_wrap(labels_hifG, width = 40))+
  my_theme
```
### Peroxidase/catalase transcript isolation
```{r}
toMatch <- c("peroxidase", "catalase")

perox_GCM <- annot_GCM %>%
  filter(grepl(paste(toMatch,collapse="|"), Protein.names, ignore.case = TRUE))

perox_GCM <- perox_GCM %>%
  distinct(uniprot_accession.y, .keep_all = TRUE)

perox_GCM_1 <- perox_GCM %>%
  pivot_longer(!1:6, names_to = "sample", values_to = "count")


perox_GCM_2 <- left_join(perox_GCM_1, treatmentinfo, by = "sample")

write.csv(perox_GCM_2, file = "/home/shared/8TB_HDD_02/graceleuchtenberger/byssus-exp-analysis/output/perox_GCM.csv")


```

# Graphing peroxidase/catalase transcript counts in foot tissue
```{r}
genes_to_plot_peroxF <- perox_GCM_2 %>%
  filter(tissue == "F", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 1) %>%
  pull(Protein.names) 

genes_to_plot_peroxF <- genes_to_plot_peroxF[-c(8,10)]

labels_peroxF <- c(
  "Byssal peroxidase-like protein 4", 
  "Catalase", 
  "Chloride intracellular channel protein 2 (Glutaredoxin-like peroxidase CLIC2)", 
  "Dye-decolorizing peroxidase YfeX", 
  "Glutathione peroxidase (GPX)", 
  "Gluthathione peroxidase (GPx) (GSHPx)", 
  "Glutathione peroxidase 1", 
  "Glutathione peroxidase 7",
  "Glutathione S-transferase 3, mitochondrial (Glutathione peroxidase MGST3)", 
  "Peroxiredoxin 4 (Thioredoxin peroxidase AO372)", 
  "Peroxiredoxin 5, mitochondrial (Thioredoxin peroxidase PMP20)", 
  "Peroxiredoxin 5, mitochondrial (Thioredoxin peroxidase)", 
  "Peroxidredoxin 6 (Glutathione peroxidase 4)", 
  "Phospholipid hydroperoxide glutathione peroxidase ", 
  "Thioredoxin reductase 1, cytoplasmic (Peroxidase TXNRD1)")


perox_GCM_2 %>%
  filter(tissue == "F", day == 3, Protein.names %in% genes_to_plot_peroxF) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot(alpha = 5/10) +
  theme_classic() +
  labs(fill = "Treatment",
       x = "Peroxidase proteins")+
  scale_fill_manual(labels = c("Treatment control", "Hypoxia", "Ocean acidification", "Ocean warming"), values = c("purple", "green", "red", "orange"))+
  theme(axis.text.x = element_text(color="black",angle=45, hjust = 1),
        plot.margin = unit(c(0.5, 0.5, 0.5, 2), "inches"))+
  scale_x_discrete(labels = str_wrap(labels_peroxF, width = 40))+
  #scale_x_discrete(labels = function(x) str_wrap(x, width = 40))+
  my_theme
```

# Graphing peroxidase/catalase transcript counts in gill tissue
```{r}
genes_to_plot_peroxG <- perox_GCM_2 %>%
  filter(tissue == "G", day == 3, treatment == "control") %>%
  group_by(Protein.names) %>%
  summarise(max_count = max(count, na.rm = TRUE)) %>%
  filter(max_count > 1) %>%
  pull(Protein.names) 

genes_to_plot_peroxG <- genes_to_plot_peroxG[-c(8,12)]

labels_peroxG <- c(
  "Byssal peroxidase-like protein 4", 
  "Catalase", 
  "Chloride intracellular channel protein 2 (Glutaredoxin-like peroxidase CLIC2)", 
  "Dye-decolorizing peroxidase YfeX", 
  "Glutathione peroxidase (GPX)", 
  "Gluthathione peroxidase (GPx) (GSHPx)", 
  "Glutathione peroxidase 1", 
  "Glutathione peroxidase 3", 
  "Glutathione peroxidase 7",
  "Glutathione S-transferase 3, mitochondrial (Glutathione peroxidase MGST3)",
  "Peroxiredoxin-4 (Thioredoxin peroxidase AO372)", 
  "Peroxiredoxin-5, mitochondrial (Thioredoxin peroxidase PMP20)", 
  "Peroxiredoxin-5, mitochondrial (Thioredoxin peroxidase)", 
  "Peroxidredoxin-6 (Glutathione peroxidase 4)", 
  "Phospholipid hydroperoxide glutathione peroxidase", 
  "Thioredoxin reductase 1, cytoplasmic (Peroxidase TXNRD1)")


perox_GCM_2 %>%
  filter(tissue == "G", day == 3, Protein.names %in% genes_to_plot_peroxG) %>%
  ggplot(aes(x = Protein.names, y = log(count), fill = treatment)) +
  geom_boxplot(alpha = 5/10) +
  theme_classic() +
  labs(fill = "Treatment",
       x = "Peroxidase proteins")+
  scale_fill_manual(labels = c("Treatment control", "Hypoxia", "Ocean acidification", "Ocean warming"), values = c("purple", "green", "red", "orange"))+
  theme(axis.text.x = element_text(color="black",angle=45, hjust = 1),
        plot.margin = unit(c(0.5, 0.5, 0.5, 2), "inches"))+
  #scale_x_discrete(labels = function(x) str_wrap(x, width = 40))
  scale_x_discrete(labels = str_wrap(labels_peroxG, width = 40))+
  my_theme
```
