---
title: "1_extract_tensometer_data"
author: "Matt George"
date: "2026-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What this script does

Walks every subfolder of `02_data/tensometer_output/`, reads each raw force/displacement
trace, joins the mussel key, and writes **one** trace-level table to
`03_analyses/extract-tensometer-data/thread-summary-raw-output.xlsx`.

Nothing is excluded. Every trace that parses is exported alongside a QC plot, so bad runs
get curated later against the evidence rather than dropped silently here. That file is the
input to the manual curation step that produces `03_analyses/thread-summary.xlsx`.

Replaces the previous two-pass `control` / `treatment` layout, which no longer exists on disk.

### The label model

Three facts, kept deliberately separate, because **a mussel and its threads did not
necessarily experience the same thing**:

| column | grain | source | meaning |
|---|---|---|---|
| `thread_trt` | thread | folder name | what the **thread** was built in |
| `phase` | thread | folder name | `lab` / `pre` / `post` |
| `mussel_trt` | mussel | `02_data/mussel-treatment-key.csv` | the arm the **animal** was assigned to |

`mussel_trt` is a *destiny* label, not an experienced-condition label. For a `pre` thread it
describes the animal's future, not its past: a baseline thread from an OA animal is **not**
an OA thread. That is the one place the two grains diverge, and keeping them in separate
columns from separate sources is what stops them being conflated.

`phase` has three levels, not two, because the lab-reference animals (day 0, never in the
experimental system) are not the same thing as the paired pre-exposure baselines. A binary
before/after would silently pool them.

`group` from the older `thread-summary.xlsx` is deliberately **not** reproduced: it was fully
recoverable from `thread_trt`, and its value `control` collided with both
`mussel_trt == "control"` and `thread_trt == "treatment_control"`.

Run `0_build_mussel_key.Rmd` first if the morphometrics workbook has changed. If the key is
missing this script still runs; it just warns and leaves the mussel-level columns out.

## Load packages; set ggplot theme

```{r message=FALSE, warning=FALSE}

## Run within "thread-strength.Rproj" so here::here() resolves to the analysis root.

## clear workspace
rm(list = ls())

## Load Packages
load.lib    <- c("readr", "tidyverse", "openxlsx", "EnvStats", "here")  # required packages
install.lib <- load.lib[!load.lib %in% installed.packages()]            # select missing packages
for (lib in install.lib) install.packages(lib, dependencies = TRUE)     # install missing + deps
sapply(load.lib, require, character = TRUE)                             # load all packages

## Set ggplot theme
my_theme <- theme(line              = element_line(linewidth = 1.2),
                  rect              = element_rect(linewidth = 1.2),
                  text              = element_text(size = 14, color = "black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(),
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size = 16, color = "black"),
                  axis.text.y       = element_text(size = 16, color = "black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color = "black"),
                  axis.ticks.y      = element_line(color = "black"),
                  panel.border      = element_rect(color = "black", fill = NA, linewidth = 1.2),
                  legend.key        = element_blank(),
                  legend.position   = "none")
```

## Paths

```{r paths}
tens_root <- here::here("02_data", "tensometer_output")
key_csv   <- here::here("02_data", "mussel-treatment-key.csv")
qc_root   <- here::here("03_analyses", "extract-tensometer-data", "QC_plots")
xlsx_out  <- here::here("03_analyses", "extract-tensometer-data", "thread-summary-raw-output.xlsx")

if (!dir.exists(tens_root))
  stop("Cannot find:\n  ", tens_root,
       "\nOpen `thread-strength.Rproj` (not the repository-root .Rproj) before running.")

dir.create(dirname(xlsx_out), recursive = TRUE, showWarnings = FALSE)
```

## Folder -> thread label map

The subfolder name is the only place the thread treatment is recorded, so this table is the
single point of interpretation in the script. It is deliberately explicit and editable.

`expect_mussel_trt` is not exported. It exists so the script can check the folder against the
morphometrics record: for a `post` folder the thread condition and the animal's arm must
agree, and a disagreement means a file is filed under the wrong folder. It is `NA` for `pre`,
where the two legitimately differ.

The legacy rows keep this script working **before** the folder rename is applied. Delete them
once `rename_tensometer_folders.sh` has been run and pushed.

```{r}
folder_labels <- tibble::tribble(
  ~source_folder,          ~thread_trt,          ~phase, ~day, ~expect_mussel_trt,
  # --- current layout ---------------------------------------------------------------
  "00_lab_reference",      "lab_control",        "lab",   0L,  "lab_control",
  "01_pre_exposure",       "baseline",           "pre",   1L,  NA_character_,
  "02_post_control",       "treatment_control",  "post",  3L,  "control",
  "03_post_OA",            "OA",                 "post",  3L,  "OA",
  "04_post_OW",            "OW",                 "post",  3L,  "OW",
  "05_post_DO",            "DO",                 "post",  3L,  "DO",
  # --- legacy names, pre-rename (safe to delete once renamed) ------------------------
  "00_laboratory_control", "lab_control",        "lab",   0L,  "lab_control",
  "00_baseline",           "baseline",           "pre",   1L,  NA_character_,
  "01_treatment_control",  "treatment_control",  "post",  3L,  "control",
  "02_OA_treatment",       "OA",                 "post",  3L,  "OA",
  "03_OW_treatment",       "OW",                 "post",  3L,  "OW",
  "04_DO_treatment",       "DO",                 "post",  3L,  "DO"
)

# Fallback for a folder that is not in the map. Deliberately crude, and always warned about,
# so a new folder is labelled provisionally rather than silently mislabelled.
derive_label <- function(folder) {
  bare <- sub("^[0-9]+_", "", folder)
  tibble(source_folder     = folder,
         thread_trt        = sub("^(pre|post|lab)_", "", sub("_treatment$", "", bare)),
         phase             = dplyr::case_when(grepl("^lab", bare)  ~ "lab",
                                              grepl("^pre", bare)  ~ "pre",
                                              TRUE                 ~ "post"),
         day               = NA_integer_,
         expect_mussel_trt = NA_character_)
}

folders <- sort(list.dirs(tens_root, full.names = FALSE, recursive = FALSE))
folders <- folders[nzchar(folders)]
if (length(folders) == 0) stop("No subfolders under ", tens_root)

novel <- setdiff(folders, folder_labels$source_folder)
if (length(novel) > 0) {
  warning("Unmapped tensometer subfolder(s): ", paste(novel, collapse = ", "),
          ". Labels were guessed from the folder name and `day` is NA; add them to ",
          "`folder_labels` and confirm the phase.")
  folder_labels <- bind_rows(folder_labels, bind_rows(lapply(novel, derive_label)))
}

# Keep only folders that exist, in on-disk order. This is what makes the legacy rows harmless:
# the old and new names never coexist.
folder_labels <- folder_labels %>%
  filter(source_folder %in% folders) %>%
  arrange(match(source_folder, folders))

if (any(duplicated(folder_labels$source_folder)))
  stop("Duplicate rows in `folder_labels` for: ",
       paste(unique(folder_labels$source_folder[duplicated(folder_labels$source_folder)]),
             collapse = ", "))

knitr::kable(folder_labels)
```

## Parse trace filenames into mussel ID, thread number, and note

Tensometer files are named `<prefix><digits>_<digits><optional note>` and may or may not
carry a `.txt` extension (the instrument exports them without one; the extension was added by
hand for some files and missed for others). Any trailing text is the technician's note
(`try2`, `fluke`, `thirdTry`, ...). It is kept as its own field so every run is imported and
can be checked against its QC plot before curation. Nothing is dropped or merged here.

`mussel` + `thread` is **not** unique across the dataset: 45 animals were pulled before and
after exposure, so the same pair appears in two folders. `source_folder` is part of the key.

```{r}
# Extension-agnostic and width-agnostic (works for 2-digit IDs like G18_01 and 3-digit IDs
# like T037_03). Returns NULL for a name that is not a trace.
parse_trace_name <- function(file_name) {
  base <- sub("\\.txt$", "", file_name, ignore.case = TRUE)
  if (!grepl("^[A-Za-z]+[0-9]+_[0-9]+", base)) return(NULL)
  list(mussel = sub("^([A-Za-z]+[0-9]+)_.*$", "\\1", base),
       thread = as.integer(sub("^[A-Za-z]+[0-9]+_([0-9]+).*$", "\\1", base)),
       note   = sub("^[A-Za-z]+[0-9]+_[0-9]+", "", base))
}

# List every trace in a folder regardless of extension. Anything not matching the naming
# convention is reported via warning() rather than skipped silently, so a stray file
# (a spreadsheet, a .DS_Store) can never be read as data unnoticed.
list_trace_files <- function(dir) {
  all_files   <- list.files(dir)
  trace_files <- all_files[grepl("^[A-Za-z]+[0-9]+_[0-9]+", all_files)]
  skipped     <- setdiff(all_files, trace_files)
  if (length(skipped) > 0)
    warning(basename(dir), ": ", length(skipped),
            " file(s) did not match the trace pattern and were skipped: ",
            paste(skipped, collapse = ", "))
  sort(trace_files)
}
```

## Read one trace

The instrument writes each trace **wide**: a `Time / Displacement / Force` header line
followed by three tab-separated rows, one per channel. Reading the lines directly is both
faster and safer than the old `read.delim()` + transpose, which had to infer a column count
from a 773-field row.

A missing **first** Force sample (a literal `NaN`) is normal instrument behaviour at the start
of a pull and is set to zero. A missing sample anywhere else is also zeroed, but warns with
the file and index, because that is a dropout rather than a start-up artifact and should be
checked against the QC plot.

```{r}
read_trace <- function(path) {
  lines <- readLines(path, warn = FALSE)
  lines <- lines[trimws(lines) != ""]

  # Drop the channel-name header if present. Detect it by row count, not by content:
  # as.numeric("NaN") is a valid number, so a content test would misfire on the leading NaN.
  if (length(lines) == 4L) {
    lines <- lines[-1]
  } else if (length(lines) != 3L) {
    stop(basename(path), ": expected 3 channel rows (optionally after a header), found ",
         length(lines), ". This file is not in the instrument's wide format.")
  }

  to_num <- function(l) {
    tok <- strsplit(l, "\t", fixed = TRUE)[[1]]
    suppressWarnings(as.numeric(tok[trimws(tok) != ""]))
  }
  time <- to_num(lines[1]); disp <- to_num(lines[2]); force <- to_num(lines[3])

  n <- length(time)
  if (length(disp) != n || length(force) != n)
    stop(basename(path), ": channel rows have unequal lengths (",
         n, "/", length(disp), "/", length(force), ").")
  if (n < 3L)
    stop(basename(path), ": only ", n, " sample(s); too short to integrate.")

  # Count missing values before filling, so the QC record survives into the output.
  n_na <- c(time = sum(is.na(time)), disp = sum(is.na(disp)), force = sum(is.na(force)))

  zero_fill <- function(v, channel) {
    bad <- which(is.na(v))
    if (length(bad) > 0) {
      non_leading <- bad[bad != 1L]
      if (length(non_leading) > 0)
        warning(basename(path), ": ", channel, " has ", length(non_leading),
                " missing sample(s) away from the start (index ",
                paste(utils::head(non_leading, 10), collapse = ", "),
                "); zero-filled - check the QC plot.", call. = FALSE)
      v[bad] <- 0
    }
    v
  }
  time  <- zero_fill(time,  "time")
  disp  <- zero_fill(disp,  "displacement")
  force <- zero_fill(force, "force")

  # Integration assumes increasing time. Sort rather than trust, and say so if it was not.
  if (is.unsorted(time, strictly = FALSE)) {
    warning(basename(path), ": time was not monotonic; samples reordered before integration.",
            call. = FALSE)
    o <- order(time); time <- time[o]; disp <- disp[o]; force <- force[o]
  }

  list(time = time, displacement = disp, force = force, n_na = n_na)
}

# Area under the force/time curve, by trapezoid, on the raw trace.
#
# The previous version called loess() then integrated `loess()$y`, which is the response, not
# `$fitted` - so it was already integrating the raw trace, but on a grid from
# approx(n = length(x)) that did not line up with the diff(x) it was multiplied by. This is the
# same quantity, computed correctly. Across the current 375 traces the two agree to a median of
# 0.006% and a maximum of 2.1%. loess() is still fitted below, for the QC plot only.
trapz <- function(y, x) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
```

## Extract every trace

```{r warning=TRUE}
records <- list()

for (folder in folders) {
  fdir  <- file.path(tens_root, folder)
  files <- list_trace_files(fdir)
  lab   <- folder_labels[folder_labels$source_folder == folder, ]

  qc_dir <- file.path(qc_root, folder)
  dir.create(qc_dir, recursive = TRUE, showWarnings = FALSE)

  message("Reading ", folder, ": ", length(files), " trace(s)")

  for (file_name in files) {
    ids <- parse_trace_name(file_name)
    if (is.null(ids)) next          # already reported by list_trace_files()
    tr  <- read_trace(file.path(fdir, file_name))

    # QC plot, named after the source file so retries and flukes never overwrite each other.
    qc_name <- sub("\\.txt$", "", file_name, ignore.case = TRUE)
    fit <- tryCatch(loess(force ~ time, data = data.frame(time = tr$time, force = tr$force)),
                    error = function(e) NULL)
    if (is.null(fit))
      warning(file_name, ": loess fit failed; QC plot drawn without a smoother.", call. = FALSE)

    jpeg(file.path(qc_dir, paste0(qc_name, ".jpg")))
    tryCatch({
      plot(tr$time, tr$force,
           main = paste0(qc_name, "  (", folder, ")"),
           xlab = "Time (s)", ylab = "Force (N)", col = "blue")
      if (!is.null(fit)) lines(fit, col = "red")
    }, finally = dev.off())   # never leave a graphics device open if plotting fails

    records[[length(records) + 1L]] <- tibble(
      source_folder    = folder,
      thread_trt       = lab$thread_trt,
      phase            = lab$phase,
      day              = lab$day,
      mussel           = ids$mussel,
      thread           = ids$thread,
      note             = ids$note,
      max_force        = max(tr$force),                  # N
      integral         = trapz(tr$force, tr$time),       # N*s, area under force/time
      max_displacement = max(tr$displacement),           # mm at peak extension
      n_points         = length(tr$time),
      duration_s       = max(tr$time) - min(tr$time),
      n_na_force       = unname(tr$n_na["force"]),       # missing force samples, zero-filled
      file             = file_name)
  }
}

thread_raw <- bind_rows(records) %>%
  arrange(match(source_folder, folders), mussel, thread, note)

message("Extracted ", nrow(thread_raw), " trace(s) from ", length(folders), " folder(s); ",
        n_distinct(thread_raw$mussel), " unique mussels.")
```

## Join the mussel key

Mussel-level facts come from `02_data/mussel-treatment-key.csv`, built by
`0_build_mussel_key.Rmd` from the morphometrics workbook. They are joined, never inferred
from the folder, so there is exactly one source of truth for what an animal was assigned to.

```{r}
# Canonical tag: prefix + zero-padded 3 digits (T37 -> T037), so both sides join regardless
# of how the tag was written.
norm_id <- function(x) {
  x   <- trimws(as.character(x))
  pre <- sub("^([A-Za-z]+).*$", "\\1", x)
  num <- suppressWarnings(as.integer(sub("^[A-Za-z]+0*([0-9]+).*$", "\\1", x)))
  ifelse(is.na(num) | pre == x, NA_character_, sprintf("%s%03d", pre, num))
}

thread_raw <- thread_raw %>% mutate(mussel = norm_id(mussel))

if (file.exists(key_csv)) {
  mussel_key <- readr::read_csv(key_csv, show_col_types = FALSE) %>%
    mutate(mussel = norm_id(mussel)) %>%
    select(mussel, species, mussel_trt, rna_sequenced)

  thread_raw <- thread_raw %>% left_join(mussel_key, by = "mussel")

  orphan <- thread_raw %>% filter(is.na(mussel_trt)) %>% distinct(mussel) %>% pull(mussel)
  if (length(orphan) > 0)
    warning(length(orphan), " mussel(s) with traces but no key row: ",
            paste(orphan, collapse = ", "),
            ". Re-run 0_build_mussel_key.Rmd, or check the tag.")

  # Integrity check: in a `post` folder the thread condition and the animal's arm must agree.
  # A mismatch means a trace is filed under the wrong folder, which no downstream script can
  # detect on its own.
  mismatch <- thread_raw %>%
    left_join(folder_labels %>% select(source_folder, expect_mussel_trt), by = "source_folder") %>%
    filter(!is.na(expect_mussel_trt), !is.na(mussel_trt), expect_mussel_trt != mussel_trt)

  if (nrow(mismatch) > 0) {
    warning(nrow(mismatch), " trace(s) sit in a folder that disagrees with morphometrics.")
    mismatch %>% count(source_folder, expect_mussel_trt, mussel_trt) %>% knitr::kable()
  } else {
    message("Folder arm agrees with morphometrics on all post-exposure traces.")
  }
} else {
  warning("Mussel key not found at ", key_csv,
          ". Run 0_build_mussel_key.Rmd. Mussel-level columns are omitted from the output.")
}

# Final column order: identity, thread labels, mussel labels, measurements, QC.
thread_raw <- thread_raw %>%
  select(any_of(c("source_folder", "thread_trt", "phase", "day",
                  "mussel", "thread", "note",
                  "species", "mussel_trt", "rna_sequenced",
                  "max_force", "integral", "max_displacement",
                  "n_points", "duration_s", "n_na_force", "file")))
```

## Coverage

Written as a second sheet so a short count in an arm is visible without re-deriving it.
`na_force_samples` counts leading-`NaN` start-up artifacts, which are expected.

```{r}
coverage <- thread_raw %>%
  group_by(source_folder, thread_trt, phase, day) %>%
  summarise(traces                 = n(),
            mussels                = n_distinct(mussel),
            threads_per_mussel_min = min(table(mussel)),
            threads_per_mussel_max = max(table(mussel)),
            retry_traces           = sum(note != ""),
            na_force_samples       = sum(n_na_force),
            .groups = "drop") %>%
  arrange(match(source_folder, folders))

knitr::kable(coverage)
```

### Pairing

Which animals have both a pre-exposure and a post-exposure pull. Unpaired animals are not an
error, but they are what makes a per-arm marginal mean unreliable, so the count is worth
seeing before any model is fitted.

```{r}
pairing <- thread_raw %>%
  filter(phase %in% c("pre", "post")) %>%
  distinct(mussel, phase) %>%
  count(mussel, name = "n_phases") %>%
  mutate(paired = n_phases == 2L)

# `mussel_trt` is only present when the key was found; fall back to an ungrouped count.
if ("mussel_trt" %in% names(thread_raw)) {
  pairing <- pairing %>%
    left_join(thread_raw %>% distinct(mussel, mussel_trt), by = "mussel")
  pairing %>%
    count(mussel_trt, paired) %>%
    pivot_wider(names_from = paired, values_from = n,
                names_prefix = "paired_", values_fill = 0) %>%
    knitr::kable()
} else {
  pairing %>% count(paired) %>% knitr::kable()
}
```

## Outlier screen

Peak force by thread treatment, with each point labelled, plus a Rosner test for extreme
values. This flags candidates for curation; it does not remove anything.

```{r fig.width=9, fig.height=5}
plot_dat <- thread_raw %>%
  mutate(ID         = paste(mussel, thread),
         thread_trt = factor(thread_trt, levels = unique(folder_labels$thread_trt)))

ggplot(plot_dat, aes(x = thread_trt, y = max_force, label = ID)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_text(check_overlap = TRUE, position = position_jitter(width = 0.18), size = 3) +
  labs(x = NULL, y = "Peak force (N)") +
  my_theme +
  theme(axis.text.x = element_text(size = 12, angle = 25, hjust = 1))
```

```{r}
# Pooled screen. Rosner assumes an approximately normal bulk and wants n >= 25.
print(rosnerTest(thread_raw$max_force, k = 5)$all.stats)

# Per thread treatment, where the group is large enough to support it.
for (trt in levels(plot_dat$thread_trt)) {
  v <- plot_dat$max_force[plot_dat$thread_trt == trt]
  if (length(v) < 25) {
    message(trt, ": n = ", length(v), ", too small for a Rosner test; skipped.")
    next
  }
  res <- tryCatch(rosnerTest(v, k = min(5, length(v) - 2)), error = function(e) NULL)
  if (is.null(res)) { message(trt, ": Rosner test failed."); next }
  message("--- ", trt, " (n = ", length(v), ") ---")
  print(res$all.stats)
}
```

## Write the combined output

One workbook, three sheets: `data` (one row per trace), `coverage`, and `pairing`. `data` is
sheet 1, so `read_excel(path)` with no sheet argument gets the trace table.

```{r}
write.xlsx(list(data = thread_raw, coverage = coverage, pairing = pairing),
           file = xlsx_out, colNames = TRUE, overwrite = TRUE)

message("Wrote ", xlsx_out, " (", nrow(thread_raw), " rows).")
message("QC plots under ", qc_root)
print("done")
```
