---
title: "1_extract_tensometer_data"
author: "Matt George"
date: "2023-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages; set ggplot theme
```{r message=TRUE, warning=TRUE}

## Run within "thread-strength.Rproj" so here::here() resolves to the analysis root.

## clear workspace
rm(list=ls())

## Load Packages
load.lib<-c("readxl","tidyverse","openxlsx","EnvStats","here") # List of required packages
install.lib <- load.lib[!load.lib %in% installed.packages()] # Select missing packages
for(lib in install.lib) install.packages(lib,dependencies=TRUE) # Install missing packages + dependencies
sapply(load.lib,require,character=TRUE) # Load all packages.

## Set ggplot theme
my_theme <- theme(line              = element_line(linewidth=1.2),
                  rect              = element_rect(linewidth=1.2),
                  text              = element_text(size=14,color="black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(), 
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size=16,color="black"), #,angle=90),
                  axis.text.y       = element_text(size=16,color="black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color="black"),
                  axis.ticks.y      = element_line(color="black"),
                  # axis.line         = element_line(color = "black", size = 0.1),
                  panel.border      = element_rect(color = "black", fill=NA, size=1.2),
                  legend.key        = element_blank(), # removes background of legend bullets
                  legend.position   = "none"
                  )

```


## Parse trace filenames into mussel ID, thread number, and note

Tensometer files are named `<prefix><digits>_<digits><optional note>` and may or may not carry a
`.txt` extension (the instrument exports them without one; the extension was added by hand for some
files and missed for others). The trailing text is the technician's note (`try2`, `fluke`, `tryTwo`,
`thirdTry`, `firstTry`, ...). We keep that note as its own field so every run is imported and can be
checked against its QC plot before any curation. Nothing is dropped or silently merged here.

```{r}
# Split a filename into mussel ID, thread number, and note. Extension-agnostic and width-agnostic
# (works for 2-digit control IDs like G18_01 and 3-digit treatment IDs like T037_03).
parse_trace_name <- function(file_name) {
  base   <- sub("\\.txt$", "", file_name)                    # drop a trailing .txt if present
  mussel <- sub("^([A-Za-z]+[0-9]+)_.*$", "\\1", base)       # prefix + digits before the "_"
  thread <- sub("^[A-Za-z]+[0-9]+_([0-9]+).*$", "\\1", base) # digits immediately after the "_"
  note   <- sub("^[A-Za-z]+[0-9]+_[0-9]+", "", base)         # anything trailing = technician note
  list(mussel = mussel, thread = thread, note = note)
}

# List every tensometer trace in a folder regardless of extension. Anything that does not match the
# trace naming convention is reported via warning() rather than silently skipped, so a stray file
# (e.g. a spreadsheet or .DS_Store dropped in the folder) can never be read as data unnoticed.
list_trace_files <- function(dir) {
  all_files   <- list.files(dir)
  trace_files <- all_files[grepl("^[A-Za-z]+[0-9]+_[0-9]+", all_files)]
  skipped     <- setdiff(all_files, trace_files)
  if (length(skipped) > 0)
    warning(basename(dir), ": ", length(skipped),
            " file(s) did not match the trace pattern and were skipped: ",
            paste(skipped, collapse = ", "))
  trace_files
}
```

## CONTROLS

```{r warning=FALSE}

# extract data from all .txt files in a folder located at 02_data/tensometer_output/control/ and add it to a nested list object
# the data has three column headers: Time, displacement, force. 
# Add a third data column called "mussel" that is the first three characters of the file name.
# Folder with the control .txt traces, relative to the project root
control_dir <- here::here("02_data", "tensometer_output", "control")

# Create an empty list to hold the data
nested_list <- list()

# Get all tensometer traces in the directory, with or without a .txt extension
txt_files <- list_trace_files(control_dir)

# Loop through each txt file, extract the data, and add it to the nested list
for (file_name in txt_files) {
    # Read in the data from the txt file
    data <- read.delim(file.path(control_dir, file_name),
                       header = FALSE,
                       sep = "\t", # how is the data separated?
                       dec = ".")
    
    # transpose
    data_transpose = t(data) 
    
    # Snip data, Fix headers
    data_snipped <- as.data.frame(data_transpose[,2:4]) # keep all rows, grab only column 2 through 4
    colnames(data_snipped) <- c('time', 'displacement', 'force')
    
    # Fix column data types
    data_snipped$time         <- as.numeric(data_snipped$time)
    data_snipped$displacement <- as.numeric(data_snipped$displacement)
    data_snipped$force        <- as.numeric(data_snipped$force)
    
    # Fix NA
    data_snipped$time[is.na(data_snipped$time)] <- 0
    data_snipped$displacement[is.na(data_snipped$displacement)] <- 0
    data_snipped$force[is.na(data_snipped$force)] <- 0
  
    # Extract mussel ID, thread number, and technician note from the file name
    ids <- parse_trace_name(file_name)
    data_snipped$mussel <- ids$mussel # Saving mussel ID
    data_snipped$thread <- ids$thread # Saving thread number
    data_snipped$note   <- ids$note   # Saving technician note (try2, fluke, ...; "" if none)
  
    # Add the data to the nested list using the file name as the list name
    nested_list[[file_name]] <- data_snipped
}

```


```{r}
# Loop through the list object "nested_list"
# calculate the maximum value from the "force" column
# Save the maximum into a dataframe, along with the first values in the "mussel" column and "thread" column

# create a data frame with three columns: mussel, thread, and max_force
max_force_control <- data.frame(mussel = character(),
                        thread = character(),
                        note = character(),
                        max_force = numeric())

# loop through the nested list
for(i in seq_along(nested_list)){
  # create a data frame for each list in the nested list
  current_df <- nested_list[[i]]
  # add a row to the max_force data frame with the mussel, thread, note, and max_force
  max_force_control <- rbind(max_force_control, data.frame(mussel = current_df$mussel[1],
                                            thread = current_df$thread[1],
                                            note = current_df$note[1],
                                            max_force = max(current_df$force)))

}

```


```{r}

require(stats)

# Loop through the list object "nested_list"
# extract the "time" and "force" column
# remove rows that contain zeros before first non-zero row. Apply to time column as well.
# remove rows that contain zeros after last non-zero row. Apply to time column as well.
# fit loess function to "time" and "force" after trimmed. Time is x, force is y.
# calculate area under the curve and save in dataframe, along with the first values in the "mussel" column and "thread" column from "nested_list"
# use a for loop to iterate through the nested list object "nested_list"

# Output folder for control QC plots (created if needed)
qc_dir <- here::here("03_analyses", "extract-tensometer-data", "QC_plots", "control")
dir.create(qc_dir, recursive = TRUE, showWarnings = FALSE)

# create a data frame with three columns: mussel, thread, and max_force
integral_control <- data.frame(mussel = character(),
                        thread = character(),
                        note = character(),
                        auc = numeric())

for(i in seq_along(nested_list)){
  # extract time and force column from the current nested list element using $ notation
  current_df <- nested_list[[i]][c("time", "force")]
  
  # fit loess function to "time" and "force" using loess() and save as an object
  current_loess <- loess(force ~ time, data = current_df)
  
  # plot the loess function with the corresponding time and force data points.
  # Name the QC plot after the source file (extension stripped) so each run, including retries and
  # flukes that share a mussel/thread, gets its own plot instead of overwriting another's.
  qc_name <- sub("\\.txt$", "", names(nested_list)[i])
  jpeg(file.path(qc_dir, paste0(qc_name, ".jpg")))
  plot(current_df$time, current_df$force, main= "Force over Time Plot with Loess Function", 
       xlab = "Time", ylab = "Force", col = "blue")
  lines(current_loess, col = "red") # add the loess function to the plot in red
  dev.off()
  
  # calculate area under the curve using approx() to approximate the integral and save to a variable
  auc <- sum(diff(current_loess$x) * (approx(current_loess$x, current_loess$y, n = length(current_loess$x)))$y[-1])
  
  # save the first values in the "mussel" and "thread" columns to variables
  mussel_first_val <- nested_list[[i]]$mussel[1]
  thread_first_val <- nested_list[[i]]$thread[1]
  
  # create a dataframe to store the results, and save the variables to the corresponding columns in the dataframe
  integral_control <- rbind(integral_control, data.frame(mussel = nested_list[[i]]$mussel[1],
                                           thread = nested_list[[i]]$thread[1],
                                           note = nested_list[[i]]$note[1],
                                           auc = auc))
}


```

```{r}
library(ggplot2)
hist(max_force_control$max_force)
max_force_control$ID <- paste(max_force_control$mussel, max_force_control$thread)
  ggplot(data=max_force_control, aes(x=1, y=max_force, label = ID))+
  geom_boxplot(width=.5)+
  # jittered text with geom_text
  geom_text(check_overlap = TRUE,
            position=position_jitter(width=0.15))+
  theme(legend.position="none")
```

```{r}
library(EnvStats)
test <- rosnerTest(max_force_control$max_force,
  k = 5
)
test
```

```{r}
sum_dir <- here::here("03_analyses", "extract-tensometer-data", "summarized_data")
dir.create(sum_dir, recursive = TRUE, showWarnings = FALSE)
write.xlsx(max_force_control, file = file.path(sum_dir, "max_force_control.xlsx"), colNames = TRUE)
write.xlsx(integral_control,  file = file.path(sum_dir, "integral_control.xlsx"),  colNames = TRUE)
print("done")
```




## TREATMENT

```{r warning=FALSE}

# extract data from all .txt files in a folder located at 02_data/tensometer_output/control/ and add it to a nested list object
# the data has three column headers: Time, displacement, force. 
# Add a third data column called "mussel" that is the first three characters of the file name.

# Folder with the treatment .txt traces, relative to the project root
treatment_dir <- here::here("02_data", "tensometer_output", "treatment")

# Create an empty list to hold the data
nested_list <- list()

# Get all tensometer traces in the directory, with or without a .txt extension
txt_files <- list_trace_files(treatment_dir)

# Loop through each txt file, extract the data, and add it to the nested list
for (file_name in txt_files) {
    # Read in the data from the txt file
    data <- read.delim(file.path(treatment_dir, file_name),
                       header = FALSE,
                       sep = "\t", # how is the data separated?
                       dec = ".")
    
    # transpose
    data_transpose = t(data) 
    
    # Snip data, Fix headers
    data_snipped <- as.data.frame(data_transpose[,2:4]) # keep all rows, grab only column 2 through 4
    colnames(data_snipped) <- c('time', 'displacement', 'force')
    
    # Fix column data types
    data_snipped$time         <- as.numeric(data_snipped$time)
    data_snipped$displacement <- as.numeric(data_snipped$displacement)
    data_snipped$force        <- as.numeric(data_snipped$force)
    
    # Fix NA
    data_snipped$time[is.na(data_snipped$time)] <- 0
    data_snipped$displacement[is.na(data_snipped$displacement)] <- 0
    data_snipped$force[is.na(data_snipped$force)] <- 0
  
    # Extract mussel ID, thread number, and technician note from the file name
    ids <- parse_trace_name(file_name)
    data_snipped$mussel <- ids$mussel # Saving mussel ID
    data_snipped$thread <- ids$thread # Saving thread number
    data_snipped$note   <- ids$note   # Saving technician note (try2, fluke, ...; "" if none)
  
    # Add the data to the nested list using the file name as the list name
    nested_list[[file_name]] <- data_snipped
}

```


```{r}
# Loop through the list object "nested_list"
# calculate the maximum value from the "force" column
# Save the maximum into a dataframe, along with the first values in the "mussel" column and "thread" column

# create a data frame with three columns: mussel, thread, and max_force
max_force_treatment <- data.frame(mussel = character(),
                        thread = character(),
                        note = character(),
                        max_force = numeric())

# loop through the nested list
for(i in seq_along(nested_list)){
  # create a data frame for each list in the nested list
  current_df <- nested_list[[i]]
  # add a row to the max_force data frame with the mussel, thread, note, and max_force
  max_force_treatment <- rbind(max_force_treatment, data.frame(mussel = current_df$mussel[1],
                                            thread = current_df$thread[1],
                                            note = current_df$note[1],
                                            max_force = max(current_df$force)))

}

```


```{r}

require(stats)

# Loop through the list object "nested_list"
# extract the "time" and "force" column
# remove rows that contain zeros before first non-zero row. Apply to time column as well.
# remove rows that contain zeros after last non-zero row. Apply to time column as well.
# fit loess function to "time" and "force" after trimmed. Time is x, force is y.
# calculate area under the curve and save in dataframe, along with the first values in the "mussel" column and "thread" column from "nested_list"
# use a for loop to iterate through the nested list object "nested_list"

# Output folder for treatment QC plots (created if needed)
qc_dir <- here::here("03_analyses", "extract-tensometer-data", "QC_plots", "treatment")
dir.create(qc_dir, recursive = TRUE, showWarnings = FALSE)

# create a data frame with three columns: mussel, thread, and max_force
integral <- data.frame(mussel = numeric(),
                        thread = numeric(),
                        auc = numeric())

# create a data frame with columns: mussel, thread, note, auc
integral_treatment <- data.frame(mussel = character(),
                        thread = character(),
                        note = character(),
                        auc = numeric())

for(i in seq_along(nested_list)){
  # extract time and force column from the current nested list element using $ notation
  current_df <- nested_list[[i]][c("time", "force")]
  
  # fit loess function to "time" and "force" using loess() and save as an object
  current_loess <- loess(force ~ time, data = current_df)
  
  # plot the loess function with the corresponding time and force data points.
  # Name the QC plot after the source file (extension stripped) so each run, including retries and
  # flukes that share a mussel/thread, gets its own plot instead of overwriting another's.
  qc_name <- sub("\\.txt$", "", names(nested_list)[i])
  jpeg(file.path(qc_dir, paste0(qc_name, ".jpg")))
  plot(current_df$time, current_df$force, main= "Force over Time Plot with Loess Function", 
       xlab = "Time", ylab = "Force", col = "blue")
  lines(current_loess, col = "red") # add the loess function to the plot in red
  dev.off()
  
  # calculate area under the curve using approx() to approximate the integral and save to a variable
  auc <- sum(diff(current_loess$x) * (approx(current_loess$x, current_loess$y, n = length(current_loess$x)))$y[-1])
  
  # save the first values in the "mussel" and "thread" columns to variables
  mussel_first_val <- nested_list[[i]]$mussel[1]
  thread_first_val <- nested_list[[i]]$thread[1]
  
  # create a dataframe to store the results, and save the variables to the corresponding columns in the dataframe
  integral_treatment <- rbind(integral_treatment, data.frame(mussel = nested_list[[i]]$mussel[1],
                                           thread = nested_list[[i]]$thread[1],
                                           note = nested_list[[i]]$note[1],
                                           auc = auc))
}


```

```{r}
library(ggplot2)
hist(max_force_treatment$max_force)
max_force_treatment$ID <- paste(max_force_treatment$mussel, max_force_treatment$thread)
  ggplot(data=max_force_treatment, aes(x=1, y=max_force, label = ID))+
  geom_boxplot(width=.5)+
  # jittered text with geom_text
  geom_text(check_overlap = TRUE,
            position=position_jitter(width=0.15))+
  theme(legend.position="none")
```

```{r}
library(EnvStats)
test <- rosnerTest(max_force_treatment$max_force,
  k = 5
)
test
```

```{r}
sum_dir <- here::here("03_analyses", "extract-tensometer-data", "summarized_data")
dir.create(sum_dir, recursive = TRUE, showWarnings = FALSE)
write.xlsx(max_force_treatment, file = file.path(sum_dir, "max_force_treatment.xlsx"), colNames = TRUE)
write.xlsx(integral_treatment,  file = file.path(sum_dir, "integral_treatment.xlsx"),  colNames = TRUE)
print("done")
```
