---
title: "Assemble thread summary"
author: "PSMFC mytilus byssus pilot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

This script assembles the initial, pre-curation thread summary by joining the tensometer extraction
outputs (script 1) with per-mussel metadata. It does not exclude any data; every extracted trace is
carried through so bad runs can be reviewed against their QC plots and removed by hand. The curated
result is what you save as `thread-summary-clean.xlsx` for script 3.

Sources and join keys:

  - max_force, integral, note         from the extraction outputs   <- the trace itself
  - species, assigned treatment (trt) from the morphometrics `key`   <- mussel tag
  - pad_area, failure, technician force from the tech GOOGLESHEET     <- mussel tag + thread + before/after
  - rna-sequenced flag                from `rna_isolation_info`       <- mussel tag

The `treatment` label follows the experiment: a control-folder trace is a before-stress (baseline)
thread, so its label is "control" regardless of the mussel's eventual assignment; a treatment-folder
trace takes the mussel's assigned treatment (OA/OW/DO).

## Load packages

```{r}
## Run within "thread-strength.Rproj" so here::here() resolves to the analysis root.
load.lib <- c("readxl", "tidyverse", "openxlsx", "here") # List of required packages
install.lib <- load.lib[!load.lib %in% installed.packages()] # Select missing packages
for (lib in install.lib) install.packages(lib, dependencies = TRUE) # Install missing packages
sapply(load.lib, require, character = TRUE) # Load all packages
```

## Paths

```{r paths}
# Within-analysis inputs and outputs are anchored to the thread-strength .Rproj.
extract_dir <- here::here("03_analyses", "extract-tensometer-data", "summarized_data")
data_dir    <- here::here("02_data")
out_dir     <- here::here("03_analyses", "assemble-thread-summary")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Cross-analysis input: per-mussel metadata lives in the morphometrics folder.
repo_root  <- normalizePath(file.path(here::here(), ".."))
morph_xlsx <- file.path(repo_root, "morphometrics", "02_data", "morphometrics.xlsx")
```

## Helper: normalize a mussel tag to a single 3-digit form

```{r}
# Tensometer tags are 2-digit in the control folder (G18, T37) and 3-digit in the treatment folder
# (G013, T037); the metadata sheets mix both. Collapse everything to prefix + zero-padded 3-digit
# (T37 -> T037) so the same physical mussel joins across sources regardless of how it was written.
norm_id <- function(x) {
  pre <- sub("^([A-Za-z]+).*$", "\\1", x)
  num <- as.integer(sub("^[A-Za-z]+0*([0-9]+).*$", "\\1", x))
  ifelse(is.na(num), NA_character_, sprintf("%s%03d", pre, num))
}
```

## 1. Combine the extraction outputs into one trace-level table

```{r}
# Each extracted trace is uniquely identified by mussel, thread, and the technician note (the note
# distinguishes a clean run from its retries/flukes). Join max_force to integral on all three.
read_group <- function(group) {
  mf <- read_excel(file.path(extract_dir, paste0("max_force_", group, ".xlsx")))
  ig <- read_excel(file.path(extract_dir, paste0("integral_",  group, ".xlsx")))
  mf <- mf %>% transmute(mussel,
                         thread = as.integer(thread),
                         note   = replace_na(as.character(note), ""),
                         max_force)
  ig <- ig %>% transmute(mussel,
                         thread = as.integer(thread),
                         note   = replace_na(as.character(note), ""),
                         integral = auc)
  full_join(mf, ig, by = c("mussel", "thread", "note")) %>% mutate(group = group)
}

traces <- bind_rows(read_group("control"), read_group("treatment")) %>%
  mutate(mkey = norm_id(mussel))

message("Combined traces: ", nrow(traces),
        " | unique mussels: ", n_distinct(traces$mkey))
```

## 2. Join species and assigned treatment from the morphometrics key

```{r}
key <- read_excel(morph_xlsx, sheet = "key") %>%
  transmute(mkey = norm_id(ID), species, trt) %>%
  distinct(mkey, .keep_all = TRUE)

traces <- traces %>%
  left_join(key, by = "mkey") %>%
  mutate(treatment = if_else(group == "control", "control", trt))

unmatched <- traces %>% filter(is.na(species))
if (nrow(unmatched) > 0) {
  warning(nrow(unmatched), " trace(s) have no morphometrics key match: ",
          paste(unique(unmatched$mussel), collapse = ", "))
}
```

## 3. Join pad area, failure mode, and technician-recorded force from the tech sheet

```{r}
# The tech sheet's own `treatment` column is already "control"/"treatment", matching our group label.
tech <- read_excel(file.path(data_dir, "GOOGLESHEET-PSMFC-mytilus-byssus-pilot-threads.xlsx")) %>%
  transmute(mkey   = norm_id(mussel_ID),
            thread = as.integer(thread_number),
            group  = treatment,
            pad_area       = `pad_cross-sectional_area_mm^2`,
            failure        = failure_mode,
            tech_max_force = maximum_force,
            tech_notes     = notes)

# A few mussel/thread/group keys have retest duplicates (e.g. T029). Keep the first of each for the
# initial join and report the rest so they can be checked during curation.
tech_dups <- tech %>% group_by(mkey, thread, group) %>% filter(n() > 1) %>% ungroup()
if (nrow(tech_dups) > 0) {
  message("Tech sheet retest duplicates on ",
          n_distinct(tech_dups %>% select(mkey, thread, group)),
          " mussel/thread/group key(s); using the first of each.")
}
tech1 <- tech %>% distinct(mkey, thread, group, .keep_all = TRUE)

traces <- traces %>% left_join(tech1, by = c("mkey", "thread", "group"))
```

## 4. Derive adhesion, the technician-vs-computed comparison, and the RNA flag

```{r}
seq_mussels <- read_excel(morph_xlsx, sheet = "rna_isolation_info") %>%
  pull(mussel) %>% norm_id() %>% unique()

summary_tbl <- traces %>%
  mutate(
    adhesion_kpa  = max_force / pad_area * 1000,
    force_diff    = max_force - tech_max_force,                 # computed peak minus technician value
    rna_sequenced = if_else(mkey %in% seq_mussels, "yes", "no"),
    sort_ID       = paste0(mussel, "_", thread)
  ) %>%
  arrange(species, mkey, thread, group) %>%
  transmute(sort_ID, mussel, thread, note, group, species,
            treatment, mussel_trt = trt,
            max_force, integral, pad_area, adhesion_kpa,
            failure, tech_max_force, force_diff, tech_notes, rna_sequenced)
```

## 5. Split by species and write the initial summaries

```{r}
trossulus <- summary_tbl %>% filter(species == "tross")
gallo     <- summary_tbl %>% filter(species == "gallo")

write.xlsx(list(data = trossulus),
           file = file.path(out_dir, "thread-summary-trossulus.xlsx"), colNames = TRUE)
write.xlsx(list(data = gallo),
           file = file.path(out_dir, "thread-summary-gallo.xlsx"), colNames = TRUE)

message("Wrote trossulus (", nrow(trossulus), " rows) and gallo (", nrow(gallo), " rows).")
```

These are the initial, everything-included summaries. Review them with the QC plots from script 1,
using the `note`, `failure`, and `force_diff` columns to spot bad runs, remove what should be
excluded, and save the curated trossulus table as `02_data/thread-summary-clean.xlsx` (sheet "data")
for script 3.

## 6. QC: technician-recorded vs computed peak force

```{r}
compare <- summary_tbl %>% filter(!is.na(tech_max_force), !is.na(max_force))
message("Force comparison available for ", nrow(compare), " of ", nrow(summary_tbl), " traces.")
message("  median |computed - technician|: ", round(median(abs(compare$force_diff)), 4), " N")

flagged <- compare %>% filter(abs(force_diff) > 0.05) %>% arrange(desc(abs(force_diff)))
message("  traces differing by more than 0.05 N: ", nrow(flagged))

p <- ggplot(compare, aes(tech_max_force, max_force, colour = species)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.6) +
  labs(x = "Technician-recorded maximum force (N)",
       y = "Computed peak force (N)",
       title = "Computed vs technician-recorded thread force") +
  theme_bw()
ggsave(file.path(out_dir, "force_computed_vs_technician.jpg"), p,
       width = 6, height = 5, dpi = 150)
```

## 7. Coverage diagnostics

```{r}
summary_tbl %>%
  summarise(
    traces             = n(),
    trossulus          = sum(species == "tross", na.rm = TRUE),
    gallo              = sum(species == "gallo", na.rm = TRUE),
    no_species         = sum(is.na(species)),
    before_threads     = sum(group == "control"),
    after_threads      = sum(group == "treatment"),
    retry_traces       = sum(note != ""),
    missing_pad_area   = sum(is.na(pad_area)),
    missing_tech_force = sum(is.na(tech_max_force))
  ) %>%
  knitr::kable()
```
