---
title: "Assemble thread summary"
author: "PSMFC mytilus byssus pilot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

## What this script does

Takes the trace-level output of `1_extract_tensometer_data.Rmd` and produces a
**curation-ready candidate** in the column shape of `03_analyses/thread-summary.xlsx`, so the
manual curation step is a review rather than a retype.

It carries forward the two columns that cannot be computed from a trace, `pad_area` and
`failure`, from whatever rows already exist in `thread-summary.xlsx`, and reports exactly
which traces still need a plaque measurement.

```
1_extract_tensometer_data.Rmd  ->  03_analyses/thread-summary-raw-output.xlsx   (375 traces)
2_assemble_thread_summary.Rmd  ->  03_analyses/assemble-thread-summary/
                                     thread-summary-candidate.xlsx
   (manual curation: add pad_area / failure, drop bad runs)
                               ->  03_analyses/thread-summary.xlsx
3_analyze_thread_strength.Rmd  ->  03_analyses/analyze-thread-strength/
```

### What changed, and why

The previous version joined four extraction files
(`max_force_{control,treatment}.xlsx`, `integral_{control,treatment}.xlsx`) plus
`morphometrics/02_data/morphometrics.xlsx` and
`02_data/GOOGLESHEET-PSMFC-mytilus-byssus-pilot-threads.xlsx`. None of those paths exist any
more. The mussel-level metadata now arrives already joined by script 1 from
`02_data/mussel-treatment-key.csv`, and there is a single extraction output instead of four.

`group` is gone. It was fully recoverable from `thread_trt`, and its value `control` meant
"day 0 or 1" while `mussel_trt == "control"` means the control tank arm, which is a different
thing. `phase` (`lab` / `pre` / `post`) replaces it and does not collide.

**Nothing is excluded here.** Every extracted trace is carried through so bad runs can be
reviewed against their QC plots and removed by hand.

## Load packages

```{r}
## Run within "thread-strength.Rproj" so here::here() resolves to the analysis root.
load.lib    <- c("readxl", "tidyverse", "openxlsx", "here")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for (lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)
```

## Paths

```{r paths}
raw_xlsx     <- here::here("03_analyses", "thread-summary-raw-output.xlsx")
curated_xlsx <- here::here("03_analyses", "thread-summary.xlsx")   # the hand-curated file
out_dir      <- here::here("03_analyses", "assemble-thread-summary")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(raw_xlsx))
  stop("Not found:\n  ", raw_xlsx, "\nRun 1_extract_tensometer_data.Rmd first.")
```

## Helper: canonical mussel tag

```{r}
# Collapse any spelling of a tag to prefix + zero-padded 3 digits (T37 -> T037) so the same
# physical animal joins across sources regardless of how it was written.
norm_id <- function(x) {
  x   <- trimws(as.character(x))
  pre <- sub("^([A-Za-z]+).*$", "\\1", x)
  num <- suppressWarnings(as.integer(sub("^[A-Za-z]+0*([0-9]+).*$", "\\1", x)))
  ifelse(is.na(num) | pre == x, NA_character_, sprintf("%s%03d", pre, num))
}
```

## 1. Read the extraction output

```{r}
traces <- read_excel(raw_xlsx, sheet = "data") %>%
  mutate(mussel = norm_id(mussel),
         thread = as.integer(thread),
         note   = replace_na(as.character(note), ""))

message("Traces: ", nrow(traces), " | unique mussels: ", n_distinct(traces$mussel))

if (!all(c("species", "mussel_trt", "rna_sequenced") %in% names(traces)))
  warning("The extraction output has no mussel-level columns. Run 0_build_mussel_key.Rmd ",
          "and re-run script 1, or those columns will be blank in the candidate file.")

traces %>% count(source_folder, thread_trt, phase, day) %>% knitr::kable()
```

## 2. Carry forward plaque area and failure mode

`pad_area` (plaque cross-sectional area, mm^2) and `failure` are measured from the microscope
images in `02_data/pictures/`; they cannot be derived from a force trace. Any row already
present in `thread-summary.xlsx` is reused so those measurements are not lost or re-entered.

The join key is `mussel` + `thread` + `thread_trt`. That triple is unique on both sides, and
it has to include `thread_trt`: 45 animals were pulled before and after exposure, so
`mussel` + `thread` alone matches two different traces.

```{r}
have_curated <- file.exists(curated_xlsx)

if (have_curated) {
  curated <- read_excel(curated_xlsx, sheet = "data") %>%
    transmute(mussel        = norm_id(mussel_ID),
              thread        = as.integer(thread_num),
              thread_trt    = as.character(thread_trt),
              pad_area      = as.numeric(pad_area),
              failure       = as.character(failure),
              prior_max_force = as.numeric(max_force))

  dup <- curated %>% count(mussel, thread, thread_trt) %>% filter(n > 1)
  if (nrow(dup) > 0)
    warning(nrow(dup), " duplicated key(s) in ", basename(curated_xlsx),
            "; the first of each is used.")
  curated <- curated %>% distinct(mussel, thread, thread_trt, .keep_all = TRUE)

  traces <- traces %>% left_join(curated, by = c("mussel", "thread", "thread_trt"))

  orphan <- curated %>%
    anti_join(traces, by = c("mussel", "thread", "thread_trt"))
  if (nrow(orphan) > 0)
    warning(nrow(orphan), " row(s) in ", basename(curated_xlsx),
            " have no matching trace: ",
            paste(paste0(orphan$mussel, "_", orphan$thread, " (", orphan$thread_trt, ")"),
                  collapse = ", "))
} else {
  warning("No ", basename(curated_xlsx), " found; pad_area and failure will be blank.")
  traces <- traces %>% mutate(pad_area = NA_real_, failure = NA_character_,
                              prior_max_force = NA_real_)
}
```

### Does the freshly extracted peak force match what the curated file recorded?

A disagreement here is not a rounding artifact. It means the curated value came from a
different file, or was edited by hand, and it propagated into `adhesion_kpa`. Check any row
listed below against its QC plot before trusting the curated number.

```{r}
force_check <- traces %>%
  filter(!is.na(prior_max_force)) %>%
  mutate(diff = max_force - prior_max_force) %>%
  filter(abs(diff) > 1e-9) %>%
  arrange(desc(abs(diff))) %>%
  select(source_folder, file, mussel, thread, thread_trt,
         max_force_new = max_force, max_force_curated = prior_max_force, diff)

message("Curated rows compared: ", sum(!is.na(traces$prior_max_force)),
        " | disagreeing: ", nrow(force_check))
if (nrow(force_check) > 0) knitr::kable(force_check)
```

## 3. Derive adhesion

```{r}
# If script 1 ran without the mussel key, these columns are absent. Create them empty rather
# than failing, so the candidate file still builds and the gap is visible as blank cells.
for (col in c("species", "mussel_trt", "rna_sequenced"))
  if (!col %in% names(traces)) traces[[col]] <- NA_character_

summary_tbl <- traces %>%
  mutate(
    adhesion_kpa = max_force / pad_area * 1000,       # N / mm^2 -> kPa
    sort_ID      = paste0(mussel, "_", thread)
  ) %>%
  arrange(match(source_folder, unique(traces$source_folder)), mussel, thread) %>%
  transmute(species, sort_ID,
            mussel_ID  = mussel,
            thread_num = thread,
            note, phase, day,
            mussel_trt,                                # the arm the ANIMAL was assigned to
            thread_trt,                                # what the THREAD was built in
            max_force, integral, max_displacement,
            pad_area, adhesion_kpa, failure,
            rna_sequenced,
            source_folder, file)                       # provenance, drop before pasting
```

## 4. What still needs measuring

```{r}
summary_tbl %>%
  group_by(source_folder, thread_trt) %>%
  summarise(traces          = n(),
            have_pad_area   = sum(!is.na(pad_area)),
            need_pad_area   = sum(is.na(pad_area)),
            have_failure    = sum(!is.na(failure)),
            .groups = "drop") %>%
  knitr::kable()

message("Total traces needing a plaque measurement: ", sum(is.na(summary_tbl$pad_area)))
```

## 5. Pairing, by arm

Adhesion contrasts rest on animals measured at both `pre` and `post`. An unbalanced arm is
what makes a per-arm marginal mean from a random-intercept model unreliable, so look here
before fitting one.

```{r}
summary_tbl %>%
  filter(phase %in% c("pre", "post")) %>%
  distinct(mussel_ID, mussel_trt, phase) %>%
  count(mussel_ID, mussel_trt, name = "n_phases") %>%
  count(mussel_trt, paired = n_phases == 2L) %>%
  pivot_wider(names_from = paired, values_from = n,
              names_prefix = "paired_", values_fill = 0) %>%
  knitr::kable()
```

## 6. Write the candidate

Columns match `thread-summary.xlsx` apart from the retired `group` (replaced by `phase`) and
three additions: `note`, `max_displacement`, and the `source_folder` / `file` provenance pair.
Drop the last two before pasting if you want an exact column match.

```{r}
write.xlsx(list(data = summary_tbl),
           file = file.path(out_dir, "thread-summary-candidate.xlsx"),
           colNames = TRUE, overwrite = TRUE)

message("Wrote ", file.path(out_dir, "thread-summary-candidate.xlsx"),
        " (", nrow(summary_tbl), " rows).")
```

Review this against the QC plots in
`03_analyses/extract-tensometer-data/QC_plots/<source_folder>/<file>.jpg`, using `note`,
`failure` and the force-disagreement table in section 2 to spot bad runs. Remove what should
be excluded and save the result as `03_analyses/thread-summary.xlsx` (sheet `data`) for
script 3.
