---
title: "Assemble thread summary"
author: "PSMFC mytilus byssus pilot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

## What this script does

Joins the trace-level output of `1_extract_tensometer_data.Rmd` to the plaque measurements in
`02_data/pad_area_measurements.xlsx`, computes adhesion, and writes a **curation-ready
candidate** so the manual step is a review rather than a retype.

It also writes two worklists: which traces still need a plaque measured (and whether a
microscope image for them exists in the repository), and which animals that ought to pair
before/after still do not.

```
1_extract_tensometer_data.Rmd  ->  03_analyses/extract-tensometer-data/thread-summary-raw-output.xlsx
              +  02_data/pad_area_measurements.xlsx                            (hand-measured)
2_assemble_thread_summary.Rmd  ->  03_analyses/assemble-thread-summary/
                                     thread-summary-candidate.xlsx
                                     pad-area-worklist.xlsx
   (manual: drop bad runs against the QC plots)
                               ->  03_analyses/thread-summary.xlsx
3_analyze_thread_strength.Rmd  ->  03_analyses/analyze-thread-strength/
```

### Where each fact comes from

One fact, one source. Nothing is copied into two files where the copies could drift.

| fact | grain | source |
|---|---|---|
| `max_force`, `integral`, `max_displacement` | thread | the trace itself, via script 1 |
| `thread_trt`, `phase`, `day` | thread | the tensometer subfolder, via script 1 |
| `species`, `mussel_trt`, `rna_sequenced` | mussel | `02_data/mussel-treatment-key.csv`, via script 1 |
| `pad_area`, `failure` | thread | `02_data/pad_area_measurements.xlsx` |

`pad_area_measurements.xlsx` therefore carries only what it is the source of truth for:
`mussel`, `thread`, `thread_trt`, `pad_area`, `failure`, `notes`. Its previous `species`,
`group`, `mussel_trt` and `day` columns were second copies of facts owned elsewhere and have
been removed. They were verified to agree before removal.

**Nothing is excluded here.** Every extracted trace is carried through so bad runs can be
reviewed against their QC plots and removed deliberately.

## Load packages

```{r}
## Run within "thread-strength.Rproj" so here::here() resolves to the analysis root.
load.lib    <- c("readxl", "tidyverse", "openxlsx", "here")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for (lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)
```

## Paths

```{r paths}
raw_xlsx <- here::here("03_analyses", "extract-tensometer-data", "thread-summary-raw-output.xlsx")
pad_xlsx <- here::here("02_data", "pad_area_measurements.xlsx")
pic_root <- here::here("02_data", "pictures")
out_dir  <- here::here("03_analyses", "assemble-thread-summary")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(raw_xlsx))
  stop("Not found:\n  ", raw_xlsx, "\nRun 1_extract_tensometer_data.Rmd first.")
if (!file.exists(pad_xlsx))
  stop("Not found:\n  ", pad_xlsx)
```

## Helper: canonical mussel tag

```{r}
# Collapse any spelling of a tag to prefix + zero-padded 3 digits (T37 -> T037) so the same
# physical animal joins across sources regardless of how it was written.
norm_id <- function(x) {
  x   <- trimws(as.character(x))
  pre <- sub("^([A-Za-z]+).*$", "\\1", x)
  num <- suppressWarnings(as.integer(sub("^[A-Za-z]+0*([0-9]+).*$", "\\1", x)))
  ifelse(is.na(num) | pre == x, NA_character_, sprintf("%s%03d", pre, num))
}
```

## 1. Read the extraction output

```{r}
traces <- read_excel(raw_xlsx, sheet = "data") %>%
  mutate(mussel = norm_id(mussel),
         thread = as.integer(thread),
         note   = replace_na(as.character(note), ""))

folder_order <- unique(traces$source_folder)

message("Traces: ", nrow(traces), " | unique mussels: ", n_distinct(traces$mussel))

if (!all(c("species", "mussel_trt", "rna_sequenced") %in% names(traces)))
  warning("The extraction output has no mussel-level columns. Run 0_build_mussel_key.Rmd ",
          "and re-run script 1, or those columns will be blank in the candidate file.")

print(knitr::kable(traces %>% count(source_folder, thread_trt, phase, day)))
```

## 2. Join the plaque measurements

`pad_area` (plaque cross-sectional area, mm^2) and `failure` are measured by hand from the
microscope images in `02_data/pictures/`; they cannot be derived from a force trace.

The join key is `mussel` + `thread` + `thread_trt`. It has to include `thread_trt`: 45
animals were pulled before and after exposure, so `mussel` + `thread` alone matches two
different traces and would silently duplicate rows.

```{r}
pad <- read_excel(pad_xlsx, sheet = "data")

# Accept the old column spellings so an older copy of the file still joins.
ren <- c(mussel_ID = "mussel", thread_num = "thread")
for (from in names(ren)) {
  to <- ren[[from]]
  if (from %in% names(pad) && !to %in% names(pad))
    names(pad)[names(pad) == from] <- to
}

need_cols <- c("mussel", "thread", "thread_trt", "pad_area")
if (length(setdiff(need_cols, names(pad))) > 0)
  stop(basename(pad_xlsx), " is missing column(s): ",
       paste(setdiff(need_cols, names(pad)), collapse = ", "),
       "\nColumns present: ", paste(names(pad), collapse = ", "))

# Optional columns: create them empty if the file does not carry them, so the transmute
# below never has to branch on the schema.
if (!"failure" %in% names(pad)) pad$failure <- NA_character_
if (!"notes"   %in% names(pad)) pad$notes   <- NA_character_

pad <- pad %>%
  transmute(mussel     = norm_id(mussel),
            thread     = as.integer(thread),
            thread_trt = as.character(thread_trt),
            pad_area   = suppressWarnings(as.numeric(pad_area)),
            failure    = as.character(failure),
            pad_notes  = as.character(notes))

dup <- pad %>% count(mussel, thread, thread_trt) %>% filter(n > 1)
if (nrow(dup) > 0) {
  warning(nrow(dup), " duplicated key(s) in ", basename(pad_xlsx),
          "; the first of each is used. Fix the source file.")
  knitr::kable(dup) %>% print()
  pad <- pad %>% distinct(mussel, thread, thread_trt, .keep_all = TRUE)
}

# Rows in the measurement file that match no trace: a typo, or a trace that was deleted.
orphan <- pad %>% anti_join(traces, by = c("mussel", "thread", "thread_trt"))
if (nrow(orphan) > 0) {
  warning(nrow(orphan), " row(s) in ", basename(pad_xlsx), " match no trace.")
  knitr::kable(orphan %>% select(mussel, thread, thread_trt, pad_area)) %>% print()
}

# Traces with no row at all in the measurement file (as opposed to a row with a blank value).
missing_rows <- traces %>% anti_join(pad, by = c("mussel", "thread", "thread_trt"))
if (nrow(missing_rows) > 0)
  message(nrow(missing_rows), " trace(s) have no row in ", basename(pad_xlsx),
          "; they appear in the worklist below and can be pasted in.")

traces <- traces %>% left_join(pad, by = c("mussel", "thread", "thread_trt"))

message("Plaque measurements joined: ", sum(!is.na(traces$pad_area)), " of ", nrow(traces))
```

## 3. Which microscope image belongs to each trace

The picture folders are split `control` / `treatment`, matching the pre/post split, and a
single image sometimes covers several threads (`T020_01-02-03.png`). Both are handled here so
the worklist can say whether a trace is measurable from what is in the repository.

```{r}
scan_pictures <- function(root) {
  sides <- c("control", "treatment")
  rows  <- list()
  for (side in sides) {
    d <- file.path(root, side)
    if (!dir.exists(d)) next
    for (f in list.files(d)) {
      m <- regmatches(f, regexec("^([A-Za-z]+[0-9]+)_([0-9-]+)\\.", f))[[1]]
      if (length(m) == 0) next
      threads <- suppressWarnings(as.integer(strsplit(m[3], "-", fixed = TRUE)[[1]]))
      threads <- threads[!is.na(threads)]
      if (length(threads) == 0) next
      rows[[length(rows) + 1L]] <- tibble(mussel   = norm_id(m[2]),
                                          thread   = threads,
                                          pic_side = side,
                                          picture  = file.path("02_data", "pictures", side, f))
    }
  }
  if (length(rows) == 0) return(tibble(mussel = character(), thread = integer(),
                                       pic_side = character(), picture = character()))
  bind_rows(rows) %>%
    group_by(mussel, thread, pic_side) %>%
    summarise(picture = paste(unique(picture), collapse = "; "), .groups = "drop")
}

pictures <- scan_pictures(pic_root)
message("Indexed ", nrow(pictures), " mussel/thread/side picture entries.")

# Pre-exposure threads are photographed into pictures/control/, day-3 threads into
# pictures/treatment/.
traces <- traces %>%
  mutate(pic_side = if_else(thread_trt %in% c("baseline", "lab_control"),
                            "control", "treatment")) %>%
  left_join(pictures, by = c("mussel", "thread", "pic_side")) %>%
  mutate(picture_status = if_else(is.na(picture), "missing", "available"))
```

## 4. Derive adhesion and assemble the candidate

```{r}
# If script 1 ran without the mussel key, these columns are absent. Create them empty rather
# than failing, so the candidate still builds and the gap is visible as blank cells.
for (col in c("species", "mussel_trt", "rna_sequenced"))
  if (!col %in% names(traces)) traces[[col]] <- NA_character_

summary_tbl <- traces %>%
  mutate(adhesion_kpa = max_force / pad_area * 1000,       # N / mm^2 -> kPa
         sort_ID      = paste0(mussel, "_", thread)) %>%
  arrange(match(source_folder, folder_order), mussel, thread) %>%
  transmute(species, sort_ID,
            mussel_ID  = mussel,
            thread_num = thread,
            note, phase, day,
            mussel_trt,                                # the arm the ANIMAL was assigned to
            thread_trt,                                # what the THREAD was built in
            max_force, integral, max_displacement,
            pad_area, adhesion_kpa, failure,
            rna_sequenced,
            pad_notes, picture_status,
            source_folder, file)                       # provenance, drop before pasting

message("Candidate rows: ", nrow(summary_tbl),
        " | with adhesion_kpa: ", sum(!is.na(summary_tbl$adhesion_kpa)))
```

## 5. Worklist: what still needs a plaque measured

```{r}
worklist <- traces %>%
  filter(is.na(pad_area)) %>%
  arrange(match(source_folder, folder_order), mussel, thread) %>%
  transmute(mussel, thread, thread_trt, phase, mussel_trt,
            source_folder, file,
            qc_plot = file.path("03_analyses", "extract-tensometer-data", "QC_plots",
                                source_folder, sub("\\.txt$", ".jpg", file, ignore.case = TRUE)),
            picture_status, picture)

cat("Traces still needing a plaque measurement:", nrow(worklist), "\n")
print(knitr::kable(worklist %>%
  count(source_folder, thread_trt, picture_status) %>%
  pivot_wider(names_from = picture_status, values_from = n, values_fill = 0)))

no_pic <- worklist %>% filter(picture_status == "missing")
if (nrow(no_pic) > 0) {
  message(nrow(no_pic), " of those have NO microscope image in the repository and cannot be ",
          "measured from what is committed here.")
  no_pic %>% count(mussel, name = "traces_without_an_image") %>% knitr::kable() %>% print()
}
```

## 6. Pairing: which animals that should pair still do not

Three distinct situations, and only the first is fixable by measuring.

  - `pair_blocked_unmeasured` : the animal has traces at BOTH pre and post, but at least one
    side has no plaque measurement, so it cannot enter an adhesion pair. Fixable, provided an
    image exists.
  - `missing_pre_trace`  : day-3 threads exist, but no pre-exposure trace was ever recorded.
  - `missing_post_trace` : pre-exposure threads exist, but no day-3 trace was ever recorded.

```{r}
by_side <- traces %>%
  filter(phase %in% c("pre", "post")) %>%
  group_by(mussel, mussel_trt, phase) %>%
  summarise(traces = n(), measured = sum(!is.na(pad_area)), .groups = "drop")

wide <- by_side %>%
  pivot_wider(names_from = phase, values_from = c(traces, measured), values_fill = 0)
for (col in c("traces_pre", "traces_post", "measured_pre", "measured_post"))
  if (!col %in% names(wide)) wide[[col]] <- 0L

wide <- wide %>%
  mutate(trace_paired = traces_pre > 0 & traces_post > 0,
         pair_usable  = measured_pre > 0 & measured_post > 0,
         # Held as its own column rather than recomputed inside summarise(): within one
         # summarise(), a later expression sees the value an earlier one just created, so
         # `sum(trace_paired & !pair_usable)` there would silently use the scalar count.
         blocked_flag = trace_paired & !pair_usable,
         gap_type = case_when(
           trace_paired & !pair_usable          ~ "pair_blocked_unmeasured",
           traces_post > 0 & traces_pre  == 0   ~ "missing_pre_trace",
           traces_pre  > 0 & traces_post == 0   ~ "missing_post_trace",
           TRUE                                 ~ "ok"))

# Can the blocked ones actually be fixed? Only if the unmeasured side has an image.
pic_by_side <- traces %>%
  filter(phase %in% c("pre", "post"), is.na(pad_area)) %>%
  group_by(mussel, phase) %>%
  summarise(images = sum(picture_status == "available"), .groups = "drop") %>%
  pivot_wider(names_from = phase, values_from = images,
              names_prefix = "images_missing_side_", values_fill = 0)

pairing_gaps <- wide %>%
  select(-blocked_flag) %>%
  left_join(pic_by_side, by = "mussel") %>%
  mutate(across(starts_with("images_missing_side_"), ~ replace_na(.x, 0L))) %>%
  filter(gap_type != "ok") %>%
  arrange(gap_type, mussel_trt, mussel)

cat("\nPairing status by arm:\n")
print(knitr::kable(wide %>%
  group_by(mussel_trt) %>%
  summarise(mussels      = n(),
            trace_paired = sum(trace_paired),
            usable_pairs = sum(pair_usable),
            blocked      = sum(blocked_flag),
            .groups = "drop")))

cat("\nGaps:\n")
print(knitr::kable(pairing_gaps %>% count(gap_type, mussel_trt) %>%
  pivot_wider(names_from = mussel_trt, values_from = n, values_fill = 0)))

blocked <- pairing_gaps %>% filter(gap_type == "pair_blocked_unmeasured")
if (nrow(blocked) > 0) {
  cat("\nAnimals with traces on both sides but no usable adhesion pair:\n")
  blocked %>%
    select(mussel, mussel_trt, traces_pre, measured_pre, traces_post, measured_post,
           images_missing_side_pre, images_missing_side_post) %>%
    knitr::kable() %>% print()
}
```

## 7. Write the outputs

```{r}
write.xlsx(list(data = summary_tbl),
           file = file.path(out_dir, "thread-summary-candidate.xlsx"),
           colNames = TRUE, overwrite = TRUE)

write.xlsx(list(to_measure = worklist, pairing_gaps = pairing_gaps),
           file = file.path(out_dir, "pad-area-worklist.xlsx"),
           colNames = TRUE, overwrite = TRUE)

message("Wrote ", file.path(out_dir, "thread-summary-candidate.xlsx"),
        " (", nrow(summary_tbl), " rows).")
message("Wrote ", file.path(out_dir, "pad-area-worklist.xlsx"),
        " (", nrow(worklist), " to measure, ", nrow(pairing_gaps), " pairing gaps).")
```

Review the candidate against the QC plots in
`03_analyses/extract-tensometer-data/QC_plots/<source_folder>/<file>.jpg`, using `note`,
`failure` and `picture_status` to spot bad runs. Remove what should be excluded and save the
result as `03_analyses/thread-summary.xlsx` (sheet `data`) for script 3. Drop
`source_folder`, `file`, `pad_notes` and `picture_status` before saving if you want the
narrower column set script 3 previously used; script 3 ignores extra columns either way.
