---
title: "3_analyze_thread_strength"
author: "Matt George"
date: "2023-04-25"
output: html_document
---

<!--
PURPOSE
  Analyze byssal-thread plaque adhesion (kPa) for Mytilus trossulus before and
  after a 3-day stress exposure (ocean acidification OA, ocean warming OW,
  hypoxia DO), relative to a pre-exposure baseline and a day-3 control arm.

INPUT
  03_analyses/thread-summary.xlsx, sheet "data" - the hand-curated table.
  Produced by curating 03_analyses/assemble-thread-summary/thread-summary-candidate.xlsx
  (script 2), which in turn reads script 1's extraction output.

DATA MODEL (read this before editing the filters below)
  Three columns, three different grains. They are easy to confuse, so each is named
  for what it describes:
    - thread_trt : what the THREAD was built in.
                   lab_control | baseline | treatment_control | OA | OW | DO
    - phase      : lab  = never entered the experimental system (day 0)
                   pre  = shared holding system, before exposure (day 1)
                   post = during the exposure (day 3)
                   Derived from thread_trt if the column is absent.
    - mussel_trt : the arm the ANIMAL was assigned to (its fate).
                   lab_control | control | OA | OW | DO
                   This is a DESTINY label. For a `pre` thread it describes the
                   animal's future, not its past: a baseline thread from an OA
                   animal is NOT an OA thread.

  `phase` has THREE levels on purpose. The previous version derived a binary
  timepoint from a `group` column whose "control" value pooled the day-0
  lab-reference animals with the day-1 pre-exposure baselines. That put 11
  animals which never entered the experimental system into a 54-animal
  "baseline" reference pool. See INCLUDE_LAB_REFERENCE below.

  The retired `group` column is not used. It was fully recoverable from
  thread_trt, and its value "control" collided with both
  mussel_trt == "control" and thread_trt == "treatment_control".

FIGURE CONVENTION FOR THE LINE+BOX PANELS (one per arm)
  - Left cluster  = the baseline reference pool (see INCLUDE_LAB_REFERENCE);
    the SAME mussels appear on every panel.
  - Right cluster = only mussels that produced threads in that arm at day 3.
  - A connecting line is drawn only for mussels present on BOTH sides (true
    repeated measures); unpaired mussels appear as a single point with no line.
  - A boxplot of the per-mussel means flanks each cluster, offset so it never
    overlaps the points or crosses the connecting lines.
  - A panel is now built for the CONTROL arm too, not only OA/OW/DO. Whether it
    has connecting lines depends on whether the T126-T135 baselines have been
    curated with a plaque area; the script reports the pair count either way.
  - Far-left grey cluster = the day-0 lab-reference animals, labelled "lab (day 0)"
    (SHOW_LAB_REFERENCE_IN_PANELS). They are shown for comparison only: never in the
    baseline pool, never paired, never in a model.

THE TWO CONTROLS (see also 03_analyses/README.md)
  - treatment_control / mussel_trt == "control": day-3 animals held three days under
    ambient conditions in the same system as the stressor arms. This is the reference
    for a STRESSOR effect (mixed-model interaction, vs_ctrl_* columns in 1b, DESeq2 TC).
  - lab_control / phase == "lab": day-0 animals that never entered the experimental
    system. They measure time in system + handling (+ secretion state, on the
    expression side) and are confounded with all three; a descriptive comparison
    group only (Statistics 3a), never a baseline.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages; set ggplot theme
```{r message=FALSE, warning=FALSE}
## Clear workspace
rm(list = ls())

## Load packages (auto-install any that are missing).
## bestNormalize, agricolae, nlme, multcomp and rstatix were declared by the previous
## version but never called; they are dropped so a fresh machine does not install five
## packages this script has no use for.
load.lib <- c("readxl", "tidyverse", "lme4", "lmerTest", "emmeans",
              "broom", "broom.mixed", "colorspace", "here")
              # emmeans: lmer post-hoc contrasts; broom/broom.mixed: tidy stats output;
              # colorspace: lightened violin fills in p2
install.lib <- load.lib[!load.lib %in% installed.packages()]
for (lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)

## ggplot theme (ggplot2 >= 3.4: line/rect widths use `linewidth`, not `size`)
my_theme <- theme(line             = element_line(linewidth = 1.2),
                  rect             = element_rect(linewidth = 1.2),
                  text             = element_text(size = 14, color = "black"),
                  panel.background = element_blank(),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.text.x      = element_text(size = 16, color = "black"),
                  axis.text.y      = element_text(size = 16, color = "black"),
                  axis.title.x     = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y     = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x     = element_line(color = "black"),
                  axis.ticks.y     = element_line(color = "black"),
                  panel.border     = element_rect(color = "black", fill = NA, linewidth = 1.2),
                  legend.key       = element_blank(),
                  legend.position  = "none")

## Palette, keyed on thread_trt. `baseline` keeps the blue the old pooled "control"
## level used, so the pre-exposure cluster reads the same as before; the day-3 control
## arm gets its own lighter blue so the two are never mistaken for each other.
treatment_colors <- c("lab_control"       = "#B0BEC5",   # grey: outside the experiment
                      "baseline"          = "#5DADE2",   # blue
                      "treatment_control" = "#7FB3D5",   # lighter blue
                      "OW"                = "#fdae61",   # orange
                      "OA"                = "#4CAF50",   # green
                      "DO"                = "#9884a3")   # purple
```

```{r}
# Create this script's output folder under 03_analyses (idempotent)
out_dir <- here::here("03_analyses", "analyze-thread-strength")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

### Configuration
```{r}
# Should the day-0 lab-reference animals (phase == "lab", mussel_trt == "lab_control")
# count as part of the BASELINE REFERENCE POOL?
#
#   FALSE (default) - the pool is the day-1 pre-exposure threads only. These animals
#                     went on to be exposed, so a before/after contrast against them
#                     isolates the stressor.
#   TRUE            - reproduces the previous behaviour, where the pool was 43
#                     pre-exposure animals PLUS 11 lab-reference animals that never
#                     entered the experimental system. That confounds "effect of the
#                     stressor" with "effect of being in the system at all".
#
# This toggle affects the line+box left cluster and the two-sample tests in 1b only.
# It does NOT affect the paired tests (lab animals have no day-3 pull, so they never
# pair) and it does NOT affect the mixed model, which is always restricted to
# phase %in% c("pre", "post").
INCLUDE_LAB_REFERENCE <- FALSE
if (isTRUE(INCLUDE_LAB_REFERENCE))
  warning("INCLUDE_LAB_REFERENCE = TRUE: the day-0 lab-reference animals are being POOLED ",
          "into the baseline reference. Every two-sample p in 1b and the left cluster of ",
          "every line+box panel then confound 'stressor' with 'being in the system at all'. ",
          "This is the legacy behaviour and is not the analysis of record.", call. = FALSE)

# Show the day-0 lab-reference animals in the line+box panels as their OWN labelled group
# (grey cluster at the far left, "lab (day 0)"), never merged into the baseline cluster.
# Ignored when INCLUDE_LAB_REFERENCE is TRUE (they are then inside the baseline pool).
SHOW_LAB_REFERENCE_IN_PANELS <- TRUE
```

### Load data set; derive adhesion and timepoint
```{r}
thread_plot <- read_excel(here::here("03_analyses", "thread-summary.xlsx"),
                          sheet = "data", col_names = TRUE)

# --- Schema normalisation -------------------------------------------------------------
# Accepts the curated file under either the old or the new column names, so this script
# keeps working across the rename. `mussel`/`thread`/`thread_trt` are the names used below.
ren <- c(mussel_ID = "mussel", thread_num = "thread", treatment = "thread_trt")
for (from in names(ren)) {
  to <- ren[[from]]
  if (from %in% names(thread_plot) && !to %in% names(thread_plot))
    names(thread_plot)[names(thread_plot) == from] <- to
}

need <- c("mussel", "thread", "thread_trt", "mussel_trt", "max_force", "pad_area")
if (length(setdiff(need, names(thread_plot))) > 0)
  stop("thread-summary.xlsx is missing required column(s): ",
       paste(setdiff(need, names(thread_plot)), collapse = ", "),
       "\nColumns present: ", paste(names(thread_plot), collapse = ", "))

# `phase` is fully determined by thread_trt, so derive it if the curated file predates it.
if (!"phase" %in% names(thread_plot)) {
  thread_plot$phase <- dplyr::case_when(
    thread_plot$thread_trt == "lab_control" ~ "lab",
    thread_plot$thread_trt == "baseline"    ~ "pre",
    TRUE                                    ~ "post")
  message("No `phase` column found; derived from thread_trt.")
}

# Canonical tag, so `mussel` joins regardless of how it was written.
norm_id <- function(x) {
  x   <- trimws(as.character(x))
  pre <- sub("^([A-Za-z]+).*$", "\\1", x)
  num <- suppressWarnings(as.integer(sub("^[A-Za-z]+0*([0-9]+).*$", "\\1", x)))
  ifelse(is.na(num) | pre == x, NA_character_, sprintf("%s%03d", pre, num))
}
thread_plot$mussel <- norm_id(thread_plot$mussel)

# --- Adhesion -------------------------------------------------------------------------
# kPa = N / mm^2 * 1000. Recomputed here so the script does not depend on the
# spreadsheet's cached formula column.
thread_plot$adhesion_kpa <- thread_plot$max_force / thread_plot$pad_area * 1000

# Drop rows with no plaque measurement, LOUDLY. Previously these passed through as NA and
# were silently discarded by each model, so the n behind a contrast was never visible.
n_all  <- nrow(thread_plot)
no_pad <- thread_plot %>% filter(is.na(adhesion_kpa))
if (nrow(no_pad) > 0) {
  message(nrow(no_pad), " of ", n_all,
          " thread(s) have no pad_area and cannot yield adhesion_kpa; excluded.")
  no_pad %>% count(thread_trt, name = "threads_dropped") %>% knitr::kable() %>% print()
}
thread_plot <- thread_plot %>% filter(!is.na(adhesion_kpa))

# --- Factors and the before/after axis -------------------------------------------------
# `timepoint` is the two-level axis the models use, and it is deliberately NOT defined for
# phase == "lab": those animals were never in the experimental system, so they have no
# before/after position. They stay in the data as their own descriptive group.
thread_plot$timepoint <- factor(dplyr::case_when(thread_plot$phase == "pre"  ~ "baseline",
                                                 thread_plot$phase == "post" ~ "post",
                                                 TRUE                        ~ NA_character_),
                                levels = c("baseline", "post"))

thread_trt_levels <- c("lab_control", "baseline", "treatment_control", "OW", "OA", "DO")
thread_plot$thread_trt <- factor(thread_plot$thread_trt,
                                 levels = intersect(thread_trt_levels,
                                                    unique(thread_plot$thread_trt)))

# `mussel_trt` uses control as the reference level so the lmer interaction reads as
# "change relative to the control arm".
mussel_trt_levels <- c("control", "OA", "OW", "DO", "lab_control")
unknown_arm <- setdiff(unique(thread_plot$mussel_trt), mussel_trt_levels)
if (length(unknown_arm) > 0)
  stop("Unrecognised mussel_trt value(s): ", paste(unknown_arm, collapse = ", "),
       "\nAdd them to `mussel_trt_levels` rather than letting factor() silently NA them.")
thread_plot$mussel_trt <- factor(thread_plot$mussel_trt,
                                 levels = intersect(mussel_trt_levels,
                                                    unique(thread_plot$mussel_trt)))

thread_plot$mussel <- factor(thread_plot$mussel)
thread_plot$thread <- factor(thread_plot$thread)

# Structural check: every mussel maps to exactly one fate.
bad_fate <- thread_plot %>% group_by(mussel) %>%
  summarise(n_fate = n_distinct(mussel_trt), .groups = "drop") %>% filter(n_fate != 1)
if (nrow(bad_fate) > 0)
  stop("Mussel(s) carrying more than one mussel_trt: ",
       paste(bad_fate$mussel, collapse = ", "))

cat("Threads analysed:", nrow(thread_plot), "| mussels:", n_distinct(thread_plot$mussel), "\n")
thread_plot %>% count(phase, thread_trt, mussel_trt) %>% knitr::kable()
```

### Per-mussel summaries (the unit of replication for figures and classic tests)
```{r message=FALSE}
# Collapse threads to one adhesion summary per mussel x thread_trt. Within a phase a mussel
# carries a single thread_trt, so this yields one row per mussel per phase.
# SE is NA for singletons (1 thread); those points get no error bar.
per_mussel_means <- thread_plot %>%
  group_by(mussel, phase, timepoint, thread_trt, mussel_trt) %>%
  summarise(adh_mean = mean(adhesion_kpa),
            adh_sd   = sd(adhesion_kpa),
            n        = n(),
            adh_se   = adh_sd / sqrt(n),
            .groups  = "drop")

# THE BASELINE REFERENCE POOL. One definition, used by the line+box panels and the
# two-sample tests, so the figures and the p-values can never disagree about who is in it.
baseline_pool <- per_mussel_means %>%
  filter(phase == "pre" | (INCLUDE_LAB_REFERENCE & phase == "lab"))

cat("Baseline reference pool: ", n_distinct(baseline_pool$mussel), " mussels (",
    if (INCLUDE_LAB_REFERENCE) "pre-exposure + lab-reference" else "pre-exposure only",
    ").\n", sep = "")
baseline_pool %>% count(phase, thread_trt, name = "mussels") %>% knitr::kable()

# Fixed horizontal offset per mussel, computed ONCE over all mussels so the baseline
# cluster looks identical across the arm panels. Used only for point/line placement.
set.seed(1)
mussel_offsets <- tibble(mussel = levels(thread_plot$mussel),
                         xoff   = runif(nlevels(thread_plot$mussel), -0.08, 0.08))
```

### Distribution plots: adhesion by thread treatment
```{r warning=FALSE}
# These four panels summarize the distribution of adhesion. p1 (all threads) and
# p3 (per-mussel means) correspond to the two manuscript panels; p2 and p4 restrict to
# mussels measured at BOTH timepoints (true repeated measures).
#
# x is `thread_trt`, so each bar is one condition a thread was actually built in. The
# previous version plotted a `treatment` column whose "control" bar POOLED day-0/day-1
# baselines with day-3 control-arm threads; `baseline` and `treatment_control` are now
# separate bars. Compare against the old figures with that in mind.

# Mussels measured at both baseline and post-stress (repeated-measures subset).
thread_plot_repeated <- thread_plot %>%
  group_by(mussel) %>%
  filter(n_distinct(timepoint[!is.na(timepoint)]) == 2) %>%
  ungroup()

thread_plot_all_mean <- per_mussel_means
thread_plot_repeated_mean <- per_mussel_means %>%
  group_by(mussel) %>%
  filter(n_distinct(timepoint[!is.na(timepoint)]) == 2) %>%
  ungroup()

cat("Repeated-measures mussels (both timepoints):",
    n_distinct(thread_plot_repeated$mussel), "\n")

# p1: all threads, violin + box + jittered points
p1 <- ggplot(thread_plot, aes(x = thread_trt, y = adhesion_kpa)) +
      geom_violin(aes(fill = thread_trt), color = "black", linewidth = 1.2, bw = 12) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = NA,
                   width = 0.2, notch = FALSE, varwidth = FALSE, alpha = 0.8, coef = 0) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5, alpha = .1) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5,
                 stroke = .7, shape = 1, color = "black") +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      scale_fill_manual(values = treatment_colors) +
      xlab(NULL) + ylab(NULL) + my_theme +
      theme(axis.text.x = element_text(size = 12, angle = 25, hjust = 1))
p1

# p2: repeated-measures mussels only, all threads
p2 <- ggplot(thread_plot_repeated, aes(x = thread_trt, y = adhesion_kpa)) +
      geom_violin(aes(fill = thread_trt,
                      fill = after_scale(colorspace::lighten(fill, .5))),
                  color = "black", linewidth = 1.2, bw = 12) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = NA,
                   width = 0.2, notch = FALSE, varwidth = FALSE, alpha = 0.8, coef = 0) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5, alpha = .3) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5,
                 stroke = .7, shape = 1, color = "black") +
      scale_fill_manual(values = treatment_colors) +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      xlab(NULL) + ylab(NULL) + my_theme +
      theme(axis.text.x = element_text(size = 12, angle = 25, hjust = 1))
p2

# p3: all mussels, per-mussel means
p3 <- ggplot(thread_plot_all_mean, aes(x = thread_trt, y = adh_mean)) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = 16,
                   notch = FALSE, varwidth = TRUE, alpha = 0.8) +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      xlab(NULL) + ylab(NULL) + my_theme +
      theme(axis.text.x = element_text(size = 11, angle = 25, hjust = 1))
p3

# p4: repeated-measures mussels only, per-mussel means
p4 <- ggplot(thread_plot_repeated_mean, aes(x = thread_trt, y = adh_mean)) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = 16,
                   notch = FALSE, varwidth = TRUE, alpha = 0.8) +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      xlab(NULL) + ylab(NULL) + my_theme +
      theme(axis.text.x = element_text(size = 11, angle = 25, hjust = 1))
p4

ggsave(file.path(out_dir, "BP_AllMussels_AllThreads.png"),
       plot = p1, dpi = 1200, device = "png", width = 7, height = 5, units = "in")
ggsave(file.path(out_dir, "BP_OnlyRepeatedMussels_AllThreads.png"),
       plot = p2, dpi = 1200, device = "png", width = 8, height = 6, units = "in")
ggsave(file.path(out_dir, "BP_AllMussels_MeanThreads.png"),
       plot = p3, dpi = 1200, device = "png", width = 4.5, height = 4, units = "in")
ggsave(file.path(out_dir, "BP_OnlyRepeatedMussels_MeanThreads.png"),
       plot = p4, dpi = 1200, device = "png", width = 4.5, height = 4, units = "in")
```

### Line + box panels: baseline vs each day-3 arm
```{r warning=FALSE}
# One reusable builder, called once per day-3 arm.
#
# x-geometry (continuous axis):
#   baseline points/lines/error bars live near x = 1 (within +/- 0.08)
#   arm      points/lines/error bars live near x = 2 (within +/- 0.08)
#   baseline box sits at x = 0.60, arm box at x = 2.40 (width 0.22)
# so the boxes are always clear of the point clusters and the connecting lines.
#
# Coloring: baseline points/box in baseline blue; arm points/box in the arm color;
# connecting lines neutral grey so they read as links rather than adding new colors.
#
# The control arm gets a panel too. Whether it shows connecting lines depends on whether
# the day-3 control animals' own baseline threads have been curated with a plaque area;
# the pair count is printed for every panel so an empty pairing is never silent.

make_line_box <- function(trt_code, trt_label = trt_code,
                          y_breaks = seq(0, 200, 25), y_lim = c(0, 210)) {

  base_col <- treatment_colors[["baseline"]]
  lab_col  <- treatment_colors[["lab_control"]]
  trt_col  <- if (trt_code %in% names(treatment_colors)) treatment_colors[[trt_code]] else "grey60"
  box_w <- 0.22; box_ctrl_x <- 0.60; box_trt_x <- 2.40; box_lab_x <- -0.40

  ctrl <- baseline_pool %>% mutate(side = "control", xbase = 1, xbox = box_ctrl_x)
  trt  <- per_mussel_means %>% filter(phase == "post", thread_trt == trt_code) %>%
          mutate(side = "treatment", xbase = 2, xbox = box_trt_x)
  # Day-0 lab-reference animals as their own cluster at the far left (x = 0). They are a
  # descriptive group only: never pooled with the baseline, never paired, never modelled.
  show_lab <- isTRUE(SHOW_LAB_REFERENCE_IN_PANELS) && !isTRUE(INCLUDE_LAB_REFERENCE)
  lab  <- if (show_lab) per_mussel_means %>% filter(phase == "lab") %>%
            mutate(side = "lab", xbase = 0, xbox = box_lab_x) else per_mussel_means[0, ]
  show_lab <- show_lab && nrow(lab) > 0

  if (nrow(trt) == 0) {
    warning("No day-3 threads for arm '", trt_code, "'; panel skipped.", call. = FALSE)
    return(NULL)
  }

  dat <- bind_rows(lab, ctrl, trt) %>%
         left_join(mussel_offsets, by = "mussel") %>%
         mutate(xpos = xbase + xoff)

  # Connecting lines only for mussels present on both sides (true repeated measures).
  paired_ids <- intersect(as.character(ctrl$mussel), as.character(trt$mussel))
  lines_dat  <- dat %>% filter(side != "lab", as.character(mussel) %in% paired_ids)
  message(trt_code, ": ", nrow(trt), " day-3 mussels, ", length(paired_ids),
          " paired to a baseline", if (show_lab) paste0("; ", nrow(lab), " lab-reference animals shown") else "", ".")

  x_breaks <- if (show_lab) c(0, 1, 2) else c(1, 2)
  x_labels <- if (show_lab) c("lab\n(day 0)", "baseline", trt_label) else c("baseline", trt_label)
  x_limits <- if (show_lab) c(-0.6, 2.6) else c(0.4, 2.6)

  ggplot() +
    { if (show_lab) geom_boxplot(data = filter(dat, side == "lab"),
                                 aes(x = xbox, y = adh_mean, group = 1),
                                 width = box_w, fill = lab_col, color = "black",
                                 linewidth = 0.7, alpha = 0.55, outlier.shape = NA) } +
    { if (show_lab) geom_errorbar(data = filter(dat, side == "lab"),
                                  aes(x = xpos, ymin = adh_mean - adh_se, ymax = adh_mean + adh_se),
                                  width = 0.04, linewidth = 0.6, color = "black", na.rm = TRUE) } +
    { if (show_lab) geom_point(data = filter(dat, side == "lab"),
                               aes(x = xpos, y = adh_mean), size = 3, color = lab_col) } +
    { if (show_lab) geom_point(data = filter(dat, side == "lab"),
                               aes(x = xpos, y = adh_mean), size = 3, shape = 1, stroke = 0.6, color = "black") } +
    geom_line(data = lines_dat,
              aes(x = xpos, y = adh_mean, group = mussel),
              color = "grey55", linewidth = 0.6, alpha = 0.8) +
    geom_boxplot(data = filter(dat, side == "control"),
                 aes(x = xbox, y = adh_mean, group = 1),
                 width = box_w, fill = base_col, color = "black",
                 linewidth = 0.7, alpha = 0.55, outlier.shape = NA) +
    geom_boxplot(data = filter(dat, side == "treatment"),
                 aes(x = xbox, y = adh_mean, group = 1),
                 width = box_w, fill = trt_col, color = "black",
                 linewidth = 0.7, alpha = 0.55, outlier.shape = NA) +
    geom_errorbar(data = filter(dat, side == "control"),
                  aes(x = xpos, ymin = adh_mean - adh_se, ymax = adh_mean + adh_se),
                  width = 0.04, linewidth = 0.6, color = "black", na.rm = TRUE) +
    geom_point(data = filter(dat, side == "control"),
               aes(x = xpos, y = adh_mean), size = 3, color = base_col) +
    geom_point(data = filter(dat, side == "control"),
               aes(x = xpos, y = adh_mean), size = 3, shape = 1, stroke = 0.6, color = "black") +
    geom_errorbar(data = filter(dat, side == "treatment"),
                  aes(x = xpos, ymin = adh_mean - adh_se, ymax = adh_mean + adh_se),
                  width = 0.04, linewidth = 0.6, color = "black", na.rm = TRUE) +
    geom_point(data = filter(dat, side == "treatment"),
               aes(x = xpos, y = adh_mean), size = 3, color = trt_col) +
    geom_point(data = filter(dat, side == "treatment"),
               aes(x = xpos, y = adh_mean), size = 3, shape = 1, stroke = 0.6, color = "black") +
    scale_x_continuous(breaks = x_breaks, labels = x_labels, limits = x_limits) +
    scale_y_continuous(breaks = y_breaks, limits = y_lim) +
    xlab(NULL) + ylab(NULL) + my_theme
}

# Build a panel for every day-3 arm actually present in the data.
day3_arms <- per_mussel_means %>% filter(phase == "post") %>%
             pull(thread_trt) %>% droplevels() %>% levels()

panel_files <- c(treatment_control = "LineBox_CONTROL_BaselineVsDay3.png",
                 OA = "LineBox_OA_BaselineVsTreatment.png",
                 OW = "LineBox_OW_BaselineVsTreatment.png",
                 DO = "LineBox_DO_BaselineVsTreatment.png")

line_box_panels <- list()
for (arm in day3_arms) {
  pnl <- make_line_box(arm, trt_label = if (arm == "treatment_control") "control" else arm)
  if (is.null(pnl)) next
  line_box_panels[[arm]] <- pnl
  print(pnl)
  fn <- if (arm %in% names(panel_files)) panel_files[[arm]] else paste0("LineBox_", arm, ".png")
  ggsave(file.path(out_dir, fn), plot = pnl,
         dpi = 1200, device = "png", width = 4.5, height = 4.5, units = "in")
}
```

<!--
STATISTICS

  The previous version's tests were not valid for this design and were replaced:
    - The "all treatments" model was aov(response ~ trt) with NO error term, so it treated
      every thread as independent (pseudoreplication) while its comment claimed it accounted
      for repeated sampling.
    - The per-arm models used aov(response ~ trt + Error(ID/thread)); `thread` (1,2,3) is just
      a within-mussel index, not a meaningful nesting stratum, and the data fed in mixed all
      arms' baselines into the "control" level. aov(... Error()) also requires balanced,
      complete cells, which this partially paired design does not have.

  Replacement, in two parts:
    (1) CLASSIC tests on the per-mussel means (one value per mussel per phase), which is
        the correct unit of replication and removes pseudoreplication without a mixed model.
    (2) A MIXED-MODEL section (lmer) on the full thread-level data, which uses every thread and
        every animal (paired and unpaired) and tests the arm-by-timepoint interaction directly.

  CHANGED IN THIS VERSION
    - The baseline reference pool is governed by INCLUDE_LAB_REFERENCE and is FALSE by
      default, so the day-0 lab-reference animals no longer sit in the reference group for
      the two-sample tests. This moves welch_p and mannwhit_p relative to earlier figures.
    - The control arm is no longer hardcoded as unpairable. The previous comment asserted
      that its baseline cohort was T001-T012 and its day-3 cohort T126-T135, so no animal
      appeared in both. That was an artifact of pooling the lab-reference animals into the
      baseline: T126-T135 DO have baseline threads on disk. Pairing is now computed from the
      data, per arm, and reported. It will be zero until those baseline threads are curated
      with a plaque area.
    - The mixed model is always restricted to phase %in% c("pre", "post"). Lab-reference
      animals have no before/after position, so including them would make the
      arm-by-timepoint interaction unestimable for their level.
    - Every table that was previously print-only is now written to disk.
    - 1b now also compares each arm's day-3 animals with the DAY-3 TREATMENT CONTROL
      (vs_ctrl_welch_p, vs_ctrl_mannwhit_p, delta_vs_ctrl), the reference the mixed model
      and the expression contrasts use; welch_p / mannwhit_p keep comparing with the pooled
      pre-exposure baseline, which is a different question.
    - Statistics 3 writes the lab-reference check, the baseline balance by future arm and
      the repeatability (ICC, per-arm pre/post correlation) that the paired design relies on.

  Parametric and non-parametric versions are reported side by side rather than relying on an
  automatic transform, so the reader can see both. Sample sizes are small; results are
  descriptive of this pilot, not confirmatory.
-->

### Statistics 1a: between-arm comparison of day-3 adhesion (per-mussel means)
```{r}
# Each mussel contributes one day-3 mean, so groups are independent (no pseudoreplication).
post_means <- per_mussel_means %>%
  filter(phase == "post") %>%
  mutate(thread_trt = droplevels(thread_trt))

cat("Day-3 per-mussel means, n per arm:\n")
print(post_means %>% count(thread_trt))

if (n_distinct(post_means$thread_trt) < 2) {
  message("Fewer than two day-3 arms with data; between-arm tests skipped.")
} else {
  # Parametric: one-way ANOVA + Tukey HSD
  aov_between <- aov(adh_mean ~ thread_trt, data = post_means)
  print(summary(aov_between))
  tukey_between <- TukeyHSD(aov_between)
  print(tukey_between)

  # Non-parametric backup: Kruskal-Wallis (robust to the small, skewed samples)
  print(kruskal.test(adh_mean ~ thread_trt, data = post_means))

  write.csv(broom::tidy(aov_between),
            file.path(out_dir, "STATS_between_arm_day3_ANOVA.csv"), row.names = FALSE)
  write.csv(broom::tidy(tukey_between),
            file.path(out_dir, "STATS_between_arm_day3_TukeyHSD.csv"), row.names = FALSE)
  write.csv(broom::tidy(kruskal.test(adh_mean ~ thread_trt, data = post_means)),
            file.path(out_dir, "STATS_between_arm_day3_KruskalWallis.csv"), row.names = FALSE)
}
```

### Statistics 1b: within-arm before/after, per arm (per-mussel means)
```{r warning=FALSE}
# Three complementary questions per arm:
#   PAIRED    : did the animals measured at both timepoints change? (matches the connecting
#               lines; uses only paired mussels, so n is small)
#   2-SAMPLE  : does the arm's day-3 distribution differ from the baseline reference pool?
#               (matches the boxes; uses all pool animals and all arm animals). NOTE the
#               reference here is the pooled pre-exposure baseline of ALL arms, not the
#               day-3 control arm; it is the "before" of the before/after picture.
#   VS CONTROL: does the arm's day-3 distribution differ from the DAY-3 TREATMENT CONTROL
#               (the reference the mixed model's interaction and the DESeq2 contrasts use)?
#               Columns vs_ctrl_*; NA for the control arm itself.
#
# All are computed for EVERY day-3 arm, including the control arm. Nothing is assumed
# about which arms can be paired; n_pairs is reported and the paired columns are NA when
# there are fewer than two pairs.

baseline_means <- baseline_pool
control_post   <- per_mussel_means %>% filter(phase == "post", thread_trt == "treatment_control")

run_arm_tests <- function(trt_code) {
  post_x <- per_mussel_means %>% filter(phase == "post", thread_trt == trt_code)
  paired <- inner_join(baseline_means %>%
                         transmute(mussel = as.character(mussel), base = adh_mean),
                       post_x %>%
                         transmute(mussel = as.character(mussel), post = adh_mean),
                       by = "mussel")

  res <- tibble(arm        = as.character(trt_code),
                n_pairs    = nrow(paired),
                n_base     = nrow(baseline_means),
                n_post     = nrow(post_x),
                n_ctrl_post = nrow(control_post),
                delta_mean = if (nrow(paired) >= 1) mean(paired$post - paired$base) else NA_real_,
                # arm day-3 mean minus the day-3 control mean (the between-arm effect at day 3)
                delta_vs_ctrl = if (trt_code != "treatment_control" && nrow(control_post) >= 1)
                                  mean(post_x$adh_mean) - mean(control_post$adh_mean) else NA_real_)

  # Paired tests (need >= 2 pairs)
  if (nrow(paired) >= 2) {
    res$paired_t_p <- t.test(paired$post, paired$base, paired = TRUE)$p.value
    res$wilcoxon_p <- suppressWarnings(wilcox.test(paired$post, paired$base, paired = TRUE)$p.value)
  } else {
    res$paired_t_p <- NA_real_; res$wilcoxon_p <- NA_real_
  }

  # Two-sample tests: baseline pool vs this arm's day-3 (need >= 2 per side)
  if (nrow(post_x) >= 2 && nrow(baseline_means) >= 2) {
    res$welch_p    <- t.test(post_x$adh_mean, baseline_means$adh_mean, var.equal = FALSE)$p.value
    res$mannwhit_p <- suppressWarnings(
      wilcox.test(post_x$adh_mean, baseline_means$adh_mean, paired = FALSE)$p.value)
  } else {
    res$welch_p <- NA_real_; res$mannwhit_p <- NA_real_
  }

  # Two-sample tests: day-3 treatment control vs this arm's day-3 (need >= 2 per side)
  if (trt_code != "treatment_control" && nrow(post_x) >= 2 && nrow(control_post) >= 2) {
    res$vs_ctrl_welch_p    <- t.test(post_x$adh_mean, control_post$adh_mean, var.equal = FALSE)$p.value
    res$vs_ctrl_mannwhit_p <- suppressWarnings(
      wilcox.test(post_x$adh_mean, control_post$adh_mean, paired = FALSE)$p.value)
  } else {
    res$vs_ctrl_welch_p <- NA_real_; res$vs_ctrl_mannwhit_p <- NA_real_
  }
  res
}

within_arm <- bind_rows(lapply(day3_arms, run_arm_tests)) %>%
  mutate(baseline_pool     = if (INCLUDE_LAB_REFERENCE) "pre + lab" else "pre only",
         vs_ctrl_reference = "day-3 treatment_control (per-mussel means)")
print(as.data.frame(within_arm))

unpaired_arms <- within_arm %>% filter(n_pairs < 2) %>% pull(arm)
if (length(unpaired_arms) > 0)
  message("Arm(s) with fewer than 2 within-subject pairs, so no paired test: ",
          paste(unpaired_arms, collapse = ", "),
          ". Check whether their baseline threads simply lack a plaque measurement.")

write.csv(within_arm, file.path(out_dir, "STATS_within_arm_beforeafter.csv"), row.names = FALSE)
```

### Statistics 2: mixed model on all threads (lmer)
```{r warning=FALSE}
# Full thread-level model. Random intercept for mussel absorbs (a) multiple threads per
# mussel x timepoint and (b) the within-mussel correlation across timepoints for paired
# animals. Fixed effects: arm, timepoint, and their interaction (the interaction is the
# scientific target: does the baseline -> day-3 change differ by arm?).
#
# Restricted to phase pre/post. Lab-reference animals have no before/after position, so
# their level of the interaction is unestimable; excluding them is not optional.
#
# CAVEAT: for any arm whose animals are not paired within-subject, "timepoint" is a
# between-cohort contrast rather than a within-subject one. The clean within-subject
# evidence is the paired tests in 1b. n_pairs there tells you which arms qualify.

dat_lmer <- thread_plot %>%
  filter(phase %in% c("pre", "post"), !is.na(timepoint)) %>%
  droplevels()

cat("lmer data:", nrow(dat_lmer), "threads,", n_distinct(dat_lmer$mussel), "mussels\n")
design_cells <- dat_lmer %>% count(mussel_trt, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0)
print(design_cells)

# Design-balance check. An arm with an empty baseline cell makes the interaction rank
# deficient: lme4 drops a coefficient, lmerTest warns about missing cells, and emmeans
# returns `nonEst` for that arm's within-arm change. That is correct behaviour, but it is
# easy to mistake for a model failure, so name the cause here.
empty_cells <- design_cells %>%
  filter(if_any(any_of(c("baseline", "post")), ~ .x == 0))
if (nrow(empty_cells) > 0) {
  warning("Arm(s) with an empty design cell, so their interaction term is not estimable: ",
          paste(empty_cells$mussel_trt, collapse = ", "),
          ". Expect a rank-deficiency message and `nonEst` contrasts for them. ",
          "This is a curation gap, not a model error, if those threads exist on disk ",
          "but have no plaque measurement.", call. = FALSE)
  knitr::kable(empty_cells) %>% print()
}

mod <- lmer(adhesion_kpa ~ mussel_trt * timepoint + (1 | mussel), data = dat_lmer)

cat("Type III ANOVA (lmerTest default, Satterthwaite df):\n")
aov_mod <- anova(mod)
print(aov_mod)
cat("\nFixed-effect estimates:\n")
print(summary(mod)$coefficients)

# Residual diagnostics (adhesion is right-skewed; check before trusting the Gaussian fit).
# Saved, not just printed, so the diagnostic travels with the result.
png(file.path(out_dir, "DIAG_lmer_residuals.png"), width = 1600, height = 800, res = 150)
par(mfrow = c(1, 2))
qqnorm(resid(mod), main = "Normal Q-Q, adhesion model"); qqline(resid(mod))
plot(fitted(mod), resid(mod), xlab = "Fitted", ylab = "Residuals",
     main = "Residuals vs fitted"); abline(h = 0, lty = 2)
par(mfrow = c(1, 1)); dev.off()

qqnorm(resid(mod)); qqline(resid(mod))
plot(fitted(mod), resid(mod), xlab = "Fitted", ylab = "Residuals"); abline(h = 0, lty = 2)

# Robustness: same model on log adhesion. Compare the interaction term against the raw fit.
mod_log <- lmer(log(adhesion_kpa) ~ mussel_trt * timepoint + (1 | mussel), data = dat_lmer)
cat("\nType III ANOVA, log(adhesion) model:\n")
aov_mod_log <- anova(mod_log)
print(aov_mod_log)

# Interpretable contrasts via estimated marginal means
emm_change <- emmeans(mod, ~ timepoint | mussel_trt)   # post vs baseline, within each arm
cat("\nWithin-arm change (post - baseline):\n")
print(pairs(emm_change))

emm_post <- emmeans(mod, ~ mussel_trt, at = list(timepoint = "post"))   # arm means at day 3
cat("\nArm marginal means at day 3 (post):\n")
print(emm_post)
cat("\nArm contrasts at day 3 (post):\n")
print(pairs(emm_post, adjust = "tukey"))

# Randomization check: arms were assigned at random, so baseline adhesion should be similar.
emm_base <- emmeans(mod, ~ mussel_trt, at = list(timepoint = "baseline"))
cat("\nArm marginal means at baseline (randomization check):\n")
print(emm_base)

# --- Save everything. The previous version printed the ANOVA tables, the log-adhesion
# --- model and the marginal means to console and wrote none of them (register items 5, 6).
tidy_anova <- function(a, label) {
  as.data.frame(a) %>% tibble::rownames_to_column("term") %>% mutate(model = label)
}
write.csv(bind_rows(tidy_anova(aov_mod, "adhesion_kpa"),
                    tidy_anova(aov_mod_log, "log(adhesion_kpa)")),
          file.path(out_dir, "STATS_lmer_typeIII_ANOVA.csv"), row.names = FALSE)
write.csv(broom.mixed::tidy(mod, effects = "fixed"),
          file.path(out_dir, "STATS_lmer_fixed_effects.csv"), row.names = FALSE)
write.csv(broom.mixed::tidy(mod_log, effects = "fixed"),
          file.path(out_dir, "STATS_lmer_fixed_effects_log.csv"), row.names = FALSE)
write.csv(as.data.frame(pairs(emm_change)),
          file.path(out_dir, "STATS_lmer_within_arm_change.csv"), row.names = FALSE)
write.csv(as.data.frame(pairs(emm_post, adjust = "tukey")),
          file.path(out_dir, "STATS_lmer_arm_contrasts.csv"), row.names = FALSE)
write.csv(bind_rows(as.data.frame(emm_post)  %>% mutate(timepoint = "post"),
                    as.data.frame(emm_base)  %>% mutate(timepoint = "baseline")),
          file.path(out_dir, "STATS_lmer_marginal_means.csv"), row.names = FALSE)
```

### Statistics 3: the two controls, baseline balance, and repeatability
```{r warning=FALSE}
# Three checks that justify the design, written out as tables rather than left as
# by-products of the figures. All on PER-MUSSEL MEANS (one value per mussel per phase), for
# adhesion, peak force, plaque area and extension.
#
#   3a  LAB-REFERENCE CHECK (STATS_lab_reference_check.csv)
#       The day-0 lab-reference animals (T001-T012) never entered the experimental system;
#       the day-1 pre-exposure animals had been in the shared holding system for a day; the
#       day-3 treatment-control animals spent three days in a tank under ambient conditions.
#       If handling and time in the system did not move attachment, the three groups should
#       agree. Welch t and Mann-Whitney per metric: lab vs pre, day-3 control vs pre,
#       lab vs day-3 control.
#   3b  BASELINE BALANCE (STATS_baseline_balance.csv)
#       Arms were assigned at random after the baseline pull, so the per-mussel baseline
#       should not differ by FUTURE arm. One-way ANOVA and Kruskal-Wallis per metric. This is
#       the per-mussel companion of the lmer `emm_base` randomization check above and the
#       justification for reading the within-animal change.
#   3c  REPEATABILITY (STATS_repeatability.csv)
#       Pairing only buys power if an animal's attachment is repeatable. Two views:
#       - ICC from the mixed model's variance components, var(mussel) / (var(mussel) +
#         var(residual)), for the same arm x timepoint model as Statistics 2 (thread level;
#         logged metrics as in script 4);
#       - per-arm Pearson and Spearman correlation of the paired animals' baseline and day-3
#         per-mussel means.

metric_cols <- intersect(c("adhesion_kpa", "max_force", "pad_area", "max_displacement"),
                         names(thread_plot))   # extension is optional in the curated table
pm_metrics <- thread_plot %>%
  group_by(mussel, phase, mussel_trt) %>%
  summarise(n_threads = n(), across(all_of(metric_cols), ~ mean(.x, na.rm = TRUE)),
            .groups = "drop") %>%
  mutate(across(all_of(metric_cols), ~ ifelse(is.nan(.x), NA_real_, .x)))

two_sample <- function(a, b) {
  ok <- length(a[!is.na(a)]) >= 2 && length(b[!is.na(b)]) >= 2
  tibble(n_a = sum(!is.na(a)), n_b = sum(!is.na(b)),
         mean_a = mean(a, na.rm = TRUE), mean_b = mean(b, na.rm = TRUE),
         median_a = median(a, na.rm = TRUE), median_b = median(b, na.rm = TRUE),
         diff_a_minus_b = mean(a, na.rm = TRUE) - mean(b, na.rm = TRUE),
         welch_p    = if (ok) t.test(a, b, var.equal = FALSE)$p.value else NA_real_,
         wilcoxon_p = if (ok) suppressWarnings(wilcox.test(a, b, paired = FALSE)$p.value) else NA_real_)
}

# ---- 3a lab-reference check ------------------------------------------------------------
grp_lab  <- pm_metrics %>% filter(phase == "lab")
grp_pre  <- pm_metrics %>% filter(phase == "pre")
grp_ctrl <- pm_metrics %>% filter(phase == "post", mussel_trt == "control")
lab_check <- bind_rows(lapply(metric_cols, function(m) bind_rows(
  two_sample(grp_lab[[m]],  grp_pre[[m]])  %>% mutate(comparison = "lab_day0_vs_pre_day1"),
  two_sample(grp_ctrl[[m]], grp_pre[[m]])  %>% mutate(comparison = "control_day3_vs_pre_day1"),
  two_sample(grp_lab[[m]],  grp_ctrl[[m]]) %>% mutate(comparison = "lab_day0_vs_control_day3")) %>%
  mutate(metric = m, .before = 1))) %>%
  mutate(unit = "per-mussel mean", .after = comparison)
print(as.data.frame(lab_check %>% mutate(across(where(is.numeric), ~ signif(.x, 3)))))
write.csv(lab_check, file.path(out_dir, "STATS_lab_reference_check.csv"), row.names = FALSE)

# ---- 3b baseline balance by future arm -----------------------------------------------
base_by_arm <- grp_pre %>% filter(mussel_trt %in% c("control", "OA", "OW", "DO")) %>%
  mutate(mussel_trt = droplevels(mussel_trt))
balance <- bind_rows(lapply(metric_cols, function(m) {
  d <- base_by_arm %>% filter(!is.na(.data[[m]]))
  if (n_distinct(d$mussel_trt) < 2) return(NULL)
  a <- anova(lm(reformulate("mussel_trt", response = m), data = d))
  k <- kruskal.test(reformulate("mussel_trt", response = m), data = d)
  d %>% group_by(future_arm = mussel_trt) %>%
    summarise(n = n(), mean = mean(.data[[m]]), sd = sd(.data[[m]]), median = median(.data[[m]]),
              .groups = "drop") %>%
    mutate(metric = m, .before = 1) %>%
    mutate(anova_F = a[["F value"]][1], anova_df = paste(a$Df, collapse = ","), anova_p = a[["Pr(>F)"]][1],
           kruskal_H = unname(k$statistic), kruskal_p = k$p.value)
}))
print(as.data.frame(balance %>% mutate(across(where(is.numeric), ~ signif(.x, 3)))))
write.csv(balance, file.path(out_dir, "STATS_baseline_balance.csv"), row.names = FALSE)

# ---- 3c repeatability: ICC and per-arm pre/post correlation ---------------------------
icc_from <- function(fit) {
  vc <- as.data.frame(VarCorr(fit))
  v_m <- vc$vcov[vc$grp == "mussel"]; v_r <- vc$vcov[vc$grp == "Residual"]
  c(icc = v_m / (v_m + v_r), var_mussel = v_m, var_residual = v_r)
}
icc_specs <- tibble::tribble(
  ~metric,            ~scale, ~response,
  "adhesion_kpa",     "raw",  "adhesion_kpa",
  "adhesion_kpa",     "log",  "log(adhesion_kpa)",
  "max_force",        "log",  "log(max_force)",
  "pad_area",         "log",  "log(pad_area)",
  "max_displacement", "raw",  "max_displacement"
)
icc_tbl <- bind_rows(lapply(seq_len(nrow(icc_specs)), function(i) {
  sp <- icc_specs[i, ]
  if (!sp$metric %in% names(dat_lmer)) return(NULL)
  d  <- dat_lmer %>% filter(!is.na(.data[[sp$metric]]))
  fit <- tryCatch(suppressWarnings(lmer(as.formula(paste0(sp$response, " ~ mussel_trt * timepoint + (1 | mussel)")),
                                        data = d)), error = function(e) NULL)
  if (is.null(fit)) return(NULL)
  v <- icc_from(fit)
  tibble(metric = sp$metric, scale = sp$scale, statistic = "ICC_mussel_lmer", arm = "all",
         n = n_distinct(d$mussel), n_threads = nrow(d),
         value = unname(v["icc"]), var_mussel = unname(v["var_mussel"]),
         var_residual = unname(v["var_residual"]), p = NA_real_)
}))

paired_pm <- pm_metrics %>% filter(phase %in% c("pre", "post")) %>%
  select(mussel, mussel_trt, phase, all_of(metric_cols)) %>%
  pivot_wider(names_from = phase, values_from = all_of(metric_cols)) %>%
  filter(!is.na(adhesion_kpa_pre), !is.na(adhesion_kpa_post))
corr_tbl <- bind_rows(lapply(metric_cols, function(m) {
  bind_rows(lapply(c(levels(droplevels(paired_pm$mussel_trt)), "all"), function(a) {
    d <- if (a == "all") paired_pm else paired_pm %>% filter(mussel_trt == a)
    x <- d[[paste0(m, "_pre")]]; y <- d[[paste0(m, "_post")]]
    ok <- !is.na(x) & !is.na(y); x <- x[ok]; y <- y[ok]
    if (length(x) < 3) return(NULL)
    pe <- suppressWarnings(cor.test(x, y, method = "pearson"))
    sp <- suppressWarnings(cor.test(x, y, method = "spearman", exact = FALSE))
    bind_rows(tibble(metric = m, scale = "raw", statistic = "corr_pre_post_pearson", arm = a,
                     n = length(x), n_threads = NA_integer_, value = unname(pe$estimate),
                     var_mussel = NA_real_, var_residual = NA_real_, p = pe$p.value),
              tibble(metric = m, scale = "raw", statistic = "corr_pre_post_spearman", arm = a,
                     n = length(x), n_threads = NA_integer_, value = unname(sp$estimate),
                     var_mussel = NA_real_, var_residual = NA_real_, p = sp$p.value))
  }))
}))
repeatability <- bind_rows(icc_tbl, corr_tbl)
print(as.data.frame(repeatability %>% mutate(across(where(is.numeric), ~ signif(.x, 3)))))
write.csv(repeatability, file.path(out_dir, "STATS_repeatability.csv"), row.names = FALSE)
```

### Provenance
```{r}
# Which configuration produced the files in this folder.
writeLines(c(paste0("generated       : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
             paste0("input           : 03_analyses/thread-summary.xlsx"),
             paste0("INCLUDE_LAB_REFERENCE : ", INCLUDE_LAB_REFERENCE),
             paste0("SHOW_LAB_REFERENCE_IN_PANELS : ", SHOW_LAB_REFERENCE_IN_PANELS),
             paste0("threads analysed: ", nrow(thread_plot)),
             paste0("threads dropped (no pad_area): ", nrow(no_pad)),
             paste0("baseline pool   : ", n_distinct(baseline_pool$mussel), " mussels"),
             paste0("lab reference   : ", n_distinct(grp_lab$mussel), " mussels (day 0, shown, never pooled)"),
             paste0("day-3 control   : ", n_distinct(grp_ctrl$mussel), " mussels"),
             paste0("lmer data       : ", nrow(dat_lmer), " threads, ",
                    n_distinct(dat_lmer$mussel), " mussels"),
             paste0("paired animals  : ", nrow(paired_pm), " (pre and post per-mussel means)"),
             paste0("R / lme4 / lmerTest / emmeans : ", R.version.string, " / ",
                    as.character(packageVersion("lme4")), " / ",
                    as.character(packageVersion("lmerTest")), " / ",
                    as.character(packageVersion("emmeans")))),
           file.path(out_dir, "RUN_provenance.txt"))
```
