---
title: "3_analyze_thread_strength"
author: "Matt George"
date: "2023-04-25"
output: html_document
---

<!--
PURPOSE
  Analyze byssal-thread plaque adhesion (kPa) for Mytilus trossulus before and
  after a 3-day stress exposure (ocean acidification OA, ocean warming OW,
  hypoxia DO), relative to a day-0 baseline and a day-3 control arm.

DATA MODEL (read this before editing the filters below)
  The input table mixes three columns that are easy to confuse:
    - group       : "control" = day-0 BASELINE (pre-stress); "treatment" = day-3 (post-stress).
                    This is the real before/after axis.
    - mussel_trt  : the arm each mussel was ASSIGNED to (its fate): control, OA, OW, DO.
                    Each mussel maps to exactly one fate (verified).
    - treatment   : the condition under which a given THREAD was made. Every baseline
                    thread is labeled "control" regardless of fate, and the control-arm
                    day-3 threads are ALSO labeled "control". So `treatment == "control"`
                    is a mixture of (i) day-0 baselines from all arms and (ii) day-3
                    control-arm threads (mussels T126-T135).
  Because of that overload, all filtering in this script is done on `group` + `mussel_trt`,
  and a clean `timepoint` factor ("baseline"/"post") is derived from `group`.

FIGURE CONVENTION FOR THE LINE+BOX PANELS (one per stress arm)
  - Left cluster  ("control") = ALL day-0 baselines (group == "control"); the SAME 54 mussels
    appear on every panel (this is the chosen "population baseline" reference).
  - Right cluster (the arm)   = only mussels that produced threads in that arm at day 3.
  - A connecting line is drawn only for mussels present on BOTH sides (true repeated measures);
    unpaired mussels appear as a single point with no line.
  - A boxplot of the per-mussel means flanks each cluster: control box to the LEFT of the
    control points, treatment box to the RIGHT of the treatment points, both offset so they
    never overlap the points or cross the connecting lines.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages; set ggplot theme
```{r message=FALSE, warning=FALSE}
## Clear workspace
rm(list = ls())

## Load packages (auto-install any that are missing)
load.lib <- c("readxl", "tidyverse", "bestNormalize", "agricolae",
              "nlme", "multcomp", "rstatix", "lme4", "lmerTest",
              "emmeans", "broom", "broom.mixed", "colorspace", "here")
              # emmeans: lmer post-hoc contrasts; broom/broom.mixed: tidy stats output;
              # colorspace: lightened violin fills in p2
install.lib <- load.lib[!load.lib %in% installed.packages()]
for (lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)

## ggplot theme (ggplot2 >= 3.4: line/rect widths use `linewidth`, not `size`)
my_theme <- theme(line             = element_line(linewidth = 1.2),
                  rect             = element_rect(linewidth = 1.2),
                  text             = element_text(size = 14, color = "black"),
                  panel.background = element_blank(),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.text.x      = element_text(size = 16, color = "black"),
                  axis.text.y      = element_text(size = 16, color = "black"),
                  axis.title.x     = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y     = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x     = element_line(color = "black"),
                  axis.ticks.y     = element_line(color = "black"),
                  panel.border     = element_rect(color = "black", fill = NA, linewidth = 1.2),
                  legend.key       = element_blank(),
                  legend.position  = "none")

## Treatment palette (kept from prior version)
treatment_colors <- c("control" = "#5DADE2",   # blue
                      "OW"      = "#fdae61",    # orange
                      "OA"      = "#4CAF50",    # green
                      "DO"      = "#9884a3")    # purple
```

```{r}
# Create this script's output folder under 03_analyses (idempotent)
out_dir <- here::here("03_analyses", "analyze-thread-strength")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

### Load data set; derive adhesion and timepoint
```{r}
# Load thread-level table
thread_plot <- read_excel(here::here("02_data", "thread-summary-trossulus-clean.xlsx"),
                          sheet = "data", col_names = TRUE)

# Adhesion strength (kPa) = peak force (N) / pad area (mm^2) * 1000.
# Recomputed here so the script does not depend on the spreadsheet's cached formula column.
thread_plot$adhesion_kpa <- thread_plot$max_force / thread_plot$pad_area * 1000

# Derive a clean before/after axis from `group` (see DATA MODEL note at top).
thread_plot$timepoint <- factor(ifelse(thread_plot$group == "control", "baseline", "post"),
                                levels = c("baseline", "post"))

# Factors. `treatment` keeps the analysis/plot order; `mussel_trt` uses control as the
# reference level so the lmer interaction reads as "change relative to the control arm".
thread_plot$treatment  <- factor(thread_plot$treatment,  levels = c("control", "OW", "OA", "DO"))
thread_plot$mussel_trt <- factor(thread_plot$mussel_trt, levels = c("control", "OA", "OW", "DO"))
thread_plot$mussel     <- factor(thread_plot$mussel)
thread_plot$thread     <- factor(thread_plot$thread)

# Quick structural check (printed once): every mussel should map to exactly one fate.
stopifnot(all(thread_plot %>% group_by(mussel) %>%
                summarise(n_fate = n_distinct(mussel_trt), .groups = "drop") %>% pull(n_fate) == 1))
```

### Per-mussel summaries (the unit of replication for figures and classic tests)
```{r message=FALSE}
# Collapse threads to one adhesion summary per mussel x timepoint x treatment.
# Within a timepoint a mussel carries a single treatment label, so this yields one row
# per mussel per timepoint. SE is NA for singletons (1 thread); those points get no error bar.
per_mussel_means <- thread_plot %>%
  group_by(mussel, timepoint, treatment) %>%
  summarise(adh_mean = mean(adhesion_kpa),
            adh_sd   = sd(adhesion_kpa),
            n        = n(),
            adh_se   = adh_sd / sqrt(n),
            .groups  = "drop")

# Fixed horizontal offset per mussel, computed ONCE over all mussels so the control
# cluster looks identical across the three arm panels. Used only for point/line placement.
set.seed(1)
mussel_offsets <- tibble(mussel = levels(thread_plot$mussel),
                         xoff   = runif(nlevels(thread_plot$mussel), -0.08, 0.08))
```

### Distribution plots: adhesion by treatment (manuscript Fig. "all threads" / "by individual")
```{r warning=FALSE}
# These four panels summarize the distribution of adhesion. p1 (all threads) and
# p3 (per-mussel means) correspond to the two manuscript panels; p2 and p4 restrict to
# mussels measured at BOTH timepoints (true repeated measures).
#
# NOTE: x here is the `treatment` column, so its "control" bar pools day-0 baselines (all
# arms) with day-3 control-arm threads. These panels are descriptive; the before/after
# structure is shown properly by the line+box panels further down.

# Mussels measured at both baseline and post-stress (repeated-measures subset).
thread_plot_repeated <- thread_plot %>%
  group_by(mussel) %>%
  filter(n_distinct(timepoint) == 2) %>%
  ungroup()

# Per-mussel means, all mussels and repeated-only, for the mean-based panels.
thread_plot_all_mean <- per_mussel_means
thread_plot_repeated_mean <- per_mussel_means %>%
  group_by(mussel) %>%
  filter(n_distinct(timepoint) == 2) %>%
  ungroup()

# p1: all threads, violin + box + jittered points
p1 <- ggplot(thread_plot, aes(x = treatment, y = adhesion_kpa)) +
      geom_violin(aes(fill = treatment), color = "black", linewidth = 1.2, bw = 12) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = NA,
                   width = 0.2, notch = FALSE, varwidth = FALSE, alpha = 0.8, coef = 0) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5, alpha = .1) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5,
                 stroke = .7, shape = 1, color = "black") +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      scale_fill_manual(values = treatment_colors) +
      xlab(NULL) + ylab(NULL) + my_theme
p1

# p2: repeated-measures mussels only, all threads
p2 <- ggplot(thread_plot_repeated, aes(x = treatment, y = adhesion_kpa)) +
      geom_violin(aes(fill = treatment,
                      fill = after_scale(colorspace::lighten(fill, .5))),
                  color = "black", linewidth = 1.2, bw = 12) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = NA,
                   width = 0.2, notch = FALSE, varwidth = FALSE, alpha = 0.8, coef = 0) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5, alpha = .3) +
      geom_point(position = position_jitter(width = .03, seed = 0), size = 5,
                 stroke = .7, shape = 1, color = "black") +
      scale_fill_manual(values = treatment_colors) +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      xlab(NULL) + ylab(NULL) + my_theme
p2

# p3: all mussels, per-mussel means
p3 <- ggplot(thread_plot_all_mean, aes(x = treatment, y = adh_mean)) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = 16,
                   notch = FALSE, varwidth = TRUE, alpha = 0.8) +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      xlab(NULL) + ylab(NULL) + my_theme
p3

# p4: repeated-measures mussels only, per-mussel means
p4 <- ggplot(thread_plot_repeated_mean, aes(x = treatment, y = adh_mean)) +
      geom_boxplot(colour = "black", linewidth = 0.8, outlier.shape = 16,
                   notch = FALSE, varwidth = TRUE, alpha = 0.8) +
      scale_y_continuous(breaks = seq(0, 200, 25), limits = c(0, 210)) +
      xlab(NULL) + ylab(NULL) + my_theme
p4

ggsave(file.path(out_dir, "BP_AllMussels_AllThreads.png"),
       plot = p1, dpi = 1200, device = "png", width = 7, height = 5, units = "in")
ggsave(file.path(out_dir, "BP_OnlyRepeatedMussels_AllThreads.png"),
       plot = p2, dpi = 1200, device = "png", width = 8, height = 6, units = "in")
ggsave(file.path(out_dir, "BP_AllMussels_MeanThreads.png"),
       plot = p3, dpi = 1200, device = "png", width = 4, height = 4, units = "in")
ggsave(file.path(out_dir, "BP_OnlyRepeatedMussels_MeanThreads.png"),
       plot = p4, dpi = 1200, device = "png", width = 4, height = 4, units = "in")
```

### Line + box panels: baseline vs each stress arm
```{r warning=FALSE}
# One reusable builder, called once per arm (removes the previous copy-paste blocks).
#
# x-geometry (continuous axis):
#   control points/lines/error bars live near x = 1 (within +/- 0.08)
#   arm     points/lines/error bars live near x = 2 (within +/- 0.08)
#   control box sits at x = 0.60, arm box at x = 2.40 (width 0.22)
# so the boxes are always clear of the point clusters and the connecting lines.
#
# Coloring: control points/box in control blue; arm points/box in the arm color;
# connecting lines neutral grey so they read as links rather than adding new colors.

make_line_box <- function(trt_code, trt_label = trt_code,
                          y_breaks = seq(0, 200, 25), y_lim = c(0, 210)) {

  control_col <- treatment_colors[["control"]]
  trt_col     <- treatment_colors[[trt_code]]
  box_w <- 0.22; box_ctrl_x <- 0.60; box_trt_x <- 2.40

  # Control side = ALL day-0 baselines (same 54 mussels on every panel).
  ctrl <- per_mussel_means %>% filter(timepoint == "baseline") %>%
          mutate(side = "control", xbase = 1, xbox = box_ctrl_x)
  # Arm side = mussels that produced threads in this arm at day 3.
  trt  <- per_mussel_means %>% filter(timepoint == "post", treatment == trt_code) %>%
          mutate(side = "treatment", xbase = 2, xbox = box_trt_x)

  dat <- bind_rows(ctrl, trt) %>%
         left_join(mussel_offsets, by = "mussel") %>%
         mutate(xpos = xbase + xoff)

  # Connecting lines only for mussels present on both sides (true repeated measures).
  paired_ids <- intersect(ctrl$mussel, trt$mussel)
  lines_dat  <- dat %>% filter(mussel %in% paired_ids)

  ggplot() +
    # connecting lines (paired only)
    geom_line(data = lines_dat,
              aes(x = xpos, y = adh_mean, group = mussel),
              color = "grey55", linewidth = 0.6, alpha = 0.8) +
    # flanking boxplots of the per-mussel means (outliers hidden; points shown separately)
    geom_boxplot(data = filter(dat, side == "control"),
                 aes(x = xbox, y = adh_mean, group = 1),
                 width = box_w, fill = control_col, color = "black",
                 linewidth = 0.7, alpha = 0.55, outlier.shape = NA) +
    geom_boxplot(data = filter(dat, side == "treatment"),
                 aes(x = xbox, y = adh_mean, group = 1),
                 width = box_w, fill = trt_col, color = "black",
                 linewidth = 0.7, alpha = 0.55, outlier.shape = NA) +
    # control points + SE
    geom_errorbar(data = filter(dat, side == "control"),
                  aes(x = xpos, ymin = adh_mean - adh_se, ymax = adh_mean + adh_se),
                  width = 0.04, linewidth = 0.6, color = "black", na.rm = TRUE) +
    geom_point(data = filter(dat, side == "control"),
               aes(x = xpos, y = adh_mean), size = 3, color = control_col) +
    geom_point(data = filter(dat, side == "control"),
               aes(x = xpos, y = adh_mean), size = 3, shape = 1, stroke = 0.6, color = "black") +
    # arm points + SE
    geom_errorbar(data = filter(dat, side == "treatment"),
                  aes(x = xpos, ymin = adh_mean - adh_se, ymax = adh_mean + adh_se),
                  width = 0.04, linewidth = 0.6, color = "black", na.rm = TRUE) +
    geom_point(data = filter(dat, side == "treatment"),
               aes(x = xpos, y = adh_mean), size = 3, color = trt_col) +
    geom_point(data = filter(dat, side == "treatment"),
               aes(x = xpos, y = adh_mean), size = 3, shape = 1, stroke = 0.6, color = "black") +
    scale_x_continuous(breaks = c(1, 2), labels = c("control", trt_label),
                       limits = c(0.4, 2.6)) +
    scale_y_continuous(breaks = y_breaks, limits = y_lim) +
    xlab(NULL) + ylab(NULL) + my_theme
}

p_OA <- make_line_box("OA"); p_OA
p_OW <- make_line_box("OW"); p_OW
p_DO <- make_line_box("DO"); p_DO

ggsave(file.path(out_dir, "LineBox_OA_BaselineVsTreatment.png"),
       plot = p_OA, dpi = 1200, device = "png", width = 4.5, height = 4.5, units = "in")
ggsave(file.path(out_dir, "LineBox_OW_BaselineVsTreatment.png"),
       plot = p_OW, dpi = 1200, device = "png", width = 4.5, height = 4.5, units = "in")
ggsave(file.path(out_dir, "LineBox_DO_BaselineVsTreatment.png"),
       plot = p_DO, dpi = 1200, device = "png", width = 4.5, height = 4.5, units = "in")
```

<!--
STATISTICS

  The previous version's tests were not valid for this design and have been replaced:
    - The "all treatments" model was aov(response ~ trt) with NO error term, so it treated
      every thread as independent (pseudoreplication) while its comment claimed it accounted
      for repeated sampling.
    - The per-arm models used aov(response ~ trt + Error(ID/thread)); `thread` (1,2,3) is just
      a within-mussel index, not a meaningful nesting stratum, and the data fed in mixed all
      arms' baselines into the "control" level. aov(... Error()) also requires balanced,
      complete cells, which this partially paired design does not have.

  Replacement, in two parts:
    (1) CLASSIC tests on the per-mussel means (one value per mussel per timepoint), which is
        the correct unit of replication and removes pseudoreplication without a mixed model.
    (2) A MIXED-MODEL section (lmer) on the full thread-level data, which uses every thread and
        every animal (paired and unpaired) and tests the arm-by-timepoint interaction directly.

  Parametric and non-parametric versions are reported side by side rather than relying on an
  automatic transform, so the reader can see both. Sample sizes are small; results are
  descriptive of this pilot, not confirmatory.
-->

### Statistics 1a: between-arm comparison of day-3 adhesion (per-mussel means)
```{r}
# Each mussel contributes one day-3 mean, so groups are independent (no pseudoreplication).
post_means <- per_mussel_means %>%
  filter(timepoint == "post") %>%
  mutate(treatment = factor(treatment, levels = c("control", "OA", "OW", "DO")))

cat("Day-3 per-mussel means, n per arm:\n")
print(post_means %>% count(treatment))

# Parametric: one-way ANOVA + Tukey HSD
aov_between <- aov(adh_mean ~ treatment, data = post_means)
print(summary(aov_between))
tukey_between <- TukeyHSD(aov_between)
print(tukey_between)

# Non-parametric backup: Kruskal-Wallis (robust to the small, skewed samples)
print(kruskal.test(adh_mean ~ treatment, data = post_means))

# Save
write.csv(broom::tidy(aov_between),
          file.path(out_dir, "STATS_between_arm_day3_ANOVA.csv"), row.names = FALSE)
write.csv(broom::tidy(tukey_between),
          file.path(out_dir, "STATS_between_arm_day3_TukeyHSD.csv"), row.names = FALSE)
```

### Statistics 1b: within-arm before/after, per arm (per-mussel means)
```{r warning=FALSE}
# Two complementary questions per arm:
#   PAIRED  : did the animals measured at both timepoints change? (matches the connecting lines;
#             uses only paired mussels, so n is small)
#   2-SAMPLE: does the arm's day-3 distribution differ from the full day-0 baseline pool?
#             (matches the boxes; uses all baseline and all arm animals)
# The control arm has no within-subject pairs (baseline cohort T001-T012 differs from the
# day-3 control cohort T126-T135), so it has no paired test.

baseline_means <- per_mussel_means %>% filter(timepoint == "baseline")

run_arm_tests <- function(trt_code) {
  post_x <- per_mussel_means %>% filter(timepoint == "post", treatment == trt_code)
  paired <- inner_join(baseline_means %>% select(mussel, base = adh_mean),
                       post_x        %>% select(mussel, post = adh_mean),
                       by = "mussel")

  res <- tibble(arm = trt_code,
                n_pairs   = nrow(paired),
                n_base    = nrow(baseline_means),
                n_post    = nrow(post_x),
                delta_mean = if (nrow(paired) >= 1) mean(paired$post - paired$base) else NA_real_)

  # Paired tests (need >= 2 pairs)
  if (nrow(paired) >= 2) {
    res$paired_t_p   <- t.test(paired$post, paired$base, paired = TRUE)$p.value
    res$wilcoxon_p   <- suppressWarnings(wilcox.test(paired$post, paired$base, paired = TRUE)$p.value)
  } else {
    res$paired_t_p <- NA_real_; res$wilcoxon_p <- NA_real_
  }

  # Two-sample tests: baseline pool vs this arm's day-3
  res$welch_p   <- t.test(post_x$adh_mean, baseline_means$adh_mean, var.equal = FALSE)$p.value
  res$mannwhit_p <- suppressWarnings(
    wilcox.test(post_x$adh_mean, baseline_means$adh_mean, paired = FALSE)$p.value)
  res
}

within_arm <- bind_rows(lapply(c("OA", "OW", "DO"), run_arm_tests))
print(as.data.frame(within_arm))

write.csv(within_arm, file.path(out_dir, "STATS_within_arm_beforeafter.csv"), row.names = FALSE)
```

### Statistics 2: mixed model on all threads (lmer)
```{r warning=FALSE}
# Full thread-level model. Random intercept for mussel absorbs (a) multiple threads per
# mussel x timepoint and (b) the within-mussel correlation across timepoints for paired
# animals. Fixed effects: arm, timepoint, and their interaction (the interaction is the
# scientific target: does the baseline -> day-3 change differ by arm?).
#
# CAVEAT: the control arm is not paired within-subject (different baseline and day-3 cohorts),
# so for the control arm "timepoint" is a between-cohort contrast. The clean within-subject
# evidence is the paired tests in 1b; this model adds the full-data, all-animals view.

dat_lmer <- thread_plot %>% droplevels()

mod <- lmer(adhesion_kpa ~ mussel_trt * timepoint + (1 | mussel), data = dat_lmer)

cat("Type III ANOVA (lmerTest default, Satterthwaite df):\n")
print(anova(mod))
cat("\nFixed-effect estimates:\n")
print(summary(mod)$coefficients)

# Residual diagnostics (adhesion is right-skewed; check before trusting the Gaussian fit)
qqnorm(resid(mod)); qqline(resid(mod))
plot(fitted(mod), resid(mod), xlab = "Fitted", ylab = "Residuals"); abline(h = 0, lty = 2)

# Robustness: same model on log adhesion. Compare the interaction term against the raw fit.
mod_log <- lmer(log(adhesion_kpa) ~ mussel_trt * timepoint + (1 | mussel), data = dat_lmer)
cat("\nType III ANOVA, log(adhesion) model:\n")
print(anova(mod_log))

# Interpretable contrasts via estimated marginal means
emm_change <- emmeans(mod, ~ timepoint | mussel_trt)   # post vs baseline, within each arm
cat("\nWithin-arm change (post - baseline):\n")
print(pairs(emm_change))

emm_post <- emmeans(mod, ~ mussel_trt, at = list(timepoint = "post"))   # arm means at day 3
cat("\nArm marginal means at day 3 (post):\n")
print(emm_post)
cat("\nArm contrasts at day 3 (post):\n")
print(pairs(emm_post, adjust = "tukey"))

# Randomization check: arms were assigned at random, so baseline adhesion should be similar.
emm_base <- emmeans(mod, ~ mussel_trt, at = list(timepoint = "baseline"))
cat("\nArm marginal means at baseline (randomization check):\n")
print(emm_base)

# Save
write.csv(broom.mixed::tidy(mod, effects = "fixed"),
          file.path(out_dir, "STATS_lmer_fixed_effects.csv"), row.names = FALSE)
write.csv(as.data.frame(pairs(emm_change)),
          file.path(out_dir, "STATS_lmer_within_arm_change.csv"), row.names = FALSE)
write.csv(as.data.frame(pairs(emm_post, adjust = "tukey")),
          file.path(out_dir, "STATS_lmer_arm_contrasts.csv"), row.names = FALSE)
```
