---
title: "4_decompose_adhesion"
author: "Matt George"
date: "2026-09-16"
output: html_document
---

<!--
PURPOSE
  Adhesion (kPa) is a ratio: peak force / plaque area. Script 3 models the ratio. This
  script models the two parts separately, plus extension at break, so that an arm whose
  force collapses while its plaques shrink is not read as "no different" from an arm whose
  force holds while its plaques grow, just because the two ratios happen to fall by the
  same fraction.

  On the log scale the decomposition is exact:
      log(adhesion) = log(max_force) - log(pad_area)
  so the fitted change in log-adhesion is the difference of the fitted changes in
  log-force and log-area. Every model below is the same model as script 3's mixed model,
  with a different response.

INPUT
  03_analyses/thread-summary.xlsx, sheet "data" - the same curated table script 3 reads.
  Same schema normalisation, same phase logic, same exclusion of the lab-reference animals
  from the models (they have no before/after position).

OUTPUT  03_analyses/decompose-adhesion/
  STATS_decomp_typeIII_ANOVA.csv       arm, timepoint, interaction - one block per response
  STATS_decomp_within_arm_change.csv   post vs baseline within each arm, per response,
                                       back-transformed to a ratio for the logged responses
  STATS_decomp_paired_by_arm.csv       per-animal paired change, per arm, per response
  STATS_failure_mode.csv               failure-mode counts by thread treatment + chi-square
                                       + Fisher's exact (Monte Carlo) p for the day-3 arms
  FIG_paired_change_by_metric.png      paired % change with 95% CI, one panel per response
  FIG_failure_mode_by_arm.png          failure-mode composition by thread treatment
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages; theme; palette
```{r message=FALSE, warning=FALSE}
rm(list = ls())

load.lib <- c("readxl", "tidyverse", "lme4", "lmerTest", "emmeans", "broom", "here")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for (lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)

my_theme <- theme(line             = element_line(linewidth = 1.0),
                  rect             = element_rect(linewidth = 1.0),
                  text             = element_text(size = 13, color = "black"),
                  panel.background = element_blank(),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.text        = element_text(size = 12, color = "black"),
                  panel.border     = element_rect(color = "black", fill = NA, linewidth = 1.0),
                  strip.background = element_rect(fill = "grey92", color = "black"),
                  strip.text       = element_text(size = 12, face = "bold"),
                  legend.key       = element_blank(),
                  legend.position  = "none")

# Same palette as script 3, keyed on the ARM here (mussel_trt), because every panel below is
# a within-animal contrast and the animal's arm is the grouping that matters.
arm_colors <- c("control" = "#7FB3D5", "OA" = "#4CAF50", "OW" = "#fdae61", "DO" = "#9884a3")

out_dir <- here::here("03_analyses", "decompose-adhesion")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

### Load and normalise (identical to script 3)
```{r}
thread_plot <- read_excel(here::here("03_analyses", "thread-summary.xlsx"),
                          sheet = "data", col_names = TRUE)

ren <- c(mussel_ID = "mussel", thread_num = "thread", treatment = "thread_trt")
for (from in names(ren)) {
  to <- ren[[from]]
  if (from %in% names(thread_plot) && !to %in% names(thread_plot))
    names(thread_plot)[names(thread_plot) == from] <- to
}
need <- c("mussel", "thread", "thread_trt", "mussel_trt", "max_force", "pad_area",
          "max_displacement")
if (length(setdiff(need, names(thread_plot))) > 0)
  stop("thread-summary.xlsx is missing: ", paste(setdiff(need, names(thread_plot)), collapse = ", "))

if (!"phase" %in% names(thread_plot))
  thread_plot$phase <- dplyr::case_when(thread_plot$thread_trt == "lab_control" ~ "lab",
                                        thread_plot$thread_trt == "baseline"    ~ "pre",
                                        TRUE                                    ~ "post")

norm_id <- function(x) {
  x   <- trimws(as.character(x))
  pre <- sub("^([A-Za-z]+).*$", "\\1", x)
  num <- suppressWarnings(as.integer(sub("^[A-Za-z]+0*([0-9]+).*$", "\\1", x)))
  ifelse(is.na(num) | pre == x, NA_character_, sprintf("%s%03d", pre, num))
}

dat <- thread_plot %>%
  mutate(mussel       = norm_id(mussel),
         adhesion_kpa = max_force / pad_area * 1000) %>%
  filter(!is.na(adhesion_kpa), phase %in% c("pre", "post")) %>%
  mutate(timepoint  = factor(if_else(phase == "pre", "baseline", "post"),
                             levels = c("baseline", "post")),
         mussel_trt = factor(mussel_trt, levels = c("control", "OA", "OW", "DO")),
         mussel     = factor(mussel))

if (any(is.na(dat$mussel_trt)))
  stop("Unrecognised mussel_trt value(s): ",
       paste(unique(thread_plot$mussel_trt[!thread_plot$mussel_trt %in% levels(dat$mussel_trt)]),
             collapse = ", "))

cat("Threads in models:", nrow(dat), "| mussels:", n_distinct(dat$mussel), "\n")
dat %>% count(mussel_trt, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0) %>% knitr::kable()
```

### The four responses
```{r}
# Logged where the quantity is positive and right-skewed, so that a change is a ratio and
# the force/area decomposition of adhesion is additive. Extension stays on its own scale.
responses <- tibble::tribble(
  ~response,           ~column,            ~transform, ~label,
  "adhesion_kpa",      "adhesion_kpa",     "log",      "Adhesion (kPa)",
  "max_force",         "max_force",        "log",      "Peak force (N)",
  "pad_area",          "pad_area",         "log",      "Plaque area (mm^2)",
  "max_displacement",  "max_displacement", "none",     "Extension at break (mm)"
)
```

### Mixed model per response: arm x timepoint, random intercept for mussel
```{r warning=FALSE}
fit_one <- function(col, transform) {
  y <- if (transform == "log") paste0("log(", col, ")") else col
  f <- as.formula(paste0(y, " ~ mussel_trt * timepoint + (1 | mussel)"))
  lmer(f, data = dat)
}

fits <- pmap(responses, function(response, column, transform, label)
  list(response = response, label = label, transform = transform,
       fit = fit_one(column, transform)))
names(fits) <- responses$response

# Type III ANOVA, one block per response
anova_tbl <- map_dfr(fits, function(x)
  as.data.frame(anova(x$fit)) %>%
    tibble::rownames_to_column("term") %>%
    mutate(response = x$response, scale = x$transform, .before = 1))
knitr::kable(anova_tbl %>% select(response, scale, term, NumDF, DenDF, `F value`, `Pr(>F)`),
             digits = 4)
write.csv(anova_tbl, file.path(out_dir, "STATS_decomp_typeIII_ANOVA.csv"), row.names = FALSE)

# Within-arm change, post vs baseline. For logged responses `type = "response"` returns the
# post/baseline RATIO with its CI; for extension it is the raw difference.
change_tbl <- map_dfr(fits, function(x) {
  # The formula is built programmatically inside fit_one(), so the model call holds the symbol
  # `f` rather than the literal `log(...)`, and emmeans cannot see the transformation on its
  # own. Declare it explicitly, otherwise `type = "response"` silently returns the log scale.
  emm <- if (x$transform == "log")
           emmeans(x$fit, ~ timepoint | mussel_trt, tran = "log")
         else
           emmeans(x$fit, ~ timepoint | mussel_trt)
  ctr <- if (x$transform == "log")
           contrast(emm, method = "revpairwise", type = "response")   # post / baseline, a ratio
         else
           contrast(emm, method = "revpairwise")                      # post - baseline
  out <- as.data.frame(ctr)
  if (x$transform == "log") {
    ci  <- as.data.frame(confint(ctr, type = "response"))
    out <- out %>% mutate(lower.CL = ci$lower.CL, upper.CL = ci$upper.CL,
                          pct_change = 100 * (ratio - 1))
  }
  out %>% mutate(response = x$response, scale = x$transform, .before = 1)
})
knitr::kable(change_tbl, digits = 4)
write.csv(change_tbl, file.path(out_dir, "STATS_decomp_within_arm_change.csv"), row.names = FALSE)
```

### Per-animal paired change, per arm, per response
```{r warning=FALSE}
# One value per animal per timepoint (the mean of its threads), then post - pre. This is the
# same unit of replication as script 3's paired tests; it is repeated here for every response
# so the decomposition can be read off one table.
per_mussel <- dat %>%
  group_by(mussel, mussel_trt, timepoint) %>%
  summarise(across(c(adhesion_kpa, max_force, pad_area, max_displacement), mean),
            .groups = "drop")

paired <- per_mussel %>%
  pivot_wider(names_from = timepoint,
              values_from = c(adhesion_kpa, max_force, pad_area, max_displacement)) %>%
  filter(!is.na(adhesion_kpa_baseline), !is.na(adhesion_kpa_post))

paired_tbl <- map_dfr(responses$response, function(r) {
  paired %>%
    group_by(mussel_trt) %>%
    summarise(response   = r,
              n_pairs    = n(),
              mean_pre   = mean(.data[[paste0(r, "_baseline")]]),
              mean_post  = mean(.data[[paste0(r, "_post")]]),
              delta      = mean(.data[[paste0(r, "_post")]] - .data[[paste0(r, "_baseline")]]),
              pct_change = 100 * delta / mean_pre,
              ci_low     = if (n() >= 2) t.test(.data[[paste0(r, "_post")]] - .data[[paste0(r, "_baseline")]])$conf.int[1] else NA_real_,
              ci_high    = if (n() >= 2) t.test(.data[[paste0(r, "_post")]] - .data[[paste0(r, "_baseline")]])$conf.int[2] else NA_real_,
              paired_t_p = if (n() >= 2) t.test(.data[[paste0(r, "_post")]], .data[[paste0(r, "_baseline")]], paired = TRUE)$p.value else NA_real_,
              wilcoxon_p = if (n() >= 2) suppressWarnings(wilcox.test(.data[[paste0(r, "_post")]], .data[[paste0(r, "_baseline")]], paired = TRUE)$p.value) else NA_real_,
              .groups = "drop") %>%
    mutate(pct_ci_low  = 100 * ci_low  / mean_pre,
           pct_ci_high = 100 * ci_high / mean_pre)
})

knitr::kable(paired_tbl %>%
               select(response, mussel_trt, n_pairs, mean_pre, mean_post, delta, pct_change,
                      paired_t_p, wilcoxon_p),
             digits = 3)
write.csv(paired_tbl, file.path(out_dir, "STATS_decomp_paired_by_arm.csv"), row.names = FALSE)
```

### Figure: paired % change by arm, one panel per response
```{r fig.width=9, fig.height=6, warning=FALSE}
plot_dat <- paired_tbl %>%
  mutate(response   = factor(response, levels = responses$response,
                             labels = responses$label),
         mussel_trt = factor(mussel_trt, levels = c("control", "OA", "OW", "DO")))

p_change <- ggplot(plot_dat, aes(x = mussel_trt, y = pct_change)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(aes(ymin = pct_ci_low, ymax = pct_ci_high),
                width = 0.18, linewidth = 0.7, color = "black") +
  geom_point(aes(fill = mussel_trt), shape = 21, size = 4, stroke = 0.8, color = "black") +
  geom_text(aes(label = paste0("n=", n_pairs), y = pct_ci_high), vjust = -0.8, size = 3.2) +
  facet_wrap(~ response, scales = "free_y", nrow = 2) +
  scale_fill_manual(values = arm_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.22))) +
  labs(x = NULL, y = "Paired change, day 3 vs baseline (%)") +
  my_theme
p_change

ggsave(file.path(out_dir, "FIG_paired_change_by_metric.png"), plot = p_change,
       dpi = 600, width = 9, height = 6, units = "in")
```

### Per-animal response classification (handoff to gene-mechanics-correlation)

Which animals made **weaker** threads at day 3 than at baseline, and which made
**stronger**, on each of the four metrics above. One row per animal that produced day-3
threads; the change columns are populated only for animals that also have baseline threads
(`paired = TRUE`), because a change needs a before.

Two reference frames are given side by side, because they answer different questions:

  - `<metric>_direction`     : raw sign of the animal's own change, `decreased` / `increased`.
                               This is "did this animal's threads get weaker".
  - `<metric>_vs_control`    : the animal's % change minus the control arm's mean % change,
                               and its sign as `below_control` / `above_control`.
                               Every arm fell ~20% in adhesion, so an animal that fell 20% is
                               typical, not a poor responder. This frame asks "weaker than the
                               control trajectory", which is the stress-specific question.

The composite `response_class` uses the two metrics that define attachment strength, peak
force and adhesion: `weaker` if both decreased, `stronger` if both increased, `mixed`
otherwise. `response_score` is the mean of the animal's standardised log-ratios across
force, area and adhesion (higher = held up better), a single continuous number for the
correlation pipeline. Extension is reported but kept out of the composite; its direction is
not a strength direction.

Written to `03_analyses/decompose-adhesion/mussel_response_classification.csv`. Script 20 in
`gene-mechanics-correlation/` reads it.

```{r}
# Per-animal means at each timepoint, every animal with day-3 threads
resp_cols <- c("adhesion_kpa", "max_force", "pad_area", "max_displacement")
pm_all <- dat %>%
  group_by(mussel, mussel_trt, timepoint) %>%
  summarise(n_threads = n(), across(all_of(resp_cols), mean), .groups = "drop")

post_side <- pm_all %>% filter(timepoint == "post") %>%
  select(mussel, mussel_trt, n_threads_post = n_threads, all_of(resp_cols)) %>%
  rename_with(~ paste0(.x, "_post"), all_of(resp_cols))
pre_side  <- pm_all %>% filter(timepoint == "baseline") %>%
  select(mussel, n_threads_pre = n_threads, all_of(resp_cols)) %>%
  rename_with(~ paste0(.x, "_pre"), all_of(resp_cols))

resp <- post_side %>%
  left_join(pre_side, by = "mussel") %>%
  mutate(paired = !is.na(n_threads_pre),
         n_threads_pre = replace_na(n_threads_pre, 0L))

# Change columns. Log-ratio for the positive, right-skewed metrics (matches the models above);
# plain difference for extension. % change for all four, on the raw scale.
for (m in resp_cols) {
  pre <- resp[[paste0(m, "_pre")]]; post <- resp[[paste0(m, "_post")]]
  resp[[paste0(m, "_pct_change")]] <- 100 * (post - pre) / pre
  resp[[paste0(m, "_change")]] <- if (m == "max_displacement") post - pre else log(post / pre)
  resp[[paste0(m, "_direction")]] <- case_when(is.na(pre) ~ NA_character_,
                                               post < pre ~ "decreased",
                                               post > pre ~ "increased",
                                               TRUE       ~ "unchanged")
}

# Control-referenced frame: subtract the control arm's mean % change on each metric
ctrl_mean <- resp %>% filter(paired, mussel_trt == "control") %>%
  summarise(across(ends_with("_pct_change"), ~ mean(.x, na.rm = TRUE)))
if (nrow(ctrl_mean) == 0 || any(is.na(ctrl_mean[1, ])))
  warning("Control arm has no paired animals; *_vs_control columns will be NA.", call. = FALSE)
for (m in resp_cols) {
  ref <- if (nrow(ctrl_mean)) ctrl_mean[[paste0(m, "_pct_change")]] else NA_real_
  resp[[paste0(m, "_vs_control")]] <- resp[[paste0(m, "_pct_change")]] - ref
  resp[[paste0(m, "_rel_direction")]] <- case_when(
    is.na(resp[[paste0(m, "_vs_control")]]) ~ NA_character_,
    resp[[paste0(m, "_vs_control")]] < 0    ~ "below_control",
    TRUE                                    ~ "above_control")
}

# Composite class and score
zs <- function(x) if (sum(!is.na(x)) > 1) as.numeric(scale(x)) else x * NA_real_
resp <- resp %>%
  mutate(response_class = case_when(
           !paired ~ NA_character_,
           max_force_direction == "decreased" & adhesion_kpa_direction == "decreased" ~ "weaker",
           max_force_direction == "increased" & adhesion_kpa_direction == "increased" ~ "stronger",
           TRUE ~ "mixed"),
         response_score = rowMeans(cbind(zs(max_force_change), zs(pad_area_change),
                                         zs(adhesion_kpa_change)), na.rm = FALSE)) %>%
  arrange(mussel_trt, mussel)

# Tidy column order: identity, pairing, then one block per metric, then the composite
ordered_cols <- c("mussel", "mussel_trt", "paired", "n_threads_pre", "n_threads_post",
  unlist(lapply(resp_cols, function(m) paste0(m, c("_pre", "_post", "_change", "_pct_change",
                                                    "_direction", "_vs_control", "_rel_direction")))),
  "response_class", "response_score")
resp <- resp %>% select(all_of(ordered_cols))

write_csv(resp, file.path(out_dir, "mussel_response_classification.csv"), na = "")
message("Wrote mussel_response_classification.csv: ", nrow(resp), " animals with day-3 threads, ",
        sum(resp$paired), " paired.")

cat("\nResponse class by arm (paired animals):\n")
resp %>% filter(paired) %>% count(mussel_trt, response_class) %>%
  pivot_wider(names_from = response_class, values_from = n, values_fill = 0) %>%
  knitr::kable() %>% print()

cat("\nDirection by arm and metric (paired animals, n decreased / n paired):\n")
resp %>% filter(paired) %>%
  group_by(mussel_trt) %>%
  summarise(n = n(),
            adhesion  = sum(adhesion_kpa_direction == "decreased"),
            force     = sum(max_force_direction == "decreased"),
            area      = sum(pad_area_direction == "decreased"),
            extension = sum(max_displacement_direction == "decreased"),
            .groups = "drop") %>%
  knitr::kable() %>% print()
```

### Figure: every paired animal, change in force against change in plaque area
```{r fig.width=7.5, fig.height=6, warning=FALSE}
# The animal-level view of the dichotomy in the panel figure above. Bottom-left = smaller
# plaques holding less force; top-right = larger plaques holding more.
resp_p <- resp %>% filter(paired) %>%
  mutate(mussel_trt = factor(mussel_trt, levels = c("control", "OA", "OW", "DO")))

p_animal <- ggplot(resp_p, aes(x = 100 * (exp(pad_area_change) - 1),
                                y = 100 * (exp(max_force_change) - 1))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(fill = mussel_trt, shape = response_class), size = 3.6, stroke = 0.7,
             color = "black") +
  geom_text(aes(label = mussel), size = 2.4, vjust = -1.1, check_overlap = TRUE) +
  scale_fill_manual(values = arm_colors, name = "Arm",
                    guide = guide_legend(override.aes = list(shape = 21))) +
  scale_shape_manual(values = c(weaker = 25, stronger = 24, mixed = 21), name = "Response") +
  labs(x = "Plaque area, day 3 vs baseline (%)",
       y = "Peak force, day 3 vs baseline (%)") +
  my_theme + theme(legend.position = "right")
p_animal

ggsave(file.path(out_dir, "FIG_animal_response_force_vs_area.png"), plot = p_animal,
       dpi = 600, width = 7.5, height = 6, units = "in")
```

### Failure mode by thread treatment
```{r fig.width=8, fig.height=4.5, warning=FALSE}
fm <- thread_plot %>%
  filter(!is.na(failure)) %>%
  mutate(failure    = factor(failure, levels = c("cohesive", "peeling", "tearing", "thread")),
         thread_trt = factor(thread_trt, levels = c("lab_control", "baseline",
                                                    "treatment_control", "OA", "OW", "DO")))

fm_counts <- fm %>% count(thread_trt, failure) %>%
  group_by(thread_trt) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()
knitr::kable(fm_counts %>% select(-pct) %>%
               pivot_wider(names_from = failure, values_from = n, values_fill = 0))

# Does failure-mode composition differ among the four day-3 arms?
post_fm <- fm %>% filter(phase == "post") %>% droplevels()   # drop the empty pre/lab levels
fm_table <- table(post_fm$thread_trt, post_fm$failure)
chi <- suppressWarnings(chisq.test(fm_table))
cat("Chi-square, failure mode x day-3 arm: X2 =", round(chi$statistic, 2),
    " df =", chi$parameter, " p =", signif(chi$p.value, 3), "\n")
# Fisher's exact test (Monte Carlo, B draws) is always computed and written beside the
# chi-square, because several expected counts are below 5 and the chi-square approximation
# is the weaker of the two here. FISHER_B and the seed are recorded in the output row.
FISHER_B <- 10000
set.seed(20260916)
fisher_p <- fisher.test(fm_table, simulate.p.value = TRUE, B = FISHER_B)$p.value
cat("Fisher's exact (Monte Carlo, B =", FISHER_B, ") p =", signif(fisher_p, 3),
    if (any(chi$expected < 5)) " <- some expected counts < 5, prefer this p" else "", "\n")

# Adhesion by failure mode, for interpretation of the composition shift
adh_by_fm <- thread_plot %>%
  mutate(adhesion_kpa = max_force / pad_area * 1000) %>%
  filter(!is.na(failure), !is.na(adhesion_kpa)) %>%
  group_by(failure) %>%
  summarise(threads = n(), mean_kpa = mean(adhesion_kpa), median_kpa = median(adhesion_kpa),
            .groups = "drop")
knitr::kable(adh_by_fm, digits = 1)

write.csv(bind_rows(fm_counts %>% mutate(table = "counts_by_thread_trt"),
                    tibble(table = "chi_square_day3", statistic = unname(chi$statistic),
                           df = unname(chi$parameter), p = chi$p.value,
                           min_expected_count = min(chi$expected),
                           note = ifelse(any(chi$expected < 5),
                                         "expected counts < 5: read fisher_exact_day3", "")),
                    tibble(table = "fisher_exact_day3", statistic = NA_real_, df = NA_real_,
                           p = fisher_p, min_expected_count = min(chi$expected),
                           note = sprintf("Monte Carlo, B = %d, seed 20260916; %d threads, %d arms x %d modes",
                                          FISHER_B, sum(fm_table), nrow(fm_table), ncol(fm_table)))),
          file.path(out_dir, "STATS_failure_mode.csv"), row.names = FALSE)

failure_fill <- c("cohesive" = "#4C72B0", "peeling" = "#DD8452",
                  "tearing"  = "#8C8C8C", "thread"  = "#2F2F2F")
p_fm <- ggplot(fm_counts, aes(x = thread_trt, y = pct, fill = failure)) +
  geom_col(width = 0.7, color = "white", linewidth = 0.6) +
  geom_text(data = fm_counts %>% filter(pct >= 8),
            aes(label = paste0(round(pct), "%")),
            position = position_stack(vjust = 0.5), size = 3.2, color = "white") +
  scale_fill_manual(values = failure_fill, name = "Failure mode") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100.5)) +
  labs(x = NULL, y = "Threads (%)") +
  my_theme + theme(legend.position = "right",
                   axis.text.x = element_text(angle = 25, hjust = 1))
p_fm

ggsave(file.path(out_dir, "FIG_failure_mode_by_arm.png"), plot = p_fm,
       dpi = 600, width = 8, height = 4.5, units = "in")
```

### Provenance
```{r}
writeLines(c(paste0("generated : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
             paste0("input     : 03_analyses/thread-summary.xlsx"),
             paste0("threads in models : ", nrow(dat)),
             paste0("mussels in models : ", n_distinct(dat$mussel)),
             paste0("paired animals    : ", nrow(paired)),
             paste0("classified animals: ", nrow(resp), " (", sum(resp$paired), " paired)")),
           file.path(out_dir, "RUN_provenance.txt"))
print("done")
```
