---
title: "20 - Gene expression x byssal thread mechanics correlation (paired samples)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"   # "F" foot (default) or "G" gill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# Purpose

A subset of *M. trossulus* that were mechanically tested for byssal plaque attachment
were also sampled for Tag-seq. This script links those two datasets at the level of the
individual animal and asks: **does foot (or gill) gene expression at day 3 of stress
covary with plaque attachment performance?**

It produces:

1. A paired-sample manifest (exactly which animals/samples were used).
2. Per-gene association statistics: one baseline-adjusted regression (ANCOVA) of each
   animal's day-3 level of a mechanical metric on its expression, controlling for treatment
   and for the animal's own pre-exposure level, with multiple-testing correction.
3. Two gene universes: a biologically motivated **candidate set** (byssal / collagen /
   HSP / HIF / peroxidase / aminoacyl-tRNA-synthetase genes) and the **data-driven DEG
   union** (genes already found to respond to any stressor).
4. Diagnostic plots (top-hit scatterplots, candidate-gene correlation heatmap).

> **READ THIS FIRST - power.** The paired set is small: with the full thread dataset and the
> day-3 control arm included, n = 45 foot animals (control = 11, OA = 12, OW = 12, DO = 10),
> 44 of them with baseline threads (gill: 46 / 45), and the baseline-adjusted test uses only
> the animals with a baseline. Within any single arm n is 9-12. This analysis is
> **exploratory and hypothesis-generating, not confirmatory.** Treat raw p-values as ranking
> signals, expect few or no hits to survive FDR correction, and do not write causal language
> from it. Tag-seq sampling was destructive (foot/gill removed), so expression and the
> *post-stress* threads come from the same animal at the same timepoint; this is association,
> not a demonstrated mechanism.

> **The search is declared here, not in the write-up.** `METRICS` below names the metrics
> that form the confirmatory family (candidate genes x primary metrics); every other metric
> is labelled `exploratory` in every output table (`tier` column), and `q_family` is the BH
> correction over each gene set x tier family, so the primary family carries its own
> search-corrected q.

> **Which metric to read first.** On the full dataset the stressor effect is in **peak force
> and plaque area separately** more than in their ratio: on the log scale the arm x
> timepoint interaction is p = 2.9e-5 for force (post / baseline: control 0%, OA -33%,
> OW -55%, DO -46%) and p < 1e-10 for area (control +10%, OA +7%, OW -18%, DO -37%),
> against p = 0.0057 for adhesion, whose within-arm changes (control -9%, OA -36%, OW -45%,
> DO -14%) mix the two (`thread-strength/01_code/4_decompose_adhesion.Rmd`). A gene
> associated with plaque size is invisible against the ratio. `max_force` and `pad_area` are
> therefore the primary targets here; `adhesion_kpa` is retained as the material-strength
> readout. Every metric is tested as the animal's day-3 level with its own baseline level as
> a covariate, so "does expression track attachment" is asked net of where the animal
> started.

---

# 0. Configuration

Set these two paths to wherever the repos live on your machine. The script reads thread
data from the **main** repo and expression data from the **expression** repo.

```{r config}
## ---- analysis options ---------------------------------------------------------------
## Tissue comes from the knit parameter (`params$tissue`, default "F") so the same script
## serves foot and gill; the driver 24-run_gene_mechanics_by_tissue.Rmd renders both.
## When run chunk-by-chunk outside a knit, `params` may not exist; fall back to foot.
TISSUE <- if (exists("params") && !is.null(params$tissue)) toupper(params$tissue) else "F"
stopifnot(TISSUE %in% c("F", "G"))
MIN_COUNT     <- 5            # gene kept if count > MIN_COUNT in >= 1/3 of paired samples
FDR_ALPHA     <- 0.10         # BH threshold for flagging
## Detection floor. A gene whose VST value sits at the floor (zero counts) in many paired
## animals gives an association driven by presence/absence, not by expression level. Every
## tested gene gets frac_at_floor and floor_flag (exclude > FLOOR_EXCLUDE, caution >
## FLOOR_CAUTION); with EXCLUDE_FLOOR_GENES = TRUE the `exclude` genes are dropped from the
## candidate and DEG-union families BEFORE testing, so BH runs on the family that can be
## interpreted. They stay listed (floor_flag, tested = FALSE) and they still contribute to
## script 21's module scores.
FLOOR_EXCLUDE <- 0.40
FLOOR_CAUTION <- 0.20
EXCLUDE_FLOOR_GENES <- TRUE
INCLUDE_CONTROL_ARM <- TRUE   # TRUE = the day-3 common-garden control animals (T126-T137)
                              #        are part of the paired set, as a fourth treatment
                              #        level. They have foot RNA, day-3 threads and baseline
                              #        threads. Their own three-day trajectory (adhesion -9%,
                              #        force 0%, area +10%; thread-strength script 4) is the
                              #        null a stress-specific expression signal must beat.
                              # FALSE = stressor arms only.
USE_SECRETION_STATE <- FALSE  # TRUE = add each animal's byssal SECRETION STATE (on/off; the
                              #        foot plaque-protein module, computed per foot sample by
                              #        differential-expression/01_code/01_7-secretion_state.Rmd
                              #        and read from SECRETION_STATE_FILE) as a nuisance
                              #        covariate in every per-gene regression, beside
                              #        treatment. With 15 of 45 foot animals "on", any
                              #        candidate that co-varies with that module would
                              #        otherwise pick up secretion state, not mechanics.
                              #        Script 21 reads the same choice from metrics_config.
                              #        Gill animals take the state of their own foot sample.
                              # FALSE (default) = treatment is the only covariate; the state
                              #        is still joined into the manifest for inspection when
                              #        the file exists, but enters no model.

## ---- mechanical metrics and the reported test ----------------------------------------
## One test per gene x metric: an analysis of covariance of the animal's day-3 level on
## its expression, with the arm [and the secretion state] and the animal's OWN pre-exposure
## level as covariates:
##     level_day3 ~ expression + treatment [+ secretion_state] + level_baseline
## Force, area and adhesion are modelled on the LOG scale (per-animal geometric means of
## the plaques; multiplicative, right-skewed quantities), extension on its raw scale. Animals
## without baseline threads are not in the fits (44 of 45 foot, 45 of 46 gill).
## `tier`: force and area carry the stressor signal (thread-strength script 4), so they are
## the pre-declared confirmatory family (candidate genes x primary metrics); adhesion (their
## ratio) and extension are exploratory. Every output table carries the tier, and `q_family`
## corrects within each gene set x tier.
## Order here is the reporting order. Force and area first, deliberately (see above).
METRICS <- data.frame(
  metric = c("max_force",      "pad_area",           "adhesion_kpa",   "max_displacement"),
  scale  = c("log",            "log",                "log",            "raw"),
  tier   = c("primary",        "primary",            "exploratory",    "exploratory"),
  label  = c("Peak force (N)", "Plaque area (mm^2)", "Adhesion (kPa)", "Extension at break (mm)"),
  stringsAsFactors = FALSE)
metrics         <- METRICS$metric
PRIMARY_METRICS <- METRICS$metric[METRICS$tier == "primary"]
scale_of <- function(m) METRICS$scale[match(m, METRICS$metric)]
tier_of  <- function(m) METRICS$tier[match(m, METRICS$metric)]
label_of <- function(m) METRICS$label[match(m, METRICS$metric)]
on_scale <- function(m, v) if (scale_of(m) == "log") log(v) else v    # a value on its model scale

## ---- candidate gene keywords (regex, case-insensitive) ------------------------------
## Grouped by biology; edit freely. Matched against UniProt "protein_name".
CANDIDATE_KEYWORDS <- paste(
  # byssal structural / adhesive proteins (plaque proteins, thread core, cuticle)
  "byss", "foot protein", "Mefp", "adhesive plaque", "polyphenolic",
  "byssal protein-", "preCol", "precollagen", "byssal EP", "ACDC", "byssal tyrosinase",
  "byssal peroxidase", "byssal HSP", "byssal calumenin", "byssal protease inhibitor",
  # collagens / preCols (byssal thread core)
  "collagen",
  # plaque-curing / cross-linking enzymes
  "tyrosinase", "peroxidase", "amine oxidase", "catechol oxidase", "lysyl oxidase",
  # proteostasis / heat-shock
  "heat shock", "Hsp", "stress-70", "BiP", "chaperone",
  "endoplasmic reticulum chaperone", "hypoxia up-regulated",
  # hypoxia signalling
  "hypoxia-inducible", "Egl nine", "prolyl 4-hydroxylase", "prolyl hydroxylase",
  # translational machinery (foot-OA enrichment signal)
  "--tRNA", "tRNA ligase", "aminoacyl",
  # oxidative stress / detox
  "superoxide", "glutathione", "catalase", "peroxiredoxin", "dual oxidase",
  sep = "|"
)
```


```{r}
# Get the correct Bioconductor repository link for your R version
# bioc_version <- BiocManager::version()
# bioc_repo <- paste0("https://bioconductor.org/packages/", bioc_version, "/bioc")
# 
# # Install DESeq2 directly as a standard package 
# install.packages("GenomeInfoDbData", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))
# install.packages("DESeq2", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))

```




```{r libs}
## Load Packages
## DESeq2 first, tidyverse LAST: S4Vectors/IRanges (pulled in by DESeq2) mask dplyr's
## rename(), count(), first() and desc(). Loading tidyverse after them restores dplyr's.
load.lib <- c("DESeq2","readxl","openxlsx","here","tidyverse")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for(lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)


`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a
```

```{r paths}
## Resolve the repository root (the folder holding the analysis subfolders), robust to whether
## this is knit from gene-mechanics-correlation.Rproj or the top-level project.
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
repo_root   <- find_repo_root()
DE_DIR      <- file.path(repo_root, "differential-expression")          # expression-side inputs
## The curated thread table produced by thread-strength scripts 1-2 and hand-reviewed; and
## the per-animal response classification produced by thread-strength script 4.
THREAD_FILE   <- file.path(repo_root, "thread-strength", "03_analyses", "thread-summary.xlsx")
RESPONSE_FILE <- file.path(repo_root, "thread-strength", "03_analyses", "decompose-adhesion",
                           "mussel_response_classification.csv")
OUT_DIR     <- file.path(repo_root, "gene-mechanics-correlation", "03_analyses", "gene_mechanics")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
## Per-animal byssal secretion state (differential-expression script 01_7). Joined into the
## manifest whenever it exists; used as a covariate only when USE_SECRETION_STATE is TRUE.
SECRETION_STATE_FILE <- file.path(DE_DIR, "02_data", "secretion_state.csv")
## ---- annotation map and candidate universe ------------------------------------------
## CANDIDATE_ANNOTATION:
##   "genome" (default) = every gene in the count matrix is annotated with its best UniProt
##            hit (highest bitscore) from the genome-wide BLAST table MASTER_ANNOT_FILE, so
##            any EXPRESSED gene whose name matches CANDIDATE_KEYWORDS is a candidate.
##   "TC_DEG" = only genes that carry a name in the treatment-vs-control DEG tables
##            (GOterms_genome/*_sigs_ID.csv) can be candidates, plus the byssal structural
##            genes. That was the behaviour until 2026-09-18; it made "already a DEG in some
##            contrast" a hidden entry condition for the candidate family.
## The two sources agree on the name for 947 of the 969 DEG-annotated genes; the 22 that
## differ are LOCs whose transcripts hit different proteins, and the genome map takes the
## highest-scoring hit (the same rule script 23 uses).
## CANDIDATE_MAX_EVALUE / CANDIDATE_MIN_PIDENT: BLAST-quality floor for a gene to be a
## candidate under "genome" (the map itself keeps every hit, with pident and evalue columns).
## On this table every keyword match already has evalue < 1e-10; pident >= 40 would cut the
## foot candidates from ~226 to ~144 (script 23's HI_CONF_ONLY rule).
CANDIDATE_ANNOTATION <- "genome"
CANDIDATE_MAX_EVALUE <- 1e-10
CANDIDATE_MIN_PIDENT <- 0
MASTER_ANNOT_FILE <- file.path(repo_root, "blast", "03_analyses", "genome-foot", "LOC_GO_list.txt")
BYSSAL_STRUCTURAL_REGEX <- paste("foot protein", "adhesive plaque", "polyphenolic", "byssal protein-",
                                 "preCol", "precollagen", "byssal EP", "ACDC", "byssal tyrosinase",
                                 sep = "|")

if (!file.exists(THREAD_FILE))
  stop("Thread summary not found:\n  ", THREAD_FILE,
       "\nRun thread-strength scripts 1 and 2 and curate 03_analyses/thread-summary.xlsx.")

## Canonicalize a mussel tag to one 3-digit form (T55 -> T055) so the thread side and the
## expression side cannot split on zero-padding when joined.
norm_id <- function(x) {
  x   <- as.character(x)
  pre <- gsub("[0-9].*$", "", x)                                        # leading letters
  num <- suppressWarnings(as.integer(gsub("[^0-9].*$", "", gsub("^[^0-9]+", "", x))))
  ifelse(is.na(num), NA_character_, sprintf("%s%03d", pre, num))
}

## Read the curated thread table under either the current or the older column names, and
## make sure `phase` exists. thread_trt = what the THREAD was built in; mussel_trt = the arm
## the ANIMAL was assigned to; phase = lab / pre / post. The older `treatment == "control"`
## convention pooled baseline threads with the day-3 control arm and is not used here.
read_thread_summary <- function(path) {
  d <- readxl::read_excel(path, sheet = "data")
  ren <- c(mussel_ID = "mussel", thread_num = "thread", treatment = "thread_trt")
  for (from in names(ren)) {
    to <- ren[[from]]
    if (from %in% names(d) && !to %in% names(d)) names(d)[names(d) == from] <- to
  }
  need <- c("mussel", "thread_trt", "mussel_trt", "max_force", "pad_area")
  if (length(setdiff(need, names(d))) > 0)
    stop("thread-summary.xlsx is missing: ", paste(setdiff(need, names(d)), collapse = ", "))
  if (!"phase" %in% names(d))
    d$phase <- dplyr::case_when(d$thread_trt == "lab_control" ~ "lab",
                                d$thread_trt == "baseline"    ~ "pre",
                                TRUE                          ~ "post")
  if (!"max_displacement" %in% names(d)) d$max_displacement <- NA_real_
  d %>%
    mutate(mussel       = norm_id(mussel),
           adhesion_kpa = max_force / pad_area * 1000) %>%   # recomputed, never trusted from the sheet
    filter(!is.na(adhesion_kpa))
}
```

---

# 1. Build the paired-sample table

Each animal gets one value per metric and timepoint: the **geometric mean** of its plaques
for force, area and adhesion (the mean on the log scale the model uses) and the arithmetic
mean for extension; the same summary of its pre-exposure threads is its baseline. The
Tag-seq treatment assignment (the authoritative one) comes from the foot `treatmentinfo`
file. With `INCLUDE_CONTROL_ARM = TRUE` the day-3 control animals enter as their own
treatment level; the day-0 lab-reference animals never do (they have no day-3 threads).

```{r paired}
## ---- thread mechanics: one value per animal at day 3 and at baseline ------------------
thr_raw <- read_thread_summary(THREAD_FILE)

## per-animal summary on each metric's model scale: geometric mean for the log metrics
## (max_force N, pad_area mm^2, adhesion_kpa kPa), arithmetic mean for extension (mm)
summarise_animal <- function(d, prefix = "") {
  out <- d %>% group_by(mussel) %>% summarise(n = n(), .groups = "drop")
  for (m in metrics) {
    v <- d %>% group_by(mussel) %>%
      summarise(val = if (scale_of(m) == "log") exp(mean(log(.data[[m]]), na.rm = TRUE))
                      else mean(.data[[m]], na.rm = TRUE), .groups = "drop")
    out[[paste0(prefix, m)]] <- v$val[match(out$mussel, v$mussel)]
  }
  names(out)[names(out) == "n"] <- paste0("n_plaques", if (nzchar(prefix)) "_base" else "")
  out
}

arm_of <- thr_raw %>% filter(phase == "post") %>% distinct(mussel, treatment = mussel_trt)
stopifnot(!any(duplicated(arm_of$mussel)))                        # one arm per animal
post_stress <- thr_raw %>% filter(phase == "post") %>% summarise_animal() %>%
  left_join(arm_of, by = "mussel") %>% relocate(treatment, .after = mussel)

## pre-exposure baseline: the same animals' own threads before the exposure began
baseline <- thr_raw %>% filter(phase == "pre") %>% summarise_animal(prefix = "base_")

## ---- Tag-seq treatment assignment (definitive), for the chosen tissue ----------------
## The gill file codes the day-3 control arm as "control_3" and has no leading index column;
## read_treatmentinfo() normalises both so the two tissues look identical downstream.
read_treatmentinfo <- function(tissue) {
  f <- file.path(DE_DIR, "03_analyses", "DEG_lists",
                 ifelse(tissue == "F", "Foot", "Gill"),
                 sprintf("%s_treatmentinfo.csv", tissue))
  if (!file.exists(f)) stop("treatmentinfo not found: ", f)
  d <- read.table(f, header = TRUE, sep = " ", stringsAsFactors = FALSE)
  names(d)[1] <- "sample_id"
  d %>%
    mutate(sample_id = as.character(sample_id),
           mussel    = norm_id(sub(paste0(tissue, "$"), "", sample_id)),
           treatment = sub("_3$", "", treatment),          # control_3 -> control
           day       = as.integer(day)) %>%
    select(sample_id, mussel, treatment, day)
}
arms_kept <- if (INCLUDE_CONTROL_ARM) c("control", "OA", "OW", "DO") else c("OA", "OW", "DO")
fti <- read_treatmentinfo(TISSUE) %>%
  filter(day == 3, treatment %in% arms_kept) %>%       # day-3 animals only
  distinct(mussel, treatment)

## ---- assemble paired table ----------------------------------------------------------
paired <- post_stress %>%
  inner_join(fti, by = c("mussel", "treatment")) %>%   # the join also checks the two
  left_join(baseline, by = "mussel") %>%               # sources agree on the arm
  mutate(
    sample    = paste0(mussel, TISSUE),                # expression column name for chosen tissue
    treatment = factor(treatment, levels = arms_kept)
  )

## Arm disagreement between the thread key and the Tag-seq metadata would drop an animal
## silently at the inner_join above; make that loud.
arm_clash <- post_stress %>%
  inner_join(fti %>% dplyr::rename(rna_treatment = treatment), by = "mussel") %>%
  dplyr::filter(treatment != rna_treatment)
if (nrow(arm_clash) > 0) {
  warning(nrow(arm_clash), " animal(s) carry different arms in thread-summary and F_treatmentinfo; ",
          "excluded: ", paste(arm_clash$mussel, collapse = ", "))
  print(arm_clash)
}

## ---- per-animal response classification from thread-strength script 4 ----------------
## Carried in the manifest for inspection (which animals weakened relative to control);
## it enters no model here.
if (file.exists(RESPONSE_FILE)) {
  resp <- read_csv(RESPONSE_FILE, show_col_types = FALSE) %>%
    mutate(mussel = norm_id(mussel)) %>%
    select(mussel, response_class, response_score)
  paired <- paired %>% left_join(resp, by = "mussel")
} else {
  warning("Response classification not found (", RESPONSE_FILE,
          "); run thread-strength/01_code/4_decompose_adhesion.Rmd. Continuing without it.")
}

## An animal can be in treatmentinfo without a column in the count matrix (a library that
## was dropped at QC). Drop it here, loudly, rather than fail at the VST step.
cm_cols <- colnames(read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
                             nrows = 1, check.names = FALSE))
no_expr <- paired %>% filter(!sample %in% cm_cols)
if (nrow(no_expr) > 0) {
  message(nrow(no_expr), " paired animal(s) have no ", TISSUE,
          " column in the count matrix and are dropped: ",
          paste(no_expr$mussel, collapse = ", "))
  paired <- paired %>% filter(sample %in% cm_cols)
}

## ---- byssal secretion state (foot plaque-protein module, script 01_7) -----------------
## One state per ANIMAL, taken from its foot sample, so the gill analysis uses the same
## label. Joined for inspection whenever the file exists; a model covariate only when
## USE_SECRETION_STATE is TRUE.
if (file.exists(SECRETION_STATE_FILE)) {
  sec <- read_csv(SECRETION_STATE_FILE, show_col_types = FALSE) %>%
    mutate(mussel = norm_id(mussel)) %>%
    filter(!is.na(mussel)) %>%
    distinct(mussel, .keep_all = TRUE) %>%
    select(mussel, secretion_score, secretion_state)
  paired <- paired %>% left_join(sec, by = "mussel")
  cat("Secretion state joined:", sum(!is.na(paired$secretion_state)), "of", nrow(paired),
      "animals labelled (", sum(paired$secretion_state == "on", na.rm = TRUE), "on ).\n")
} else {
  paired$secretion_score <- NA_real_; paired$secretion_state <- NA_character_
  if (USE_SECRETION_STATE)
    stop("USE_SECRETION_STATE is TRUE but ", SECRETION_STATE_FILE, " does not exist; ",
         "run differential-expression/01_code/01_7-secretion_state.Rmd first.")
}
if (USE_SECRETION_STATE) {
  no_state <- paired %>% filter(is.na(secretion_state))
  if (nrow(no_state) > 0) {
    warning(nrow(no_state), " paired animal(s) have no secretion state and are dropped ",
            "because USE_SECRETION_STATE = TRUE: ", paste(no_state$mussel, collapse = ", "))
    paired <- paired %>% filter(!is.na(secretion_state))
  }
  paired <- paired %>% mutate(secretion_state = factor(secretion_state, levels = c("off", "on")))
  print(table(arm = paired$treatment, secretion_state = paired$secretion_state))
}
## the nuisance terms every regression below controls for; script 21 reads the same list
COVARIATES <- if (USE_SECRETION_STATE) c("treatment", "secretion_state") else "treatment"

cat("Paired animals:", nrow(paired), "\n")
print(table(paired$treatment))
cat("With baseline threads (the animals in every ANCOVA fit):",
    sum(!is.na(paired$base_max_force)), "\n")
print(table(paired$treatment[!is.na(paired$base_max_force)]))
if ("response_class" %in% names(paired)) print(table(paired$treatment, paired$response_class))
write_csv(paired, file.path(OUT_DIR, sprintf("paired_sample_manifest_%s.csv", TISSUE)))
```

---

# 2. Load counts and VST-normalise

Variance-stabilising transformation is estimated across **all** samples of the chosen
tissue (more stable dispersion estimates), then we subset to the paired animals. Gene IDs
in the matrix look like `gene-LOC134696364|LOC134696364`; we key on the `LOC######` id.

```{r counts}
cm <- read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
               row.names = 1, check.names = FALSE)

## keep only columns for the chosen tissue (F = ...F, but NOT ...FX; G = ...G)
tissue_cols <- if (TISSUE == "F") {
  grep("F$", colnames(cm), value = TRUE)        # excludes FX (ends in X) and G
} else {
  grep("G$", colnames(cm), value = TRUE)
}
cm_tissue <- cm[, tissue_cols, drop = FALSE]

## minimal colData (treatment per sample, for dispersion modelling)
mussel_of  <- norm_id(sub(paste0(TISSUE, "$"), "", colnames(cm_tissue)))
all_meta <- read_treatmentinfo(TISSUE)
coldata <- data.frame(mussel = mussel_of) %>%
  left_join(all_meta %>% distinct(mussel, treatment, day), by = "mussel") %>%
  mutate(grp = ifelse(is.na(treatment), "unk", paste0(treatment, "_d", day)))
rownames(coldata) <- colnames(cm_tissue)
coldata$grp <- factor(coldata$grp)

## DESeq2 + VST
dds <- DESeqDataSetFromMatrix(round(as.matrix(cm_tissue)), coldata, design = ~ 1)
dds <- dds[rowSums(counts(dds)) > 0, ]
vsd  <- vst(dds, blind = TRUE)
vmat <- assay(vsd)                        # rownames = full gene IDs, same order as counts(dds)
rownames(vmat) <- sub(".*\\|", "", sub("^gene-", "", rownames(vmat)))  # -> LOC key

stopifnot(all(paired$sample %in% colnames(vmat)))
vp       <- vmat[,        paired$sample, drop = FALSE]
counts_p <- counts(dds)[, paired$sample, drop = FALSE]   # identical row order to vp
keep <- rowSums(counts_p > MIN_COUNT) >= ceiling(nrow(paired) / 3)
vp <- vp[keep, , drop = FALSE]
cat("Genes tested (after expression filter):", nrow(vp), "\n")
cat("Duplicate LOC keys among tested genes:", sum(duplicated(rownames(vp))), "\n")  # should be ~0

## detection floor per gene: fraction of paired animals at the VST minimum (zero counts)
floor_val   <- min(vp)
frac_floor  <- rowMeans(abs(vp - floor_val) < 1e-6)
floor_flag_of <- function(ids) {
  fr <- unname(frac_floor[match(ids, rownames(vp))])
  tibble(LOC_ID = ids, frac_at_floor = round(fr, 3),
         floor_flag = dplyr::case_when(fr > FLOOR_EXCLUDE ~ "exclude", fr > FLOOR_CAUTION ~ "caution", TRUE ~ "ok"),
         tested = !(EXCLUDE_FLOOR_GENES & fr > FLOOR_EXCLUDE))
}
```

---

# 3. Annotation map (portable LOC -> protein name)

The map is the genome-wide BLAST table, best UniProt hit per LOC by bitscore
(`CANDIDATE_ANNOTATION = "genome"`), with the hit's `pident` and `evalue` carried along so a
weak annotation is visible wherever a name is printed. The treatment-vs-control DEG tables
(`GOterms_genome/*_sigs_ID.csv`) are read as well: `in_TC_DEG_annotation` marks the genes
they name, and with `CANDIDATE_ANNOTATION = "TC_DEG"` only those (plus the byssal structural
genes) can enter the candidate set.

```{r annot}
## names carried by the TC DEG tables (the pre-2026-09-18 universe)
sig_files <- list.files(file.path(DE_DIR, "03_analyses", "DEG_lists", "GOterms_genome"),
                        pattern = "_sigs_ID\\.csv$", full.names = TRUE)
annot_deg <- map_dfr(sig_files, ~ {
  d <- tryCatch(read.csv(.x, stringsAsFactors = FALSE), error = function(e) NULL)
  if (is.null(d) || !all(c("LOC_ID", "protein_name") %in% names(d))) return(NULL)
  d %>% select(LOC_ID, protein_name)
}) %>% distinct(LOC_ID, .keep_all = TRUE) %>% as_tibble()
cat("LOCs named in the TC DEG tables:", nrow(annot_deg), "\n")

## genome-wide map: best hit per LOC
if (!file.exists(MASTER_ANNOT_FILE))
  stop("Master annotation not found: ", MASTER_ANNOT_FILE,
       "\nIt is the source of the annotation map (CANDIDATE_ANNOTATION = \"genome\").")
master <- read_delim(MASTER_ANNOT_FILE, delim = "\t", show_col_types = FALSE,
                     col_select = c("gene", "pident", "evalue", "bitscore", "Protein.names")) %>%
  ## key: the LOC id, or the gene symbol for the 13 mitochondrial genes (ND1, COX1, ...),
  ## which is how the count matrix names them too
  transmute(LOC_ID = dplyr::coalesce(str_extract(as.character(gene), "LOC\\d+"),
                                     na_if(trimws(as.character(gene)), "")),
            protein_name = as.character(Protein.names),
            blast_pident = suppressWarnings(as.numeric(pident)),
            blast_evalue = suppressWarnings(as.numeric(evalue)),
            bitscore     = suppressWarnings(as.numeric(bitscore))) %>%
  filter(!is.na(LOC_ID), !is.na(protein_name), nzchar(protein_name)) %>%
  arrange(desc(bitscore)) %>% distinct(LOC_ID, .keep_all = TRUE) %>%           # best hit per LOC
  select(-bitscore)
cat("LOCs with a genome-wide best hit:", nrow(master), "\n")

if (CANDIDATE_ANNOTATION == "genome") {
  annot <- master %>%
    mutate(annotation_source = "genome_blast",
           in_TC_DEG_annotation = LOC_ID %in% annot_deg$LOC_ID)
  cand_ok <- with(annot, blast_evalue <= CANDIDATE_MAX_EVALUE & blast_pident >= CANDIDATE_MIN_PIDENT)
} else if (CANDIDATE_ANNOTATION == "TC_DEG") {
  ## the DEG-table names, plus the byssal structural genes from the genome map (they are not
  ## DEGs in any treatment-vs-control contrast, so the DEG tables never reach them)
  byssal_add <- master %>%
    filter(grepl(BYSSAL_STRUCTURAL_REGEX, protein_name, ignore.case = TRUE),
           LOC_ID %in% rownames(vp), !LOC_ID %in% annot_deg$LOC_ID) %>%
    mutate(annotation_source = "genome_blast_byssal")
  annot <- bind_rows(annot_deg %>% mutate(annotation_source = "TC_DEG_sigs") %>%
                       left_join(master %>% select(LOC_ID, blast_pident, blast_evalue), by = "LOC_ID"),
                     byssal_add) %>%
    mutate(in_TC_DEG_annotation = annotation_source == "TC_DEG_sigs")
  cand_ok <- rep(TRUE, nrow(annot))
} else stop("CANDIDATE_ANNOTATION must be \"genome\" or \"TC_DEG\"")
annot$blast_ok <- !is.na(cand_ok) & cand_ok   # passes the BLAST-quality floor; script 21's modules use it too
cat("Annotation map (", CANDIDATE_ANNOTATION, "):", nrow(annot), "LOCs, of which", sum(annot$blast_ok),
    "pass the BLAST-quality floor\n")

## candidate gene set = annotated, expressed genes whose protein name matches the keyword list
candidates <- annot %>% filter(blast_ok) %>%
  filter(grepl(CANDIDATE_KEYWORDS, protein_name, ignore.case = TRUE)) %>%
  filter(LOC_ID %in% rownames(vp)) %>%
  mutate(byssal_structural = grepl(BYSSAL_STRUCTURAL_REGEX, protein_name, ignore.case = TRUE))
candidates <- candidates %>% left_join(floor_flag_of(candidates$LOC_ID), by = "LOC_ID")
cat("Candidate (byssal/stress) genes present & expressed:", nrow(candidates),
    "( byssal structural:", sum(candidates$byssal_structural),
    "; named in the TC DEG tables:", sum(candidates$in_TC_DEG_annotation),
    "; at the detection floor:", sum(candidates$floor_flag == "exclude"),
    if (EXCLUDE_FLOOR_GENES) "excluded from the tested family" else "kept", ")\n")
cat("Candidate genes tested:", sum(candidates$tested), "\n")
print(as.data.frame(candidates %>% select(LOC_ID, protein_name, blast_pident, in_TC_DEG_annotation,
                                          byssal_structural, frac_at_floor, floor_flag, tested) %>%
                      mutate(protein_name = substr(protein_name, 1, 50))))
```

```{r exports}
## ---- handoff CSVs for downstream scripts (21 reads all; 22 reads annotation_map) -----
## paired_sample_manifest_<T>.csv was written above; these add the rest so 21 can run on its
## own instead of sharing this R session's objects.
write_csv(data.frame(LOC_ID = rownames(vp), vp, check.names = FALSE),
          file.path(OUT_DIR, sprintf("vst_paired_%s.csv", TISSUE)))            # genes x paired samples (VST)
write_csv(annot,      file.path(OUT_DIR, "annotation_map.csv"))                # LOC_ID -> protein_name
write_csv(candidates, file.path(OUT_DIR, sprintf("candidate_genes_%s.csv", TISSUE)))  # candidate set
write_csv(thr_raw %>% filter(phase == "post", mussel %in% paired$mussel) %>%
            transmute(mussel, treatment = mussel_trt, phase, thread_trt,
                      max_force, pad_area, adhesion_kpa, max_displacement),
          file.path(OUT_DIR, sprintf("thread_plaques_paired_%s.csv", TISSUE)))  # raw day-3 plaques, paired animals
write_csv(METRICS %>%                                                            # so 21 fits the same
            mutate(arms                = paste(arms_kept, collapse = ";"),        # model on the same
                   covariates          = paste(COVARIATES, collapse = ";"),       # metrics and scales
                   use_secretion_state = USE_SECRETION_STATE,
                   exclude_floor_genes = EXCLUDE_FLOOR_GENES),
          file.path(OUT_DIR, sprintf("metrics_config_%s.csv", TISSUE)))
```

---

# 4. DEG union (data-driven gene universe)

Union of significant genes from all stressor contrasts for this tissue (treatment-control
baseline). These are genes that already responded to stress; the question is whether,
among the paired animals, their expression *level* tracks attachment.

```{r deg_union}
deg_dir <- file.path(DE_DIR, "03_analyses", "DEG_lists", ifelse(TISSUE == "F", "Foot", "Gill"))
deg_files <- list.files(deg_dir, pattern = "_TC_siggene.*\\.csv$", full.names = TRUE)
deg_union <- map(deg_files, ~ {
  d <- tryCatch(read.table(.x, header = TRUE, sep = " ", stringsAsFactors = FALSE),
                error = function(e) NULL)
  if (is.null(d)) return(character(0))
  g <- d[[grep("gene", names(d), ignore.case = TRUE)[1]]]
  sub(".*\\|", "", sub("^gene-", "", g))
}) %>% unlist() %>% unique()
deg_union <- intersect(deg_union, rownames(vp))
deg_union_tbl <- floor_flag_of(deg_union)
cat("DEG-union genes present & expressed:", length(deg_union),
    "| at the detection floor:", sum(deg_union_tbl$floor_flag == "exclude"),
    "| tested:", sum(deg_union_tbl$tested), "\n")

## the floor table for both universes, so downstream scripts and the report can read it
write_csv(bind_rows(candidates %>% transmute(LOC_ID, gene_set = "candidate", frac_at_floor, floor_flag, tested),
                    deg_union_tbl %>% filter(!LOC_ID %in% candidates$LOC_ID) %>%
                      transmute(LOC_ID, gene_set = "DEG_union", frac_at_floor, floor_flag, tested)) %>%
            left_join(annot %>% select(LOC_ID, protein_name), by = "LOC_ID") %>%
            mutate(n_samples = nrow(paired), n_at_floor = round(frac_at_floor * nrow(paired))) %>%
            select(LOC_ID, protein_name, gene_set, n_samples, n_at_floor, frac_at_floor, floor_flag, tested) %>%
            arrange(desc(frac_at_floor), LOC_ID),
          file.path(OUT_DIR, sprintf("detection_floor_flags_%s.csv", TISSUE)))
```

---

# 5. Association testing

One reported test per gene x metric, an analysis of covariance on the animal's day-3 level:

    level_day3 ~ expression + treatment [+ secretion_state] + level_baseline

Force, area and adhesion enter as log(per-animal geometric mean) on both sides of the
model, extension on its raw scale (`scale` column). `slope` is the coefficient on VST
expression with the arm and the animal's own baseline held constant and `p_lm` its t-test;
`baseline_slope` is the coefficient on the baseline. Beside the parametric slope,
`rho_partial` is the Spearman correlation of expression and day-3 level with the same
nuisance terms (arm, state, baseline) partialled out of the ranks, computed on exactly the
animals the model used. BH correction is applied **within each gene set x metric**
(`q_lm`) and **within each gene set x tier** (`q_family`); the candidate x primary family is
the declared confirmatory one.

```{r assoc_fns}
partial_spearman <- function(x, y, nuis) {
  # rank-based partial correlation: the nuisance columns (factors and numerics) are
  # regressed out of the ranks of x and of y, complete cases only
  ok <- stats::complete.cases(x, y, nuis)
  if (sum(ok) < 6) return(NA_real_)
  X  <- model.matrix(reformulate(names(nuis)), data = droplevels(nuis[ok, , drop = FALSE]))
  rx <- residuals(lm(rank(x[ok]) ~ X - 1))
  ry <- residuals(lm(rank(y[ok]) ~ X - 1))
  suppressWarnings(cor(rx, ry, method = "pearson"))   # corr of ranks' residuals
}

## the ANCOVA data for one gene x metric: response and baseline on the metric's model scale,
## complete cases only (animals without baseline threads drop here)
ancova_data <- function(loc, metric) {
  dat <- data.frame(y    = on_scale(metric, paired[[metric]]),
                    x    = as.numeric(vp[match(loc, rownames(vp)), paired$sample]),
                    base = on_scale(metric, paired[[paste0("base_", metric)]]),
                    paired[, COVARIATES, drop = FALSE])
  dat[stats::complete.cases(dat), , drop = FALSE]
}
ANCOVA_FORMULA <- reformulate(c("x", COVARIATES, "base"), response = "y")
NUISANCE_TERMS <- c(COVARIATES, "base")

test_gene <- function(loc, metric) {
  dat <- ancova_data(loc, metric)
  if (nrow(dat) < 6) return(NULL)
  fm <- tryCatch(lm(ANCOVA_FORMULA, data = dat), error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!all(c("x", "base") %in% rownames(co))) return(NULL)
  tibble(
    LOC_ID         = loc,
    metric         = metric,
    scale          = scale_of(metric),
    tier           = tier_of(metric),
    n              = nrow(dat),
    slope          = co["x", "Estimate"],
    se             = co["x", "Std. Error"],
    p_lm           = co["x", "Pr(>|t|)"],
    baseline_slope = co["base", "Estimate"],
    rho_partial    = partial_spearman(dat$x, dat$y, dat[, NUISANCE_TERMS, drop = FALSE])
  )
}

run_set <- function(loci, set_name) {
  res <- map_dfr(metrics, function(mt) map_dfr(loci, ~ test_gene(.x, mt)))
  res %>%
    group_by(metric) %>% mutate(q_lm = p.adjust(p_lm, "BH")) %>% ungroup() %>%      # within metric
    group_by(tier)   %>% mutate(q_family = p.adjust(p_lm, "BH")) %>% ungroup() %>%  # within tier family
    left_join(annot, by = "LOC_ID") %>%
    mutate(gene_set = set_name) %>%
    arrange(factor(metric, levels = metrics), p_lm)
}
```

```{r run}
res_cand <- run_set(candidates$LOC_ID[candidates$tested], "candidate") %>%
  left_join(candidates %>% select(LOC_ID, frac_at_floor, floor_flag, byssal_structural), by = "LOC_ID")
res_deg  <- run_set(deg_union_tbl$LOC_ID[deg_union_tbl$tested], "DEG_union") %>%
  left_join(deg_union_tbl %>% select(LOC_ID, frac_at_floor, floor_flag), by = "LOC_ID")

write_csv(res_cand, file.path(OUT_DIR, sprintf("assoc_candidate_%s.csv", TISSUE)))
write_csv(res_deg,  file.path(OUT_DIR, sprintf("assoc_DEGunion_%s.csv",  TISSUE)))

cat("\n================ CANDIDATE GENES: top associations per metric ================\n")
res_cand %>%
  mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_lm, n = 6) %>% ungroup() %>%
  mutate(across(c(slope, baseline_slope, rho_partial), ~round(., 3)),
         across(c(p_lm, q_lm, q_family), ~signif(., 2))) %>%
  select(metric, tier, n, LOC_ID, protein_name, slope, p_lm, q_lm, q_family, rho_partial) %>%
  print(n = 24)

## the declared confirmatory family: candidate genes x primary metrics
fam <- res_cand %>% filter(tier == "primary")
cat(sprintf("\nCandidate x primary family (%s): %d tests, %d with q_family < %s; smallest p = %s\n",
            paste(PRIMARY_METRICS, collapse = ", "), nrow(fam), sum(fam$q_family < FDR_ALPHA),
            FDR_ALPHA, signif(min(fam$p_lm), 2)))

cat("\n================ DEG-UNION: genes passing BH <", FDR_ALPHA, "================\n")
hits <- res_deg %>% filter(q_lm < FDR_ALPHA) %>% arrange(factor(metric, levels = metrics), p_lm)
if (nrow(hits) == 0) {
  cat("None survive FDR at this sample size (expected). Inspect raw-p ranking below.\n")
  res_deg %>% mutate(metric = factor(metric, levels = metrics)) %>%
    group_by(metric) %>% slice_min(p_lm, n = 5) %>% ungroup() %>%
    mutate(across(c(slope, rho_partial), ~round(., 3)), across(c(p_lm, q_lm, q_family), ~signif(., 2))) %>%
    select(metric, LOC_ID, protein_name, slope, p_lm, q_lm, q_family, rho_partial) %>%
    as.data.frame() %>% head(24) %>% print()
} else {
  print(hits, n = nrow(hits))
}
```

---

# 6. Plots

## 6.1 Candidate-gene correlation heatmap

Partial Spearman (arm and baseline partialled out) of candidate genes against each
metric's day-3 level. With the genome-wide candidate set (~230 genes) the panel shows the
`HEATMAP_MAX_GENES` genes with the strongest association on any metric (smallest `p_lm`);
the full table is `assoc_candidate_<T>.csv`.

```{r heatmap, fig.height=7}
HEATMAP_MAX_GENES <- 60
hm_keep <- res_cand %>% group_by(LOC_ID) %>% summarise(best_p = min(p_lm), .groups = "drop") %>%
  slice_min(best_p, n = HEATMAP_MAX_GENES, with_ties = FALSE) %>% pull(LOC_ID)
hm <- res_cand %>%
  filter(LOC_ID %in% hm_keep) %>%
  mutate(label = paste0(substr(protein_name, 1, 34), " (", LOC_ID, ")")) %>%
  select(label, metric, rho_partial) %>%
  pivot_wider(names_from = metric, values_from = rho_partial) %>%
  arrange(desc(abs(max_force)))

hm_long <- hm %>% pivot_longer(-label, names_to = "metric", values_to = "rho") %>%
  mutate(metric = factor(metric, levels = metrics))
short_label <- function(m) sub(" \\(.*\\)$", "", label_of(m))    # "Peak force (N)" -> "Peak force"
ggplot(hm_long, aes(metric, factor(label, levels = rev(hm$label)), fill = rho)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1)) +
  scale_x_discrete(labels = short_label, position = "top") +
  labs(x = NULL, y = NULL, fill = "partial\nSpearman",
       title = "Candidate gene expression vs plaque mechanics",
       subtitle = sprintf("%s, n = %d animals with baseline; top %d of %d candidate genes by smallest p",
                          ifelse(TISSUE == "F", "foot", "gill"), sum(!is.na(paired$base_max_force)),
                          nrow(hm), nrow(candidates))) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 9, angle = 20, hjust = 0),
        plot.title.position = "plot")
ggsave(file.path(OUT_DIR, sprintf("candidate_heatmap_%s.png", TISSUE)),
       width = 7.5, height = 8.5, dpi = 150, bg = "white")   # theme_minimal has no background: force one
```

## 6.2 Scatterplots for the strongest candidate hits (vs peak force)

The y axis is the day-3 level on the model scale, adjusted to the mean baseline: the
residual of `log(level) ~ log(baseline)` plus the grand mean. The reported slopes come from
the full ANCOVA (arm and baseline together); this is the display version of the same
adjustment, kept free of the arm so that colouring by arm still shows where each arm sits.

```{r scatter, fig.height=8}
## day-3 level adjusted to the mean baseline, on the metric's model scale (NA without baseline)
adjusted_level <- function(metric) {
  d  <- data.frame(y = on_scale(metric, paired[[metric]]),
                   base = on_scale(metric, paired[[paste0("base_", metric)]]))
  ok <- stats::complete.cases(d)
  out <- rep(NA_real_, nrow(d))
  if (sum(ok) >= 6) out[ok] <- mean(d$y[ok]) + residuals(lm(y ~ base, data = d[ok, ]))
  out
}
adjusted_label <- function(metric)
  sprintf("%s%s, baseline-adjusted", if (scale_of(metric) == "log") "log " else "", label_of(metric))

top_loci <- res_cand %>% filter(metric == "max_force") %>%
  slice_min(p_lm, n = 6, with_ties = FALSE) %>% pull(LOC_ID)

plot_df <- map_dfr(top_loci, function(loc) {
  tibble(LOC_ID = loc,
         protein = annot$protein_name[match(loc, annot$LOC_ID)],
         expr = as.numeric(vp[match(loc, rownames(vp)), paired$sample]),
         y = adjusted_level("max_force"),
         treatment = paired$treatment)
}) %>% filter(!is.na(y)) %>%
  mutate(facet = factor(paste0(substr(protein, 1, 30), "\n", LOC_ID),
                        levels = unique(paste0(substr(protein, 1, 30), "\n", LOC_ID))))

arm_cols <- c(control = "#7FB3D5", OA = "#4CAF50", OW = "#fdae61", DO = "#9884a3")   # shared by 6.2 and 6.3
ggplot(plot_df, aes(expr, y)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linewidth = 0.6) +
  geom_point(aes(fill = treatment), shape = 21, size = 2.4, stroke = 0.5, color = "black") +
  facet_wrap(~ facet, scales = "free_x") +
  scale_fill_manual(values = arm_cols, name = "Arm") +
  labs(x = "VST expression", y = adjusted_label("max_force"),
       title = "Top candidate genes vs peak force (line = pooled fit)") +
  theme_bw(base_size = 11)
ggsave(file.path(OUT_DIR, sprintf("top_candidate_scatter_%s.png", TISSUE)),
       width = 9, height = 7, dpi = 150, bg = "white")
```

## 6.3 The strongest candidate hit for each metric, coloured by arm

One panel per metric, the single best candidate gene for that metric, y baseline-adjusted
as above. Colouring by arm is the check that matters: an association driven by one arm
sitting at one end of both axes is an arm effect, not a gene effect.

```{r scatter_best, fig.height=8}
best_per_metric <- res_cand %>%
  mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  arrange(metric)

best_df <- pmap_dfr(best_per_metric %>% select(LOC_ID, metric, p_lm, q_lm, q_family),
                    function(LOC_ID, metric, p_lm, q_lm, q_family) {
  m <- as.character(metric)
  tibble(metric = m,
         panel  = sprintf("%s (%s)\n%s\np = %s, q = %s, q_family = %s", m, tier_of(m),
                          substr(annot$protein_name[match(LOC_ID, annot$LOC_ID)], 1, 34),
                          signif(p_lm, 2), signif(q_lm, 2), signif(q_family, 2)),
         expr   = as.numeric(vp[match(LOC_ID, rownames(vp)), paired$sample]),
         y      = adjusted_level(m),
         treatment = paired$treatment)
}) %>% filter(!is.na(y)) %>%
  mutate(panel = factor(panel, levels = unique(panel)))

ggplot(best_df, aes(expr, y)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linewidth = 0.6) +
  geom_point(aes(fill = treatment), shape = 21, size = 2.6, stroke = 0.5, color = "black") +
  facet_wrap(~ panel, scales = "free", ncol = 2) +
  scale_fill_manual(values = arm_cols, name = "Arm") +
  labs(x = "VST expression", y = "Day-3 level on the model scale, baseline-adjusted",
       title = sprintf("Best candidate gene per metric (%s, n = %d with baseline)",
                       ifelse(TISSUE == "F", "foot", "gill"), sum(!is.na(paired$base_max_force)))) +
  theme_bw(base_size = 10) + theme(strip.text = element_text(size = 7.5))
ggsave(file.path(OUT_DIR, sprintf("best_hit_per_metric_scatter_%s.png", TISSUE)),
       width = 9, height = 8.5, dpi = 150, bg = "white")
```

---

# 7. Session info & interpretation notes

```{r notes, echo=FALSE, results='asis'}
cat("
**How to read the outputs**

- `assoc_candidate_*.csv` / `assoc_DEGunion_*.csv`: one row per gene x metric from the
  ANCOVA `level_day3 ~ expression + treatment [+ secretion_state] + level_baseline`.
  `slope` (with `se`) is the change in the day-3 level per unit VST expression with the arm
  and the animal's own baseline held constant: log units for force, area and adhesion, so
  exp(slope) is the multiplicative change per VST unit; mm for extension. `p_lm` is its
  t-test; `q_lm` is BH within the gene set x metric and `q_family` BH within the gene set x
  tier (candidate x primary is the declared confirmatory family); `baseline_slope` is the
  coefficient on the baseline level; `rho_partial` is the Spearman correlation with the
  same nuisance terms partialled out of the ranks.
- A negative slope/rho means higher expression goes with *weaker* attachment.

**Caveats baked into the design (do not skip in the manuscript):**

1. n = 44 foot / 45 gill animals are in the fits (the paired animals with baseline threads;
   control 11 / OA 12 / OW 12 / DO 9-10). Within-arm correlations are not interpretable on
   their own; the pooled-with-covariate model borrows strength across arms but confounds
   arm-level expression shifts with attachment differences.
2. Multiple testing: the candidate set (~230 foot / ~280 gill genes under the genome-wide
   map) x PRIMARY_METRICS is the declared confirmatory family (`tier = primary`,
   `q_family`); the remaining metrics and the DEG-union (hundreds) are exploratory /
   screening. Quote q beside p, never p alone.
3. Baseline adjustment: the covariate is the animal's own pre-exposure level, the same
   summary on the same scale. It absorbs the between-animal baseline differences that the
   arm assignment did not balance (thread-strength script 3, 3b: baseline area and
   extension differ by future arm) without adding the baseline's measurement noise to the
   response. The one animal per tissue without baseline threads is not in the fits.
4. `USE_SECRETION_STATE` (default FALSE) adds the animal's byssal secretion state (on/off,
   from differential-expression script 01_7) as a nuisance covariate. It is off by default
   because the state is itself an expression-derived quantity measured on the same sample;
   turn it on as a sensitivity analysis when a candidate co-varies with the plaque-protein
   module (`byssal_structural = TRUE` in candidate_genes_<T>.csv).
5. Destructive sampling: expression and post-stress threads are the same animal/timepoint,
   so this is contemporaneous association, not a predictive or causal link.
6. `pad_area` and `max_force` carry the stressor signal more cleanly than `adhesion_kpa`
   (their ratio): on this dataset the arm x timepoint interaction is p = 2.9e-5 for force
   and p < 1e-10 for area against p = 0.0057 for adhesion, and the area response (larger
   plaques under OA, smaller under OW and DO) is invisible in the ratio. Read force and
   area first.
7. With `INCLUDE_CONTROL_ARM = TRUE` the control animals contribute a fourth treatment level.
   An expression-mechanics association that holds within the control arm as well is about
   attachment biology in general, not about the stress response; that distinction is worth
   checking by hand for any top hit (scatter by arm).
")
sessionInfo()
```

# 8. Provenance

`RUN_provenance_<T>.txt` records the configuration behind every file in
`03_analyses/gene_mechanics/` (the thread-strength scripts keep the same kind of file), and
the git commit that last touched the thread input, so a result can always be traced to the
thread table it was computed from. Script 21 appends its own block.

```{r provenance}
git_last_commit <- function(path) {
  ## guarded: no git, not a repo, or an untracked file all yield "unavailable"
  if (!nzchar(Sys.which("git")) || !file.exists(path)) return("unavailable")
  out <- tryCatch(suppressWarnings(
           system2("git", c("-C", shQuote(dirname(path)), "log", "-n", "1",
                            shQuote("--format=%h %cs"), "--", shQuote(basename(path))),
                   stdout = TRUE, stderr = FALSE)),
         error = function(e) character(0))
  if (length(out) == 0 || !is.null(attr(out, "status")) || !nzchar(out[1])) "unavailable" else out[1]
}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
writeLines(c(
  "[20-gene_mechanics_correlation.Rmd]",
  paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
  paste0("tissue               : ", TISSUE),
  paste0("thread input         : ", THREAD_FILE),
  paste0("thread input commit  : ", git_last_commit(THREAD_FILE)),
  paste0("response input       : ", RESPONSE_FILE, " (", ifelse(file.exists(RESPONSE_FILE), "present", "absent"), ")"),
  paste0("count matrix         : ", file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv")),
  paste0("INCLUDE_CONTROL_ARM  : ", INCLUDE_CONTROL_ARM),
  paste0("USE_SECRETION_STATE  : ", USE_SECRETION_STATE),
  paste0("reported test        : ANCOVA level_day3 ~ expression + ", paste(COVARIATES, collapse = " + "),
         " + level_baseline (per-animal geometric mean, log scale; extension raw)"),
  paste0("metrics [scale, tier]: ", paste(sprintf("%s [%s, %s]", METRICS$metric, METRICS$scale, METRICS$tier),
                                          collapse = ", ")),
  paste0("PRIMARY_METRICS      : ", paste(PRIMARY_METRICS, collapse = ", ")),
  paste0("MIN_COUNT / FDR_ALPHA: ", MIN_COUNT, " / ", FDR_ALPHA),
  paste0("paired animals       : ", nrow(paired), " (", paste(sprintf("%s %d", names(table(paired$treatment)),
                                                                     as.integer(table(paired$treatment))), collapse = ", "), ")"),
  paste0("with baseline (in fits): ", sum(!is.na(paired$base_max_force)), " (",
         paste(sprintf("%s %d", names(table(paired$treatment[!is.na(paired$base_max_force)])),
                       as.integer(table(paired$treatment[!is.na(paired$base_max_force)]))), collapse = ", "), ")"),
  paste0("genes tested         : ", nrow(vp)),
  paste0("CANDIDATE_ANNOTATION : ", CANDIDATE_ANNOTATION, " (evalue <= ", CANDIDATE_MAX_EVALUE,
         ", pident >= ", CANDIDATE_MIN_PIDENT, ")"),
  paste0("annotation map       : ", nrow(annot), " LOCs (", MASTER_ANNOT_FILE, ")"),
  paste0("candidate genes      : ", nrow(candidates), " (byssal structural ", sum(candidates$byssal_structural),
         "; named in TC DEG tables ", sum(candidates$in_TC_DEG_annotation), ")"),
  paste0("EXCLUDE_FLOOR_GENES  : ", EXCLUDE_FLOOR_GENES, " (exclude > ", FLOOR_EXCLUDE, ", caution > ", FLOOR_CAUTION, ")"),
  paste0("candidates tested    : ", sum(candidates$tested), " (", sum(!candidates$tested), " at the detection floor)"),
  paste0("DEG union tested     : ", sum(deg_union_tbl$tested), " of ", nrow(deg_union_tbl)),
  paste0("DEG union genes      : ", length(deg_union)),
  paste0("R / DESeq2           : ", R.version.string, " / ", as.character(packageVersion("DESeq2")))),
  prov)
cat(readLines(prov), sep = "\n")
```
