---
title: "20 - Gene expression x byssal thread mechanics correlation (paired samples)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# Purpose

A subset of *M. trossulus* that were mechanically tested for byssal plaque attachment
were also sampled for Tag-seq. This script links those two datasets at the level of the
individual animal and asks: **does foot (or gill) gene expression at day 3 of stress
covary with plaque attachment performance?**

It produces:

1. A paired-sample manifest (exactly which animals/samples were used).
2. Per-gene association statistics (expression vs. each mechanical metric), controlling
   for treatment, with multiple-testing correction.
3. Two gene universes: a biologically motivated **candidate set** (byssal / collagen /
   HSP / HIF / peroxidase / aminoacyl-tRNA-synthetase genes) and the **data-driven DEG
   union** (genes already found to respond to any stressor).
4. Diagnostic plots (top-hit scatterplots, candidate-gene correlation heatmap).

> **READ THIS FIRST - power.** The paired set is small: n = 31 animals (OA = 12, OW = 12,
> DO = 7). Within any single treatment n is 7-12. This analysis is **exploratory and
> hypothesis-generating, not confirmatory.** Treat raw p-values as ranking signals, expect
> few or no hits to survive FDR correction, and do not write causal language from it.
> Tag-seq sampling was destructive (foot/gill removed), so expression and the *post-stress*
> threads come from the same animal at the same timepoint; this is association, not a
> demonstrated mechanism.

---

# 0. Configuration

Set these two paths to wherever the repos live on your machine. The script reads thread
data from the **main** repo and expression data from the **expression** repo.

```{r config}
## ---- analysis options ---------------------------------------------------------------
TISSUE        <- "F"          # "F" = foot (byssus-producing organ; default), "G" = gill
MIN_COUNT     <- 5            # gene kept if count > MIN_COUNT in >= 1/3 of paired samples
FDR_ALPHA     <- 0.10         # BH threshold for flagging
ADJUST_FOR_BASELINE <- FALSE  # TRUE = add each animal's pre-stress thread metric as a
                              #        covariate (n drops to animals with baseline threads)

## ---- candidate gene keywords (regex, case-insensitive) ------------------------------
## Grouped by biology; edit freely. Matched against UniProt "protein_name".
CANDIDATE_KEYWORDS <- paste(
  # byssal structural / adhesive proteins
  "byss", "foot protein", "Mefp", "adhesive plaque", "polyphenolic",
  # collagens / preCols (byssal thread core)
  "collagen",
  # plaque-curing / cross-linking enzymes
  "tyrosinase", "peroxidase", "amine oxidase", "catechol oxidase", "lysyl oxidase",
  # proteostasis / heat-shock
  "heat shock", "Hsp", "stress-70", "BiP", "chaperone",
  "endoplasmic reticulum chaperone", "hypoxia up-regulated",
  # hypoxia signalling
  "hypoxia-inducible", "Egl nine", "prolyl 4-hydroxylase", "prolyl hydroxylase",
  # translational machinery (foot-OA enrichment signal)
  "--tRNA", "tRNA ligase", "aminoacyl",
  # oxidative stress / detox
  "superoxide", "glutathione", "catalase", "peroxiredoxin", "dual oxidase",
  sep = "|"
)
```


```{r}
# Get the correct Bioconductor repository link for your R version
bioc_version <- BiocManager::version()
bioc_repo <- paste0("https://bioconductor.org/packages/", bioc_version, "/bioc")

# Install DESeq2 directly as a standard package 
install.packages("GenomeInfoDbData", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))
install.packages("DESeq2", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))

```




```{r libs}
## Load Packages
load.lib <- c("dbplyr","tidyverse","readxl","openxlsx","lubridate","DESeq2","here")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for(lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)


`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a
```

```{r paths}
## Resolve the repository root (the folder holding the analysis subfolders), robust to whether
## this is knit from gene-mechanics-correlation.Rproj or the top-level project.
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
repo_root   <- find_repo_root()
DE_DIR      <- file.path(repo_root, "differential-expression")          # expression-side inputs
THREAD_FILE <- file.path(repo_root, "thread-strength", "02_data",
                         "thread-summary-trossulus-clean.xlsx")         # curated thread summary (trossulus)
OUT_DIR     <- file.path(repo_root, "gene-mechanics-correlation", "03_analyses", "gene_mechanics")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

## Canonicalize a mussel tag to one 3-digit form (T55 -> T055) so the thread side and the
## expression side cannot split on zero-padding when joined.
norm_id <- function(x) {
  x   <- as.character(x)
  pre <- gsub("[0-9].*$", "", x)                                        # leading letters
  num <- suppressWarnings(as.integer(gsub("[^0-9].*$", "", gsub("^[^0-9]+", "", x))))
  ifelse(is.na(num), NA_character_, sprintf("%s%03d", pre, num))
}
```

---

# 1. Build the paired-sample table

Thread metrics are averaged per animal across its 3-5 post-stress plaques. The Tag-seq
treatment assignment (the authoritative one) comes from the foot `treatmentinfo` file;
control animals (day 0 and day 3) are not part of the correlation because they were not
stressed.

```{r paired}
## ---- thread mechanics: per-animal post-stress means ---------------------------------
thread_path <- THREAD_FILE
thr_raw <- read_excel(thread_path, sheet = "data") %>% mutate(mussel = norm_id(mussel))

post_stress <- thr_raw %>%
  filter(treatment %in% c("OA", "OW", "DO")) %>%
  group_by(mussel, treatment) %>%
  summarise(
    n_plaques    = n(),
    max_force    = mean(max_force,    na.rm = TRUE),  # N, absolute holding force
    adhesion_kpa = mean(adhesion_kpa, na.rm = TRUE),  # kPa, force / plaque area
    pad_area     = mean(pad_area,     na.rm = TRUE),  # mm^2, plaque size
    .groups = "drop"
  )

## pre-stress ("control" rows are the same animals' baseline threads) - optional covariate
baseline <- thr_raw %>%
  filter(treatment == "control") %>%
  group_by(mussel) %>%
  summarise(
    base_max_force    = mean(max_force,    na.rm = TRUE),
    base_adhesion_kpa = mean(adhesion_kpa, na.rm = TRUE),
    base_pad_area     = mean(pad_area,     na.rm = TRUE),
    .groups = "drop"
  )

## ---- Tag-seq treatment assignment (definitive) --------------------------------------
fti <- read.table(file.path(DE_DIR, "03_analyses", "DEG_lists", "Foot", "F_treatmentinfo.csv"),
                  header = TRUE, sep = " ", stringsAsFactors = FALSE)
names(fti)[1] <- "sample_id"
fti <- fti %>%
  mutate(mussel = norm_id(sub("F$", "", sample_id))) %>%
  filter(day == 3, treatment != "control") %>%        # day-3 stressor animals only
  distinct(mussel, treatment)

## ---- assemble paired table ----------------------------------------------------------
paired <- post_stress %>%
  inner_join(fti, by = c("mussel", "treatment")) %>%
  left_join(baseline, by = "mussel") %>%
  mutate(sample = paste0(mussel, TISSUE),            # expression column name for chosen tissue
         treatment = factor(treatment, levels = c("OA", "OW", "DO")))

cat("Paired animals:", nrow(paired), "\n")
print(table(paired$treatment))
write_csv(paired, file.path(OUT_DIR, "paired_sample_manifest.csv"))
```

---

# 2. Load counts and VST-normalise

Variance-stabilising transformation is estimated across **all** samples of the chosen
tissue (more stable dispersion estimates), then we subset to the paired animals. Gene IDs
in the matrix look like `gene-LOC134696364|LOC134696364`; we key on the `LOC######` id.

```{r counts}
cm <- read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
               row.names = 1, check.names = FALSE)

## keep only columns for the chosen tissue (F = ...F, but NOT ...FX; G = ...G)
tissue_cols <- if (TISSUE == "F") {
  grep("F$", colnames(cm), value = TRUE)        # excludes FX (ends in X) and G
} else {
  grep("G$", colnames(cm), value = TRUE)
}
cm_tissue <- cm[, tissue_cols, drop = FALSE]

## minimal colData (treatment per sample, for dispersion modelling)
mussel_of  <- norm_id(sub(paste0(TISSUE, "$"), "", colnames(cm_tissue)))
all_meta <- read.table(file.path(DE_DIR, "03_analyses", "DEG_lists", "Foot", "F_treatmentinfo.csv"),
                        header = TRUE, sep = " ", stringsAsFactors = FALSE)
names(all_meta)[1] <- "sample_id"
all_meta <- all_meta %>% mutate(mussel = norm_id(sub("F$", "", sample_id)))
coldata <- data.frame(mussel = mussel_of) %>%
  left_join(all_meta %>% distinct(mussel, treatment, day), by = "mussel") %>%
  mutate(grp = ifelse(is.na(treatment), "unk", paste0(treatment, "_d", day)))
rownames(coldata) <- colnames(cm_tissue)
coldata$grp <- factor(coldata$grp)

## DESeq2 + VST
dds <- DESeqDataSetFromMatrix(round(as.matrix(cm_tissue)), coldata, design = ~ 1)
dds <- dds[rowSums(counts(dds)) > 0, ]
vsd <- vst(dds, blind = TRUE)
vmat <- assay(vsd)                       # genes x samples (VST)
rownames(vmat) <- sub(".*\\|", "", sub("^gene-", "", rownames(vmat)))  # -> LOC id (or gene-... if no pipe)
rownames(vmat) <- sub("^gene-", "", rownames(vmat))
vmat <- vmat[!duplicated(rownames(vmat)), ]   # safety: keep one row per LOC key

## subset to paired samples; filter low-expression genes within the paired set
stopifnot(all(paired$sample %in% colnames(vmat)))
vp <- vmat[, paired$sample, drop = FALSE]
counts_p <- counts(dds)[, paired$sample, drop = FALSE]
keep <- rowSums(counts_p > MIN_COUNT) >= ceiling(nrow(paired) / 3)
vp <- vp[keep, ]
cat("Genes tested (after expression filter):", nrow(vp), "\n")
```

---

# 3. Annotation map (portable LOC -> protein name)

Built from the genome-based annotated DEG files (`GOterms_genome/*_sigs_ID.csv`), which
carry both the `LOC_ID` and the UniProt `protein_name`. This covers stress-responsive
genes; unannotated genes are kept in the DEG-wide test but cannot be keyword-filtered.

```{r annot}
sig_files <- list.files(file.path(DE_DIR, "03_analyses", "DEG_lists", "GOterms_genome"),
                        pattern = "_sigs_ID\\.csv$", full.names = TRUE)
annot <- map_dfr(sig_files, ~ {
  d <- tryCatch(read.csv(.x, stringsAsFactors = FALSE), error = function(e) NULL)
  if (is.null(d) || !all(c("LOC_ID", "protein_name") %in% names(d))) return(NULL)
  d %>% select(LOC_ID, protein_name)
}) %>% distinct(LOC_ID, .keep_all = TRUE)

cat("Annotated LOCs:", nrow(annot), "\n")

## candidate gene set = annotated genes whose protein name matches the keyword list
candidates <- annot %>%
  filter(grepl(CANDIDATE_KEYWORDS, protein_name, ignore.case = TRUE)) %>%
  filter(LOC_ID %in% rownames(vp))
cat("Candidate (byssal/stress) genes present & expressed:", nrow(candidates), "\n")
print(candidates, n = nrow(candidates))
```

---

# 4. DEG union (data-driven gene universe)

Union of significant genes from all stressor contrasts for this tissue (treatment-control
baseline). These are genes that already responded to stress; the question is whether,
among the paired animals, their expression *level* tracks attachment.

```{r deg_union}
deg_dir <- file.path(DE_DIR, "03_analyses", "DEG_lists", ifelse(TISSUE == "F", "Foot", "Gill"))
deg_files <- list.files(deg_dir, pattern = "_TC_siggene.*\\.csv$", full.names = TRUE)
deg_union <- map(deg_files, ~ {
  d <- tryCatch(read.table(.x, header = TRUE, sep = " ", stringsAsFactors = FALSE),
                error = function(e) NULL)
  if (is.null(d)) return(character(0))
  g <- d[[grep("gene", names(d), ignore.case = TRUE)[1]]]
  sub(".*\\|", "", sub("^gene-", "", g))
}) %>% unlist() %>% unique()
deg_union <- intersect(deg_union, rownames(vp))
cat("DEG-union genes present & expressed:", length(deg_union), "\n")
```

---

# 5. Association testing

For every gene we fit, controlling for treatment:

* **Parametric:** `lm(metric ~ expression + treatment)` -> slope, t-test p-value.
* **Robust:** partial Spearman correlation of expression vs metric, treatment partialled
  out (rank-residualisation). Reported as `rho_partial`.

Run across three mechanical metrics (`max_force`, `adhesion_kpa`, `pad_area`). BH
correction is applied **within each gene set x metric** combination.

```{r assoc_fns}
metrics <- c("max_force", "adhesion_kpa", "pad_area")

# design matrix for treatment (used to residualise)
trtX <- model.matrix(~ treatment, data = paired)

partial_spearman <- function(x, y) {
  # rank-based partial correlation, treatment partialled out
  rx <- residuals(lm(rank(x) ~ trtX - 1))
  ry <- residuals(lm(rank(y) ~ trtX - 1))
  suppressWarnings(cor(rx, ry, method = "pearson"))   # corr of ranks' residuals
}

test_gene <- function(loc, metric) {
  x <- as.numeric(vp[loc, paired$sample])
  y <- paired[[metric]]
  dat <- data.frame(y = y, x = x, treatment = paired$treatment)
  if (ADJUST_FOR_BASELINE) {
    bcol <- paste0("base_", metric)
    dat$base <- paired[[bcol]]
    dat <- dat[stats::complete.cases(dat), ]
    fm <- tryCatch(lm(y ~ x + treatment + base, data = dat), error = function(e) NULL)
  } else {
    fm <- tryCatch(lm(y ~ x + treatment, data = dat), error = function(e) NULL)
  }
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"x" %in% rownames(co)) return(NULL)
  tibble(
    LOC_ID     = loc,
    metric     = metric,
    n          = nrow(dat),
    slope      = co["x", "Estimate"],
    p_lm       = co["x", "Pr(>|t|)"],
    rho_partial = partial_spearman(x, y)
  )
}

run_set <- function(loci, set_name) {
  res <- map_dfr(metrics, function(mt) {
    map_dfr(loci, ~ test_gene(.x, mt))
  })
  res %>%
    group_by(metric) %>%
    mutate(q_lm = p.adjust(p_lm, "BH")) %>%
    ungroup() %>%
    left_join(annot, by = "LOC_ID") %>%
    mutate(gene_set = set_name) %>%
    arrange(metric, p_lm)
}
```

```{r run}
res_cand <- run_set(candidates$LOC_ID, "candidate")
res_deg  <- run_set(deg_union,        "DEG_union")

write_csv(res_cand, file.path(OUT_DIR, sprintf("assoc_candidate_%s.csv", TISSUE)))
write_csv(res_deg,  file.path(OUT_DIR, sprintf("assoc_DEGunion_%s.csv",  TISSUE)))

cat("\n================ CANDIDATE GENES: top associations per metric ================\n")
res_cand %>%
  group_by(metric) %>% slice_min(p_lm, n = 8) %>% ungroup() %>%
  mutate(across(c(slope, rho_partial), ~round(., 3)),
         p_lm = signif(p_lm, 2), q_lm = signif(q_lm, 2)) %>%
  select(metric, LOC_ID, protein_name, slope, p_lm, q_lm, rho_partial) %>%
  print(n = 24)

cat("\n================ DEG-UNION: genes passing BH <", FDR_ALPHA, "================\n")
hits <- res_deg %>% filter(q_lm < FDR_ALPHA) %>% arrange(metric, p_lm)
if (nrow(hits) == 0) {
  cat("None survive FDR at this sample size (expected). Inspect raw-p ranking below.\n")
  res_deg %>% group_by(metric) %>% slice_min(p_lm, n = 5) %>% ungroup() %>%
    mutate(p_lm = signif(p_lm, 2)) %>%
    select(metric, LOC_ID, protein_name, slope, p_lm, rho_partial) %>% print(n = 15)
} else {
  print(hits, n = nrow(hits))
}
```

---

# 6. Plots

## 6.1 Candidate-gene correlation heatmap

Partial Spearman (treatment-controlled) of every candidate gene against each metric.

```{r heatmap, fig.height=7}
hm <- res_cand %>%
  mutate(label = paste0(substr(protein_name, 1, 34), " (", LOC_ID, ")")) %>%
  select(label, metric, rho_partial) %>%
  pivot_wider(names_from = metric, values_from = rho_partial) %>%
  arrange(desc(abs(max_force)))

hm_long <- hm %>% pivot_longer(-label, names_to = "metric", values_to = "rho")
ggplot(hm_long, aes(metric, factor(label, levels = rev(hm$label)), fill = rho)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "partial\nSpearman",
       title = sprintf("Candidate gene expression vs plaque mechanics (%s, n=%d)",
                       ifelse(TISSUE=="F","foot","gill"), nrow(paired))) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 8))
ggsave(file.path(OUT_DIR, sprintf("candidate_heatmap_%s.png", TISSUE)),
       width = 7, height = 8, dpi = 150)
```

## 6.2 Scatterplots for the strongest candidate hits (vs max force)

```{r scatter, fig.height=8}
top_loci <- res_cand %>% filter(metric == "max_force") %>%
  slice_min(p_lm, n = 6) %>% pull(LOC_ID)

plot_df <- map_dfr(top_loci, function(loc) {
  tibble(LOC_ID = loc,
         protein = annot$protein_name[match(loc, annot$LOC_ID)],
         expr = as.numeric(vp[loc, paired$sample]),
         max_force = paired$max_force,
         treatment = paired$treatment)
}) %>% mutate(facet = paste0(substr(protein, 1, 30), "\n", LOC_ID))

ggplot(plot_df, aes(expr, max_force)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linewidth = 0.6) +
  geom_point(aes(color = treatment), size = 2) +
  facet_wrap(~ facet, scales = "free_x") +
  labs(x = "VST expression", y = "Plaque max force (N)",
       title = "Top candidate genes vs attachment force (line = pooled fit)") +
  theme_bw(base_size = 11)
ggsave(file.path(OUT_DIR, sprintf("top_candidate_scatter_%s.png", TISSUE)),
       width = 9, height = 7, dpi = 150)
```

---

# 7. Session info & interpretation notes

```{r notes, echo=FALSE, results='asis'}
cat("
**How to read the outputs**

- `assoc_candidate_*.csv` / `assoc_DEGunion_*.csv`: one row per gene x metric. `slope` is
  the change in the mechanical metric per unit VST expression (treatment held constant);
  `p_lm`/`q_lm` are the raw and BH-adjusted p-values for that slope; `rho_partial` is the
  treatment-controlled Spearman rank correlation.
- A negative slope/rho means higher expression goes with *weaker* attachment.

**Caveats baked into the design (do not skip in the manuscript):**

1. n = 31 (OA 12 / OW 12 / DO 7). Within-treatment correlations are not interpretable on
   their own; the pooled-with-covariate model borrows strength across treatments but
   confounds treatment-level expression shifts with attachment differences.
2. Multiple testing: the candidate set (~tens of genes) is the defensible test; the
   DEG-union (hundreds) is screening only.
3. Destructive sampling: expression and post-stress threads are the same animal/timepoint,
   so this is contemporaneous association, not a predictive or causal link.
4. `pad_area` and `max_force` move together; `adhesion_kpa` (area-normalised) is the
   stricter material-strength readout and tends to show weaker associations - that pattern
   is itself informative (stress acts more on plaque size than on intrinsic strength).
5. Set `ADJUST_FOR_BASELINE <- TRUE` to re-run using each animal's pre-stress thread
   strength as a covariate (tests whether expression tracks the *change* in attachment);
   this drops the few animals lacking baseline threads.
")
sessionInfo()
```
