---
title: "20 - Gene expression x byssal thread mechanics correlation (paired samples)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"   # "F" foot (default) or "G" gill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# Purpose

A subset of *M. trossulus* that were mechanically tested for byssal plaque attachment
were also sampled for Tag-seq. This script links those two datasets at the level of the
individual animal and asks: **does foot (or gill) gene expression at day 3 of stress
covary with plaque attachment performance?**

It produces:

1. A paired-sample manifest (exactly which animals/samples were used).
2. Per-gene association statistics (expression vs. each mechanical metric), controlling
   for treatment, with multiple-testing correction.
3. Two gene universes: a biologically motivated **candidate set** (byssal / collagen /
   HSP / HIF / peroxidase / aminoacyl-tRNA-synthetase genes) and the **data-driven DEG
   union** (genes already found to respond to any stressor).
4. Diagnostic plots (top-hit scatterplots, candidate-gene correlation heatmap).

> **READ THIS FIRST - power.** The paired set is small: with the full thread dataset and the
> day-3 control arm included, n = 45 foot animals (control = 11, OA = 12, OW = 12, DO = 10),
> 44 of them with baseline threads (gill: 46 / 45). Within any single arm n is 10-12. This
> analysis is **exploratory and hypothesis-generating, not confirmatory.** Treat raw p-values
> as ranking signals, expect few or no hits to survive FDR correction, and do not write causal
> language from it. Tag-seq sampling was destructive (foot/gill removed), so expression and the
> *post-stress* threads come from the same animal at the same timepoint; this is association,
> not a demonstrated mechanism.

> **The search is declared here, not in the write-up.** `PRIMARY_METRICS` below names the
> metrics that form the confirmatory family (candidate genes x primary metrics); every other
> metric is labelled `exploratory` in every output table (`tier` column), and script 21's
> permutation reports a separate search-corrected p for the primary family.

> **Which metric to read first.** On the full dataset the stressor effect is in **peak force
> and plaque area separately**, not in their ratio: adhesion (kPa) falls ~20% in every arm
> including control (arm x timepoint p = 0.55), while force (p = 0.0005) and area
> (p < 0.0001) split warming/hypoxia from control/acidification
> (`thread-strength/01_code/4_decompose_adhesion.Rmd`). A gene associated with plaque size
> is invisible against the ratio. `max_force` and `pad_area` are therefore the primary
> targets here; `adhesion_kpa` is retained as the material-strength readout. The
> change-from-baseline metrics (`dlog_*`) ask the sharper question: does expression track
> how much *this animal's* attachment moved from its own baseline.

---

# 0. Configuration

Set these two paths to wherever the repos live on your machine. The script reads thread
data from the **main** repo and expression data from the **expression** repo.

```{r config}
## ---- analysis options ---------------------------------------------------------------
## Tissue comes from the knit parameter (`params$tissue`, default "F") so the same script
## serves foot and gill; the driver 24-run_gene_mechanics_by_tissue.Rmd renders both.
## When run chunk-by-chunk outside a knit, `params` may not exist; fall back to foot.
TISSUE <- if (exists("params") && !is.null(params$tissue)) toupper(params$tissue) else "F"
stopifnot(TISSUE %in% c("F", "G"))
MIN_COUNT     <- 5            # gene kept if count > MIN_COUNT in >= 1/3 of paired samples
FDR_ALPHA     <- 0.10         # BH threshold for flagging
## Detection floor. A gene whose VST value sits at the floor (zero counts) in many paired
## animals gives an association driven by presence/absence, not by expression level. Every
## tested gene gets frac_at_floor and floor_flag (exclude > FLOOR_EXCLUDE, caution >
## FLOOR_CAUTION); with EXCLUDE_FLOOR_GENES = TRUE the `exclude` genes are dropped from the
## candidate and DEG-union families BEFORE testing, so BH and script 21's permutation run on
## the family that can be interpreted. They stay listed (floor_flag, tested = FALSE) and they
## still contribute to script 21's module scores.
FLOOR_EXCLUDE <- 0.40
FLOOR_CAUTION <- 0.20
EXCLUDE_FLOOR_GENES <- TRUE
ADJUST_FOR_BASELINE <- TRUE   # TRUE = add each animal's pre-stress thread metric as a
                              #        covariate for the LEVEL metrics (n drops to animals
                              #        with baseline threads). The dlog_* CHANGE metrics
                              #        never use it; the baseline is inside the response.
INCLUDE_CONTROL_ARM <- TRUE   # TRUE = the day-3 common-garden control animals (T126-T137)
                              #        are part of the paired set, as a fourth treatment
                              #        level. They have foot RNA, day-3 threads and baseline
                              #        threads. Their own ~20% decline is the null trajectory
                              #        a stress-specific expression signal must beat.
                              # FALSE = stressor arms only.
USE_SECRETION_STATE <- FALSE  # TRUE = add each animal's byssal SECRETION STATE (on/off; the
                              #        foot plaque-protein module, computed per foot sample by
                              #        differential-expression/01_code/01_7-secretion_state.Rmd
                              #        and read from SECRETION_STATE_FILE) as a nuisance
                              #        covariate in every per-gene regression, beside
                              #        treatment. With ~10 of 46 day-3 animals "on", any
                              #        candidate that co-varies with that module would
                              #        otherwise pick up secretion state, not mechanics.
                              #        Script 21 reads the same choice from metrics_config and
                              #        permutes within arm x state. Gill animals take the
                              #        state of their own foot sample.
                              # FALSE (default) = treatment is the only covariate; the state
                              #        is still joined into the manifest for inspection when
                              #        the file exists, but enters no model.

## ---- mechanical metrics -------------------------------------------------------------
## LEVEL metrics: per-animal mean of its day-3 plaques.
## CHANGE metrics: log(day-3 mean / baseline mean) per animal; NA where no baseline.
## Order here is the reporting order. Force and area first, deliberately (see above).
LEVEL_METRICS  <- c("max_force", "pad_area", "adhesion_kpa", "max_displacement")
CHANGE_METRICS <- c("dlog_max_force", "dlog_pad_area", "dlog_adhesion_kpa")
## PRIMARY metrics: the pre-declared confirmatory family. Force and area carry the stressor
## signal (script 4), so their day-3 levels and their changes from baseline are primary;
## adhesion (their ratio), extension and the adhesion change are exploratory. Every output
## table carries a `tier` column with this label; script 21 permutes the primary family
## separately from the whole search.
PRIMARY_METRICS <- c("max_force", "pad_area", "dlog_max_force", "dlog_pad_area")
stopifnot(all(PRIMARY_METRICS %in% c(LEVEL_METRICS, CHANGE_METRICS)))

## ---- candidate gene keywords (regex, case-insensitive) ------------------------------
## Grouped by biology; edit freely. Matched against UniProt "protein_name".
CANDIDATE_KEYWORDS <- paste(
  # byssal structural / adhesive proteins (plaque proteins, thread core, cuticle)
  "byss", "foot protein", "Mefp", "adhesive plaque", "polyphenolic",
  "byssal protein-", "preCol", "precollagen", "byssal EP", "ACDC", "byssal tyrosinase",
  "byssal peroxidase", "byssal HSP", "byssal calumenin", "byssal protease inhibitor",
  # collagens / preCols (byssal thread core)
  "collagen",
  # plaque-curing / cross-linking enzymes
  "tyrosinase", "peroxidase", "amine oxidase", "catechol oxidase", "lysyl oxidase",
  # proteostasis / heat-shock
  "heat shock", "Hsp", "stress-70", "BiP", "chaperone",
  "endoplasmic reticulum chaperone", "hypoxia up-regulated",
  # hypoxia signalling
  "hypoxia-inducible", "Egl nine", "prolyl 4-hydroxylase", "prolyl hydroxylase",
  # translational machinery (foot-OA enrichment signal)
  "--tRNA", "tRNA ligase", "aminoacyl",
  # oxidative stress / detox
  "superoxide", "glutathione", "catalase", "peroxiredoxin", "dual oxidase",
  sep = "|"
)
```


```{r}
# Get the correct Bioconductor repository link for your R version
# bioc_version <- BiocManager::version()
# bioc_repo <- paste0("https://bioconductor.org/packages/", bioc_version, "/bioc")
# 
# # Install DESeq2 directly as a standard package 
# install.packages("GenomeInfoDbData", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))
# install.packages("DESeq2", repos = c(Bioc = bioc_repo, CRAN = "https://cloud.r-project.org"))

```




```{r libs}
## Load Packages
## DESeq2 first, tidyverse LAST: S4Vectors/IRanges (pulled in by DESeq2) mask dplyr's
## rename(), count(), first() and desc(). Loading tidyverse after them restores dplyr's.
load.lib <- c("DESeq2","readxl","openxlsx","here","tidyverse")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for(lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)


`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a
```

```{r paths}
## Resolve the repository root (the folder holding the analysis subfolders), robust to whether
## this is knit from gene-mechanics-correlation.Rproj or the top-level project.
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
repo_root   <- find_repo_root()
DE_DIR      <- file.path(repo_root, "differential-expression")          # expression-side inputs
## The curated thread table produced by thread-strength scripts 1-2 and hand-reviewed; and
## the per-animal response classification produced by thread-strength script 4.
THREAD_FILE   <- file.path(repo_root, "thread-strength", "03_analyses", "thread-summary.xlsx")
RESPONSE_FILE <- file.path(repo_root, "thread-strength", "03_analyses", "decompose-adhesion",
                           "mussel_response_classification.csv")
OUT_DIR     <- file.path(repo_root, "gene-mechanics-correlation", "03_analyses", "gene_mechanics")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
## Per-animal byssal secretion state (differential-expression script 01_7). Joined into the
## manifest whenever it exists; used as a covariate only when USE_SECRETION_STATE is TRUE.
SECRETION_STATE_FILE <- file.path(DE_DIR, "02_data", "secretion_state.csv")
## ---- annotation map and candidate universe ------------------------------------------
## CANDIDATE_ANNOTATION:
##   "genome" (default) = every gene in the count matrix is annotated with its best UniProt
##            hit (highest bitscore) from the genome-wide BLAST table MASTER_ANNOT_FILE, so
##            any EXPRESSED gene whose name matches CANDIDATE_KEYWORDS is a candidate.
##   "TC_DEG" = only genes that carry a name in the treatment-vs-control DEG tables
##            (GOterms_genome/*_sigs_ID.csv) can be candidates, plus the byssal structural
##            genes. That was the behaviour until 2026-09-18; it made "already a DEG in some
##            contrast" a hidden entry condition for the candidate family.
## The two sources agree on the name for 947 of the 969 DEG-annotated genes; the 22 that
## differ are LOCs whose transcripts hit different proteins, and the genome map takes the
## highest-scoring hit (the same rule script 23 uses).
## CANDIDATE_MAX_EVALUE / CANDIDATE_MIN_PIDENT: BLAST-quality floor for a gene to be a
## candidate under "genome" (the map itself keeps every hit, with pident and evalue columns).
## On this table every keyword match already has evalue < 1e-10; pident >= 40 would cut the
## foot candidates from ~226 to ~144 (script 23's HI_CONF_ONLY rule).
CANDIDATE_ANNOTATION <- "genome"
CANDIDATE_MAX_EVALUE <- 1e-10
CANDIDATE_MIN_PIDENT <- 0
MASTER_ANNOT_FILE <- file.path(repo_root, "blast", "03_analyses", "genome-foot", "LOC_GO_list.txt")
BYSSAL_STRUCTURAL_REGEX <- paste("foot protein", "adhesive plaque", "polyphenolic", "byssal protein-",
                                 "preCol", "precollagen", "byssal EP", "ACDC", "byssal tyrosinase",
                                 sep = "|")

if (!file.exists(THREAD_FILE))
  stop("Thread summary not found:\n  ", THREAD_FILE,
       "\nRun thread-strength scripts 1 and 2 and curate 03_analyses/thread-summary.xlsx.")

## Canonicalize a mussel tag to one 3-digit form (T55 -> T055) so the thread side and the
## expression side cannot split on zero-padding when joined.
norm_id <- function(x) {
  x   <- as.character(x)
  pre <- gsub("[0-9].*$", "", x)                                        # leading letters
  num <- suppressWarnings(as.integer(gsub("[^0-9].*$", "", gsub("^[^0-9]+", "", x))))
  ifelse(is.na(num), NA_character_, sprintf("%s%03d", pre, num))
}

## Read the curated thread table under either the current or the older column names, and
## make sure `phase` exists. thread_trt = what the THREAD was built in; mussel_trt = the arm
## the ANIMAL was assigned to; phase = lab / pre / post. The older `treatment == "control"`
## convention pooled baseline threads with the day-3 control arm and is not used here.
read_thread_summary <- function(path) {
  d <- readxl::read_excel(path, sheet = "data")
  ren <- c(mussel_ID = "mussel", thread_num = "thread", treatment = "thread_trt")
  for (from in names(ren)) {
    to <- ren[[from]]
    if (from %in% names(d) && !to %in% names(d)) names(d)[names(d) == from] <- to
  }
  need <- c("mussel", "thread_trt", "mussel_trt", "max_force", "pad_area")
  if (length(setdiff(need, names(d))) > 0)
    stop("thread-summary.xlsx is missing: ", paste(setdiff(need, names(d)), collapse = ", "))
  if (!"phase" %in% names(d))
    d$phase <- dplyr::case_when(d$thread_trt == "lab_control" ~ "lab",
                                d$thread_trt == "baseline"    ~ "pre",
                                TRUE                          ~ "post")
  if (!"max_displacement" %in% names(d)) d$max_displacement <- NA_real_
  d %>%
    mutate(mussel       = norm_id(mussel),
           adhesion_kpa = max_force / pad_area * 1000) %>%   # recomputed, never trusted from the sheet
    filter(!is.na(adhesion_kpa))
}
```

---

# 1. Build the paired-sample table

Thread metrics are averaged per animal across its day-3 plaques. The Tag-seq treatment
assignment (the authoritative one) comes from the foot `treatmentinfo` file. With
`INCLUDE_CONTROL_ARM = TRUE` the day-3 control animals enter as their own treatment level;
the day-0 lab-reference animals never do (they have no day-3 threads).

```{r paired}
## ---- thread mechanics: per-animal day-3 means and per-animal baseline means ----------
thr_raw <- read_thread_summary(THREAD_FILE)

post_stress <- thr_raw %>%
  filter(phase == "post") %>%
  group_by(mussel, treatment = mussel_trt) %>%
  summarise(
    n_plaques        = n(),
    max_force        = mean(max_force,        na.rm = TRUE),  # N, absolute holding force
    pad_area         = mean(pad_area,         na.rm = TRUE),  # mm^2, plaque size
    adhesion_kpa     = mean(adhesion_kpa,     na.rm = TRUE),  # kPa, force / plaque area
    max_displacement = mean(max_displacement, na.rm = TRUE),  # mm, extension at break
    .groups = "drop"
  )

## pre-exposure baseline: the same animals' own threads before the exposure began
baseline <- thr_raw %>%
  filter(phase == "pre") %>%
  group_by(mussel) %>%
  summarise(
    n_plaques_base        = n(),
    base_max_force        = mean(max_force,        na.rm = TRUE),
    base_pad_area         = mean(pad_area,         na.rm = TRUE),
    base_adhesion_kpa     = mean(adhesion_kpa,     na.rm = TRUE),
    base_max_displacement = mean(max_displacement, na.rm = TRUE),
    .groups = "drop"
  )

## ---- Tag-seq treatment assignment (definitive), for the chosen tissue ----------------
## The gill file codes the day-3 control arm as "control_3" and has no leading index column;
## read_treatmentinfo() normalises both so the two tissues look identical downstream.
read_treatmentinfo <- function(tissue) {
  f <- file.path(DE_DIR, "03_analyses", "DEG_lists",
                 ifelse(tissue == "F", "Foot", "Gill"),
                 sprintf("%s_treatmentinfo.csv", tissue))
  if (!file.exists(f)) stop("treatmentinfo not found: ", f)
  d <- read.table(f, header = TRUE, sep = " ", stringsAsFactors = FALSE)
  names(d)[1] <- "sample_id"
  d %>%
    mutate(sample_id = as.character(sample_id),
           mussel    = norm_id(sub(paste0(tissue, "$"), "", sample_id)),
           treatment = sub("_3$", "", treatment),          # control_3 -> control
           day       = as.integer(day)) %>%
    select(sample_id, mussel, treatment, day)
}
arms_kept <- if (INCLUDE_CONTROL_ARM) c("control", "OA", "OW", "DO") else c("OA", "OW", "DO")
fti <- read_treatmentinfo(TISSUE) %>%
  filter(day == 3, treatment %in% arms_kept) %>%       # day-3 animals only
  distinct(mussel, treatment)

## ---- assemble paired table ----------------------------------------------------------
paired <- post_stress %>%
  inner_join(fti, by = c("mussel", "treatment")) %>%   # the join also checks the two
  left_join(baseline, by = "mussel") %>%               # sources agree on the arm
  mutate(
    dlog_max_force    = log(max_force    / base_max_force),     # CHANGE metrics: log ratio
    dlog_pad_area     = log(pad_area     / base_pad_area),      # day-3 / baseline, per animal
    dlog_adhesion_kpa = log(adhesion_kpa / base_adhesion_kpa),  # NA where no baseline
    sample    = paste0(mussel, TISSUE),                # expression column name for chosen tissue
    treatment = factor(treatment, levels = arms_kept)
  )

## Arm disagreement between the thread key and the Tag-seq metadata would drop an animal
## silently at the inner_join above; make that loud.
arm_clash <- post_stress %>%
  inner_join(fti %>% dplyr::rename(rna_treatment = treatment), by = "mussel") %>%
  dplyr::filter(treatment != rna_treatment)
if (nrow(arm_clash) > 0) {
  warning(nrow(arm_clash), " animal(s) carry different arms in thread-summary and F_treatmentinfo; ",
          "excluded: ", paste(arm_clash$mussel, collapse = ", "))
  print(arm_clash)
}

## ---- per-animal response classification from thread-strength script 4 ----------------
if (file.exists(RESPONSE_FILE)) {
  resp <- read_csv(RESPONSE_FILE, show_col_types = FALSE) %>%
    mutate(mussel = norm_id(mussel)) %>%
    select(mussel, response_class, response_score,
           ends_with("_direction"), ends_with("_vs_control"))
  paired <- paired %>% left_join(resp, by = "mussel")
} else {
  warning("Response classification not found (", RESPONSE_FILE,
          "); run thread-strength/01_code/4_decompose_adhesion.Rmd. Continuing without it.")
}

## An animal can be in treatmentinfo without a column in the count matrix (a library that
## was dropped at QC). Drop it here, loudly, rather than fail at the VST step.
cm_cols <- colnames(read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
                             nrows = 1, check.names = FALSE))
no_expr <- paired %>% filter(!sample %in% cm_cols)
if (nrow(no_expr) > 0) {
  message(nrow(no_expr), " paired animal(s) have no ", TISSUE,
          " column in the count matrix and are dropped: ",
          paste(no_expr$mussel, collapse = ", "))
  paired <- paired %>% filter(sample %in% cm_cols)
}

## ---- byssal secretion state (foot plaque-protein module, script 01_7) -----------------
## One state per ANIMAL, taken from its foot sample, so the gill analysis uses the same
## label. Joined for inspection whenever the file exists; a model covariate only when
## USE_SECRETION_STATE is TRUE.
if (file.exists(SECRETION_STATE_FILE)) {
  sec <- read_csv(SECRETION_STATE_FILE, show_col_types = FALSE) %>%
    mutate(mussel = norm_id(mussel)) %>%
    filter(!is.na(mussel)) %>%
    distinct(mussel, .keep_all = TRUE) %>%
    select(mussel, secretion_score, secretion_state)
  paired <- paired %>% left_join(sec, by = "mussel")
  cat("Secretion state joined:", sum(!is.na(paired$secretion_state)), "of", nrow(paired),
      "animals labelled (", sum(paired$secretion_state == "on", na.rm = TRUE), "on ).\n")
} else {
  paired$secretion_score <- NA_real_; paired$secretion_state <- NA_character_
  if (USE_SECRETION_STATE)
    stop("USE_SECRETION_STATE is TRUE but ", SECRETION_STATE_FILE, " does not exist; ",
         "run differential-expression/01_code/01_7-secretion_state.Rmd first.")
}
if (USE_SECRETION_STATE) {
  no_state <- paired %>% filter(is.na(secretion_state))
  if (nrow(no_state) > 0) {
    warning(nrow(no_state), " paired animal(s) have no secretion state and are dropped ",
            "because USE_SECRETION_STATE = TRUE: ", paste(no_state$mussel, collapse = ", "))
    paired <- paired %>% filter(!is.na(secretion_state))
  }
  paired <- paired %>% mutate(secretion_state = factor(secretion_state, levels = c("off", "on")))
  print(table(arm = paired$treatment, secretion_state = paired$secretion_state))
}
## the nuisance terms every regression below controls for; script 21 reads the same list
COVARIATES <- if (USE_SECRETION_STATE) c("treatment", "secretion_state") else "treatment"

cat("Paired animals:", nrow(paired), "\n")
print(table(paired$treatment))
cat("With baseline threads (for CHANGE metrics and ADJUST_FOR_BASELINE):",
    sum(!is.na(paired$base_max_force)), "\n")
if ("response_class" %in% names(paired)) print(table(paired$treatment, paired$response_class))
write_csv(paired, file.path(OUT_DIR, sprintf("paired_sample_manifest_%s.csv", TISSUE)))
```

---

# 2. Load counts and VST-normalise

Variance-stabilising transformation is estimated across **all** samples of the chosen
tissue (more stable dispersion estimates), then we subset to the paired animals. Gene IDs
in the matrix look like `gene-LOC134696364|LOC134696364`; we key on the `LOC######` id.

```{r counts}
cm <- read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
               row.names = 1, check.names = FALSE)

## keep only columns for the chosen tissue (F = ...F, but NOT ...FX; G = ...G)
tissue_cols <- if (TISSUE == "F") {
  grep("F$", colnames(cm), value = TRUE)        # excludes FX (ends in X) and G
} else {
  grep("G$", colnames(cm), value = TRUE)
}
cm_tissue <- cm[, tissue_cols, drop = FALSE]

## minimal colData (treatment per sample, for dispersion modelling)
mussel_of  <- norm_id(sub(paste0(TISSUE, "$"), "", colnames(cm_tissue)))
all_meta <- read_treatmentinfo(TISSUE)
coldata <- data.frame(mussel = mussel_of) %>%
  left_join(all_meta %>% distinct(mussel, treatment, day), by = "mussel") %>%
  mutate(grp = ifelse(is.na(treatment), "unk", paste0(treatment, "_d", day)))
rownames(coldata) <- colnames(cm_tissue)
coldata$grp <- factor(coldata$grp)

## DESeq2 + VST
dds <- DESeqDataSetFromMatrix(round(as.matrix(cm_tissue)), coldata, design = ~ 1)
dds <- dds[rowSums(counts(dds)) > 0, ]
vsd  <- vst(dds, blind = TRUE)
vmat <- assay(vsd)                        # rownames = full gene IDs, same order as counts(dds)
rownames(vmat) <- sub(".*\\|", "", sub("^gene-", "", rownames(vmat)))  # -> LOC key

stopifnot(all(paired$sample %in% colnames(vmat)))
vp       <- vmat[,        paired$sample, drop = FALSE]
counts_p <- counts(dds)[, paired$sample, drop = FALSE]   # identical row order to vp
keep <- rowSums(counts_p > MIN_COUNT) >= ceiling(nrow(paired) / 3)
vp <- vp[keep, , drop = FALSE]
cat("Genes tested (after expression filter):", nrow(vp), "\n")
cat("Duplicate LOC keys among tested genes:", sum(duplicated(rownames(vp))), "\n")  # should be ~0

## detection floor per gene: fraction of paired animals at the VST minimum (zero counts)
floor_val   <- min(vp)
frac_floor  <- rowMeans(abs(vp - floor_val) < 1e-6)
floor_flag_of <- function(ids) {
  fr <- unname(frac_floor[match(ids, rownames(vp))])
  tibble(LOC_ID = ids, frac_at_floor = round(fr, 3),
         floor_flag = dplyr::case_when(fr > FLOOR_EXCLUDE ~ "exclude", fr > FLOOR_CAUTION ~ "caution", TRUE ~ "ok"),
         tested = !(EXCLUDE_FLOOR_GENES & fr > FLOOR_EXCLUDE))
}
```

---

# 3. Annotation map (portable LOC -> protein name)

The map is the genome-wide BLAST table, best UniProt hit per LOC by bitscore
(`CANDIDATE_ANNOTATION = "genome"`), with the hit's `pident` and `evalue` carried along so a
weak annotation is visible wherever a name is printed. The treatment-vs-control DEG tables
(`GOterms_genome/*_sigs_ID.csv`) are read as well: `in_TC_DEG_annotation` marks the genes
they name, and with `CANDIDATE_ANNOTATION = "TC_DEG"` only those (plus the byssal structural
genes) can enter the candidate set.

```{r annot}
## names carried by the TC DEG tables (the pre-2026-09-18 universe)
sig_files <- list.files(file.path(DE_DIR, "03_analyses", "DEG_lists", "GOterms_genome"),
                        pattern = "_sigs_ID\\.csv$", full.names = TRUE)
annot_deg <- map_dfr(sig_files, ~ {
  d <- tryCatch(read.csv(.x, stringsAsFactors = FALSE), error = function(e) NULL)
  if (is.null(d) || !all(c("LOC_ID", "protein_name") %in% names(d))) return(NULL)
  d %>% select(LOC_ID, protein_name)
}) %>% distinct(LOC_ID, .keep_all = TRUE) %>% as_tibble()
cat("LOCs named in the TC DEG tables:", nrow(annot_deg), "\n")

## genome-wide map: best hit per LOC
if (!file.exists(MASTER_ANNOT_FILE))
  stop("Master annotation not found: ", MASTER_ANNOT_FILE,
       "\nIt is the source of the annotation map (CANDIDATE_ANNOTATION = \"genome\").")
master <- read_delim(MASTER_ANNOT_FILE, delim = "\t", show_col_types = FALSE,
                     col_select = c("gene", "pident", "evalue", "bitscore", "Protein.names")) %>%
  ## key: the LOC id, or the gene symbol for the 13 mitochondrial genes (ND1, COX1, ...),
  ## which is how the count matrix names them too
  transmute(LOC_ID = dplyr::coalesce(str_extract(as.character(gene), "LOC\\d+"),
                                     na_if(trimws(as.character(gene)), "")),
            protein_name = as.character(Protein.names),
            blast_pident = suppressWarnings(as.numeric(pident)),
            blast_evalue = suppressWarnings(as.numeric(evalue)),
            bitscore     = suppressWarnings(as.numeric(bitscore))) %>%
  filter(!is.na(LOC_ID), !is.na(protein_name), nzchar(protein_name)) %>%
  arrange(desc(bitscore)) %>% distinct(LOC_ID, .keep_all = TRUE) %>%           # best hit per LOC
  select(-bitscore)
cat("LOCs with a genome-wide best hit:", nrow(master), "\n")

if (CANDIDATE_ANNOTATION == "genome") {
  annot <- master %>%
    mutate(annotation_source = "genome_blast",
           in_TC_DEG_annotation = LOC_ID %in% annot_deg$LOC_ID)
  cand_ok <- with(annot, blast_evalue <= CANDIDATE_MAX_EVALUE & blast_pident >= CANDIDATE_MIN_PIDENT)
} else if (CANDIDATE_ANNOTATION == "TC_DEG") {
  ## the DEG-table names, plus the byssal structural genes from the genome map (they are not
  ## DEGs in any treatment-vs-control contrast, so the DEG tables never reach them)
  byssal_add <- master %>%
    filter(grepl(BYSSAL_STRUCTURAL_REGEX, protein_name, ignore.case = TRUE),
           LOC_ID %in% rownames(vp), !LOC_ID %in% annot_deg$LOC_ID) %>%
    mutate(annotation_source = "genome_blast_byssal")
  annot <- bind_rows(annot_deg %>% mutate(annotation_source = "TC_DEG_sigs") %>%
                       left_join(master %>% select(LOC_ID, blast_pident, blast_evalue), by = "LOC_ID"),
                     byssal_add) %>%
    mutate(in_TC_DEG_annotation = annotation_source == "TC_DEG_sigs")
  cand_ok <- rep(TRUE, nrow(annot))
} else stop("CANDIDATE_ANNOTATION must be \"genome\" or \"TC_DEG\"")
annot$blast_ok <- !is.na(cand_ok) & cand_ok   # passes the BLAST-quality floor; script 21's modules use it too
cat("Annotation map (", CANDIDATE_ANNOTATION, "):", nrow(annot), "LOCs, of which", sum(annot$blast_ok),
    "pass the BLAST-quality floor\n")

## candidate gene set = annotated, expressed genes whose protein name matches the keyword list
candidates <- annot %>% filter(blast_ok) %>%
  filter(grepl(CANDIDATE_KEYWORDS, protein_name, ignore.case = TRUE)) %>%
  filter(LOC_ID %in% rownames(vp)) %>%
  mutate(byssal_structural = grepl(BYSSAL_STRUCTURAL_REGEX, protein_name, ignore.case = TRUE))
candidates <- candidates %>% left_join(floor_flag_of(candidates$LOC_ID), by = "LOC_ID")
cat("Candidate (byssal/stress) genes present & expressed:", nrow(candidates),
    "( byssal structural:", sum(candidates$byssal_structural),
    "; named in the TC DEG tables:", sum(candidates$in_TC_DEG_annotation),
    "; at the detection floor:", sum(candidates$floor_flag == "exclude"),
    if (EXCLUDE_FLOOR_GENES) "excluded from the tested family" else "kept", ")\n")
cat("Candidate genes tested:", sum(candidates$tested), "\n")
print(as.data.frame(candidates %>% select(LOC_ID, protein_name, blast_pident, in_TC_DEG_annotation,
                                          byssal_structural, frac_at_floor, floor_flag, tested) %>%
                      mutate(protein_name = substr(protein_name, 1, 50))))
```

```{r exports}
## ---- handoff CSVs for downstream scripts (21 reads all; 22 reads annotation_map) -----
## paired_sample_manifest_<T>.csv was written above; these add the rest so 21 can run on its
## own instead of sharing this R session's objects.
write_csv(data.frame(LOC_ID = rownames(vp), vp, check.names = FALSE),
          file.path(OUT_DIR, sprintf("vst_paired_%s.csv", TISSUE)))            # genes x paired samples (VST)
write_csv(annot,      file.path(OUT_DIR, "annotation_map.csv"))                # LOC_ID -> protein_name
write_csv(candidates, file.path(OUT_DIR, sprintf("candidate_genes_%s.csv", TISSUE)))  # candidate set
write_csv(thr_raw %>% filter(phase == "post", mussel %in% paired$mussel) %>%
            transmute(mussel, treatment = mussel_trt, phase, thread_trt,
                      max_force, pad_area, adhesion_kpa, max_displacement),
          file.path(OUT_DIR, sprintf("thread_plaques_paired_%s.csv", TISSUE)))  # raw day-3 plaques, paired animals
write_csv(tibble(metric = c(LEVEL_METRICS, CHANGE_METRICS),
                 type   = c(rep("level", length(LEVEL_METRICS)), rep("change", length(CHANGE_METRICS))),
                 tier   = ifelse(c(LEVEL_METRICS, CHANGE_METRICS) %in% PRIMARY_METRICS,
                                 "primary", "exploratory"),
                 arms   = paste(arms_kept, collapse = ";"),
                 covariates          = paste(COVARIATES, collapse = ";"),
                 use_secretion_state = USE_SECRETION_STATE,
                 adjust_for_baseline = ADJUST_FOR_BASELINE,
                 exclude_floor_genes = EXCLUDE_FLOOR_GENES),
          file.path(OUT_DIR, sprintf("metrics_config_%s.csv", TISSUE)))          # so 21 uses the same list
```

---

# 4. DEG union (data-driven gene universe)

Union of significant genes from all stressor contrasts for this tissue (treatment-control
baseline). These are genes that already responded to stress; the question is whether,
among the paired animals, their expression *level* tracks attachment.

```{r deg_union}
deg_dir <- file.path(DE_DIR, "03_analyses", "DEG_lists", ifelse(TISSUE == "F", "Foot", "Gill"))
deg_files <- list.files(deg_dir, pattern = "_TC_siggene.*\\.csv$", full.names = TRUE)
deg_union <- map(deg_files, ~ {
  d <- tryCatch(read.table(.x, header = TRUE, sep = " ", stringsAsFactors = FALSE),
                error = function(e) NULL)
  if (is.null(d)) return(character(0))
  g <- d[[grep("gene", names(d), ignore.case = TRUE)[1]]]
  sub(".*\\|", "", sub("^gene-", "", g))
}) %>% unlist() %>% unique()
deg_union <- intersect(deg_union, rownames(vp))
deg_union_tbl <- floor_flag_of(deg_union)
cat("DEG-union genes present & expressed:", length(deg_union),
    "| at the detection floor:", sum(deg_union_tbl$floor_flag == "exclude"),
    "| tested:", sum(deg_union_tbl$tested), "\n")

## the floor table for both universes, so downstream scripts and the report can read it
write_csv(bind_rows(candidates %>% transmute(LOC_ID, gene_set = "candidate", frac_at_floor, floor_flag, tested),
                    deg_union_tbl %>% filter(!LOC_ID %in% candidates$LOC_ID) %>%
                      transmute(LOC_ID, gene_set = "DEG_union", frac_at_floor, floor_flag, tested)) %>%
            left_join(annot %>% select(LOC_ID, protein_name), by = "LOC_ID") %>%
            mutate(n_samples = nrow(paired), n_at_floor = round(frac_at_floor * nrow(paired))) %>%
            select(LOC_ID, protein_name, gene_set, n_samples, n_at_floor, frac_at_floor, floor_flag, tested) %>%
            arrange(desc(frac_at_floor), LOC_ID),
          file.path(OUT_DIR, sprintf("detection_floor_flags_%s.csv", TISSUE)))
```

---

# 5. Association testing

For every gene we fit, controlling for treatment:

* **Parametric:** `lm(metric ~ expression + treatment)` -> slope, t-test p-value.
* **Robust:** partial Spearman correlation of expression vs metric, treatment partialled
  out (rank-residualisation). Reported as `rho_partial`.

Run across the LEVEL metrics (`max_force`, `pad_area`, `adhesion_kpa`, `max_displacement`)
and the CHANGE metrics (`dlog_*`, each animal's log day-3/baseline ratio). BH correction is
applied **within each gene set x metric** combination. The partial Spearman is computed on
exactly the animals the linear model used.

```{r assoc_fns}
metrics <- c(LEVEL_METRICS, CHANGE_METRICS)
is_change <- function(metric) metric %in% CHANGE_METRICS
tier_of   <- function(metric) ifelse(metric %in% PRIMARY_METRICS, "primary", "exploratory")

## nuisance design shared by every fit: treatment, plus the secretion state when enabled
nuis_formula <- reformulate(COVARIATES)                       # ~ treatment [+ secretion_state]

partial_spearman <- function(x, y, nuis) {
  # rank-based partial correlation, nuisance covariates partialled out, complete cases only
  ok <- stats::complete.cases(x, y, nuis)
  if (sum(ok) < 6) return(NA_real_)
  X  <- model.matrix(nuis_formula, data = droplevels(nuis[ok, , drop = FALSE]))
  rx <- residuals(lm(rank(x[ok]) ~ X - 1))
  ry <- residuals(lm(rank(y[ok]) ~ X - 1))
  suppressWarnings(cor(rx, ry, method = "pearson"))   # corr of ranks' residuals
}

test_gene <- function(loc, metric) {
  x <- as.numeric(vp[match(loc, rownames(vp)), paired$sample])
  y <- paired[[metric]]
  dat <- data.frame(y = y, x = x, paired[, COVARIATES, drop = FALSE])
  if (ADJUST_FOR_BASELINE && !is_change(metric)) {
    dat$base <- paired[[paste0("base_", metric)]]
    dat <- dat[stats::complete.cases(dat), ]
    fm <- tryCatch(lm(reformulate(c("x", COVARIATES, "base"), response = "y"), data = dat),
                   error = function(e) NULL)
  } else {
    dat <- dat[stats::complete.cases(dat), ]
    fm <- tryCatch(lm(reformulate(c("x", COVARIATES), response = "y"), data = dat),
                   error = function(e) NULL)
  }
  if (is.null(fm) || nrow(dat) < 6) return(NULL)
  co <- summary(fm)$coefficients
  if (!"x" %in% rownames(co)) return(NULL)
  tibble(
    LOC_ID      = loc,
    metric      = metric,
    metric_type = if (is_change(metric)) "change" else "level",
    tier        = tier_of(metric),
    n           = nrow(dat),
    slope       = co["x", "Estimate"],
    p_lm        = co["x", "Pr(>|t|)"],
    rho_partial = partial_spearman(dat$x, dat$y, dat[, COVARIATES, drop = FALSE])
  )
}

run_set <- function(loci, set_name) {
  res <- map_dfr(metrics, function(mt) {
    map_dfr(loci, ~ test_gene(.x, mt))
  })
  res %>%
    group_by(metric) %>%
    mutate(q_lm = p.adjust(p_lm, "BH")) %>%
    ungroup() %>%
    left_join(annot, by = "LOC_ID") %>%
    mutate(gene_set = set_name) %>%
    arrange(metric, p_lm)
}
```

```{r run}
res_cand <- run_set(candidates$LOC_ID[candidates$tested], "candidate") %>%
  left_join(candidates %>% select(LOC_ID, frac_at_floor, floor_flag, byssal_structural), by = "LOC_ID")
res_deg  <- run_set(deg_union_tbl$LOC_ID[deg_union_tbl$tested], "DEG_union") %>%
  left_join(deg_union_tbl %>% select(LOC_ID, frac_at_floor, floor_flag), by = "LOC_ID")

write_csv(res_cand, file.path(OUT_DIR, sprintf("assoc_candidate_%s.csv", TISSUE)))
write_csv(res_deg,  file.path(OUT_DIR, sprintf("assoc_DEGunion_%s.csv",  TISSUE)))

cat("\n================ CANDIDATE GENES: top associations per metric ================\n")
res_cand %>%
  mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_lm, n = 6) %>% ungroup() %>%
  mutate(across(c(slope, rho_partial), ~round(., 3)),
         p_lm = signif(p_lm, 2), q_lm = signif(q_lm, 2)) %>%
  select(metric, metric_type, tier, n, LOC_ID, protein_name, slope, p_lm, q_lm, rho_partial) %>%
  print(n = 42)

cat("\n================ DEG-UNION: genes passing BH <", FDR_ALPHA, "================\n")
hits <- res_deg %>% filter(q_lm < FDR_ALPHA) %>% arrange(metric, p_lm)
if (nrow(hits) == 0) {
  cat("None survive FDR at this sample size (expected). Inspect raw-p ranking below.\n")
  res_deg %>% group_by(metric) %>% slice_min(p_lm, n = 5) %>% ungroup() %>%
    mutate(p_lm = signif(p_lm, 2)) %>%
    select(metric, LOC_ID, protein_name, slope, p_lm, q_lm, rho_partial) %>%
    as.data.frame() %>% head(24) %>% print()
} else {
  print(hits, n = nrow(hits))
}
```

---

# 6. Plots

## 6.1 Candidate-gene correlation heatmap

Partial Spearman (treatment-controlled) of candidate genes against each metric. With the
genome-wide candidate set (~230 genes) the panel shows the `HEATMAP_MAX_GENES` genes with
the strongest association on any metric (smallest `p_lm`); the full table is
`assoc_candidate_<T>.csv`.

```{r heatmap, fig.height=7}
HEATMAP_MAX_GENES <- 60
hm_keep <- res_cand %>% group_by(LOC_ID) %>% summarise(best_p = min(p_lm), .groups = "drop") %>%
  slice_min(best_p, n = HEATMAP_MAX_GENES, with_ties = FALSE) %>% pull(LOC_ID)
hm <- res_cand %>%
  filter(LOC_ID %in% hm_keep) %>%
  mutate(label = paste0(substr(protein_name, 1, 34), " (", LOC_ID, ")")) %>%
  select(label, metric, rho_partial) %>%
  pivot_wider(names_from = metric, values_from = rho_partial) %>%
  arrange(desc(abs(max_force)))

hm_long <- hm %>% pivot_longer(-label, names_to = "metric", values_to = "rho") %>%
  mutate(metric = factor(metric, levels = metrics))
ggplot(hm_long, aes(metric, factor(label, levels = rev(hm$label)), fill = rho)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "partial\nSpearman",
       title = sprintf("Candidate gene expression vs plaque mechanics (%s, n=%d; top %d of %d genes)",
                       ifelse(TISSUE=="F","foot","gill"), nrow(paired), nrow(hm), nrow(candidates))) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 8))
ggsave(file.path(OUT_DIR, sprintf("candidate_heatmap_%s.png", TISSUE)),
       width = 7, height = 8, dpi = 150)
```

## 6.2 Scatterplots for the strongest candidate hits (vs max force)

```{r scatter, fig.height=8}
top_loci <- res_cand %>% filter(metric == "max_force") %>%
  slice_min(p_lm, n = 6) %>% pull(LOC_ID)

plot_df <- map_dfr(top_loci, function(loc) {
  tibble(LOC_ID = loc,
         protein = annot$protein_name[match(loc, annot$LOC_ID)],
         expr = as.numeric(vp[match(loc, rownames(vp)), paired$sample]),
         max_force = paired$max_force,
         treatment = paired$treatment)
}) %>% mutate(facet = paste0(substr(protein, 1, 30), "\n", LOC_ID))

ggplot(plot_df, aes(expr, max_force)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linewidth = 0.6) +
  geom_point(aes(color = treatment), size = 2) +
  facet_wrap(~ facet, scales = "free_x") +
  labs(x = "VST expression", y = "Plaque max force (N)",
       title = "Top candidate genes vs attachment force (line = pooled fit)") +
  theme_bw(base_size = 11)
ggsave(file.path(OUT_DIR, sprintf("top_candidate_scatter_%s.png", TISSUE)),
       width = 9, height = 7, dpi = 150)
```

## 6.3 The strongest candidate hit for each metric, coloured by arm

One panel per metric, the single best candidate gene for that metric. Colouring by arm is
the check that matters: an association driven by one arm sitting at one end of both axes is
an arm effect, not a gene effect.

```{r scatter_best, fig.height=8}
best_per_metric <- res_cand %>%
  mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup()

best_df <- pmap_dfr(best_per_metric %>% select(LOC_ID, metric, p_lm, q_lm), function(LOC_ID, metric, p_lm, q_lm) {
  m <- as.character(metric)
  tibble(metric = m,
         panel  = sprintf("%s\n%s\np = %s, q = %s", m,
                          substr(annot$protein_name[match(LOC_ID, annot$LOC_ID)], 1, 34),
                          signif(p_lm, 2), signif(q_lm, 2)),
         expr   = as.numeric(vp[match(LOC_ID, rownames(vp)), paired$sample]),
         y      = paired[[m]],
         treatment = paired$treatment)
}) %>% filter(!is.na(y)) %>%
  mutate(panel = factor(panel, levels = unique(panel)))

arm_cols <- c(control = "#7FB3D5", OA = "#4CAF50", OW = "#fdae61", DO = "#9884a3")
ggplot(best_df, aes(expr, y)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linewidth = 0.6) +
  geom_point(aes(fill = treatment), shape = 21, size = 2.6, stroke = 0.5, color = "black") +
  facet_wrap(~ panel, scales = "free", ncol = 3) +
  scale_fill_manual(values = arm_cols, name = "Arm") +
  labs(x = "VST expression", y = "Mechanical metric",
       title = sprintf("Best candidate gene per metric (%s, n = %d)",
                       ifelse(TISSUE == "F", "foot", "gill"), nrow(paired))) +
  theme_bw(base_size = 10) + theme(strip.text = element_text(size = 7.5))
ggsave(file.path(OUT_DIR, sprintf("best_hit_per_metric_scatter_%s.png", TISSUE)),
       width = 10, height = 8.5, dpi = 150)
```

---

# 7. Session info & interpretation notes

```{r notes, echo=FALSE, results='asis'}
cat("
**How to read the outputs**

- `assoc_candidate_*.csv` / `assoc_DEGunion_*.csv`: one row per gene x metric. `slope` is
  the change in the mechanical metric per unit VST expression (treatment held constant);
  `p_lm`/`q_lm` are the raw and BH-adjusted p-values for that slope; `rho_partial` is the
  treatment-controlled Spearman rank correlation.
- A negative slope/rho means higher expression goes with *weaker* attachment.

**Caveats baked into the design (do not skip in the manuscript):**

1. n = 45 foot / 46 gill with the control arm (control 11 / OA 12 / OW 12 / DO 10-11), 44 / 45
   with baselines. Within-arm correlations are not interpretable on their own; the
   pooled-with-covariate model borrows strength across arms but confounds arm-level
   expression shifts with attachment differences.
2. Multiple testing: the candidate set (~230 foot / ~280 gill genes under the genome-wide
   map) x PRIMARY_METRICS is the declared confirmatory family (`tier = primary`); the
   remaining metrics and the DEG-union (hundreds) are exploratory / screening. Script 21
   gives every family a search-corrected permutation p; quote q only beside it.
7. `USE_SECRETION_STATE` (default FALSE) adds the animal's byssal secretion state (on/off,
   from differential-expression script 01_7) as a nuisance covariate. It is off by default
   because the state is itself an expression-derived quantity measured on the same sample;
   turn it on as a sensitivity analysis when a candidate co-varies with the plaque-protein
   module (`byssal_structural = TRUE` in candidate_genes_<T>.csv).
3. Destructive sampling: expression and post-stress threads are the same animal/timepoint,
   so this is contemporaneous association, not a predictive or causal link.
4. `pad_area` and `max_force` carry the stressor signal; `adhesion_kpa` (their ratio) does
   not, because on this dataset every arm including control falls ~20% in kPa while force
   and area split warming/hypoxia from control/acidification. Read force and area first.
5. The `dlog_*` CHANGE metrics are each animal's log(day-3 / baseline) and need no baseline
   covariate; they are the direct test of whether expression tracks how much attachment
   moved. `ADJUST_FOR_BASELINE` applies to the LEVEL metrics only (ANCOVA on the day-3 level
   with baseline as covariate), a related but different question.
6. With `INCLUDE_CONTROL_ARM = TRUE` the control animals contribute a fourth treatment level.
   An expression-mechanics association that holds within the control arm as well is about
   attachment biology in general, not about the stress response; that distinction is worth
   checking by hand for any top hit (scatter by arm).
")
sessionInfo()
```

# 8. Provenance

`RUN_provenance_<T>.txt` records the configuration behind every file in
`03_analyses/gene_mechanics/` (the thread-strength scripts keep the same kind of file), and
the git commit that last touched the thread input, so a result can always be traced to the
thread table it was computed from. Script 21 appends its own block.

```{r provenance}
git_last_commit <- function(path) {
  ## guarded: no git, not a repo, or an untracked file all yield "unavailable"
  if (!nzchar(Sys.which("git")) || !file.exists(path)) return("unavailable")
  out <- tryCatch(suppressWarnings(
           system2("git", c("-C", shQuote(dirname(path)), "log", "-n", "1",
                            shQuote("--format=%h %cs"), "--", shQuote(basename(path))),
                   stdout = TRUE, stderr = FALSE)),
         error = function(e) character(0))
  if (length(out) == 0 || !is.null(attr(out, "status")) || !nzchar(out[1])) "unavailable" else out[1]
}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
writeLines(c(
  "[20-gene_mechanics_correlation.Rmd]",
  paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
  paste0("tissue               : ", TISSUE),
  paste0("thread input         : ", THREAD_FILE),
  paste0("thread input commit  : ", git_last_commit(THREAD_FILE)),
  paste0("response input       : ", RESPONSE_FILE, " (", ifelse(file.exists(RESPONSE_FILE), "present", "absent"), ")"),
  paste0("count matrix         : ", file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv")),
  paste0("INCLUDE_CONTROL_ARM  : ", INCLUDE_CONTROL_ARM),
  paste0("ADJUST_FOR_BASELINE  : ", ADJUST_FOR_BASELINE),
  paste0("USE_SECRETION_STATE  : ", USE_SECRETION_STATE),
  paste0("covariates           : ", paste(COVARIATES, collapse = " + ")),
  paste0("PRIMARY_METRICS      : ", paste(PRIMARY_METRICS, collapse = ", ")),
  paste0("LEVEL_METRICS        : ", paste(LEVEL_METRICS, collapse = ", ")),
  paste0("CHANGE_METRICS       : ", paste(CHANGE_METRICS, collapse = ", ")),
  paste0("MIN_COUNT / FDR_ALPHA: ", MIN_COUNT, " / ", FDR_ALPHA),
  paste0("paired animals       : ", nrow(paired), " (", paste(sprintf("%s %d", names(table(paired$treatment)),
                                                                     as.integer(table(paired$treatment))), collapse = ", "), ")"),
  paste0("with baseline        : ", sum(!is.na(paired$base_max_force))),
  paste0("genes tested         : ", nrow(vp)),
  paste0("CANDIDATE_ANNOTATION : ", CANDIDATE_ANNOTATION, " (evalue <= ", CANDIDATE_MAX_EVALUE,
         ", pident >= ", CANDIDATE_MIN_PIDENT, ")"),
  paste0("annotation map       : ", nrow(annot), " LOCs (", MASTER_ANNOT_FILE, ")"),
  paste0("candidate genes      : ", nrow(candidates), " (byssal structural ", sum(candidates$byssal_structural),
         "; named in TC DEG tables ", sum(candidates$in_TC_DEG_annotation), ")"),
  paste0("EXCLUDE_FLOOR_GENES  : ", EXCLUDE_FLOOR_GENES, " (exclude > ", FLOOR_EXCLUDE, ", caution > ", FLOOR_CAUTION, ")"),
  paste0("candidates tested    : ", sum(candidates$tested), " (", sum(!candidates$tested), " at the detection floor)"),
  paste0("DEG union tested     : ", sum(deg_union_tbl$tested), " of ", nrow(deg_union_tbl)),
  paste0("DEG union genes      : ", length(deg_union)),
  paste0("R / DESeq2           : ", R.version.string, " / ", as.character(packageVersion("DESeq2")))),
  prov)
cat(readLines(prov), sep = "\n")
```
