---
title: "21 - Gene x mechanics: expanded analysis (thread-level mixed ANCOVA, modules, diagnostics)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"   # "F" foot (default) or "G" gill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# What this adds to script 20

Script 20 fits the reported test, one baseline-adjusted regression (ANCOVA) per gene x
metric on the per-animal values:

    level_day3 ~ expression + treatment [+ secretion_state] + level_baseline

with force, area and adhesion on the log scale (per-animal geometric means of the plaques)
and extension on its raw scale. The animal is the unit, so there is no pseudoreplication,
but every animal counts the same whether it built one plaque or five. This script keeps
that test as the reported one and adds what a reader needs to trust or discount a hit:

- **A. Thread-level mixed ANCOVA (sensitivity)** - the same model on every plaque, with
  `mussel` as a random intercept, so an animal's weight follows the precision of its mean.
  The expression and baseline terms vary only between animals, so their tests reflect
  animal-level degrees of freedom (Satterthwaite), not the thread count. A hit whose
  `p_mixed` disagrees badly with script 20's `p_lm` is being carried by the plaque count of a
  few animals.
- **B. Module eigengenes** - collapse co-regulated pathways (HIF, oxidative, HSP,
  tRNA-synthetase, byssal/collagen, and the byssal structural module) into one score each
  and put the score through the same ANCOVA (and its mixed version), cutting the
  multiple-testing burden from hundreds of genes to six modules.
- **C. Diagnostics** - echoes the detection-floor flags, refits the reported ANCOVA for the
  best hits without their most influential animal (Cook's distance), and writes
  `best_hits_<T>.csv`: one row per metric and gene set with the reported p, both q's, the
  mixed-model p, the floor flag and the influence flag side by side.

> **Prerequisite:** run script 20 first. This script does not share 20's R session; it
> reads 20's CSV handoffs (`paired_sample_manifest_<T>`, `vst_paired_<T>`, `annotation_map`,
> `candidate_genes_<T>`, `thread_plaques_paired_<T>`, `metrics_config_<T>`,
> `assoc_candidate_<T>`, `assoc_DEGunion_<T>`, `detection_floor_flags_<T>`) from
> `03_analyses/gene_mechanics/`.

```{r setup_inputs}
suppressPackageStartupMessages({ library(tidyverse); library(lmerTest); library(here) })
## lmerTest::lmer gives Satterthwaite p-values

## ---- config -------------------------------------------------------------------------
## Tissue comes from the knit parameter (`params$tissue`, default "F") so the same script
## serves foot and gill; the driver 24-run_gene_mechanics_by_tissue.Rmd renders both.
## When run chunk-by-chunk outside a knit, `params` may not exist; fall back to foot.
TISSUE <- if (exists("params") && !is.null(params$tissue)) toupper(params$tissue) else "F"
stopifnot(TISSUE %in% c("F", "G"))
FDR_ALPHA        <- 0.10
N_INFLUENCE_HITS <- 3     # candidate hits per metric that get the leave-one-animal-out refit
## The detection-floor filter (FLOOR_EXCLUDE / EXCLUDE_FLOOR_GENES) lives in script 20, which
## writes detection_floor_flags_<T>.csv and marks each candidate `tested`; this script works
## on the same family script 20 tested.

## The metrics, their model scale (log / raw) and tier (primary / exploratory), the treatment
## levels and the nuisance covariates are read from script 20's metrics_config handoff, so
## the two scripts fit the same model on the same animals.

## ---- read script 20's CSV handoffs (run 20 first) ----------------------------------
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
IN_DIR  <- file.path(find_repo_root(), "gene-mechanics-correlation", "03_analyses", "gene_mechanics")
OUT_DIR <- IN_DIR                                  # 21 writes alongside 20's outputs
need <- function(f) { p <- file.path(IN_DIR, f)
  if (!file.exists(p)) stop("Missing ", f, " - run script 20 first (it writes this)."); p }

mcfg <- read_csv(need(sprintf("metrics_config_%s.csv", TISSUE)), show_col_types = FALSE)
if (!all(c("metric", "scale", "tier", "arms", "covariates") %in% names(mcfg)))
  stop("metrics_config_", TISSUE, ".csv predates the ANCOVA version of script 20; re-run script 20.")
metrics         <- mcfg$metric
PRIMARY_METRICS <- mcfg$metric[mcfg$tier == "primary"]
arms_kept       <- strsplit(mcfg$arms[1], ";", fixed = TRUE)[[1]]
COVARIATES      <- strsplit(mcfg$covariates[1], ";", fixed = TRUE)[[1]]   # "treatment" [+ "secretion_state"]
USE_SECRETION_STATE <- "secretion_state" %in% COVARIATES
scale_of <- function(m) mcfg$scale[match(m, mcfg$metric)]
tier_of  <- function(m) mcfg$tier[match(m, mcfg$metric)]
on_scale <- function(m, v) if (scale_of(m) == "log") log(v) else v    # a value on its model scale

paired <- read_csv(need(sprintf("paired_sample_manifest_%s.csv", TISSUE)), show_col_types = FALSE) %>%
  mutate(treatment = factor(treatment, levels = arms_kept))
if (USE_SECRETION_STATE) {
  ## script 20 joined the per-animal byssal secretion state (differential-expression script
  ## 01_7) into the manifest; it is a nuisance covariate in every fit below, as `treatment` is.
  if (!"secretion_state" %in% names(paired) || anyNA(paired$secretion_state))
    stop("USE_SECRETION_STATE is on in script 20 but the manifest has no complete `secretion_state` column.")
  paired <- paired %>% mutate(secretion_state = factor(secretion_state, levels = c("off", "on")))
}
candidates_all <- read_csv(need(sprintf("candidate_genes_%s.csv", TISSUE)), show_col_types = FALSE)
candidates <- candidates_all %>% filter(tested)
floor_tbl  <- read_csv(need(sprintf("detection_floor_flags_%s.csv", TISSUE)), show_col_types = FALSE) %>%
  select(LOC_ID, frac_at_floor, floor_flag)
annot      <- read_csv(need("annotation_map.csv"), show_col_types = FALSE)
thr_raw    <- read_csv(need(sprintf("thread_plaques_paired_%s.csv", TISSUE)), show_col_types = FALSE)
vp_df <- read_csv(need(sprintf("vst_paired_%s.csv", TISSUE)), show_col_types = FALSE)
vp <- as.matrix(vp_df[, -1]); rownames(vp) <- vp_df[[1]]   # LOC_ID column -> matrix rownames
stopifnot(all(paired$sample %in% colnames(vp)))
vp <- vp[, paired$sample, drop = FALSE]                      # column order = manifest order
expr_of <- function(loc) as.numeric(vp[match(loc, rownames(vp)), paired$sample])

## script 20's reported test, one row per gene x metric, for the two gene universes
res_cand <- read_csv(need(sprintf("assoc_candidate_%s.csv", TISSUE)), show_col_types = FALSE)
if (!all(c("p_lm", "q_lm", "q_family", "baseline_slope") %in% names(res_cand)))
  stop("assoc_candidate_", TISSUE, ".csv predates the ANCOVA version of script 20; re-run script 20.")
deg_union_file <- file.path(IN_DIR, sprintf("assoc_DEGunion_%s.csv", TISSUE))
res_deg <- if (file.exists(deg_union_file)) read_csv(deg_union_file, show_col_types = FALSE) else NULL

## ---- the reported ANCOVA, exactly as script 20 fits it --------------------------------
## y and base on the metric's model scale; complete cases only (animals without baseline
## threads drop here); `mussel` rides along so the influence block can name an animal.
ancova_data <- function(x, metric) {          # x = one value per animal, manifest order
  dat <- data.frame(y      = on_scale(metric, paired[[metric]]),
                    x      = x,
                    base   = on_scale(metric, paired[[paste0("base_", metric)]]),
                    mussel = paired$mussel,
                    paired[, COVARIATES, drop = FALSE])
  dat[stats::complete.cases(dat) & is.finite(dat$y) & is.finite(dat$base), , drop = FALSE]
}
ANCOVA_FORMULA <- reformulate(c("x", COVARIATES, "base"), response = "y")
fit_ancova <- function(x, metric) {
  dat <- ancova_data(x, metric)
  if (nrow(dat) < 6) return(NULL)
  fm <- tryCatch(lm(ANCOVA_FORMULA, data = dat), error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!all(c("x", "base") %in% rownames(co))) return(NULL)
  list(fit = fm, data = dat, n = nrow(dat),
       slope = co["x", "Estimate"], se = co["x", "Std. Error"], p = co["x", "Pr(>|t|)"])
}

cat("Paired animals:", nrow(paired), "| in the fits (with baseline):", sum(!is.na(paired$base_max_force)),
    "| arms:", paste(arms_kept, collapse = ", "),
    "\nMetrics [scale, tier]:", paste(sprintf("%s [%s, %s]", mcfg$metric, mcfg$scale, mcfg$tier), collapse = ", "),
    "\nNuisance covariates:", paste(COVARIATES, collapse = " + "),
    "| candidates tested:", nrow(candidates), "of", nrow(candidates_all),
    "| DEG union tested:", if (is.null(res_deg)) 0L else dplyr::n_distinct(res_deg$LOC_ID), "\n")
```

---

# A. Thread-level mixed ANCOVA (sensitivity check on the reported test)

Every day-3 plaque of the paired animals, joined to the animal's expression and to its
baseline level (the same per-animal summary script 20 uses, on the same scale):

    plaque_day3 ~ expression + treatment [+ secretion_state] + level_baseline + (1 | mussel)

with `plaque_day3` = log(plaque value) for force, area and adhesion and the raw value for
extension. The expression and baseline terms vary only between animals, so their tests
reflect animal-level degrees of freedom (Satterthwaite); each animal contributes in
proportion to the precision of its mean, which is what the per-animal model cannot do.
Singular-fit warnings are expected when between-animal variance is near zero and do not
invalidate the fixed-effect estimate.

```{r mixed}
## raw day-3 plaque-level mechanics for the paired animals (script 20 exports phase == "post"),
## with the animal's baseline levels and any non-arm covariate joined by animal
thr_threads <- thr_raw %>%
  filter(phase == "post", mussel %in% paired$mussel, treatment %in% arms_kept) %>%
  transmute(mussel, treatment = factor(treatment, levels = arms_kept),
            max_force, pad_area, adhesion_kpa, max_displacement) %>%
  left_join(paired %>% select(mussel, starts_with("base_"), all_of(setdiff(COVARIATES, "treatment"))),
            by = "mussel")

MIXED_FORMULA <- reformulate(c("x", COVARIATES, "base", "(1|mussel)"), response = "y")
fit_mixed <- function(x, metric) {           # x = one value per animal, manifest order
  d0  <- thr_threads %>% inner_join(data.frame(mussel = paired$mussel, x = x), by = "mussel")
  dat <- data.frame(y      = on_scale(metric, d0[[metric]]),
                    x      = d0$x,
                    base   = on_scale(metric, d0[[paste0("base_", metric)]]),
                    mussel = d0$mussel,
                    d0[, COVARIATES, drop = FALSE])
  dat <- dat[stats::complete.cases(dat) & is.finite(dat$y) & is.finite(dat$base), , drop = FALSE]
  if (dplyr::n_distinct(dat$mussel) < 6) return(NULL)
  fm <- tryCatch(suppressWarnings(lmerTest::lmer(MIXED_FORMULA, data = dat, REML = TRUE)),
                 error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"x" %in% rownames(co)) return(NULL)
  tibble(n_threads = nrow(dat), n_mussel = dplyr::n_distinct(dat$mussel),
         slope_mixed = co["x", "Estimate"], p_mixed = co["x", "Pr(>|t|)"])
}

run_mixed <- function(loci, set_name) {
  map_dfr(metrics, function(mt) map_dfr(loci, function(loc) {
    r <- fit_mixed(expr_of(loc), mt)
    if (is.null(r)) return(NULL)
    bind_cols(tibble(LOC_ID = loc, metric = mt, scale = scale_of(mt), tier = tier_of(mt)), r)
  })) %>%
    group_by(metric) %>% mutate(q_mixed = p.adjust(p_mixed, "BH")) %>% ungroup() %>%
    left_join(annot, by = "LOC_ID") %>%
    mutate(gene_set = set_name) %>%
    arrange(factor(metric, levels = metrics), p_mixed)
}

res_cand_mixed <- run_mixed(candidates$LOC_ID, "candidate")
write_csv(res_cand_mixed, file.path(OUT_DIR, sprintf("assoc_candidate_MIXED_%s.csv", TISSUE)))

## the reported test (script 20) and the sensitivity check side by side
cand_compare <- res_cand %>%
  select(LOC_ID, metric, tier, n, protein_name, slope, p_lm, q_lm, q_family) %>%
  left_join(res_cand_mixed %>% select(LOC_ID, metric, n_mussel, slope_mixed, p_mixed, q_mixed),
            by = c("LOC_ID", "metric"))

cat("Candidate genes: reported ANCOVA (script 20, p_lm) beside the thread-level mixed ANCOVA (p_mixed), top 6 per metric:\n")
cand_compare %>% mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_lm, n = 6, with_ties = FALSE) %>% ungroup() %>%
  mutate(across(c(slope, slope_mixed), ~round(., 4)),
         across(c(p_lm, q_lm, q_family, p_mixed), ~signif(., 2)),
         protein_name = substr(protein_name, 1, 34)) %>%
  select(metric, tier, LOC_ID, protein_name, n, slope, p_lm, q_lm, q_family, slope_mixed, p_mixed) %>%
  as.data.frame() %>% print()

cat("\nAgreement between the two fits across the whole candidate family, per metric:\n")
cand_compare %>% filter(!is.na(p_mixed)) %>%
  mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>%
  summarise(n_genes         = n(),
            spearman_of_p   = round(cor(p_lm, p_mixed, method = "spearman"), 3),
            same_sign       = round(mean(sign(slope) == sign(slope_mixed)), 3),
            both_p_below_05 = sum(p_lm < 0.05 & p_mixed < 0.05),
            lm_only         = sum(p_lm < 0.05 & p_mixed >= 0.05),
            mixed_only      = sum(p_lm >= 0.05 & p_mixed < 0.05), .groups = "drop") %>%
  as.data.frame() %>% print()
```

---

# B. Module eigengenes (power-preserving)

Each module = PC1 of its members' VST values (per mussel), oriented so higher score = higher
module expression. Six modules x four metrics is FDR-tractable, and a coordinated pathway
score is biologically closer to "is the HIF response engaged" than any single gene. The
score goes through the same ANCOVA as a single gene (reported) and through the thread-level
mixed ANCOVA (sensitivity).

`byssal_structural` is the plaque/thread protein module (foot proteins, preCols, byssal
EP/ACDC): the genes that switch on together when an animal is actively secreting a thread.
It is kept separate from the broader `byssal_collagen` regex (which also catches every
collagen and every "byss*" annotation) so that the direct test of "these genes track thread
building, not strength" is its own row.

```{r modules}
modules <- list(
  HIF_hypoxia       = "hypoxia-inducible|Egl nine|prolyl 4-hydroxylase|prolyl hydroxylase|hypoxia up-regulated",
  oxidative_detox   = "superoxide|glutathione|catalase|peroxiredoxin|dual oxidase|peroxidase",
  HSP_proteostasis  = "heat shock|Hsp|stress-70|chaperone|BiP|peptidyl-prolyl",
  tRNA_translation  = "--tRNA|tRNA ligase|aminoacyl",
  byssal_collagen   = "collagen|byss|foot protein|adhesive plaque|polyphenolic",
  byssal_structural = "foot protein|adhesive plaque|polyphenolic|byssal protein-|preCol|precollagen|byssal EP|ACDC"
)

if (!"blast_ok" %in% names(annot)) annot$blast_ok <- TRUE   # older script-20 handoff
module_score <- function(rx) {
  locs <- annot %>% filter(blast_ok, grepl(rx, protein_name, ignore.case = TRUE)) %>% pull(LOC_ID)
  locs <- intersect(locs, rownames(vp))
  if (length(locs) < 3) return(NULL)
  z <- t(scale(t(vp[locs, paired$sample, drop = FALSE])))   # standardise each gene
  z <- z[stats::complete.cases(z), , drop = FALSE]
  pc <- prcomp(t(z), center = TRUE, scale. = FALSE)
  sc <- pc$x[, 1]
  if (suppressWarnings(cor(sc, colMeans(vp[locs, paired$sample]))) < 0) sc <- -sc
  list(score = sc, n_genes = length(locs), members = locs,
       var_expl = summary(pc)$importance[2, 1])
}
mod_scores <- purrr::compact(map(modules, module_score))
write_csv(map_dfr(names(mod_scores), ~ tibble(module = .x, LOC_ID = mod_scores[[.x]]$members)) %>%
            left_join(annot, by = "LOC_ID"),
          file.path(OUT_DIR, sprintf("module_members_%s.csv", TISSUE)))

## the same ANCOVA as script 20 on the module score (reported), plus the thread-level mixed
## ANCOVA as the sensitivity column
mod_fits <- imap_dfr(mod_scores, function(ms, nm) {
  map_dfr(metrics, function(mt) {
    a <- fit_ancova(ms$score, mt)
    if (is.null(a)) return(NULL)
    m <- fit_mixed(ms$score, mt)
    tibble(module = nm, n_genes = ms$n_genes, pc1_var = round(ms$var_expl, 2),
           metric = mt, scale = scale_of(mt), tier = tier_of(mt), n = a$n,
           slope = a$slope, se = a$se, p_lm = a$p,
           slope_mixed = if (is.null(m)) NA_real_ else m$slope_mixed,
           p_mixed     = if (is.null(m)) NA_real_ else m$p_mixed)
  })
}) %>%
  group_by(metric) %>% mutate(q_lm = p.adjust(p_lm, "BH"), q_mixed = p.adjust(p_mixed, "BH")) %>% ungroup() %>%
  group_by(tier)   %>% mutate(q_family = p.adjust(p_lm, "BH")) %>% ungroup() %>%
  arrange(p_lm)
write_csv(mod_fits, file.path(OUT_DIR, sprintf("module_associations_%s.csv", TISSUE)))

cat("MODULE-level associations (ANCOVA on the module score; q_lm = BH across the ",
    n_distinct(mod_fits$module), " modules within each metric, q_family = BH within each tier):\n", sep = "")
mod_fits %>% mutate(across(c(slope, slope_mixed), ~round(., 4)),
                    across(c(p_lm, q_lm, q_family, p_mixed), ~signif(., 2))) %>%
  select(module, n_genes, pc1_var, metric, tier, n, slope, p_lm, q_lm, q_family, p_mixed) %>%
  as.data.frame() %>% print()
```

---

# C. Diagnostics: detection floor and influence; the best-hits table

Genes where many paired samples sit at the VST minimum are effectively absent/zero-inflated;
their correlations are driven by detection, not biology. Script 20 flags them and, by
default, keeps them out of the tested family; the flags are echoed here. Also report the
most influential animal per top hit (Cook's distance) so you know if one mussel drives a fit.

The tables are written to `03_analyses/gene_mechanics/` so the report and any downstream
filter can use them without re-running this script:

- `detection_floor_flags_<T>.csv` (written by script 20): one row per gene in the candidate
  set and the DEG union, with the fraction of paired samples at the VST floor, a
  `floor_flag` of `exclude` (> 0.4), `caution` (> 0.2) or `ok`, and `tested` (FALSE for the
  `exclude` genes when script 20's `EXCLUDE_FLOOR_GENES` is TRUE: they are out of the family
  before BH and appear in no association table).
- `influence_top_hits_<T>.csv`: the three best candidate hits and the best DEG-union hit per
  metric (ranked by the reported `p_lm`) with the most influential animal, its Cook's
  distance, the slope and p after dropping that animal, and an `influence_flag` (`fragile`
  when the hit loses p < 0.05 or its slope moves by more than half).
- `best_hits_<T>.csv`: the single best hit per metric for each gene set (candidate, module,
  DEG union), with `p_lm`, `q_lm`, `q_family`, `p_mixed`, the floor flag and the influence
  flag in one row. Nothing should be quoted from a p alone.

```{r diagnostics}
## detection floor: computed and applied by script 20 (detection_floor_flags_<T>.csv); shown
## here so the report carries it beside the influence table
floor_flag <- read_csv(need(sprintf("detection_floor_flags_%s.csv", TISSUE)), show_col_types = FALSE)
cat("Detection-floor summary (", nrow(floor_flag), " genes in the two universes; `tested` = in the family):\n", sep = "")
print(table(gene_set = floor_flag$gene_set, floor_flag = floor_flag$floor_flag, tested = floor_flag$tested))

## influence: the best hits per metric, which mussel has the largest Cook's distance in the
## reported ANCOVA, and what the fit looks like without it
top_hits <- bind_rows(
  res_cand %>% group_by(metric) %>% slice_min(p_lm, n = N_INFLUENCE_HITS, with_ties = FALSE) %>% ungroup() %>%
    transmute(gene_set = "candidate", metric, LOC_ID, p_lm, q_lm, q_family),
  if (!is.null(res_deg)) res_deg %>% group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
    transmute(gene_set = "DEG_union", metric, LOC_ID, p_lm, q_lm, q_family))

infl <- pmap_dfr(top_hits, function(gene_set, metric, LOC_ID, p_lm, q_lm, q_family) {
  a <- fit_ancova(expr_of(LOC_ID), metric)
  if (is.null(a)) return(NULL)
  cd   <- cooks.distance(a$fit)
  drop <- which.max(cd)
  fm_drop <- lm(ANCOVA_FORMULA, data = a$data[-drop, ])
  b_all <- a$slope; b_drop <- unname(coef(fm_drop)["x"])
  p_drop <- summary(fm_drop)$coefficients["x", "Pr(>|t|)"]
  tibble(gene_set, metric, LOC_ID,
         p_lm = signif(p_lm, 3), q_lm = signif(q_lm, 3), q_family = signif(q_family, 3),
         n                       = a$n,
         most_influential_mussel = a$data$mussel[drop],
         max_cooks_D             = round(max(cd), 3),
         cook_threshold_4n       = round(4 / a$n, 3),
         ## the max of n Cook's distances exceeds 4/n in almost every fit of this size, so
         ## that screen alone is uninformative here; D > 1 is the conventional "one point
         ## moves the fit" level (Cook & Weisberg 1982)
         cooks_flag              = case_when(max(cd) > 1       ~ "D>1",
                                             max(cd) > 4 / a$n ~ "D>4/n",
                                             TRUE              ~ "ok"),
         slope                   = signif(b_all, 4),
         slope_without_mussel    = signif(b_drop, 4),
         p_without_mussel        = signif(p_drop, 3),
         slope_change_frac       = round(abs(b_drop - b_all) / abs(b_all), 3),
         ## "fragile": the result depends on one animal (loses p < 0.05, or the slope moves
         ## by more than half) - report such a hit with the leave-one-out slope beside it
         influence_flag          = ifelse((p_lm < 0.05 & p_drop >= 0.05) |
                                            abs(b_drop - b_all) / abs(b_all) > 0.5,
                                          "fragile", "robust"))
}) %>%
  left_join(annot %>% select(LOC_ID, protein_name, blast_pident), by = "LOC_ID") %>%
  mutate(tier = tier_of(metric),
         metric = factor(metric, levels = metrics),
         gene_set = factor(gene_set, levels = c("candidate", "DEG_union"))) %>%
  arrange(gene_set, metric, p_lm) %>%
  mutate(metric = as.character(metric), gene_set = as.character(gene_set)) %>%
  select(gene_set, metric, tier, LOC_ID, protein_name, n, p_lm, q_lm, q_family, everything())
write_csv(infl, file.path(OUT_DIR, sprintf("influence_top_hits_%s.csv", TISSUE)))

cat("\nMost influential animal for the top", N_INFLUENCE_HITS, "candidate hits and the best DEG-union hit per metric",
    "('fragile' = loses p < 0.05 or slope moves > 50% when that animal is dropped):\n")
infl %>% select(gene_set, metric, LOC_ID, p_lm, q_lm, most_influential_mussel, max_cooks_D,
                cooks_flag, slope, slope_without_mussel, p_without_mussel, influence_flag) %>%
  as.data.frame() %>% print()

## ---- best hit per metric and gene set, everything on one row --------------------------
best_cand <- res_cand %>% group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  left_join(res_cand_mixed %>% select(LOC_ID, metric, p_mixed), by = c("LOC_ID", "metric")) %>%
  transmute(gene_set = "candidate", metric, tier, feature = LOC_ID, protein_name,
            n, slope, p_lm, q_lm, q_family, p_mixed)
best_mod <- mod_fits %>% group_by(metric) %>% slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  transmute(gene_set = "module", metric, tier, feature = module,
            protein_name = paste0(n_genes, "-gene module"),
            n, slope, p_lm, q_lm, q_family, p_mixed)
best_deg <- if (!is.null(res_deg)) res_deg %>% group_by(metric) %>%
  slice_min(p_lm, n = 1, with_ties = FALSE) %>% ungroup() %>%
  mutate(p_mixed = map2_dbl(LOC_ID, metric, ~ { m <- fit_mixed(expr_of(.x), .y)
                                                if (is.null(m)) NA_real_ else m$p_mixed })) %>%
  transmute(gene_set = "DEG_union", metric, tier, feature = LOC_ID, protein_name,
            n, slope, p_lm, q_lm, q_family, p_mixed) else NULL
best_hits <- bind_rows(best_cand, best_mod, best_deg) %>%
  left_join(floor_tbl %>% select(feature = LOC_ID, floor_flag), by = "feature") %>%
  left_join(infl %>% select(feature = LOC_ID, metric, most_influential_mussel, influence_flag),
            by = c("feature", "metric")) %>%
  mutate(metric = factor(metric, levels = metrics),
         gene_set = factor(gene_set, levels = c("candidate", "module", "DEG_union"))) %>%
  arrange(gene_set, metric) %>% mutate(metric = as.character(metric), gene_set = as.character(gene_set)) %>%
  select(gene_set, metric, tier, feature, protein_name, n, slope, p_lm, q_lm, q_family, p_mixed,
         floor_flag, most_influential_mussel, influence_flag)
write_csv(best_hits, file.path(OUT_DIR, sprintf("best_hits_%s.csv", TISSUE)))

cat("\nBest hit per metric and gene set (the reported p beside both q's and the mixed-model p):\n")
best_hits %>% mutate(across(c(slope), ~signif(., 3)),
                     across(c(p_lm, q_lm, q_family, p_mixed), ~signif(., 2)),
                     protein_name = substr(protein_name, 1, 30)) %>%
  select(gene_set, metric, tier, feature, protein_name, n, slope, p_lm, q_lm, q_family, p_mixed,
         floor_flag, influence_flag) %>%
  as.data.frame() %>% print()
```

---

# D. Provenance

Script 20 opens `RUN_provenance_<T>.txt` with its settings and the git commit of the thread
input; this block appends this script's settings, so the folder records the configuration
behind every file in it.

```{r provenance}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
cat(c("",
      "[21-gene_mechanics_expanded.Rmd]",
      paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
      paste0("reported test        : script 20's ANCOVA (assoc_candidate_", TISSUE, ".csv, assoc_DEGunion_", TISSUE, ".csv)"),
      paste0("sensitivity model    : plaque ~ expression + ", paste(COVARIATES, collapse = " + "),
             " + level_baseline + (1|mussel), thread level, Satterthwaite df"),
      paste0("metrics [scale, tier]: ", paste(sprintf("%s [%s, %s]", mcfg$metric, mcfg$scale, mcfg$tier), collapse = ", ")),
      paste0("primary metrics      : ", paste(PRIMARY_METRICS, collapse = ", ")),
      paste0("nuisance covariates  : ", paste(COVARIATES, collapse = " + ")),
      paste0("modules              : ", paste(sprintf("%s (%d)", names(mod_scores), sapply(mod_scores, `[[`, "n_genes")), collapse = ", ")),
      paste0("FDR_ALPHA            : ", FDR_ALPHA),
      paste0("candidates tested    : ", nrow(candidates), " of ", nrow(candidates_all), " (floor filter in script 20)"),
      paste0("N_INFLUENCE_HITS     : ", N_INFLUENCE_HITS),
      paste0("R / lme4 / lmerTest  : ", R.version.string, " / ",
             as.character(packageVersion("lme4")), " / ", as.character(packageVersion("lmerTest")))),
    file = prov, sep = "\n", append = file.exists(prov))
cat(readLines(prov), sep = "\n")
```
