---
title: "21 - Gene x mechanics: expanded analysis (mixed models, modules, permutation)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# What this adds to script 20

Script 20 correlated each mussel's **unweighted mean** plaque mechanics against expression.
That uses the mussel as the unit (correct, no pseudoreplication) but ignores that some
mussels have only 1-2 plaques, which inflates the apparent significance of the best hits.
This script:

- **A. Thread-level mixed model** - keeps every plaque, models `mussel` as a random
  intercept, so the expression effect is estimated between mussels and weighted by each
  mussel's plaque count. **This is the headline fix for the pooling issue.**
- **B. Precision-weighted regression** - a transparent, fast alternative (weights =
  n_plaques) used for the permutation loop.
- **C. Module eigengenes** - collapse co-regulated pathways (HIF, oxidative, HSP,
  tRNA-synthetase, byssal/collagen) into one score each, cutting the multiple-testing
  burden from hundreds of genes to ~5 modules and matching the biology.
- **D. Permutation null** - an honest "is the best hit better than chance?" p-value that
  accounts for searching across many genes AND their correlation structure.
- **E. Diagnostics** - flags genes sitting at the detection floor (whose correlations are
  driven by zeros) and influential animals.

> **Prerequisite:** run script 20 first. This script no longer shares 20's R session; it
> reads 20's CSV handoffs (`paired_sample_manifest`, `vst_paired_<T>`, `annotation_map`,
> `candidate_genes_<T>`, `thread_plaques_paired_<T>`) from `03_analyses/gene_mechanics/`.

```{r setup_inputs}
suppressPackageStartupMessages({ library(tidyverse); library(lmerTest); library(here) })
## lmerTest::lmer gives Satterthwaite p-values

## ---- config: must match the values script 20 ran with ------------------------------
TISSUE    <- "F"                                   # "F" foot (default) or "G" gill
FDR_ALPHA <- 0.10
## The metric list and the treatment levels are read from script 20's metrics_config.csv
## handoff below, so the two scripts cannot drift apart.

## ---- read script 20's CSV handoffs (run 20 first) ----------------------------------
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
IN_DIR  <- file.path(find_repo_root(), "gene-mechanics-correlation", "03_analyses", "gene_mechanics")
OUT_DIR <- IN_DIR                                  # 21 writes alongside 20's outputs
need <- function(f) { p <- file.path(IN_DIR, f)
  if (!file.exists(p)) stop("Missing ", f, " - run script 20 first (it writes this)."); p }

mcfg <- read_csv(need("metrics_config.csv"), show_col_types = FALSE)
metrics        <- mcfg$metric
LEVEL_METRICS  <- mcfg$metric[mcfg$type == "level"]
CHANGE_METRICS <- mcfg$metric[mcfg$type == "change"]
arms_kept      <- strsplit(mcfg$arms[1], ";", fixed = TRUE)[[1]]

paired <- read_csv(need("paired_sample_manifest.csv"), show_col_types = FALSE) %>%
  mutate(treatment = factor(treatment, levels = arms_kept))
candidates <- read_csv(need(sprintf("candidate_genes_%s.csv", TISSUE)), show_col_types = FALSE)
annot      <- read_csv(need("annotation_map.csv"), show_col_types = FALSE)
thr_raw    <- read_csv(need(sprintf("thread_plaques_paired_%s.csv", TISSUE)), show_col_types = FALSE)
vp_df <- read_csv(need(sprintf("vst_paired_%s.csv", TISSUE)), show_col_types = FALSE)
vp <- as.matrix(vp_df[, -1]); rownames(vp) <- vp_df[[1]]   # LOC_ID column -> matrix rownames

cat("Paired animals:", nrow(paired), "| arms:", paste(arms_kept, collapse = ", "),
    "| level metrics:", paste(LEVEL_METRICS, collapse = ", "),
    "| change metrics:", paste(CHANGE_METRICS, collapse = ", "), "\n")
```

---

# A. Thread-level mixed model (the pooling fix)

Raw plaque-level data, joined to each mussel's expression. Model:
`metric ~ expression + treatment + (1 | mussel)`. The expression term varies only between
mussels, so its test reflects ~31 mussel-level df (Satterthwaite), not the thread count;
mussels with more/precise plaques contribute more. Singular-fit warnings are expected when
between-mussel variance is near zero and do not invalidate the fixed-effect estimate.

The thread-level model applies to the LEVEL metrics only: a CHANGE metric (`dlog_*`) is one
number per animal by construction, so for those the per-animal weighted regression in block
B is the right tool and is what is reported.

```{r mixed}
## raw day-3 plaque-level mechanics for the paired animals (script 20 exports phase == "post")
thr_threads <- thr_raw %>%
  filter(phase == "post", mussel %in% paired$mussel, treatment %in% arms_kept) %>%
  transmute(mussel,
            treatment = factor(treatment, levels = arms_kept),
            max_force, pad_area, adhesion_kpa, max_displacement)

test_gene_mixed <- function(loc, metric) {
  ev  <- data.frame(mussel = paired$mussel,
                    expr = as.numeric(vp[match(loc, rownames(vp)), paired$sample]))
  dat <- thr_threads %>% inner_join(ev, by = "mussel")
  dat <- dat[stats::complete.cases(dat[, c(metric, "expr", "treatment", "mussel")]), ]
  if (dplyr::n_distinct(dat$mussel) < 6) return(NULL)
  fm <- tryCatch(suppressWarnings(
          lmerTest::lmer(reformulate(c("expr", "treatment", "(1|mussel)"),
                                     response = metric),
                         data = dat, REML = TRUE)),
        error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"expr" %in% rownames(co)) return(NULL)
  tibble(LOC_ID = loc, metric = metric,
         n_threads = nrow(dat), n_mussel = dplyr::n_distinct(dat$mussel),
         slope_mixed = co["expr", "Estimate"], p_mixed = co["expr", "Pr(>|t|)"])
}

run_mixed <- function(loci, set_name) {
  map_dfr(LEVEL_METRICS, function(mt) map_dfr(loci, ~ test_gene_mixed(.x, mt))) %>%
    group_by(metric) %>% mutate(q_mixed = p.adjust(p_mixed, "BH")) %>% ungroup() %>%
    left_join(annot, by = "LOC_ID") %>% mutate(gene_set = set_name) %>%
    arrange(metric, p_mixed)
}

## candidate set: already loaded from script 20's candidate_genes_<T>.csv handoff above

res_cand_mixed <- run_mixed(candidates$LOC_ID, "candidate")
write_csv(res_cand_mixed, file.path(OUT_DIR, sprintf("assoc_candidate_MIXED_%s.csv", TISSUE)))

cat("Candidate genes, MIXED model, top per LEVEL metric (compare p_mixed with script-20 p_lm):\n")
res_cand_mixed %>% mutate(metric = factor(metric, levels = LEVEL_METRICS)) %>%
  group_by(metric) %>% slice_min(p_mixed, n = 6) %>% ungroup() %>%
  mutate(across(c(slope_mixed), ~round(., 4)),
         p_mixed = signif(p_mixed, 2), q_mixed = signif(q_mixed, 2)) %>%
  select(metric, LOC_ID, protein_name, n_mussel, slope_mixed, p_mixed, q_mixed) %>%
  as.data.frame() %>% print()
```

---

# B. Precision-weighted regression (fast, for permutation)

Per-mussel means weighted by plaque count. Nearly equivalent to the mixed model but cheap
to refit thousands of times.

For a CHANGE metric the weight is the smaller of the animal's baseline and day-3 plaque
counts, since the log ratio is only as precise as its less-replicated side.

```{r weighted}
wls_weight <- function(metric) {
  if (metric %in% CHANGE_METRICS) pmin(paired$n_plaques, paired$n_plaques_base) else paired$n_plaques
}
test_gene_wls <- function(loc, metric) {
  x <- as.numeric(vp[match(loc, rownames(vp)), paired$sample])
  dat <- data.frame(y = paired[[metric]], x = x,
                    treatment = paired$treatment, w = wls_weight(metric))
  dat <- dat[stats::complete.cases(dat), ]
  if (nrow(dat) < 6) return(NULL)
  fm <- tryCatch(lm(y ~ x + treatment, data = dat, weights = w), error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"x" %in% rownames(co)) return(NULL)
  tibble(LOC_ID = loc, metric = metric, n = nrow(dat),
         slope_wls = co["x", "Estimate"], p_wls = co["x", "Pr(>|t|)"])
}
res_cand_wls <- map_dfr(metrics, function(mt) map_dfr(candidates$LOC_ID, ~ test_gene_wls(.x, mt))) %>%
  group_by(metric) %>% mutate(q_wls = p.adjust(p_wls, "BH")) %>% ungroup() %>%
  left_join(annot, by = "LOC_ID")
write_csv(res_cand_wls, file.path(OUT_DIR, sprintf("assoc_candidate_WLS_%s.csv", TISSUE)))

cat("Candidate genes, weighted per-animal regression, top per metric (this is the reported\n",
    "test for the CHANGE metrics):\n")
res_cand_wls %>% mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_wls, n = 4) %>% ungroup() %>%
  mutate(slope_wls = round(slope_wls, 4), p_wls = signif(p_wls, 2), q_wls = signif(q_wls, 2)) %>%
  select(metric, n, LOC_ID, protein_name, slope_wls, p_wls, q_wls) %>%
  as.data.frame() %>% print()
```

---

# C. Module eigengenes (power-preserving)

Each module = PC1 of its members' VST values (per mussel), oriented so higher score = higher
module expression. ~5 modules x 3 metrics is FDR-tractable, and a coordinated pathway score
is biologically closer to "is the HIF response engaged" than any single gene.

```{r modules}
modules <- list(
  HIF_hypoxia      = "hypoxia-inducible|Egl nine|prolyl 4-hydroxylase|prolyl hydroxylase|hypoxia up-regulated",
  oxidative_detox  = "superoxide|glutathione|catalase|peroxiredoxin|dual oxidase|peroxidase",
  HSP_proteostasis = "heat shock|Hsp|stress-70|chaperone|BiP|peptidyl-prolyl",
  tRNA_translation = "--tRNA|tRNA ligase|aminoacyl",
  byssal_collagen  = "collagen|byss|foot protein|adhesive plaque|polyphenolic"
)

module_score <- function(rx) {
  locs <- annot %>% filter(grepl(rx, protein_name, ignore.case = TRUE)) %>% pull(LOC_ID)
  locs <- intersect(locs, rownames(vp))
  if (length(locs) < 3) return(NULL)
  z <- t(scale(t(vp[locs, paired$sample, drop = FALSE])))   # standardise each gene
  z <- z[stats::complete.cases(z), , drop = FALSE]
  pc <- prcomp(t(z), center = TRUE, scale. = FALSE)
  sc <- pc$x[, 1]
  if (suppressWarnings(cor(sc, colMeans(vp[locs, paired$sample]))) < 0) sc <- -sc
  list(score = sc, n_genes = length(locs),
       var_expl = summary(pc)$importance[2, 1])
}

mod_fits <- imap_dfr(modules, function(rx, nm) {
  ms <- module_score(rx); if (is.null(ms)) return(NULL)
  dat0 <- tibble(mussel = paired$mussel, score = ms$score)
  map_dfr(metrics, function(mt) {
    if (mt %in% LEVEL_METRICS) {
      ## thread-level mixed model, random intercept for animal
      dat <- thr_threads %>% inner_join(dat0, by = "mussel")
      fm <- tryCatch(suppressWarnings(
              lmerTest::lmer(reformulate(c("score", "treatment", "(1|mussel)"),
                                         response = mt), data = dat, REML = TRUE)),
            error = function(e) NULL)
      if (is.null(fm)) return(NULL)
      co <- summary(fm)$coefficients
      tibble(module = nm, n_genes = ms$n_genes, pc1_var = round(ms$var_expl, 2),
             metric = mt, model = "mixed", n = dplyr::n_distinct(dat$mussel),
             slope = co["score", "Estimate"], p = co["score", "Pr(>|t|)"])
    } else {
      ## change metric: one value per animal, weighted regression
      dat <- paired %>% select(mussel, treatment, y = all_of(mt)) %>%
        inner_join(dat0, by = "mussel") %>%
        mutate(w = wls_weight(mt)[match(mussel, paired$mussel)]) %>%
        filter(!is.na(y))
      if (nrow(dat) < 6) return(NULL)
      fm <- tryCatch(lm(y ~ score + treatment, data = dat, weights = w), error = function(e) NULL)
      if (is.null(fm)) return(NULL)
      co <- summary(fm)$coefficients
      tibble(module = nm, n_genes = ms$n_genes, pc1_var = round(ms$var_expl, 2),
             metric = mt, model = "wls", n = nrow(dat),
             slope = co["score", "Estimate"], p = co["score", "Pr(>|t|)"])
    }
  })
}) %>% group_by(metric) %>% mutate(q = p.adjust(p, "BH")) %>% ungroup() %>% arrange(p)

write_csv(mod_fits, file.path(OUT_DIR, sprintf("module_associations_%s.csv", TISSUE)))
cat("MODULE-level associations (mixed model, BH across the ", n_distinct(mod_fits$module),
    " modules x metric):\n", sep = "")
mod_fits %>% mutate(slope = round(slope, 4), p = signif(p, 2), q = signif(q, 2)) %>%
  as.data.frame() %>% print()
```

---

# D. Permutation null for the best candidate hit

Shuffle the expression-to-animal assignment **within treatment** (preserving treatment
structure and gene-gene correlation), recompute the minimum p across the candidate set,
repeat. The empirical p tells you whether your best observed hit beats what the search
itself produces by chance. Uses the fast weighted lm.

```{r permutation}
set.seed(1)
NPERM    <- 1000
PMETRICS <- c("max_force", "pad_area", "dlog_adhesion_kpa")   # the two primary levels + the
                                                              # change metric with the best raw hit

perm_test <- function(pm) {
  ok  <- !is.na(paired[[pm]])
  E   <- vp[candidates$LOC_ID, paired$sample[ok], drop = FALSE]   # genes x mussels
  y   <- paired[[pm]][ok]; w <- wls_weight(pm)[ok]; trt <- droplevels(paired$treatment[ok])
  idx_by_trt <- split(seq_along(y), trt)
  min_p_for_order <- function(perm) {
    Ep <- E[, perm, drop = FALSE]
    ps <- apply(Ep, 1, function(xx) {
      f <- tryCatch(lm(y ~ xx + trt, weights = w), error = function(e) NULL)
      if (is.null(f)) return(NA_real_)
      cc <- summary(f)$coefficients
      if ("xx" %in% rownames(cc)) cc["xx", "Pr(>|t|)"] else NA_real_
    })
    min(ps, na.rm = TRUE)
  }
  obs_min_p <- min_p_for_order(seq_along(y))
  null_min_p <- replicate(NPERM, {
    perm <- seq_along(y)
    for (g in idx_by_trt) perm[g] <- sample(g)
    min_p_for_order(perm)
  })
  tibble(metric = pm, n = sum(ok), n_genes = nrow(E), n_perm = NPERM,
         observed_min_p = obs_min_p,
         p_permutation  = (sum(null_min_p <= obs_min_p) + 1) / (NPERM + 1))
}

perm_res <- map_dfr(PMETRICS, perm_test)
write_csv(perm_res, file.path(OUT_DIR, sprintf("permutation_best_hit_%s.csv", TISSUE)))
print(as.data.frame(perm_res))
cat("Interpretation: p_permutation is the chance of a candidate-set minimum p this small when",
    "expression is shuffled among animals within arm. If it is >> 0.05, the best hit is",
    "consistent with noise given the search.\n")
```

---

# E. Diagnostics: detection floor and influence

Genes where many paired samples sit at the VST minimum are effectively absent/zero-inflated;
their correlations are driven by detection, not biology (e.g. the multidrug-resistance gene
in the script-20 scatter pile-up at x ~ 3.5). Flag and optionally drop them. Also report the
most influential animal per top hit (Cook's distance) so you know if one mussel drives a fit.

```{r diagnostics}
floor_val <- min(vp)
frac_at_floor <- apply(vp[candidates$LOC_ID, paired$sample, drop = FALSE], 1,
                       function(x) mean(abs(x - floor_val) < 1e-6))
floor_flag <- tibble(LOC_ID = candidates$LOC_ID,
                     protein_name = candidates$protein_name,
                     frac_samples_at_floor = round(frac_at_floor, 2)) %>%
  arrange(desc(frac_samples_at_floor))
cat("Candidate genes most affected by detection floor (consider excluding > ~0.4):\n")
floor_flag %>% filter(frac_samples_at_floor > 0.2) %>% as.data.frame() %>% print()

## influence: top hit per metric, which mussel has the largest Cook's distance
top_by_metric <- res_cand_mixed %>% group_by(metric) %>% slice_min(p_mixed, n = 1) %>% ungroup()
infl <- pmap_dfr(top_by_metric, function(LOC_ID, metric, ...) {
  x <- as.numeric(vp[match(LOC_ID, rownames(vp)), paired$sample])
  fm <- lm(paired[[metric]] ~ x + paired$treatment)
  cd <- cooks.distance(fm)
  tibble(metric = metric, LOC_ID = LOC_ID,
         most_influential_mussel = paired$mussel[which.max(cd)],
         max_cooks_D = round(max(cd), 2),
         cook_threshold_4n = round(4 / nrow(paired), 2))
})
cat("\nMost influential animal for each top hit (flag if max_cooks_D > 4/n):\n")
as.data.frame(infl) %>% print()
```

---

# F. Baseline-adjusted level models

Adding each animal's baseline as a covariate asks whether expression relates to the day-3
level *given where the animal started*. The direct test of whether expression tracks the
*change* is the `dlog_*` metrics in blocks B and C, which need no covariate. Both are kept
because they answer slightly different questions (ANCOVA versus change score) and can
disagree when baseline and change are correlated.

```{r baseline-adjusted}
base_lookup <- paired %>% select(mussel, starts_with("base_"))
test_gene_mixed_adj <- function(loc, metric) {
  bcol <- paste0("base_", metric)
  ev <- data.frame(mussel = paired$mussel,
                   expr = as.numeric(vp[match(loc, rownames(vp)), paired$sample]))
  dat <- thr_threads %>% inner_join(ev, by = "mussel") %>%
    left_join(base_lookup, by = "mussel")
  dat$base_val <- dat[[bcol]]                       # was: rename(base = !!bcol)
  keep_cols <- c(metric, "expr", "treatment", "mussel", "base_val")
  dat <- dat[stats::complete.cases(dat[, keep_cols]), ]
  if (dplyr::n_distinct(dat$mussel) < 6) return(NULL)
  fm <- tryCatch(suppressWarnings(
          lmerTest::lmer(reformulate(c("expr", "treatment", "base_val", "(1|mussel)"),
                                     response = metric), data = dat, REML = TRUE)),
        error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"expr" %in% rownames(co)) return(NULL)
  tibble(LOC_ID = loc, metric = metric, n_mussel = dplyr::n_distinct(dat$mussel),
         slope = co["expr", "Estimate"], p = co["expr", "Pr(>|t|)"])
}

res_adj <- map_dfr(LEVEL_METRICS, function(mt) map_dfr(candidates$LOC_ID, ~ test_gene_mixed_adj(.x, mt))) %>%
  group_by(metric) %>% mutate(q = p.adjust(p, "BH")) %>% ungroup() %>%
  left_join(annot, by = "LOC_ID") %>% arrange(metric, p)
write_csv(res_adj, file.path(OUT_DIR, sprintf("assoc_candidate_BASELINEADJ_%s.csv", TISSUE)))
cat("Baseline-adjusted candidate top hits (expression vs CHANGE in attachment):\n")
res_adj %>% group_by(metric) %>% slice_min(p, n = 5) %>% ungroup() %>%
  mutate(slope = round(slope, 4), p = signif(p, 2), q = signif(q, 2)) %>%
  select(metric, LOC_ID, protein_name, n_mussel, slope, p, q) %>%
  as.data.frame() %>% print()
```

```{r close, echo=FALSE, results='asis'}
cat("
**Reading these results together**

- Compare `p_mixed` (block A) with script-20's `p_lm` for the LEVEL metrics. Where they
  diverge, trust the mixed model; divergence means single-plaque animals were driving the
  original p. For the CHANGE metrics the reported test is the weighted regression in block B.
- The module table (block C) is your best-powered test. If, say, `HIF_hypoxia` vs
  `max_force` is significant after BH while no single gene is, that is a stronger and more
  defensible statement than any individual correlation.
- The permutation p (block D) is the number to quote when asked 'is your best hit real?'.
- Genes flagged at the detection floor (block E) should be dropped before interpretation.
- None of this raises the ceiling set by n = 44; it makes the inference honest and tells you
  which signals merit a powered follow-up (paired RNA + mechanics on more animals, or qPCR
  of the HIF/oxidative module against thread strength in a larger cohort).
")
```
