---
title: "21 - Gene x mechanics: expanded analysis (mixed models, modules, permutation)"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  tissue: "F"   # "F" foot (default) or "G" gill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
```

# What this adds to script 20

Script 20 correlated each mussel's **unweighted mean** plaque mechanics against expression.
That uses the mussel as the unit (correct, no pseudoreplication) but ignores that some
mussels have only 1-2 plaques, which inflates the apparent significance of the best hits.
This script:

- **A. Thread-level mixed model** - keeps every plaque, models `mussel` as a random
  intercept, so the expression effect is estimated between mussels and weighted by each
  mussel's plaque count. Reported as a *sensitivity* check on the level metrics.
- **B. Precision-weighted regression (the reported test)** - per-animal means weighted by
  plaque count (weights = n_plaques; for a change metric the smaller of the two plaque
  counts). This is the statistic that is ranked, BH-corrected **and permuted**, for every
  metric, so a reported p and its permutation companion always refer to the same test.
- **C. Module eigengenes** - collapse co-regulated pathways (HIF, oxidative, HSP,
  tRNA-synthetase, byssal/collagen, and the byssal structural module) into one score each,
  cutting the multiple-testing burden from hundreds of genes to six modules.
- **D. Permutation null** - "is the best hit better than chance?" p-values that account for
  searching across genes, across metrics AND their correlation structure: per metric, across
  the whole search (genes x metrics), and for the pre-declared primary family (candidate
  genes x primary metrics). Run for the candidate set, the modules and the DEG union, with
  the same shuffles, and the null vectors are written to disk.
- **E. Diagnostics** - flags genes sitting at the detection floor (whose correlations are
  driven by zeros) and influential animals; writes `best_hits_<T>.csv`, one row per metric
  and gene set, with q and every permutation p side by side.

> **Prerequisite:** run script 20 first. This script no longer shares 20's R session; it
> reads 20's CSV handoffs (`paired_sample_manifest_<T>`, `vst_paired_<T>`, `annotation_map`,
> `candidate_genes_<T>`, `thread_plaques_paired_<T>`, `metrics_config_<T>`,
> `assoc_DEGunion_<T>`) from `03_analyses/gene_mechanics/`.

```{r setup_inputs}
suppressPackageStartupMessages({ library(tidyverse); library(lmerTest); library(here) })
## lmerTest::lmer gives Satterthwaite p-values

## ---- config -------------------------------------------------------------------------
## Tissue comes from the knit parameter (`params$tissue`, default "F") so the same script
## serves foot and gill; the driver 24-run_gene_mechanics_by_tissue.Rmd renders both.
## When run chunk-by-chunk outside a knit, `params` may not exist; fall back to foot.
TISSUE <- if (exists("params") && !is.null(params$tissue)) toupper(params$tissue) else "F"
stopifnot(TISSUE %in% c("F", "G"))
FDR_ALPHA <- 0.10
NPERM     <- 10000   # permutations per family. At 1,000 draws a p near 0.05 has a Monte Carlo
                     # SE of ~0.007; at 10,000 it is ~0.002. The matrix engine in block D makes
                     # 10,000 x 7 metrics x (58 + 6 + ~1,000) genes a matter of a minute.
PERM_SEED <- 1
N_INFLUENCE_HITS <- 3
## The detection-floor filter (FLOOR_EXCLUDE / EXCLUDE_FLOOR_GENES) lives in script 20, which
## writes detection_floor_flags_<T>.csv and marks each candidate `tested`; this script tests
## the same family script 20 tested.

## The metric list, the metric tiers (primary / exploratory), the treatment levels and the
## nuisance covariates are read from script 20's metrics_config.csv handoff below, so the
## two scripts cannot drift apart. The permutation runs for EVERY metric (no PMETRICS).

## ---- read script 20's CSV handoffs (run 20 first) ----------------------------------
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
IN_DIR  <- file.path(find_repo_root(), "gene-mechanics-correlation", "03_analyses", "gene_mechanics")
OUT_DIR <- IN_DIR                                  # 21 writes alongside 20's outputs
need <- function(f) { p <- file.path(IN_DIR, f)
  if (!file.exists(p)) stop("Missing ", f, " - run script 20 first (it writes this)."); p }

mcfg <- read_csv(need(sprintf("metrics_config_%s.csv", TISSUE)), show_col_types = FALSE)
if (!"tier" %in% names(mcfg)) {
  message("metrics_config has no `tier` column (script 20 predates PRIMARY_METRICS); ",
          "assuming force/area and their change metrics are primary. Re-run script 20.")
  mcfg$tier <- ifelse(mcfg$metric %in% c("max_force", "pad_area", "dlog_max_force", "dlog_pad_area"),
                      "primary", "exploratory")
}
if (!"covariates" %in% names(mcfg)) mcfg$covariates <- "treatment"
metrics         <- mcfg$metric
LEVEL_METRICS   <- mcfg$metric[mcfg$type == "level"]
CHANGE_METRICS  <- mcfg$metric[mcfg$type == "change"]
PRIMARY_METRICS <- mcfg$metric[mcfg$tier == "primary"]
arms_kept       <- strsplit(mcfg$arms[1], ";", fixed = TRUE)[[1]]
COVARIATES      <- strsplit(mcfg$covariates[1], ";", fixed = TRUE)[[1]]   # "treatment" [+ "secretion_state"]
USE_SECRETION_STATE <- "secretion_state" %in% COVARIATES
tier_of <- function(m) mcfg$tier[match(m, mcfg$metric)]

paired <- read_csv(need(sprintf("paired_sample_manifest_%s.csv", TISSUE)), show_col_types = FALSE) %>%
  mutate(treatment = factor(treatment, levels = arms_kept))
if (USE_SECRETION_STATE) {
  ## script 20 joined the per-animal byssal secretion state (differential-expression script
  ## 01_7) into the manifest; it is a nuisance covariate in every fit below and a stratum of
  ## the permutation, exactly as `treatment` is.
  if (!"secretion_state" %in% names(paired) || anyNA(paired$secretion_state))
    stop("USE_SECRETION_STATE is on in script 20 but the manifest has no complete `secretion_state` column.")
  paired <- paired %>% mutate(secretion_state = factor(secretion_state, levels = c("off", "on")))
}
candidates_all <- read_csv(need(sprintf("candidate_genes_%s.csv", TISSUE)), show_col_types = FALSE)
if (!"tested" %in% names(candidates_all)) candidates_all$tested <- TRUE   # older script-20 handoff
candidates <- candidates_all %>% filter(tested)
floor_tbl  <- read_csv(need(sprintf("detection_floor_flags_%s.csv", TISSUE)), show_col_types = FALSE) %>%
  select(LOC_ID, frac_at_floor, floor_flag)
annot      <- read_csv(need("annotation_map.csv"), show_col_types = FALSE)
thr_raw    <- read_csv(need(sprintf("thread_plaques_paired_%s.csv", TISSUE)), show_col_types = FALSE)
vp_df <- read_csv(need(sprintf("vst_paired_%s.csv", TISSUE)), show_col_types = FALSE)
vp <- as.matrix(vp_df[, -1]); rownames(vp) <- vp_df[[1]]   # LOC_ID column -> matrix rownames
stopifnot(all(paired$sample %in% colnames(vp)))
vp <- vp[, paired$sample, drop = FALSE]                      # column order = manifest order

## DEG union (data-driven universe) from script 20's per-gene table
deg_union_file <- file.path(IN_DIR, sprintf("assoc_DEGunion_%s.csv", TISSUE))
deg_union_ids <- if (file.exists(deg_union_file))
  intersect(unique(read_csv(deg_union_file, show_col_types = FALSE)$LOC_ID), rownames(vp)) else character(0)

## nuisance design shared by every fit: y ~ x + treatment [+ secretion_state]
nuis_terms <- COVARIATES
cat("Paired animals:", nrow(paired), "| arms:", paste(arms_kept, collapse = ", "),
    "| level metrics:", paste(LEVEL_METRICS, collapse = ", "),
    "| change metrics:", paste(CHANGE_METRICS, collapse = ", "),
    "\nPrimary metrics:", paste(PRIMARY_METRICS, collapse = ", "),
    "| nuisance covariates:", paste(nuis_terms, collapse = " + "),
    "| candidates tested:", nrow(candidates), "of", nrow(candidates_all),
    "| DEG union tested:", length(deg_union_ids), "\n")
```

---

# A. Thread-level mixed model (sensitivity check on the level metrics)

Raw plaque-level data, joined to each mussel's expression. Model:
`metric ~ expression + treatment + (1 | mussel)`. The expression term varies only between
mussels, so its test reflects ~31 mussel-level df (Satterthwaite), not the thread count;
mussels with more/precise plaques contribute more. Singular-fit warnings are expected when
between-mussel variance is near zero and do not invalidate the fixed-effect estimate.

The thread-level model applies to the LEVEL metrics only: a CHANGE metric (`dlog_*`) is one
number per animal by construction. It is kept as a check that the reported per-animal
weighted regression (block B) is not an artefact of averaging; it is not the reported test,
because it cannot be permuted at a useful cost and would then have no search-corrected p.

```{r mixed}
## raw day-3 plaque-level mechanics for the paired animals (script 20 exports phase == "post")
thr_threads <- thr_raw %>%
  filter(phase == "post", mussel %in% paired$mussel, treatment %in% arms_kept) %>%
  transmute(mussel,
            treatment = factor(treatment, levels = arms_kept),
            max_force, pad_area, adhesion_kpa, max_displacement)
if (USE_SECRETION_STATE)
  thr_threads <- thr_threads %>%
    left_join(paired %>% select(mussel, secretion_state), by = "mussel")

test_gene_mixed <- function(loc, metric) {
  ev  <- data.frame(mussel = paired$mussel,
                    expr = as.numeric(vp[match(loc, rownames(vp)), paired$sample]))
  dat <- thr_threads %>% inner_join(ev, by = "mussel")
  dat <- dat[stats::complete.cases(dat[, c(metric, "expr", nuis_terms, "mussel")]), ]
  if (dplyr::n_distinct(dat$mussel) < 6) return(NULL)
  fm <- tryCatch(suppressWarnings(
          lmerTest::lmer(reformulate(c("expr", nuis_terms, "(1|mussel)"),
                                     response = metric),
                         data = dat, REML = TRUE)),
        error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"expr" %in% rownames(co)) return(NULL)
  tibble(LOC_ID = loc, metric = metric,
         n_threads = nrow(dat), n_mussel = dplyr::n_distinct(dat$mussel),
         slope_mixed = co["expr", "Estimate"], p_mixed = co["expr", "Pr(>|t|)"])
}

run_mixed <- function(loci, set_name) {
  map_dfr(LEVEL_METRICS, function(mt) map_dfr(loci, ~ test_gene_mixed(.x, mt))) %>%
    group_by(metric) %>% mutate(q_mixed = p.adjust(p_mixed, "BH")) %>% ungroup() %>%
    left_join(annot, by = "LOC_ID") %>%
    mutate(gene_set = set_name, tier = tier_of(metric)) %>%
    arrange(metric, p_mixed)
}

## candidate set: already loaded from script 20's candidate_genes_<T>.csv handoff above

res_cand_mixed <- run_mixed(candidates$LOC_ID, "candidate")
write_csv(res_cand_mixed, file.path(OUT_DIR, sprintf("assoc_candidate_MIXED_%s.csv", TISSUE)))

cat("Candidate genes, MIXED model, top per LEVEL metric (sensitivity; compare p_mixed with p_wls below):\n")
res_cand_mixed %>% mutate(metric = factor(metric, levels = LEVEL_METRICS)) %>%
  group_by(metric) %>% slice_min(p_mixed, n = 6) %>% ungroup() %>%
  mutate(across(c(slope_mixed), ~round(., 4)),
         p_mixed = signif(p_mixed, 2), q_mixed = signif(q_mixed, 2)) %>%
  select(metric, tier, LOC_ID, protein_name, n_mussel, slope_mixed, p_mixed, q_mixed) %>%
  as.data.frame() %>% print()
```

---

# B. Precision-weighted regression (the reported test)

Per-mussel means weighted by plaque count, `metric ~ expression + treatment` (plus the
secretion state when script 20 enabled it). Nearly equivalent to the mixed model for the
level metrics, and the only sensible model for the change metrics. It is cheap to refit
thousands of times, which is why it is the statistic the permutation in block D calibrates,
and therefore the statistic reported for every metric.

For a CHANGE metric the weight is the smaller of the animal's baseline and day-3 plaque
counts, since the log ratio is only as precise as its less-replicated side.

```{r weighted}
wls_weight <- function(metric) {
  if (metric %in% CHANGE_METRICS) pmin(paired$n_plaques, paired$n_plaques_base) else paired$n_plaques
}
wls_formula <- reformulate(c("x", nuis_terms), response = "y")
test_gene_wls <- function(loc, metric) {
  x <- as.numeric(vp[match(loc, rownames(vp)), paired$sample])
  dat <- data.frame(y = paired[[metric]], x = x, w = wls_weight(metric),
                    paired[, nuis_terms, drop = FALSE])
  dat <- dat[stats::complete.cases(dat), ]
  if (nrow(dat) < 6) return(NULL)
  fm <- tryCatch(lm(wls_formula, data = dat, weights = w), error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"x" %in% rownames(co)) return(NULL)
  tibble(LOC_ID = loc, metric = metric, n = nrow(dat),
         slope_wls = co["x", "Estimate"], p_wls = co["x", "Pr(>|t|)"])
}
run_wls <- function(loci, set_name) {
  map_dfr(metrics, function(mt) map_dfr(loci, ~ test_gene_wls(.x, mt))) %>%
    group_by(metric) %>% mutate(q_wls = p.adjust(p_wls, "BH")) %>% ungroup() %>%
    left_join(annot, by = "LOC_ID") %>%
    mutate(gene_set = set_name, tier = tier_of(metric))
}
res_cand_wls <- run_wls(candidates$LOC_ID, "candidate") %>%
  ## the mixed-model p for the level metrics sits beside the reported p as a sensitivity column
  left_join(res_cand_mixed %>% select(LOC_ID, metric, slope_mixed, p_mixed, q_mixed),
            by = c("LOC_ID", "metric"))
## (the permutation columns are appended to this table in block D before it is written)

cat("Candidate genes, weighted per-animal regression (the reported test), top per metric:\n")
res_cand_wls %>% mutate(metric = factor(metric, levels = metrics)) %>%
  group_by(metric) %>% slice_min(p_wls, n = 4) %>% ungroup() %>%
  mutate(slope_wls = round(slope_wls, 4), p_wls = signif(p_wls, 2), q_wls = signif(q_wls, 2),
         p_mixed = signif(p_mixed, 2)) %>%
  select(metric, tier, n, LOC_ID, protein_name, slope_wls, p_wls, q_wls, p_mixed) %>%
  as.data.frame() %>% print()
```

---

# C. Module eigengenes (power-preserving)

Each module = PC1 of its members' VST values (per mussel), oriented so higher score = higher
module expression. Six modules x seven metrics is FDR-tractable, and a coordinated pathway
score is biologically closer to "is the HIF response engaged" than any single gene.

`byssal_structural` is the plaque/thread protein module (foot proteins, preCols, byssal
EP/ACDC): the genes that switch on together when an animal is actively secreting a thread.
It is kept separate from the broader `byssal_collagen` regex (which also catches every
collagen and every "byss*" annotation) so that the direct test of "these genes track thread
building, not strength" is its own row.

```{r modules}
modules <- list(
  HIF_hypoxia       = "hypoxia-inducible|Egl nine|prolyl 4-hydroxylase|prolyl hydroxylase|hypoxia up-regulated",
  oxidative_detox   = "superoxide|glutathione|catalase|peroxiredoxin|dual oxidase|peroxidase",
  HSP_proteostasis  = "heat shock|Hsp|stress-70|chaperone|BiP|peptidyl-prolyl",
  tRNA_translation  = "--tRNA|tRNA ligase|aminoacyl",
  byssal_collagen   = "collagen|byss|foot protein|adhesive plaque|polyphenolic",
  byssal_structural = "foot protein|adhesive plaque|polyphenolic|byssal protein-|preCol|precollagen|byssal EP|ACDC"
)

if (!"blast_ok" %in% names(annot)) annot$blast_ok <- TRUE   # older script-20 handoff
module_score <- function(rx) {
  locs <- annot %>% filter(blast_ok, grepl(rx, protein_name, ignore.case = TRUE)) %>% pull(LOC_ID)
  locs <- intersect(locs, rownames(vp))
  if (length(locs) < 3) return(NULL)
  z <- t(scale(t(vp[locs, paired$sample, drop = FALSE])))   # standardise each gene
  z <- z[stats::complete.cases(z), , drop = FALSE]
  pc <- prcomp(t(z), center = TRUE, scale. = FALSE)
  sc <- pc$x[, 1]
  if (suppressWarnings(cor(sc, colMeans(vp[locs, paired$sample]))) < 0) sc <- -sc
  list(score = sc, n_genes = length(locs), members = locs,
       var_expl = summary(pc)$importance[2, 1])
}
mod_scores <- purrr::compact(map(modules, module_score))
MOD <- do.call(rbind, map(mod_scores, "score"))          # modules x animals, manifest order
colnames(MOD) <- paired$sample
write_csv(map_dfr(names(mod_scores), ~ tibble(module = .x, LOC_ID = mod_scores[[.x]]$members)) %>%
            left_join(annot, by = "LOC_ID"),
          file.path(OUT_DIR, sprintf("module_members_%s.csv", TISSUE)))

## the same weighted regression as block B (reported, permuted), plus the thread-level mixed
## model for the level metrics as the sensitivity column
mod_fits <- imap_dfr(mod_scores, function(ms, nm) {
  dat0 <- tibble(mussel = paired$mussel, score = ms$score)
  map_dfr(metrics, function(mt) {
    dat <- data.frame(y = paired[[mt]], x = ms$score, w = wls_weight(mt),
                      paired[, nuis_terms, drop = FALSE])
    dat <- dat[stats::complete.cases(dat), ]
    if (nrow(dat) < 6) return(NULL)
    fm <- tryCatch(lm(wls_formula, data = dat, weights = w), error = function(e) NULL)
    if (is.null(fm)) return(NULL)
    co <- summary(fm)$coefficients
    out <- tibble(module = nm, n_genes = ms$n_genes, pc1_var = round(ms$var_expl, 2),
                  metric = mt, tier = tier_of(mt), n = nrow(dat),
                  slope_wls = co["x", "Estimate"], p_wls = co["x", "Pr(>|t|)"],
                  slope_mixed = NA_real_, p_mixed = NA_real_)
    if (mt %in% LEVEL_METRICS) {
      datm <- thr_threads %>% inner_join(dat0, by = "mussel")
      fmm <- tryCatch(suppressWarnings(
               lmerTest::lmer(reformulate(c("score", nuis_terms, "(1|mussel)"),
                                          response = mt), data = datm, REML = TRUE)),
             error = function(e) NULL)
      if (!is.null(fmm)) {
        com <- summary(fmm)$coefficients
        out$slope_mixed <- com["score", "Estimate"]; out$p_mixed <- com["score", "Pr(>|t|)"]
      }
    }
    out
  })
}) %>% group_by(metric) %>%
  mutate(q_wls = p.adjust(p_wls, "BH"), q_mixed = p.adjust(p_mixed, "BH")) %>%
  ungroup() %>% arrange(p_wls)
## (permutation columns are appended and the file written in block D)

cat("MODULE-level associations (weighted regression, BH across the ", n_distinct(mod_fits$module),
    " modules within each metric):\n", sep = "")
mod_fits %>% mutate(across(c(slope_wls, slope_mixed), ~round(., 4)),
                    across(c(p_wls, q_wls, p_mixed, q_mixed), ~signif(., 2))) %>%
  select(module, n_genes, pc1_var, metric, tier, n, slope_wls, p_wls, q_wls, p_mixed) %>%
  as.data.frame() %>% print()
```

---

# D. Permutation null: per metric, family-wide, and for the primary family

**Why a permutation at all.** The search is 58 candidate genes x 7 metrics (plus six modules
and ~500-1,000 DEG-union genes), the genes are correlated with one another and the metrics
with one another, and n is 45. BH within a metric corrects for the genes but not for the
metrics, and treats correlated genes as independent tests. The permutation asks the question
directly: shuffle the expression-to-animal assignment **within treatment arm** (so arm-level
expression shifts and the gene-gene correlation structure are preserved), recompute the
whole search, and see how often chance produces a minimum p as small as the observed one.

Three numbers come out of the same shuffles:

- `p_perm_metric`: the chance of a minimum p this small among the genes of ONE metric
  (what the old `permutation_best_hit` file reported, now for every metric);
- `p_perm_family`: the chance across the whole search, genes x all seven metrics; this is the
  number that answers "did the search find anything";
- `p_perm_family_primary`: the same for the pre-declared primary family (candidate genes x
  `PRIMARY_METRICS` from script 20), the confirmatory test.

Each observed gene x metric p also gets a single-step min-p adjusted p (`p_perm_famwise`,
Westfall & Young 1993): the fraction of shuffles whose family-wide minimum p is at or below
that gene's observed p. It is monotone in p and reads like a Bonferroni-Holm value that knows
the correlation structure.

**Engine.** For a weighted regression `y ~ x + Z` (Z = the nuisance design: intercept, arm
dummies, and the secretion state when enabled), the t statistic of x equals the t statistic
of the weighted, Z-residualised x regressed on the weighted, Z-residualised y
(Frisch-Waugh-Lovell): `t = r sqrt(df) / sqrt(1 - r^2)`, with r the uncentred correlation of
the two residual vectors and `df = n - ncol(Z) - 1`. The residualiser M depends only on Z
and the weights, so a permutation is one column permutation of the expression matrix and
one matrix multiply, for every gene at once. That is what makes 10,000 shuffles across every
gene set and metric take about a minute instead of a day of `lm()` calls. The engine is
checked against `lm()` on the observed data before it is used.

```{r permutation}
## ---- the engine ---------------------------------------------------------------------
make_engine <- function(y, w, Z) {
  ok <- !is.na(y) & !is.na(w) & stats::complete.cases(Z)
  sw <- sqrt(w[ok]); Zt <- Z[ok, , drop = FALSE] * sw
  ZtZ_inv <- tryCatch(chol2inv(chol(crossprod(Zt))),
                      error = function(e) stop("nuisance design is singular within the animals ",
                                               "that have this metric: ", conditionMessage(e)))
  M  <- diag(sum(ok)) - Zt %*% ZtZ_inv %*% t(Zt)
  yr <- as.vector(M %*% (y[ok] * sw))
  list(ok = ok, sw = sw,
       SM = sw * M,                       # diag(sqrt(w)) %*% M: weight and residualise in one multiply
       yr = yr, yr_norm = sqrt(sum(yr^2)),
       df = sum(ok) - ncol(Zt) - 1, n = sum(ok),
       key = paste(c(ok, signif(sw, 8)), collapse = ","))   # metrics sharing ok & weights share SM
}
## weighted, residualised expression for every row of E (genes x animals, manifest order),
## optionally with the animals' expression columns permuted first
resid_X <- function(eng, E, perm = seq_len(ncol(E)))
  E[, perm, drop = FALSE][, eng$ok, drop = FALSE] %*% eng$SM
## two-sided p for the x coefficient from the residualised rows (rn = their norms)
p_from_Xr <- function(eng, Xr, rn = sqrt(rowSums(Xr^2))) {
  r <- as.vector(Xr %*% eng$yr) / (rn * eng$yr_norm)
  r[!is.finite(r)] <- 0
  tt <- r * sqrt(eng$df) / sqrt(pmax(1 - r^2, .Machine$double.eps))
  2 * pt(abs(tt), eng$df, lower.tail = FALSE)
}
engine_p <- function(eng, E, perm = seq_len(ncol(E))) p_from_Xr(eng, resid_X(eng, E, perm))

Z_nuis  <- model.matrix(reformulate(nuis_terms), data = droplevels(paired[, nuis_terms, drop = FALSE]))
engines <- setNames(lapply(metrics, function(m) make_engine(paired[[m]], wls_weight(m), Z_nuis)), metrics)
## metrics with the same animals and weights (all level metrics; all change metrics) share
## one residualisation per shuffle
engine_groups <- split(metrics, sapply(metrics, function(m) engines[[m]]$key))

## ---- check the engine against lm() on the observed data -------------------------------
E_cand <- vp[candidates$LOC_ID, , drop = FALSE]
chk <- map_dfr(metrics, function(m) {
  tibble(metric = m, LOC_ID = rownames(E_cand), p_engine = engine_p(engines[[m]], E_cand))
}) %>% inner_join(res_cand_wls %>% select(metric, LOC_ID, p_wls), by = c("metric", "LOC_ID"))
stopifnot(nrow(chk) > 0, max(abs(chk$p_engine - chk$p_wls)) < 1e-8)
cat("Engine check: max |p_engine - p_lm| over", nrow(chk), "candidate fits =",
    signif(max(abs(chk$p_engine - chk$p_wls)), 2), "\n")

## ---- the shuffles: within arm (and within secretion state when it is a covariate) -----
strata <- interaction(droplevels(paired[, nuis_terms, drop = FALSE]), drop = TRUE)
idx_by_stratum <- split(seq_len(nrow(paired)), strata)
perm_within <- function() {
  p <- seq_len(nrow(paired))
  for (g in idx_by_stratum) if (length(g) > 1) p[g] <- g[sample.int(length(g))]
  p
}

## ---- families: candidate genes, modules, DEG union ----------------------------------
families <- list(candidate = E_cand, module = MOD)
if (length(deg_union_ids) > 0) families$DEG_union <- vp[deg_union_ids, , drop = FALSE]

## observed and null minimum p per family x metric (one shared set of shuffles, so the
## family-wide minimum is taken inside each shuffle and metric-metric correlation is kept)
set.seed(PERM_SEED)
perms <- replicate(NPERM, perm_within())               # animals x NPERM
t0 <- Sys.time()
perm_min <- imap(families, function(E, fam) {
  obs  <- sapply(metrics, function(m) min(engine_p(engines[[m]], E), na.rm = TRUE))
  null <- matrix(NA_real_, NPERM, length(metrics), dimnames = list(NULL, metrics))
  for (b in seq_len(NPERM)) {
    perm <- perms[, b]
    for (g in engine_groups) {
      Xr <- resid_X(engines[[g[1]]], E, perm); rn <- sqrt(rowSums(Xr^2))
      for (m in g) null[b, m] <- min(p_from_Xr(engines[[m]], Xr, rn), na.rm = TRUE)
    }
  }
  list(obs = obs, null = null)
})
cat("Permutations:", NPERM, "shuffles x", length(metrics), "metrics x",
    paste(sprintf("%s (%d genes)", names(families), sapply(families, nrow)), collapse = ", "),
    "in", round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1), "min\n")

## ---- summaries ------------------------------------------------------------------------
perm_p <- function(null_vec, obs) (sum(null_vec <= obs) + 1) / (length(null_vec) + 1)

perm_summary <- imap_dfr(perm_min, function(pm, fam) {
  fam_all  <- apply(pm$null, 1, min)                                       # family-wide null
  fam_prim <- apply(pm$null[, intersect(PRIMARY_METRICS, metrics), drop = FALSE], 1, min)
  obs_all  <- min(pm$obs); obs_prim <- min(pm$obs[intersect(PRIMARY_METRICS, metrics)])
  bind_rows(
    tibble(gene_set = fam, family = "per_metric", metric = metrics, tier = tier_of(metrics),
           n = sapply(metrics, function(m) engines[[m]]$n),
           n_genes = nrow(families[[fam]]), n_tests = nrow(families[[fam]]), n_perm = NPERM,
           observed_min_p = unname(pm$obs),
           p_permutation  = sapply(metrics, function(m) perm_p(pm$null[, m], pm$obs[[m]]))),
    tibble(gene_set = fam, family = "all_metrics", metric = "all", tier = "all",
           n = NA_integer_, n_genes = nrow(families[[fam]]),
           n_tests = nrow(families[[fam]]) * length(metrics), n_perm = NPERM,
           observed_min_p = obs_all, p_permutation = perm_p(fam_all, obs_all)),
    tibble(gene_set = fam, family = "primary_metrics", metric = paste(PRIMARY_METRICS, collapse = ";"),
           tier = "primary", n = NA_integer_, n_genes = nrow(families[[fam]]),
           n_tests = nrow(families[[fam]]) * length(intersect(PRIMARY_METRICS, metrics)), n_perm = NPERM,
           observed_min_p = obs_prim, p_permutation = perm_p(fam_prim, obs_prim)))
})
write_csv(perm_summary, file.path(OUT_DIR, sprintf("permutation_summary_%s.csv", TISSUE)))
## backward-compatible file: the per-metric candidate rows in the old column layout
write_csv(perm_summary %>% filter(gene_set == "candidate", family == "per_metric") %>%
            transmute(metric, n, n_genes, n_perm, observed_min_p, p_permutation),
          file.path(OUT_DIR, sprintf("permutation_best_hit_%s.csv", TISSUE)))

## null vectors, so the tails can be inspected and the p re-derived without re-running
null_df <- imap_dfc(perm_min, function(pm, fam) {
  d <- as.data.frame(signif(pm$null, 4)); names(d) <- paste0(fam, "__", metrics)
  d[[paste0(fam, "__all_metrics")]]     <- signif(apply(pm$null, 1, min), 4)
  d[[paste0(fam, "__primary_metrics")]] <- signif(apply(pm$null[, intersect(PRIMARY_METRICS, metrics), drop = FALSE], 1, min), 4)
  d
}) %>% mutate(perm = seq_len(NPERM), .before = 1)
write_csv(null_df, file.path(OUT_DIR, sprintf("permutation_null_%s.csv.gz", TISSUE)))

cat("\nPermutation p per family (observed minimum p and how often the shuffled search beats it):\n")
perm_summary %>% filter(family != "per_metric" | tier == "primary") %>%
  mutate(observed_min_p = signif(observed_min_p, 3), p_permutation = signif(p_permutation, 3)) %>%
  as.data.frame() %>% print()

## ---- single-step adjusted p for every observed gene x metric (Westfall-Young min-p) ------
## p_perm_famwise: fraction of shuffles whose whole-search minimum p <= this observed p
## p_perm_metricwise: the same within the gene's own metric (genes only, not metrics)
adj_p <- function(obs_p, null_sorted)            # null_sorted: ascending null minimum p
  (findInterval(obs_p, null_sorted) + 1) / (length(null_sorted) + 1)
add_perm_cols <- function(tab, fam) {
  pm <- perm_min[[fam]]
  null_metric <- apply(pm$null, 2, sort)                                             # per metric
  null_all    <- sort(apply(pm$null, 1, min))                                        # whole search
  null_prim   <- sort(apply(pm$null[, intersect(PRIMARY_METRICS, metrics), drop = FALSE], 1, min))
  tab %>% mutate(
    p_perm_metricwise = mapply(function(p, m) adj_p(p, null_metric[, m]), p_wls, metric),
    p_perm_famwise    = adj_p(p_wls, null_all),
    p_perm_primary    = ifelse(metric %in% PRIMARY_METRICS, adj_p(p_wls, null_prim), NA_real_))
}
res_cand_wls <- add_perm_cols(res_cand_wls, "candidate") %>% arrange(metric, p_wls)
write_csv(res_cand_wls, file.path(OUT_DIR, sprintf("assoc_candidate_WLS_%s.csv", TISSUE)))
mod_fits <- add_perm_cols(mod_fits, "module") %>% arrange(p_wls)
write_csv(mod_fits, file.path(OUT_DIR, sprintf("module_associations_%s.csv", TISSUE)))
if ("DEG_union" %in% names(families)) {
  res_deg_wls <- run_wls(deg_union_ids, "DEG_union") %>% add_perm_cols("DEG_union") %>% arrange(metric, p_wls)
  write_csv(res_deg_wls, file.path(OUT_DIR, sprintf("assoc_DEGunion_WLS_%s.csv", TISSUE)))
}

cat("\nInterpretation: p_permutation is the chance of a minimum p this small when expression is",
    "shuffled among animals within arm. If it is >> 0.05, the best hit is consistent with noise",
    "given the search. p_perm_famwise on a gene row is that gene's search-corrected p.\n")
```

---

# E. Diagnostics: detection floor and influence; the best-hits table

Genes where many paired samples sit at the VST minimum are effectively absent/zero-inflated;
their correlations are driven by detection, not biology. Script 20 flags them and, by
default, keeps them out of the tested family; the flags are echoed here. Also report the
most influential animal per top hit (Cook's distance) so you know if one mussel drives a fit.

The tables are written to `03_analyses/gene_mechanics/` so the report and any downstream
filter can use them without re-running this script:

- `detection_floor_flags_<T>.csv` (written by script 20): one row per gene in the candidate
  set and the DEG union, with the fraction of paired samples at the VST floor, a
  `floor_flag` of `exclude` (> 0.4), `caution` (> 0.2) or `ok`, and `tested` (FALSE for the
  `exclude` genes when script 20's `EXCLUDE_FLOOR_GENES` is TRUE: they are out of the family
  before BH and before the permutation, and appear in no association table).
- `influence_top_hits_<T>.csv`: the three best candidate hits per metric (ranked by the
  reported test, `p_wls`, for every metric) with the most influential animal, its Cook's
  distance, the slope and p after dropping that animal, an `influence_flag` (`fragile` when
  the hit loses p < 0.05 or its slope moves by more than half), and the q and permutation p
  of the hit.
- `best_hits_<T>.csv`: the single best hit per metric for each gene set (candidate, module,
  DEG union), with `q_wls`, `p_perm_metric`, `p_perm_famwise`, `p_perm_primary`, the floor
  flag and the influence flag in one row. Nothing should be quoted from a q alone.

```{r diagnostics}
## detection floor: computed and applied by script 20 (detection_floor_flags_<T>.csv); shown
## here so the report carries it beside the influence table
floor_flag <- read_csv(need(sprintf("detection_floor_flags_%s.csv", TISSUE)), show_col_types = FALSE)
cat("Detection-floor summary (", nrow(floor_flag), " genes in the two universes; `tested` = in the family):\n", sep = "")
print(table(gene_set = floor_flag$gene_set, floor_flag = floor_flag$floor_flag, tested = floor_flag$tested))

## influence: best hits per metric, which mussel has the largest Cook's distance.
## Ranking is by the reported test (p_wls) for every metric, so the leave-one-out refit, the
## q and the permutation p on a row all describe the same fit.
top_hits <- res_cand_wls %>%
  group_by(metric) %>% slice_min(p_wls, n = N_INFLUENCE_HITS, with_ties = FALSE) %>% ungroup() %>%
  transmute(metric, LOC_ID, ranking_model = "wls", p_rank = p_wls, q_wls,
            p_perm_metricwise, p_perm_famwise, p_perm_primary)

infl <- pmap_dfr(top_hits %>% select(metric, LOC_ID, p_rank), function(metric, LOC_ID, p_rank) {
  dat <- data.frame(y = paired[[metric]],
                    x = as.numeric(vp[match(LOC_ID, rownames(vp)), paired$sample]),
                    w = wls_weight(metric), mussel = paired$mussel,
                    paired[, nuis_terms, drop = FALSE])
  dat <- dat[stats::complete.cases(dat), ]
  fm  <- lm(wls_formula, data = dat, weights = w)
  cd  <- cooks.distance(fm)
  ## slope after dropping the most influential animal: how much of the fit is one mussel
  fm_drop <- lm(wls_formula, data = dat[-which.max(cd), ], weights = w)
  b_all  <- unname(coef(fm)["x"]); b_drop <- unname(coef(fm_drop)["x"])
  p_drop <- summary(fm_drop)$coefficients["x", "Pr(>|t|)"]
  tibble(metric, LOC_ID, p_rank = signif(p_rank, 3),
         n                       = nrow(dat),
         most_influential_mussel = dat$mussel[which.max(cd)],
         max_cooks_D             = round(max(cd), 3),
         cook_threshold_4n       = round(4 / nrow(dat), 3),
         ## the max of n Cook's distances exceeds 4/n in almost every fit of this size, so
         ## that screen alone is uninformative here; D > 1 is the conventional "one point
         ## moves the fit" level (Cook & Weisberg 1982)
         cooks_flag              = case_when(max(cd) > 1            ~ "D>1",
                                             max(cd) > 4 / nrow(dat) ~ "D>4/n",
                                             TRUE                     ~ "ok"),
         slope                   = signif(b_all, 4),
         slope_without_mussel    = signif(b_drop, 4),
         p_without_mussel        = signif(p_drop, 3),
         slope_change_frac       = round(abs(b_drop - b_all) / abs(b_all), 3),
         ## "fragile": the result depends on one animal (loses p < 0.05, or the slope moves
         ## by more than half) - report such a hit with the leave-one-out slope beside it
         influence_flag          = ifelse((p_rank < 0.05 & p_drop >= 0.05) |
                                            abs(b_drop - b_all) / abs(b_all) > 0.5,
                                          "fragile", "robust"))
}) %>%
  left_join(top_hits %>% select(metric, LOC_ID, ranking_model, q_wls, p_perm_metricwise,
                                p_perm_famwise, p_perm_primary), by = c("metric", "LOC_ID")) %>%
  left_join(annot, by = "LOC_ID") %>%
  mutate(tier = tier_of(metric), metric = factor(metric, levels = metrics)) %>%
  arrange(metric, p_rank) %>% mutate(metric = as.character(metric)) %>%
  select(metric, tier, LOC_ID, protein_name, ranking_model, n, p_rank, q_wls,
         p_perm_metricwise, p_perm_famwise, p_perm_primary, everything())
write_csv(infl, file.path(OUT_DIR, sprintf("influence_top_hits_%s.csv", TISSUE)))

cat("\nMost influential animal for the top", N_INFLUENCE_HITS, "hits per metric",
    "('fragile' = loses p < 0.05 or slope moves > 50% when that animal is dropped):\n")
infl %>% select(metric, LOC_ID, p_rank, q_wls, p_perm_famwise, most_influential_mussel, max_cooks_D,
                cooks_flag, slope, slope_without_mussel, p_without_mussel, influence_flag) %>%
  as.data.frame() %>% print()

## ---- best hit per metric and gene set, everything on one row --------------------------
best_cand <- res_cand_wls %>% group_by(metric) %>% slice_min(p_wls, n = 1, with_ties = FALSE) %>%
  ungroup() %>% transmute(gene_set = "candidate", metric, tier, feature = LOC_ID, protein_name,
                          n, slope_wls, p_wls, q_wls, p_mixed, p_perm_metricwise, p_perm_famwise, p_perm_primary)
best_mod <- mod_fits %>% group_by(metric) %>% slice_min(p_wls, n = 1, with_ties = FALSE) %>%
  ungroup() %>% transmute(gene_set = "module", metric, tier, feature = module,
                          protein_name = paste0(n_genes, "-gene module"),
                          n, slope_wls, p_wls, q_wls, p_mixed, p_perm_metricwise, p_perm_famwise, p_perm_primary)
best_deg <- if (exists("res_deg_wls")) res_deg_wls %>% group_by(metric) %>%
  slice_min(p_wls, n = 1, with_ties = FALSE) %>% ungroup() %>%
  transmute(gene_set = "DEG_union", metric, tier, feature = LOC_ID, protein_name,
            n, slope_wls, p_wls, q_wls, p_mixed = NA_real_, p_perm_metricwise, p_perm_famwise, p_perm_primary) else NULL
best_hits <- bind_rows(best_cand, best_mod, best_deg) %>%
  left_join(floor_tbl %>% select(feature = LOC_ID, floor_flag), by = "feature") %>%
  left_join(infl %>% select(feature = LOC_ID, metric, most_influential_mussel, influence_flag),
            by = c("feature", "metric")) %>%
  left_join(perm_summary %>% filter(family == "per_metric") %>%
              select(gene_set, metric, p_perm_metric = p_permutation), by = c("gene_set", "metric")) %>%
  left_join(perm_summary %>% filter(family == "all_metrics") %>%
              select(gene_set, p_perm_family = p_permutation), by = "gene_set") %>%
  left_join(perm_summary %>% filter(family == "primary_metrics") %>%
              select(gene_set, p_perm_family_primary = p_permutation), by = "gene_set") %>%
  mutate(metric = factor(metric, levels = metrics),
         gene_set = factor(gene_set, levels = c("candidate", "module", "DEG_union"))) %>%
  arrange(gene_set, metric) %>% mutate(metric = as.character(metric), gene_set = as.character(gene_set)) %>%
  select(gene_set, metric, tier, feature, protein_name, n, slope_wls, p_wls, q_wls, p_mixed,
         p_perm_metric, p_perm_metricwise, p_perm_famwise, p_perm_primary,
         p_perm_family, p_perm_family_primary, floor_flag, most_influential_mussel, influence_flag)
write_csv(best_hits, file.path(OUT_DIR, sprintf("best_hits_%s.csv", TISSUE)))

cat("\nBest hit per metric and gene set (q beside its permutation companions):\n")
best_hits %>% mutate(across(c(p_wls, q_wls, p_mixed, p_perm_metric, p_perm_famwise, p_perm_primary,
                              p_perm_family, p_perm_family_primary), ~signif(., 2)),
                     protein_name = substr(protein_name, 1, 30)) %>%
  select(gene_set, metric, tier, feature, protein_name, p_wls, q_wls, p_perm_metric,
         p_perm_famwise, p_perm_primary, p_perm_family, floor_flag, influence_flag) %>%
  as.data.frame() %>% print()
```

---

# F. Baseline-adjusted level models

Adding each animal's baseline as a covariate asks whether expression relates to the day-3
level *given where the animal started*. The direct test of whether expression tracks the
*change* is the `dlog_*` metrics in blocks B and C, which need no covariate. Both are kept
because they answer slightly different questions (ANCOVA versus change score) and can
disagree when baseline and change are correlated.

```{r baseline-adjusted}
base_lookup <- paired %>% select(mussel, starts_with("base_"))
test_gene_mixed_adj <- function(loc, metric) {
  bcol <- paste0("base_", metric)
  ev <- data.frame(mussel = paired$mussel,
                   expr = as.numeric(vp[match(loc, rownames(vp)), paired$sample]))
  dat <- thr_threads %>% inner_join(ev, by = "mussel") %>%
    left_join(base_lookup, by = "mussel")
  dat$base_val <- dat[[bcol]]                       # was: rename(base = !!bcol)
  keep_cols <- c(metric, "expr", nuis_terms, "mussel", "base_val")
  dat <- dat[stats::complete.cases(dat[, keep_cols]), ]
  if (dplyr::n_distinct(dat$mussel) < 6) return(NULL)
  fm <- tryCatch(suppressWarnings(
          lmerTest::lmer(reformulate(c("expr", nuis_terms, "base_val", "(1|mussel)"),
                                     response = metric), data = dat, REML = TRUE)),
        error = function(e) NULL)
  if (is.null(fm)) return(NULL)
  co <- summary(fm)$coefficients
  if (!"expr" %in% rownames(co)) return(NULL)
  tibble(LOC_ID = loc, metric = metric, n_mussel = dplyr::n_distinct(dat$mussel),
         slope = co["expr", "Estimate"], p = co["expr", "Pr(>|t|)"])
}

res_adj <- map_dfr(LEVEL_METRICS, function(mt) map_dfr(candidates$LOC_ID, ~ test_gene_mixed_adj(.x, mt))) %>%
  group_by(metric) %>% mutate(q = p.adjust(p, "BH")) %>% ungroup() %>%
  left_join(annot, by = "LOC_ID") %>% mutate(tier = tier_of(metric)) %>% arrange(metric, p)
write_csv(res_adj, file.path(OUT_DIR, sprintf("assoc_candidate_BASELINEADJ_%s.csv", TISSUE)))
cat("Baseline-adjusted candidate top hits (expression vs CHANGE in attachment):\n")
res_adj %>% group_by(metric) %>% slice_min(p, n = 5) %>% ungroup() %>%
  mutate(slope = round(slope, 4), p = signif(p, 2), q = signif(q, 2)) %>%
  select(metric, tier, LOC_ID, protein_name, n_mussel, slope, p, q) %>%
  as.data.frame() %>% print()
```

---

# G. Provenance

Script 20 opens `RUN_provenance_<T>.txt` with its settings and the git commit of the thread
input; this block appends the permutation settings, so the folder records the configuration
behind every file in it.

```{r provenance}
prov <- file.path(OUT_DIR, sprintf("RUN_provenance_%s.txt", TISSUE))
cat(c("",
      "[21-gene_mechanics_expanded.Rmd]",
      paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
      paste0("NPERM                : ", NPERM),
      paste0("PERM_SEED            : ", PERM_SEED),
      paste0("permuted metrics     : ", paste(metrics, collapse = ", "), " (all)"),
      paste0("primary metrics      : ", paste(PRIMARY_METRICS, collapse = ", ")),
      paste0("nuisance covariates  : ", paste(nuis_terms, collapse = " + ")),
      paste0("permutation strata   : ", paste(names(idx_by_stratum), collapse = ", ")),
      paste0("families             : ", paste(sprintf("%s (%d)", names(families), sapply(families, nrow)), collapse = ", ")),
      paste0("modules              : ", paste(names(mod_scores), collapse = ", ")),
      paste0("FDR_ALPHA            : ", FDR_ALPHA),
      paste0("candidates tested    : ", nrow(candidates), " of ", nrow(candidates_all), " (floor filter in script 20)"),
      paste0("N_INFLUENCE_HITS     : ", N_INFLUENCE_HITS),
      paste0("R / lme4 / lmerTest  : ", R.version.string, " / ",
             as.character(packageVersion("lme4")), " / ", as.character(packageVersion("lmerTest")))),
    file = prov, sep = "\n", append = file.exists(prov))
cat(readLines(prov), sep = "\n")
```
