---
title: "22 - RNA x thread manifest + top-25 up/down expression tables"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

1. **Manifest:** per treatment, which RNA-sequenced mussels also have thread data, with a
   reconciliation against the raw tensometer files so you can see exactly where the summary
   is missing entries.
2. **Expression tables:** for each tissue x stressor contrast, the top 25 up- and top 25
   down-regulated DEGs (by log2 fold change), with their expression across **every
   sequenced mussel that has thread data**, plus a companion table of each mussel's thread
   mechanics so you can cross-reference expression against attachment in one place.

This script reads script 20's `annotation_map.csv` handoff, so run 20 first. All paths resolve automatically.

```{r config}
suppressPackageStartupMessages({ library(tidyverse); library(readxl); library(DESeq2); library(here) })

## ---- analysis options --------------------------------------------------------------
TISSUE      <- "F"                  # "F" foot (default) or "G" gill
TREATMENTS  <- c("OA", "OW", "DO")  # contrasts to tabulate
N_TOP       <- 25                   # genes per direction
SIG_ONLY    <- TRUE                 # rank within significant DEGs (padj<0.05); FALSE = all genes
USE_RAW_THREAD_SET <- FALSE         # FALSE = mussels in the clean summary; TRUE = any raw (T) tensometer file

## ---- portable paths ----------------------------------------------------------------
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
repo_root   <- find_repo_root()
DE_DIR      <- file.path(repo_root, "differential-expression")
THREAD_FILE <- file.path(repo_root, "thread-strength", "02_data", "thread-summary-trossulus-clean.xlsx")
RAW_DIR     <- file.path(repo_root, "thread-strength", "02_data", "tensometer_output")
GM_DIR      <- file.path(repo_root, "gene-mechanics-correlation", "03_analyses", "gene_mechanics")  # script 20 outputs
OUT_DIR     <- file.path(repo_root, "gene-mechanics-correlation", "03_analyses", "expr_tables")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

## canonical 3-digit trossulus mussel tag, shared across the chain
norm_id <- function(x) {
  x   <- as.character(x)
  pre <- gsub("[0-9].*$", "", x)
  num <- suppressWarnings(as.integer(gsub("[^0-9].*$", "", gsub("^[^0-9]+", "", x))))
  ifelse(is.na(num), NA_character_, sprintf("%s%03d", pre, num))
}
```

---

# 1. RNA x thread manifest and raw-file reconciliation

```{r manifest}
## RNA samples present (chosen-tissue + the other, to know what was sequenced)
cm_head <- read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
                    nrows = 1, check.names = FALSE)
rna_cols <- colnames(cm_head)[-1]
rna_tab <- tibble(col = rna_cols) %>%
  mutate(mussel = str_extract(col, "^T\\d+"),
         tissue = str_extract(col, "(FX|F|G)$")) %>%
  filter(!is.na(mussel)) %>%
  group_by(mussel) %>% summarise(rna_tissues = paste(sort(unique(tissue)), collapse = ""),
                                 .groups = "drop")

## RNA treatment/day (foot treatmentinfo is authoritative; same animal => same treatment)
ti <- read.table(file.path(DE_DIR, "03_analyses", "DEG_lists", "Foot", "F_treatmentinfo.csv"),
                 header = TRUE, sep = " ") %>%
  mutate(mussel = sub("F$", "", `...1`)) %>% distinct(mussel, treatment, day)

## thread summary presence
thr_raw <- read_excel(THREAD_FILE,
                      sheet = "data") %>% mutate(mn = norm_id(mussel))
post_n <- thr_raw %>% filter(treatment %in% c("OA","OW","DO")) %>% count(mn, name = "in_summary_post")
base_n <- thr_raw %>% filter(treatment == "control") %>% count(mn, name = "in_summary_base")

## raw tensometer file counts (mussel -> n files), normalised IDs
count_raw <- function(sub) {
  fp <- file.path(RAW_DIR, sub)
  ids <- str_extract(list.files(fp), "^T\\d+")
  ids <- ids[!is.na(ids)]
  tibble(mn = vapply(ids, norm_id, character(1))) %>% count(mn, name = paste0("raw_", sub))
}
raw_treat <- count_raw("treatment"); raw_ctrl <- count_raw("control")

manifest <- rna_tab %>%
  mutate(mn = norm_id(mussel)) %>%
  left_join(ti, by = "mussel") %>%
  left_join(post_n, by = "mn") %>% left_join(base_n, by = "mn") %>%
  left_join(raw_treat, by = "mn") %>% left_join(raw_ctrl, by = "mn") %>%
  mutate(across(c(in_summary_post, in_summary_base, raw_treatment, raw_control),
                ~replace_na(., 0)),
         poststress_missing = raw_treatment > 0 & in_summary_post == 0,
         baseline_missing   = raw_control  > 0 & in_summary_base == 0) %>%
  arrange(treatment, mn)

write_csv(manifest, file.path(OUT_DIR, "rna_thread_manifest.csv"))

cat("RNA-sequenced mussels with post-stress thread data, by treatment:\n")
manifest %>% filter(in_summary_post > 0) %>%
  group_by(treatment) %>% summarise(mussels = paste(mussel, collapse = ", "), .groups="drop") %>%
  as.data.frame() %>% print()

cat("\nRNA-sequenced, MISSING post-stress threads (check raw files):\n")
manifest %>% filter(!is.na(treatment), treatment != "control", in_summary_post == 0) %>%
  select(mussel, treatment, rna_tissues, raw_treatment) %>% as.data.frame() %>% print()

cat("\nBaseline thread data in RAW but MISSING from summary (summarisation gap):\n")
manifest %>% filter(baseline_missing) %>%
  select(mussel, treatment, rna_tissues, raw_control) %>% as.data.frame() %>% print()
```

---

# 2. Define the "mussels with thread data" sample set

```{r threadset}
if (USE_RAW_THREAD_SET) {
  thread_mussels <- manifest %>% filter(raw_treatment > 0 | raw_control > 0) %>% pull(mussel)
} else {
  thread_mussels <- unique(thr_raw$mussel)            # present in summary 'data' sheet
}
samples_wanted <- paste0(thread_mussels, TISSUE)

## per-mussel thread mechanics (post-stress means) for cross-reference
mech <- thr_raw %>%
  filter(treatment %in% c("OA","OW","DO")) %>%
  group_by(mussel, treatment) %>%
  summarise(n_plaques = n(),
            max_force = mean(max_force, na.rm = TRUE),
            adhesion_kpa = mean(adhesion_kpa, na.rm = TRUE),
            pad_area = mean(pad_area, na.rm = TRUE), .groups = "drop")
```

---

# 3. VST expression matrix (chosen tissue)

```{r vst}
cm <- read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
               row.names = 1, check.names = FALSE)
tissue_cols <- if (TISSUE == "F") grep("F$", colnames(cm), value = TRUE) else
                                  grep("G$", colnames(cm), value = TRUE)   # F$ excludes FX
cm_t <- cm[, tissue_cols, drop = FALSE]
dds <- DESeqDataSetFromMatrix(round(as.matrix(cm_t)),
                              DataFrame(row.names = colnames(cm_t)), design = ~ 1)
dds <- dds[rowSums(counts(dds)) > 0, ]
vmat <- assay(vst(dds, blind = TRUE))
rownames(vmat) <- sub("^gene-", "", sub(".*\\|", "", sub("^gene-", "", rownames(vmat))))

samples_have <- intersect(samples_wanted, colnames(vmat))
cat("Thread-having mussels:", length(thread_mussels),
    "| with", TISSUE, "expression:", length(samples_have), "\n")
```

---

# 4. Annotation map

```{r annot}
## annotation map: read script 20's handoff (one source of truth across the chain)
am_path <- file.path(GM_DIR, "annotation_map.csv")
if (!file.exists(am_path))
  stop("annotation_map.csv not found in ", GM_DIR, " - run script 20 first.")
annot <- read_csv(am_path, show_col_types = FALSE)
```

---

# 5. Build top-25 up / down expression tables per contrast

For each contrast, rank DEGs by log2 fold change (within significant genes if `SIG_ONLY`),
take the top/bottom `N_TOP`, and emit a gene x sample VST table with LFC/padj and the
direction label. Column order is sorted by the contrast's own treatment first so the
stressed animals cluster, but every thread-having mussel is included.

```{r tables}
find_sig <- function(tcode) {                       # locate the siggene file for this contrast
  d <- file.path(DE_DIR, "03_analyses", "DEG_lists", ifelse(TISSUE == "F", "Foot", "Gill"))
  f <- list.files(d, pattern = sprintf("^%s%s_TC_siggene.*\\.csv$", TISSUE, tcode),
                  full.names = TRUE)
  if (!length(f)) return(NULL)
  read.table(f[1], header = TRUE, sep = " ")
}

## sample column order: chosen contrast animals first, then others, by treatment
mussel_trt <- manifest %>% select(mussel, treatment) %>% deframe()
order_samples <- function(tcode) {
  s_mussel <- sub(paste0(TISSUE, "$"), "", samples_have)
  ord <- tibble(sample = samples_have, mussel = s_mussel,
                trt = mussel_trt[s_mussel]) %>%
    mutate(trt = replace_na(trt, "control"),
           pri = ifelse(trt == tcode, 0, 1)) %>%
    arrange(pri, trt, mussel)
  ord$sample
}

build_table <- function(tcode) {
  sg <- find_sig(tcode); if (is.null(sg)) { message("no siggene for ", tcode); return(NULL) }
  gcol <- grep("gene", names(sg), ignore.case = TRUE, value = TRUE)[1]
  sg <- sg %>%
    mutate(LOC_ID = sub("^gene-", "", sub(".*\\|", "", sub("^gene-", "", .data[[gcol]]))))
  if (SIG_ONLY && "padj" %in% names(sg)) sg <- filter(sg, padj < 0.05)
  sg <- sg %>% filter(LOC_ID %in% rownames(vmat)) %>%
    arrange(desc(log2FoldChange))
  top <- bind_rows(
    head(sg, N_TOP) %>% mutate(direction = "up"),
    tail(sg, N_TOP) %>% mutate(direction = "down"))
  cols <- order_samples(tcode)
  expr <- vmat[top$LOC_ID, cols, drop = FALSE]
  out <- top %>%
    left_join(annot, by = "LOC_ID") %>%
    transmute(direction, LOC_ID, protein_name,
              log2FoldChange = round(log2FoldChange, 3),
              padj = signif(padj, 3)) %>%
    bind_cols(as.data.frame(round(expr, 3)))
  write_csv(out, file.path(OUT_DIR, sprintf("top%d_updown_%s_%s.csv", N_TOP, TISSUE, tcode)))
  out
}

tables <- set_names(TREATMENTS) %>% map(build_table)
cat("Wrote", sum(!map_lgl(tables, is.null)), "expression tables to", OUT_DIR, "\n")

## companion: per-sample metadata + thread mechanics (one row per thread-having mussel)
sample_meta <- tibble(sample = samples_have,
                      mussel = sub(paste0(TISSUE, "$"), "", samples_have)) %>%
  left_join(manifest %>% select(mussel, rna_treatment = treatment, rna_day = day), by = "mussel") %>%
  left_join(mech, by = "mussel") %>%
  arrange(rna_treatment, mussel)
write_csv(sample_meta, file.path(OUT_DIR, sprintf("sample_metadata_%s.csv", TISSUE)))

cat("\nExample - top up/down for", TREATMENTS[1], "(first few columns):\n")
if (!is.null(tables[[1]]))
  tables[[1]] %>% select(direction, protein_name, log2FoldChange, padj) %>%
    group_by(direction) %>% slice_head(n = 5) %>% ungroup() %>% as.data.frame() %>% print()
```

---

# 6. Notes

```{r notes, echo=FALSE, results='asis'}
cat("
**Outputs (in ", OUT_DIR, "):**

- `rna_thread_manifest.csv` - every sequenced mussel: RNA tissues, treatment/day, summary
  presence (post-stress & baseline), raw-file counts, and `baseline_missing` /
  `poststress_missing` flags. Filter `baseline_missing == TRUE` to get your audit list.
- `top25_updown_<tissue>_<treatment>.csv` - 25 up + 25 down DEGs (by LFC) with LFC, padj,
  direction, and VST expression across all thread-having mussels.
- `sample_metadata_<tissue>.csv` - per mussel: treatment, day, and post-stress thread
  mechanics, to cross-reference against the expression columns above (match on mussel ID;
  expression column `T046F` -> mussel `T046`).

**Caveats:**

- `SIG_ONLY = TRUE` ranks within padj<0.05 DEGs; for contrasts with few DEGs (e.g. foot-OA,
  ~80 genes) you may get fewer than 25 in a direction. Set FALSE to rank all expressed genes.
- With `USE_RAW_THREAD_SET = FALSE`, columns are limited to mussels currently in the summary.
  Recover the missing baseline records (manifest `baseline_missing`) and re-run with TRUE to
  pull the day-0 control animals (T003-T012 etc.) into the tables.
- Genes are ranked by fold change, not by association with mechanics; this is a descriptive
  cross-reference table, not a test.
")
```
