---
title: "23 - Byssus/foot gene list + expression across thread-tested mussels"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Build a gene list **narrowed to byssus/foot-relevant proteins and processes** (structural
byssal/foot proteins, byssal collagens, cross-linking and curing enzymes, collagen
biosynthesis, and named ECM/adhesion proteins), and tabulate their expression across every
sequenced mussel that has thread data, with each gene's stress-responsiveness annotated.

This refines the repo's existing approach (script 11), which greps only "foot"/"byss" and
so misses collagens, plaque proteins, tyrosinases, and cross-linkers. It also **excludes**
generic antioxidant peroxidases / peroxiredoxins and monoamine/polyamine oxidases, which a
naive "peroxidase|amine oxidase" filter wrongly pulls in.

Reads script 22's `rna_thread_manifest.csv` for the mussel set, so run 22 first. All paths resolve automatically.

```{r config}
suppressPackageStartupMessages({ library(tidyverse); library(readxl); library(DESeq2); library(here) })

## ---- analysis options --------------------------------------------------------------
TISSUE      <- "F"                 # "F" foot (default) or "G" gill
HI_CONF_ONLY <- FALSE             # TRUE = keep only confident blast annotations (pident>=40 & evalue<1e-10)
KEEP_CATEGORIES <- c("byssal_structural","byssal_accessory","collagen",
                     "crosslink_curing","collagen_biosynth","ecm_adhesion")
USE_RAW_THREAD_SET <- FALSE        # FALSE = mussels in the clean summary; TRUE = any raw (T) thread file
MIN_COUNT <- 5                     # gene kept if count > MIN_COUNT in >= 1/3 of thread-having samples

## ---- portable paths ----------------------------------------------------------------
find_repo_root <- function(start = here::here()) {
  d <- normalizePath(start, winslash = "/")
  for (i in 1:6) {
    if (all(dir.exists(file.path(d, c("differential-expression", "thread-strength"))))) return(d)
    d <- dirname(d)
  }
  stop("Repo root not found: need a parent containing differential-expression/ and thread-strength/.")
}
repo_root    <- find_repo_root()
DE_DIR       <- file.path(repo_root, "differential-expression")
THREAD_FILE  <- file.path(repo_root, "thread-strength", "02_data", "thread-summary-trossulus-clean.xlsx")
EXPR_DIR     <- file.path(repo_root, "gene-mechanics-correlation", "03_analyses", "expr_tables")  # script 22 outputs
LOC_GO_FILE  <- file.path(repo_root, "blast", "03_analyses", "genome-foot", "LOC_GO_list.txt")
OUT_DIR      <- file.path(repo_root, "gene-mechanics-correlation", "03_analyses", "byssus_genes")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

# 1. Build the byssus/foot gene list (categorised)

```{r genelist}
ann <- read_delim(LOC_GO_FILE, delim = "\t",
                  show_col_types = FALSE) %>%
  mutate(LOC_ID = str_extract(as.character(gene), "LOC\\d+")) %>%
  filter(!is.na(LOC_ID)) %>%
  mutate(across(c(pident, evalue, bitscore), as.numeric)) %>%
  arrange(desc(bitscore)) %>% distinct(LOC_ID, .keep_all = TRUE)   # best hit per LOC

PN <- replace_na(ann$`Protein.names`, "")
g  <- function(rx, x = PN) grepl(rx, x, ignore.case = TRUE, perl = TRUE)

category <- dplyr::case_when(
  g("foot protein|adhesive plaque|polyphenolic|byssal protein-?\\d|\\bMGFP|\\bMCFP|Mefp|byssal EP|\\bACDC\\b|mussel byssus|mussel adhesive|preCol") ~ "byssal_structural",
  g("byssal peroxidase|peroxidasin|\\btyrosinase|catechol oxidase|lysyl oxidase|transglutaminase|gamma-glutamyltransferase|DOPA decarboxylase|aromatic-L-amino-acid decarboxylase|tyrosine decarboxylase") &
    !g("glutathione|thioredoxin|peroxiredoxin|monoamine|diamine|polyamine|spermine|histone demethylase|renalase|chloride intracellular|lysine-specific") ~ "crosslink_curing",
  g("prolyl 4-hydroxylase|prolyl 3-hydroxylase|lysyl hydroxylase|procollagen-lysine|procollagen-proline|protein disulfide-isomerase") ~ "collagen_biosynth",
  g("\\bcollagen\\b|procollagen") ~ "collagen",
  g("byssal|byssus") ~ "byssal_accessory",
  g("\\bADAMTS|thrombospondin|fibrillin|\\blaminin|von Willebrand factor|fibronectin|matrilin|tenascin|nidogen|perlecan|basement membrane-specific") ~ "ecm_adhesion",
  TRUE ~ NA_character_)

byss <- ann %>% mutate(category = category) %>% filter(!is.na(category)) %>%
  mutate(hi_conf = pident >= 40 & evalue < 1e-10)

cat("Byssus/foot genes:", nrow(byss), "| high-confidence:", sum(byss$hi_conf, na.rm = TRUE), "\n")
byss %>% count(category) %>% as.data.frame() %>% print()
```

---

# 2. Stress-responsiveness (foot/gill DEG contrasts)

```{r de}
deg_dir <- file.path(DE_DIR, "03_analyses", "DEG_lists", ifelse(TISSUE == "F", "Foot", "Gill"))
add_lfc <- function(tcode) {
  f <- list.files(deg_dir, pattern = sprintf("^%s%s_TC_siggene.*\\.csv$", TISSUE, tcode),
                  full.names = TRUE)
  if (!length(f)) return(tibble(LOC_ID = character(), lfc = numeric()))
  d <- read.table(f[1], header = TRUE, sep = " ")
  gcol <- grep("gene", names(d), ignore.case = TRUE, value = TRUE)[1]
  tibble(LOC_ID = sub("^gene-", "", sub(".*\\|", "", sub("^gene-", "", d[[gcol]]))),
         lfc = d$log2FoldChange)
}
for (tc in c("OA","OW","DO")) {
  lf <- add_lfc(tc) %>% distinct(LOC_ID, .keep_all = TRUE)
  byss[[paste0(TISSUE, "_", tc, "_log2FC")]] <- round(lf$lfc[match(byss$LOC_ID, lf$LOC_ID)], 2)
}
byss$n_DE <- rowSums(!is.na(byss[, grep("_log2FC$", names(byss))]))
```

---

# 3. Thread-having mussels and VST expression

```{r expr}
## thread-having mussel set: from script 22's reconciliation handoff (run 22 first)
man_path <- file.path(EXPR_DIR, "rna_thread_manifest.csv")
if (!file.exists(man_path))
  stop("rna_thread_manifest.csv not found in ", EXPR_DIR, " - run script 22 first.")
manifest <- read_csv(man_path, show_col_types = FALSE)
thread_mussels <- if (USE_RAW_THREAD_SET) {
  manifest %>% filter(raw_treatment > 0 | raw_control > 0) %>% pull(mussel)
} else {
  manifest %>% filter(in_summary_post > 0 | in_summary_base > 0) %>% pull(mussel)
}
thread_mussels <- unique(thread_mussels)

## per-mussel post-stress thread mechanics for the companion table (canonical clean summary)
thr_raw <- read_excel(THREAD_FILE, sheet = "data")
samples_wanted <- paste0(thread_mussels, TISSUE)

cm <- read.csv(file.path(DE_DIR, "02_data", "gene_count_matrix_clean.csv"),
               row.names = 1, check.names = FALSE)
tissue_cols <- if (TISSUE=="F") grep("F$", colnames(cm), value=TRUE) else grep("G$", colnames(cm), value=TRUE)
dds <- DESeqDataSetFromMatrix(round(as.matrix(cm[, tissue_cols])),
                              DataFrame(row.names = tissue_cols), design = ~ 1)
dds <- dds[rowSums(counts(dds)) > 0, ]
vmat <- assay(vst(dds, blind = TRUE))
rownames(vmat) <- sub("^gene-", "", sub(".*\\|", "", sub("^gene-", "", rownames(vmat))))

samples_have <- intersect(samples_wanted, colnames(vmat))
counts_named <- counts(dds); rownames(counts_named) <- rownames(vmat)
cat("Thread-having mussels:", length(thread_mussels), "| with", TISSUE, "expression:",
    length(samples_have), "\n")
```

---

# 4. Assemble the narrowed expression table

```{r assemble}
genes <- byss
if (HI_CONF_ONLY)       genes <- filter(genes, hi_conf)
genes <- filter(genes, category %in% KEEP_CATEGORIES, LOC_ID %in% rownames(vmat))

## keep genes actually expressed in the thread-having samples
cmat <- counts_named[genes$LOC_ID, samples_have, drop = FALSE]
expressed <- rowSums(cmat > MIN_COUNT) >= ceiling(length(samples_have)/3)
genes <- genes[expressed, ]
cat("Byssus genes expressed in", TISSUE, "across thread-having mussels:", nrow(genes), "\n")

expr_block <- as.data.frame(round(vmat[genes$LOC_ID, samples_have, drop = FALSE], 3))
table_out <- genes %>%
  transmute(category, LOC_ID, protein_name = `Protein.names`, gene_name = `Gene.Names`,
            pident = round(pident,1), evalue, hi_conf,
            !!!set_names(map(grep("_log2FC$", names(genes), value = TRUE),
                             ~ genes[[.x]]), grep("_log2FC$", names(genes), value = TRUE)),
            n_DE) %>%
  bind_cols(expr_block) %>%
  arrange(factor(category, levels = KEEP_CATEGORIES), desc(n_DE), protein_name)

write_csv(table_out, file.path(OUT_DIR, sprintf("byssus_gene_expression_%s.csv", TISSUE)))

## companion: per-mussel treatment + thread mechanics
mech <- thr_raw %>% filter(treatment %in% c("OA","OW","DO")) %>%
  group_by(mussel) %>% summarise(treatment = first(treatment), n_plaques = n(),
            max_force = round(mean(max_force, na.rm=TRUE),3),
            adhesion_kpa = round(mean(adhesion_kpa, na.rm=TRUE),1),
            pad_area = round(mean(pad_area, na.rm=TRUE),2), .groups="drop")
sample_meta <- tibble(sample = samples_have, mussel = sub(paste0(TISSUE,"$"),"",samples_have)) %>%
  left_join(mech, by = "mussel")
write_csv(sample_meta, file.path(OUT_DIR, sprintf("sample_metadata_%s.csv", TISSUE)))

cat("\nGene counts in final table by category:\n")
table_out %>% count(category) %>% as.data.frame() %>% print()
cat("\nByssus genes that respond to stress (n_DE>0):\n")
table_out %>% filter(n_DE > 0) %>%
  select(category, protein_name, ends_with("_log2FC")) %>% as.data.frame() %>% print()
```

---

# 5. Optional: per-category module score per mussel

A compact summary, mean VST per category per mussel, useful for plotting category-level
expression against thread mechanics (feeds naturally into the script-21 mixed model with
`category mean` as the predictor).

```{r catscore}
cat_score <- map_dfr(unique(genes$category), function(cc) {
  locs <- genes$LOC_ID[genes$category == cc]
  if (length(locs) < 2) return(NULL)
  tibble(category = cc, sample = samples_have,
         mean_vst = colMeans(vmat[locs, samples_have, drop = FALSE]))
}) %>%
  pivot_wider(names_from = category, values_from = mean_vst) %>%
  mutate(mussel = sub(paste0(TISSUE,"$"), "", sample)) %>%
  left_join(mech, by = "mussel")
write_csv(cat_score, file.path(OUT_DIR, sprintf("byssus_category_scores_%s.csv", TISSUE)))
cat("Wrote per-category mean-expression scores (", ncol(cat_score), "cols ).\n")
```

---

# 6. Notes

```{r notes, echo=FALSE, results='asis'}
cat("
**Outputs (", OUT_DIR, "):**

- `byssus_gene_expression_<tissue>.csv` - the narrowed list: category, protein, blast
  quality (pident/evalue/hi_conf), per-contrast log2FC, n_DE, then VST expression across
  every thread-having mussel (columns like `T046F` = mussel `T046`).
- `sample_metadata_<tissue>.csv` - per mussel: treatment and post-stress thread mechanics,
  to align against the expression columns.
- `byssus_category_scores_<tissue>.csv` - mean VST per category per mussel + mechanics.

**Interpretation / caveats:**

- Categories were built from UniProt names + a tight enzyme filter; `hi_conf` flags
  confident blast hits (pident>=40, evalue<1e-10). Set `HI_CONF_ONLY=TRUE` to drop weak
  annotations (e.g. distant paralog hits among the many byssal-protein gene copies).
- Byssal proteins are a highly duplicated gene family, so one protein name maps to many
  LOCs; that is real, not a bug, but treat the paralog set with care.
- In foot, the canonical *structural* byssal proteins are not differentially expressed; the
  stress signal sits in the curing/biosynthesis machinery (prolyl-4-hydroxylase, byssal
  HSP-like, transglutaminase). The table lets you confirm this per gene per animal.
- Columns are limited to mussels currently in the summary. Once the missing baseline
  records are recovered (script 22 audit), set `USE_RAW_THREAD_SET=TRUE` to include the
  day-0 control animals.
")
```
