---
title: "24 - Run the gene-mechanics chain for foot and gill"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output: html_document
params:
  tissues: ["F", "G"]     # which tissues to run; one or both
  scripts: [20, 21, 22, 23]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Scripts 20 to 23 are parameterised by tissue (`params$tissue`, default `"F"`). This driver
renders them in order for each tissue requested, so the gill analysis is the same code as the
foot analysis with one parameter changed, and the two cannot drift apart.

Every output of 20 to 23 is tissue-suffixed (`_F` / `_G`), except `annotation_map.csv`,
which is a tissue-independent LOC-to-protein map. Rendered HTML reports land in
`03_analyses/knit_html/`.

Knit this file, or run the chunk below, from inside `gene-mechanics-correlation.Rproj`.
Each render starts a fresh environment, exactly as knitting the script by hand would.

```{r run}
suppressPackageStartupMessages(library(here))

code_dir <- here::here("01_code")
html_dir <- here::here("03_analyses", "knit_html")
dir.create(html_dir, showWarnings = FALSE, recursive = TRUE)

script_files <- c(
  `20` = "20-gene_mechanics_correlation.Rmd",
  `21` = "21-gene_mechanics_expanded.Rmd",
  `22` = "22-rna_thread_manifest_and_expression_tables.Rmd",
  `23` = "23-byssus_foot_gene_list_expression.Rmd"
)
tissues <- toupper(params$tissues)
scripts <- script_files[as.character(params$scripts)]
stopifnot(all(tissues %in% c("F", "G")), all(!is.na(scripts)))

## Each script is rendered in its OWN R process. Calling rmarkdown::render() from inside a
## knit does not work (the child's `setup` chunk collides with this document's in knitr's
## label registry), and a fresh process is anyway the faithful equivalent of knitting each
## script by hand: nothing leaks between scripts or between tissues.
rscript <- file.path(R.home("bin"), if (.Platform$OS.type == "windows") "Rscript.exe" else "Rscript")

render_in_new_process <- function(rmd, tissue, out_html, log_file) {
  job <- tempfile(fileext = ".R")
  writeLines(sprintf(
    'rmarkdown::render(%s, params = list(tissue = %s), output_file = %s, output_dir = %s, quiet = TRUE)',
    shQuote(normalizePath(rmd, winslash = "/")), shQuote(tissue),
    shQuote(out_html), shQuote(normalizePath(html_dir, winslash = "/"))), job)
  status <- system2(rscript, shQuote(job), stdout = log_file, stderr = log_file)
  unlink(job)
  status == 0
}

log <- list()
for (tissue in tissues) {
  for (s in scripts) {
    stem <- tools::file_path_sans_ext(s)
    out  <- sprintf("%s_%s.html", stem, tissue)
    logf <- file.path(html_dir, sprintf("%s_%s.log", stem, tissue))
    message(sprintf("==== %s  tissue = %s ====", s, tissue))
    t0 <- Sys.time()
    ok <- render_in_new_process(file.path(code_dir, s), tissue, out, logf)
    if (!ok) message("FAILED; see ", logf)
    log[[length(log) + 1]] <- data.frame(
      tissue = tissue, script = s, ok = ok,
      minutes = round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1),
      html = if (ok) file.path("03_analyses", "knit_html", out) else NA_character_)
    if (!ok) { message("Stopping the ", tissue, " chain; later scripts depend on this one."); break }
  }
}
run_log <- do.call(rbind, log)
print(run_log)
write.csv(run_log, file.path(html_dir, "run_log.csv"), row.names = FALSE)
```

# What differs between the tissues

Nothing in the code. What differs is the data:

- The gill count matrix has 59 samples against 58 for foot, and T047 has gill but no foot
  RNA, so the gill paired set is 46 animals (45 with baselines) against 45 (44) for foot.
  T051G is listed in `Gill/G_treatmentinfo.csv` but was removed from the count matrix for QC
  (`01_5-gene_count_matrix.Rmd`); script 20 drops it with a message.
- The byssal secretion state (`differential-expression/02_data/secretion_state.csv`, from
  DE script `01_7`) is measured on the foot sample and joined to both tissues by animal;
  run the differential-expression batch driver first when that file is missing or stale.
- `Gill/G_treatmentinfo.csv` codes the day-3 control arm as `control_3` and has no leading
  index column; script 20 and 22 normalise both.
- The DEG union is built from the tissue's own `*_TC_siggene*` files, which
  `differential-expression/01_code/03_5-DEG_table_provenance_check.Rmd` verifies are
  reproduced exactly from the committed inputs for all six contrasts.
- The foot is the byssus-producing organ. A gill association is a systemic-state readout,
  not a mechanism at the site of thread production, and should be framed that way.
