---
title: "Assemble thread summary"
author: "PSMFC mytilus byssus pilot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

## What this script does

Joins the trace-level output of `1_extract_tensometer_data.Rmd` to the plaque measurements in
`02_data/pad_area_measurements.xlsx`, computes adhesion, and writes the thread summary that
scripts 3 and 4, and the `09_gene-mechanics-correlation` scripts, read.

```
1_extract_tensometer_data.Rmd  ->  03_analyses/01_extract-tensometer-data/thread-summary-raw-output.xlsx
              +  02_data/pad_area_measurements.xlsx                            (hand-measured)
2_assemble_thread_summary.Rmd  ->  03_analyses/02_assemble-thread-summary/thread-summary.xlsx
3_analyze_thread_strength.Rmd  ->  03_analyses/03_analyze-thread-strength/
4_decompose_adhesion.Rmd       ->  03_analyses/04_decompose-adhesion/
```

### Where each fact comes from

One fact, one source. Nothing is copied into two files where the copies could drift.

| fact | grain | source |
|---|---|---|
| `max_force`, `integral`, `max_displacement` | thread | the trace itself, via script 1 |
| `thread_trt`, `phase`, `day` | thread | the tensometer subfolder, via script 1 |
| `species`, `mussel_trt`, `rna_sequenced` | mussel | `02_data/mussel-treatment-key.csv`, via script 1 |
| `pad_area`, `failure` | thread | `02_data/pad_area_measurements.xlsx` |

`pad_area_measurements.xlsx` therefore carries only what it is the source of truth for:
`mussel`, `thread`, `thread_trt`, `pad_area`, `failure`, `notes`. Its previous `species`,
`group`, `mussel_trt` and `day` columns were second copies of facts owned elsewhere and have
been removed. They were verified to agree before removal.

**Nothing is excluded here.** Every extracted trace is carried through, so bad runs can be
reviewed against their QC plots.

## Load packages

```{r}
## Run within "thread-strength.Rproj" so here::here() resolves to the analysis root.
load.lib    <- c("readxl", "tidyverse", "openxlsx", "here")
install.lib <- load.lib[!load.lib %in% installed.packages()]
for (lib in install.lib) install.packages(lib, dependencies = TRUE)
sapply(load.lib, require, character = TRUE)
```

## Paths

```{r paths}
raw_xlsx <- here::here("03_analyses", "01_extract-tensometer-data", "thread-summary-raw-output.xlsx")
pad_xlsx <- here::here("02_data", "pad_area_measurements.xlsx")
out_dir  <- here::here("03_analyses", "02_assemble-thread-summary")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(raw_xlsx))
  stop("Not found:\n  ", raw_xlsx, "\nRun 1_extract_tensometer_data.Rmd first.")
if (!file.exists(pad_xlsx))
  stop("Not found:\n  ", pad_xlsx)
```

## Helper: canonical mussel tag

```{r}
# Collapse any spelling of a tag to prefix + zero-padded 3 digits (T37 -> T037) so the same
# physical animal joins across sources regardless of how it was written.
norm_id <- function(x) {
  x   <- trimws(as.character(x))
  pre <- sub("^([A-Za-z]+).*$", "\\1", x)
  num <- suppressWarnings(as.integer(sub("^[A-Za-z]+0*([0-9]+).*$", "\\1", x)))
  ifelse(is.na(num) | pre == x, NA_character_, sprintf("%s%03d", pre, num))
}
```

## 1. Read the extraction output

```{r}
traces <- read_excel(raw_xlsx, sheet = "data") %>%
  mutate(mussel = norm_id(mussel),
         thread = as.integer(thread),
         note   = replace_na(as.character(note), ""))

folder_order <- unique(traces$source_folder)

message("Traces: ", nrow(traces), " | unique mussels: ", n_distinct(traces$mussel))

if (!all(c("species", "mussel_trt", "rna_sequenced") %in% names(traces)))
  warning("The extraction output has no mussel-level columns. Run 0_build_mussel_key.Rmd ",
          "and re-run script 1, or those columns will be blank in the thread summary.")

print(knitr::kable(traces %>% count(source_folder, thread_trt, phase, day)))
```

## 2. Join the plaque measurements

`pad_area` (plaque cross-sectional area, mm^2) and `failure` are measured by hand from
microscope images of each plaque; they cannot be derived from a force trace.

The join key is `mussel` + `thread` + `thread_trt`. It has to include `thread_trt`: for an
animal pulled before and after exposure, `mussel` + `thread` alone matches two different
traces and would silently duplicate rows.

```{r}
pad <- read_excel(pad_xlsx, sheet = "data")

# Accept the old column spellings so an older copy of the file still joins.
ren <- c(mussel_ID = "mussel", thread_num = "thread")
for (from in names(ren)) {
  to <- ren[[from]]
  if (from %in% names(pad) && !to %in% names(pad))
    names(pad)[names(pad) == from] <- to
}

need_cols <- c("mussel", "thread", "thread_trt", "pad_area")
if (length(setdiff(need_cols, names(pad))) > 0)
  stop(basename(pad_xlsx), " is missing column(s): ",
       paste(setdiff(need_cols, names(pad)), collapse = ", "),
       "\nColumns present: ", paste(names(pad), collapse = ", "))

# Optional columns: create them empty if the file does not carry them, so the transmute
# below never has to branch on the schema.
if (!"failure" %in% names(pad)) pad$failure <- NA_character_
if (!"notes"   %in% names(pad)) pad$notes   <- NA_character_

pad <- pad %>%
  transmute(mussel     = norm_id(mussel),
            thread     = as.integer(thread),
            thread_trt = as.character(thread_trt),
            pad_area   = suppressWarnings(as.numeric(pad_area)),
            failure    = as.character(failure),
            pad_notes  = as.character(notes))

dup <- pad %>% count(mussel, thread, thread_trt) %>% filter(n > 1)
if (nrow(dup) > 0) {
  warning(nrow(dup), " duplicated key(s) in ", basename(pad_xlsx),
          "; the first of each is used. Fix the source file.")
  knitr::kable(dup) %>% print()
  pad <- pad %>% distinct(mussel, thread, thread_trt, .keep_all = TRUE)
}

# Rows in the measurement file that match no trace: a typo, or a trace that was deleted.
orphan <- pad %>% anti_join(traces, by = c("mussel", "thread", "thread_trt"))
if (nrow(orphan) > 0) {
  warning(nrow(orphan), " row(s) in ", basename(pad_xlsx), " match no trace.")
  knitr::kable(orphan %>% select(mussel, thread, thread_trt, pad_area)) %>% print()
}

# Traces with no row at all in the measurement file (as opposed to a row with a blank value).
missing_rows <- traces %>% anti_join(pad, by = c("mussel", "thread", "thread_trt"))
if (nrow(missing_rows) > 0)
  message(nrow(missing_rows), " trace(s) have no row in ", basename(pad_xlsx),
          "; they have no pad_area, so scripts 3 and 4 exclude them until a row is added.")

traces <- traces %>% left_join(pad, by = c("mussel", "thread", "thread_trt"))

message("Plaque measurements joined: ", sum(!is.na(traces$pad_area)), " of ", nrow(traces))
```


## 3. Derive adhesion and assemble the thread summary

```{r}
# If script 1 ran without the mussel key, these columns are absent. Create them empty rather
# than failing, so the summary still builds and the gap is visible as blank cells.
for (col in c("species", "mussel_trt", "rna_sequenced"))
  if (!col %in% names(traces)) traces[[col]] <- NA_character_

summary_tbl <- traces %>%
  mutate(adhesion_kpa = max_force / pad_area * 1000,       # N / mm^2 -> kPa
         sort_ID      = paste0(mussel, "_", thread)) %>%
  arrange(match(source_folder, folder_order), mussel, thread) %>%
  transmute(species, sort_ID,
            mussel_ID  = mussel,
            thread_num = thread,
            note, phase, day,
            mussel_trt,                                # the arm the ANIMAL was assigned to
            thread_trt,                                # what the THREAD was built in
            max_force, integral, max_displacement,
            pad_area, adhesion_kpa, failure,
            rna_sequenced,
            pad_notes,
            source_folder, file)                       # provenance, drop before pasting

message("Candidate rows: ", nrow(summary_tbl),
        " | with adhesion_kpa: ", sum(!is.na(summary_tbl$adhesion_kpa)))
```


## 4. Write the outputs

```{r}
write.xlsx(list(data = summary_tbl),
           file = file.path(out_dir, "thread-summary.xlsx"),
           colNames = TRUE, overwrite = TRUE)

message("Wrote ", file.path(out_dir, "thread-summary.xlsx"),
        " (", nrow(summary_tbl), " rows).")
```

Review the output against the QC plots in
`03_analyses/01_extract-tensometer-data/QC_plots/<source_folder>/<file>.jpg`, using `note` and
`failure` to spot bad runs. Scripts 3 and 4 read
`03_analyses/02_assemble-thread-summary/thread-summary.xlsx` (sheet `data`) and ignore extra
columns. Re-running this script overwrites that file, so any hand edits made to it are lost.
