---
title: "00 - Run the Iso-Seq branch"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output: html_document
params:
  online: false         # steps 02-03 and 05-06 download inputs and run the aligners; false runs steps 04 and 07
  minimap2: "minimap2"  # binary paths for steps 02 and 03
  salmon: "salmon"
  bin: ""               # step 06: folder with hisat2, samtools, stringtie and featureCounts ("" = PATH)
  threads: 4
  steps: "all"          # "all", or step numbers, e.g. [4]
  stop_on_fail: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs the Iso-Seq branch: the same differential expression as `06_differential-expression`,
but with the Tag-seq reads quantified against the Iso-Seq transcriptome instead of the genome,
and a comparison of the two. Each step runs in its own R process with an HTML report and a
log in `03_analyses/knit_html/` (git-ignored) and a `run_log.csv`.

| step | script | writes to `03_analyses/` | needs |
|---|---|---|---|
| 01 | `01_isoseq_transcriptome_check.Rmd` | nothing (a length QC; chunks not evaluated) | owl |
| 02 | `02_isoform_gene_map.Rmd` | `02_isoform-gene-map/`: each isoform's genome gene | owl, NCBI, minimap2; about 1.5 hours, 9 GB of memory |
| 03 | `03_salmon_quant.Rmd` | `03_salmon/`: mapping summary and gene counts | gannet, salmon; about 15 minutes for the index plus 3 to 4 minutes per library |
| 04 | `04_isoseq_de_comparison.Rmd` | `04_isoseq-de/`: TC contrasts on the Iso-Seq counts and the comparison with the genome | committed files only |
| 05 | `05_augmented_annotation.Rmd` | `05_augmented-annotation/`: option B, the RefSeq annotation augmented with the isoforms (3' extension; full models) | step 02's alignment; a few minutes |
| 06 | `06_genome_recount.Rmd` | `06_genome-recount/`: option B, the reads realigned (HISAT2) and counted on each annotation (StringTie + prepDE, featureCounts) | gannet, NCBI, HISAT2, StringTie, Subread; about 15 minutes for the index plus 4 minutes per library |
| 07 | `07_augmented_de_comparison.Rmd` | `07_augmented-de/`: option B, the TC contrasts on each recount, compared with the record | committed files only |

With `online: false` (the default) steps 04 and 07 run, from the committed counts. Set
`online: true`, with `minimap2`, `salmon` and `bin` pointing at the binaries (minimap2 2.31,
salmon 1.10.3, HISAT2 2.2.1, StringTie 2.2.1, Subread 2.1.1, samtools 1.24 were used, from
bioconda), to rebuild steps 02 to 07.

```{r run}
suppressPackageStartupMessages(library(here))
source(file.path(here::here(), "..", "tools", "run_steps.R"))
html_dir <- here::here("03_analyses", "knit_html")
steps <- if (!identical(params$steps, "all")) params$steps else if (isTRUE(params$online)) 2:7 else c(4, 7)
tool_params <- list("02" = list(minimap2 = params$minimap2, threads = params$threads),
                    "03" = list(salmon = params$salmon, threads = params$threads),
                    "05" = list(threads = params$threads),
                    "06" = list(bin = params$bin, threads = params$threads),
                    "07" = list(cores = params$threads))
run_log <- run_steps(here::here("01_code"), html_dir, steps = steps, step_params = tool_params,
                     stop_on_fail = isTRUE(params$stop_on_fail))
finish_run(run_log, html_dir, "Iso-Seq")
```
