---
title: "02 - Iso-Seq isoforms to genome genes"
output: html_document
params:
  minimap2: "minimap2"     # path to the minimap2 binary (2.31 used here)
  threads: 4
  min_identity: 0.90       # matching bases / alignment block length
  min_aln: 100             # aligned bases
  ambiguous_ratio: 0.98    # a second gene scoring at least this share of the best flags the isoform
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

The Iso-Seq transcriptome (`Mtros-hq_transcripts.fasta`, 411,251 high-quality isoforms named
`MT_Pool_HQ_transcript/N`) was clustered but never collapsed to genes, so its isoforms carry
no gene. This step assigns each isoform to a gene of the genome annotation
(GCF_036588685.1), the genes the genome branch (`05`, `06`) counts, so the two branches can
be compared gene by gene.

**Why the CDS and not the genome.** The usual route aligns isoforms to the genome with a
spliced aligner (minimap2 `-ax splice:hq`; Li 2018, *Bioinformatics* 34:3094) and assigns
each to the gene it overlaps. The genome FASTA is not reachable from the environment this was
built in (NCBI and ENA are blocked), but the annotation's CDS sequences are (gannet copy of
`cds_from_genomic.fasta`, 53,269 CDS of 34,235 genes, each header naming its gene). Each
isoform is therefore aligned to the CDS set (minimap2 `-cx asm20`, which tolerates the few
percent of allelic divergence between the Iso-Seq animals and the genome animal) and takes the
gene of its best alignment:

- on the same strand (Iso-Seq isoforms and CDS are both sense);
- at least `min_identity` identity over at least `min_aln` aligned bases;
- `ambiguous` when a different gene scores at least `ambiguous_ratio` of the best (paralogs);
  such isoforms keep the best gene and are flagged.

Isoforms without such a hit (non-coding transcripts, and 3' fragments that end before the
CDS starts) stay unassigned and are kept as their own features, named `isoseq:<isoform>`.
Genes without a CDS (non-coding genes) cannot receive isoforms this way, which limits the
comparison to coding genes. A genome-based map would remove both limits; it needs the genome
FASTA (`GCF_036588685.1`) and the StringTie coordinates already in
`05_sequence-alignment/03_analyses/hisat/t_data.ctab`.

Writes, in `03_analyses/02_isoform-gene-map/`:

- `isoform_gene_map.csv.gz`: one row per isoform: `isoform`, `gene` (LOC key via
  `gene_key()`, or `isoseq:<isoform>`), `status` (`assigned`, `ambiguous`, `unassigned`), the
  best CDS, identity, aligned bases, isoform and CDS lengths, and the runner-up gene;
- `isoform_gene_map_summary.csv`: isoforms and genes by status;
- `RUN_provenance.txt`: settings, minimap2 version and input MD5s.

The minimap2 alignments (`minimap2_isoforms_vs_cds.paf.gz`) are kept beside them but are
git-ignored (large, rebuilt by this script).

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))         # gene_key()
source(file.path(repo_root, "tools", "pipeline_checks.R"))  # provenance_lines()
suppressPackageStartupMessages({ library(data.table) })
```

# Inputs

```{r inputs}
if (!file.exists(isoseq_fa)) download.file(ISOSEQ_URL, isoseq_fa, mode = "wb")
if (!file.exists(cds_fa))    download.file(CDS_URL, cds_fa, mode = "wb")
cds_hdr <- system2("grep", c("'^>'", shQuote(cds_fa)), stdout = TRUE)
cds <- data.table(cds  = sub("^>(\\S+).*$", "\\1", cds_hdr),
                  gene = gene_key(sub("^.*\\[gene=([^]]+)\\].*$", "\\1", cds_hdr)))
stopifnot(!anyDuplicated(cds$cds), !anyNA(cds$gene))
cat(nrow(cds), "CDS of", uniqueN(cds$gene), "genes\n")
```

# Align

```{r align}
paf <- file.path(map_dir, "minimap2_isoforms_vs_cds.paf.gz")
mm2_version <- system2(params$minimap2, "--version", stdout = TRUE)
if (!file.exists(paf)) {
  cmd <- sprintf("%s -c -x asm20 -N 5 -t %d %s %s 2> %s | gzip > %s", params$minimap2, params$threads,
                 shQuote(cds_fa), shQuote(isoseq_fa), shQuote(file.path(map_dir, "minimap2.log")), shQuote(paf))
  stopifnot(system(cmd) == 0)
}
## PAF columns 1-11 and the alignment score (AS:i) of every alignment
aln <- fread(cmd = sprintf("zcat %s | awk -F'\\t' '{as=\"\"; for (i=13; i<=NF; i++) if ($i ~ /^AS:i:/) as=substr($i,6); print $1\"\\t\"$2\"\\t\"$5\"\\t\"$6\"\\t\"$7\"\\t\"$10\"\\t\"$11\"\\t\"as}'",
                             shQuote(paf)),
             col.names = c("isoform", "isoform_len", "strand", "cds", "cds_len", "matches", "block", "AS"),
             sep = "\t", header = FALSE)
aln[, identity := matches / block]
aln <- merge(aln, cds, by = "cds", all.x = TRUE)
stopifnot(!anyNA(aln$gene))
cat(nrow(aln), "alignments of", uniqueN(aln$isoform), "isoforms\n")
```

# Assign

```{r assign}
ok <- aln[strand == "+" & identity >= params$min_identity & matches >= params$min_aln]
setorder(ok, isoform, -AS)
ok[, best_gene := gene[1], by = isoform]
best   <- ok[, .SD[1], by = isoform]
## best alignment of a different gene, for the ambiguity flag
second <- ok[gene != best_gene, .(runner_up_gene = gene[1], runner_up_AS = as.numeric(AS[1])), by = isoform]
best <- merge(best, second, by = "isoform", all.x = TRUE)
best[, status := ifelse(!is.na(runner_up_AS) & runner_up_AS >= params$ambiguous_ratio * AS, "ambiguous", "assigned")]

all_iso <- data.table(isoform = sub("^>(\\S+).*$", "\\1",
                                    system2("grep", c("'^>'", shQuote(isoseq_fa)), stdout = TRUE)))
stopifnot(!anyDuplicated(all_iso$isoform))
map <- merge(all_iso, best[, .(isoform, gene, status, cds, identity = round(identity, 4), aligned = matches,
                               isoform_len, cds_len, AS, runner_up_gene, runner_up_AS)],
             by = "isoform", all.x = TRUE, sort = FALSE)
map[is.na(gene), `:=`(gene = paste0("isoseq:", isoform), status = "unassigned")]
## antisense-only isoforms, as a check on the strand rule
anti <- setdiff(unique(aln[strand == "-" & identity >= params$min_identity & matches >= params$min_aln, isoform]),
                best$isoform)
fwrite(map, file.path(map_dir, "isoform_gene_map.csv.gz"))

summ <- rbind(map[, .(isoforms = .N, genes = uniqueN(gene)), by = status],
              data.table(status = "all", isoforms = nrow(map), genes = uniqueN(map$gene)))
summ <- rbind(summ, data.table(status = "unassigned with only an antisense hit", isoforms = length(anti), genes = NA))
summ <- rbind(summ, data.table(status = "genes with a CDS that received an isoform", isoforms = NA,
                               genes = uniqueN(map[status != "unassigned", gene])))
fwrite(summ, file.path(map_dir, "isoform_gene_map_summary.csv"))
knitr::kable(summ)
```

# Provenance

```{r provenance}
settings <- c("[02_isoform_gene_map.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              paste0("minimap2             : ", mm2_version, " (-c -x asm20 -N 5)"),
              paste0("assignment           : same strand, identity >= ", params$min_identity, ", aligned >= ",
                     params$min_aln, " bp; ambiguous when another gene scores >= ", params$ambiguous_ratio, " of the best"),
              paste0("isoforms / CDS       : ", nrow(map), " / ", nrow(cds)))
writeLines(provenance_lines(settings, c(isoseq_fa, cds_fa, here::here("01_code", "02_isoform_gene_map.Rmd")),
                            repo_root, packages = c("data.table")),
           file.path(map_dir, "RUN_provenance.txt"))
```
