---
title: "02 - Iso-Seq isoforms to genome genes"
output: html_document
params:
  minimap2: "minimap2"     # path to the minimap2 binary (2.31 used here)
  threads: 4
  min_identity: 0.90       # gap-compressed identity, 1 - minimap2's de (each gap counts once)
  min_overlap: 50          # exonic bases shared with a gene, same strand
  ambiguous_ratio: 0.98    # a second gene sharing at least this share of the best overlap flags the isoform
  mt_min_aln: 100          # aligned bases to the mitogenome that flag an isoform (mt_hit)
  mt_min_cov: 0.5          # share of an isoform aligned to the mitogenome that makes it mitochondrial
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

The Iso-Seq transcriptome (`Mtros-hq_transcripts.fasta`, 411,251 high-quality isoforms named
`MT_Pool_HQ_transcript/N`) was clustered but never collapsed to genes, so its isoforms carry
no gene. This step places each isoform on the genome (RefSeq GCF_036588685.1, annotation
release RS_2024_02, the annotation the genome branch `05`/`06` counts against) and gives it
the gene whose exons it overlaps, so the two branches can be compared gene by gene.

1. **Align.** minimap2 spliced alignment of each isoform to the genome (`-x splice:hq -uf`:
   the preset for accurate full-length transcripts, aligned on their own strand; Li 2018,
   *Bioinformatics* 34:3094), keeping the best alignment of each isoform.
2. **Overlap.** The aligned blocks (exons) of the isoform against the exons of each annotated
   gene (all its transcripts merged), on the same strand. Every gene type counts (coding,
   lncRNA, tRNA, pseudogene), as in the genome branch.
3. **Assign.** An isoform aligned at `min_identity` or better (gap-compressed identity, one
   minus minimap2's `de`: each insertion or deletion counts once whatever its length, so a
   long unspliced gap does not swamp an otherwise exact alignment as it does in matching
   bases over block length) takes the gene with which it
   shares the most exonic bases, if at least `min_overlap`. It is `ambiguous` (it keeps the
   gene, flagged) when a second gene shares at least `ambiguous_ratio` of that overlap, or when
   minimap2 gives the alignment mapping quality 0 (an equally good placement elsewhere).
4. **Isoforms outside annotated genes.** A well-aligned isoform that overlaps no gene joins a
   `novel_locus`: isoforms whose genomic spans overlap on the same strand form one feature,
   named `novel:<sequence>:<start>-<end>:<strand>`. Unaligned isoforms and those below
   `min_identity` stay `unassigned`, each its own feature, `isoseq:<isoform>`.
5. **Mitochondrial transcripts.** An isoform is `mitochondrial` (its own feature,
   `mito:<isoform>`, which step 04 leaves out as 06 leaves out the mitochondrial loci) when its
   genome alignment lies on the mitogenome or shares most exonic bases with a mitochondrial
   locus (the genes of `06 .../count_matrix/mitochondrial_loci.csv`), or when at least
   `mt_min_cov` of it aligns to the mitogenome sequence itself (minimap2 `-x asm20`, any
   identity, either strand). The last test catches mitochondrial transcripts that the genome
   alignment placed on unannotated copies of mitochondrial sequence on unplaced scaffolds, or
   on a nuclear gene. This is decided before novel loci are formed, so those loci hold no
   mitochondrial isoforms. `mt_hit` also marks isoforms with a smaller mitochondrial segment
   (at least `mt_min_aln` bases aligned to the mitogenome, or any exonic overlap with a
   mitochondrial locus); they keep their gene, and step 04 leaves out any feature outside
   annotated genes that holds one.

An earlier version of this step, written while the genome was not reachable, assigned isoforms
through their alignment to the annotation's CDS (`_superseded/02_isoform_gene_map_cds.Rmd`).
That route reaches coding genes only and misses isoforms that end before the CDS. Its map is
kept (`03_analyses/_superseded/02_isoform-gene-map_cds/`) and compared with this one below.

Writes, in `03_analyses/02_isoform-gene-map/`:

- `isoform_gene_map.csv.gz`: one row per isoform: `isoform`, `gene` (a gene key via
  `gene_key()`, `novel:...`, `mito:<isoform>` or `isoseq:<isoform>`), `status` (`assigned`,
  `ambiguous`, `novel_locus`, `mitochondrial`, `unassigned`), the alignment (`seqid`, `strand`, `start`, `end`, `identity`
  (gap-compressed),
  `mapq`, `n_alignments`), `exonic_overlap` with the gene, the runner-up gene and its overlap,
  `mt_hit`, `mito_cov` (share of the isoform aligned to the mitogenome), `isoform_len` and
  `three_prime_beyond` (bases by which an assigned isoform's 3' end passes the last annotated
  exon of its gene; 0 when it ends within it);
- `isoform_gene_map_summary.csv`: isoforms and features by status, and the genome-branch genes
  reached;
- `cds_map_agreement.csv`: agreement with the retired CDS-based map;
- `RUN_provenance.txt`: settings, minimap2 version, input MD5s.

The alignments (`minimap2_isoforms_vs_genome.paf.gz`, `minimap2.log`, and
`mitogenome.fa` with `minimap2_isoforms_vs_mitogenome.paf.gz`) are kept beside them,
git-ignored (rebuilt by this script in about 1.5 hours on four cores, with about 7 GB of
memory for the genome index).

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))         # gene_key()
source(file.path(repo_root, "tools", "pipeline_checks.R"))  # provenance_lines()
suppressPackageStartupMessages({
  library(data.table); library(GenomicRanges); library(GenomicAlignments); library(rtracklayer)
})
```

# Inputs

```{r inputs}
if (!file.exists(isoseq_fa)) download.file(ISOSEQ_URL, isoseq_fa, mode = "wb")
for (f in c(genome_fa, gff_gz)) if (!file.exists(f)) download.file(paste0(NCBI_DIR, basename(f)), f, mode = "wb")
## NCBI's checksums for the two files
ncbi_md5 <- fread(cmd = paste("curl -sS", shQuote(paste0(NCBI_DIR, "md5checksums.txt"))), header = FALSE,
                  col.names = c("md5", "file"))[, file := basename(file)]
stopifnot(all(unname(tools::md5sum(c(genome_fa, gff_gz))) == ncbi_md5[match(basename(c(genome_fa, gff_gz)), file), md5]))

mt <- fread(file.path(de06, "count_matrix", "mitochondrial_loci.csv"))
mt_genes <- unique(mt$LOC_ID)
mt_seqids <- unique(mt[source == "mitogenome", seqid])
stopifnot(length(mt_genes) == nrow(mt), length(mt_seqids) == 1)
```

# Genes: exons of every annotated gene

```{r genes}
gff <- import(gff_gz)
parent1 <- sub(",.*$", "", as.character(unstrsplit(gff$Parent, sep = ",")))
parent1[!nzchar(parent1)] <- NA
## each feature's top-level gene: follow Parent from exon to transcript to gene
has_id  <- !is.na(gff$ID) & !duplicated(gff$ID) & !gff$ID %in% gff$ID[duplicated(gff$ID)]
up      <- setNames(parent1[has_id], gff$ID[has_id])
ex      <- gff[gff$type == "exon"]
top     <- parent1[gff$type == "exon"]
for (i in 1:5) { nxt <- up[top]; move <- !is.na(nxt); if (!any(move)) break; top[move] <- nxt[move] }
## only the mitogenome's rRNAs and tRNAs have no gene feature above them in RefSeq's GFF
stopifnot(!anyNA(top), all(startsWith(top, "gene-") | as.character(seqnames(ex)) %in% mt_seqids))
ex$gene <- gene_key(top)
gene_exons <- unlist(reduce(split(granges(ex), ex$gene)))            # each gene's exons merged
gene_exons$gene <- names(gene_exons); names(gene_exons) <- NULL
## each gene's annotated extent (first to last exon of any of its transcripts), for the 3' end test below
gene_span <- unlist(range(split(granges(ex), ex$gene)))
gene_span <- data.table(gene = names(gene_span), g_start = start(gene_span), g_end = end(gene_span),
                        g_strand = as.character(strand(gene_span)))[!duplicated(gene)]

## the genome branch's genes are all here
g06 <- gene_key(fread(file.path(de06, "count_matrix", "gene_count_matrix_clean.csv"), select = 1)[[1]])
stopifnot(!anyDuplicated(g06))
cat(length(unique(ex$gene)), "annotated genes with exons;", sum(g06 %in% ex$gene), "of the", length(g06),
    "genome-branch genes among them\n")
stopifnot(all(mt_genes %in% ex$gene))
```

# Align

```{r align}
paf <- file.path(map_dir, "minimap2_isoforms_vs_genome.paf.gz")
mm2_version <- system2(params$minimap2, "--version", stdout = TRUE)
mm2_args <- "-c -x splice:hq -uf --secondary=no"
if (!file.exists(paf)) {
  cmd <- sprintf("%s %s -t %d %s %s 2> %s | gzip > %s", params$minimap2, mm2_args, params$threads,
                 shQuote(genome_fa), shQuote(isoseq_fa), shQuote(file.path(map_dir, "minimap2.log")), shQuote(paf))
  stopifnot(system(cmd) == 0)
}
## PAF columns and the CIGAR (cg:Z) of every alignment
awk_prog <- tempfile(fileext = ".awk")
writeLines('BEGIN{FS=OFS="\\t"} {cg=""; de=""; for (i=13; i<=NF; i++) { if ($i ~ /^cg:Z:/) cg=substr($i,6); if ($i ~ /^de:f:/) de=substr($i,6) }; print $1,$2,$5,$6,$8,$9,$10,$11,$12,de,cg}',
           awk_prog)
aln <- fread(cmd = sprintf("zcat %s | awk -f %s", shQuote(paf), shQuote(awk_prog)), sep = "\t", header = FALSE,
             col.names = c("isoform", "isoform_len", "strand", "seqid", "tstart", "tend", "matches", "block", "mapq", "de", "cigar"))
stopifnot(all(nzchar(aln$cigar)), !anyNA(aln$de))
aln[, n_alignments := .N, by = isoform]
## chimeric isoforms can give more than one alignment: keep the one with most matching bases
setorder(aln, isoform, -matches)
aln <- aln[!duplicated(isoform)]
aln[, identity := 1 - de]                                                      # gap-compressed identity
cat(nrow(aln), "isoforms aligned;", aln[n_alignments > 1, .N], "with more than one alignment\n")
```

# Overlap with gene exons

```{r overlap}
blocks <- extractAlignmentRangesOnReference(aln$cigar, pos = aln$tstart + 1L)   # 1-based, introns (N) split
blk <- GRanges(rep(aln$seqid, lengths(blocks)), unlist(blocks), strand = rep(aln$strand, lengths(blocks)))
blk$i <- rep(seq_len(nrow(aln)), lengths(blocks))
## the blocks start and end where the alignment does
stopifnot(all(min(splitAsList(start(blk), blk$i)) == aln$tstart + 1L), all(max(splitAsList(end(blk), blk$i)) == aln$tend))

hits <- findOverlaps(blk, gene_exons)                                          # same strand
ov <- data.table(i = blk$i[queryHits(hits)], gene = gene_exons$gene[subjectHits(hits)],
                 bp = width(pintersect(blk[queryHits(hits)], gene_exons[subjectHits(hits)])))
ov <- ov[, .(bp = sum(bp)), by = .(i, gene)]
setorder(ov, i, -bp)
best   <- ov[!duplicated(i)]
second <- ov[duplicated(i)][!duplicated(i), .(i, runner_up_gene = gene, runner_up_overlap = bp)]
aln[, i := .I]
aln <- merge(aln, best[, .(i, gene, exonic_overlap = bp)], by = "i", all.x = TRUE)
aln <- merge(aln, second, by = "i", all.x = TRUE)
aln[is.na(exonic_overlap), exonic_overlap := 0L]

## mitochondrial: on the mitogenome, or any exonic overlap with a mitochondrial locus, at any identity
mt_iso <- unique(ov[gene %in% mt_genes, i])
aln[, mt_hit := seqid %in% mt_seqids | i %in% mt_iso]

## mitochondrial by sequence: isoforms aligned to the mitogenome itself
mito_fa  <- file.path(map_dir, "mitogenome.fa")
mito_paf <- file.path(map_dir, "minimap2_isoforms_vs_mitogenome.paf.gz")
if (!file.exists(mito_paf)) {
  g <- Biostrings::readDNAStringSet(genome_fa)
  Biostrings::writeXStringSet(g[sub(" .*$", "", names(g)) %in% mt_seqids], mito_fa)
  rm(g); invisible(gc())
  stopifnot(system(sprintf("%s -c -x asm20 -t %d %s %s 2>/dev/null | gzip > %s", params$minimap2, params$threads,
                           shQuote(mito_fa), shQuote(isoseq_fa), shQuote(mito_paf))) == 0)
}
mito_aln <- fread(cmd = sprintf("zcat %s | cut -f1,2,10", shQuote(mito_paf)), header = FALSE, col.names = c("isoform", "qlen", "matches"))
mito_tab <- mito_aln[, .(mito_bp = max(matches), mito_cov = max(matches) / qlen[1]), by = isoform]
mt_seq_iso <- mito_tab[mito_bp >= params$mt_min_aln, isoform]
cat(length(mt_seq_iso), "isoforms with >=", params$mt_min_aln, "bases aligned to the mitogenome;",
    sum(!mt_seq_iso %in% aln[mt_hit == TRUE, isoform]), "of them not caught by the genome alignment\n")
```

# Assign

```{r assign}
aln[, status := fifelse(identity < params$min_identity, "unassigned",
               fifelse(exonic_overlap >= params$min_overlap, "assigned", "novel_locus"))]
aln[status == "assigned" & (mapq == 0 | (!is.na(runner_up_overlap) & runner_up_overlap >= params$ambiguous_ratio * exonic_overlap)),
    status := "ambiguous"]
aln[!status %in% c("assigned", "ambiguous"), `:=`(gene = NA_character_)]

## mitochondrial isoforms become their own features before novel loci are formed
aln <- merge(aln, mito_tab[, .(isoform, mito_cov)], by = "isoform", all.x = TRUE)
aln[is.na(mito_cov), mito_cov := 0]
aln[, mt_hit := mt_hit | isoform %in% mt_seq_iso]
aln[seqid %in% mt_seqids | (status %in% c("assigned", "ambiguous") & gene %in% mt_genes) | mito_cov >= params$mt_min_cov,
    `:=`(status = "mitochondrial", gene = paste0("mito:", isoform), mt_hit = TRUE)]

## novel loci: overlapping genomic spans of well-aligned isoforms outside genes, per strand
nv <- aln[status == "novel_locus"]
spans <- GRanges(nv$seqid, IRanges(nv$tstart + 1L, nv$tend), strand = nv$strand)
loci  <- reduce(spans)
loc_name <- sprintf("novel:%s:%d-%d:%s", seqnames(loci), start(loci), end(loci), strand(loci))
aln[status == "novel_locus", gene := loc_name[findOverlaps(spans, loci, select = "first")]]

all_iso <- data.table(isoform = sub("^>(\\S+).*$", "\\1",
                                    system2("grep", c("'^>'", shQuote(isoseq_fa)), stdout = TRUE)))
stopifnot(!anyDuplicated(all_iso$isoform), all(aln$isoform %in% all_iso$isoform))
map <- merge(all_iso, aln[, .(isoform, gene, status, seqid, strand, start = tstart + 1L, end = tend,
                              identity = round(identity, 4), mapq, n_alignments, exonic_overlap,
                              runner_up_gene, runner_up_overlap, mt_hit, mito_cov = round(mito_cov, 3), isoform_len)],
             by = "isoform", all.x = TRUE, sort = FALSE)
map[is.na(status), `:=`(status = "unassigned", mt_hit = FALSE, n_alignments = 0L, mito_cov = 0)]
## how far an assigned isoform's 3' end lies beyond its gene's annotated 3' end (bases; 0 if within)
map <- merge(map, gene_span, by = "gene", all.x = TRUE, sort = FALSE)
map[status %in% c("assigned", "ambiguous"),
    three_prime_beyond := pmax(0L, fifelse(strand == "+", end - g_end, g_start - start))]
map[, c("g_start", "g_end", "g_strand") := NULL]
## unaligned isoforms that are mitochondrial by sequence
um <- map$status == "unassigned" & map$n_alignments == 0 & map$isoform %in% mito_tab[mito_cov >= params$mt_min_cov, isoform]
map[um, `:=`(status = "mitochondrial", gene = paste0("mito:", isoform), mt_hit = TRUE)]
map[um, mito_cov := round(mito_tab$mito_cov[match(map$isoform[um], mito_tab$isoform)], 3)]
map[status != "mitochondrial" & isoform %in% mt_seq_iso, mt_hit := TRUE]
map[is.na(gene), gene := paste0("isoseq:", isoform)]
map <- map[match(all_iso$isoform, isoform)]                                    # FASTA order
stopifnot(!anyNA(map$gene), nrow(map) == nrow(all_iso))
fwrite(map, file.path(map_dir, "isoform_gene_map.csv.gz"))

summ <- rbind(map[, .(isoforms = .N, features = uniqueN(gene)), by = status],
              data.table(status = "all", isoforms = nrow(map), features = uniqueN(map$gene)),
              data.table(status = "unaligned (within unassigned)", isoforms = map[n_alignments == 0, .N], features = NA),
              data.table(status = "flagged mt_hit (mitochondrial, or with a mitochondrial segment)", isoforms = map[mt_hit == TRUE, .N], features = NA),
              data.table(status = "genome-branch genes that received an isoform", isoforms = NA,
                         features = sum(g06 %in% map[status %in% c("assigned", "ambiguous"), gene])))
fwrite(summ, file.path(map_dir, "isoform_gene_map_summary.csv"))
knitr::kable(summ)
```

# Agreement with the CDS-based map

```{r cds-agreement}
old <- fread(cds_map, select = c("isoform", "gene", "status"))
both <- merge(map[, .(isoform, gene_genome = gene, status_genome = status)],
              old[, .(isoform, gene_cds = gene, status_cds = status)], by = "isoform")
## nuclear genes only: leave out isoforms either map calls mitochondrial
both <- both[status_genome != "mitochondrial" & !(status_cds %in% c("assigned", "ambiguous") & gene_cds %in% mt_genes)]
on_gene <- function(s) s %in% c("assigned", "ambiguous")
agree <- data.table(
  measure = c("isoforms on a gene in both maps", "  of which the same gene", "  of which the same gene, both unambiguous",
              "isoforms on a gene in the CDS map only", "isoforms on a gene in the genome map only",
              "genes reached by the CDS map", "genes reached by the genome map", "genes reached by both"),
  n = c(both[on_gene(status_genome) & on_gene(status_cds), .N],
        both[on_gene(status_genome) & on_gene(status_cds) & gene_genome == gene_cds, .N],
        both[status_genome == "assigned" & status_cds == "assigned" & gene_genome == gene_cds, .N],
        both[!on_gene(status_genome) & on_gene(status_cds), .N],
        both[on_gene(status_genome) & !on_gene(status_cds), .N],
        uniqueN(both[on_gene(status_cds), gene_cds]), uniqueN(both[on_gene(status_genome), gene_genome]),
        length(intersect(both[on_gene(status_cds), gene_cds], both[on_gene(status_genome), gene_genome]))))
fwrite(agree, file.path(map_dir, "cds_map_agreement.csv"))
knitr::kable(agree)
```

# Provenance

```{r provenance}
settings <- c("[02_isoform_gene_map.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              paste0("minimap2             : ", mm2_version, " (", mm2_args, ")"),
              paste0("genome / annotation  : ", ASSEMBLY, " (RefSeq annotation RS_2024_02), MD5s checked against NCBI"),
              paste0("assignment           : gap-compressed identity (1 - de) >= ", params$min_identity, "; most exonic bases shared with a gene, same strand, >= ",
                     params$min_overlap, "; ambiguous when mapq 0 or a second gene shares >= ", params$ambiguous_ratio, " of that"),
              paste0("isoforms             : ", nrow(map)))
writeLines(provenance_lines(settings, c(isoseq_fa, genome_fa, gff_gz, file.path(de06, "count_matrix", "mitochondrial_loci.csv"),
                                        file.path(de06, "count_matrix", "gene_count_matrix_clean.csv"), cds_map,
                                        here::here("01_code", "02_isoform_gene_map.Rmd")),
                            repo_root, packages = c("data.table", "GenomicRanges", "GenomicAlignments", "rtracklayer")),
           file.path(map_dir, "RUN_provenance.txt"))
```
