---
title: "02 - Quantify the Tag-seq reads against the Iso-Seq transcriptome (salmon)"
output: html_document
params:
  salmon: "salmon"          # path to the salmon binary (1.10.3 used here)
  threads: 4
  k: 31                     # index k-mer length (trimmed reads are 20-95 bp, mean about 63)
  ext_bp: 100               # an isoform ending this far past its gene's annotated 3' end counts as extended
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Pseudo-aligns each library's trimmed reads to the Iso-Seq isoforms with salmon (selective
alignment; Patro et al. 2017, *Nature Methods* 14:417) and sums the isoforms of each gene
(`05_sequence-alignment` step 04's map) with tximport (Soneson, Love and Robinson 2015,
*F1000Research* 4:1521).

- **Reads:** the 131 trimmed, lane-merged files the genome branch aligned with HISAT2
  (`*_L099_R1_cmb.trim.fastq.gz`, gannet `panopea/PSMFC-mytilus-byssus-pilot/byssus-exp-analysis/data/raw-trimmed/`;
  leader-clipped and deduplicated, see `05_sequence-alignment/README.md`, "The reads").
  Each is downloaded to a temporary file, quantified and deleted.
- **No length correction.** 3' Tag-seq yields about one read per transcript regardless of its
  length, so `salmon quant --noLengthCorrection`, and tximport with
  `countsFromAbundance = "no"`: the gene counts are summed read counts, passed to DESeq2
  without a length offset, as the tximport vignette advises for 3' tagged RNA-seq.
- **Library type** is detected by salmon (`-l A`); the detected type is reported.
- **No decoys.** Selective alignment can use the genome as decoy sequence to absorb reads from
  unannotated loci; the genome FASTA is not available here, so none is used.

Writes, in `03_analyses/02_salmon/`:

- `salmon_mapping_summary.csv`: per library, reads processed, mapped, the mapping rate and the
  detected library type;
- `gene_counts.csv.gz`: genes (the map's `gene`) x libraries, summed read counts; column
  names are library IDs as in `06`'s count matrix (`T001F`);
- `read_classes_by_library.csv`: per library, the assigned reads by what their isoforms are
  (annotated gene within or beyond its annotated 3' end, novel locus, unassigned isoform,
  mitochondrial), as reads and as a share (see "Where the reads go");
- `RUN_provenance.txt`.

The index (`index/`) and the per-library salmon output (`quant/<library>/`) are git-ignored:
large, and rebuilt by this script (about 15 minutes for the index and 3 to 4 minutes per library
on four cores).

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # provenance_lines()
suppressPackageStartupMessages({ library(data.table); library(tximport); library(jsonlite) })
idx      <- file.path(salmon_dir, "index")
quant_root <- file.path(salmon_dir, "quant")
dir.create(quant_root, showWarnings = FALSE)
salmon_version <- system2(params$salmon, "--version", stdout = TRUE)
```

# Index

```{r index}
if (!file.exists(file.path(idx, "versionInfo.json"))) {
  stopifnot(file.exists(isoseq_fa))
  stopifnot(system2(params$salmon, c("index", "-t", shQuote(isoseq_fa), "-i", shQuote(idx),
                                     "-k", params$k, "-p", params$threads),
                    stdout = file.path(salmon_dir, "index.log"), stderr = file.path(salmon_dir, "index.log")) == 0)
}
dups <- file.path(idx, "duplicate_clusters.tsv")
cat("isoforms collapsed as exact duplicates by salmon:",
    if (file.exists(dups)) max(0, length(readLines(dups)) - 1) else 0, "\n")
```

# Quantify every library

```{r quant}
listing <- readLines(url(READS_URL), warn = FALSE)
files <- unique(regmatches(listing, regexpr("[A-H][0-9]+-T[0-9]+[A-Z]*_S[0-9]+_L099_R1_cmb\\.trim\\.fastq\\.gz", listing)))
lib_of <- function(f) sub("^[A-H][0-9]+-", "", sub("_S[0-9]+_L099.*$", "", f))
stopifnot(length(files) == 131, !anyDuplicated(lib_of(files)))
for (f in files) {
  out <- file.path(quant_root, lib_of(f))
  if (file.exists(file.path(out, "quant.sf"))) next
  tmp <- tempfile(fileext = ".fastq.gz")
  download.file(paste0(READS_URL, f), tmp, mode = "wb", quiet = TRUE)
  rc <- system2(params$salmon, c("quant", "-i", shQuote(idx), "-l", "A", "-r", shQuote(tmp),
                                 "--noLengthCorrection", "-p", params$threads, "-o", shQuote(out)),
                stdout = FALSE, stderr = FALSE)
  unlink(tmp)
  if (rc != 0) stop("salmon quant failed for ", f)
}
```

# Mapping summary

```{r summary}
summ <- rbindlist(lapply(files, function(f) {
  q <- file.path(quant_root, lib_of(f))
  m <- fromJSON(file.path(q, "aux_info", "meta_info.json"))
  l <- fromJSON(file.path(q, "lib_format_counts.json"))
  data.table(library = lib_of(f), file = f, reads = m$num_processed, mapped = m$num_mapped,
             percent_mapped = round(m$percent_mapped, 2), library_type = l$expected_format)
}))
setorder(summ, library)
fwrite(summ, file.path(salmon_dir, "salmon_mapping_summary.csv"))
knitr::kable(summ[, .(libraries = .N, median_reads = median(reads), median_percent_mapped = median(percent_mapped),
                      min_percent_mapped = min(percent_mapped)), by = library_type])
```

# Gene counts

```{r tximport}
map <- fread(file.path(map_dir, "isoform_gene_map.csv.gz"), select = c("isoform", "gene"))
qf <- setNames(file.path(quant_root, summ$library, "quant.sf"), summ$library)
txi <- tximport(qf, type = "salmon", tx2gene = as.data.frame(map[, .(isoform, gene)]),
                countsFromAbundance = "no")
counts <- txi$counts
cat(nrow(counts), "genes and unassigned isoforms;", sum(rowSums(counts) > 0), "with any read\n")
fwrite(data.table(gene = rownames(counts), round(counts, 3)), file.path(salmon_dir, "gene_counts.csv.gz"))
```

# Where the reads go

Per library, the reads salmon assigned, split by what the map made of their isoforms: an
annotated gene, with the isoform ending within `ext_bp` of the gene's annotated 3' end or
further out (Tag-seq reads sit at the 3' end, so reads of an extended isoform probably fall
partly beyond the annotation, where the genome branch, which counts annotated exons, does not
count them); a novel locus; an unassigned isoform; or a mitochondrial feature. Writes
`read_classes_by_library.csv`. This is the evidence for whether an annotation extended with the
Iso-Seq isoforms would add reads to the genome branch.

```{r read-classes}
cls_map <- fread(file.path(map_dir, "isoform_gene_map.csv.gz"), select = c("isoform", "status", "three_prime_beyond"))
cls_map[, class := fifelse(status %in% c("assigned", "ambiguous"),
                           fifelse(three_prime_beyond > params$ext_bp, "gene_3prime_extended", "gene_within_annotation"),
                           fifelse(status == "novel_locus", "novel_locus",
                                   fifelse(status == "mitochondrial", "mitochondrial", "unassigned_isoform")))]
classes <- c("gene_within_annotation", "gene_3prime_extended", "novel_locus", "unassigned_isoform", "mitochondrial")
rc <- rbindlist(lapply(summ$library, function(lib) {
  q <- fread(file.path(quant_root, lib, "quant.sf"), select = c("Name", "NumReads"))
  q[, class := cls_map$class[match(Name, cls_map$isoform)]]
  stopifnot(!anyNA(q$class))
  s <- q[, .(reads = sum(NumReads)), by = class]
  data.table(library = lib, t(setNames(s$reads[match(classes, s$class)], classes)))
}))
for (k in classes) set(rc, which(is.na(rc[[k]])), k, 0)
rc[, assigned := rowSums(.SD), .SDcols = classes]
rc <- merge(summ[, .(library, reads_processed = reads, percent_mapped)], rc, by = "library")
for (k in classes) rc[, paste0("pct_", k) := round(100 * get(k) / assigned, 2)]
rc[, (classes) := lapply(.SD, round, 1), .SDcols = classes]
fwrite(rc, file.path(salmon_dir, "read_classes_by_library.csv"))
knitr::kable(rc[, lapply(.SD, function(x) round(median(x), 1)), .SDcols = paste0("pct_", classes)],
             caption = "median share of assigned reads, per class (%)")
```

# Provenance

```{r provenance}
settings <- c("[02_salmon_quant.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              paste0("salmon               : ", salmon_version, " (index -k ", params$k,
                     "; quant -l A --noLengthCorrection, no decoys)"),
              paste0("reads                : ", READS_URL, " (", length(files), " files)"),
              "gene counts          : tximport countsFromAbundance = \"no\" over 05_sequence-alignment step 04's isoform-to-gene map")
writeLines(provenance_lines(settings, c(isoseq_fa, file.path(map_dir, "isoform_gene_map.csv.gz"),
                                        here::here("01_code", "02_salmon_quant.Rmd")),
                            repo_root, packages = c("tximport", "data.table")),
           file.path(salmon_dir, "RUN_provenance.txt"))
```
