---
title: "03 - Quantify the Tag-seq reads against the Iso-Seq transcriptome (salmon)"
output: html_document
params:
  salmon: "salmon"          # path to the salmon binary (1.10.3 used here)
  threads: 4
  k: 31                     # index k-mer length (trimmed reads are 20-95 bp, mean about 63)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Pseudo-aligns each library's trimmed reads to the Iso-Seq isoforms with salmon (selective
alignment; Patro et al. 2017, *Nature Methods* 14:417) and sums the isoforms of each gene
(step 02's map) with tximport (Soneson, Love and Robinson 2015, *F1000Research* 4:1521).

- **Reads:** the 131 trimmed, lane-merged files the genome branch aligned with HISAT2
  (`*_L099_R1_cmb.trim.fastq.gz`, gannet `panopea/PSMFC-mytilus-byssus-pilot/byssus-exp-analysis/data/raw-trimmed/`;
  leader-clipped and deduplicated, see `05_sequence-alignment/README.md`, "The reads").
  Each is downloaded to a temporary file, quantified and deleted.
- **No length correction.** 3' Tag-seq yields about one read per transcript regardless of its
  length, so `salmon quant --noLengthCorrection`, and tximport with
  `countsFromAbundance = "no"`: the gene counts are summed read counts, passed to DESeq2
  without a length offset, as the tximport vignette advises for 3' tagged RNA-seq.
- **Library type** is detected by salmon (`-l A`); the detected type is reported.
- **No decoys.** Selective alignment can use the genome as decoy sequence to absorb reads from
  unannotated loci; the genome FASTA is not available here, so none is used.

Writes, in `03_analyses/03_salmon/`:

- `salmon_mapping_summary.csv`: per library, reads processed, mapped, the mapping rate and the
  detected library type;
- `gene_counts.csv.gz`: genes (step 02's `gene`) x libraries, summed read counts; column
  names are library IDs as in `06`'s count matrix (`T001F`);
- `RUN_provenance.txt`.

The index (`index/`) and the per-library salmon output (`quant/<library>/`) are git-ignored:
large, and rebuilt by this script (about one hour for the index and a few minutes per library
on four cores).

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # provenance_lines()
suppressPackageStartupMessages({ library(data.table); library(tximport); library(jsonlite) })
idx      <- file.path(salmon_dir, "index")
quant_root <- file.path(salmon_dir, "quant")
dir.create(quant_root, showWarnings = FALSE)
salmon_version <- system2(params$salmon, "--version", stdout = TRUE)
```

# Index

```{r index}
if (!file.exists(file.path(idx, "versionInfo.json"))) {
  stopifnot(file.exists(isoseq_fa))
  stopifnot(system2(params$salmon, c("index", "-t", shQuote(isoseq_fa), "-i", shQuote(idx),
                                     "-k", params$k, "-p", params$threads),
                    stdout = file.path(salmon_dir, "index.log"), stderr = file.path(salmon_dir, "index.log")) == 0)
}
dups <- file.path(idx, "duplicate_clusters.tsv")
cat("isoforms collapsed as exact duplicates by salmon:",
    if (file.exists(dups)) max(0, length(readLines(dups)) - 1) else 0, "\n")
```

# Quantify every library

```{r quant}
listing <- readLines(url(READS_URL), warn = FALSE)
files <- unique(regmatches(listing, regexpr("[A-H][0-9]+-T[0-9]+[A-Z]*_S[0-9]+_L099_R1_cmb\\.trim\\.fastq\\.gz", listing)))
lib_of <- function(f) sub("^[A-H][0-9]+-", "", sub("_S[0-9]+_L099.*$", "", f))
stopifnot(length(files) == 131, !anyDuplicated(lib_of(files)))
for (f in files) {
  out <- file.path(quant_root, lib_of(f))
  if (file.exists(file.path(out, "quant.sf"))) next
  tmp <- tempfile(fileext = ".fastq.gz")
  download.file(paste0(READS_URL, f), tmp, mode = "wb", quiet = TRUE)
  rc <- system2(params$salmon, c("quant", "-i", shQuote(idx), "-l", "A", "-r", shQuote(tmp),
                                 "--noLengthCorrection", "-p", params$threads, "-o", shQuote(out)),
                stdout = FALSE, stderr = FALSE)
  unlink(tmp)
  if (rc != 0) stop("salmon quant failed for ", f)
}
```

# Mapping summary

```{r summary}
summ <- rbindlist(lapply(files, function(f) {
  q <- file.path(quant_root, lib_of(f))
  m <- fromJSON(file.path(q, "aux_info", "meta_info.json"))
  l <- fromJSON(file.path(q, "lib_format_counts.json"))
  data.table(library = lib_of(f), file = f, reads = m$num_processed, mapped = m$num_mapped,
             percent_mapped = round(m$percent_mapped, 2), library_type = l$expected_format)
}))
setorder(summ, library)
fwrite(summ, file.path(salmon_dir, "salmon_mapping_summary.csv"))
knitr::kable(summ[, .(libraries = .N, median_reads = median(reads), median_percent_mapped = median(percent_mapped),
                      min_percent_mapped = min(percent_mapped)), by = library_type])
```

# Gene counts

```{r tximport}
map <- fread(file.path(map_dir, "isoform_gene_map.csv.gz"), select = c("isoform", "gene"))
qf <- setNames(file.path(quant_root, summ$library, "quant.sf"), summ$library)
txi <- tximport(qf, type = "salmon", tx2gene = as.data.frame(map[, .(isoform, gene)]),
                countsFromAbundance = "no")
counts <- txi$counts
cat(nrow(counts), "genes and unassigned isoforms;", sum(rowSums(counts) > 0), "with any read\n")
fwrite(data.table(gene = rownames(counts), round(counts, 3)), file.path(salmon_dir, "gene_counts.csv.gz"))
```

# Provenance

```{r provenance}
settings <- c("[03_salmon_quant.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              paste0("salmon               : ", salmon_version, " (index -k ", params$k,
                     "; quant -l A --noLengthCorrection, no decoys)"),
              paste0("reads                : ", READS_URL, " (", length(files), " files)"),
              "gene counts          : tximport countsFromAbundance = \"no\" over step 02's isoform-to-gene map")
writeLines(provenance_lines(settings, c(isoseq_fa, file.path(map_dir, "isoform_gene_map.csv.gz"),
                                        here::here("01_code", "03_salmon_quant.Rmd")),
                            repo_root, packages = c("tximport", "data.table")),
           file.path(salmon_dir, "RUN_provenance.txt"))
```
