---
title: "04 - Differential expression on the Iso-Seq reference, compared with the genome"
output: html_document
params:
  padj_cutoff: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Fits the treatment-control (TC) contrasts of record to the Iso-Seq gene counts of step 03,
exactly as `06_differential-expression` fits them to the genome counts, and compares the two
references gene by gene. It answers whether the DEG results depend on the reference.

- **Same samples, model, order and thresholds as 06:** the samples of each contrast in
  `06 .../DEG_lists/contrast_samples.csv`, `~ treatment` with the day-3 control as reference,
  fitted on every feature, then features kept with at least 10 counts in a third of the
  samples, apeglm shrinkage (Zhu, Ibrahim and Love 2019, *Bioinformatics* 35:2084), DEG at
  padj < 0.05. Counts are summed read counts (step 03), rounded, with no length offset.
- **Mitochondrial genes** are left out, as in 06: the genes of
  `06 .../count_matrix/mitochondrial_loci.csv`, the mitochondrial isoforms of step 02 (features
  `mito:<isoform>`), and every feature outside annotated genes (novel locus or unassigned
  isoform) holding an isoform with a mitochondrial segment (`mt_hit`).
- **Comparison:** genes are matched by their key (`gene_key()`); step 02 placed the isoforms
  on the same annotation the genome branch counts. For each contrast: DEGs in both, in one
  reference only, and genome DEGs whose gene received no Iso-Seq isoform; the correlation of
  log2 fold changes over genes tested in both; and the DEGs among features outside annotated
  genes (novel loci and unassigned isoforms).

Writes, in `03_analyses/04_isoseq-de/`:

- `<code>_isoseq_apeglm.csv`: every tested feature with its apeglm log2 fold change, Wald p
  and padj (`feature` is a gene key, `novel:<locus>` or `isoseq:<isoform>`);
- `isoseq_DEG_counts.csv`: features tested and DEGs (up, down) per contrast, with the genome
  numbers beside them;
- `reference_agreement.csv`: per contrast, DEG overlap and fold-change agreement;
- `counts_per_gene_both_references.csv`: for every nuclear gene counted on both references,
  its total count over the day-3 libraries on each, and their ratio;
- `isoseq_only_DEGs.csv`: the DEGs found on the Iso-Seq reference only, with their mean count
  and fold change on each reference (`tested_on_genome` FALSE when the genome filter dropped
  the gene);
- `FIG_isoseq_vs_genome_lfc.png`: log2 fold change on each reference for the genes tested in
  both, per contrast;
- `FIG_mapping_rates.png`: per library, the share of reads salmon assigned to the Iso-Seq
  isoforms against the HISAT2 alignment rate to the genome;
- `RUN_provenance.txt`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
source(file.path(repo_root, "tools", "pipeline_checks.R"))
suppressPackageStartupMessages({ library(DESeq2); library(apeglm); library(data.table); library(ggplot2) })
counts_dt <- fread(file.path(salmon_dir, "gene_counts.csv.gz"))
counts <- as.matrix(counts_dt[, -1]); rownames(counts) <- counts_dt$gene
contrasts <- fread(file.path(de06, "DEG_lists", "contrasts.csv"))[family == "TC"]
cs   <- fread(file.path(de06, "DEG_lists", "contrast_samples.csv"))
ti   <- fread(file.path(de06, "count_matrix", "treatmentinfo_clean.csv"))
mt   <- fread(file.path(de06, "count_matrix", "mitochondrial_loci.csv"))
hisat_qc <- file.path(repo_root, "05_sequence-alignment", "03_analyses", "hisat", "multiqc_data", "multiqc_bowtie2.txt")
stopifnot(nrow(contrasts) == 6, all(cs[code %in% contrasts$code, sample.1] %in% colnames(counts)))
mt_keys <- unique(mt$LOC_ID)
iso_map <- fread(file.path(map_dir, "isoform_gene_map.csv.gz"), select = c("gene", "status", "mt_hit"))
off_gene <- function(f) grepl("^(novel|isoseq|mito):", f)               # features outside annotated genes
mt_like <- unique(c(iso_map[status == "mitochondrial", gene], iso_map[off_gene(gene) & mt_hit == TRUE, gene]))
cat("Iso-Seq features:", nrow(counts), "| left out: mitochondrial genes", sum(rownames(counts) %in% mt_keys),
    ", mitochondrial isoforms and features outside genes holding one", sum(rownames(counts) %in% mt_like), "\n")
counts <- counts[!rownames(counts) %in% c(mt_keys, mt_like), ]
```

# Fit each TC contrast

```{r fit}
## the order of 06 steps 03-04: fit every feature of the contrast's samples, then keep features
## with >= 10 counts in a third of them, then apeglm on the kept ones, whose pvalue and padj
## lfcShrink takes from results(). Features with no read in these samples are dropped first;
## DESeq2 ignores all-zero rows, so this changes nothing but the run time.
fit_one <- function(row) {
  smp <- cs[code == row$code, sample.1]
  cd  <- as.data.frame(ti[match(smp, sample.1), .(sample.1, treatment)]); rownames(cd) <- cd$sample.1
  cd$treatment <- factor(cd$treatment, levels = c(row$ref_level, row$test_level))
  stopifnot(!anyNA(cd$treatment))
  m <- round(counts[, smp])
  m <- m[rowSums(m) > 0, ]
  dds <- DESeq(DESeqDataSetFromMatrix(m, cd, ~ treatment), quiet = TRUE)
  stopifnot(resultsNames(dds)[2] == sprintf("treatment_%s_vs_%s", row$test_level, row$ref_level))
  keep <- rowSums(counts(dds) >= 10) >= ncol(dds) / 3
  shr <- lfcShrink(dds[keep, ], coef = 2, type = "apeglm", quiet = TRUE)
  out <- data.table(feature = rownames(shr), baseMean = shr$baseMean, log2FoldChange = shr$log2FoldChange,
                    lfcSE = shr$lfcSE, pvalue = shr$pvalue, padj = shr$padj)
  fwrite(out, file.path(de_dir, sprintf("%s_isoseq_apeglm.csv", row$code)))
  out[, code := row$code][]
}
iso <- rbindlist(lapply(seq_len(nrow(contrasts)), function(i) fit_one(contrasts[i])))
iso[, deg := !is.na(padj) & padj < params$padj_cutoff]
```

# Compare with the genome

```{r compare}
genome <- rbindlist(lapply(seq_len(nrow(contrasts)), function(i) {
  r <- contrasts[i]
  g <- fread(file.path(de06, "DEG_lists", r$dir, sprintf("%s_apeglm.csv", r$code)))
  g[, .(code = r$code, feature = gene_key(gene), log2FoldChange, padj)]
}))
genome[, deg := !is.na(padj) & padj < params$padj_cutoff]
stopifnot(!anyDuplicated(genome[, .(code, feature)]))
isoform_genes <- unique(iso_map$gene)

agree <- rbindlist(lapply(contrasts$code, function(cd) {
  g <- genome[code == cd]; i <- iso[code == cd]
  both <- merge(g, i, by = "feature", suffixes = c("_genome", "_isoseq"))
  data.table(code = cd,
             genome_DEGs = sum(g$deg), isoseq_DEGs = sum(i$deg),
             isoseq_DEGs_on_genes = sum(i$deg & !off_gene(i$feature)),
             isoseq_DEGs_novel_loci = sum(i$deg & startsWith(i$feature, "novel:")),
             isoseq_DEGs_unassigned_isoforms = sum(i$deg & startsWith(i$feature, "isoseq:")),
             genes_tested_in_both = nrow(both),
             DEG_in_both = sum(both$deg_genome & both$deg_isoseq),
             same_direction_of_shared = sum(both$deg_genome & both$deg_isoseq &
                                              sign(both$log2FoldChange_genome) == sign(both$log2FoldChange_isoseq)),
             genome_DEG_only = sum(both$deg_genome & !both$deg_isoseq),
             isoseq_DEG_only = sum(!both$deg_genome & both$deg_isoseq),
             genome_DEGs_not_tested_on_isoseq = sum(g$deg & !g$feature %in% i$feature),
             genome_DEGs_without_isoform = sum(g$deg & !g$feature %in% isoform_genes),
             lfc_spearman = round(cor(both$log2FoldChange_genome, both$log2FoldChange_isoseq, method = "spearman"), 3))
}))
fwrite(agree, file.path(de_dir, "reference_agreement.csv"))
knitr::kable(agree)

deg_counts <- merge(iso[, .(features_tested = .N, isoseq_DEGs = sum(deg), isoseq_up = sum(deg & log2FoldChange > 0),
                            isoseq_down = sum(deg & log2FoldChange < 0)), by = code],
                    genome[, .(genome_tested = .N, genome_DEGs = sum(deg)), by = code], by = "code")
fwrite(deg_counts, file.path(de_dir, "isoseq_DEG_counts.csv"))
```

# Why the references differ

Counts per gene: the genome counts are `prepDE` estimates (coverage x transcript length / 75,
the read length the original run assumed, while the trimmed reads average about 63 bases), so a
ratio near 0.84 means equal read numbers; what matters is how the ratio spreads. The Iso-Seq-only
DEGs are compared on their mean count on each reference, which shows whether they are genes the
genome branch counted thinly.

```{r why-differ}
g06 <- fread(file.path(de06, "count_matrix", "gene_count_matrix_clean.csv"))
g06m <- as.matrix(g06[, -1]); rownames(g06m) <- gene_key(g06[[1]])
day3 <- ti[day == "3", sample.1]
both_genes <- setdiff(intersect(rownames(g06m), rownames(counts)), mt_keys)
cpg <- data.table(gene = both_genes,
                  genome_total = rowSums(g06m[both_genes, intersect(day3, colnames(g06m))]),
                  isoseq_total = round(rowSums(counts[both_genes, intersect(day3, colnames(counts))]), 1))
cpg[, ratio := round(isoseq_total / genome_total, 3)]
fwrite(cpg, file.path(de_dir, "counts_per_gene_both_references.csv"))
print(cpg[genome_total > 0 & isoseq_total > 0, .(genes = .N, median_ratio = median(ratio),
                                                 q25 = quantile(ratio, .25), q75 = quantile(ratio, .75),
                                                 more_than_double = sum(ratio > 2 * median(ratio)),
                                                 less_than_half = sum(ratio < median(ratio) / 2))])

io <- merge(iso[deg == TRUE], genome, by = c("code", "feature"), all.x = TRUE, suffixes = c("_isoseq", "_genome"))
io <- io[is.na(deg_genome) | deg_genome == FALSE]
io[, tested_on_genome := !is.na(log2FoldChange_genome)]
gbm <- rbindlist(lapply(seq_len(nrow(contrasts)), function(i) {
  r <- contrasts[i]
  g <- fread(file.path(de06, "DEG_lists", r$dir, sprintf("%s_apeglm.csv", r$code)))
  g[, .(code = r$code, feature = gene_key(gene), baseMean_genome = baseMean)]
}))
io <- merge(io, gbm, by = c("code", "feature"), all.x = TRUE)
io <- io[, .(code, feature, on_annotated_gene = !grepl("^(novel|isoseq|mito):", feature), tested_on_genome,
             baseMean_isoseq = round(baseMean, 1), baseMean_genome = round(baseMean_genome, 1),
             log2FoldChange_isoseq = round(log2FoldChange_isoseq, 3), log2FoldChange_genome = round(log2FoldChange_genome, 3),
             padj_isoseq = signif(padj_isoseq, 3), padj_genome = signif(padj_genome, 3))]
fwrite(io, file.path(de_dir, "isoseq_only_DEGs.csv"))
knitr::kable(io[, .(isoseq_only_DEGs = .N, on_genes = sum(on_annotated_gene),
                    not_tested_on_genome = sum(on_annotated_gene & !tested_on_genome),
                    median_baseMean_isoseq = median(baseMean_isoseq[on_annotated_gene & tested_on_genome]),
                    median_baseMean_genome = median(baseMean_genome[on_annotated_gene & tested_on_genome], na.rm = TRUE),
                    median_padj_genome = median(padj_genome[on_annotated_gene & tested_on_genome], na.rm = TRUE)), by = code])
```

# Figures

```{r fig-lfc, fig.width = 10, fig.height = 6.5}
lab <- setNames(sprintf("%s %s", ifelse(substr(contrasts$code, 1, 1) == "F", "Foot", "Gill"),
                        substr(contrasts$code, 2, 3)), contrasts$code)
pd <- rbindlist(lapply(contrasts$code, function(cd)
  merge(genome[code == cd], iso[code == cd], by = c("code", "feature"), suffixes = c("_genome", "_isoseq"))))
classes <- c("DEG on neither", "DEG on the genome only", "DEG on Iso-Seq only", "DEG on both")
pd[, class := factor(fifelse(deg_genome & deg_isoseq, classes[4], fifelse(deg_genome, classes[2],
                     fifelse(deg_isoseq, classes[3], classes[1]))), levels = classes)]
pd[, panel := factor(lab[code], levels = lab)]
setorder(pd, class)
## neutral greys for set categories (tools/plot_style.R), with shape so the two one-sided
## classes do not rest on grey level alone
p <- ggplot(pd, aes(log2FoldChange_genome, log2FoldChange_isoseq, colour = class, shape = class)) +
  geom_hline(yintercept = 0, colour = "grey80") + geom_vline(xintercept = 0, colour = "grey80") +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = 2) +
  geom_point(size = 0.8, alpha = 0.85, show.legend = TRUE) +
  facet_wrap(~ panel, nrow = 2) +
  scale_colour_manual(values = setNames(c("#d9d9d9", "#7f7f7f", "#7f7f7f", "#1a1a1a"), classes),
                      name = NULL, drop = FALSE) +
  scale_shape_manual(values = setNames(c(16, 17, 15, 16), classes), name = NULL, drop = FALSE) +
  guides(colour = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  labs(x = "log2 fold change, genome reference (06)", y = "log2 fold change, Iso-Seq reference",
       title = "Stressor vs day-3 control on the two references (genes tested on both)",
       caption = "apeglm log2 fold changes; DEG: padj < 0.05. Dashed line: equal fold change.") +
  theme_psmfc(10) + theme(legend.position = "top", plot.caption = element_text(hjust = 0))
ggsave(file.path(de_dir, "FIG_isoseq_vs_genome_lfc.png"), p, width = 10, height = 6.5, dpi = 300)
p
```

```{r fig-mapping, fig.width = 6, fig.height = 4.5}
sal <- fread(file.path(salmon_dir, "salmon_mapping_summary.csv"))
hs  <- fread(hisat_qc)                                          # HISAT2 summaries, as MultiQC tabulated them
hs[, library := sub("^[A-H][0-9]+-", "", sub("_S[0-9]+.*$", "", Sample))]
stopifnot(!anyDuplicated(hs$library))
mr <- merge(sal, hs[, .(library, hisat2 = overall_alignment_rate)], by = "library")
stopifnot(nrow(mr) == nrow(sal))
mr[, tissue := TISSUE_LABELS[ifelse(grepl("G$", library), "G", "F")]]
p2 <- ggplot(mr, aes(hisat2, percent_mapped, colour = tissue)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = 2) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = setNames(TISSUE_COLORS, TISSUE_LABELS), name = NULL) +
  labs(x = "HISAT2 alignment to the genome (%)", y = "salmon mapping to Iso-Seq isoforms (%)",
       title = "Reads placed on each reference, per library",
       caption = sprintf("%d libraries; dashed line: equal rates.", nrow(mr))) +
  theme_psmfc(10) + theme(plot.caption = element_text(hjust = 0))
ggsave(file.path(de_dir, "FIG_mapping_rates.png"), p2, width = 6, height = 4.5, dpi = 300)
p2
```

# Provenance

```{r provenance}
settings <- c("[04_isoseq_de_comparison.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              "model                : as 06: ~ treatment (day-3 control reference), >= 10 counts in a third of samples, apeglm, Wald padj",
              paste0("DEG                  : padj < ", params$padj_cutoff),
              paste0("contrasts            : ", paste(contrasts$code, collapse = ", ")))
inputs <- c(file.path(salmon_dir, c("gene_counts.csv.gz", "salmon_mapping_summary.csv")),
            file.path(map_dir, "isoform_gene_map.csv.gz"),
            file.path(de06, "DEG_lists", c("contrasts.csv", "contrast_samples.csv")),
            file.path(de06, "DEG_lists", contrasts$dir, paste0(contrasts$code, "_apeglm.csv")),
            file.path(de06, "count_matrix", c("treatmentinfo_clean.csv", "mitochondrial_loci.csv", "gene_count_matrix_clean.csv")), hisat_qc,
            here::here("01_code", "04_isoseq_de_comparison.Rmd"))
writeLines(provenance_lines(settings, inputs, repo_root, packages = c("DESeq2", "apeglm", "data.table")),
           file.path(de_dir, "RUN_provenance.txt"))
```
