---
title: "05 - Genome annotation augmented with the Iso-Seq isoforms (option B)"
output: html_document
params:
  threads: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Steps 02 to 04 showed that the RefSeq gene models often end before the 3' ends that Tag-seq
reads (18% of the reads salmon assigned fell on isoforms running more than 100 bases past
their gene's last annotated exon, and 14.5% on loci with no annotation). Option B, run here as
a parallel analysis beside the method of record, recounts the Tag-seq reads on the genome
with the RefSeq annotation augmented by the Iso-Seq isoforms. This step builds the
annotations; step 06 realigns the reads and counts them on each; step 07 repeats `06`'s
contrasts on every count matrix and compares them with the record.

Three annotations, all on GCF_036588685.1 (RefSeq annotation release RS_2024_02):

1. **`refseq`**: the RefSeq GFF unchanged, the annotation the record counted against. Counting
   on it after the realignment here is the control: it shows how far step 06 reproduces the
   record before any annotation changes.
2. **`ext3`** (3' extension only): each RefSeq transcript's last exon is extended to the 3' end
   of the furthest Iso-Seq isoform whose last aligned block overlaps that exon, on the same
   strand and assigned to the same gene, when that end lies beyond the exon. The extension
   stops before any other gene on the same strand. Nothing else changes: same genes, same
   keys, same names and GO annotation. An isoform whose last block lies downstream of the
   transcript's last exon (an extra exon, or an alignment across a long intron) extends
   nothing, so `ext3` only lengthens 3' untranslated regions in place.
3. **`full`**: the RefSeq annotation plus every such isoform as an extra transcript of its gene
   (its aligned blocks as exons), plus the novel loci of step 02 as new genes
   (`novel:<sequence>:<start>-<end>:<strand>`, each isoform a transcript). A novel locus that
   overlaps the exons of exactly one gene on the same strand, counting that gene's added
   isoforms, joins that gene; one overlapping several genes or a mitochondrial locus is left
   out (below).

Which isoforms are used: those step 02 `assigned` to a gene with no exon overlap with a second
gene (readthrough isoforms, which would merge two genes, are left out, as are `ambiguous`
placements), and the isoforms of novel loci. Mitochondrial isoforms are never added, and a
novel locus holding an isoform with a mitochondrial segment (`mt_hit`) is left out, as step 04
leaves those features out; the 310 mitochondrial loci stay in every annotation exactly as
RefSeq has them, and step 07 leaves them out of the fits as `06` does.

Writes, in `03_analyses/05_augmented-annotation/`:

- `ext3_extensions.csv.gz`: every extended transcript: `transcript_id`, `gene` (gene key),
  `seqid`, `strand`, the old and new 3' end, `extension_bp`, `isoforms` (how many isoforms
  overlap the last exon), `capped` (stopped before another gene);
- `full_added_transcripts.bed.gz`: the transcripts `full` adds, as BED12 (name =
  `<gene>|<isoform>`), so `full` can be rebuilt without step 02's alignment;
- `novel_loci_fate.csv`: how many of step 02's novel loci joined a gene, were left out, or
  stay novel loci in `full`;
- `annotation_summary.csv`: per annotation, genes, transcripts, exonic bases (exons of each
  gene merged) and genes changed;
- `RUN_provenance.txt`: rules, input MD5s.

and, git-ignored (rebuilt here in a few minutes): `annotation_<refseq|ext3|full>.gff`
(StringTie's input) and `annotation_<refseq|ext3|full>.saf` (featureCounts's input: the
exons of each gene, keyed by `gene_key()`).

Needs step 02's genome alignment of the isoforms (`02_isoform-gene-map/minimap2_isoforms_vs_genome.paf.gz`,
git-ignored) and the RefSeq GFF in `02_data/`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))         # gene_key()
source(file.path(repo_root, "tools", "pipeline_checks.R"))  # provenance_lines()
suppressPackageStartupMessages({
  library(data.table); library(GenomicRanges); library(GenomicAlignments)
})
paf <- file.path(map_dir, "minimap2_isoforms_vs_genome.paf.gz")
map_file <- file.path(map_dir, "isoform_gene_map.csv.gz")
stopifnot(file.exists(gff_gz), file.exists(paf), file.exists(map_file))
```

# The RefSeq GFF, line by line

The GFF is read as text, so the augmented files differ from RefSeq's only where this step
changes them.

```{r gff}
gff_lines <- readLines(gzfile(gff_gz))
is_com <- startsWith(gff_lines, "#")
body <- fread(text = gff_lines[!is_com], sep = "\t", header = FALSE, quote = "",
              colClasses = c("character", "character", "character", "integer", "integer",
                             "character", "character", "character", "character"))
stopifnot(nrow(body) == sum(!is_com), ncol(body) == 9)
setnames(body, c("seqid", "source", "type", "start", "end", "score", "strand", "phase", "attr"))
body[, line := which(!is_com)]
attr_of <- function(key, x) {                                   # value of one GFF3 attribute, or NA
  has <- grepl(paste0("(^|;)", key, "="), x, perl = TRUE)
  out <- rep(NA_character_, length(x))
  out[has] <- sub(paste0("^(?:.*;)?", key, "=([^;]*).*$"), "\\1", x[has], perl = TRUE)
  out
}
body[, ID := attr_of("ID", attr)]
body[, Parent := sub(",.*$", "", attr_of("Parent", attr))]

## each feature's top-level gene, following Parent upwards (as step 02)
ids <- body[!is.na(ID) & !duplicated(ID) & !ID %in% ID[duplicated(ID)]]
up <- setNames(ids$Parent, ids$ID)
ex <- body[type == "exon"]
ex[, top := Parent]
for (i in 1:5) { nxt <- up[ex$top]; move <- !is.na(nxt); if (!any(move)) break; ex[move, top := nxt[move]] }
stopifnot(!anyNA(ex$top))
ex[, gene := gene_key(top)]
top_of <- unique(ex[, .(gene, top)])
stopifnot(!anyDuplicated(top_of$gene))                       # one GFF gene per key
cat(nrow(ex), "exons in", uniqueN(ex$Parent), "transcripts of", uniqueN(ex$gene), "genes\n")
```

# The isoforms and their aligned blocks

```{r isoforms}
map <- fread(map_file)
use <- map[(status == "assigned" & is.na(runner_up_overlap)) | status == "novel_locus"]
mt_loci <- unique(map[status == "novel_locus" & mt_hit == TRUE, gene])
use <- use[!(status == "novel_locus" & gene %in% mt_loci)]
stopifnot(all(use[status == "assigned", gene] %in% top_of$gene))

## the alignment of each isoform (the best one, as step 02 kept)
awk_prog <- tempfile(fileext = ".awk")
writeLines('BEGIN{FS=OFS="\\t"} {cg=""; for (i=13; i<=NF; i++) if ($i ~ /^cg:Z:/) cg=substr($i,6); print $1,$5,$6,$8,$9,$10,cg}', awk_prog)
aln <- fread(cmd = sprintf("zcat %s | awk -f %s", shQuote(paf), shQuote(awk_prog)), sep = "\t", header = FALSE,
             col.names = c("isoform", "strand", "seqid", "tstart", "tend", "matches", "cigar"))
setorder(aln, isoform, -matches)
aln <- aln[!duplicated(isoform)][isoform %in% use$isoform]
aln <- merge(aln, use[, .(isoform, gene, status, m_seqid = seqid, m_strand = strand, m_start = start, m_end = end)], by = "isoform")
## the same alignment step 02 used
stopifnot(nrow(aln) == nrow(use), all(aln$seqid == aln$m_seqid), all(aln$strand == aln$m_strand),
          all(aln$tstart + 1L == aln$m_start), all(aln$tend == aln$m_end))
blocks <- extractAlignmentRangesOnReference(aln$cigar, pos = aln$tstart + 1L)
blk <- data.table(isoform = rep(aln$isoform, lengths(blocks)), gene = rep(aln$gene, lengths(blocks)),
                  status = rep(aln$status, lengths(blocks)), seqid = rep(aln$seqid, lengths(blocks)),
                  strand = rep(aln$strand, lengths(blocks)), start = start(unlist(blocks)), end = end(unlist(blocks)))
## an insertion between two introns leaves an empty range: drop those
blk <- blk[end >= start]
## blocks are split at introns (N) only; merge any that touch, per isoform
rb <- reduce(GRanges(blk$isoform, IRanges(blk$start, blk$end)), min.gapwidth = 1L)
if (length(rb) < nrow(blk)) {
  info <- unique(blk[, .(isoform, gene, status, seqid, strand)])
  blk <- merge(data.table(isoform = as.character(seqnames(rb)), start = start(rb), end = end(rb)), info, by = "isoform")
  setcolorder(blk, c("isoform", "gene", "status", "seqid", "strand", "start", "end"))
}
cat(uniqueN(blk[status == "assigned", isoform]), "assigned isoforms on", uniqueN(blk[status == "assigned", gene]), "genes;",
    uniqueN(blk[status == "novel_locus", isoform]), "isoforms in", uniqueN(blk[status == "novel_locus", gene]), "novel loci\n")
```

# Novel loci against the augmented genes

Step 02 called a locus novel when its isoforms share fewer than 50 exonic bases with the
RefSeq exons. Many such loci lie where a gene's own isoforms add exons (an extended 3' end,
for instance), so in `full` they would overlap that gene and their reads would be split
between the two or, for featureCounts, dropped as ambiguous. A novel locus whose exons overlap
the exons of exactly one gene on the same strand (its RefSeq exons or its added isoforms) joins
that gene; one that overlaps more than one gene, or a mitochondrial locus, is left out.

```{r loci}
mt_genes <- unique(fread(file.path(de06, "count_matrix", "mitochondrial_loci.csv"))$LOC_ID)
stopifnot(all(mt_genes %in% ex$gene))
gene_ex <- rbind(ex[, .(gene, seqid, strand, start, end)], blk[status == "assigned", .(gene, seqid, strand, start, end)])
nb <- blk[status == "novel_locus"]
ho <- findOverlaps(GRanges(nb$seqid, IRanges(nb$start, nb$end), nb$strand),
                   GRanges(gene_ex$seqid, IRanges(gene_ex$start, gene_ex$end), gene_ex$strand))
lg <- unique(data.table(locus = nb$gene[queryHits(ho)], gene = gene_ex$gene[subjectHits(ho)]))
lg_n <- lg[, .(genes = .N, mt = any(gene %in% mt_genes)), by = locus]
join <- lg[locus %in% lg_n[genes == 1 & !mt, locus]]
drop <- lg_n[genes > 1 | mt, locus]
loci_fate <- data.table(fate = c("joined one gene", "left out (several genes or mitochondrial)", "kept as novel loci"),
                        loci = c(nrow(join), length(drop), uniqueN(nb$gene) - nrow(join) - length(drop)),
                        isoforms = c(nb[gene %in% join$locus, uniqueN(isoform)], nb[gene %in% drop, uniqueN(isoform)],
                                     nb[!gene %in% c(join$locus, drop), uniqueN(isoform)]))
blk[status == "novel_locus" & gene %in% join$locus, `:=`(gene = join$gene[match(gene, join$locus)], status = "joined")]
blk <- blk[!(status == "novel_locus" & gene %in% drop)]
fwrite(loci_fate, file.path(ann_dir, "novel_loci_fate.csv"))
knitr::kable(loci_fate)
```

# `ext3`: last exons extended to the isoforms' 3' ends

```{r ext3}
## each transcript's last exon (by strand)
ex[, is_last := fifelse(strand == "+", end == max(end), start == min(start)), by = Parent]
te <- ex[is_last == TRUE]
te <- te[!duplicated(Parent)]                                 # one last exon per transcript
## each assigned isoform's last block
tb <- blk[status == "assigned"][order(isoform, fifelse(strand == "+", -end, start))][!duplicated(isoform)]
hits <- findOverlaps(GRanges(te$seqid, IRanges(te$start, te$end), te$strand),
                     GRanges(tb$seqid, IRanges(tb$start, tb$end), tb$strand))
h <- data.table(r = queryHits(hits), b = subjectHits(hits))
h <- h[te$gene[r] == tb$gene[b]]
h[, `:=`(strand = te$strand[r], iso3 = fifelse(te$strand[r] == "+", tb$end[b], tb$start[b]))]
ext <- h[, .(new3 = if (strand[1] == "+") max(iso3) else min(iso3), isoforms = .N), by = r]
ext[, `:=`(transcript_id = te$Parent[r], gene = te$gene[r], seqid = te$seqid[r], strand = te$strand[r],
           old3 = fifelse(te$strand[r] == "+", te$end[r], te$start[r]), line = te$line[r])]
ext <- ext[fifelse(strand == "+", new3 > old3, new3 < old3)]

## stop before any other gene on the same strand (gene extents from their exons)
span <- ex[, .(start = min(start), end = max(end)), by = .(gene, seqid, strand)]
reg <- GRanges(ext$seqid, IRanges(pmin(ext$old3, ext$new3) + (ext$strand == "+"), pmax(ext$old3, ext$new3) - (ext$strand == "-")), ext$strand)
oh <- findOverlaps(reg, GRanges(span$seqid, IRanges(span$start, span$end), span$strand))
o <- data.table(r = queryHits(oh), s = subjectHits(oh))[ext$gene[r] != span$gene[s]]
o[, cap := fifelse(ext$strand[r] == "+", span$start[s] - 1L, span$end[s] + 1L)]
cap <- o[, .(cap = if (ext$strand[r[1]] == "+") min(cap) else max(cap)), by = r]
ext[, capped := FALSE]
ext[cap$r, `:=`(new3 = fifelse(strand == "+", pmin(new3, cap$cap), pmax(new3, cap$cap)), capped = TRUE)]
ext <- ext[fifelse(strand == "+", new3 > old3, new3 < old3)]   # a gene right after the exon leaves no room
ext[, extension_bp := abs(new3 - old3)]
cat(nrow(ext), "transcripts of", uniqueN(ext$gene), "genes extended; median", median(ext$extension_bp),
    "bases (interquartile", paste(quantile(ext$extension_bp, c(.25, .75)), collapse = " to "), "); capped:", sum(ext$capped), "\n")
fwrite(ext[, .(transcript_id, gene, seqid, strand, old3, new3, extension_bp, isoforms, capped)],
       file.path(ann_dir, "ext3_extensions.csv.gz"))
```

# Writing the annotations

```{r write}
## GFF: change the extended exons and widen their transcript and gene rows to cover them
dup_id <- body[!is.na(ID) & duplicated(ID), unique(ID)]
widen <- function(b, rows) {                                   # rows: data.table(ID, lo, hi)
  w <- rows[, .(lo = min(lo), hi = max(hi)), by = ID]
  i <- match(w$ID, b$ID); stopifnot(!anyNA(i), !any(w$ID %in% dup_id))
  b[i, `:=`(start = pmin(start, w$lo), end = pmax(end, w$hi))]
}
b_ext <- copy(body)
i <- match(ext$line, b_ext$line)
b_ext[i[ext$strand == "+"], end := ext$new3[ext$strand == "+"]]
b_ext[i[ext$strand == "-"], start := ext$new3[ext$strand == "-"]]
ext_rng <- ext[, .(lo = pmin(old3, new3), hi = pmax(old3, new3), transcript_id, gene)]
widen(b_ext, ext_rng[, .(ID = transcript_id, lo, hi)])
widen(b_ext, ext_rng[, .(ID = top_of$top[match(gene, top_of$gene)], lo, hi)])

## full: the isoforms as transcripts of their gene, novel loci as genes
iso_id <- function(x) paste0("iso-", sub("^MT_Pool_HQ_transcript/", "", x))
tx <- blk[, .(start = min(start), end = max(end)), by = .(isoform, gene, status, seqid, strand)]
tx[, parent := fifelse(status %in% c("assigned", "joined"), top_of$top[match(gene, top_of$gene)], gene)]
loci <- tx[status == "novel_locus", .(start = min(start), end = max(end)), by = .(gene, seqid, strand)]
new_rows <- rbind(
  loci[, .(seqid, source = "IsoSeq", type = "gene", start, end, score = ".", strand, phase = ".",
           attr = paste0("ID=", gene, ";gene_biotype=isoseq_novel_locus"))],
  tx[, .(seqid, source = "IsoSeq", type = "transcript", start, end, score = ".", strand, phase = ".",
         attr = paste0("ID=", iso_id(isoform), ";Parent=", parent))],
  blk[, .(seqid, source = "IsoSeq", type = "exon", start, end, score = ".", strand, phase = ".",
          attr = paste0("ID=exon-", iso_id(isoform), "-", seq_len(.N), ";Parent=", iso_id(isoform))), by = isoform][, isoform := NULL])
b_full <- copy(body)
widen(b_full, tx[status %in% c("assigned", "joined"), .(ID = parent, lo = start, hi = end)])

write_gff <- function(b, extra, path) {
  out <- gff_lines
  out[b$line] <- do.call(paste, c(b[, .(seqid, source, type, start, end, score, strand, phase, attr)], sep = "\t"))
  if (!is.null(extra)) out <- c(out, do.call(paste, c(extra, sep = "\t")))
  writeLines(out, path)
}
write_gff(body, NULL, file.path(ann_dir, "annotation_refseq.gff"))
write_gff(b_ext, NULL, file.path(ann_dir, "annotation_ext3.gff"))
write_gff(b_full, new_rows, file.path(ann_dir, "annotation_full.gff"))
## the refseq file is RefSeq's GFF, byte for byte
stopifnot(identical(unname(tools::md5sum(file.path(ann_dir, "annotation_refseq.gff"))),
                    system(paste("gzip -dc", shQuote(gff_gz), "| md5sum | cut -c1-32"), intern = TRUE)))

## SAF for featureCounts: the exons of each gene, by gene key
saf_of <- function(e) e[, .(GeneID = gene, Chr = seqid, Start = start, End = end, Strand = strand)]
ex_ext <- copy(ex); ex_ext[match(ext$line, line), `:=`(start = b_ext$start[match(ext$line, b_ext$line)],
                                                        end = b_ext$end[match(ext$line, b_ext$line)])]
saf <- list(refseq = saf_of(ex), ext3 = saf_of(ex_ext),
            full = rbind(saf_of(ex), blk[, .(GeneID = gene, Chr = seqid, Start = start, End = end, Strand = strand)]))
for (a in names(saf)) fwrite(unique(saf[[a]]), file.path(ann_dir, paste0("annotation_", a, ".saf")), sep = "\t")

## BED12 of the transcripts full adds, so full can be rebuilt without step 02's alignment
bed <- blk[order(isoform, start), .(seqid = seqid[1], start0 = min(start) - 1L, end = max(end), strand = strand[1], gene = gene[1],
                                    n = .N, sizes = paste0(paste(end - start + 1L, collapse = ","), ","),
                                    starts = paste0(paste(start - min(start), collapse = ","), ",")), by = isoform]
fwrite(bed[, .(seqid, start0, end, name = paste0(gene, "|", isoform), score = 0, strand, start0, end, rgb = 0, n, sizes, starts)],
       file.path(ann_dir, "full_added_transcripts.bed.gz"), sep = "\t", col.names = FALSE)
```

# Summary

```{r summary}
exonic_bp <- function(s) sum(width(reduce(GRanges(s$Chr, IRanges(s$Start, s$End), s$Strand) |> split(s$GeneID)) |> unlist()))
summ <- data.table(annotation = names(saf),
                   genes = vapply(saf, function(s) uniqueN(s$GeneID), 1L),
                   transcripts = c(uniqueN(ex$Parent), uniqueN(ex$Parent), uniqueN(ex$Parent) + nrow(tx)),
                   exonic_bases = vapply(saf, exonic_bp, 1),
                   genes_changed = c(0L, uniqueN(ext$gene), uniqueN(tx$gene)))
summ[, exonic_bases_vs_refseq := round(exonic_bases / exonic_bases[1], 3)]
fwrite(summ, file.path(ann_dir, "annotation_summary.csv"))
knitr::kable(summ)
```

```{r provenance}
settings <- c("[05_augmented_annotation.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              paste0("genome / annotation  : ", ASSEMBLY, " (RefSeq annotation RS_2024_02)"),
              "isoforms used        : step 02 'assigned' with no exon overlap with a second gene; novel loci without an mt_hit isoform",
              "ext3                 : last exon of each transcript extended to the furthest 3' end of a same-gene isoform whose last block overlaps it; stops before any other gene on the same strand",
              "full                 : RefSeq plus the isoforms as transcripts of their gene, plus novel loci as genes; a locus overlapping one gene's exons (RefSeq or added) on its strand joins it, one overlapping several genes or a mitochondrial locus is left out",
              paste0("novel loci           : ", paste(paste(loci_fate$fate, loci_fate$loci), collapse = "; ")),
              paste0("transcripts extended : ", nrow(ext), " (", uniqueN(ext$gene), " genes)"),
              paste0("transcripts added    : ", nrow(tx), " (", nrow(loci), " novel loci)"))
writeLines(provenance_lines(settings, c(gff_gz, map_file, paf, here::here("01_code", "05_augmented_annotation.Rmd")),
                            repo_root, packages = c("data.table", "GenomicRanges", "GenomicAlignments")),
           file.path(ann_dir, "RUN_provenance.txt"))
```
