---
title: "07 - Differential expression on the recounted genome, compared with the record (option B)"
output: html_document
params:
  padj_cutoff: 0.05
  cores: 4                 # parallel fits (forked processes; one on Windows)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Fits the six treatment-control (TC) contrasts of record to each of step 06's six count
matrices (two counters x three annotations), exactly as `06_differential-expression` fits them
to the record's counts, and compares the results three ways:

1. **Control against the record.** StringTie + prepDE on the unchanged RefSeq annotation
   differs from the record only in the alignment (step 06). How close its DEGs come to the
   record's is the yardstick for every other difference below.
2. **Annotation effect.** Within each counter, `ext3` and `full` against that counter's RefSeq
   control: the same alignment and counter, so the differences come from the annotation alone.
3. **Each matrix against the record**, which is what adopting it would change.

Same samples, model, order and thresholds as `06` (and step 04): the samples of each contrast
in `06 .../DEG_lists/contrast_samples.csv`, `~ treatment` with the day-3 control as reference,
fitted on every gene, then genes kept with at least 10 counts in a third of the samples, apeglm
shrinkage (Zhu, Ibrahim and Love 2019, *Bioinformatics* 35:2084), DEG at padj < 0.05. The 310
mitochondrial loci (`06 .../count_matrix/mitochondrial_loci.csv`) are left out of every fit;
step 05 kept them unchanged in every annotation. Genes are matched by `gene_key()`; the novel
loci of `full` (`novel:...`) exist only there.

Writes, in `03_analyses/07_augmented-de/`:

- `<counter>_<annotation>_apeglm.csv.gz`: every tested gene of every contrast (`code`,
  `feature`, `baseMean`, apeglm `log2FoldChange`, `lfcSE`, Wald `pvalue`, `padj`);
- `deg_summary.csv`: per matrix and contrast, genes tested, DEGs (up, down), and the agreement
  with the record: DEGs shared, in the same direction, record DEGs lost (and of those not
  tested here), new DEGs (and of those on novel loci), and the Spearman correlation of fold
  changes over genes tested in both;
- `annotation_effect.csv`: per counter and contrast, `ext3` and `full` against that counter's
  RefSeq control (DEGs shared, gained, lost, fold-change correlation);
- `counts_per_gene.csv`: per gene, the total count over the day-3 libraries on each matrix and
  in the record;
- `control_vs_record.csv`: the control against the record: per-library totals, the
  mitochondrial share of each library, gene counts;
- `new_DEGs.csv`: DEGs of `ext3` and `full` (StringTie + prepDE) that are not DEGs in the
  record, with the record's padj and both counts, and whether `ext3` extended the gene;
- `byssal_genes.csv`: the byssal structural genes (best BLAST hit named as in `09`) that are a
  DEG in any matrix or in the record;
- `FIG_deg_counts.png`, `FIG_lfc_vs_record.png`;
- `RUN_provenance.txt`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
source(file.path(repo_root, "tools", "pipeline_checks.R"))
suppressPackageStartupMessages({ library(DESeq2); library(apeglm); library(data.table); library(ggplot2); library(parallel) })
contrasts <- fread(file.path(de06, "DEG_lists", "contrasts.csv"))[family == "TC"]
cs <- fread(file.path(de06, "DEG_lists", "contrast_samples.csv"))
ti <- fread(file.path(de06, "count_matrix", "treatmentinfo_clean.csv"))
mt_keys <- unique(fread(file.path(de06, "count_matrix", "mitochondrial_loci.csv"))$LOC_ID)
stopifnot(nrow(contrasts) == 6)
counters <- c(stringtie = "StringTie + prepDE", featurecounts = "featureCounts")
anns <- c(refseq = "RefSeq (control)", ext3 = "3' extension", full = "full models")
mats <- list(); mt_share <- list()
for (cn in names(counters)) for (a in names(anns)) {
  d <- fread(file.path(recount_dir, sprintf("%s_%s_gene_counts.csv.gz", cn, a)), sep = ",")
  m <- as.matrix(d[, -1]); rownames(m) <- gene_key(d$gene)
  stopifnot(!anyDuplicated(rownames(m)), all(cs[code %in% contrasts$code, sample.1] %in% colnames(m)))
  mt_share[[paste(cn, a)]] <- colSums(m[rownames(m) %in% mt_keys, ]) / colSums(m)
  mats[[paste(cn, a)]] <- m[!rownames(m) %in% mt_keys, ]
}
sapply(mats, nrow)
```

# Fit each TC contrast on each matrix

```{r fit}
## as 04 step 04 and 06 steps 03-04: fit every gene of the contrast's samples, keep genes with
## >= 10 counts in a third of them, apeglm on those (pvalue and padj from results())
fit_one <- function(counts, row) {
  smp <- cs[code == row$code, sample.1]
  cd  <- as.data.frame(ti[match(smp, sample.1), .(sample.1, treatment)]); rownames(cd) <- cd$sample.1
  cd$treatment <- factor(cd$treatment, levels = c(row$ref_level, row$test_level))
  stopifnot(!anyNA(cd$treatment))
  m <- round(counts[, smp]); m <- m[rowSums(m) > 0, ]
  dds <- DESeq(DESeqDataSetFromMatrix(m, cd, ~ treatment), quiet = TRUE)
  stopifnot(resultsNames(dds)[2] == sprintf("treatment_%s_vs_%s", row$test_level, row$ref_level))
  keep <- rowSums(counts(dds) >= 10) >= ncol(dds) / 3
  shr <- lfcShrink(dds[keep, ], coef = 2, type = "apeglm", quiet = TRUE)
  data.table(code = row$code, feature = rownames(shr), baseMean = shr$baseMean, log2FoldChange = shr$log2FoldChange,
             lfcSE = shr$lfcSE, pvalue = shr$pvalue, padj = shr$padj)
}
jobs <- CJ(matrix = names(mats), i = seq_len(nrow(contrasts)), sorted = FALSE)
cores <- if (.Platform$OS.type == "windows") 1L else params$cores
fits <- mclapply(seq_len(nrow(jobs)), function(j) fit_one(mats[[jobs$matrix[j]]], contrasts[jobs$i[j]]),
                 mc.cores = cores, mc.preschedule = FALSE)
stopifnot(!vapply(fits, inherits, NA, "try-error"))
res <- rbindlist(Map(function(f, mx) f[, matrix := mx], fits, jobs$matrix))
res[, deg := !is.na(padj) & padj < params$padj_cutoff]
for (mx in names(mats)) {
  out <- res[matrix == mx, .(code, feature, baseMean, log2FoldChange, lfcSE, pvalue, padj)]
  fwrite(out, file.path(aug_de_dir, sprintf("%s_apeglm.csv.gz", sub(" ", "_", mx))))
}
```

# Against the record

```{r record}
record <- rbindlist(lapply(seq_len(nrow(contrasts)), function(i) {
  r <- contrasts[i]
  g <- fread(file.path(de06, "DEG_lists", r$dir, sprintf("%s_apeglm.csv", r$code)), sep = ",")
  g[, .(code = r$code, feature = gene_key(gene), baseMean, log2FoldChange, padj)]
}))
record[, deg := !is.na(padj) & padj < params$padj_cutoff]
stopifnot(!anyDuplicated(record[, .(code, feature)]))

compare <- function(x, y) {                                     # x: a result set, y: the reference set
  rbindlist(lapply(contrasts$code, function(cd) {
    a <- x[code == cd]; b <- y[code == cd]
    both <- merge(a, b, by = "feature", suffixes = c("", "_ref"))
    data.table(code = cd, tested = nrow(a), DEGs = sum(a$deg), up = sum(a$deg & a$log2FoldChange > 0),
               down = sum(a$deg & a$log2FoldChange < 0), ref_DEGs = sum(b$deg),
               shared = sum(both$deg & both$deg_ref),
               same_direction = sum(both$deg & both$deg_ref & sign(both$log2FoldChange) == sign(both$log2FoldChange_ref)),
               ref_DEGs_lost = sum(b$deg) - sum(both$deg & both$deg_ref),
               ref_DEGs_not_tested = sum(b$deg & !b$feature %in% a$feature),
               new_DEGs = sum(a$deg) - sum(both$deg & both$deg_ref),
               new_DEGs_novel_loci = sum(a$deg & startsWith(a$feature, "novel:")),
               lfc_spearman = round(cor(both$log2FoldChange, both$log2FoldChange_ref, method = "spearman"), 3))
  }))
}
summ <- rbindlist(lapply(names(mats), function(mx) {
  s <- compare(res[matrix == mx], record)
  s[, `:=`(counter = sub(" .*$", "", mx), annotation = sub("^.* ", "", mx))]
}))
setcolorder(summ, c("counter", "annotation"))
setnames(summ, c("ref_DEGs", "ref_DEGs_lost", "ref_DEGs_not_tested"), c("record_DEGs", "record_DEGs_lost", "record_DEGs_not_tested"))
fwrite(summ, file.path(aug_de_dir, "deg_summary.csv"))
knitr::kable(summ)
```

# Annotation effect within each counter

```{r annotation-effect}
eff <- rbindlist(lapply(names(counters), function(cn) rbindlist(lapply(c("ext3", "full"), function(a) {
  s <- compare(res[matrix == paste(cn, a)], res[matrix == paste(cn, "refseq")])
  s[, .(counter = cn, annotation = a, code, DEGs, control_DEGs = ref_DEGs, shared, gained = new_DEGs,
        gained_novel_loci = new_DEGs_novel_loci, lost = ref_DEGs_lost, lfc_spearman)]
}))))
fwrite(eff, file.path(aug_de_dir, "annotation_effect.csv"))
knitr::kable(eff)
```

# The control against the record

```{r control}
g06 <- fread(file.path(de06, "count_matrix", "gene_count_matrix_clean.csv"), sep = ",")
rec_m <- as.matrix(g06[, -1]); rownames(rec_m) <- gene_key(g06[[1]])
ctrl <- fread(file.path(recount_dir, "stringtie_refseq_gene_counts.csv.gz"), sep = ",")
ctrl_m <- as.matrix(ctrl[, -1]); rownames(ctrl_m) <- gene_key(ctrl$gene)
libs <- intersect(colnames(rec_m), colnames(ctrl_m))
genes <- intersect(rownames(rec_m), rownames(ctrl_m))
rec_share <- colSums(rec_m[rownames(rec_m) %in% mt_keys, libs]) / colSums(rec_m[, libs])
day3 <- intersect(ti[day == "3", sample.1], libs)
gt <- data.table(gene = genes, record = rowSums(rec_m[genes, day3]), control = rowSums(ctrl_m[genes, day3]))
g10 <- gt[(record >= 10 | control >= 10) & !gene %in% mt_keys]
cvr <- data.table(
  measure = c("libraries compared", "genes in both matrices", "genes only in the record (StringTie IDs)",
              "per-library total count, control / record: median", "range",
              "per-library mitochondrial share, Spearman control vs record",
              "nuclear genes with >= 10 counts over the day-3 libraries in either",
              "of those, day-3 total within 10% of the record", "Spearman of day-3 totals (those genes)"),
  value = c(length(libs), length(genes), sum(!rownames(rec_m) %in% genes),
            round(median(colSums(ctrl_m[, libs]) / colSums(rec_m[, libs])), 4),
            paste(round(range(colSums(ctrl_m[, libs]) / colSums(rec_m[, libs])), 4), collapse = " to "),
            round(cor(mt_share[["stringtie refseq"]][libs], rec_share, method = "spearman"), 3),
            nrow(g10), round(mean(abs(g10$control - g10$record) <= 0.1 * g10$record), 3),
            round(cor(g10$control, g10$record, method = "spearman"), 4)))
fwrite(cvr, file.path(aug_de_dir, "control_vs_record.csv"))
knitr::kable(cvr)

## counts per gene on every matrix (day-3 libraries), with the record
cpg <- Reduce(function(x, y) merge(x, y, by = "gene", all = TRUE),
              c(list(data.table(gene = rownames(rec_m), record = rowSums(rec_m[, day3]))),
                lapply(names(mats), function(mx) { m <- mats[[mx]]
                  setnames(data.table(gene = rownames(m), rowSums(m[, day3])), 2, sub(" ", "_", mx)) })))
cpg <- cpg[!gene %in% mt_keys]
fwrite(cpg, file.path(aug_de_dir, "counts_per_gene.csv"))
knitr::kable(cpg[, lapply(.SD, function(v) sum(v, na.rm = TRUE)), .SDcols = -"gene"], caption = "total count, day-3 libraries")
```

# What the new DEGs are

```{r new-degs}
ext <- fread(file.path(ann_dir, "ext3_extensions.csv.gz"))
nd <- rbindlist(lapply(c("ext3", "full"), function(a) {
  x <- merge(res[matrix == paste("stringtie", a) & deg == TRUE], record, by = c("code", "feature"), all.x = TRUE,
             suffixes = c("", "_record"))
  x[is.na(deg_record) | deg_record == FALSE, .(annotation = a, code, feature, novel_locus = startsWith(feature, "novel:"),
                                                extended_in_ext3 = feature %in% ext$gene,
                                                tested_in_record = !is.na(padj_record) | !is.na(log2FoldChange_record),
                                                baseMean = round(baseMean, 1), baseMean_record = round(baseMean_record, 1),
                                                log2FoldChange = round(log2FoldChange, 3), log2FoldChange_record = round(log2FoldChange_record, 3),
                                                padj = signif(padj, 3), padj_record = signif(padj_record, 3))]
}))
fwrite(nd, file.path(aug_de_dir, "new_DEGs.csv"))
knitr::kable(nd[, .(new_DEGs = .N, novel_loci = sum(novel_locus), extended_genes = sum(extended_in_ext3 & !novel_locus),
                    not_tested_in_record = sum(!tested_in_record & !novel_locus),
                    median_padj_record = median(padj_record, na.rm = TRUE),
                    record_padj_below_0.2 = round(mean(padj_record < 0.2, na.rm = TRUE), 2)), by = .(annotation, code)])
```

# Byssal genes

The byssal structural genes are named as in `09` (`BYSSAL_STRUCTURAL_REGEX`): the best BLAST hit
(highest bitscore) of the genome-wide table names a foot protein, plaque, precollagen, ACDC or
byssal tyrosinase.

```{r byssal}
byssal_rx <- paste("foot protein", "adhesive plaque", "polyphenolic", "byssal protein-", "preCol", "precollagen",
                   "byssal EP", "ACDC", "byssal tyrosinase", sep = "|")
bl <- fread(file.path(repo_root, "03_blast", "03_analyses", "genome-foot", "LOC_GO_list.txt"), sep = "\t",
            select = c("gene", "bitscore", "Protein.names"))
bl[, `:=`(gene = gene_key(trimws(gene)), bitscore = suppressWarnings(as.numeric(bitscore)))]
bl <- bl[!is.na(gene) & nzchar(gene)][order(-bitscore)][!duplicated(gene)]
byssal <- bl[grepl(byssal_rx, Protein.names, ignore.case = TRUE), .(feature = gene, protein = Protein.names)]
bd <- rbind(record[feature %in% byssal$feature, .(source = "record", code, feature, log2FoldChange, padj, deg)],
            res[feature %in% byssal$feature, .(source = matrix, code, feature, log2FoldChange, padj, deg)])
any_deg <- unique(bd[deg == TRUE, .(code, feature)])
bw <- dcast(merge(bd, any_deg, by = c("code", "feature")), code + feature ~ source, value.var = "deg", fill = NA)
bw <- merge(bw, byssal, by = "feature")
fwrite(bw, file.path(aug_de_dir, "byssal_genes.csv"))
knitr::kable(bw[, lapply(.SD, function(v) sum(v, na.rm = TRUE)), by = code, .SDcols = setdiff(names(bw), c("code", "feature", "protein"))],
             caption = "byssal structural DEGs per contrast and source")
```

# Figures

```{r fig-degs, fig.width = 10, fig.height = 5}
lab <- setNames(sprintf("%s %s", ifelse(substr(contrasts$code, 1, 1) == "F", "Foot", "Gill"), substr(contrasts$code, 2, 3)),
                contrasts$code)
levels_src <- c("record (06)", unname(anns))
pd <- rbind(summ[, .(counter = counters[counter], source = anns[annotation], code, DEGs)],
            rbindlist(lapply(counters, function(cl) record[, .(counter = cl, source = "record (06)", DEGs = sum(deg)), by = code])),
            use.names = TRUE)
pd[, `:=`(source = factor(source, levels = levels_src), contrast = factor(lab[code], levels = lab),
          counter = factor(counter, levels = counters))]
## neutral greys for set categories (tools/plot_style.R), labelled
p <- ggplot(pd, aes(contrast, DEGs, fill = source)) +
  geom_col(position = position_dodge(width = 0.82), width = 0.78, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = DEGs), position = position_dodge(width = 0.82), vjust = -0.3, size = 2.4) +
  facet_wrap(~ counter, nrow = 1) +
  scale_fill_manual(values = setNames(c("#1a1a1a", "#7f7f7f", "#a6a6a6", "#d0d0d0"), levels_src), name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = NULL, y = "DEGs (apeglm padj < 0.05)", title = "DEGs per contrast: the record and the three recounted annotations",
       caption = "Same samples, model and thresholds as 06; mitochondrial loci left out. Record: 06's DEGs (prepDE on the original alignment).") +
  theme_psmfc(10) + theme(legend.position = "top", plot.caption = element_text(hjust = 0),
                          axis.text.x = element_text(angle = 30, hjust = 1))
ggsave(file.path(aug_de_dir, "FIG_deg_counts.png"), p, width = 10, height = 5, dpi = 300)
p
```

```{r fig-lfc, fig.width = 12, fig.height = 6.5}
classes <- c("DEG in neither", "DEG in the record only", "DEG in the recount only", "DEG in both")
pl <- rbindlist(lapply(c("refseq", "ext3", "full"), function(a)
  merge(res[matrix == paste("stringtie", a)], record, by = c("code", "feature"), suffixes = c("", "_record"))[, annotation := anns[a]]))
pl[, class := factor(fifelse(deg & deg_record, classes[4], fifelse(deg_record, classes[2], fifelse(deg, classes[3], classes[1]))),
                     levels = classes)]
pl[, `:=`(contrast = factor(lab[code], levels = lab), annotation = factor(annotation, levels = anns))]
setorder(pl, class)
p2 <- ggplot(pl, aes(log2FoldChange_record, log2FoldChange, colour = class, shape = class)) +
  geom_hline(yintercept = 0, colour = "grey85") + geom_vline(xintercept = 0, colour = "grey85") +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = 2) +
  geom_point(size = 0.6, alpha = 0.85, show.legend = TRUE) +
  facet_grid(annotation ~ contrast) +
  scale_colour_manual(values = setNames(c("#d9d9d9", "#7f7f7f", "#7f7f7f", "#1a1a1a"), classes), name = NULL, drop = FALSE) +
  scale_shape_manual(values = setNames(c(16, 17, 15, 16), classes), name = NULL, drop = FALSE) +
  guides(colour = guide_legend(override.aes = list(size = 2.5, alpha = 1)), shape = guide_legend()) +
  labs(x = "log2 fold change, record (06)", y = "log2 fold change, recount (StringTie + prepDE)",
       title = "Stressor vs day-3 control: each recounted annotation against the record (genes tested in both)",
       caption = "apeglm log2 fold changes; DEG: padj < 0.05. Dashed line: equal fold change.") +
  theme_psmfc(9) + theme(legend.position = "top", plot.caption = element_text(hjust = 0))
ggsave(file.path(aug_de_dir, "FIG_lfc_vs_record.png"), p2, width = 12, height = 6.5, dpi = 300)
p2
```

# Provenance

```{r provenance}
settings <- c("[07_augmented_de_comparison.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              "model                : as 06: ~ treatment (day-3 control reference), >= 10 counts in a third of samples, apeglm, Wald padj",
              paste0("DEG                  : padj < ", params$padj_cutoff),
              paste0("matrices             : ", paste(names(mats), collapse = ", ")),
              paste0("contrasts            : ", paste(contrasts$code, collapse = ", ")))
inputs <- c(file.path(recount_dir, sprintf("%s_gene_counts.csv.gz", sub(" ", "_", names(mats)))),
            file.path(ann_dir, "ext3_extensions.csv.gz"),
            file.path(de06, "DEG_lists", c("contrasts.csv", "contrast_samples.csv")),
            file.path(de06, "DEG_lists", contrasts$dir, paste0(contrasts$code, "_apeglm.csv")),
            file.path(de06, "count_matrix", c("treatmentinfo_clean.csv", "mitochondrial_loci.csv", "gene_count_matrix_clean.csv")),
            file.path(repo_root, "03_blast", "03_analyses", "genome-foot", "LOC_GO_list.txt"),
            here::here("01_code", "07_augmented_de_comparison.Rmd"))
writeLines(provenance_lines(settings, inputs, repo_root, packages = c("DESeq2", "apeglm", "data.table")),
           file.path(aug_de_dir, "RUN_provenance.txt"))
```
