---
title: "01 - GO slim (biological process) of the treatment-control DEGs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Maps the annotated DEGs of the six treatment-control (TC) contrasts onto the generic GO slim
(biological process), counts up- and down-regulated genes per slim term, and draws them.

Inputs:

- `05_differential-expression/03_analyses/DEG_lists/GOterms_genome/<code>_sigs_ID.csv`: each
  TC DEG with a UniProt hit, its log2 fold change and its GO IDs (05 script 06);
- `05_differential-expression/03_analyses/count_matrix/mitochondrial_loci.csv`: the
  mitochondrial loci (05 script 01), checked to be absent from the DEG lists;
- `02_data/goslim_generic.obo`: the generic GO slim of GO release 2026-01-23 (see the file's
  header and `_derive_goslim.R`). The GO graph, the descendants of each slim term, comes from
  `GO.db`, which must hold the same release (checked below): a slim term made obsolete in
  `GO.db`'s release would have no descendants there and count no genes.

A gene belongs to a slim term when any of its GO IDs is that term or one of its descendants.
A gene can belong to several slim terms, so the counts do not add up to the number of DEGs.

Outputs, in `03_analyses/goslims/`:

- `<code>_sigs_ID.tab`: one row per BP slim term, with the number of genes (all, up, down),
  the GO IDs that mapped to it, and the genes' UniProt entry-name mnemonics (`geneID`, e.g.
  `CO6A3` for COL6A3) and LOC IDs;
- `<code>_sigs_ID.BP_per_gene.tab`: one row per DEG with its slim terms;
- `goslim_summary_TC.csv`: the six contrasts in one long table, the data behind the figure;
- `goslim_TC_heatmap.png`;
- `goslim_provenance.txt`: the slim and `GO.db` versions behind the run.

**One annotation per gene.** A gene with several transcripts can carry several BLAST hits;
it takes the GO IDs of its best hit (highest bitscore, ties in table order), as in
`07_enrichment`. The earlier version of this script took the first-listed hit, which gave 33
of the TC DEGs a different set of GO IDs from the one 07 tested.

**Mitochondrial genes.** The mitochondrial genome's genes and their 293 nuclear copies (126
protein-coding LOCs, 167 pseudogenes) are no longer in the genome DEG lists (05 step 03); they
are analysed on their own in 05 step 13. The check below stops if any reaches a DEG table.

**Correction to the earlier version.** `06-get_GOSlims.Rmd`, whose tables are kept in
`03_analyses/_superseded/goslims_genome/`, lost most of the annotation in two places. The GO
IDs were split on ";" without removing the space after it, and `GSEABase::GOCollection()`
drops IDs that start with a space, so each gene contributed only its first-listed GO ID
(UniProt lists them in numeric order, so often a cellular-component term). Then genes were
looked up for only the first GO ID of each slim term, so its tables listed 36-65% of the
gene-to-slim links implied even by their own `GO.IDs` column. It also downloaded the current
slim into the GSEABase package folder on every run; the slim is now pinned in `02_data/`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({
  library(GSEABase); library(GO.db); library(dplyr); library(tidyr); library(ggplot2)
})
codes    <- c("FOA", "FOW", "FDO", "GOA", "GOW", "GDO")   # tissue + stressor, TC contrasts
slim_obo <- file.path(dat, "goslim_generic.obo")
stopifnot(file.exists(slim_obo))
mt_loci  <- read.csv(file.path(repo_root, "05_differential-expression", "03_analyses", "count_matrix",
                               "mitochondrial_loci.csv"), stringsAsFactors = FALSE)
```

# Slim terms and what each covers

```{r slim}
## the slim and GO.db must be of one GO release
slim_release <- sub("^data-version: go/releases/([0-9-]+).*$", "\\1", grep("^data-version:", readLines(slim_obo), value = TRUE))
go_release   <- GO.db::GO_dbInfo()$value[GO.db::GO_dbInfo()$name == "GOSOURCEDATE"]
if (!identical(slim_release, go_release))
  stop("the slim is of GO release ", slim_release, " and GO.db of ", go_release, "; see 01_code/_derive_goslim.R")
slim_ids <- ids(getOBOCollection(slim_obo))
info <- suppressMessages(AnnotationDbi::select(GO.db, slim_ids, c("ONTOLOGY", "TERM"), "GOID"))
missing_in_godb <- info$GOID[is.na(info$ONTOLOGY)]
if (length(missing_in_godb)) stop("slim terms not in this GO.db: ", paste(missing_in_godb, collapse = ", "))
bp <- info[!is.na(info$ONTOLOGY) & info$ONTOLOGY == "BP", c("GOID", "TERM")]
bp <- bp[order(bp$GOID), ]
## each slim term covers itself and all its descendants
offspring  <- as.list(GOBPOFFSPRING[bp$GOID])
cover_long <- bind_rows(lapply(bp$GOID, function(s) {
  o <- offspring[[s]]
  data.frame(GOslim.BP = s, GO_ID = unique(c(s, o[!is.na(o)])))
}))
cat(nrow(bp), "BP slim terms of", length(slim_ids), "covering", n_distinct(cover_long$GO_ID), "BP terms\n")
```

# Map each contrast's DEGs

```{r map}
slim_one <- function(code) {
  sig <- read.csv(file.path(deg, "GOterms_genome", paste0(code, "_sigs_ID.csv")),
                  stringsAsFactors = FALSE)
  if (any(sig$gene %in% mt_loci$gene))
    stop(code, ": mitochondrial loci in the DEG table; rerun 05_differential-expression")
  genes <- sig %>% arrange(desc(bitscore)) %>% distinct(LOC_ID, .keep_all = TRUE) %>%   # best hit
    transmute(LOC_ID, geneID, log2FoldChange, GO_ID = ifelse(is.na(GO_ID), "", GO_ID))
  gene_go <- genes %>%
    separate_rows(GO_ID, sep = ";") %>%
    mutate(GO_ID = trimws(GO_ID)) %>%                 # "GO:1; GO:2" -> "GO:1", "GO:2"
    filter(grepl("^GO:[0-9]{7}$", GO_ID)) %>%
    distinct()
  hits <- inner_join(gene_go, cover_long, by = "GO_ID", relationship = "many-to-many")

  by_slim <- hits %>%
    group_by(GOslim.BP) %>%
    summarise(GO.IDs  = paste(sort(unique(GO_ID)), collapse = ";"),
              n_genes = n_distinct(LOC_ID),
              n_up    = n_distinct(LOC_ID[log2FoldChange > 0]),
              n_down  = n_distinct(LOC_ID[log2FoldChange < 0]),
              Genes   = paste(sort(unique(geneID)), collapse = ";"),
              LOC_IDs = paste(sort(unique(LOC_ID)), collapse = ";"), .groups = "drop")
  slim_tab <- bp %>% dplyr::rename(GOslim.BP = GOID, Term = TERM) %>%
    left_join(by_slim, by = "GOslim.BP") %>%
    mutate(across(c(n_genes, n_up, n_down), ~ coalesce(.x, 0L)),
           across(c(GO.IDs, Genes, LOC_IDs), ~ coalesce(.x, ""))) %>%
    select(GOslim.BP, Term, n_genes, n_up, n_down, GO.IDs, Genes, LOC_IDs)

  per_gene <- hits %>%
    left_join(dplyr::rename(bp, GOslim.BP = GOID, Term = TERM), by = "GOslim.BP") %>%
    distinct(LOC_ID, geneID, log2FoldChange, GOslim.BP, Term) %>%
    arrange(GOslim.BP) %>%
    group_by(LOC_ID, geneID, log2FoldChange) %>%
    summarise(GOslim.BP = paste(GOslim.BP, collapse = ";"), Term = paste(Term, collapse = ";"),
              .groups = "drop") %>%
    arrange(LOC_ID)

  write.table(slim_tab, file.path(goslims, paste0(code, "_sigs_ID.tab")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(per_gene, file.path(goslims, paste0(code, "_sigs_ID.BP_per_gene.tab")),
              sep = "\t", quote = FALSE, row.names = FALSE)

  list(slim = mutate(slim_tab, code = code),
       counts = data.frame(code = code, DEGs_annotated = nrow(genes),
                           with_GO = n_distinct(gene_go$LOC_ID),
                           in_a_BP_slim_term = n_distinct(hits$LOC_ID),
                           GO_IDs = n_distinct(gene_go$GO_ID),
                           GO_IDs_under_a_BP_slim_term = n_distinct(hits$GO_ID)))
}

res <- lapply(codes, slim_one)
coverage <- bind_rows(lapply(res, `[[`, "counts"))
knitr::kable(coverage, caption = "annotated TC DEGs and how many reach a BP slim term")
```

# Summary table and figure

```{r summary}
tissue_of   <- c(F = "Foot", G = "Gill")
summary_tc <- bind_rows(lapply(res, `[[`, "slim")) %>%
  mutate(tissue = unname(tissue_of[substr(code, 1, 1)]), stressor = substr(code, 2, 3)) %>%
  left_join(select(coverage, code, in_a_BP_slim_term), by = "code") %>%
  select(code, tissue, stressor, GOslim.BP, Term, n_genes, n_up, n_down, in_a_BP_slim_term)
write.csv(summary_tc, file.path(goslims, "goslim_summary_TC.csv"), row.names = FALSE)

writeLines(c(paste0("generated : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
             paste0("slim      : ", sub("^data-version: ", "", grep("^data-version:", readLines(slim_obo), value = TRUE))),
             paste0("GO.db     : ", packageVersion("GO.db"), " (GO source date ",
                    GO.db::GO_dbInfo()$value[GO.db::GO_dbInfo()$name == "GOSOURCEDATE"], ")"),
             paste0("GSEABase  : ", packageVersion("GSEABase")),
             paste0("BP slim terms: ", nrow(bp))),
           file.path(goslims, "goslim_provenance.txt"))
```

Slim terms with fewer than five DEGs over the six contrasts are left out of the figure (they
are in `goslim_summary_TC.csv`).

```{r figure, fig.width = 9, fig.height = 9}
MIN_TOTAL <- 5
n_terms_hit <- n_distinct(summary_tc$GOslim.BP[summary_tc$n_genes > 0])
long <- summary_tc %>%
  group_by(GOslim.BP) %>% filter(sum(n_genes) >= MIN_TOTAL) %>% ungroup() %>%
  pivot_longer(c(n_up, n_down), names_to = "direction", values_to = "n") %>%
  mutate(direction = factor(ifelse(direction == "n_up", "Up-regulated", "Down-regulated"),
                            levels = c("Up-regulated", "Down-regulated")),
         contrast = factor(paste(tissue, stressor), levels = paste(rep(c("Foot", "Gill"), each = 3),
                                                                   c("OA", "OW", "DO"))))
term_order <- long %>% group_by(Term) %>% summarise(total = sum(n)) %>% arrange(total, Term) %>% pull(Term)
long$Term <- factor(long$Term, levels = term_order)
## one single-hue ramp per direction (light = few genes, saturated = many)
ramp <- function(n, hue) ifelse(n == 0, "#f4f4f4",
                                scales::seq_gradient_pal("#f7f7f7", hue)(sqrt(n / max(long$n))))
long$fill <- ifelse(long$direction == "Up-regulated", ramp(long$n, DIRECTION_COLORS[["up"]]),
                    ramp(long$n, DIRECTION_COLORS[["down"]]))
p <- ggplot(long, aes(contrast, Term)) +
  geom_tile(aes(fill = fill), colour = "white", linewidth = 0.6) +
  geom_text(data = filter(long, n > 0), aes(label = n), size = 2.8, colour = "grey10") +
  scale_fill_identity() +
  facet_wrap(~ direction) +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "Treatment-control DEGs by GO slim term (biological process)",
       caption = paste(strwrap(paste0(
         "Numbers are DEGs (apeglm padj < 0.05) with a UniProt GO annotation under the slim term; ",
         "a gene can count under several terms. ", nlevels(long$Term), " of the ", n_terms_hit,
         " slim terms with any DEG are shown (at least ", MIN_TOTAL, " DEGs over all contrasts). ",
         "Generic GO slim, GO release ", slim_release, ". Mitochondrial genes and their nuclear copies are ",
         "analysed separately (05 step 13)."), width = 120), collapse = "\n")) +
  theme_psmfc(10) +
  theme(panel.grid = element_blank(), axis.ticks = element_blank(),
        axis.text.x.top = element_text(angle = 45, hjust = 0), plot.caption = element_text(hjust = 0),
        plot.caption.position = "plot", plot.title.position = "plot")
h <- 1.6 + 0.19 * nlevels(long$Term)
ggsave(file.path(goslims, "goslim_TC_heatmap.png"), p, width = 9, height = h, dpi = 300, limitsize = FALSE)
p
```
