---
title: "02 - NCBI gene summaries for the top-50 DEGs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

For every UniProt accession in the six TC top-50 lists
(`06_differential-expression/03_analyses/top_DEGs/Top_50_genes/<code>_topgenes.csv`), finds
the NCBI Gene record of that protein and its summary, through NCBI E-utilities (rentrez; Winter
2017, *The R Journal* 9:520):

1. the UniProt accession in NCBI Protein (`<accession>[Accession]`; Swiss-Prot entries are
   mirrored there);
2. the protein's link to NCBI Gene (`protein_gene`);
3. that gene's summary.

The first version of this step searched NCBI Gene with the accession as free text and took the
first hit, which is sometimes another gene whose record merely mentions the accession (for
O95238, SPDEF, it returned AR). The linked route gives the gene of that protein or nothing.

Writes, in `03_analyses/Top_gene_summaries/`, one file per contrast, `<code>_topgene_summs.csv`:
one row per accession in the list, `UniProt_ID`, `uniprot_gene` (the UniProt gene name, from
the list), `ncbi_gene_id`, `ncbi_symbol`, `ncbi_organism`, `symbol_matches` (the NCBI symbol is
one of the UniProt gene names, ignoring case) and `Summary` (the NCBI summary, or why there is
none, or `Error: <message>` when a request failed). And `RUN_provenance_summaries.txt`.

NCBI allows 3 requests per second without an API key, counted per IP address; a shared address
can exceed it even at that pace, so requests are spaced and retried with growing waits on a
429 or a server error. An API key, if you have one, goes in the `ENTREZ_KEY` environment
variable (for example `ENTREZ_KEY=<your key>` in `~/.Renviron`), never in this file.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "pipeline_checks.R"))   # provenance_lines()
## rentrez is on CRAN; install once if missing (a headless knit cannot ask for a mirror)
if (!requireNamespace("rentrez", quietly = TRUE))
  install.packages("rentrez", repos = "https://cloud.r-project.org")
library(rentrez)
codes <- c("FOA", "FOW", "FDO", "GOA", "GOW", "GDO")
```

# Look up each accession once

```{r lookup}
lists <- setNames(lapply(codes, function(cd) read.csv(file.path(top50_in, paste0(cd, "_topgenes.csv")))), codes)
acc <- unique(unlist(lapply(lists, function(d) d$uniprot_accession)))
acc <- acc[!is.na(acc) & nzchar(acc)]
cat(length(acc), "distinct accessions in the six lists\n")

n_retry <- 0
ncbi <- function(f, ...) {                       # one spaced, retried E-utilities call
  for (k in 1:6) {
    Sys.sleep(if (nzchar(Sys.getenv("ENTREZ_KEY"))) 0.12 else 0.4)
    r <- tryCatch(f(...), error = function(e) e)
    if (!inherits(r, "error")) return(r)
    if (!grepl("429|HTTP failure: 5|[Tt]imeout|timed out", conditionMessage(r))) stop(r)
    n_retry <<- n_retry + 1
    Sys.sleep(2^k)
  }
  stop("no answer after 6 tries")
}
lookup <- function(a) {
  row <- function(id = NA_character_, sym = NA_character_, org = NA_character_, summary)
    data.frame(UniProt_ID = a, ncbi_gene_id = id, ncbi_symbol = sym, ncbi_organism = org, Summary = summary)
  tryCatch({
    p <- ncbi(entrez_search, db = "protein", term = paste0(a, "[Accession]"), retmax = 1)$ids
    if (!length(p)) return(row(summary = "No NCBI protein record"))
    g <- ncbi(entrez_link, dbfrom = "protein", id = p[1], db = "gene")$links$protein_gene
    if (!length(g)) return(row(summary = "No NCBI gene linked to the protein"))
    s <- ncbi(entrez_summary, db = "gene", id = g[1])
    row(g[1], s$name, s$organism$scientificname, if (nzchar(s$summary)) s$summary else "No summary in NCBI Gene")
  }, error = function(e) row(summary = paste("Error:", conditionMessage(e))))
}
found <- do.call(rbind, lapply(acc, lookup))
stopifnot(nrow(found) == length(acc), !anyDuplicated(found$UniProt_ID))
knitr::kable(as.data.frame(table(outcome = ifelse(!is.na(found$ncbi_gene_id), "gene found",
                                                  sub(":.*$", "", found$Summary)))))
cat("requests retried:", n_retry, "\n")
```

# One table per contrast

```{r write}
for (cd in codes) {
  d <- lists[[cd]]
  d <- d[!is.na(d$uniprot_accession) & nzchar(d$uniprot_accession) & !duplicated(d$uniprot_accession), ]
  out <- merge(data.frame(UniProt_ID = d$uniprot_accession, uniprot_gene = d$gene_name, order = seq_len(nrow(d))),
               found, by = "UniProt_ID", all.x = TRUE)
  stopifnot(nrow(out) == nrow(d))
  out <- out[order(out$order), ]
  names_up <- strsplit(toupper(ifelse(is.na(out$uniprot_gene), "", out$uniprot_gene)), " +")
  out$symbol_matches <- ifelse(is.na(out$ncbi_symbol), NA,
                               mapply(function(s, n) s %in% n, toupper(out$ncbi_symbol), names_up))
  out <- out[, c("UniProt_ID", "uniprot_gene", "ncbi_gene_id", "ncbi_symbol", "ncbi_organism", "symbol_matches", "Summary")]
  write.csv(out, file.path(topgenes, paste0(cd, "_topgene_summs.csv")), row.names = FALSE)
}
knitr::kable(do.call(rbind, lapply(codes, function(cd) {
  o <- read.csv(file.path(topgenes, paste0(cd, "_topgene_summs.csv")))
  data.frame(contrast = cd, accessions = nrow(o), gene_found = sum(!is.na(o$ncbi_gene_id)),
             symbol_matches = sum(o$symbol_matches %in% TRUE), errors = sum(startsWith(o$Summary, "Error")))
})))
```

# Provenance

```{r provenance}
writeLines(provenance_lines(
  c("[02_uniprot_summaries.Rmd]",
    paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
    "route                : UniProt accession -> NCBI Protein [Accession] -> protein_gene link -> Gene summary",
    paste0("NCBI API key         : ", if (nzchar(Sys.getenv("ENTREZ_KEY"))) "yes" else "no"),
    paste0("accessions           : ", length(acc), " (", sum(!is.na(found$ncbi_gene_id)), " with a gene; ", n_retry, " requests retried)")),
  c(file.path(top50_in, paste0(codes, "_topgenes.csv")), here::here("01_code", "02_uniprot_summaries.Rmd")),
  repo_root, packages = "rentrez"),
  file.path(topgenes, "RUN_provenance_summaries.txt"))
```
