---
title: "01 - Clean gene count matrix and sample table"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# What this does

Starts the differential-expression analysis from the gene count matrix of record written by
`04_sequence-alignment/01_code/04_count_matrix_of_record.Rmd` (featureCounts on the RefSeq
annotation with Iso-Seq-extended 3' ends; unique reads, sense strand), and the raw sample sheet
in `02_data/`:

1. orders the libraries, removes T051F and T051G (QC), and sorts the genes by ID. The counts
   are unchanged. (The previous matrix, StringTie + prepDE in
   `04_sequence-alignment/03_analyses/prepDE/`, named its columns by read file, as
   `A10-T015F_S172`; the renaming below still handles that form.);
2. builds the sample table (`sample.1`, `tissue`, `treatment`, `day`, `region`) from the
   sample sheet; treatment and day come from `trt` (`T_OA_d3` -> OA, 3);
3. checks every library against the RNA isolation log and writes the crosswalk between the
   two naming schemes (below);
4. lists the mitochondrial loci (below), which the genome analysis leaves out and step 13
   analyses on their own.

**Foot regions.** Two pieces of the foot were sequenced, and the sheets name them
inconsistently:

| library ID | `tissue` | `region` | sampled in | isolation log name (tissue label) |
|---|---|---|---|---|
| `T001F` | F | foot (phenol gland to tip) | every animal | `T01-F_PG` ("phenol gland") |
| `T001FX` | F | foot (without phenol gland) | the 12 day-0 animals only | `T01-F` ("foot") |
| `T001G` | G | gill | every animal | `T01-G` ("gill") |

Both are foot tissue. The sample sheet's free-text columns call the FX libraries "gill" and
"foot_control"; both are wrong and are not used. The FX libraries stay in the count matrix and
the sample table but enter no contrast (`02_define_contrasts.Rmd`), because the two regions
differ strongly in expression, byssal genes included.

**Day-0 lab controls.** The day-0 feet were cut into two pieces (`F_PG` and `F` in the
isolation log); the day-3 feet were sampled as one (`F_PG` only). The part of the foot behind a
day-0 library is therefore not the part behind a day-3 library, and the experimenter confirms
the day-0 animals are not true lab controls. No contrast uses the day-0 libraries; they stay in
the count matrix and appear only in the all-sample PCA (step 03).

**Mitochondrial loci.** The reference holds the female-type mitochondrial genome
(NC_007687.1: 12 protein genes and their tRNAs), and 293 loci on unplaced scaffolds are copies
of its protein genes (126 protein-coding LOCs and 167 pseudogenes), which take a large and
variable share of the mitochondrial reads (`tools/mt_encoded.R`). `mitochondrial_loci.csv` lists all of them; steps 03 and 05 fit the
genome without them, and step 13 analyses one summed count per mitochondrial protein.

Writes `03_analyses/count_matrix/gene_count_matrix_clean.csv`, `treatmentinfo_clean.csv`,
`library_crosswalk.csv` and `mitochondrial_loci.csv`, which later steps read. Replaces `01_5-gene_count_matrix.Rmd`, which
could not run (it read a matrix that was not in the repository and its sample-table step was
never finished).

```{r paths}
source(here::here("01_code", "_paths.R"))
raw_counts <- file.path(repo_root, "04_sequence-alignment", "03_analyses", "featurecounts", "gene_count_matrix.csv")
stopifnot(file.exists(raw_counts))
dir.create(cnt, showWarnings = FALSE, recursive = TRUE)
QC_REMOVED <- c("T051F", "T051G")
```

# Count matrix

```{r counts}
gcm <- read.csv(raw_counts, check.names = FALSE, row.names = 1)
## read-file base name -> sample ID: drop the plate well ("A10-") and the sequencing index ("_S172")
ids <- sub("^[A-H][0-9]+-", "", sub("_S[0-9]+$", "", colnames(gcm)))
stopifnot(!anyDuplicated(ids), all(grepl("^T[0-9]{3}[A-Z]+$", ids)))
colnames(gcm) <- ids
gcm <- gcm[order(rownames(gcm), method = "radix"), setdiff(ids[order(ids, method = "radix")], QC_REMOVED)]
cat("genes:", nrow(gcm), "| samples:", ncol(gcm), "(", paste(QC_REMOVED, collapse = ", "), "removed )\n")
write.csv(gcm, file.path(cnt, "gene_count_matrix_clean.csv"), row.names = TRUE)
```

# Sample table

```{r samples}
sheet <- read.csv(file.path(dat, "PSMFC-mytilus-byssus-pilot-RNA-tagseq_raw.csv"))
sheet <- sheet[!is.na(sheet$trt) & sheet$trt != "", ]          # drop the sheet's blank rows
stopifnot(all(grepl("^T_[A-Za-z]+_d[0-9]+$", sheet$trt)))
## library-ID suffix -> tissue and foot region (see the table above)
suffix  <- sub("^T[0-9]+", "", sheet$sample.1)
regions <- c(F = "foot (phenol gland to tip)", FX = "foot (without phenol gland)", G = "gill")
stopifnot(all(suffix %in% names(regions)))
ti <- data.frame(sample.1  = sheet$sample.1,
                 tissue    = ifelse(suffix == "G", "G", "F"),
                 treatment = sub("^T_(.*)_d[0-9]+$", "\\1", sheet$trt),
                 day       = sub("^.*_d", "", sheet$trt),
                 region    = unname(regions[suffix]))

## T047F has no library; T047G does. The count matrix is the source of which samples exist.
not_sequenced <- setdiff(ti$sample.1, c(colnames(gcm), QC_REMOVED))
cat("in the sample sheet but not in the count matrix:", paste(not_sequenced, collapse = ", "), "\n")
ti <- ti[match(colnames(gcm), ti$sample.1), ]
stopifnot(!anyNA(ti$sample.1))
rownames(ti) <- ti$sample.1
write.csv(ti, file.path(cnt, "treatmentinfo_clean.csv"), row.names = TRUE)
knitr::kable(as.data.frame(table(paste(ti$treatment, "day", ti$day), ti$region)) |>
               tidyr::pivot_wider(names_from = Var2, values_from = Freq), caption = "samples by group and region")
```

# Crosswalk to the RNA isolation log

The isolation log (`02_data/psmfc_mussel_rna_summary.csv`) names the phenol-gland-to-tip
extractions `T01-F_PG` and the rest of the foot `T01-F`, so its "F" is the library suffix
"FX", not "F". Each library is matched to its isolation row on the mussel tag, RNA
concentration, volume and yield, which are copied between the two sheets, and the match
must agree with the region assigned above.

```{r crosswalk}
iso <- read.csv(file.path(dat, "psmfc_mussel_rna_summary.csv"), check.names = FALSE)
tag <- function(x) sprintf("T%03d", as.integer(sub("^T0*([0-9]+).*$", "\\1", x)))
key <- function(mussel, conc, vol, yield) paste(mussel, conc, vol, yield)
iso_key <- key(tag(iso$sample), iso$`concentration(ng/uL)`, iso$`volume(uL)`, iso$`yield(ng)`)
lib_key <- key(tag(sheet$sample.1), sheet$conc_ng_uL.1, sheet$vol_uL, sheet$yield_ng)
stopifnot(!anyDuplicated(iso_key))
hit <- match(lib_key, iso_key)
xw <- data.frame(sample.1 = sheet$sample.1, region = unname(regions[suffix]),
                 in_count_matrix = ifelse(sheet$sample.1 %in% colnames(gcm), "yes",
                                          ifelse(sheet$sample.1 %in% QC_REMOVED, "removed at QC", "not sequenced")),
                 isolation_log_sample = iso$sample[hit], isolation_log_tissue = iso$tissue[hit],
                 isolation_date = iso$isolation_date[hit])
expected_label <- c("foot (phenol gland to tip)" = "phenol gland", "foot (without phenol gland)" = "foot",
                    gill = "gill")
bad <- xw[is.na(hit) | expected_label[xw$region] != xw$isolation_log_tissue, ]
if (nrow(bad)) { print(bad); stop(nrow(bad), " library(ies) do not match the isolation log as expected.") }
xw <- xw[order(xw$sample.1, method = "radix"), ]
write.csv(xw, file.path(cnt, "library_crosswalk.csv"), row.names = FALSE)
knitr::kable(table(region = xw$region, isolation_log_tissue = xw$isolation_log_tissue))
```

# Mitochondrial loci

One row per matrix row that is mitochondrial: on the mitogenome (`source = mitogenome`,
from the StringTie table of `04_sequence-alignment`), or a copy (`source = nuclear copy`): a
LOC whose best BLAST hit (highest bitscore, `03_blast/03_analyses/genome-foot-uniprot2026_03/LOC_GO_list.txt`)
is a mitochondrially encoded protein, or a locus NCBI names after one
(`04_sequence-alignment/02_data/annotation_mt_like_loci.csv`; this catches the pseudogene
copies, which have no CDS to BLAST). The two routes agree on every coding copy, protein
included; `evidence` says which found each row. `mt_protein` is the protein the row counts towards in
step 13; the share of each sample's reads the two sources hold is printed.

```{r mitochondrial}
source(file.path(repo_root, "tools", "gene_ids.R"))     # gene_key()
source(file.path(repo_root, "tools", "mt_encoded.R"))
mtl <- mitochondrial_loci(rownames(gcm), file.path(blast_go, "LOC_GO_list.txt"), t_data, mt_annot)
stopifnot(!anyDuplicated(mtl$gene), all(mtl$gene %in% rownames(gcm)),
          sum(mtl$source == "mitogenome" & mtl$mt_protein != "tRNA/rRNA") == 12)
write.csv(mtl, file.path(cnt, "mitochondrial_loci.csv"), row.names = FALSE)
knitr::kable(table(mt_protein = mtl$mt_protein, source = mtl$source), caption = "mitochondrial rows")
nuclear_total <- colSums(gcm[!rownames(gcm) %in% mtl$gene, ])
share <- sapply(c("mitogenome", "nuclear copy"), function(s)
  100 * colSums(gcm[mtl$gene[mtl$source == s], , drop = FALSE]) / nuclear_total)
knitr::kable(apply(share, 2, function(x) round(quantile(x, c(0, 0.25, 0.5, 0.75, 1)), 3)),
             caption = "mitochondrial reads as % of each sample's other reads")
```
