---
title: "02 - Define the differential-expression contrasts"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# What this does

Defines every pairwise DESeq2 contrast from the sample table written by script 01, so the
sample set behind each result is set by rule in one place. (Before, the TC sample lists were
written only by two scripts kept as superseded.) Two families, both on day-3 animals only:

| family | contrast | samples | design | reference |
|---|---|---|---|---|
| **TC, of record** | `<T><X>_TC`, X = OA, OW, DO | one tissue: stressor X at day 3 + control at day 3 | `~ treatment` | `control` |
| FG | `FG_TC` | foot + gill of the day-3 controls | `~ tissue` | `F` (foot) |

T is `F` (foot) or `G` (gill). Samples are listed in sample-ID order.

**No day-0 contrasts.** The day-0 lab-control animals are used as a control nowhere: their feet
were dissected differently from the day-3 feet (two pieces, `F_PG` and `F`, against one), so
a day-0 vs day-3 difference would mix the sampled region into the effect (script 01). The
former LC family (each stressor, and the day-3 control, against the day-0 control) and the
day-0 foot-vs-gill contrast `FG_LC` are retired; their tables are kept in
`03_analyses/_superseded/LC_contrasts/`.

Writes `03_analyses/DEG_lists/contrasts.csv` (one row per contrast) and
`contrast_samples.csv` (one row per contrast x sample), read by script 03.

```{r paths}
source(here::here("01_code", "_paths.R"))
ti <- read.csv(file.path(cnt, "treatmentinfo_clean.csv"), row.names = 1, colClasses = "character")
stopifnot(all(ti$region %in% c("foot (phenol gland to tip)", "foot (without phenol gland)", "gill")))
ti <- ti[ti$region != "foot (without phenol gland)" & ti$day == "3", ]   # day-3 libraries only (see above)
```

# Contrasts

```{r define}
tissues   <- c(F = "Foot", G = "Gill")
stressors <- c("OA", "OW", "DO")
defs <- list()
add <- function(code, family, dir, design_var, ref, test, keep, description)
  defs[[code]] <<- list(code = code, family = family, dir = dir, design_var = design_var,
                        ref_level = ref, test_level = test, keep = keep, description = description)

for (t in names(tissues)) {
  for (x in stressors) {
    add(paste0(t, x, "_TC"), "TC", tissues[[t]], "treatment", "control", x,
        with(ti, tissue == t & day == "3" & treatment %in% c("control", x)),
        sprintf("%s: %s vs day-3 treatment control", tissues[[t]], x))
  }
}
add("FG_TC", "FG", "Foot_vs_Gill", "tissue", "F", "G",
    with(ti, tissue %in% c("F", "G") & treatment == "control" & day == "3"),
    "Gill vs foot, day-3 treatment controls")

contrast_samples <- do.call(rbind, lapply(defs, function(d) {
  s <- ti[d$keep, ]
  s <- s[order(s$sample.1, method = "radix"), ]
  data.frame(code = d$code, sample.1 = s$sample.1, tissue = s$tissue, treatment = s$treatment,
             day = s$day, group = ifelse(s[[d$design_var]] == d$test_level, "test", "reference"))
}))
contrasts <- do.call(rbind, lapply(defs, function(d) {
  s <- contrast_samples[contrast_samples$code == d$code, ]
  data.frame(code = d$code, family = d$family, dir = d$dir, design_var = d$design_var,
             ref_level = d$ref_level, test_level = d$test_level,
             n_reference = sum(s$group == "reference"), n_test = sum(s$group == "test"),
             description = d$description)
}))
rownames(contrasts) <- rownames(contrast_samples) <- NULL

## every contrast has two groups, and the levels it names are the ones its samples carry
stopifnot(all(contrasts$n_reference > 0), all(contrasts$n_test > 0),
          !anyDuplicated(paste(contrast_samples$code, contrast_samples$sample.1)))
write.csv(contrasts, file.path(deg, "contrasts.csv"), row.names = FALSE)
write.csv(contrast_samples, file.path(deg, "contrast_samples.csv"), row.names = FALSE)
knitr::kable(contrasts[, c("code", "family", "design_var", "ref_level", "test_level", "n_reference", "n_test", "description")])
```
