---
title: "06 - Join DEG lists to the genome annotation"
output: html_document
date: "2024-02-20"
---

```{r paths, include=FALSE}
source(here::here("01_code", "_paths.R"))
```

# What this does

For each treatment-vs-treatment-control contrast, join the DESeq2 DEG list
(`<X>_TC_siggene*.csv`, one row per gene, `padj < 0.05`) to the BLAST/UniProt/GO master
table (`03_blast/03_analyses/genome-foot/LOC_GO_list.txt`) and write three files to
`03_analyses/DEG_lists/GOterms_genome/`:

- `<X>_sigs_merged.csv`: every DEG with every annotation row it matched (a gene with several
  BLAST hits appears several times, so **this is not a gene count**);
- `<X>_sigs_ID.csv`: the merged rows that carry an annotation;
- `<X>_sigs_unID.csv`: DEGs with no annotation (one row each).

It also writes `03_analyses/DEG_lists/DEG_join_summary.csv` with, per contrast, the number
of DEGs (genes), merged rows, annotated and unannotated genes, and how many of the DEGs are
mitochondrial (a check: script 03 leaves the mitochondrial loci out of the genome fits, so it
must be 0; their results are in `03_analyses/mitochondrial/`, script 13). Quote `n_DEG_genes`
from that file, not the merged-row count.

## Join key and the mitochondrial genes

Nuclear genes are named `gene-LOC134721619|LOC134721619` in the count matrix (or
`gene-LOC...` without the part after `|`, or `STRG.10|LOC...` for a StringTie gene over a
reference gene) and the master table keys them by the `LOC` identifier. The 12 mitochondrial
protein-coding genes are named `gene-ND2|ND2`, `gene-CYTB|CYTB`, and so on and are keyed by
that symbol. `gene_key()` (`tools/gene_ids.R`, shared with 07 and 09) returns the `LOC`
identifier when the name holds one and otherwise the gene ID without `gene-`. The
mitochondrial genes and their nuclear copies are no longer in the genome contrasts (script
03); the check below confirms that none reaches a DEG table.

The join is `left_join(..., na_matches = "never")`: the master table has about 16,800 rows
with an empty key (isoform hits already represented under another transcript of the same
gene), and a default join would attach all of them to any DEG whose key were missing.

```{r load_libraries}
suppressPackageStartupMessages({ library(tidyverse); library(stringr) })
```

```{r masterID}
# Output of blastx + UniProt + GO, one row per (transcript, hit). Columns are selected and
# renamed by their names in the file, so a reordered or extended table cannot shift them.
masterID <- read.delim(file.path(blast_go, "LOC_GO_list.txt"), sep = "\t", header = TRUE)
master_cols <- c(transcript = "transcript", database = "sp", uniprot_accession = "uniprot_accession.x",
                 geneID = "geneID", species = "species", pident = "pident", length = "length",
                 mismatch = "mismatch", gapopen = "gapopen", qstart = "qstart", qend = "qend",
                 sstart = "sstart", send = "send", evalue = "evalue", bitscore = "bitscore",
                 reviewed = "Reviewed", entryname = "Entry.Name", protein_name = "Protein.names",
                 gene_name = "Gene.Names", organism = "Organism", prot_length = "Length",
                 GO_ID = "Gene.Ontology.IDs", GO_BP = "Gene.Ontology..biological.process.",
                 GO_MF = "Gene.Ontology..molecular.function.",
                 GO_CC = "Gene.Ontology..cellular.component.",
                 Genename_orderedlocus = "Gene.Names..ordered.locus.", Induction = "Induction",
                 Pathwaycommons = "PathwayCommons", Reactome = "Reactome", Unipathway = "UniPathway",
                 Interacts_with = "Interacts.with", LOC_ID = "gene")
missing_cols <- setdiff(master_cols, names(masterID))
if (length(missing_cols)) stop("LOC_GO_list.txt lacks column(s): ", paste(missing_cols, collapse = ", "))
# the three helper columns left by the BLAST step (V11, V2.x, uniprot_accession.y) are dropped
masterID <- setNames(masterID[, master_cols], names(master_cols))
stopifnot(identical(sub("_.*$", "", masterID$entryname[!is.na(masterID$entryname)]),
                    masterID$geneID[!is.na(masterID$entryname)]))   # entry name = geneID_species
masterID$LOC_ID[masterID$LOC_ID == ""] <- NA
cat("Master table:", nrow(masterID), "rows;", sum(is.na(masterID$LOC_ID)), "with no key;",
    dplyr::n_distinct(na.omit(masterID$LOC_ID)), "distinct keys\n")
```

`geneID` is the first half of the UniProt entry name (`TYRO` in `TYRO_HUMAN`), a mnemonic
that often differs from the gene symbol (`CO6A3` for COL6A3); `gene_name` holds the gene
symbols UniProt lists.

```{r key}
# Join key for a count-matrix gene name (tools/gene_ids.R): the LOC identifier when there is
# one, otherwise the gene_id without "gene-" (the 12 mitochondrial genes: ND2, CYTB, ...).
source(file.path(repo_root, "tools", "gene_ids.R"))
mt_loci <- read.csv(file.path(cnt, "mitochondrial_loci.csv"), stringsAsFactors = FALSE)  # step 01
```

# Contrasts

```{r contrasts}
contrasts <- tribble(
  ~code,  ~dir,   ~siggene,
  "FDO",  "Foot", "FDO_TC_siggene.csv",
  "GDO",  "Gill", "GDO_TC_siggene.csv",
  "FOA",  "Foot", "FOA_TC_siggene.csv",
  "GOA",  "Gill", "GOA_TC_siggene.csv",
  "FOW",  "Foot", "FOW_TC_siggene.csv",
  "GOW",  "Gill", "GOW_TC_siggene.csv"
)
out_dir <- file.path(deg, "GOterms_genome")
dir.create(out_dir, showWarnings = FALSE)
```

# Join and write

```{r join}
join_contrast <- function(code, dir, siggene) {
  sig <- read.csv(file.path(deg, dir, siggene), header = TRUE) %>%
    mutate(LOC_ID = gene_key(gene))
  if (anyDuplicated(sig$gene) > 0)
    stop(code, ": duplicated gene names in ", siggene)

  merged <- left_join(sig, masterID, by = "LOC_ID", na_matches = "never",
                      relationship = "many-to-many")
  unID <- merged[is.na(merged$geneID), ]
  ID   <- merged[!is.na(merged$geneID), ]

  write_csv(merged, file.path(out_dir, sprintf("%s_sigs_merged.csv", code)))
  write_csv(unID,   file.path(out_dir, sprintf("%s_sigs_unID.csv", code)))
  write_csv(ID,     file.path(out_dir, sprintf("%s_sigs_ID.csv", code)))

  tibble(code = code, tissue = dir,
         n_DEG_genes         = nrow(sig),                       # the number to report
         n_merged_rows       = nrow(merged),                    # one row per gene x hit
         n_ID_rows           = nrow(ID),
         n_unID_rows         = nrow(unID),
         n_genes_annotated   = dplyr::n_distinct(ID$gene),
         n_genes_unannotated = dplyr::n_distinct(unID$gene),
         n_mito_DEGs         = sum(sig$gene %in% mt_loci$gene),
         mito_DEGs           = paste(sig$LOC_ID[sig$gene %in% mt_loci$gene], collapse = ";"))
}

summary_tbl <- pmap_dfr(contrasts, join_contrast)
stopifnot(all(summary_tbl$n_genes_annotated + summary_tbl$n_genes_unannotated == summary_tbl$n_DEG_genes),
          all(summary_tbl$n_mito_DEGs == 0))   # mitochondrial loci are analysed in script 13
write_csv(summary_tbl, file.path(deg, "DEG_join_summary.csv"))
knitr::kable(summary_tbl)
```

`n_DEG_genes` is what the manuscript should report as the number of DEGs. `n_merged_rows`
is larger wherever a gene matched more than one BLAST hit; the earlier draft of the results
section quoted those row counts as DEG numbers.
