---
title: "00 - Run the Iso-Seq branch"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output: html_document
params:
  online: false         # true: steps 02-04 (02 downloads the reads and runs salmon); false: steps 03 and 04
  salmon: "salmon"      # step 02
  threads: 4
  steps: "all"          # "all", or step numbers, e.g. [3]
  stop_on_fail: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs the Iso-Seq branch, a sensitivity analysis: the same differential expression as
`05_differential-expression`, but with the Tag-seq reads quantified against the Iso-Seq
transcriptome instead of the genome, and a comparison of the two; and the comparison of the
genome recounts (`04_sequence-alignment` step 06) with the previous record, which is why the
count matrix of record changed. Each step runs in its own R process with an HTML report and a
log in `03_analyses/knit_html/` (git-ignored) and a `run_log.csv`.

| step | script | writes to `03_analyses/` | needs |
|---|---|---|---|
| 01 | `01_isoseq_transcriptome_check.Rmd` | nothing in `03_analyses/` (a length QC knit by hand; not run by this runner) | owl |
| 02 | `02_salmon_quant.Rmd` | `02_salmon/`: mapping summary and gene counts | gannet, salmon, `04_sequence-alignment` step 04's map; about 15 minutes for the index plus 3 to 4 minutes per library |
| 03 | `03_isoseq_de_comparison.Rmd` | `03_isoseq-de/`: TC contrasts on the Iso-Seq counts and the comparison with the genome | committed files only |
| 04 | `04_augmented_de_comparison.Rmd` | `04_augmented-de/`: option B, the TC contrasts on each recount of `04_sequence-alignment` step 06, compared with the previous record | committed files only |

With `online: false` (the default) steps 03 and 04 run, from the committed counts. Set
`online: true`, with `salmon` pointing at the binary (salmon 1.10.3 was used, from bioconda),
to rebuild step 02 as well. The isoform-to-gene map, the augmented annotations and the genome
recounts, which until 2026-10-03 were steps 02, 05 and 06 here, are steps 04 to 06 of
`04_sequence-alignment`, because the count matrix of record is built from them.

```{r run}
suppressPackageStartupMessages(library(here))
source(file.path(here::here(), "..", "tools", "run_steps.R"))
html_dir <- here::here("03_analyses", "knit_html")
steps <- if (!identical(params$steps, "all")) params$steps else if (isTRUE(params$online)) 2:4 else c(3, 4)
tool_params <- list("02" = list(salmon = params$salmon, threads = params$threads),
                    "04" = list(cores = params$threads))
run_log <- run_steps(here::here("01_code"), html_dir, steps = steps, step_params = tool_params,
                     stop_on_fail = isTRUE(params$stop_on_fail))
finish_run(run_log, html_dir, "Iso-Seq")
```
