---
title: "04 - Differential expression on the recounted genome, compared with the previous record (option B)"
output: html_document
params:
  padj_cutoff: 0.05
  cores: 4                 # parallel fits (forked processes; one on Windows)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Option B, adopted on 2026-10-02: the count matrix of record is now featureCounts on the `ext3`
annotation (`04_sequence-alignment` step 07 takes it from that folder's step 06). This step keeps
the evidence for that choice. It fits the six treatment-control (TC) contrasts to each of
`04_sequence-alignment` step 06's six count matrices (two counters x three annotations), and to
the previous record (StringTie +
prepDE on the original HPC alignment, `04_sequence-alignment/03_analyses/prepDE/`), exactly as
`05_differential-expression` fits them, and compares the results three ways:

1. **Control against the previous record.** StringTie + prepDE on the unchanged RefSeq
   annotation differs from the previous record only in the alignment (`04_sequence-alignment`
   step 06). How close its DEGs come to the previous record's is the yardstick for every other
   difference below.
2. **Annotation effect.** Within each counter, `ext3` and `full` against that counter's RefSeq
   control: the same alignment and counter, so the differences come from the annotation alone.
3. **Each matrix against the previous record**, which is what adopting it changes; for
   featureCounts `ext3`, what the adoption changed.

It also checks that its fit of featureCounts `ext3` is `05`'s: same genes tested, same DEGs,
the same fold changes (the record's DEG lists in `05 .../DEG_lists/`).

Same samples, model, order and thresholds as `05` (and step 03): the samples of each contrast
in `05 .../DEG_lists/contrast_samples.csv`, `~ treatment` with the day-3 control as reference,
fitted on every gene, then genes kept with at least 10 counts in a third of the samples, apeglm
shrinkage (Zhu, Ibrahim and Love 2019, *Bioinformatics* 35:2084), DEG at padj < 0.05. The 331
mitochondrial loci (`05 .../count_matrix/mitochondrial_loci.csv`) are left out of every fit;
`04_sequence-alignment` step 05 kept them unchanged in every annotation. (The previous matrix
filed 21 mitochondrial tRNAs and rRNAs under StringTie's own `STRG` numbers, so they stayed in
its fits; they stay in the refit here too, which therefore reproduces the previous record.) Genes are matched by
`gene_key()`; the novel loci of `full` (`novel:...`) exist only there.

Writes, in `03_analyses/04_augmented-de/`:

- `<counter>_<annotation>_apeglm.csv.gz` and `previous_record_apeglm.csv.gz`: every tested
  gene of every contrast (`code`, `feature`, `baseMean`, apeglm `log2FoldChange`, `lfcSE`, Wald
  `pvalue`, `padj`);
- `deg_summary.csv`: per matrix and contrast, genes tested, DEGs (up, down), and the agreement
  with the previous record: DEGs shared, in the same direction, previous DEGs lost (and of
  those not tested here), new DEGs (and of those on novel loci), and the Spearman correlation
  of fold changes over genes tested in both;
- `annotation_effect.csv`: per counter and contrast, `ext3` and `full` against that counter's
  RefSeq control (DEGs shared, gained, lost, fold-change correlation);
- `record_change.csv`: every gene that is a DEG in the previous record or in the record
  (featureCounts `ext3`), per contrast, with both fold changes, padj and mean counts, and
  whether `ext3` extended it;
- `counts_per_gene.csv`: per gene, the total count over the day-3 libraries on each matrix and
  in the previous record;
- `control_vs_previous.csv`: the control against the previous record: per-library totals, the
  mitochondrial share of each library, gene counts;
- `mitochondrial_share.csv`: per matrix, the median mitochondrial share of the libraries and
  its rank correlation with the previous record's;
- `new_DEGs.csv`: DEGs of each matrix that are not DEGs in the previous record, with the
  previous padj and both counts, and whether `ext3` extended the gene;
- `byssal_genes.csv`: the byssal structural genes (best BLAST hit named as in `09`) that are a
  DEG in any matrix or in the previous record;
- `FIG_deg_counts.png`, `FIG_lfc_vs_previous.png`;
- `RUN_provenance.txt`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
source(file.path(repo_root, "tools", "pipeline_checks.R"))
suppressPackageStartupMessages({ library(DESeq2); library(apeglm); library(data.table); library(ggplot2); library(parallel) })
contrasts <- fread(file.path(de05, "DEG_lists", "contrasts.csv"))[family == "TC"]
cs <- fread(file.path(de05, "DEG_lists", "contrast_samples.csv"))
ti <- fread(file.path(de05, "count_matrix", "treatmentinfo_clean.csv"))
mt_keys <- unique(fread(file.path(de05, "count_matrix", "mitochondrial_loci.csv"))$LOC_ID)
prev_file <- file.path(repo_root, "04_sequence-alignment", "03_analyses", "prepDE", "gene_count_matrix.csv")
stopifnot(nrow(contrasts) == 6, file.exists(prev_file))
counters <- c(stringtie = "StringTie + prepDE", featurecounts = "featureCounts")
anns <- c(refseq = "RefSeq (control)", ext3 = "3' extension", full = "full models")
RECORD <- "featurecounts ext3"                                   # the matrix of record since 2026-10-02
## one row per gene key. StringTie files a few hundred unnamed RefSeq genes under its own
## STRG IDs in some libraries and not in others, so 04_sequence-alignment step 06 names them two ways
## (gene-LOC... and gene-LOC...|LOC...); no library has counts under both, so the rows are summed.
by_key <- function(d) { m <- as.matrix(d[, -1]); rowsum(m, gene_key(d$gene), reorder = FALSE) }
mats <- list(); mt_share <- list()
for (cn in names(counters)) for (a in names(anns)) {
  d <- fread(file.path(recount_dir, sprintf("%s_%s_gene_counts.csv.gz", cn, a)), sep = ",")
  m <- by_key(d)
  stopifnot(!anyDuplicated(rownames(m)), all(cs[code %in% contrasts$code, sample.1] %in% colnames(m)))
  mt_share[[paste(cn, a)]] <- colSums(m[rownames(m) %in% mt_keys, ]) / colSums(m)
  mats[[paste(cn, a)]] <- m[!rownames(m) %in% mt_keys, ]
}
## the previous record: StringTie + prepDE on the HPC alignment, columns named by read file
## ("A10-T015F_S172"), renamed to the sample ID as 05 step 01 did
pd0 <- fread(prev_file, sep = ",")
setnames(pd0, c("gene", sub("^[A-H][0-9]+-", "", sub("_S[0-9]+$", "", names(pd0)[-1]))))
prev_m <- by_key(pd0)
stopifnot(!anyDuplicated(rownames(prev_m)), !anyDuplicated(colnames(prev_m)))
prev_share <- colSums(prev_m[rownames(prev_m) %in% mt_keys, ]) / colSums(prev_m)
mats[["previous record"]] <- prev_m[!rownames(prev_m) %in% mt_keys, ]
sapply(mats, nrow)
```

# Fit each TC contrast on each matrix

```{r fit}
## as step 03 and 05 steps 03-04: fit every gene of the contrast's samples, keep genes with
## >= 10 counts in a third of them, apeglm on those (pvalue and padj from results())
fit_one <- function(counts, row) {
  smp <- cs[code == row$code, sample.1]
  cd  <- as.data.frame(ti[match(smp, sample.1), .(sample.1, treatment)]); rownames(cd) <- cd$sample.1
  cd$treatment <- factor(cd$treatment, levels = c(row$ref_level, row$test_level))
  stopifnot(!anyNA(cd$treatment))
  m <- round(counts[, smp]); m <- m[rowSums(m) > 0, ]
  dds <- DESeq(DESeqDataSetFromMatrix(m, cd, ~ treatment), quiet = TRUE)
  stopifnot(resultsNames(dds)[2] == sprintf("treatment_%s_vs_%s", row$test_level, row$ref_level))
  keep <- rowSums(counts(dds) >= 10) >= ncol(dds) / 3
  shr <- lfcShrink(dds[keep, ], coef = 2, type = "apeglm", quiet = TRUE)
  data.table(code = row$code, feature = rownames(shr), baseMean = shr$baseMean, log2FoldChange = shr$log2FoldChange,
             lfcSE = shr$lfcSE, pvalue = shr$pvalue, padj = shr$padj)
}
jobs <- CJ(matrix = names(mats), i = seq_len(nrow(contrasts)), sorted = FALSE)
cores <- if (.Platform$OS.type == "windows") 1L else params$cores
fits <- mclapply(seq_len(nrow(jobs)), function(j) fit_one(mats[[jobs$matrix[j]]], contrasts[jobs$i[j]]),
                 mc.cores = cores, mc.preschedule = FALSE)
stopifnot(!vapply(fits, inherits, NA, "try-error"))
res <- rbindlist(Map(function(f, mx) f[, matrix := mx], fits, jobs$matrix))
res[, deg := !is.na(padj) & padj < params$padj_cutoff]
for (mx in names(mats)) {
  out <- res[matrix == mx, .(code, feature, baseMean, log2FoldChange, lfcSE, pvalue, padj)]
  fwrite(out, file.path(aug_de_dir, sprintf("%s_apeglm.csv.gz", sub(" ", "_", mx))))
}
previous <- res[matrix == "previous record"]
recounts <- setdiff(names(mats), "previous record")
```

# The record: this fit against 05's

```{r record-check}
record06 <- rbindlist(lapply(seq_len(nrow(contrasts)), function(i) {
  r <- contrasts[i]
  g <- fread(file.path(de05, "DEG_lists", r$dir, sprintf("%s_apeglm.csv", r$code)), sep = ",")
  g[, .(code = r$code, feature = gene_key(gene), log2FoldChange, padj)]
}))
chk <- merge(res[matrix == RECORD, .(code, feature, log2FoldChange, padj)], record06,
             by = c("code", "feature"), all = TRUE, suffixes = c("", "_06"))
rc <- chk[, .(genes_here = sum(!is.na(log2FoldChange)), genes_06 = sum(!is.na(log2FoldChange_06)),
              DEGs_here = sum(padj < params$padj_cutoff, na.rm = TRUE), DEGs_06 = sum(padj_06 < params$padj_cutoff, na.rm = TRUE),
              max_abs_lfc_diff = max(abs(log2FoldChange - log2FoldChange_06), na.rm = TRUE)), by = code]
knitr::kable(rc, caption = "featureCounts ext3 fitted here against 05's DEG lists")
stopifnot(all(rc$genes_here == rc$genes_06), all(rc$DEGs_here == rc$DEGs_06), all(rc$max_abs_lfc_diff < 1e-8),
          identical(chk[!is.na(padj) & padj < params$padj_cutoff, paste(code, feature)],
                    chk[!is.na(padj_06) & padj_06 < params$padj_cutoff, paste(code, feature)]))
```

# Against the previous record

```{r previous}
compare <- function(x, y) {                                     # x: a result set, y: the reference set
  rbindlist(lapply(contrasts$code, function(cd) {
    a <- x[code == cd]; b <- y[code == cd]
    both <- merge(a, b, by = "feature", suffixes = c("", "_ref"))
    data.table(code = cd, tested = nrow(a), DEGs = sum(a$deg), up = sum(a$deg & a$log2FoldChange > 0),
               down = sum(a$deg & a$log2FoldChange < 0), ref_DEGs = sum(b$deg),
               shared = sum(both$deg & both$deg_ref),
               same_direction = sum(both$deg & both$deg_ref & sign(both$log2FoldChange) == sign(both$log2FoldChange_ref)),
               ref_DEGs_lost = sum(b$deg) - sum(both$deg & both$deg_ref),
               ref_DEGs_not_tested = sum(b$deg & !b$feature %in% a$feature),
               new_DEGs = sum(a$deg) - sum(both$deg & both$deg_ref),
               new_DEGs_novel_loci = sum(a$deg & startsWith(a$feature, "novel:")),
               lfc_spearman = round(cor(both$log2FoldChange, both$log2FoldChange_ref, method = "spearman"), 3))
  }))
}
summ <- rbindlist(lapply(recounts, function(mx) {
  s <- compare(res[matrix == mx], previous)
  s[, `:=`(counter = sub(" .*$", "", mx), annotation = sub("^.* ", "", mx))]
}))
setcolorder(summ, c("counter", "annotation"))
setnames(summ, c("ref_DEGs", "ref_DEGs_lost", "ref_DEGs_not_tested"), c("previous_DEGs", "previous_DEGs_lost", "previous_DEGs_not_tested"))
fwrite(summ, file.path(aug_de_dir, "deg_summary.csv"))
knitr::kable(summ)

## every gene that is a DEG in the previous record or in the record
ext <- fread(file.path(ann_dir, "ext3_extensions.csv.gz"))
rch <- merge(previous[, .(code, feature, baseMean, log2FoldChange, padj, deg)],
             res[matrix == RECORD, .(code, feature, baseMean, log2FoldChange, padj, deg)],
             by = c("code", "feature"), all = TRUE, suffixes = c("_previous", "_record"))
rch <- rch[deg_previous %in% TRUE | deg_record %in% TRUE]
rch[, `:=`(status = fifelse(deg_previous %in% TRUE & deg_record %in% TRUE, "DEG in both",
                            fifelse(deg_previous %in% TRUE, "previous record only", "record only")),
           extended_in_ext3 = feature %in% ext$gene)]
rch[, c("baseMean_previous", "baseMean_record") := lapply(.SD, round, 1), .SDcols = c("baseMean_previous", "baseMean_record")]
rch[, c("log2FoldChange_previous", "log2FoldChange_record") := lapply(.SD, round, 3), .SDcols = c("log2FoldChange_previous", "log2FoldChange_record")]
rch[, c("padj_previous", "padj_record") := lapply(.SD, signif, 3), .SDcols = c("padj_previous", "padj_record")]
rch[, c("deg_previous", "deg_record") := NULL]
setcolorder(rch, c("code", "feature", "status", "extended_in_ext3"))
fwrite(rch[order(code, status, feature)], file.path(aug_de_dir, "record_change.csv"))
knitr::kable(dcast(rch[, .N, by = .(code, status)], code ~ status, value.var = "N", fill = 0),
             caption = "DEGs: previous record (StringTie + prepDE) and record (featureCounts ext3)")
```

# Annotation effect within each counter

```{r annotation-effect}
eff <- rbindlist(lapply(names(counters), function(cn) rbindlist(lapply(c("ext3", "full"), function(a) {
  s <- compare(res[matrix == paste(cn, a)], res[matrix == paste(cn, "refseq")])
  s[, .(counter = cn, annotation = a, code, DEGs, control_DEGs = ref_DEGs, shared, gained = new_DEGs,
        gained_novel_loci = new_DEGs_novel_loci, lost = ref_DEGs_lost, lfc_spearman)]
}))))
fwrite(eff, file.path(aug_de_dir, "annotation_effect.csv"))
knitr::kable(eff)
```

# The control against the previous record

```{r control}
rec_m <- prev_m                                                  # the previous record, all genes
ctrl_m <- by_key(fread(file.path(recount_dir, "stringtie_refseq_gene_counts.csv.gz"), sep = ","))
libs <- intersect(colnames(rec_m), colnames(ctrl_m))
genes <- intersect(rownames(rec_m), rownames(ctrl_m))
rec_share <- prev_share[libs]
day3 <- intersect(ti[day == "3", sample.1], libs)
gt <- data.table(gene = genes, previous = rowSums(rec_m[genes, day3]), control = rowSums(ctrl_m[genes, day3]))
g10 <- gt[(previous >= 10 | control >= 10) & !gene %in% mt_keys]
cvr <- data.table(
  measure = c("libraries compared", "genes in both matrices", "genes only in the previous record (StringTie IDs)",
              "per-library total count, control / previous record: median", "range",
              "per-library mitochondrial share, Spearman control vs previous record",
              "nuclear genes with >= 10 counts over the day-3 libraries in either",
              "of those, day-3 total within 10% of the previous record", "Spearman of day-3 totals (those genes)"),
  value = c(length(libs), length(genes), sum(!rownames(rec_m) %in% genes),
            round(median(colSums(ctrl_m[, libs]) / colSums(rec_m[, libs])), 4),
            paste(round(range(colSums(ctrl_m[, libs]) / colSums(rec_m[, libs])), 4), collapse = " to "),
            round(cor(mt_share[["stringtie refseq"]][libs], rec_share, method = "spearman"), 3),
            nrow(g10), round(mean(abs(g10$control - g10$previous) <= 0.1 * g10$previous), 3),
            round(cor(g10$control, g10$previous, method = "spearman"), 4)))
fwrite(cvr, file.path(aug_de_dir, "control_vs_previous.csv"))
knitr::kable(cvr)

## the mitochondrial share of each library on every matrix, against the previous record
mts <- rbindlist(lapply(names(mt_share), function(mx) {
  s <- mt_share[[mx]][libs]
  data.table(matrix = mx, median_share_pct = round(100 * median(s), 2), previous_median_share_pct = round(100 * median(rec_share), 2),
             spearman_with_previous = round(cor(s, rec_share, method = "spearman"), 3))
}))
fwrite(mts, file.path(aug_de_dir, "mitochondrial_share.csv"))
knitr::kable(mts)

## counts per gene on every matrix (day-3 libraries), with the previous record
cpg <- Reduce(function(x, y) merge(x, y, by = "gene", all = TRUE),
              c(list(data.table(gene = rownames(rec_m), previous_record = rowSums(rec_m[, day3]))),
                lapply(recounts, function(mx) { m <- mats[[mx]]
                  setnames(data.table(gene = rownames(m), rowSums(m[, day3])), 2, sub(" ", "_", mx)) })))
cpg <- cpg[!gene %in% mt_keys]
fwrite(cpg, file.path(aug_de_dir, "counts_per_gene.csv"))
knitr::kable(cpg[, lapply(.SD, function(v) sum(v, na.rm = TRUE)), .SDcols = -"gene"], caption = "total count, day-3 libraries")
```

# What the new DEGs are

```{r new-degs}
nd <- rbindlist(lapply(recounts, function(mx) {
  x <- merge(res[matrix == mx & deg == TRUE], previous, by = c("code", "feature"), all.x = TRUE,
             suffixes = c("", "_previous"))
  x[is.na(deg_previous) | deg_previous == FALSE, .(matrix = mx, code, feature, novel_locus = startsWith(feature, "novel:"),
                                                    extended_in_ext3 = feature %in% ext$gene,
                                                    tested_in_previous = !is.na(padj_previous) | !is.na(log2FoldChange_previous),
                                                    baseMean = round(baseMean, 1), baseMean_previous = round(baseMean_previous, 1),
                                                    log2FoldChange = round(log2FoldChange, 3), log2FoldChange_previous = round(log2FoldChange_previous, 3),
                                                    padj = signif(padj, 3), padj_previous = signif(padj_previous, 3))]
}))
fwrite(nd, file.path(aug_de_dir, "new_DEGs.csv"))
knitr::kable(nd[, .(new_DEGs = .N, novel_loci = sum(novel_locus), extended_genes = sum(extended_in_ext3 & !novel_locus),
                    not_tested_in_previous = sum(!tested_in_previous & !novel_locus),
                    median_padj_previous = median(padj_previous, na.rm = TRUE),
                    previous_padj_below_0.2 = round(mean(padj_previous < 0.2, na.rm = TRUE), 2)), by = .(matrix, code)])
```

# Byssal genes

The byssal structural genes are named as in `09` (`BYSSAL_STRUCTURAL_REGEX`): the best BLAST hit
(highest bitscore) of the genome-wide table names a foot protein, plaque, precollagen, ACDC,
byssal tyrosinase, a thread matrix protein or preCOL-NG (UniProt's "Nongradient byssal").

```{r byssal}
byssal_rx <- paste("foot protein", "adhesive plaque", "polyphenolic", "byssal protein-", "preCol", "precollagen",
                   "byssal EP", "ACDC", "byssal tyrosinase", "thread matrix protein", "nongradient byssal", sep = "|")
bl <- fread(file.path(repo_root, "03_blast", "03_analyses", "genome-foot-sprot2026_03", "LOC_GO_list.txt"), sep = "\t",
            select = c("gene", "bitscore", "Protein.names"))
bl[, `:=`(gene = gene_key(trimws(gene)), bitscore = suppressWarnings(as.numeric(bitscore)))]
bl <- bl[!is.na(gene) & nzchar(gene)][order(-bitscore)][!duplicated(gene)]
byssal <- bl[grepl(byssal_rx, Protein.names, ignore.case = TRUE), .(feature = gene, protein = Protein.names)]
bd <- res[feature %in% byssal$feature, .(source = matrix, code, feature, log2FoldChange, padj, deg)]
any_deg <- unique(bd[deg == TRUE, .(code, feature)])
bw <- dcast(merge(bd, any_deg, by = c("code", "feature")), code + feature ~ source, value.var = "deg", fill = NA)
bw <- merge(bw, byssal, by = "feature")
fwrite(bw, file.path(aug_de_dir, "byssal_genes.csv"))
knitr::kable(bw[, lapply(.SD, function(v) sum(v, na.rm = TRUE)), by = code, .SDcols = setdiff(names(bw), c("code", "feature", "protein"))],
             caption = "byssal structural DEGs per contrast and source")
```

# Figures

```{r fig-degs, fig.width = 10, fig.height = 5}
lab <- setNames(sprintf("%s %s", ifelse(substr(contrasts$code, 1, 1) == "F", "Foot", "Gill"), substr(contrasts$code, 2, 3)),
                contrasts$code)
levels_src <- c("previous record", unname(anns))
pd <- rbind(summ[, .(counter = counters[counter], source = anns[annotation], code, DEGs)],
            rbindlist(lapply(counters, function(cl) previous[, .(counter = cl, source = "previous record", DEGs = sum(deg)), by = code])),
            use.names = TRUE)
pd[, `:=`(source = factor(source, levels = levels_src), contrast = factor(lab[code], levels = lab),
          counter = factor(counter, levels = counters))]
## neutral greys for set categories (tools/plot_style.R), labelled
p <- ggplot(pd, aes(contrast, DEGs, fill = source)) +
  geom_col(position = position_dodge(width = 0.82), width = 0.78, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = DEGs), position = position_dodge(width = 0.82), vjust = -0.3, size = 2.4) +
  facet_wrap(~ counter, nrow = 1) +
  scale_fill_manual(values = setNames(c("#1a1a1a", "#7f7f7f", "#a6a6a6", "#d0d0d0"), levels_src), name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = NULL, y = "DEGs (apeglm padj < 0.05)", title = "DEGs per contrast: the previous record and the three recounted annotations",
       caption = paste("Same samples, model and thresholds as 05; mitochondrial loci left out. Previous record: StringTie + prepDE on the",
                       "original alignment.\nRecord since 2026-10-02: featureCounts, 3' extension.")) +
  theme_psmfc(10) + theme(legend.position = "top", plot.caption = element_text(hjust = 0),
                          axis.text.x = element_text(angle = 30, hjust = 1))
ggsave(file.path(aug_de_dir, "FIG_deg_counts.png"), p, width = 10, height = 5, dpi = 300)
p
```

```{r fig-lfc, fig.width = 12, fig.height = 6.5}
classes <- c("DEG in neither", "DEG in the previous record only", "DEG in the recount only", "DEG in both")
pl <- rbindlist(lapply(c("refseq", "ext3", "full"), function(a)
  merge(res[matrix == paste("featurecounts", a)], previous, by = c("code", "feature"), suffixes = c("", "_previous"))[, annotation := anns[a]]))
pl[, class := factor(fifelse(deg & deg_previous, classes[4], fifelse(deg_previous, classes[2], fifelse(deg, classes[3], classes[1]))),
                     levels = classes)]
pl[, `:=`(contrast = factor(lab[code], levels = lab), annotation = factor(annotation, levels = anns))]
setorder(pl, class)
p2 <- ggplot(pl, aes(log2FoldChange_previous, log2FoldChange, colour = class, shape = class)) +
  geom_hline(yintercept = 0, colour = "grey85") + geom_vline(xintercept = 0, colour = "grey85") +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = 2) +
  geom_point(size = 0.6, alpha = 0.85, show.legend = TRUE) +
  facet_grid(annotation ~ contrast) +
  scale_colour_manual(values = setNames(c("#d9d9d9", "#7f7f7f", "#7f7f7f", "#1a1a1a"), classes), name = NULL, drop = FALSE) +
  scale_shape_manual(values = setNames(c(16, 17, 15, 16), classes), name = NULL, drop = FALSE) +
  guides(colour = guide_legend(override.aes = list(size = 2.5, alpha = 1)), shape = guide_legend()) +
  labs(x = "log2 fold change, previous record (StringTie + prepDE)", y = "log2 fold change, recount (featureCounts)",
       title = "Stressor vs day-3 control: each featureCounts annotation against the previous record (genes tested in both)",
       caption = "apeglm log2 fold changes; DEG: padj < 0.05. Dashed line: equal fold change. The 3' extension row is the record.") +
  theme_psmfc(9) + theme(legend.position = "top", plot.caption = element_text(hjust = 0))
ggsave(file.path(aug_de_dir, "FIG_lfc_vs_previous.png"), p2, width = 12, height = 6.5, dpi = 300)
p2
```

# Provenance

```{r provenance}
settings <- c("[04_augmented_de_comparison.Rmd]",
              paste0("generated            : ", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
              "model                : as 05: ~ treatment (day-3 control reference), >= 10 counts in a third of samples, apeglm, Wald padj",
              paste0("DEG                  : padj < ", params$padj_cutoff),
              paste0("matrices             : ", paste(names(mats), collapse = ", ")),
              paste0("record               : ", RECORD, " (05's DEG lists reproduced)"),
              paste0("contrasts            : ", paste(contrasts$code, collapse = ", ")))
inputs <- c(file.path(recount_dir, sprintf("%s_gene_counts.csv.gz", sub(" ", "_", recounts))), prev_file,
            file.path(ann_dir, "ext3_extensions.csv.gz"),
            file.path(de05, "DEG_lists", c("contrasts.csv", "contrast_samples.csv")),
            file.path(de05, "DEG_lists", contrasts$dir, paste0(contrasts$code, "_apeglm.csv")),
            file.path(de05, "count_matrix", c("treatmentinfo_clean.csv", "mitochondrial_loci.csv")),
            file.path(repo_root, "03_blast", "03_analyses", "genome-foot-sprot2026_03", "LOC_GO_list.txt"),
            here::here("01_code", "04_augmented_de_comparison.Rmd"))
writeLines(provenance_lines(settings, inputs, repo_root, packages = c("DESeq2", "apeglm", "data.table")),
           file.path(aug_de_dir, "RUN_provenance.txt"))
```
