---
title: "7_group-gene-plots"
author: "Matt George"
date: "2023-05-29"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE, cache = TRUE)
```

```{bash, engine.opts='-l'}
echo $PATH
```

### Load libraries
```{r load_libraries, inlcude = TRUE}

## clear
rm(list=ls())

# List of packages we want to install (run every time)
load.lib<-c("tidyverse", "readxl", "gptstudio", "plotly", "ggbreak", "patchwork")

# Select only the packages that aren't currently installed (run every time)
install.lib <- load.lib[!load.lib %in% installed.packages()]

# And finally we install the missing packages, including their dependency.
for(lib in install.lib) install.packages(lib,dependencies=TRUE)
# After the installation process completes, we load all packages.
sapply(load.lib,require,character=TRUE)
                        
```

#Set ggplot theme
```{r ggplot_theme, include=FALSE}

my_theme <- theme(line              = element_line(linewidth=1.5),
                  rect              = element_rect(linewidth=1.5),
                  text              = element_text(size=14,color="black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(), 
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size=16,color="black"),
                  axis.text.y       = element_text(size=16,color="black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color="black"),
                  axis.ticks.y      = element_line(color="black"),
                  # axis.line         = element_line(color = "black", size = 0.1),
                  panel.border      = element_rect(color = "black", fill=NA, linewidth=1.5),
                  legend.key        = element_blank()) # removes background of legend bullets
```

### Load data
```{r}

chaperones_2n        <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "MC_list_2n", col_names = TRUE)
chaperones_2n$trt    <- factor(chaperones_2n$trt, levels = c("ctrl","SS","MS"))
chaperones_2n$group  <- factor(chaperones_2n$group)

chaperones_3n        <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "MC_list_3n", col_names = TRUE)
chaperones_3n$trt    <- factor(chaperones_3n$trt, levels = c("ctrl","SS","MS"))
chaperones_3n$group  <- factor(chaperones_3n$group)


chaperones_2n_limited <- chaperones_2n %>% filter(group == "HSP 27" | group == "HSP 60" | group == "HSP 68" | group == "HSP 70" | group == "HSP 83" | group == "HSP 90" | group == "Bag3")
chaperones_3n_limited <- chaperones_3n %>% filter(group == "HSP 27" | group == "HSP 60" | group == "HSP 68" | group == "HSP 70" | group == "HSP 83" | group == "HSP 90" | group == "Bag3")


```

### Graph
```{r}
library("RColorBrewer")
# create a line plot using the "chaperones" data frame. Plot L2FC as a function of trt, grouped by ploidy, for each transcript. Color lines according to the values in group. 



grey_scale <- c("#323232","#404040","#4f4f4f","#5e5e5e","#5e5e5f","#69696a","#747475","#7f7f80","#8a8a8b","#959596","#a0a0a1")


# create line plot
p1 <- ggplot(chaperones_2n, aes(x = trt, y = L2FC, group = transcript, col = group)) + 
      geom_line(linewidth=1) + 
      labs(x = "trt", y = "L2FC") +
      scale_color_manual(values=c("grey","#f3eb3c",rep("grey",4),"#ff9a01","grey","#4ac159","#ef5754",rep("grey",5),"#028dd8")) +
      scale_y_continuous(breaks = seq(-2,10,2),limits = c(-3, 11)) +
      scale_x_discrete(expand=c(0,1)) +
      my_theme

p1


p2 <- ggplot(chaperones_3n, aes(x = trt, y = L2FC, group = transcript, col = group)) + 
      geom_line(linewidth=1) + 
      labs(x = "trt", y = "L2FC") +
      scale_color_manual(values=c("grey","#f3eb3c",rep("grey",4),"#ff9a01","grey","#4ac159","#ef5754",rep("grey",5),"#028dd8")) +
      scale_y_continuous(breaks = seq(-2,10,2),limits = c(-3, 11)) +
      scale_x_discrete(expand=c(0,1)) +
      my_theme

p2


ggsave("analyses/chaperones/LINE_chaperones-2n.png",
       plot   = p1,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 4,
       units  = "in")

ggsave("analyses/chaperones/LINE_chaperones-3n.png",
       plot   = p2,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 4,
       units  = "in")


```

### Load data
```{r}

IAP_2n        <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "IAP_list_2n", col_names = TRUE)
IAP_2n$trt    <- factor(IAP_2n$trt, levels = c("ctrl","SS","MS"))
IAP_2n$group  <- factor(IAP_2n$group)

IAP_3n        <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "IAP_list_3n", col_names = TRUE)
IAP_3n$trt    <- factor(IAP_3n$trt, levels = c("ctrl","SS","MS"))
IAP_3n$group  <- factor(IAP_3n$group)


# IAP_2n_limited <- IAP_2n %>% filter(group == "HSP 27" | group == "HSP 60" | group == "HSP 68" | group == "HSP 70" | group == "HSP 83" | group == "HSP 90" | group == "Bag3")
# IAP_3n_limited <- IAP_3n %>% filter(group == "HSP 27" | group == "HSP 60" | group == "HSP 68" | group == "HSP 70" | group == "HSP 83" | group == "HSP 90" | group == "Bag3")


```

### Graph
```{r}
library("RColorBrewer")
# create a line plot using the "chaperones" data frame. Plot L2FC as a function of trt, grouped by ploidy, for each transcript. Color lines according to the values in group. 



grey_scale <- c("#323232","#404040","#4f4f4f","#5e5e5e","#5e5e5f","#69696a","#747475","#7f7f80","#8a8a8b","#959596","#a0a0a1")


# create line plot
p1 <- ggplot(IAP_2n, aes(x = trt, y = L2FC, group = transcript, col = group)) + 
      geom_line(linewidth=1) + 
      labs(x = "trt", y = "L2FC") +
      # scale_color_manual(values=c("grey","#f3eb3c",rep("grey",4),"#ff9a01","grey","#4ac159","#ef5754",rep("grey",5),"#028dd8")) +
      scale_y_continuous(breaks = seq(-2,10,2),limits = c(-2.3, 6)) +
      scale_x_discrete(expand=c(0,1)) +
      my_theme

p1


p2 <- ggplot(IAP_3n, aes(x = trt, y = L2FC, group = transcript, col = group)) + 
      geom_line(linewidth=1) + 
      labs(x = "trt", y = "L2FC") +
      # scale_color_manual(values=c("grey","#f3eb3c",rep("grey",4),"#ff9a01","grey","#4ac159","#ef5754",rep("grey",5),"#028dd8")) +
      scale_y_continuous(breaks = seq(-2,10,2),limits = c(-2.3, 6)) +
      scale_x_discrete(expand=c(0,1)) +
      my_theme

p2


ggsave("analyses/IAP/LINE_IAP-2n.png",
       plot   = p1,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 4,
       units  = "in")

ggsave("analyses/IAP/LINE_IAP-3n.png",
       plot   = p2,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 4,
       units  = "in")


```










### Compare the top 50 (characterized) genes (25 up / 25 down regulated) in each ploidy following SS and MS exposure
### Load data
```{r}

# load dataset and gene codes
TOP_C        <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "top_C", col_names = TRUE)
TOP_SS       <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "top_SS", col_names = TRUE)
TOP_MS       <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "top_MS", col_names = TRUE)
TOP_C_list   <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "top_C_list", col_names = TRUE)
TOP_SS_list  <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "top_SS_list", col_names = TRUE)
TOP_MS_list  <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "top_MS_list", col_names = TRUE)
TOP_color    <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "top_color", col_names = TRUE)

# filter by ploidy
TOP_SS_2n  <- TOP_SS %>% filter(ploidy == "diploid")
TOP_SS_3n  <- TOP_SS %>% filter(ploidy == "triploid")
TOP_MS_2n  <- TOP_MS %>% filter(ploidy == "diploid")
TOP_MS_3n  <- TOP_MS %>% filter(ploidy == "triploid")

# factor gene codes so x axis is in the right order
TOP_C$code     <- factor(TOP_C$code,     levels = TOP_C_list$code, ordered = TRUE)
TOP_SS_2n$code <- factor(TOP_SS_2n$code, levels = TOP_SS_list$diploid, ordered = TRUE)
TOP_SS_3n$code <- factor(TOP_SS_3n$code, levels = TOP_SS_list$triploid, ordered = TRUE)
TOP_MS_2n$code <- factor(TOP_MS_2n$code, levels = TOP_MS_list$diploid, ordered = TRUE)
TOP_MS_3n$code <- factor(TOP_MS_3n$code, levels = TOP_MS_list$triploid, ordered = TRUE)

TOP_C$category     <- factor(TOP_C$category,     levels = TOP_color$category, ordered = TRUE)
TOP_SS_2n$category <- factor(TOP_SS_2n$category, levels = TOP_color$category, ordered = TRUE)
TOP_SS_3n$category <- factor(TOP_SS_3n$category, levels = TOP_color$category, ordered = TRUE)
TOP_MS_2n$category <- factor(TOP_MS_2n$category, levels = TOP_color$category, ordered = TRUE)
TOP_MS_3n$category <- factor(TOP_MS_3n$category, levels = TOP_color$category, ordered = TRUE)

```

#Set ggplot theme
```{r ggplot_theme, include=FALSE}

TOP_theme <- theme(line          = element_line(linewidth=1.5),
                  rect              = element_rect(linewidth=1.5),
                  text              = element_text(size=12,color="black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(), 
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size=10,color="black"),
                  axis.text.y       = element_text(size=6,color="black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color="black", linewidth = 0.5),
                  axis.ticks.y      = element_line(color="black", linewidth = 0.5),
                  # axis.line         = element_line(color = "black", size = 0.1),
                  panel.border      = element_rect(color = "black", fill=NA, linewidth=1.5),
                  legend.key        = element_blank()) # removes background of legend bullets
```

### Graph
```{r}

g0 <- ggplot(TOP_C, aes(x = code, y = L2FC, fill = category)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = TOP_color$color) +
      scale_x_discrete(expand = c(0,1)) +
      # scale_y_continuous(limits = c(-3,3), breaks = seq(-3,3,1)) +
      coord_flip() +  
      TOP_theme 

g0


g1 <- ggplot(TOP_SS_2n, aes(x = code, y = L2FC, fill = category)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = TOP_color$color) +
      scale_x_discrete(expand = c(0,1)) +
      # scale_y_continuous(limits = c(-3,3), breaks = seq(-3,3,1)) +
      coord_flip() +  
      TOP_theme

g1

g2 <- ggplot(TOP_SS_3n, aes(x = code, y = L2FC, fill = category)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = TOP_color$color) +
      scale_x_discrete(expand = c(0,1)) +
      # scale_y_continuous(limits = c(-3,3), breaks = seq(-3,3,1)) +
      coord_flip() +  
      TOP_theme

g2

g3 <- ggplot(TOP_MS_2n, aes(x = code, y = L2FC, fill = category)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = TOP_color$color) +
      scale_x_discrete(expand = c(0,1)) +
      scale_y_continuous(limits = c(-3.5,9), breaks = seq(-2,10,2)) +
      # scale_y_break(c(3,4.5),scales = 0.2) +
      coord_flip() +  
      TOP_theme

g3

g4 <- ggplot(TOP_MS_3n, aes(x = code, y = L2FC, fill = category)) +
      geom_bar(stat = "identity") +
      scale_fill_manual(values = TOP_color$color) +
      scale_x_discrete(expand = c(0,1)) +
      scale_y_continuous(limits = c(-6,9), breaks = seq(-4,10,2)) +
      # scale_y_break(c(3,4.5),scales = 0.2) +
      coord_flip() +  
      TOP_theme

g4


ggsave("analyses/TOP/BAR_top_C.png",
       plot   = g0,
       dpi    = 1200,
       device = "png",
       width  = 5.5,
       height = 5,
       units  = "in")


ggsave("analyses/TOP/BAR_top_SS_2n.png",
       plot   = g1,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 5,
       units  = "in")

ggsave("analyses/TOP/BAR_top_SS_3n.png",
       plot   = g2,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 5,
       units  = "in")

ggsave("analyses/TOP/BAR_top_MS_2n_full.png",
       plot   = g3,
       dpi    = 1200,
       device = "png",
       width  = 5.5,
       height = 5,
       units  = "in")

ggsave("analyses/TOP/BAR_top_MS_3n_full.png",
       plot   = g4,
       dpi    = 1200,
       device = "png",
       width  = 5.5,
       height = 5,
       units  = "in")

```


#### IMMUNE
### Compare (characterized) immune genes with >1 LFC difference between 2n/3n; only significant DEG included (diploid control used as consistent standard)
### Load data
```{r}

# Single stressor
IMMUNE_SS         <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "immune_SS_diff", col_names = TRUE)
IMMUNE_SS_list    <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "immune_SS_diff_list", col_names = TRUE)
IMMUNE_SS$ploidy  <- factor(IMMUNE_SS$ploidy)
IMMUNE_SS$code    <- factor(IMMUNE_SS$code, levels = IMMUNE_SS_list$code, ordered = TRUE)

# Multiple stressors
IMMUNE_MS         <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "immune_MS_diff", col_names = TRUE)
IMMUNE_MS_list    <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "immune_MS_diff_list", col_names = TRUE)
IMMUNE_MS$ploidy  <- factor(IMMUNE_MS$ploidy)
IMMUNE_MS$code    <- factor(IMMUNE_MS$code, levels = IMMUNE_MS_list$code, ordered = TRUE)


#Set ggplot theme

immune_theme <- theme(line          = element_line(linewidth=1.5),
                  rect              = element_rect(linewidth=1.5),
                  text              = element_text(size=12,color="black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(), 
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size=10,color="black"),
                  axis.text.y       = element_text(size=6,color="black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color="black", linewidth = 0.5),
                  axis.ticks.y      = element_line(color="black", linewidth = 0.5),
                  # axis.line         = element_line(color = "black", size = 0.1),
                  panel.border      = element_rect(color = "black", fill=NA, linewidth=1.5),
                  legend.key        = element_blank()) # removes background of legend bullets


# Graph

g1 <- ggplot(IMMUNE_SS, aes(x = code, y = L2FC, color = ploidy)) +
     geom_line(aes(group = code), color = "black", linewidth = 1) +
     geom_point(size=2) +
     scale_color_manual(values = c("skyblue4", "orangered2")) +
     scale_x_discrete(expand = c(0,1)) +
     scale_y_continuous(limits = c(-4,4), breaks = seq(-4,10,2)) +
     coord_flip() +  
     immune_theme

ggplotly(g1)

g1


g2 <- ggplot(IMMUNE_MS, aes(x = code, y = L2FC, color = ploidy)) +
      geom_line(aes(group = code), color = "black", linewidth = 1) +
      geom_point(size=2) +
      scale_color_manual(values = c("skyblue4", "orangered2")) +
      scale_x_discrete(expand = c(0,1)) +
      # scale_y_continuous(limits = c(-4.5,4), breaks = seq(-2,10,2)) +
      coord_flip() + 
      immune_theme

ggplotly(g2)

g2

ggsave("analyses/immune/LINE_immune_SS.png",
       plot   = g1,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 5,
       units  = "in")

ggsave("analyses/immune/LINE_immune_MS.png",
       plot   = g2,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 5,
       units  = "in")


```
#### STRESS
### Compare (characterized) stress genes with >1 LFC difference between 2n/3n; only significant DEG included (diploid control used as consistent standard)
### Load data
```{r}

# Single stressor
STRESS_SS         <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "stress_SS_diff", col_names = TRUE)
STRESS_SS_list    <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "stress_SS_diff_list", col_names = TRUE)
STRESS_SS$ploidy  <- factor(STRESS_SS$ploidy)
STRESS_SS$code    <- factor(STRESS_SS$code, levels = STRESS_SS_list$code, ordered = TRUE)

# Multiple stressors
STRESS_MS         <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "stress_MS_diff", col_names = TRUE)
STRESS_MS_list    <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "stress_MS_diff_list", col_names = TRUE)
STRESS_MS$ploidy  <- factor(STRESS_MS$ploidy)
STRESS_MS$code    <- factor(STRESS_MS$code, levels = STRESS_MS_list$code, ordered = TRUE)


#Set ggplot theme

STRESS_theme <- theme(line          = element_line(linewidth=1.5),
                  rect              = element_rect(linewidth=1.5),
                  text              = element_text(size=12,color="black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(), 
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size=10,color="black"),
                  axis.text.y       = element_text(size=6,color="black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color="black", linewidth = 0.5),
                  axis.ticks.y      = element_line(color="black", linewidth = 0.5),
                  # axis.line         = element_line(color = "black", size = 0.1),
                  panel.border      = element_rect(color = "black", fill=NA, linewidth=1.5),
                  legend.key        = element_blank()) # removes background of legend bullets


# Graph

g1 <- ggplot(STRESS_SS, aes(x = code, y = L2FC, color = ploidy)) +
      geom_line(aes(group = code), color = "black", linewidth = 1) +
      geom_point(size=2) +
      scale_color_manual(values = c("skyblue4", "orangered2")) +
      scale_x_discrete(expand = c(0,1)) +
      scale_y_continuous(limits = c(-3,3.5), breaks = seq(-2,10,2)) +
      coord_flip() +  
      STRESS_theme

ggplotly(g1)

g1


g2 <- ggplot(STRESS_MS, aes(x = code, y = L2FC, color = ploidy)) +
      geom_line(aes(group = code), color = "black", linewidth = 1) +
      geom_point(size=2) +
      scale_color_manual(values = c("skyblue4", "orangered2")) +
      scale_x_discrete(expand = c(0,1)) +
      scale_y_continuous(limits = c(-3,10), breaks = seq(-2,10,2)) +
      # scale_y_break(c(3,3.5),scales = 0.5) +
      coord_flip() +  
      STRESS_theme

ggplotly(g2)

g2

ggsave("analyses/stress/LINE_stress_SS.png",
       plot   = g1,
       dpi    = 1200,
       device = "png",
       width  = 5,
       height = 5,
       units  = "in")

ggsave("analyses/stress/LINE_stress_MS.png",
       plot   = g2,
       dpi    = 1200,
       device = "png",
       width  = 4,
       height = 5,
       units  = "in")


```


#### Metabolism
### Compare (characterized) metabolic genes with >1 LFC difference between 2n/3n; only significant DEG included (diploid control used as consistent standard)
### Load data
```{r}

# Single metaor
META_SS         <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "meta_SS_diff", col_names = TRUE)
META_SS_list    <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "meta_SS_diff_list", col_names = TRUE)
META_SS$ploidy  <- factor(META_SS$ploidy)
META_SS$code    <- factor(META_SS$code, levels = META_SS_list$code, ordered = TRUE)

# Multiple metaors
META_MS         <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "meta_MS_diff", col_names = TRUE)
META_MS_list    <- read_excel("data/gene_lists-diploid-control.xlsx", sheet = "meta_MS_diff_list", col_names = TRUE)
META_MS$ploidy  <- factor(META_MS$ploidy)
META_MS$code    <- factor(META_MS$code, levels = META_MS_list$code, ordered = TRUE)


#Set ggplot theme

META_theme <- theme(line          = element_line(linewidth=1.5),
                  rect              = element_rect(linewidth=1.5),
                  text              = element_text(size=12,color="black"),
                  panel.background  = element_blank(),
                  panel.grid.major  = element_blank(), 
                  panel.grid.minor  = element_blank(),
                  axis.text.x       = element_text(size=10,color="black"),
                  axis.text.y       = element_text(size=6,color="black"),
                  axis.title.x      = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
                  axis.title.y      = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
                  axis.ticks.x      = element_line(color="black", linewidth = 0.5),
                  axis.ticks.y      = element_line(color="black", linewidth = 0.5),
                  # axis.line         = element_line(color = "black", size = 0.1),
                  panel.border      = element_rect(color = "black", fill=NA, linewidth=1.5),
                  legend.key        = element_blank()) # removes background of legend bullets


# Graph

g1 <- ggplot(META_SS, aes(x = code, y = L2FC, color = ploidy)) +
      geom_line(aes(group = code), color = "black", linewidth = 1) +
      geom_point(size=2) +
      scale_color_manual(values = c("skyblue4", "orangered2")) +
      scale_x_discrete(expand = c(0,1)) +
      scale_y_continuous(limits = c(-2.5,2.5), breaks = seq(-2,10,1)) +
      coord_flip() +  
      META_theme

ggplotly(g1)

g1


g2 <- ggplot(META_MS, aes(x = code, y = L2FC, color = ploidy)) +
      geom_line(aes(group = code), color = "black", linewidth = 1) +
      geom_point(size=2) +
      scale_color_manual(values = c("skyblue4", "orangered2")) +
      scale_x_discrete(expand = c(0,1)) +
      scale_y_continuous(limits = c(-2.5,2.5), breaks = seq(-2,10,1)) +
      # scale_y_break(c(3,3.5),scales = 0.5) +
      coord_flip() +  
      META_theme

ggplotly(g2)

g2

ggsave("analyses/meta/LINE_meta_SS.png",
       plot   = g1,
       dpi    = 1200,
       device = "png",
       width  = 4.5,
       height = 5,
       units  = "in")

ggsave("analyses/meta/LINE_meta_MS.png",
       plot   = g2,
       dpi    = 1200,
       device = "png",
       width  = 4.5,
       height = 5,
       units  = "in")


```
