---
title: "00 - Run the differential-expression analysis"
author: "Byssus expression analysis"
date: "`r Sys.Date()`"
output: html_document
params:
  steps: "all"          # "all", or step numbers, e.g. [4, 6]
  stop_on_fail: true    # stop at the first failed step
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

Runs every numbered script in `01_code/` in order, each in its own R process, with an HTML
report and a log per step in `03_analyses/knit_html/` (git-ignored) and a `run_log.csv`.
Knit this file from inside `differential-expression.Rproj`, or let the repository-level
`00_run_pipeline.Rmd` call it.

| step | script | what it does |
|---|---|---|
| 01 | `01_clean_count_matrix.Rmd` | clean counts and sample table from `05_sequence-alignment`'s prepDE matrix |
| 02 | `02_define_contrasts.Rmd` | the 16 contrasts (TC of record, LC, foot vs gill) and their samples |
| 03 | `03_deseq_contrasts.Rmd` | one DESeq2 fit per contrast, saved to `03_analyses/dds/`; PCA plots |
| 04 | `04_shrinkage_filtration.Rmd` | apeglm shrinkage, full tables, DEG lists, MA plots, DEG counts |
| 05 | `05_fourlevel_sensitivity.Rmd` | one four-level model per tissue vs the pairwise lists |
| 06 | `06_join_annotation.Rmd` | TC DEG lists joined to the BLAST / UniProt / GO annotation |
| 07 | `07_top_degs.Rmd` | top-50 DEGs per TC contrast |
| 08-10 | `08_deg_venn.Rmd`, `09_volcano_plots.Rmd`, `10_number_degs.Rmd` | TC figures |
| 11 | `11_deg_figures_all_contrasts.Rmd` | DEG counts, volcano plots and TC-vs-LC overlap for all families |
| 12 | `12_deg_list_cleanup.Rmd` | DEG lists for Zenodo |

Script 03 takes about six minutes; the rest a few minutes together. Step 08 installs
`ggvenn` from CRAN if it is missing.

```{r run}
suppressPackageStartupMessages(library(here))
source(file.path(here::here(), "..", "tools", "run_steps.R"))
html_dir <- here::here("03_analyses", "knit_html")
run_log <- run_steps(here::here("01_code"), html_dir, steps = params$steps,
                     stop_on_fail = isTRUE(params$stop_on_fail))
finish_run(run_log, html_dir, "Differential-expression")
```

# After a run

`03_analyses/DEG_lists/DEG_counts.csv` holds the DEG counts per contrast (`n_DEG`, `n_up`,
`n_down`); `DEG_join_summary.csv` the annotated / unannotated split for the TC contrasts. The
contrast table and apeglm tables feed `07_enrichment`; `GOterms_genome/` and `top_DEGs/` feed
`08_gene-annotation`; the count matrix, sample table, TC DEG lists and `GOterms_genome/` feed
`09_gene-mechanics-correlation`. Run those after any change here (the repository-level
`00_run_pipeline.Rmd` does it in order).
