---
title: "01 - Clean gene count matrix and sample table"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# What this does

Starts the differential-expression analysis from the prepDE gene count matrix written by
`05_sequence-alignment/01_code/02_prepDE.Rmd`, and the raw sample sheet in `02_data/`:

1. renames the count-matrix columns from the read-file base name to the sample ID
   (`A10-T015F_S172` -> `T015F`), orders them, removes T051F and T051G (QC), and sorts the
   genes by ID. The original matrix came from `prepDE.py` under Python 2, whose dictionaries
   have no defined order, so its row order was arbitrary; sorting makes every later table
   reproducible. The counts are unchanged;
2. builds the sample table (`sample.1`, `tissue`, `treatment`, `day`, `region`) from the
   sample sheet; treatment and day come from `trt` (`T_OA_d3` -> OA, 3);
3. checks every library against the RNA isolation log and writes the crosswalk between the
   two naming schemes (below).

**Foot regions.** Two pieces of the foot were sequenced, and the sheets name them
inconsistently:

| library ID | `tissue` | `region` | sampled in | isolation log name (tissue label) |
|---|---|---|---|---|
| `T001F` | F | foot (phenol gland to tip) | every animal | `T01-F_PG` ("phenol gland") |
| `T001FX` | F | foot (without phenol gland) | the 12 day-0 animals only | `T01-F` ("foot") |
| `T001G` | G | gill | every animal | `T01-G` ("gill") |

Both are foot tissue. The sample sheet's free-text columns call the FX libraries "gill" and
"foot_control"; both are wrong and are not used. Every contrast uses the phenol-gland-to-tip
libraries, the region sampled in every animal; the FX libraries stay in the count matrix and
the sample table but enter no contrast (`02_define_contrasts.Rmd`), because the two regions
differ strongly in expression, byssal genes included.

Writes `03_analyses/count_matrix/gene_count_matrix_clean.csv`, `treatmentinfo_clean.csv` and
`library_crosswalk.csv`, which later steps read. Replaces `01_5-gene_count_matrix.Rmd`, which
could not run (it read a matrix that was not in the repository and its sample-table step was
never finished).

```{r paths}
source(here::here("01_code", "_paths.R"))
raw_counts <- file.path(repo_root, "05_sequence-alignment", "03_analyses", "prepDE", "gene_count_matrix.csv")
stopifnot(file.exists(raw_counts))
dir.create(cnt, showWarnings = FALSE, recursive = TRUE)
QC_REMOVED <- c("T051F", "T051G")
```

# Count matrix

```{r counts}
gcm <- read.csv(raw_counts, check.names = FALSE, row.names = 1)
## read-file base name -> sample ID: drop the plate well ("A10-") and the sequencing index ("_S172")
ids <- sub("^[A-H][0-9]+-", "", sub("_S[0-9]+$", "", colnames(gcm)))
stopifnot(!anyDuplicated(ids), all(grepl("^T[0-9]{3}[A-Z]+$", ids)))
colnames(gcm) <- ids
gcm <- gcm[order(rownames(gcm), method = "radix"), setdiff(ids[order(ids, method = "radix")], QC_REMOVED)]
cat("genes:", nrow(gcm), "| samples:", ncol(gcm), "(", paste(QC_REMOVED, collapse = ", "), "removed )\n")
write.csv(gcm, file.path(cnt, "gene_count_matrix_clean.csv"), row.names = TRUE)
```

# Sample table

```{r samples}
sheet <- read.csv(file.path(dat, "PSMFC-mytilus-byssus-pilot-RNA-tagseq_raw.csv"))
sheet <- sheet[!is.na(sheet$trt) & sheet$trt != "", ]          # drop the sheet's blank rows
stopifnot(all(grepl("^T_[A-Za-z]+_d[0-9]+$", sheet$trt)))
## library-ID suffix -> tissue and foot region (see the table above)
suffix  <- sub("^T[0-9]+", "", sheet$sample.1)
regions <- c(F = "foot (phenol gland to tip)", FX = "foot (without phenol gland)", G = "gill")
stopifnot(all(suffix %in% names(regions)))
ti <- data.frame(sample.1  = sheet$sample.1,
                 tissue    = ifelse(suffix == "G", "G", "F"),
                 treatment = sub("^T_(.*)_d[0-9]+$", "\\1", sheet$trt),
                 day       = sub("^.*_d", "", sheet$trt),
                 region    = unname(regions[suffix]))

## T047F has no library; T047G does. The count matrix is the source of which samples exist.
not_sequenced <- setdiff(ti$sample.1, c(colnames(gcm), QC_REMOVED))
cat("in the sample sheet but not in the count matrix:", paste(not_sequenced, collapse = ", "), "\n")
ti <- ti[match(colnames(gcm), ti$sample.1), ]
stopifnot(!anyNA(ti$sample.1))
rownames(ti) <- ti$sample.1
write.csv(ti, file.path(cnt, "treatmentinfo_clean.csv"), row.names = TRUE)
knitr::kable(as.data.frame(table(paste(ti$treatment, "day", ti$day), ti$region)) |>
               tidyr::pivot_wider(names_from = Var2, values_from = Freq), caption = "samples by group and region")
```

# Crosswalk to the RNA isolation log

The isolation log (`02_data/psmfc_mussel_rna_summary.csv`) names the phenol-gland-to-tip
extractions `T01-F_PG` and the rest of the foot `T01-F`, so its "F" is the library suffix
"FX", not "F". Each library is matched to its isolation row on the mussel tag, RNA
concentration, volume and yield, which are copied between the two sheets, and the match
must agree with the region assigned above.

```{r crosswalk}
iso <- read.csv(file.path(dat, "psmfc_mussel_rna_summary.csv"), check.names = FALSE)
tag <- function(x) sprintf("T%03d", as.integer(sub("^T0*([0-9]+).*$", "\\1", x)))
key <- function(mussel, conc, vol, yield) paste(mussel, conc, vol, yield)
iso_key <- key(tag(iso$sample), iso$`concentration(ng/uL)`, iso$`volume(uL)`, iso$`yield(ng)`)
lib_key <- key(tag(sheet$sample.1), sheet$conc_ng_uL.1, sheet$vol_uL, sheet$yield_ng)
stopifnot(!anyDuplicated(iso_key))
hit <- match(lib_key, iso_key)
xw <- data.frame(sample.1 = sheet$sample.1, region = unname(regions[suffix]),
                 in_count_matrix = ifelse(sheet$sample.1 %in% colnames(gcm), "yes",
                                          ifelse(sheet$sample.1 %in% QC_REMOVED, "removed at QC", "not sequenced")),
                 isolation_log_sample = iso$sample[hit], isolation_log_tissue = iso$tissue[hit],
                 isolation_date = iso$isolation_date[hit])
expected_label <- c("foot (phenol gland to tip)" = "phenol gland", "foot (without phenol gland)" = "foot",
                    gill = "gill")
bad <- xw[is.na(hit) | expected_label[xw$region] != xw$isolation_log_tissue, ]
if (nrow(bad)) { print(bad); stop(nrow(bad), " library(ies) do not match the isolation log as expected.") }
xw <- xw[order(xw$sample.1, method = "radix"), ]
write.csv(xw, file.path(cnt, "library_crosswalk.csv"), row.names = FALSE)
knitr::kable(table(region = xw$region, isolation_log_tissue = xw$isolation_log_tissue))
```
