---
title: "02 - Define the differential-expression contrasts"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# What this does

Defines every pairwise DESeq2 contrast from the sample table written by script 01, so the
sample set behind each result is set by rule in one place. (Before, the TC sample lists were
written only by two scripts kept as superseded.) Three families:

| family | contrast | samples (one tissue unless stated) | design | reference |
|---|---|---|---|---|
| **TC, of record** | `<T><X>_TC`, X = OA, OW, DO | stressor X at day 3 + control at day 3 | `~ treatment` | `control` |
| LC | `<T><X>_LC`, X = OA, OW, DO | stressor X at day 3 + control at day 0 | `~ treatment` | `control` |
| LC | `<T>TC_LC` | control at day 3 + control at day 0 | `~ day` | `0` |
| FG | `FG_TC`, `FG_LC` | foot + gill controls at day 3 (TC) or day 0 (LC) | `~ tissue` | `F` (foot) |

T is `F` (foot) or `G` (gill). Foot means the phenol gland to the tip of the foot, the region
sampled in every animal. The twelve day-0 libraries of the rest of the foot (IDs ending `FX`,
region "foot (without phenol gland)") are in no contrast: there is no day-3 counterpart, and
pooling the two regions would add a large region effect to the day-0 group. Samples are
listed in sample-ID order. TC is the control of record for the stressor effect; LC contrasts
carry the stressor plus three days in the system and handling; FG compares the tissues
within one control group.

Writes `03_analyses/DEG_lists/contrasts.csv` (one row per contrast) and
`contrast_samples.csv` (one row per contrast x sample), read by script 03.

```{r paths}
source(here::here("01_code", "_paths.R"))
ti <- read.csv(file.path(cnt, "treatmentinfo_clean.csv"), row.names = 1, colClasses = "character")
stopifnot(all(ti$region %in% c("foot (phenol gland to tip)", "foot (without phenol gland)", "gill")))
ti <- ti[ti$region != "foot (without phenol gland)", ]     # FX libraries: no contrast uses them
```

# Contrasts

```{r define}
tissues   <- c(F = "Foot", G = "Gill")
stressors <- c("OA", "OW", "DO")
defs <- list()
add <- function(code, family, dir, design_var, ref, test, keep, description)
  defs[[code]] <<- list(code = code, family = family, dir = dir, design_var = design_var,
                        ref_level = ref, test_level = test, keep = keep, description = description)

for (t in names(tissues)) {
  for (x in stressors) {
    add(paste0(t, x, "_TC"), "TC", tissues[[t]], "treatment", "control", x,
        with(ti, tissue == t & day == "3" & treatment %in% c("control", x)),
        sprintf("%s: %s vs day-3 treatment control", tissues[[t]], x))
    add(paste0(t, x, "_LC"), "LC", tissues[[t]], "treatment", "control", x,
        with(ti, tissue == t & ((treatment == x & day == "3") | (treatment == "control" & day == "0"))),
        sprintf("%s: %s vs day-0 lab control", tissues[[t]], x))
  }
  add(paste0(t, "TC_LC"), "LC", tissues[[t]], "day", "0", "3",
      with(ti, tissue == t & treatment == "control"),
      sprintf("%s: day-3 treatment control vs day-0 lab control", tissues[[t]]))
}
for (fam in c("TC", "LC")) {
  d <- if (fam == "TC") "3" else "0"
  add(paste0("FG_", fam), "FG", "Foot_vs_Gill", "tissue", "F", "G",
      with(ti, tissue %in% c("F", "G") & treatment == "control" & day == d),
      sprintf("Gill vs foot, %s controls", if (fam == "TC") "day-3 treatment" else "day-0 lab"))
}

contrast_samples <- do.call(rbind, lapply(defs, function(d) {
  s <- ti[d$keep, ]
  s <- s[order(s$sample.1, method = "radix"), ]
  data.frame(code = d$code, sample.1 = s$sample.1, tissue = s$tissue, treatment = s$treatment,
             day = s$day, group = ifelse(s[[d$design_var]] == d$test_level, "test", "reference"))
}))
contrasts <- do.call(rbind, lapply(defs, function(d) {
  s <- contrast_samples[contrast_samples$code == d$code, ]
  data.frame(code = d$code, family = d$family, dir = d$dir, design_var = d$design_var,
             ref_level = d$ref_level, test_level = d$test_level,
             n_reference = sum(s$group == "reference"), n_test = sum(s$group == "test"),
             description = d$description)
}))
rownames(contrasts) <- rownames(contrast_samples) <- NULL

## every contrast has two groups, and the levels it names are the ones its samples carry
stopifnot(all(contrasts$n_reference > 0), all(contrasts$n_test > 0),
          !anyDuplicated(paste(contrast_samples$code, contrast_samples$sample.1)))
write.csv(contrasts, file.path(deg, "contrasts.csv"), row.names = FALSE)
write.csv(contrast_samples, file.path(deg, "contrast_samples.csv"), row.names = FALSE)
knitr::kable(contrasts[, c("code", "family", "design_var", "ref_level", "test_level", "n_reference", "n_test", "description")])
```
