---
title: "03 - Fit DESeq2 to each contrast"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5)
```

# What this does

Fits one DESeq2 model per contrast defined by script 02 and saves the fitted object, so the
shrinkage step (script 04) runs on its own instead of needing an interactive session. For each
contrast, exactly as the original analysis did:

1. subset the clean count matrix to the contrast's samples, leaving out the mitochondrial
   loci listed by script 01 (the mitogenome genes and their nuclear copies, analysed on their
   own in script 13), so the genome results and the GO enrichment built on them carry one
   nuclear gene per row;
2. set the design variable's levels explicitly (reference first) and fit `~ <variable>`;
3. check that the second coefficient is `<variable>_<test>_vs_<reference>`, which fixes the
   sign of every log2 fold change;
4. keep genes with at least 10 counts in at least a third of the contrast's samples (applied
   after fitting, as before).

Writes `03_analyses/dds/<contrast>.rds` (the filtered `DESeqDataSet`; git-ignored, rebuilt by
this script), `03_analyses/DEG_lists/filter_summary.csv`, and PCA plots of the variance-
stabilised counts to `03_analyses/figures/`.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({ library(DESeq2); library(ggplot2) })
gcm <- as.matrix(read.csv(file.path(cnt, "gene_count_matrix_clean.csv"), row.names = 1, check.names = FALSE))
mtl <- read.csv(file.path(cnt, "mitochondrial_loci.csv"), stringsAsFactors = FALSE)
stopifnot(all(mtl$gene %in% rownames(gcm)))
gcm <- gcm[!rownames(gcm) %in% mtl$gene, ]                 # nuclear genes only (see step 1 above)
cat("mitochondrial rows left out:", nrow(mtl), "| genes kept:", nrow(gcm), "\n")
ti  <- read.csv(file.path(cnt, "treatmentinfo_clean.csv"), row.names = 1, colClasses = "character")
contrasts <- read.csv(file.path(deg, "contrasts.csv"), colClasses = "character")
cs        <- read.csv(file.path(deg, "contrast_samples.csv"), colClasses = "character")
stopifnot(all(cs$sample.1 %in% colnames(gcm)))
```

# Fit each contrast

```{r fit}
fit_one <- function(row) {
  smp <- cs$sample.1[cs$code == row$code]
  cd  <- ti[smp, c("tissue", "treatment", "day")]
  cd[[row$design_var]] <- factor(cd[[row$design_var]], levels = c(row$ref_level, row$test_level))
  stopifnot(!anyNA(cd[[row$design_var]]))
  dds <- DESeqDataSetFromMatrix(gcm[, smp], cd, reformulate(row$design_var))
  dds <- DESeq(dds, quiet = TRUE)
  coef_name <- sprintf("%s_%s_vs_%s", row$design_var, row$test_level, row$ref_level)
  stopifnot(resultsNames(dds)[2] == coef_name)
  keep <- rowSums(counts(dds) >= 10) >= ncol(dds) / 3
  saveRDS(dds[keep, ], file.path(dds_dir, paste0(row$code, ".rds")))
  data.frame(code = row$code, family = row$family, coefficient = coef_name,
             n_samples = ncol(dds), genes_in_matrix = nrow(dds), genes_after_filter = sum(keep))
}
summary_tbl <- do.call(rbind, lapply(seq_len(nrow(contrasts)), function(i) fit_one(contrasts[i, ])))
write.csv(summary_tbl, file.path(deg, "filter_summary.csv"), row.names = FALSE)
knitr::kable(summary_tbl)
```

# PCA of the variance-stabilised counts

Blind variance-stabilising transform (`vst(blind = TRUE)`), top 500 most variable nuclear
genes, as `DESeq2::plotPCA`. The original analysis drew the day-3 panels. The all-sample panel shows
how tissue dominates the variation, which is why every contrast stays within one tissue, and
that the day-0 libraries of the rest of the foot (no phenol gland) separate from the
phenol-gland-to-tip libraries of the same animals. It is the only place the day-0 libraries
appear: they enter no contrast (scripts 01 and 02).

```{r pca}
pca_data <- function(samples) {
  cd <- ti[samples, ]
  dds <- DESeqDataSetFromMatrix(gcm[, samples], cd, ~ 1)
  v <- vst(dds, blind = TRUE)
  d <- plotPCA(v, intgroup = c("tissue", "region", "treatment", "day"), returnData = TRUE)
  d$treatment <- factor(d$treatment, levels = names(TREATMENT_COLORS))
  list(data = d, var = round(100 * attr(d, "percentVar")))
}
axis_labs <- function(p) labs(x = sprintf("PC1 (%d%% of variance)", p$var[1]),
                              y = sprintf("PC2 (%d%% of variance)", p$var[2]))
save_fig <- function(plot, name, w = 7, h = 5) {
  ggsave(file.path(figs, name), plot, width = w, height = h, dpi = 300); print(plot)
}

## day 3, one tissue at a time: treatment by colour and shape
for (tcode in c("F", "G")) {
  t <- TISSUE_LABELS[[tcode]]
  p <- pca_data(rownames(ti)[ti$tissue == tcode & ti$day == "3"])
  save_fig(ggplot(p$data, aes(PC1, PC2, colour = treatment, shape = treatment)) +
             geom_point(size = 2.6, alpha = 0.9) +
             scale_colour_manual(values = TREATMENT_COLORS, labels = TREATMENT_LABELS, name = NULL) +
             scale_shape_manual(values = TREATMENT_SHAPES, labels = TREATMENT_LABELS, name = NULL) +
             axis_labs(p) + labs(title = sprintf("%s, day 3", t)) + theme_psmfc(),
           sprintf("PCA_%s_day3.png", tolower(t)))
}

## day 3, both tissues: treatment by colour, tissue by shape
p <- pca_data(rownames(ti)[ti$tissue %in% c("F", "G") & ti$day == "3"])
save_fig(ggplot(p$data, aes(PC1, PC2, colour = treatment, shape = tissue)) +
           geom_point(size = 2.6, alpha = 0.9) +
           scale_colour_manual(values = TREATMENT_COLORS, labels = TREATMENT_LABELS, name = NULL) +
           scale_shape_manual(values = c(F = 16, G = 17), labels = TISSUE_LABELS, name = NULL) +
           axis_labs(p) + labs(title = "Foot and gill, day 3") + theme_psmfc(),
         "PCA_both_tissues_day3.png")

## every sample: sampled region by colour, day by shape
p <- pca_data(rownames(ti))
p$data$day <- factor(paste("day", p$data$day))
p$data$region <- factor(p$data$region, levels = names(REGION_COLORS))
save_fig(ggplot(p$data, aes(PC1, PC2, colour = region, shape = day)) +
           geom_point(size = 2.4, alpha = 0.85) +
           scale_colour_manual(values = REGION_COLORS, labels = REGION_LABELS, name = NULL) +
           scale_shape_manual(values = c("day 0" = 1, "day 3" = 16), name = NULL) +
           axis_labs(p) + labs(title = "All samples") + theme_psmfc(),
         "PCA_all_samples.png")
```
