---
title: "04 - Log fold change shrinkage and DEG lists"
output: html_document
params:
  padj_cutoff: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

For every contrast fitted by script 03, reads the saved `DESeqDataSet`
(`03_analyses/dds/<contrast>.rds`) and, as the original TC shrinkage script did:

- shrinks the log2 fold changes with `apeglm` (the estimates of record), and with `normal` and
  `ashr` for comparison;
- writes the full apeglm table, `<contrast>_apeglm.csv`, and the DEG list,
  `<contrast>_siggene.csv` (apeglm `padj < 0.05`);
- writes how many genes survived each step, `<contrast>_filter_counts.csv`, and MA plots of the
  three shrinkage methods, `<contrast>_MA_plots.pdf`.

Files go to `03_analyses/DEG_lists/<Foot|Gill|Foot_vs_Gill>/`, all comma-separated, plus a
summary of DEG counts by direction, `03_analyses/DEG_lists/DEG_counts.csv`. Previously this
step only worked inside an interactive session that had just run the DESeq2 scripts; it now
reads the objects script 03 saves.

```{r paths}
source(here::here("01_code", "_paths.R"))
suppressPackageStartupMessages({ library(DESeq2); library(apeglm); library(ashr) })
contrasts <- read.csv(file.path(deg, "contrasts.csv"), colClasses = "character")
summary_tbl <- read.csv(file.path(deg, "filter_summary.csv"))
```

# Shrink, filter, write

```{r shrink}
as_table <- function(res) data.frame(gene = rownames(res), as.data.frame(res), row.names = NULL)

shrink_one <- function(row) {
  ddsf <- readRDS(file.path(dds_dir, paste0(row$code, ".rds")))
  out  <- file.path(deg, row$dir); dir.create(out, showWarnings = FALSE)
  res  <- lapply(c(normal = "normal", apeglm = "apeglm", ashr = "ashr"),
                 function(type) lfcShrink(ddsf, coef = 2, type = type, quiet = TRUE))
  full <- as_table(res$apeglm)
  sig  <- full[!is.na(full$padj) & full$padj < params$padj_cutoff, ]
  write.csv(full, file.path(out, paste0(row$code, "_apeglm.csv")), row.names = FALSE)
  write.csv(sig,  file.path(out, paste0(row$code, "_siggene.csv")), row.names = FALSE)

  n_sig <- function(r) sum(!is.na(r$padj) & r$padj < params$padj_cutoff)
  s <- summary_tbl[summary_tbl$code == row$code, ]
  write.csv(data.frame(step = c("genes_in_matrix", "genes_after_filtering", "genes_dropped",
                                "results_normal", "results_apeglm", "results_ashr",
                                "DEG_apeglm_padj", "DEG_ashr_padj", "DEG_normal_padj"),
                       count = c(s$genes_in_matrix, nrow(ddsf), s$genes_in_matrix - nrow(ddsf),
                                 nrow(res$normal), nrow(res$apeglm), nrow(res$ashr),
                                 n_sig(res$apeglm), n_sig(res$ashr), n_sig(res$normal))),
            file.path(out, paste0(row$code, "_filter_counts.csv")), row.names = FALSE)

  pdf(file.path(out, paste0(row$code, "_MA_plots.pdf")), width = 10, height = 4)
  par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
  for (type in names(res)) {
    ## neutral greys: in this repository red and blue mean up and down, not significance
    plotMA(res[[type]], xlim = c(1, 1e5), ylim = c(-3, 3), main = paste(row$code, type), cex = 0.8,
           colNonSig = "grey70", colSig = "grey10", colLine = "grey40")
    abline(h = c(-1.5, 1.5), col = "grey40", lwd = 1.5, lty = 2)
  }
  dev.off()

  data.frame(code = row$code, family = row$family, tissue_dir = row$dir,
             description = row$description, genes_tested = nrow(full), n_DEG = nrow(sig),
             n_up = sum(sig$log2FoldChange > 0), n_down = sum(sig$log2FoldChange < 0))
}

deg_counts <- do.call(rbind, lapply(seq_len(nrow(contrasts)), function(i) shrink_one(contrasts[i, ])))
write.csv(deg_counts, file.path(deg, "DEG_counts.csv"), row.names = FALSE)
knitr::kable(deg_counts[, c("code", "family", "genes_tested", "n_DEG", "n_up", "n_down")])
```
