---
title: "11 - DEG figures for every contrast family (TC, LC, foot vs gill)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

Draws the results of all 16 contrasts from script 04, including the lab-control (LC) and
foot-vs-gill (FG) contrasts that had tables but no figures. Scripts 08-10 keep drawing the
manuscript figures for the six treatment-control (TC) contrasts.

1. `DEG_counts_all_contrasts.png`: up- and down-regulated DEGs per contrast, by family.
2. `volcano_<family>.png`: volcano plots (apeglm log2 fold change against -log10 adjusted p).
3. `TC_vs_LC_overlap.png` and `DEG_lists/DEG_overlap_TC_LC.csv`: for each tissue and
   stressor, how many DEGs are found against the day-3 control only (TC), against the day-0
   lab control only (LC), or both. LC lists carry the stressor effect plus three days in the
   system and handling, so most of their DEGs are not TC DEGs.

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(ggplot2) })
contrasts <- read.csv(file.path(deg, "contrasts.csv"), colClasses = "character")
counts    <- read.csv(file.path(deg, "DEG_counts.csv"))
PADJ <- 0.05
read_full <- function(code) {
  r <- contrasts[contrasts$code == code, ]
  d <- read.csv(file.path(deg, r$dir, paste0(code, "_apeglm.csv")))
  d$code <- code; d$family <- r$family
  d
}
label_of <- setNames(contrasts$description, contrasts$code)
save_fig <- function(plot, name, w, h) {
  ggsave(file.path(figs, name), plot, width = w, height = h, dpi = 300); print(plot)
}
```

# DEG counts

```{r counts, fig.width = 10, fig.height = 5}
cnt_long <- counts %>%
  mutate(family = factor(family, levels = c("TC", "LC", "FG")),
         code = factor(code, levels = rev(contrasts$code))) %>%
  pivot_longer(c(n_up, n_down), names_to = "direction", values_to = "n") %>%
  mutate(direction = ifelse(direction == "n_up", "up", "down"),
         signed = ifelse(direction == "up", n, -n))
save_fig(ggplot(cnt_long, aes(signed, code, fill = direction)) +
           geom_col(width = 0.7) +
           geom_text(aes(label = n, hjust = ifelse(signed > 0, -0.15, 1.15)), size = 3) +
           geom_vline(xintercept = 0, colour = "grey40", linewidth = 0.3) +
           facet_grid(family ~ ., scales = "free_y", space = "free_y") +
           scale_fill_manual(values = DIRECTION_COLORS[c("up", "down")],
                             labels = c(up = "Up-regulated", down = "Down-regulated"), name = NULL) +
           scale_x_continuous(labels = abs, expand = expansion(mult = 0.15)) +
           labs(x = "DEGs (apeglm padj < 0.05)", y = NULL,
                caption = "TC: vs day-3 treatment control (of record). LC: vs day-0 lab control. FG: gill vs foot within controls.") +
           theme_psmfc() + theme(legend.position = "top"),
         "DEG_counts_all_contrasts.png", 9, 6.5)
```

# Volcano plots

```{r volcano, fig.width = 11, fig.height = 7}
for (fam in c("TC", "LC", "FG")) {
  d <- bind_rows(lapply(contrasts$code[contrasts$family == fam], read_full)) %>%
    filter(!is.na(padj)) %>%
    mutate(status = case_when(padj < PADJ & log2FoldChange > 0 ~ "up",
                              padj < PADJ & log2FoldChange < 0 ~ "down", TRUE ~ "ns"),
           panel = factor(sprintf("%s\n%s", code, label_of[code]),
                          levels = sprintf("%s\n%s", contrasts$code[contrasts$family == fam],
                                           label_of[contrasts$code[contrasts$family == fam]])))
  n_lab <- d %>% group_by(panel) %>% summarise(lab = sprintf("up %d / down %d", sum(status == "up"), sum(status == "down")))
  ncol_f <- if (fam == "FG") 2 else 4
  p <- ggplot(d, aes(log2FoldChange, -log10(padj), colour = status)) +
    geom_point(data = filter(d, status == "ns"), size = 0.5, alpha = 0.4) +
    geom_point(data = filter(d, status != "ns"), size = 0.7, alpha = 0.7) +
    geom_hline(yintercept = -log10(PADJ), linetype = "dashed", linewidth = 0.3, colour = "grey40") +
    geom_text(data = n_lab, aes(x = -Inf, y = Inf, label = lab), inherit.aes = FALSE,
              hjust = -0.05, vjust = 1.4, size = 3) +
    facet_wrap(~ panel, ncol = ncol_f, scales = "free") +
    scale_colour_manual(values = DIRECTION_COLORS,
                        labels = c(up = "Up", down = "Down", ns = "Not significant"), name = NULL) +
    labs(x = "log2 fold change (apeglm)", y = "-log10 adjusted p") +
    theme_psmfc(10) + theme(legend.position = "top")
  save_fig(p, sprintf("volcano_%s.png", fam), if (fam == "FG") 8 else 12, if (fam == "FG") 4.5 else if (fam == "LC") 7 else 7)
}
```

# Treatment control vs lab control

```{r overlap, fig.width = 9, fig.height = 4.5}
sig_genes <- function(code) { d <- read_full(code); d$gene[!is.na(d$padj) & d$padj < PADJ] }
ov <- bind_rows(lapply(c("F", "G"), function(t) bind_rows(lapply(c("OA", "OW", "DO"), function(x) {
  tc <- sig_genes(paste0(t, x, "_TC")); lc <- sig_genes(paste0(t, x, "_LC"))
  data.frame(tissue = TISSUE_LABELS[[t]], stressor = x,
             TC_only = length(setdiff(tc, lc)), both = length(intersect(tc, lc)), LC_only = length(setdiff(lc, tc)),
             share_of_TC_also_LC = round(length(intersect(tc, lc)) / max(length(tc), 1), 3))
}))))
write.csv(ov, file.path(deg, "DEG_overlap_TC_LC.csv"), row.names = FALSE)
knitr::kable(ov)
ov_long <- ov %>% pivot_longer(c(TC_only, both, LC_only), names_to = "set", values_to = "n") %>%
  mutate(set = factor(set, levels = c("TC_only", "both", "LC_only")),
         stressor = factor(stressor, levels = c("OA", "OW", "DO")))
save_fig(ggplot(ov_long, aes(stressor, n, fill = set)) +
           geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
           facet_wrap(~ tissue, scales = "free_y") +
           scale_fill_manual(values = c(TC_only = "#4d4d4d", both = "#8c8c8c", LC_only = "#cccccc"),   # neutral: red/blue mean up/down
                             labels = c(TC_only = "vs day-3 control only (TC)", both = "both",
                                        LC_only = "vs day-0 lab control only (LC)"), name = NULL) +
           labs(x = NULL, y = "DEGs (padj < 0.05)") + theme_psmfc() + theme(legend.position = "top"),
         "TC_vs_LC_overlap.png", 8.5, 4.5)
```
