---
title: "13 - Mitochondrial gene expression, analysed on its own"
output: html_document
params:
  padj_cutoff: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

The genome analysis (scripts 03-12) leaves out the mitochondrial loci listed by script 01.
This script tests them on their own, for the six treatment-control (TC) contrasts of record.

**What is counted.** The reference holds the female-type mitochondrial genome (NC_007687.1)
with its 12 protein genes, and 293 loci on unplaced scaffolds that are copies of those genes:
126 protein-coding LOCs (same best UniProt hit, same identity and alignment length, about 23
identical copies each of CYTB, COX2, COX3, ND2 and ND6) and 167 pseudogenes named after them
(mostly ATP6, COX1, ND1, ND4 and CYTB). Reads from one mitochondrial transcript are split
between the mitogenome gene and its copies, in proportions that vary from sample to sample
(figure `MT_nuclear_copy_share.png`): the copies hold most of the reads of ATP6, CYTB, COX2,
ND2 and ND6. So each protein is counted once, as the sum of its mitogenome gene and its copies
(`mt_protein_counts.csv`), and `all_12` is the sum over the 12 proteins. The five mitochondrial
tRNA rows hold almost no reads (3' Tag-seq captures polyadenylated transcripts) and are not
tested.

The share of a sample's reads held by the copies varies continuously, with no split into two
groups of animals, which argues against the copies being the divergent male-type
mitochondrial genome that *Mytilus* males carry (doubly uniparental inheritance), whose
presence would follow the animal's sex.

**The test.** For each TC contrast, the contrast's samples (script 02) are fitted with DESeq2
exactly as the genome contrasts are (`~ treatment`, control as reference), on the nuclear
genes plus the 13 mitochondrial rows. The size factors come from the nuclear genes only, so
the mitochondrial rows cannot shift each other's normalisation, and their dispersions are
shrunk toward the transcriptome-wide trend. Adjusted p-values (`padj_mt`) are
Benjamini-Hochberg over the 12 proteins within each contrast; `all_12` is one pre-specified
summary endpoint and is read from its own p.

**Which fold change is drawn.** DESeq2's p-value is a Wald test of the unshrunk
(maximum-likelihood) log2 fold change; apeglm's shrunk estimate and its posterior standard
deviation are a different quantity (Zhu, Ibrahim and Love 2019, Bioinformatics 35:2084).
Drawing the apeglm estimate and interval beside the Wald p made five Foot OA proteins look
significant with intervals that crossed zero. The figure therefore shows the Wald estimate
with its 95% interval, which excludes zero whenever the unadjusted p is below 0.05 (and so
whenever `padj_mt` is); the table keeps the apeglm estimate as well (`lfc_apeglm`). The
genome DEG tables report apeglm fold changes with Wald p, the DESeq2 convention, but draw no
intervals.

The two estimates also differ where DESeq2 replaced an outlier count. With 7 or more samples
per group, `DESeq()` replaces a count whose Cook's distance is extreme and refits that gene;
the Wald estimate and p come from the refit, while `lfcShrink()` uses the original counts.
In Gill OA one library (T025G) holds about ten times the ND1 and ATP6 reads of the other OA
libraries, so the apeglm estimates of those two proteins (1.40 and 0.89) exceed the refit
Wald estimates (0.56 and 0.49). `outlier_replaced` marks these rows; the figure shows the
refit estimate, which matches its p-value.

Writes, in `03_analyses/mitochondrial/`:

- `mt_protein_counts.csv`: per sample, the summed count of each protein and of all 12;
- `mt_de_TC.csv`: per TC contrast and protein, the base mean, the Wald log2 fold change with
  its standard error and 95% interval, the apeglm log2 fold change and its posterior SD, Wald
  p, `padj_mt` and the reads' share in the nuclear copies;
- `mt_share_by_sample.csv`: per sample, mitochondrial protein reads as a percentage of its
  nuclear reads, and the share held by the copies;

and, in `03_analyses/figures/`, `MT_mitochondrial_expression.png` (the figure for the
manuscript) and `MT_nuclear_copy_share.png` (why the copies are summed).

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
suppressPackageStartupMessages({
  library(DESeq2); library(apeglm); library(dplyr); library(tidyr); library(ggplot2); library(patchwork)
})
gcm <- as.matrix(read.csv(file.path(cnt, "gene_count_matrix_clean.csv"), row.names = 1, check.names = FALSE))
ti  <- read.csv(file.path(cnt, "treatmentinfo_clean.csv"), row.names = 1, colClasses = "character")
mtl <- read.csv(file.path(cnt, "mitochondrial_loci.csv"), stringsAsFactors = FALSE)
contrasts <- read.csv(file.path(deg, "contrasts.csv"), colClasses = "character")
cs        <- read.csv(file.path(deg, "contrast_samples.csv"), colClasses = "character")
tc <- contrasts[contrasts$family == "TC", ]
stopifnot(nrow(tc) == 6)

## mitochondrial proteins in respiratory-complex order
PROTEINS <- c("ND1", "ND2", "ND3", "ND4", "ND4L", "ND5", "ND6", "CYTB", "COX1", "COX2", "COX3", "ATP6")
COMPLEX  <- c(ND1 = "Complex I", ND2 = "Complex I", ND3 = "Complex I", ND4 = "Complex I",
              ND4L = "Complex I", ND5 = "Complex I", ND6 = "Complex I", CYTB = "III",
              COX1 = "Complex IV", COX2 = "Complex IV", COX3 = "Complex IV", ATP6 = "V",
              all_12 = "All")
stopifnot(setequal(PROTEINS, unique(mtl$mt_protein[mtl$mt_protein != "tRNA/rRNA"])))
```

# Summed counts per protein

```{r sums}
nuclear <- gcm[!rownames(gcm) %in% mtl$gene, ]
mt_sum <- t(sapply(PROTEINS, function(p) colSums(gcm[mtl$gene[mtl$mt_protein == p], , drop = FALSE])))
mt_sum <- rbind(mt_sum, all_12 = colSums(mt_sum))
copy_sum <- t(sapply(PROTEINS, function(p)
  colSums(gcm[mtl$gene[mtl$mt_protein == p & mtl$source == "nuclear copy"], , drop = FALSE])))
copy_sum <- rbind(copy_sum, all_12 = colSums(copy_sum))
write.csv(mt_sum, file.path(mito, "mt_protein_counts.csv"))

share <- data.frame(sample.1 = colnames(gcm), ti[colnames(gcm), c("tissue", "treatment", "day", "region")],
                    mt_reads_pct_of_nuclear = 100 * mt_sum["all_12", ] / colSums(nuclear),
                    copy_share = copy_sum["all_12", ] / mt_sum["all_12", ], row.names = NULL)
write.csv(share, file.path(mito, "mt_share_by_sample.csv"), row.names = FALSE)
knitr::kable(share %>% group_by(region, day) %>%
               summarise(samples = n(), median_mt_pct = median(mt_reads_pct_of_nuclear),
                         median_copy_share = median(copy_share), .groups = "drop"), digits = 3)
```

# DESeq2 per TC contrast

```{r fit}
fit_mt <- function(row) {
  smp <- cs$sample.1[cs$code == row$code]
  cd  <- ti[smp, c("tissue", "treatment", "day")]
  cd$treatment <- factor(cd$treatment, levels = c(row$ref_level, row$test_level))
  m   <- rbind(nuclear[, smp], mt_sum[, smp])
  dds <- DESeqDataSetFromMatrix(m, cd, ~ treatment)
  sizeFactors(dds) <- estimateSizeFactorsForMatrix(nuclear[, smp])   # nuclear genes only
  dds <- DESeq(dds, quiet = TRUE)
  stopifnot(resultsNames(dds)[2] == sprintf("treatment_%s_vs_%s", row$test_level, row$ref_level))
  keep <- rowSums(counts(dds) >= 10) >= ncol(dds) / 3                 # the genome filter (script 03)
  keep[rownames(mt_sum)] <- TRUE                                      # every protein is reported
  dds <- dds[keep, ]
  wald <- results(dds, name = resultsNames(dds)[2])
  shr  <- lfcShrink(dds, coef = 2, type = "apeglm", quiet = TRUE)
  rows <- rownames(mt_sum)
  passes <- rowSums(counts(dds)[rows, ] >= 10) >= ncol(dds) / 3
  data.frame(code = row$code, tissue = row$dir, stressor = row$test_level, protein = rows,
             complex = unname(COMPLEX[rows]), baseMean = wald[rows, "baseMean"],
             log2FoldChange = wald[rows, "log2FoldChange"], lfcSE = wald[rows, "lfcSE"],   # Wald (MLE)
             ci_low = wald[rows, "log2FoldChange"] - qnorm(0.975) * wald[rows, "lfcSE"],
             ci_high = wald[rows, "log2FoldChange"] + qnorm(0.975) * wald[rows, "lfcSE"],
             lfc_apeglm = shr[rows, "log2FoldChange"], lfc_apeglm_sd = shr[rows, "lfcSE"],
             pvalue = wald[rows, "pvalue"], passes_count_filter = passes,
             outlier_replaced = if ("replace" %in% names(mcols(dds))) mcols(dds)[rows, "replace"] %in% TRUE else FALSE,
             copy_share = rowSums(copy_sum[rows, smp]) / rowSums(mt_sum[rows, smp]),
             row.names = NULL)
}
de <- bind_rows(lapply(seq_len(nrow(tc)), function(i) fit_mt(tc[i, ]))) %>%
  group_by(code) %>%
  mutate(padj_mt = ifelse(protein == "all_12", NA_real_,
                          p.adjust(ifelse(protein == "all_12", NA, pvalue), "BH"))) %>%
  ungroup() %>%
  mutate(significant = ifelse(protein == "all_12", pvalue < params$padj_cutoff, padj_mt < params$padj_cutoff))
## a point drawn as significant never has an interval that crosses zero
stopifnot(with(de, all(!significant | is.na(significant) | sign(ci_low) == sign(ci_high))))
write.csv(de, file.path(mito, "mt_de_TC.csv"), row.names = FALSE)
knitr::kable(de %>% select(code, protein, baseMean, log2FoldChange, lfcSE, lfc_apeglm, pvalue, padj_mt, copy_share) %>%
               mutate(across(c(baseMean, log2FoldChange, lfcSE, lfc_apeglm, copy_share), ~ round(.x, 2)),
                      across(c(pvalue, padj_mt), ~ signif(.x, 2))))
cat("Proteins with padj_mt <", params$padj_cutoff, "per contrast:\n")
print(de %>% filter(protein != "all_12") %>% group_by(code) %>%
        summarise(up = sum(significant & log2FoldChange > 0), down = sum(significant & log2FoldChange < 0),
                  .groups = "drop"))
```

# Figure for the manuscript

```{r figure, fig.width = 11, fig.height = 8.5}
d <- de %>%
  mutate(protein = factor(ifelse(protein == "all_12", "All 12", protein), levels = c(PROTEINS, "All 12")),
         complex = factor(complex, levels = c("Complex I", "III", "Complex IV", "V", "All")),
         tissue = factor(tissue, levels = c("Foot", "Gill")),
         stressor = factor(stressor, levels = c("OA", "OW", "DO")),
         fill_key = ifelse(significant, as.character(stressor), "ns"))   # open point = not significant
dodge <- position_dodge(width = 0.6)
pa <- ggplot(d, aes(protein, log2FoldChange, colour = stressor, group = stressor)) +   # group: dodge points and bars together
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, linewidth = 0.6, position = dodge) +
  geom_point(aes(fill = fill_key), shape = 21, size = 2.6, stroke = 0.9, position = dodge) +
  facet_grid(tissue ~ complex, scales = "free_x", space = "free_x") +
  scale_colour_manual(values = TREATMENT_COLORS, labels = TREATMENT_LABELS, name = NULL,
                      guide = guide_legend(override.aes = list(fill = TREATMENT_COLORS[c("OA", "OW", "DO")]))) +
  scale_fill_manual(values = c(TREATMENT_COLORS[c("OA", "OW", "DO")], ns = "white"), guide = "none") +
  labs(x = NULL, y = "log2 fold change vs day-3 control",
       title = "A. Mitochondrial protein genes, each stressor against the day-3 control") +
  theme_psmfc(10) + theme(legend.position = "top", panel.grid.major.x = element_blank())

pb <- ggplot(share %>% filter(day == "3") %>%
               mutate(treatment = factor(treatment, levels = names(TREATMENT_COLORS)),
                      tissue = factor(TISSUE_LABELS[tissue], levels = TISSUE_LABELS)),
             aes(treatment, mt_reads_pct_of_nuclear)) +
  geom_boxplot(aes(fill = treatment), alpha = 0.35, outlier.shape = NA, width = 0.6, colour = "grey30") +
  geom_point(aes(colour = treatment, shape = treatment), position = position_jitter(width = 0.12, seed = 1),
             size = 1.8) +
  facet_wrap(~ tissue, scales = "free_y") +
  scale_fill_manual(values = TREATMENT_COLORS, guide = "none") +
  scale_colour_manual(values = TREATMENT_COLORS, guide = "none") +
  scale_shape_manual(values = TREATMENT_SHAPES, guide = "none") +
  scale_x_discrete(labels = c(control = "Control", OA = "OA", OW = "OW", DO = "DO")) +
  labs(x = NULL, y = "mitochondrial protein reads\n(% of nuclear reads)",
       title = "B. Mitochondrial share of each library, day 3") +
  theme_psmfc(10)

p <- pa / pb + plot_layout(heights = c(1.6, 1)) +
  plot_annotation(caption = paste(strwrap(paste(
    "Each protein is the sum of its mitochondrial-genome gene and its nuclear-scaffold copies; All 12 sums the",
    "proteins. A: DESeq2 log2 fold change (Wald, unshrunk) with 95% interval, from separate fits per contrast",
    "(nuclear-gene size factors); filled points: BH-adjusted Wald p < 0.05 over the 12 proteins (All 12: p < 0.05),",
    "open points: not significant. B: one point per library."), width = 130), collapse = "\n"),
    theme = theme(plot.caption = element_text(hjust = 0)))
ggsave(file.path(figs, "MT_mitochondrial_expression.png"), p, width = 11, height = 8.5, dpi = 300)
p
```

# Why the copies are summed

```{r copies, fig.width = 9, fig.height = 4.5}
cs_long <- as.data.frame(t(copy_sum[PROTEINS, ] / mt_sum[PROTEINS, ])) %>%
  mutate(tissue = TISSUE_LABELS[ti[colnames(gcm), "tissue"]]) %>%
  pivot_longer(-tissue, names_to = "protein", values_to = "copy_share") %>%
  mutate(protein = factor(protein, levels = PROTEINS), tissue = factor(tissue, levels = TISSUE_LABELS))
p2 <- ggplot(cs_long, aes(protein, 100 * copy_share)) +
  geom_boxplot(aes(fill = tissue), outlier.size = 0.6, width = 0.7, alpha = 0.6,
               position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = setNames(TISSUE_COLORS, TISSUE_LABELS), name = NULL) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = NULL, y = "reads on nuclear-scaffold copies (%)",
       title = "Share of each mitochondrial protein's reads that align to its nuclear-scaffold copies",
       caption = paste(strwrap(paste("One value per library (all samples). COX1, ND3, ND4 and ND4L have few or",
                                     "no copies; ATP6, CYTB, COX2, ND2 and ND6 lose most of their reads to them,",
                                     "in proportions that vary between libraries."), width = 110), collapse = "\n")) +
  theme_psmfc(10) + theme(legend.position = "top", plot.caption = element_text(hjust = 0))
ggsave(file.path(figs, "MT_nuclear_copy_share.png"), p2, width = 9, height = 4.5, dpi = 300)
p2
```
