---
title: "13 - Mitochondrial gene expression, analysed on its own"
output: html_document
params:
  padj_cutoff: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this does

The genome analysis (scripts 03-12) leaves out the mitochondrial loci listed by script 01.
This script tests the mitochondrial protein genes on their own, for the six treatment-control
(TC) contrasts of record.

**What is counted.** Each protein's reads aligned to the mitochondrial genome alone
(NC_007687.1, the female-type genome of *M. trossulus*), counted by
`05_sequence-alignment` step 05 (`03_analyses/mitogenome/mitogenome_gene_counts.csv`): HISAT2
with default scoring against the mitogenome, one featureCounts count per read; `all_12` is the
sum over the 12 proteins. The genome alignment does not serve: there the mitochondrial reads
compete with 293 annotated nuclear copies of the mitochondrial genes (126 protein-coding LOCs
and 167 pseudogenes on unplaced scaffolds; `tools/mt_encoded.R`) and with unannotated
mitochondrial-like sequence, HISAT2 reports at most five placements of a read, and
featureCounts, which counts the nuclear genes of record, leaves out reads with several
placements altogether. On the mitogenome alone each read has one place to go. (Until
2026-10-02 this script summed each protein's mitogenome gene and its copies in the StringTie +
prepDE matrix; that count lost about half of the reads, and more for some animals, below.)
The rRNA and tRNA rows are counted but not tested.

**Mitochondrial haplotypes.** Twelve of the 59 animals carry mitochondrial haplotypes that
differ from the reference at fixed positions: every read of the animal, in every tissue, has
the non-reference base. Their reads match the nuclear copies as well as the mitogenome, so in
the genome alignment they lose their unique placements: the share of an animal's COX1 reads
that align uniquely to the mitogenome's COX1 gene in the genome alignment (the record's count)
against its count on the mitogenome alone separates three groups (`mt_haplotypes.csv`): 0.25
to 0.61 in 47 animals (reference-like), about 0.05 in 8 (group A), and under 0.01 in 4 (group
B). In the genome alignment their COX1 count fell to less than half and the 12-protein sum to
about three quarters of the reference-like animals'; on the mitogenome alone the deficit
disappears. As a check, every contrast is also fitted with the haplotype group as a covariate
(`~ haplotype + treatment`), and with the counts of a permissive alignment score (-0.6 x read
length instead of -0.2, about six mismatches in a 63-base read instead of two), which places
more of the divergent reads (`mt_de_TC_sensitivity.csv`). The groups are variants of the
female-type genome: their reads align to it with at most two mismatches, which the male-type
genome that *Mytilus* males carry under doubly uniparental inheritance (Skibinski, Gallagher and
Beynon 1994, *Genetics* 138:801; Zouros et al. 1994, *PNAS* 91:7463), far more divergent, would
not allow.

**The test.** For each TC contrast, the contrast's samples (script 02) are fitted with DESeq2
exactly as the genome contrasts are (`~ treatment`, control as reference), on the nuclear
genes of the count matrix of record plus the 13 mitochondrial rows. The size factors come from
the nuclear genes only, so the mitochondrial rows cannot shift each other's normalisation, and
their dispersions are shrunk toward the transcriptome-wide trend. Adjusted p-values
(`padj_mt`) are Benjamini-Hochberg over the 12 proteins within each contrast; `all_12` is one
pre-specified summary endpoint and is read from its own p.

**Which fold change is drawn.** DESeq2's p-value is a Wald test of the unshrunk
(maximum-likelihood) log2 fold change; apeglm's shrunk estimate and its posterior standard
deviation are a different quantity (Zhu, Ibrahim and Love 2019, Bioinformatics 35:2084).
Drawing the apeglm estimate and interval beside the Wald p made five Foot OA proteins look
significant with intervals that crossed zero. The figure therefore shows the Wald estimate
with its 95% interval, which excludes zero whenever the unadjusted p is below 0.05 (and so
whenever `padj_mt` is); the table keeps the apeglm estimate as well (`lfc_apeglm`). The
genome DEG tables report apeglm fold changes with Wald p, the DESeq2 convention, but draw no
intervals.

The two estimates also differ where DESeq2 replaced an outlier count. With 7 or more samples
per group, `DESeq()` replaces a count whose Cook's distance is extreme and refits that gene;
the Wald estimate and p come from the refit, while `lfcShrink()` uses the original counts.
`outlier_replaced` marks such rows, and the figure shows the refit estimate, which matches its
p-value.

Writes, in `03_analyses/mitochondrial/`:

- `mt_protein_counts.csv`: per sample, the count of each protein and of all 12 on the
  mitogenome;
- `mt_de_TC.csv`: per TC contrast and protein, the base mean, the Wald log2 fold change with
  its standard error and 95% interval, the apeglm log2 fold change and its posterior SD, Wald
  p and `padj_mt`;
- `mt_de_TC_sensitivity.csv`: the same test with the haplotype group as a covariate and with
  the permissive alignment's counts (Wald log2 fold change, standard error, p, `padj_mt`);
- `mt_share_by_sample.csv`: per sample, mitochondrial protein reads as a percentage of its
  nuclear reads, and the animal's haplotype group;
- `mt_haplotypes.csv`: per animal, the COX1 index above and the group;
- `mt_counts_vs_previous.csv`: per protein and haplotype group, the median ratio of the count
  here to the previous count (mitogenome gene and copies summed in the StringTie + prepDE
  matrix);

and, in `03_analyses/figures/`, `MT_mitochondrial_expression.png` (the figure for the
manuscript) and `MT_haplotypes.png` (the groups).

```{r paths}
source(here::here("01_code", "_paths.R"))
source(file.path(repo_root, "tools", "plot_style.R"))
source(file.path(repo_root, "tools", "gene_ids.R"))                 # gene_key()
suppressPackageStartupMessages({
  library(DESeq2); library(apeglm); library(dplyr); library(tidyr); library(ggplot2); library(patchwork)
})
gcm <- as.matrix(read.csv(file.path(cnt, "gene_count_matrix_clean.csv"), row.names = 1, check.names = FALSE))
ti  <- read.csv(file.path(cnt, "treatmentinfo_clean.csv"), row.names = 1, colClasses = "character")
mtl <- read.csv(file.path(cnt, "mitochondrial_loci.csv"), stringsAsFactors = FALSE)
contrasts <- read.csv(file.path(deg, "contrasts.csv"), colClasses = "character")
cs        <- read.csv(file.path(deg, "contrast_samples.csv"), colClasses = "character")
tc <- contrasts[contrasts$family == "TC", ]
stopifnot(nrow(tc) == 6)

## mitochondrial proteins in respiratory-complex order
PROTEINS <- c("ND1", "ND2", "ND3", "ND4", "ND4L", "ND5", "ND6", "CYTB", "COX1", "COX2", "COX3", "ATP6")
COMPLEX  <- c(ND1 = "Complex I", ND2 = "Complex I", ND3 = "Complex I", ND4 = "Complex I",
              ND4L = "Complex I", ND5 = "Complex I", ND6 = "Complex I", CYTB = "III",
              COX1 = "Complex IV", COX2 = "Complex IV", COX3 = "Complex IV", ATP6 = "V",
              all_12 = "All")
stopifnot(setequal(PROTEINS, unique(mtl$mt_protein[mtl$mt_protein != "tRNA/rRNA"])))
```

# Counts per protein

```{r counts}
mito_counts <- file.path(repo_root, "05_sequence-alignment", "03_analyses", "mitogenome", "mitogenome_gene_counts.csv")
mito_perm   <- file.path(repo_root, "05_sequence-alignment", "03_analyses", "mitogenome", "mitogenome_gene_counts_permissive.csv")
read_mito <- function(f) {
  m <- as.matrix(read.csv(f, row.names = 1, check.names = FALSE))
  stopifnot(all(PROTEINS %in% rownames(m)), all(colnames(gcm) %in% colnames(m)))
  m <- round(m[PROTEINS, colnames(gcm)])                            # fractional shares of the few reads on two genes
  rbind(m, all_12 = colSums(m))
}
mt_sum  <- read_mito(mito_counts)
mt_perm <- read_mito(mito_perm)
nuclear <- gcm[!rownames(gcm) %in% mtl$gene, ]
write.csv(mt_sum, file.path(mito, "mt_protein_counts.csv"))
```

# Haplotype groups

```{r haplotypes, fig.width = 7, fig.height = 3.5}
## the record's count of reads aligned uniquely to the mitogenome's COX1 gene in the genome
## alignment, against the count on the mitogenome alone; per animal, the median of its libraries
cox1_genome <- gcm[mtl$gene[mtl$LOC_ID == "COX1" & mtl$source == "mitogenome"], colnames(gcm)]
hl <- data.frame(sample.1 = colnames(gcm), animal = substr(colnames(gcm), 1, 4),
                 index = cox1_genome / mt_sum["COX1", colnames(gcm)])
ha <- hl %>% group_by(animal) %>% summarise(libraries = n(), cox1_index = median(index),
                                            spread = max(index) - min(index), .groups = "drop") %>%
  mutate(haplotype = case_when(cox1_index >= 0.15 ~ "reference-like", cox1_index >= 0.02 ~ "group A", TRUE ~ "group B"))
stopifnot(max(ha$spread) < 0.15)                                      # every library of an animal agrees
write.csv(ha, file.path(mito, "mt_haplotypes.csv"), row.names = FALSE)
knitr::kable(ha %>% group_by(haplotype) %>% summarise(animals = n(), min = min(cox1_index), median = median(cox1_index),
                                                      max = max(cox1_index), .groups = "drop"), digits = 3)
ti$haplotype <- ha$haplotype[match(substr(rownames(ti), 1, 4), ha$animal)]
hp <- ggplot(hl %>% mutate(haplotype = factor(ti[sample.1, "haplotype"], levels = c("reference-like", "group A", "group B"))),
             aes(haplotype, index)) +
  geom_point(aes(shape = haplotype), position = position_jitter(width = 0.15, seed = 1), size = 1.6, colour = "grey30") +
  scale_y_continuous(trans = "log10") +
  scale_shape_manual(values = c(16, 17, 15), guide = "none") +
  labs(x = NULL, y = "COX1 reads: unique in the genome\nalignment / on the mitogenome",
       title = "Mitochondrial haplotype groups (one point per library)") +
  theme_psmfc(10)
ggsave(file.path(figs, "MT_haplotypes.png"), hp, width = 7, height = 3.5, dpi = 300)
hp

## against the previous count: each protein's mitogenome gene and nuclear copies summed in the
## StringTie + prepDE matrix of the HPC alignment (05 .../prepDE/), as this script counted until
## 2026-10-02; per protein and haplotype group, the median ratio over the libraries
prev <- read.csv(file.path(repo_root, "05_sequence-alignment", "03_analyses", "prepDE", "gene_count_matrix.csv"),
                 row.names = 1, check.names = FALSE)
colnames(prev) <- sub("^[A-H][0-9]+-", "", sub("_S[0-9]+$", "", colnames(prev)))
stopifnot(all(colnames(gcm) %in% colnames(prev)))
pkey <- gene_key(rownames(prev))
prev_sum <- t(sapply(PROTEINS, function(p) colSums(prev[pkey %in% mtl$LOC_ID[mtl$mt_protein == p], colnames(gcm), drop = FALSE])))
prev_sum <- rbind(prev_sum, all_12 = colSums(prev_sum))
vs <- as.data.frame(as.table(mt_sum / prev_sum[rownames(mt_sum), ]), stringsAsFactors = FALSE)
names(vs) <- c("protein", "sample.1", "ratio")
vs <- vs %>% mutate(haplotype = ti[sample.1, "haplotype"]) %>% filter(is.finite(ratio)) %>%
  group_by(protein, haplotype) %>% summarise(libraries = n(), median_ratio = median(ratio), .groups = "drop") %>%
  pivot_wider(names_from = haplotype, values_from = c(libraries, median_ratio)) %>%
  arrange(match(protein, c(PROTEINS, "all_12")))
write.csv(vs, file.path(mito, "mt_counts_vs_previous.csv"), row.names = FALSE)
knitr::kable(vs, digits = 2, caption = "count on the mitogenome / previous genome count (gene plus copies), median per library")

share <- data.frame(sample.1 = colnames(gcm), ti[colnames(gcm), c("tissue", "treatment", "day", "region", "haplotype")],
                    mt_reads_pct_of_nuclear = 100 * mt_sum["all_12", ] / colSums(nuclear), row.names = NULL)
write.csv(share, file.path(mito, "mt_share_by_sample.csv"), row.names = FALSE)
knitr::kable(share %>% group_by(region, day) %>%
               summarise(samples = n(), median_mt_pct = median(mt_reads_pct_of_nuclear), .groups = "drop"), digits = 3)
```

# DESeq2 per TC contrast

```{r fit}
fit_mt <- function(row, mt = mt_sum, design = ~ treatment, shrink = TRUE) {
  smp <- cs$sample.1[cs$code == row$code]
  cd  <- ti[smp, c("tissue", "treatment", "day", "haplotype")]
  cd$treatment <- factor(cd$treatment, levels = c(row$ref_level, row$test_level))
  cd$haplotype <- factor(cd$haplotype)
  m   <- rbind(nuclear[, smp], mt[, smp])
  dds <- DESeqDataSetFromMatrix(m, cd, design)
  sizeFactors(dds) <- estimateSizeFactorsForMatrix(nuclear[, smp])   # nuclear genes only
  dds <- DESeq(dds, quiet = TRUE)
  coef <- sprintf("treatment_%s_vs_%s", row$test_level, row$ref_level)
  stopifnot(coef %in% resultsNames(dds))
  keep <- rowSums(counts(dds) >= 10) >= ncol(dds) / 3                 # the genome filter (script 03)
  keep[rownames(mt)] <- TRUE                                          # every protein is reported
  dds <- dds[keep, ]
  wald <- results(dds, name = coef)
  rows <- rownames(mt)
  if (!shrink) return(data.frame(code = row$code, protein = rows, log2FoldChange = wald[rows, "log2FoldChange"],
                                 lfcSE = wald[rows, "lfcSE"], pvalue = wald[rows, "pvalue"], row.names = NULL))
  shr  <- lfcShrink(dds, coef = coef, type = "apeglm", quiet = TRUE)
  passes <- rowSums(counts(dds)[rows, ] >= 10) >= ncol(dds) / 3
  data.frame(code = row$code, tissue = row$dir, stressor = row$test_level, protein = rows,
             complex = unname(COMPLEX[rows]), baseMean = wald[rows, "baseMean"],
             log2FoldChange = wald[rows, "log2FoldChange"], lfcSE = wald[rows, "lfcSE"],   # Wald (MLE)
             ci_low = wald[rows, "log2FoldChange"] - qnorm(0.975) * wald[rows, "lfcSE"],
             ci_high = wald[rows, "log2FoldChange"] + qnorm(0.975) * wald[rows, "lfcSE"],
             lfc_apeglm = shr[rows, "log2FoldChange"], lfc_apeglm_sd = shr[rows, "lfcSE"],
             pvalue = wald[rows, "pvalue"], passes_count_filter = passes,
             outlier_replaced = if ("replace" %in% names(mcols(dds))) mcols(dds)[rows, "replace"] %in% TRUE else FALSE,
             row.names = NULL)
}
de <- bind_rows(lapply(seq_len(nrow(tc)), function(i) fit_mt(tc[i, ]))) %>%
  group_by(code) %>%
  mutate(padj_mt = ifelse(protein == "all_12", NA_real_,
                          p.adjust(ifelse(protein == "all_12", NA, pvalue), "BH"))) %>%
  ungroup() %>%
  mutate(significant = ifelse(protein == "all_12", pvalue < params$padj_cutoff, padj_mt < params$padj_cutoff))
## a point drawn as significant never has an interval that crosses zero
stopifnot(with(de, all(!significant | is.na(significant) | sign(ci_low) == sign(ci_high))))
write.csv(de, file.path(mito, "mt_de_TC.csv"), row.names = FALSE)
knitr::kable(de %>% select(code, protein, baseMean, log2FoldChange, lfcSE, lfc_apeglm, pvalue, padj_mt) %>%
               mutate(across(c(baseMean, log2FoldChange, lfcSE, lfc_apeglm), ~ round(.x, 2)),
                      across(c(pvalue, padj_mt), ~ signif(.x, 2))))
cat("Proteins with padj_mt <", params$padj_cutoff, "per contrast:\n")
print(de %>% filter(protein != "all_12") %>% group_by(code) %>%
        summarise(up = sum(significant & log2FoldChange > 0), down = sum(significant & log2FoldChange < 0),
                  .groups = "drop"))
```

# Sensitivity: haplotype covariate and permissive alignment

```{r sensitivity}
sens <- bind_rows(
  bind_rows(lapply(seq_len(nrow(tc)), function(i) fit_mt(tc[i, ], design = ~ haplotype + treatment, shrink = FALSE))) %>%
    mutate(setting = "haplotype covariate"),
  bind_rows(lapply(seq_len(nrow(tc)), function(i) fit_mt(tc[i, ], mt = mt_perm, shrink = FALSE))) %>%
    mutate(setting = "permissive alignment")) %>%
  group_by(setting, code) %>%
  mutate(padj_mt = ifelse(protein == "all_12", NA_real_, p.adjust(ifelse(protein == "all_12", NA, pvalue), "BH"))) %>%
  ungroup() %>%
  mutate(significant = ifelse(protein == "all_12", pvalue < params$padj_cutoff, padj_mt < params$padj_cutoff)) %>%
  select(setting, code, protein, log2FoldChange, lfcSE, pvalue, padj_mt, significant)
write.csv(sens, file.path(mito, "mt_de_TC_sensitivity.csv"), row.names = FALSE)
knitr::kable(bind_rows(de %>% mutate(setting = "of record"), sens) %>% filter(protein != "all_12") %>%
               group_by(setting, code) %>% summarise(significant = paste(protein[significant %in% TRUE], collapse = ", "), .groups = "drop") %>%
               pivot_wider(names_from = setting, values_from = significant),
             caption = "significant proteins per contrast and setting")
```

# Figure for the manuscript

```{r figure, fig.width = 11, fig.height = 8.5}
d <- de %>%
  mutate(protein = factor(ifelse(protein == "all_12", "All 12", protein), levels = c(PROTEINS, "All 12")),
         complex = factor(complex, levels = c("Complex I", "III", "Complex IV", "V", "All")),
         tissue = factor(tissue, levels = c("Foot", "Gill")),
         stressor = factor(stressor, levels = c("OA", "OW", "DO")),
         fill_key = ifelse(significant, as.character(stressor), "ns"))   # open point = not significant
dodge <- position_dodge(width = 0.6)
pa <- ggplot(d, aes(protein, log2FoldChange, colour = stressor, group = stressor)) +   # group: dodge points and bars together
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0, linewidth = 0.6, position = dodge) +
  geom_point(aes(fill = fill_key), shape = 21, size = 2.6, stroke = 0.9, position = dodge) +
  facet_grid(tissue ~ complex, scales = "free_x", space = "free_x") +
  scale_colour_manual(values = TREATMENT_COLORS, labels = TREATMENT_LABELS, name = NULL,
                      guide = guide_legend(override.aes = list(fill = TREATMENT_COLORS[c("OA", "OW", "DO")]))) +
  scale_fill_manual(values = c(TREATMENT_COLORS[c("OA", "OW", "DO")], ns = "white"), guide = "none") +
  labs(x = NULL, y = "log2 fold change vs day-3 control",
       title = "A. Mitochondrial protein genes, each stressor against the day-3 control") +
  theme_psmfc(10) + theme(legend.position = "top", panel.grid.major.x = element_blank())

pb <- ggplot(share %>% filter(day == "3") %>%
               mutate(treatment = factor(treatment, levels = names(TREATMENT_COLORS)),
                      tissue = factor(TISSUE_LABELS[tissue], levels = TISSUE_LABELS)),
             aes(treatment, mt_reads_pct_of_nuclear)) +
  geom_boxplot(aes(fill = treatment), alpha = 0.35, outlier.shape = NA, width = 0.6, colour = "grey30") +
  geom_point(aes(colour = treatment, shape = treatment), position = position_jitter(width = 0.12, seed = 1),
             size = 1.8) +
  facet_wrap(~ tissue, scales = "free_y") +
  scale_fill_manual(values = TREATMENT_COLORS, guide = "none") +
  scale_colour_manual(values = TREATMENT_COLORS, guide = "none") +
  scale_shape_manual(values = TREATMENT_SHAPES, guide = "none") +
  scale_x_discrete(labels = c(control = "Control", OA = "OA", OW = "OW", DO = "DO")) +
  labs(x = NULL, y = "mitochondrial protein reads\n(% of nuclear reads)",
       title = "B. Mitochondrial share of each library, day 3") +
  theme_psmfc(10)

p <- pa / pb + plot_layout(heights = c(1.6, 1)) +
  plot_annotation(caption = paste(strwrap(paste(
    "Each protein is counted on the mitochondrial genome alone (reads aligned to NC_007687.1); All 12 sums the",
    "proteins. A: DESeq2 log2 fold change (Wald, unshrunk) with 95% interval, from separate fits per contrast",
    "(nuclear-gene size factors); filled points: BH-adjusted Wald p < 0.05 over the 12 proteins (All 12: p < 0.05),",
    "open points: not significant. B: one point per library."), width = 130), collapse = "\n"),
    theme = theme(plot.caption = element_text(hjust = 0)))
ggsave(file.path(figs, "MT_mitochondrial_expression.png"), p, width = 11, height = 8.5, dpi = 300)
p
```

